### 2017-04-25 manuscript/msv1.tex
cold open

\documentclass[onecolumn,preprintnumbers,amsmath,amssymb,superscriptaddress]{revtex4}
%\usepackage[pdftex]{graphicx}

\usepackage{amsmath,amsfonts,amssymb}
\usepackage[english]{babel}
\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{color}
\usepackage{float}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{mathtools}
\usepackage{stmaryrd}
\usepackage{anyfontsize}

%\usepackage{epstopdf}
%\usepackage{array}
%\usepackage{tabularx}
%\usepackage{multirow}
\usepackage{color}
%\usepackage{multibox}
%\usepackage{rotating}
%\usepackage{lineno}
%\usepackage[left]{lineno}
%\usepackage[comma,sort&compress]{natbib}
%\usepackage{authblk}
%\usepackage{multicol}

%\bibliographystyle{ieeetr}

%\linenumbers
%\setlength\linenumbersep{3pt}

\begin{document}

%\title{Simple rules yield complex communities: deconstructed species interactions and the assembly of communities}
%\title{Community assembly and dynamics by the deconstruction of species interactions}
\title{Salmon straying description}
%\author{Justin D. Yeakel${}^{1,2,*}$, Christopher P. Kempes${}^{2}$, \& Sidney Redner${}^{2,3}$ \\ \\
%${}^1$School of Natural Science, University of California Merced, Merced, CA \\
%${}^2$The Santa Fe Institute, Santa Fe, NM \\
%${}^3$Department of Physics, Boston University, Boston MA \\
%${}^*$To whom correspondence should be addressed: jdyeakel@gmail.com
%}

\maketitle

\section*{Introduction}

\section*{Model Description}

Here we consider a local population $N_i$ and a remote population $N_j$, each with a trait value $x_i$ and $x_j$ that determines the recruitment rate, are assumed to be normally distributed with means $\mu_i$ and $\mu_j$ and a standard deviation $\sigma$, and are subject to selection across ecological timescales.
Each population in its respective site has an optimum trait value $\theta_i$ where the rate of recruitment is maximized if $x_i = \theta_i$.
If there is no straying between these populations (such that they are independent), then the trait value should evolve towards the optimum ($x_i \rightarrow \theta_i$) and the recruitment rate will become maximized.
If there is straying between populations at rate $m$, then the traits in each respective location will be pulled away from the optimum, and recruitment rates will be lower, i.e. have lower fitness.
As $m$ becomes large, the populations should act as a single population.

Using the discrete population framework from Shelton and Mangel (2011), we have the following population dynamic equation for a local population $i$ and the population that is straying into it $j$.

\begin{align}
  N_i(t+1) &= \left(N_i(t) - m N_i(t) + m N_j(t) \right){\rm e}^{-Z} \\ \nonumber
  &+ \left((R_i[\mu_i(t)|\theta_i]+\tilde{P}) (N_i(t) - m N_i(t))+ (R_i[\mu_j(t)|\theta_i]+\tilde{P}) m N_j(t)\right) {\rm e}^{-\beta (N_i(t) - m N_i(t)+ m N_j(t))}
\end{align}

\noindent where the first line determines the number of individuals that survive from time $t$ to time $t+1$, and the second and third line accounts for the recruitment of individuals from the surviving population from time $t$ to time $t+1$.
Moreover, there is a demographic process error term included in the above dynamic equation, where $\tilde{P}\sim \mathcal{N}(0,s)$ where in the below example $s=0.001$

Survival: the probability of survival is ${\rm e}^{-Z}$... if $Z=0.5$ (Shelton and Mangel 2011), the probability of survival is 60\%. 
This survival probability is enacted on 1) The number of local individuals that return minus the number local individuals that stray ($N_i(t) - mN_i(t)$), in addition to the number of remote individuals that immigrate into the local patch $mN_j(t)$, and it is assumed that they all have equal probability of surviving.

Recruitment: The second line of the equation determines the recruitment from local individuals $N_i(t)-mN_i(t)$ and incoming strays $mN_j(t)$, as a function of their mean trait value at time $t$ and given the local trait optimum $\theta_i$, where the recruitment rate at low population densities is
\begin{align}
  R_i[\mu_i(t)|\theta_i] &= \int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i(t)-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i(t)|\mu_i,\sigma^2) {\rm d}x_i(t) \\ \nonumber
  &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-\mu_i(t))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\}.
\end{align}
This means that it is the mismatch between the trait mean $\mu_i(t)$ and the local optimum $\theta_i$ that determines the  growth rate at low population densities.
% The compensatory effects are then determined by the exponential, following the Ricker stock-recruitment relationship.

The third line of the equation determines recruitment from straying individuals (with density $mN_j(t)$), such that the recruitment rate is determined by the mismatch between the optimum trait value $\theta_i$ and the mean trait value of the incoming individuals $\mu_j$.

As shown above, the mean traits of both populations (the local mean trait value $\mu_i$ as well as the remote mean trait value $\mu_j$) also change through time.
Because individuals from the local population are being mixed with individuals from the remote population, the resulting trait distribution is a mixed normal.
We assume that the distribution that results after the remote and local populations are mixed together \emph{and reproduce} is normal with a mean value determined from a mixed-normal distribution and constant variance over time.
This assumption may be tested a few different ways if we deem it necessary down the road - as a first order approximation, it's probably fine.
The mean trait value thus changes through time according to the difference equation

\begin{equation}
  \mu_i(t+1) = w_i\mu_i(t) + (1-w_i)\mu_j(t) + \frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)|\theta_i] + (1-w_i)R_i[\mu_j(t)|\theta_i]  \right)
  \label{eq:mu}
\end{equation}

\noindent where

\begin{equation}
  w_i=\frac{N_i(t)-m N_i(t)}{N_i(t)-m N_i(t) + m N_j(t)}.
\end{equation}

The first two factors in equation \ref{eq:mu} determine the mixed normal average of the now-mixed local and remote populations.
This mixed normal is weighted by the proportion of the population that is local and remote, respectively, which depends on the stray rate $m$.
The third part of the above equation determines how the mean trait changes through time due to natural selection (from Lande), which is proportional to the change in mean fitness with respect to the mean trait (phenotype).

You can see from the elements of the difference equations that demographics directly impact the evolution of trait means due to how the local and remote trait averages at time $t$ contribute to the trait mean at time $t+1$.

%Density dependent m

\section*{Results}

% Constant m
%% Fold bifurcation @ intermediate values of m
%% PE > 1 for low values of m
%% Spike in PE at the fold bifurcation (spike in indvidual population SD)
%% Increased environmental heterogeneity leads to spikes in PE at lower stray rates
%% Effects of heritability vs. trait variation and stable states

% Density dependent m
%%

{\bf Initial results -- subject to to change:} As the proportion of strays increases, the average traits of both populations veer farther from their optimal values, and this results in the decline of the population steady state (because the recruitment rate is now sub-optimal for both populations). This is the nice, expected result.

The unexpected result are the alternative stable states that arise for lower values of $m$. Just where these alternative stable states arise depends on the variance as well as the difference in optimal trait values between sites ($\theta_1-\theta_2$), but you can see that this has big impacts on the portfolio effect.
Which trajectory goes to the higher or lower alternative stable state is random - an effect of the process noise injected into the system.
The change in the PE appears to be caused by large spikes in the variance of individual population trajectories where the alternative stable states begin (at the fold bifurcation).
This is a well-known phenomenon that forms the basis for early warning signals of phase transitions.

\begin{figure}[h]
\centering
\includegraphics[width=0.4\textwidth]{fig_Density.pdf}
\caption{}
\end{figure}

The differences in optimal trait values between sites ($\theta_1-\theta_2$) influence at what straying rate the alternative stable states appear.
Large differences in $\theta_1$ and $\theta_2$ can be interpreted as high environmental heterogeneity.
A cursory analysis shows that increasing environmental heterogeneity tends to lower the stray rate value $m^*$ at which the maximum PE is expected to occur.
Moreover, if the trait variance is much larger than the difference in optimal trait values between sites (such that selection is weaker), the alternative stable state phenomenon disappears, as does the spike in the portfolio effect.

\begin{figure}[h]
\centering
\includegraphics[width=0.8\textwidth]{fig_HSDcomb.png}
\caption{}
\end{figure}

\section*{Discussion}

\end{document}

### 2017-04-25 model/KevinSimulations.jl

@@ -1,0 +1,829 @@
+using Distributions
+using RCall
+
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve.jl")
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinJacobian.jl")
+
+
+tmax=10000;
+z=0.5;
+rmax=2.0;
+beta=0.001;
+theta1=5.0;
+thetadiff=5.0;
+tau=1.0;
+h=1.0;
+sigmaE=0.25;
+sigmaG=0.25;
+m=0.27;
+perror=0.0000000;
+n1, n2, x1, x2, w1, w2 = 
+KevinEvolve(
+  tmax,
+  z,
+  rmax,
+  beta,
+  theta1,
+  thetadiff,
+  tau,
+  h,
+  sigmaE,
+  sigmaG,
+  m, #m
+  perror
+  );
+  
+  R"""
+  library(RColorBrewer)
+  cols = brewer.pal(3,'Set1')
+  plot($(n1),type="l",col=cols[1],ylim=c(0,max(c(as.vector($n1),as.vector($n2)))))
+  lines($(n2),col=cols[2])
+  """
+  
+  R"""
+  library(RColorBrewer)
+  cols = brewer.pal(3,'Set1')
+  plot($(x1),type="l",col=cols[1],ylim=c(4,12))
+  lines($(x2),col=cols[2])
+  """
+
+
+#CalculateJacobian
+Jac = KevinJacobian(last(n1),last(n2),last(x1),last(x2),
+z,rmax,beta,theta1,thetadiff,tau,sigma,m)
+eigvals(Jac)
+
+
+
+
+
+
+#Analysis over m
+tmax=10000;
+mvec = collect(0.0:0.001:0.5);
+n1ts = zeros(Float64,length(mvec),tmax);
+n2ts = zeros(Float64,length(mvec),tmax);
+n1mean=zeros(Float64,length(mvec));
+n2mean=zeros(Float64,length(mvec));
+n1sd=zeros(Float64,length(mvec));
+n2sd=zeros(Float64,length(mvec));
+aggmean=zeros(Float64,length(mvec));
+aggsd=zeros(Float64,length(mvec));
+x1mean=zeros(Float64,length(mvec));
+x2mean=zeros(Float64,length(mvec));
+pe=zeros(Float64,length(mvec));
+eigs = Array(Array{Complex{Float64}},length(mvec));
[…]

### 2017-04-25 model/KevinSimulations_ddm.jl

@@ -1,0 +1,835 @@
+using Distributions
+using RCall
+
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve_ddm.jl")
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinJacobian.jl")
+
+
+tmax=1000;
+z=0.5;
+rmax=2.0;
+beta=0.001;
+theta1=5.0;
+thetadiff=5.0;
+tau=1.0;
+h=1.0;
+a0 = 0.6;
+C=100;
+sigmaE=0.0;
+sigmaG=1;
+perror=0.05;
+
+n1, n2, x1, x2, w1, w2 = 
+KevinEvolve_ddm(
+  tmax,
+  z,
+  rmax,
+  beta,
+  theta1,
+  thetadiff,
+  tau,
+  h,
+  a0,
+  C,
+  sigmaE,
+  sigmaG,
+  perror
+  );
+  
+  R"""
+  library(RColorBrewer)
+  cols = brewer.pal(3,'Set1')
+  plot($(n1),type="l",col=cols[1],ylim=c(0,max(c(as.vector($n1),as.vector($n2)))))
+  lines($(n2),col=cols[2])
+  """
+  
+  R"""
+  library(RColorBrewer)
+  cols = brewer.pal(3,'Set1')
+  plot($(x1),type="l",col=cols[1],ylim=c(4,12))
+  lines($(x2),col=cols[2])
+  """
+
+
+#CalculateJacobian
+Jac = KevinJacobian(last(n1),last(n2),last(x1),last(x2),
+z,rmax,beta,theta1,thetadiff,tau,sigma,m)
+eigvals(Jac)
+
+
+
+
+
+
+#Analysis over a0
+tmax=10000;
+a0vec = collect(0.0:0.002:1.0);
+n1ts = zeros(Float64,length(a0vec),tmax);
+n2ts = zeros(Float64,length(a0vec),tmax);
+n1mean=zeros(Float64,length(a0vec));
+n2mean=zeros(Float64,length(a0vec));
+n1sd=zeros(Float64,length(a0vec));
+n2sd=zeros(Float64,length(a0vec));
+aggmean=zeros(Float64,length(a0vec));
+aggsd=zeros(Float64,length(a0vec));
+x1mean=zeros(Float64,length(a0vec));
+x2mean=zeros(Float64,length(a0vec));
+pe=zeros(Float64,length(a0vec));
[…]

### 2017-04-25 model/NormalPitchfork.jl

@@ -1,0 +1,32 @@
+using Distributions
+
+tmax=10000;
+
+muvec=collect(0:0.001:2);
+xss=Array{Float64}(length(muvec));
+
+pdist = Normal(0,0.0000000001);
+
+for i=1:length(muvec)
+  x = zeros(tmax);
+  x[1] = 0.1;
+  for t=1:tmax-1
+    x[t+1] = x[t]+ muvec[i]*x[t] - x[t]^3 + rand(pdist);
+  end
+  burnin=0.9;
+  xtrim = x[Int64(floor(burnin*tmax)):tmax];
+  xss[i]=mean(xtrim);
+end
+
+R"""
+plot($muvec,$xss,pch='.')
+"""
+
+
+
+x = zeros(tmax);
+x[1] = 0.1;
+for t=1:tmax-1
+  x[t+1] = x[t]+ 0.9*x[t] - x[t]^3 + rand(pdist);
+end
+R"plot($x,type='l',ylim=c(0,2))"

### 2017-04-25 model/src/KevinEvolve.jl

@@ -1,0 +1,78 @@
+function KevinEvolve(
+  tmax, 
+  z, 
+  rmax,
+  beta,
+  theta1,
+  thetadiff,
+  tau,
+  h,
+  sigmaE,
+  sigmaG,
+  m,
+  perror
+  )
+  
+  #Initialize variables
+  n1 = zeros(Float64,tmax); n1[1]=2;
+  n2 = zeros(Float64,tmax); n2[1]=2;
+  
+  x1 = zeros(Float64,tmax); x1[1]=theta1 + rand(Normal(0,0.0001));
+  x2 = zeros(Float64,tmax); x2[1]=(theta1 + thetadiff)  + rand(Normal(0,0.0001));
+  
+  w1 = zeros(Float64,tmax-1);
+  w2 = zeros(Float64,tmax-1);
+  meanfit1 = zeros(Float64,tmax-1);
+  meanfit2 = zeros(Float64,tmax-1);
+  
+  #Process Error Distribution
+  if perror == 0
+    pdist = [0,0];
+  else
+    pdist = Normal(0,perror);
+  end
+  
+  rfunc = function(mu,theta)
+    r = rmax*tau*(exp(-((theta-mu)^2)/(2*((sigmaE^2 + sigmaG^2)+tau^2))))/(sqrt(tau^2+(sigmaE^2 + sigmaG^2)))
+  end
+  
+  for t=1:tmax-1
+    
+    n1[t+1] = ((1-m)*n1[t] + m*n2[t])*exp(-z) +
+    ((rfunc(x1[t],theta1) + rand(pdist))*(n1[t]-m*n1[t]) +
+    (rfunc(x2[t],theta1) + rand(pdist))*m*n2[t]) * 
+    exp(-beta*((1-m)*n1[t]+m*n2[t]));
+    
+    n2[t+1] = ((1-m)*n2[t] + m*n1[t])*exp(-z) +
+    ((rfunc(x2[t],(theta1 + thetadiff)) + rand(pdist))*(n2[t]-m*n2[t]) +
+    (rfunc(x1[t],(theta1 + thetadiff)) + rand(pdist))*m*n1[t]) * 
+    exp(-beta*((1-m)*n2[t]+m*n1[t]));
+    
[…]

### 2017-04-25 model/src/KevinEvolve_ddm.jl

@@ -1,0 +1,86 @@
+function KevinEvolve_ddm(
+  tmax, 
+  z, 
+  rmax,
+  beta,
+  theta1,
+  thetadiff,
+  tau,
+  h,
+  a0,
+  C,
+  sigmaE,
+  sigmaG,
+  perror
+  )
+  
+  #Initialize variables
+  n1 = zeros(Float64,tmax); n1[1]=2;
+  n2 = zeros(Float64,tmax); n2[1]=2;
+  
+  m1 = zeros(Float64,tmax-1);
+  m2 = zeros(Float64,tmax-1);
+  
+  x1 = zeros(Float64,tmax); x1[1]=theta1 + rand(Normal(0,0.0001));
+  x2 = zeros(Float64,tmax); x2[1]=(theta1 + thetadiff)  + rand(Normal(0,0.0001));
+  
+  w1 = zeros(Float64,tmax-1);
+  w2 = zeros(Float64,tmax-1);
+  meanfit1 = zeros(Float64,tmax-1);
+  meanfit2 = zeros(Float64,tmax-1);
+  
+  #Process Error Distribution
+  if perror == 0
+    pdist = [0,0];
+  else
+    pdist = Normal(0,perror);
+  end
+  
+  rfunc = function(mu,theta)
+    r = rmax*tau*(exp(-((theta-mu)^2)/(2*((sigmaE^2 + sigmaG^2)+tau^2))))/(sqrt(tau^2+(sigmaE^2 + sigmaG^2)))
+  end
+  
+  for t=1:tmax-1
+    
+    #Density dependent m
+    m1[t] = 1 - (a0 + (1-a0)*(n1[t]/(C+n1[t])));
+    m2[t] = 1 - (a0 + (1-a0)*(n2[t]/(C+n2[t])));
+    
+    n1[t+1] = ((1-m1[t])*n1[t] + m2[t]*n2[t])*exp(-z) +
+    ((rfunc(x1[t],theta1) + rand(pdist))*(n1[t]-m1[t]*n1[t]) +
+    (rfunc(x2[t],theta1) + rand(pdist))*m2[t]*n2[t]) * 
+    exp(-beta*((1-m1[t])*n1[t]+m2[t]*n2[t]));
+    
[…]

### 2017-04-25 model/src/qualsfunc.jl

@@ -1,0 +1,24 @@
+function qualsfunc(ssval1, ssval2, extinct_threshold, similarity_threshold)
+  d1 = length(ssval1[:,1,1]);
+  d2 = length(ssval1[1,:,1]);
+  d3 = length(ssval1[1,1,:]);
+  
+  out = zeros(Int64,d1,d2,d3).-1;
+  # extinct_threshold = 5.0;
+  # similarity_threshold = 10.0;
+  
+  #both are extinct
+  out[find(x->x==true,(ssval1.<extinct_threshold) .* (ssval2.<extinct_threshold))] = 0;
+  
+  #One is extinct and one is not
+  #out[find(x->(x==0.0 || x==Inf),ssval1.<extinct_threshold ./ ssval2.<extinct_threshold)] = 1;
+  
+  #Trajectories are similar
+  out[find(x->x==true,(abs(ssval1.-ssval2).<similarity_threshold) .* (ssval1 .> extinct_threshold) .* (ssval2 .> extinct_threshold))] = 1;
+  
+  #Trajectories are in alternative stable states
+  out[find(x->x==true,(abs(ssval1.-ssval2) .> similarity_threshold) .* (ssval1 .> extinct_threshold) .* (ssval2 .> extinct_threshold))] = 2;
+  
+return out
+
+end

### 2017-04-26 manuscript/msv1.tex
MDPE fig for heterogeneous environs

@@
 
-Here we consider a local population $N_i$ and a remote population $N_j$, each with a trait value $x_i$ and $x_j$ that determines the recruitment rate, are assumed to be normally distributed with means $\mu_i$ and $\mu_j$ and a standard deviation $\sigma$, and are subject to selection across ecological timescales.
+Here we consider a local population $N_i$ and a remote population $N_j$ with a trait value $x_i$ and $x_j$ that determines the recrutiment rate for the respective population.
+We assume that there is an optimum trait value $\theta_i$ and $\theta_j$ associated with each site, where recruitment is maximized if $x = \theta$.
+
 Each population in its respective site has an optimum trait value $\theta_i$ where the rate of recruitment is maximized if $x_i = \theta_i$.
+
+determines the recruitment rate, are assumed to be normally distributed with means $\mu_i$ and $\mu_j$ and a standard deviation $\sigma$, and are subject to selection across ecological timescales.
+
 If there is no straying between these populations (such that they are independent), then the trait value should evolve towards the optimum ($x_i \rightarrow \theta_i$) and the recruitment rate will become maximized.
@@
 
+%The effects of trait evolution :: trait heritability, trait variance, and habitat heterogeneity
+%% When stray rates are moderate-low, greater heritability leads to higher steady-state populations with higher PE's.
+%% Lower heritability gives way to alternative stable states at lower stray rates, which become increasingly divergent
+%% With habitat heterogeneity?
+
 % Constant m
 %% Fold bifurcation @ intermediate values of m
-%% PE > 1 for low values of m
+%% This bifurcation creates a region of depressed total biomass & depressed PE at increased stray rates and lower trait variances
+
+%% PE >> 1 for low values of m
 %% Spike in PE at the fold bifurcation (spike in indvidual population SD)

### 2017-04-26 model/KevinSimulations.jl

@@ -744,6 +744,291 @@ dev.off()
 """
 
 
+#Portfolio muli-panel plot
+namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/fig_PEpanelmh.pdf");
+
+R"""
+library(RColorBrewer)
+pal = rev(brewer.pal(9,"Blues"))
+pdf($namespace,height=4,width=14)
+par(mfrow=c(1,6))
+  image(x=$mvec,y=$hvec,z=t($(pe[1,:,:])),zlim=c(1,2),col=pal,xlab='m',ylab='h',main=paste(c('sigma_G=',$(sigmavec[1]))))
+image(x=$mvec,y=$hvec,z=t($(pe[5,:,:])),zlim=c(1,2),col=pal,xlab='m',ylab='h',main=paste(c('sigma_G=',$(sigmavec[5]))))
+image(x=$mvec,y=$hvec,z=t($(pe[10,:,:])),zlim=c(1,2),col=pal,xlab='m',ylab='h',main=paste(c('sigma_G=',$(sigmavec[10]))))
+image(x=$mvec,y=$hvec,z=t($(pe[15,:,:])),zlim=c(1,2),col=pal,xlab='m',ylab='h',main=paste(c('sigma_G=',$(sigmavec[15]))))
+image(x=$mvec,y=$hvec,z=t($(pe[20,:,:])),zlim=c(1,2),col=pal,xlab='m',ylab='h',main=paste(c('sigma_G=',$(sigmavec[20]))))
+image(x=$mvec,y=$hvec,z=t($(pe[25,:,:])),zlim=c(1,2),col=pal,xlab='m',ylab='h',main=paste(c('sigma_G=',$(sigmavec[25]))))
+legend(x=max($mvec),y=max($hvec),legend=seq(1,2,length.out=9),col=pal,pch=22,xpd=TRUE,pt.bg=pal, bty="n")
+dev.off()
+"""
+
+
+#MEAN muli-panel plot
+namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/fig_MEANpanelmh.pdf");
+
+R"""
+library(RColorBrewer)
+pal = rev(brewer.pal(9,"Blues"))
+pdf($namespace,height=4,width=14)
+par(mfrow=c(1,6))
[…]

### 2017-04-26 model/src/KevinEvolve.jl
heritability fix

@@ -60,9 +60,9 @@ function KevinEvolve(
     ((((sigmaE^2 + sigmaG^2) + tau^2)^(3/2))*(((exp(-(((theta1 + thetadiff)-x1[t])^2)/(2*((sigmaE^2 + sigmaG^2)+tau^2)))*rmax*(1-w2[t])*tau)/
     (sqrt((sigmaE^2 + sigmaG^2)+tau^2))) +  ((exp(-(((theta1 + thetadiff)-x2[t])^2)/(2*((sigmaE^2 + sigmaG^2)+tau^2)))*rmax*w2[t]*tau)/(sqrt((sigmaE^2 + sigmaG^2)+tau^2)))));
     
-    x1[t+1] = w1[t]*x1[t] + (1-w1[t])*x2[t] + (h^2)*(( sigmaG^2))*meanfit1[t];
+    x1[t+1] = w1[t]*x1[t] + (1-w1[t])*x2[t] + (h)*(( sigmaG^2))*meanfit1[t];
     
-    x2[t+1] = w2[t]*x2[t] + (1-w2[t])*x1[t] + (h^2)*(( sigmaG^2))*meanfit2[t];
+    x2[t+1] = w2[t]*x2[t] + (1-w2[t])*x1[t] + (h)*(( sigmaG^2))*meanfit2[t];
     
   end
   

### 2017-04-26 model/src/KevinEvolve_ddm.jl

@@ -68,9 +68,9 @@ function KevinEvolve_ddm(
     ((((sigmaE^2 + sigmaG^2) + tau^2)^(3/2))*(((exp(-(((theta1 + thetadiff)-x1[t])^2)/(2*((sigmaE^2 + sigmaG^2)+tau^2)))*rmax*(1-w2[t])*tau)/
     (sqrt((sigmaE^2 + sigmaG^2)+tau^2))) +  ((exp(-(((theta1 + thetadiff)-x2[t])^2)/(2*((sigmaE^2 + sigmaG^2)+tau^2)))*rmax*w2[t]*tau)/(sqrt((sigmaE^2 + sigmaG^2)+tau^2)))));
     
-    x1[t+1] = w1[t]*x1[t] + (1-w1[t])*x2[t] + (h^2)*(( sigmaG^2))*meanfit1[t];
+    x1[t+1] = w1[t]*x1[t] + (1-w1[t])*x2[t] + (h)*(( sigmaG^2))*meanfit1[t];
     
-    x2[t+1] = w2[t]*x2[t] + (1-w2[t])*x1[t] + (h^2)*(( sigmaG^2))*meanfit2[t];
+    x2[t+1] = w2[t]*x2[t] + (1-w2[t])*x1[t] + (h)*(( sigmaG^2))*meanfit2[t];
     
   end
   

### 2017-05-02 manuscript/msv1.tex
heritability fix
mtheta

--- 2017-04-25 manuscript/msv1.tex
@@
 
-Here we consider a local population $N_i$ and a remote population $N_j$, each with a trait value $x_i$ and $x_j$ that determines the recruitment rate, are assumed to be normally distributed with means $\mu_i$ and $\mu_j$ and a standard deviation $\sigma$, and are subject to selection across ecological timescales.
-Each population in its respective site has an optimum trait value $\theta_i$ where the rate of recruitment is maximized if $x_i = \theta_i$.
-If there is no straying between these populations (such that they are independent), then the trait value should evolve towards the optimum ($x_i \rightarrow \theta_i$) and the recruitment rate will become maximized.
-If there is straying between populations at rate $m$, then the traits in each respective location will be pulled away from the optimum, and recruitment rates will be lower, i.e. have lower fitness.
-As $m$ becomes large, the populations should act as a single population.
+Here we consider two populations $N_1$ and $N_2$ that are separated in space, each with trait values $x_1$ and $x_2$ determining recruitment rates.
+We assume that there is an optimum trait value $\theta_1$ and $\theta_2$ associated with each habitat, where recruitment is maximized if the trait value of the local population $x = \theta$.
+Moreover, we assume that $x_{1,2}$ are normally distributed with means $\mu_1$ and $\mu_2$ and have a shared standard deviation $\sigma$.
+As such, the recruitment rate for both populations is determined by the mean trait value of the local population, such that $r_1 = R_1[\mu_1(t)|\theta_1]$.
+Trait means for each population are subject to selection, the strength of which depends on the difference between the population mean and the local trait optimum at a given point in time.
 
-Using the discrete population framework from Shelton and Mangel (2011), we have the following population dynamic equation for a local population $i$ and the population that is straying into it $j$.
+The two populations are assumed to reproduce in spatially separate sites that are close enough such that a proportion of the population $m$ can stray into the wrong site, and where mortality occurs before individuals return to spawn.
+If there is no straying between these populations (such that they are independent), then the mean trait evolves towards the optimal value such that $x_1 \rightarrow \theta_1$, and the recruitment rate for that population will be maximized.
+If there is straying between populations at rate $m$, then the traits in each respective location will be pulled away from the optimum, and recruitment rates will be lowered.
+As $m \rightarrow 0.5$, the populations are perfectly mixed, acting as a single population.
+
+We use the discrete Ricker population dynamic framework described by Shelton and Mangel \cite{} as the basis for our two-site model, with the added effect of the local population $N_i$ mixing with a set proportion $m$ of a remote population $N_j$ that is straying into it.
+We first assume that the proportion ${\rm e}^{-Z}$ of both populations survive, and that the aggregated mix of the populations (local individuals in addition to the straying individuals) are subject to the same compensatory effects, determined by the parameter $\beta$.
+For a local site $i\in(1,2)$ that collects straying individuals from a remote site $j\in(1,2)$, if $N_i$ is the local site and $N_j$ is the remote site, the difference equation that determine changes in population size is
 
 \begin{align}
-  N_i(t+1) &= \left(N_i(t) - m N_i(t) + m N_j(t) \right){\rm e}^{-Z} \\ \nonumber
-  &+ \left((R_i[\mu_i(t)|\theta_i]+\tilde{P}) (N_i(t) - m N_i(t))+ (R_i[\mu_j(t)|\theta_i]+\tilde{P}) m N_j(t)\right) {\rm e}^{-\beta (N_i(t) - m N_i(t)+ m N_j(t))}
+  N_i(t+1) &= \left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z} \\ \nonumber
+  &+ \left((R_i[\mu_i(t)|\theta_i]+\tilde{P}) (1-m)N_i(t) + (R_i[\mu_j(t)|\theta_i]+\tilde{P}) m N_j(t)\right) {\rm e}^{-\beta ((1-m)N_i(t) + m N_j(t))},
 \end{align}
 
-\noindent where the first line determines the number of individuals that survive from time $t$ to time $t+1$, and the second and third line accounts for the recruitment of individuals from the surviving population from time $t$ to time $t+1$.
-Moreover, there is a demographic process error term included in the above dynamic equation, where $\tilde{P}\sim \mathcal{N}(0,s)$ where in the below example $s=0.001$
+\noindent where we include demographic process error with $\tilde{P}\sim \mathcal{N}(0,s)$, where $s$ is assumed to be small, and where the difference equation for $N_j$ mirrors that for $N_i$.
 
-Survival: the probability of survival is ${\rm e}^{-Z}$... if $Z=0.5$ (Shelton and Mangel 2011), the probability of survival is 60\%. 
-This survival probability is enacted on 1) The number of local individuals that return minus the number local individuals that stray ($N_i(t) - mN_i(t)$), in addition to the number of remote individuals that immigrate into the local patch $mN_j(t)$, and it is assumed that they all have equal probability of surviving.
+The combined recruitment of local individuals $(1-m)N_i(t)$ and incoming strays $mN_j(t)$, as a function of their mean trait value at time $t$ and given the local trait optimum $\theta_i$, is then
 
-Recruitment: The second line of the equation determines the recruitment from local individuals $N_i(t)-mN_i(t)$ and incoming strays $mN_j(t)$, as a function of their mean trait value at time $t$ and given the local trait optimum $\theta_i$, where the recruitment rate at low population densities is
 \begin{align}
@@
 \end{align}
-This means that it is the mismatch between the trait mean $\mu_i(t)$ and the local optimum $\theta_i$ that determines the  growth rate at low population densities.
+
+\noindent As stated previously, it is the mismatch between the local trait mean $\mu_i(t)$ and the local optimum $\theta_i$ that determines the recruitment rate for the population.
+The parameter $\tau$ controls the sensitivity of recruitment to changes in the mean trait value away from the optimum, which we set as $\tau=1$ here and throughout.
 % The compensatory effects are then determined by the exponential, following the Ricker stock-recruitment relationship.
 
-The third line of the equation determines recruitment from straying individuals (with density $mN_j(t)$), such that the recruitment rate is determined by the mismatch between the optimum trait value $\theta_i$ and the mean trait value of the incoming individuals $\mu_j$.
+Because individuals from the local population are mixed with individuals from the remote population via staying, the resulting trait distribution is a mixed normal with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and for the straying individuals, $1-w_i$, where $w_i=(1-m)N_i(t)/((1-m) N_i(t) + m N_j(t))$.
+We make two simplifying assumptions.
+First, we assume that the distribution resulting from the mix of remote and local individuals, following reproduction, is also normal with a mean value being that of the mixed-normal.
+Second, we assume that changes in trait variance through time are minimal, such that $\sigma$ is assumed to be constant.
 
-As shown above, the mean traits of both populations (the local mean trait value $\mu_i$ as well as the remote mean trait value $\mu_j$) also change through time.
-Because individuals from the local population are being mixed with individuals from the remote population, the resulting trait distribution is a mixed normal.
-We assume that the distribution that results after the remote and local populations are mixed together \emph{and reproduce} is normal with a mean value determined from a mixed-normal distribution and constant variance over time.
-This assumption may be tested a few different ways if we deem it necessary down the road - as a first order approximation, it's probably fine.
+An increasing flow of incoming strays is thus expected to pull the mean trait value of the local population away from its optimum, which will decrease its rate of recruitment.
 The mean trait value thus changes through time according to the difference equation
@@
 \begin{equation}
-  \mu_i(t+1) = w_i\mu_i(t) + (1-w_i)\mu_j(t) + \frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)|\theta_i] + (1-w_i)R_i[\mu_j(t)|\theta_i]  \right)
+  \mu_i(t+1) = w_i\mu_i(t) + (1-w_i)\mu_j(t) + \frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)|\theta_i] + (1-w_i)R_i[\mu_j(t)|\theta_i]  \right),
   \label{eq:mu}
@@
 
-\noindent where
+\noindent where the first two factors determine the mixed normal average of the now-mixed local and remote populations.
+This mixed normal is weighted by the proportion of the population that is local and remote, respectively, which depends on the stray rate $m$.
+The partial derivative in the Eq.
+\ref{eq:mu} determines how the mean trait changes through time due to natural selection (REF), which is proportional to the change in mean fitness with respect to $\mu_i$.
+
+%Density dependent m
+We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is good evidence that at least in some species the stray rate is density dependent.
+Specifically, the rate at which individuals stray has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tends to decrease the rate at which individuals stray, thus reducing the overall proportion of a population that strays.
+According to REF, given the probability that an individual strays $m_0$, the proportion of the local population $N_i(t)$ that strays is
 
 \begin{equation}
-  w_i=\frac{N_i(t)-m N_i(t)}{N_i(t)-m N_i(t) + m N_j(t)}.
+  m(t) = m_0 - m_0\left(\frac{N_i(t)}{C+N_i(t)}\right)
 \end{equation}
 
-The first two factors in equation \ref{eq:mu} determine the mixed normal average of the now-mixed local and remote populations.
-This mixed normal is weighted by the proportion of the population that is local and remote, respectively, which depends on the stray rate $m$.
-The third part of the above equation determines how the mean trait changes through time due to natural selection (from Lande), which is proportional to the change in mean fitness with respect to the mean trait (phenotype).
+\noindent where $C$ is the half-saturation value of $N_i$ where the density-dependent stray rate decreases sharply.
+We note that at the limit $C\rightarrow \infty$, the density dependent stray rate becomes constant such that $m(t) \rightarrow m_0$, and this corresponds to the original formulation where $m=m_0$.
+A similar observation shows that when the population density is very high, $m(t) \rightarrow 0$, and when it is close to extinction, $m(t) \rightarrow m_0$.
+Thus, for realistic population densities, $m(t) < m_0$.
 
-You can see from the elements of the difference equations that demographics directly impact the evolution of trait means due to how the local and remote trait averages at time $t$ contribute to the trait mean at time $t+1$.
-
-%Density dependent m
+%Linking m/m0 and thetadiff
 
@@
 
-% Constant m
-%% Fold bifurcation @ intermediate values of m
-%% PE > 1 for low values of m
+%The effects of trait evolution :: trait heritability, trait variance, and habitat heterogeneity
+%% Bifurcation to alternative stable state occurs at higher stray rates with increased trait heritability
+%% There is a peak in the portfolio effect directly before this bifurcation is crossed (early warning signal)
+%% When trait heritability is high, there is a greater sensitivity of total biomass to increasing stray rates, though a LOWER sensitivity of the difference between alternative steady state values for increasing stray rates (and vice versa)
+%% For low heritability (h = 0.01 - 0.2 or so), moderate and insensitive changes in total biomass, but large differences between the values of alternative steady states leads to a local maximum for PE at low-moderate stray rates
+%% Increased habitat heterogeneity
+  %%% 1) exaggerates lowering of mean biomass with increasing stray rates
+  %%% 2) increases the likelihood of alternative stable sites & increases the differences in alternative stable state values with increasing stray rates
+  %%% 3) decreases the PE (!!)
+
+%% PE >> 1 for low values of m
 %% Spike in PE at the fold bifurcation (spike in indvidual population SD)
@@
 The unexpected result are the alternative stable states that arise for lower values of $m$.
-Just where these alternative stable states arise depends on the variance as well as the difference in optimal trait values between sites ($\theta_1-\theta_2$), but you can see that this has big impacts on the portfolio effect.
+Just where these alternative stable states arise depends on the variance as well as the difference in optimal trait values between sites ($\theta_i-\theta_j$), but you can see that this has big impacts on the portfolio effect.
 Which trajectory goes to the higher or lower alternative stable state is random - an effect of the process noise injected into the system.
@@
 \centering
-\includegraphics[width=0.4\textwidth]{fig_Density.pdf}
+\includegraphics[width=0.4\textwidth]{figs/fig_Density.pdf}
 \caption{}
@@
 
-The differences in optimal trait values between sites ($\theta_1-\theta_2$) influence at what straying rate the alternative stable states appear.
-Large differences in $\theta_1$ and $\theta_2$ can be interpreted as high environmental heterogeneity.
+The differences in optimal trait values between sites ($\theta_i-\theta_j$) influence at what straying rate the alternative stable states appear.
+Large differences in $\theta_i$ and $\theta_j$ can be interpreted as high environmental heterogeneity.
 A cursory analysis shows that increasing environmental heterogeneity tends to lower the stray rate value $m^*$ at which the maximum PE is expected to occur.
@@
 \centering
-\includegraphics[width=0.8\textwidth]{fig_HSDcomb.png}
+\includegraphics[width=0.8\textwidth]{figs/fig_HSDcomb.png}
 \caption{}

### 2017-05-02 model/KevinSimulations_mtheta.jl

@@ -1,0 +1,553 @@
+using Distributions
+using RCall
+
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve_mtheta.jl")
+
+#Analysis over m
+tmax=10000;
+mvec = collect(0.0:0.001:0.5);
+n1ts = zeros(Float64,length(mvec),tmax);
+n2ts = zeros(Float64,length(mvec),tmax);
+n1mean=zeros(Float64,length(mvec));
+n2mean=zeros(Float64,length(mvec));
+n1sd=zeros(Float64,length(mvec));
+n2sd=zeros(Float64,length(mvec));
+aggmean=zeros(Float64,length(mvec));
+aggsd=zeros(Float64,length(mvec));
+x1mean=zeros(Float64,length(mvec));
+x2mean=zeros(Float64,length(mvec));
+pe=zeros(Float64,length(mvec));
+eigs = Array(Array{Complex{Float64}},length(mvec));
+maxeigs = Array{Float64}(length(mvec));
+maximeigs = Array{Float64}(length(mvec));
+
+z=0.5;
+rmax=2.0;
+beta=0.001;
+theta1=5.0;
+thetascale=2;
+tau=1.0;
+h=0.2;
+sigmaE=0;
+sigmaG=1;
+perror=0.01;
+
+burnin=0.99
+@time for i=1:length(mvec)
+  m=mvec[i];
+  
+  n1, n2, x1, x2, w1, w2, thetadiff = 
+  KevinEvolve_mtheta(
+    tmax,
+    z,
+    rmax,
+    beta,
+    theta1,
+    thetascale,
+    tau,
+    h,
+    sigmaE,
+    sigmaG,
+    m,
+    perror
+    );
+
+  n1trim = n1[Int64(floor(tmax*burnin)):tmax];
+  n2trim = n2[Int64(floor(tmax*burnin)):tmax];
+  x1trim = x1[Int64(floor(tmax*burnin)):tmax];
+  x2trim = x2[Int64(floor(tmax*burnin)):tmax]
+  
+  n1ts[i,:] = n1;
+  n2ts[i,:] = n2;
+  
+  n1mean[i] = mean(n1trim);
+  n2mean[i] = mean(n2trim);
+  n1sd[i] = std(n1trim);
+  n2sd[i] = std(n2trim);
+  
[…]

### 2017-05-02 model/src/KevinEvolve_mtheta_ddm.jl

@@ -1,0 +1,89 @@
+function KevinEvolve_mtheta_ddm(
+  tmax, 
+  z, 
+  rmax,
+  beta,
+  theta1,
+  thetascale,
+  tau,
+  h,
+  a0,
+  C,
+  sigmaE,
+  sigmaG,
+  perror
+  )
+  
+  #Initialize variables
+  thetadiff = (1-2*m)/(thetascale*m);
+  
+  n1 = zeros(Float64,tmax); n1[1]=2;
+  n2 = zeros(Float64,tmax); n2[1]=2;
+  
+  m1 = zeros(Float64,tmax-1);
+  m2 = zeros(Float64,tmax-1);
+  
+  x1 = zeros(Float64,tmax); x1[1]=theta1 + rand(Normal(0,0.0001));
+  x2 = zeros(Float64,tmax); x2[1]=(theta1 + thetadiff)  + rand(Normal(0,0.0001));
+  
+  w1 = zeros(Float64,tmax-1);
+  w2 = zeros(Float64,tmax-1);
+  meanfit1 = zeros(Float64,tmax-1);
+  meanfit2 = zeros(Float64,tmax-1);
+  
+  #Process Error Distribution
+  if perror == 0
+    pdist = [0,0];
+  else
+    pdist = Normal(0,perror);
+  end
+  
+  rfunc = function(mu,theta)
+    r = rmax*tau*(exp(-((theta-mu)^2)/(2*((sigmaE^2 + sigmaG^2)+tau^2))))/(sqrt(tau^2+(sigmaE^2 + sigmaG^2)))
+  end
+  
+  for t=1:tmax-1
+    
+    #Density dependent m
+    m1[t] = 1 - (a0 + (1-a0)*(n1[t]/(C+n1[t])));
+    m2[t] = 1 - (a0 + (1-a0)*(n2[t]/(C+n2[t])));
+    
+    n1[t+1] = ((1-m1[t])*n1[t] + m2[t]*n2[t])*exp(-z) +
+    ((rfunc(x1[t],theta1) + rand(pdist))*(n1[t]-m1[t]*n1[t]) +
+    (rfunc(x2[t],theta1) + rand(pdist))*m2[t]*n2[t]) * 
[…]

### 2017-05-03 manuscript/msv1.tex
mtheta ddm runs

@@
 \usepackage{anyfontsize}
+
+%\usepackage{caption}
+\usepackage{subcaption}
+\usepackage{graphicx}
 
@@
   &+ \left((R_i[\mu_i(t)|\theta_i]+\tilde{P}) (1-m)N_i(t) + (R_i[\mu_j(t)|\theta_i]+\tilde{P}) m N_j(t)\right) {\rm e}^{-\beta ((1-m)N_i(t) + m N_j(t))},
+  \label{eq:N}
 \end{align}
@@
   &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-\mu_i(t))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\}.
+  \label{eq:R}
 \end{align}
@@
 \begin{equation}
-  m(t) = m_0 - m_0\left(\frac{N_i(t)}{C+N_i(t)}\right)
+  m(t) = m_0\left(1- \frac{N_i(t)}{C+N_i(t)}\right),
+  \label{eq:ddm}
 \end{equation}
@@
 %Linking m/m0 and thetadiff
+Stray rate is intrinsically linked to the distance between the local and straying population.
+The greater the distance between two populations, the lower the expected rate of straying (REF).
+We can account for this interdependence in our model by assuming that $m$ (if the stray rate is constant) or $m_0$ (if the stray rate is density dependent) is a function of $theta_i-theta_j$, which can be assumed to be large if the remote site $j$ is a great distance away from the local site $i$.
+If sites $i$ and $j$ are very close, the stray rate is assumed to maximized at $m,m_0 = 0.5$.
+Thus, we can integrate these two variables by setting $m,m_0 = (2 + \epsilon (\theta_i-\theta_j))^{-1}$, where $\epsilon$ sets the sensitivity of a declining $m$ to increasing distance (greater values of $\theta_i-\theta_j$).
 
@@
 %% Bifurcation to alternative stable state occurs at higher stray rates with increased trait heritability
-%% There is a peak in the portfolio effect directly before this bifurcation is crossed (early warning signal)
+%% There is a peak in the portfolio effect directly before this bifurcation is crossed (early warning signal), and then a decline in PE with high m
 %% When trait heritability is high, there is a greater sensitivity of total biomass to increasing stray rates, though a LOWER sensitivity of the difference between alternative steady state values for increasing stray rates (and vice versa)
@@
   %%% 3) decreases the PE (!!)
+
+%Alternative stable states
+Straying generally lowers steady state densities for both local and remote populations. %, regardless of trait heritability and variance or habitat heterogeneity
+The decline in steady state densities is not gradual: as straying increases, the system crosses a fold bifurcation whereby the single steady state splits and becomes two alternative steady states: one at high biomass density, and one at low biomass density (Fig.
+\ref{}).
+This bifurcation occurs at lower values of the stray rate $m$ for lower levels of trait heritability $h^2$, indicating that tighter coupling between ecological and evolutionary dynamics  requires higher rates of straying to give rise to alternative stable states.
+
+Trait heritability also has a large impact on the sensitivity of both the total steady state density ($N^*_T=N^*_1+N^*_2$) as well as the difference between steady state densities ($\Delta N=\sqrt{(N^*_1-N^*_2)^2}$).
+Greater trait heritability (high $h^2$) results in a larger decline in $N_T^*$ for increasing stray rates $m$, though a moderate change in $\Delta N$ with increasing stray rates.
+Conversely, if the trait has a low heritability, an increase in the stray rate has little impact on the total biomass density, though very large impacts on the difference in population densities between sites.
+
+Together these differences in steady state population densities as a function of trait heritability and the rate of straying between populations results in nonlinear changes to the portfolio effect, quantified as
+
+\begin{equation}
+  {\rm PE} = \frac{\overline{{\rm CV}}_{i,j}}{{\rm CV}_T}.
+  \label{eq:pe}
+\end{equation}
+
+\noindent We define $\overline{{\rm CV}}_{i,j}$ as the average coefficient of variation across the local population $i$ and the remote population $j$, and ${\rm CV}_T$ as the coefficient of variation of the total or aggregated population.
+Higher ${\rm PE}$ corresponds to an increased potential for ecological rescue between populations, thus buffering the system as a whole from extinction, whereas the lowest possible value is ${\rm PE_min}=1$
+
+In the region where there is a single steady state value there is a correspondingly high portfolio effect.
+As the fold bifurcation is approached with greater rates of straying, the portfolio effect spikes due to a large increase in the standard deviation of individual trajectories.
+This explosion in variance is a well-known phenomenon that occurs near a fold bifurcation and lies at the heart of early warning signal research (REFS).
+Directly to the right of the fold bifurcation (in Fig \ref{fig:PE}C) where there are two alternative steady state densities, the portfolio effect is minimized if heritability is low, and it remains elevated if heritability is high.
+As the stray rate increases to the right of the bifurcation, increased trait heritability results in a steady decrease in the portfolio effect.
+If trait hertitability is low, the portfolio effect changes nonlinearly: after the spike associated with the fold bifurcation, {\rm PE} first increases at intermediate stray rates and then declines as $m\rightarrow0.5$.
+
+% Density dependent m
+%% Because m[t] @ steady state 0<m[t]<m0, we observe the same dynamics for higher m0 as we do for lower constant m.
+%% Higher rates of straying are tolerable (there is a not a decline in PE for very high m0 as there is for constant m)
+
+% When m and thetadiff are linked
+%%Constant m
+  %%% Low stray rate (high theta1-theta2) leads to alternative stable states and relatively low PE ~ sharp transition
+  %%% High stray rate (low theta1-theta2) increases the steady state (and no A.S.S.)... 
+  %%% This suggests that space naturally buffers against the negative effects of mixing dissimilar populations, however it is NOT SMOOTH
+  %%% This is particularly true for low heritability. High heritability eliminates the A.S.S. region
+%% Similar results for DDm
 
@@
 \includegraphics[width=0.4\textwidth]{figs/fig_Density.pdf}
-\caption{}
+\caption{} \label{fig:traj}
 \end{figure}
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figs/fig_HSDcomb.png}
-\caption{}
+\includegraphics[width=0.8\textwidth]{figs/fig_MDPE_hm.pdf}
+\caption{} \label{fig:PE}
+\end{figure}
+
+\begin{figure}
+\centering
+\begin{subfigure}[t]{0.8\textwidth}
+\centering
+\includegraphics[width=\textwidth]{figs/fig_MDPE_hm_theta3.pdf} 
+\caption{Low habitat heterogeneity ($\Delta\theta=3$)} \label{fig:hetero1}
+\end{subfigure}
+
+\begin{subfigure}[t]{0.8\textwidth}
+\centering
+\includegraphics[width=\textwidth]{figs/fig_MDPE_hm_theta8.pdf} 
+\caption{High habitat heterogeneity ($\Delta\theta=8$)} \label{fig:hetero2}
+\end{subfigure}
+
+ \caption{Changes in total means, difference in means, and the portfolio effect for different habitat heterogeneities}
+
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.5\textwidth]{figs/fig_MDPE_hm_cross358.pdf}
+\caption{} \label{fig:PE}
 \end{figure}

### 2017-05-03 model/KevinSimulations_ddm.jl

@@ -65,37 +65,37 @@ eigvals(Jac)
 
 #Analysis over a0
 tmax=10000;
-a0vec = collect(0.0:0.002:1.0);
-n1ts = zeros(Float64,length(a0vec),tmax);
-n2ts = zeros(Float64,length(a0vec),tmax);
-n1mean=zeros(Float64,length(a0vec));
-n2mean=zeros(Float64,length(a0vec));
-n1sd=zeros(Float64,length(a0vec));
-n2sd=zeros(Float64,length(a0vec));
-aggmean=zeros(Float64,length(a0vec));
-aggsd=zeros(Float64,length(a0vec));
-x1mean=zeros(Float64,length(a0vec));
-x2mean=zeros(Float64,length(a0vec));
-pe=zeros(Float64,length(a0vec));
-eigs = Array(Array{Complex{Float64}},length(a0vec));
-maxeigs = Array{Float64}(length(a0vec));
-maximeigs = Array{Float64}(length(a0vec));
+indmvec = collect(0.0:0.002:0.5);
+n1ts = zeros(Float64,length(indmvec),tmax);
+n2ts = zeros(Float64,length(indmvec),tmax);
+n1mean=zeros(Float64,length(indmvec));
+n2mean=zeros(Float64,length(indmvec));
+n1sd=zeros(Float64,length(indmvec));
+n2sd=zeros(Float64,length(indmvec));
+aggmean=zeros(Float64,length(indmvec));
+aggsd=zeros(Float64,length(indmvec));
+x1mean=zeros(Float64,length(indmvec));
+x2mean=zeros(Float64,length(indmvec));
+pe=zeros(Float64,length(indmvec));
+eigs = Array(Array{Complex{Float64}},length(indmvec));
+maxeigs = Array{Float64}(length(indmvec));
+maximeigs = Array{Float64}(length(indmvec));
 
 z=0.5;
 rmax=2.0;
 beta=0.001;
 theta1=5.0;
-thetadiff=8;
+thetadiff=5;
 tau=1.0;
 h=0.2;
-C=1000;
+C=100;
 sigmaE=0;
-sigmaG=2;
-perror=0.05;
+sigmaG=1;
+perror=0.01;
 
-burnin=0.99
-@time for i=1:length(a0vec)
-  a0 = a0vec[i];
+burnin=0.80
+@time for i=1:length(indmvec)
+  a0 = 1-indmvec[i];
[…]

### 2017-05-03 model/src/KevinEvolve_mtheta_ddm.jl

@@ -14,8 +14,9 @@ function KevinEvolve_mtheta_ddm(
   perror
   )
   
+  m0 = 1-a0;
   #Initialize variables
-  thetadiff = (1-2*m)/(thetascale*m);
+  thetadiff = (1-2*m0)/(thetascale*m0);
   
   n1 = zeros(Float64,tmax); n1[1]=2;
   n2 = zeros(Float64,tmax); n2[1]=2;

### 2017-05-09 manuscript/msv1.tex
figure and ms update

--- 2017-05-02 manuscript/msv1.tex
@@
-\documentclass[onecolumn,preprintnumbers,amsmath,amssymb,superscriptaddress]{revtex4}
+\documentclass[twocolumn,preprintnumbers,amsmath,amssymb,superscriptaddress]{revtex4}
 %\usepackage[pdftex]{graphicx}
@@
 \usepackage{anyfontsize}
+
+\usepackage{caption}
+\usepackage{subcaption}
+\captionsetup{compatibility=false}
 
@@
 %\title{Community assembly and dynamics by the deconstruction of species interactions}
-\title{Salmon straying description}
+\title{Eco-evolutionary model of salmon straying in a metapopulation (but with a cool title)}
 %\author{Justin D. Yeakel${}^{1,2,*}$, Christopher P. Kempes${}^{2}$, \& Sidney Redner${}^{2,3}$ \\ \\
@@
 \section*{Introduction}
+
+Salmony salmons
 
@@
 
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.4\textwidth]{figs/fig_Density.pdf}
+\caption{
+A) The steady state densities of $N_1$ and $N_2$ as a function of a constant stray rate $m$.
+B) The steady state trait values measured as $theta_i - x_i$, as a function of a constant stray rate $m$. 
+} \label{fig:traj}
+\end{figure}
+
 We use the discrete Ricker population dynamic framework described by Shelton and Mangel \cite{} as the basis for our two-site model, with the added effect of the local population $N_i$ mixing with a set proportion $m$ of a remote population $N_j$ that is straying into it.
@@
 \begin{align}
-  N_i(t+1) &= \left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z} \\ \nonumber
-  &+ \left((R_i[\mu_i(t)|\theta_i]+\tilde{P}) (1-m)N_i(t) + (R_i[\mu_j(t)|\theta_i]+\tilde{P}) m N_j(t)\right) {\rm e}^{-\beta ((1-m)N_i(t) + m N_j(t))},
+  &N_i(t+1) = \\ \nonumber
+  &\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z} \\ \nonumber
+  &+ \left(R_i[\mu_i(t)|\theta_i] (1-m)N_i(t) + R_i[\mu_j(t)|\theta_i] m N_j(t)\right) \\ \nonumber
+  &\times {\rm e}^{-\beta ((1-m)N_i(t) + m N_j(t))},
+  \label{eq:N}
 \end{align}
 
-\noindent where we include demographic process error with $\tilde{P}\sim \mathcal{N}(0,s)$, where $s$ is assumed to be small, and where the difference equation for $N_j$ mirrors that for $N_i$.
+\noindent where a small amount of demographic process error is added to the reproductive rate, and where the difference equation for $N_j$ mirrors that for $N_i$.
 
@@
 \begin{align}
-  R_i[\mu_i(t)|\theta_i] &= \int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i(t)-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i(t)|\mu_i,\sigma^2) {\rm d}x_i(t) \\ \nonumber
+  &R_i[\mu_i(t)|\theta_i] = \\ \nonumber
+  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i(t)-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i(t)|\mu_i,\sigma^2) {\rm d}x_i(t) \\ \nonumber
   &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-\mu_i(t))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\}.
+  \label{eq:R}
 \end{align}
@@
 
-Because individuals from the local population are mixed with individuals from the remote population via staying, the resulting trait distribution is a mixed normal with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and for the straying individuals, $1-w_i$, where $w_i=(1-m)N_i(t)/((1-m) N_i(t) + m N_j(t))$.
+\begin{figure*}
+\centering
+\includegraphics[width=0.8\textwidth]{figs/fig_MDPE_hm.pdf}
+\caption{
+Total means $N_t$, difference in means $\Delta N$, and the portfolio effect PE as a function of heritability $h^2$ and a constant stray rate $m$. Light colors = high values.
+} \label{fig:PE}
+\end{figure*}
+
+Because individuals from the local population are mixed with individuals from the remote population via staying, the resulting trait distribution is a mixed normal with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and for the straying individuals, $1-w_i$, where
+\begin{equation}
+w_i=\frac{(1-m)N_i(t)}{(1-m) N_i(t) + m N_j(t)}.
+\end{equation}
 We make two simplifying assumptions.
@@
 
-\begin{equation}
-  \mu_i(t+1) = w_i\mu_i(t) + (1-w_i)\mu_j(t) + \frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)|\theta_i] + (1-w_i)R_i[\mu_j(t)|\theta_i]  \right),
+\begin{align}
+  \mu_i(t+1) &= w_i\mu_i(t) + (1-w_i)\mu_j(t) \\ \nonumber
+  &+ \frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)|\theta_i] + (1-w_i)R_i[\mu_j(t)|\theta_i]  \right),
   \label{eq:mu}
-\end{equation}
+\end{align}
 
@@
 %% Bifurcation to alternative stable state occurs at higher stray rates with increased trait heritability
-%% There is a peak in the portfolio effect directly before this bifurcation is crossed (early warning signal)
+%% There is a peak in the portfolio effect directly before this bifurcation is crossed (early warning signal), and then a decline in PE with high m
 %% When trait heritability is high, there is a greater sensitivity of total biomass to increasing stray rates, though a LOWER sensitivity of the difference between alternative steady state values for increasing stray rates (and vice versa)
 %% For low heritability (h = 0.01 - 0.2 or so), moderate and insensitive changes in total biomass, but large differences between the values of alternative steady states leads to a local maximum for PE at low-moderate stray rates
+
+%Alternative stable states
+{\bf Nonlinear effects of straying on the Portfolio Effect} Straying generally lowers steady state densities for both local and remote populations.
+%, regardless of trait heritability and variance or habitat heterogeneity
+The decline in steady state densities is not gradual: as straying increases, the system crosses a fold bifurcation whereby the single steady state among both sites becomes two alternative steady states: one at high biomass density, and one at low biomass density (Fig.
+\ref{fig:traj}A,B).
+This bifurcation occurs at lower values of the stray rate $m$ for low trait heritability $h^2$ (Fig \ref{fig:PE}A), indicating that greater coupling between ecological and evolutionary dynamics in addition to greater rates of straying results in alternative stable states among the two sites.
+
+Trait heritability also has a large impact on the sensitivity of both the total steady state density ($N^*_T=N^*_1+N^*_2$) as well as the difference between steady state densities ($\Delta N=\sqrt{(N^*_1-N^*_2)^2}$).
+Greater trait heritability results in a larger decline in $N_T^*$ for increasing stray rates $m$, though results in only moderate changes to $\Delta N$ (Figs. \ref{fig:PE}A,B).
+Conversely, if the trait is less heritable, an increase in the stray rate has little impact on the total biomass density and contrastingly large effects on the difference in population densities between sites (Fig.
+\ref{fig:PE}A,B).
+
+\begin{figure}
+\centering
+\includegraphics[width=0.4\textwidth]{figs/fig_thetaPE.pdf}
+\caption{
+Median portfolio effect as a function of stray rate $m$ for lower trait heritability ($h^2 < 0.5$)
+} \label{fig:thetaPE}
+\end{figure}
+
+Together these changes in steady state population densities in terms of $N_T$ and $\Delta N$ as a function of trait heritability and the rate of straying between populations give rise to highly nonlinear portfolio effects, quantified here as
+
+\begin{equation}
+  {\rm PE} = \frac{\overline{{\rm CV}}_{i,j}}{{\rm CV}_T}.
+  \label{eq:pe}
+\end{equation}
+
+\noindent We define $\overline{{\rm CV}}_{i,j}$ as the average coefficient of variation across the local population $i$ and the remote population $j$ at steady state densities, and ${\rm CV}_T$ as the coefficient of variation of the total or aggregated population at steady state densities.
+The minimum portfolio effect is by definition ${\rm PE_min}=1$, whereas portfolio effects greater than unity corresponds to a greater potential for ecological rescue among populations, thus buffering the system as a whole against extinction.
+
+In the region where there is a single steady state among both populations, we find a correspondingly high portfolio effect, primarily due to the elevated mean values of $N_T$.
+As the fold bifurcation is approached with greater rates of straying, the portfolio effect spikes due to a large increase in the standard deviation of both populations.
+This explosion in variance is a well-known phenomenon that occurs near a fold bifurcation and lies at the heart of early warning signal theory (REFS).
+For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}C), where there is a high and low steady state density among the sites, if heritability is low the portfolio effect becomes minimized, is elevated and then declines as $m\rightarrow 0.5$.
+If heritability is high, the portfolio effect declines steadily after the explosion in variance associated with the development of alternative stable states.
+% As the stray rate increases to the right of the bifurcation, increased trait heritability results in a steady decrease in the portfolio effect.
+% If trait heritability is low, the portfolio effect is nonlinear over increasing stray rates: after the spike associated with the fold bifurcation, {\rm PE} first increases at intermediate stray rates and then declines as $m\rightarrow0.5$.
+
+% \begin{figure*}[htp]
+% \centering
+% \begin{subfigure}[t]{0.55\textwidth}
+% \centering
+% \includegraphics[width=\textwidth]{figs/fig_MDPE_hm_theta3.pdf} 
+% \caption{Low habitat heterogeneity ($\Delta\theta=3$)} \label{fig:thetadiff1}
+% \end{subfigure}
+% \begin{subfigure}[t]{0.55\textwidth}
+% \centering
+% \includegraphics[width=\textwidth]{figs/fig_MDPE_hm_theta8.pdf} 
+% \caption{High habitat heterogeneity ($\Delta\theta=8$)} \label{fig:thetadiff2}
+% \end{subfigure}
+% \caption{Total means $N_t$, difference in means $\Delta N$, and the portfolio effect PE for different habitat heterogeneities $\Delta\theta$. Light colors = high values.
+% }
+% \end{figure*}
+
+% Density dependent m
+%% Because m[t] @ steady state 0<m[t]<m0, we observe the same dynamics for higher m0 as we do for lower constant m.
+%% Higher rates of straying are tolerable (there is a not a decline in PE for very high m0 as there is for constant m)
+If we assume that the rate of straying is density dependent, the probability of straying at the individual level $m_0$ determines the rate of straying within the population, such that $m(t)$ becomes lower as $N(t)$ increases.
+At steady state values, $m(t)$ also becomes constant, though by definition is always less than $m_0$, such that $0 < m(t) < m_0$.
+We find that this dynamic has negligible impact on the qualitative results on our system (Fig. \ref{fig:diffddm}, Appendix Fig. \ref{fig:ddm}).
+Quantitatively, density dependent straying serves to lessen the effect of straying on $N_T$, $\Delta N$ and PE, given that the effective stray rate is effectively lessened by collective navigation.
+%Because density dependent straying does not alter that qualitative results of our model, we limit additional analyses
+
+\begin{figure}
+\centering
+\includegraphics[width=0.4\textwidth]{figs/fig_diffddm.pdf}
+\caption{
+.
+} \label{fig:diffddm}
+\end{figure}
+
 %% Increased habitat heterogeneity
@@
   %%% 3) decreases the PE (!!)
+{\bf The role of habitat heterogeneity} Increased differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater between-site differences in conditions that favor alternative physiologies, which we interpret here as increased habitat heterogeneity.
+If both populations were isolated, natural selection would direct the mean trait values of both populations towards their respective optima.
+%difference in optimal trait values between the local and remote populations, which we call .
+However when straying is allowed, we find that increasingly different trait optima generally lowers $N_T$ and exaggerates $\Delta N$, such that one population has the majority of the biomass (Appendix Fig.
+\ref{fig:thetadiff1},\ref{thetadiff2}).
+The impact of habitat heterogeneity on the portfolio effect is more complex, serving to emphasize the nonlinear relationship between the stray rate and the PE, regardless of heritability.
+%The rate of straying that gives rise to a large PE at the fold bifurcation is conditioned on the difference in optimal trait values between the local and remote populations, which we call $\Delta\theta = \sqrt{(\theta_i - \theta_j)^2}$.
+The extent of the nonlinearity between $m$ and the PE depends very much on $\Delta\theta$, particularly when trait heritability is low, such that $h^2<0.5$.
+As habitat heterogeneity increases, the PE spike generally occurs for lower values of stray rates, meaning that a smaller amount of straying can give rise to alternative stable states (Fig.
+\ref{fig:thetaPE}).
+In the region where alternative stable states are encountered (Fig.
+\ref{fig:thetaPE}), additional straying increases the portfolio effect to a local maximum before its negative effects serve to lower PE with $m\rightarrow 0.5$, and this effect is exaggerated with increasing habitat heterogeneity.
 
-%% PE >> 1 for low values of m
-%% Spike in PE at the fold bifurcation (spike in indvidual population SD)
-%% Increased environmental heterogeneity leads to spikes in PE at lower stray rates
-%% Effects of heritability vs. trait variation and stable states
+% \begin{figure}[h!]
+% \centering
+% \includegraphics[width=0.4\textwidth]{figs/fig_MDPE_hm_cross358.pdf}
+% \caption{
+% The portfolio effect over the stray rate $m$ for trait heritability $h^2=(0.1,0.5)$ for systems with different habitat heterogeneities $\Delta\theta$. Light colors = high values.
+% } \label{fig:PEcross}
+% \end{figure}
 
-% Density dependent m
-%%
+% \begin{figure*}[h!]
+% \centering
+% \includegraphics[width=0.8\textwidth]{figs/fig_MDPE_hm_ddm.pdf}
+% \caption{
+% The same simulations as presented in Figure \ref{fig:PE}, except with density dependent straying, where $m(t) = m_0\left(1-N(t)/(C+N(t)\right)$. Light colors = high values.
+% } \label{fig:ddm}
+% \end{figure*}
 
-{\bf Initial results -- subject to to change:} As the proportion of strays increases, the average traits of both populations veer farther from their optimal values, and this results in the decline of the population steady state (because the recruitment rate is now sub-optimal for both populations).
-This is the nice, expected result.
+% When m and thetadiff are linked
+%%Constant m
+  %%% Low stray rate (high theta1-theta2) leads to alternative stable states and relatively low PE ~ sharp transition
+  %%% High stray rate (low theta1-theta2) increases the steady state (and no A.S.S.)... 
+  %%% This suggests that space naturally buffers against the negative effects of mixing dissimilar populations, however it is NOT SMOOTH
+  %%% This is particularly true for low heritability. High heritability eliminates the A.S.S. region
+  %%% (!!!) Even low levels of straying from distant populations can result in highly divergent alternative steady states! THIS IS A COOL RESULT
 
-The unexpected result are the alternative stable states that arise for lower values of $m$.
-Just where these alternative stable states arise depends on the variance as well as the difference in optimal trait values between sites ($\theta_i-\theta_j$), but you can see that this has big impacts on the portfolio effect.
-Which trajectory goes to the higher or lower alternative stable state is random - an effect of the process noise injected into the system.
-The change in the PE appears to be caused by large spikes in the variance of individual population trajectories where the alternative stable states begin (at the fold bifurcation).
-This is a well-known phenomenon that forms the basis for early warning signals of phase transitions.
+  \begin{figure*}
+    \centering
+    \includegraphics[width=0.8\textwidth]{figs/fig_MDPE_hm_mtheta.pdf}
+    \caption{
+    Assuming that the rate of straying is linked directly to habitat heterogeneity.
+A low stray rate corresponds to very different (or distant) habitats (high $\Delta\theta$), whereas a higher rate of straying corresponds to very similar (or nearby) habitats (low $\Delta\theta$).
+Light colors = high values.
+    } \label{fig:mtheta}
+  \end{figure*}
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.4\textwidth]{figs/fig_Density.pdf}
-\caption{}
-\end{figure}
+{\bf Linking stray rate and habitat heterogeneity} We have so far treated $\Delta\theta$ and $m$ as independent parameters, however we may also assume that if environmental heterogeneity increases with distance -- in particular North-South difference if trait optimality is largely temperature-dependent -- the rate of straying may be expected to decline with distance.
+If we assume this interdependence of $m$ and $\Delta\theta$, low values of straying would correspond to mixing dissimilar (distant) populations, and high values of straying would correspond to mixing similar (nearby) populations.
 
-The differences in optimal trait values between sites ($\theta_i-\theta_j$) influence at what straying rate the alternative stable states appear.
-Large differences in $\theta_i$ and $\theta_j$ can be interpreted as high environmental heterogeneity.
-A cursory analysis shows that increasing environmental heterogeneity tends to lower the stray rate value $m^*$ at which the maximum PE is expected to occur.
-Moreover, if the trait variance is much larger than the difference in optimal trait values between sites (such that selection is weaker), the alternative stable state phenomenon disappears, as does the spike in the portfolio effect.
+In this case we find that alternative stable states now appear for low stray rates and low trait heritability, whereas increasing stray rates result in a single stable state with relatively high $N_T$.
+As before, there is a spike in the PE along the fold bifurcation separating the alternative stable state regime from the single stable state regime, but a sharp decline in the PE as stray rates become very low.
+This is in accordance with intuition as increasing stray rates mean that two very similar populations are mixing, resulting in little negative effect of trait dissimilarity.
 
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.8\textwidth]{figs/fig_HSDcomb.png}
-\caption{}
-\end{figure}
+However, that alternative stable states appear and that PE becomes severely depressed for very low values of PE is surprising: this means that even a small amount of straying of individuals from distant or dissimilar populations can qualitatively alter the dynamics of the metapopulation.
+(for the discussion: an example of this situation may be the salmon populations during the last glacial maximum, where any mixing would be from geographically distant populations)
 
@@
 
+\clearpage
+\setcounter{figure}{0}
+\section*{Appendix}
+
+\begin{figure*}[h!]
+\centering
+\begin{subfigure}[t]{0.55\textwidth}
+\centering
+\includegraphics[width=\textwidth]{figs/fig_MDPE_hm_theta3.pdf} 
+\caption{Low habitat heterogeneity ($\Delta\theta=3$)} \label{fig:thetadiff1}
+\end{subfigure}
+\begin{subfigure}[t]{0.55\textwidth}
+\centering
+\includegraphics[width=\textwidth]{figs/fig_MDPE_hm_theta8.pdf} 
+\caption{High habitat heterogeneity ($\Delta\theta=8$)} \label{fig:thetadiff2}
+\end{subfigure}
+\caption{Total means $N_t$, difference in means $\Delta N$, and the portfolio effect PE for different habitat heterogeneities $\Delta\theta$. Light colors = high values.
+}
+\end{figure*}
+
+\begin{figure*}[h!]
+\centering
+\includegraphics[width=0.8\textwidth]{figs/fig_MDPE_hm_ddm.pdf}
+\caption{
+The same simulations as presented in Figure \ref{fig:PE}, except with density dependent straying, where $m(t) = m_0\left(1-N(t)/(C+N(t)\right)$. Light colors = high values.
+} \label{fig:ddm}
+\end{figure*}
+
 \end{document}

### 2017-05-09 model/KevinSimulations.jl

@@ -83,12 +83,12 @@ beta=0.001;
 theta1=5.0;
 thetadiff=5;
 tau=1.0;
-h=0.2;
+h=0.1;
 sigmaE=0;
-sigmaG=0.2;
-perror=0.05;
+sigmaG=1;
+perror=0.01;
 
-burnin=0.99
+burnin=0.80
 @time for i=1:length(mvec)
   m=mvec[i];
   
@@ -812,7 +812,13 @@ image(x=$mvec,y=$hvec,z=t($(pe[10,:,:])),zlim=c(1,2),col=pal,xlab='m',ylab='h',m
 dev.off()
 """
 
-
+#Average pe over m
+medpe = Array{Float64}(length(mvec));
+pena = pe;
+pena[find(x->x==true,isnan(pe))] = 1;
+for i=1:length(mvec)
+  medpe[i] = median(pena[10,:,i])
+end
 
 
 
@@ -895,13 +901,15 @@ end
 save(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data/data_sig_h_m_theta3.jld"),"n1mean",n1mean,"n2mean",n2mean,"x1mean",x1mean,"x2mean",x2mean,"pe",pe);
 
 
-d = load(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data/data_sig_h_m_theta8.jld"));
+d = load(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data/data_sig_h_m_theta3.jld"));
 #This loads the dictionary
 n1mean = d["n1mean"];
 n2mean = d["n2mean"];
 x1mean = d["x1mean"];
 x2mean = d["x2mean"];
 pe = d["pe"];
+pena = pe;
+pena[find(x->x==true,isnan(pe))] = 1;
 
 maxPE = Array{Float64}(length(sigmavec),length(hvec));
 maxPEvalue = Array{Float64}(length(sigmavec),length(hvec));
@@ -1037,6 +1045,45 @@ dev.off()
 
 
 
+medpe3 = Array{Float64}(length(mvec));
+medpe5 = Array{Float64}(length(mvec));
+medpe8 = Array{Float64}(length(mvec));
+for i=1:length(mvec)
+  medpe3[i] = median(pena3[10,1:51,i])
+  medpe5[i] = median(pena5[10,1:51,i])
[…]

### 2017-05-09 model/KevinSimulations_ddm.jl

@@ -692,6 +692,8 @@ n2mean = d["n2mean"];
 x1mean = d["x1mean"];
 x2mean = d["x2mean"];
 pe = d["pe"];
+penaddm = pe;
+penaddm[find(x->x==true,isnan(pe))] = 1;
 
 maxPE = Array{Float64}(length(sigmavec),length(hvec));
 maxPEvalue = Array{Float64}(length(sigmavec),length(hvec));
@@ -828,6 +830,20 @@ dev.off()
 """
 
 
+dpena = pena5-penaddm;
+namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/figs/fig_diffddm.pdf");
+R"""
+pdf($namespace,height=4,width=5)
+hist($dpena,freq=FALSE,breaks=500,xlim=c(-1,1),xlab='pe-peddm',ylab='Probability',main='',col='lightgray')
+dev.off()
+"""
+dpena2 = pena5-penaddm;
+namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/figs/fig_diffddm.pdf");
+R"""
+pdf($namespace,height=4,width=5)
+hist($dpena,freq=F,breaks=300,xlim=c(-1,1),xlab='pe-peddm',ylab='Probability',main='',col='lightgray')
+dev.off()
+"""
 
 
 #THETADIFF=8

### 2017-05-12 Manuscript/msv1.tex
added

--- 2017-05-09 manuscript/msv1.tex
@@
 
-\usepackage{caption}
+\usepackage[font={small}]{caption}
 \usepackage{subcaption}
@@
 
-Salmony salmons
+%Motivation: straying
+The ability of species to follow specific migratory pathways, and in the case of many salmonids to return to their natal streams after years at sea, is among the most celebrated processes observed in the natural world (REFS).
+That individuals do not always follow these paths has long been appreciated (REFS), however only recently have the mechanisms governing `straying' behaviors been systemically investigated (REFS).
+Although the straying of individuals into sites hosting other populations provides connections within the larger metapopulation, potentially promoting ecological rescue, it also serves to introduce maladapted individuals into habitats that are host to different environmental conditions, and this may lower the mean fitness of the local (mixed) population.
+
+%The tradeoff
+These contrasting effects integrate the ecological dynamics of connected populations as well as the evolutionary dynamics of mixed trait distributions that respond to alternative local conditions.
+Population-level biodiversity is recognized to increase species persistence (REFS), and the resilience and sustainability of highly diverse systems such as the Bristol Bay sockeye populations stand in marked contrast to the comparatively homogeneous chinook salmon of California's Central Valley \cite{Anonymous:2014ku,Satterthwaite:2015ge}.
+Such diversity (or lack thereof) can manifest itself by way of asynchronous population dynamics, where the drivers that give rise to changes in population size vary across a metapopulation, decreasing the potential for synchronization, and correspondingly increasing the potential for ecological rescue (REFS).
+This statistical buffer against extinction has traditionally been quantified as the Portfolio Effect (PE), which is the ratio of the population CV to the CV of the aggregated metapopulation.
+A second measure of diversity can be quantified as genetic variability among populations, where homogenous stocks are more likely to have similar life-history structures (REFS), be at greater risk disease-induced epidemics (REFS), and recruit sub-optimally in heterogeneous environments (REFS).
+Of course, these two measures of diversity are tightly coupled: for instance, lower genetic diversity increases the likelihood that two populations respond similarly to the same stressors, and this promotes synchronized dynamics, thus lowering portfolio effects.
+
+%Eco-evolutionary dynamics in space
+%% From geographic mosaic of coevolution to...
+That evolutionary forces play out heterogeneously across geographic mosaics is now a foundational concept in ecology and evolutionary biology (REFS).
+These mosaics are in part driven by environmental differences between habitats that alter the selective forces acting on different phenotypes (REFS), and a principle underlying assumption is that there is gene flow such that individuals from different habitats mix over space (REFS).
+Although the evolutionary outcomes of these spatial processes have been explored in depth (REFS), it is less well understood how selective mosaics and their consequent evolutionary forces impact population dynamics over ecological timescales.
+How the ecological dynamics of phenotypically diverse metapopulations, wherein constituent populations are subject to different selective regimes, are altered affected by evolutionary forces is less well known.
+% and the subject of this paper.
+
+%Movement through space (link back to straying)
+Migratory populations that return to a breeding ground or natal stream to reproduce are linked to each other by some proportion of the population that stray into the `wrong' site; we might say that there is a \emph{Kevin} in every school or flock (Fig.
+\ref{fig:xkcd}).
+The rate at which individuals stray, $m$, has been subject to a growing amount of both theoretical and empirical research (REFS) and may be linked to errors made at an individual-level that are themselves diminished by larger collections of individuals (REFS).
+Regardless of the mechanism behind straying, the effect that it has on the dynamics of individual populations and the metapopulation as a whole is a topic of considerable interest that has tangible conservation implications (REFS).
+Whether and to what extent the ecological consequences of straying depend on the evolutionary dynamics that emerge from populations distributed across a selection mosaic is unknown.
+How the assumed negative evolutionary effects of straying is balanced by the positive effects of ecological rescue is the subject of this contribution.
+
+%Here we show: pattern formation, dependence on heritability, large effects of small amounts of straying
+
+\begin{figure}
+\centering
+\includegraphics[width=0.4\textwidth]{figs/fig_xkcd.png}
+\caption{
+\emph{Migrating Geese}, a comic from Randall Monroe's xkcd (https://xkcd.com/1729/). 
+A flock of geese travel in the direction of a shared destination, the lone stray named \emph{Kevin}.
+In this case, the rate of straying is $m=0.05$, which is not an uncommon rate for migrating populations of salmon \cite{Satterthwaite:2015ge}. 
+Reprinted under the Creative Commons Attribution-NonCommercial 2.5 License.
+} \label{fig:xkcd}
+\end{figure}
+
+%Alternative stable state
+Here we investigate a minimal eco-evolutionary model of two populations occupying different sites that are linked by straying, each of which with an associated trait distribution subject to natural selection determined by local conditions.
+An important component of this model is the inclusion of a parameter that defines the difference in local conditions that favor different trait optima, increasing values of which correlates to populations that mix across increasingly heterogeneous environments.
+Although our proposed model is most appropriate for modeling the dynamics of salmon populations, the framework is general and the conclusions are likely relevant to a diverse range of organisms where locally adapted populations are linked by migration.
+We first show that specific rates of straying and trait heritability can have large effects on the qualitative dynamics of populations over time, in many cases giving rise to alternative steady states where one site is pushed towards very low biomass.
+The emergence of alternative steady states results in a nonlinear response of the portfolio effect, suggesting that metapopulation persistence can be quite sensitive to the combined influences of straying and selective mosaics.
+
+%Habitat heterogeneity
+A second important finding of our minimal model reveals that systems with greater habitat heterogeneity (measured by an increased difference in the trait values that are optimal between sites) host an intermediate range of straying rates where the portfolio effect increases to a local maximum, signaling that under certain conditions moderate amounts of straying between populations can increase the likelihood of persistence despite the deleterious evolutionary effects overall.
+However, if we suppose that the rate of straying between two sites is correlated with distance, and that the difference in trait optima increases with distance as would be the case if the optima were associated with alternative temperature regimes (especially if sites are distributed latitudinally rather than longitudinally), even a very small amount of straying can drastically reduce the portfolio effect.
+Importantly, the qualitative nature of our results do not depend on whether the stray rate is density dependent or constant.
+
+%Concluding paragraph of the intro
 
@@
 
-\begin{figure}[h]
+\begin{figure}
 \centering
@@
 \caption{
-Median portfolio effect as a function of stray rate $m$ for lower trait heritability ($h^2 < 0.5$)
+Median portfolio effect as a function of stray rate $m$ for lower trait heritability ($h^2 < 0.5$).
+Vertical dashed lines represent the fold bifurcation, which occurs at lower stray rates with increased habitat heterogeneity $\Delta \theta$.
 } \label{fig:thetaPE}
@@
 
-% \begin{figure*}[htp]
-% \centering
-% \begin{subfigure}[t]{0.55\textwidth}
-% \centering
-% \includegraphics[width=\textwidth]{figs/fig_MDPE_hm_theta3.pdf} 
-% \caption{Low habitat heterogeneity ($\Delta\theta=3$)} \label{fig:thetadiff1}
-% \end{subfigure}
-% \begin{subfigure}[t]{0.55\textwidth}
-% \centering
-% \includegraphics[width=\textwidth]{figs/fig_MDPE_hm_theta8.pdf} 
-% \caption{High habitat heterogeneity ($\Delta\theta=8$)} \label{fig:thetadiff2}
-% \end{subfigure}
-% \caption{Total means $N_t$, difference in means $\Delta N$, and the portfolio effect PE for different habitat heterogeneities $\Delta\theta$. Light colors = high values.
-% }
-% \end{figure*}
-
 % Density dependent m
@@
 \caption{
-.
+Difference in PE for simulations with a constant stray rate $m$ and PE for simulations with a density dependent stray rate $m(t)$.
+The differences are minimal (mean difference in PE = -0.05) and result in no discernible qualitative difference in dynamics, though density dependent stray rates generally amplify PE.
 } \label{fig:diffddm}
 \end{figure}
-
-%% Increased habitat heterogeneity
-  %%% 1) exaggerates lowering of mean biomass with increasing stray rates
-  %%% 2) increases the likelihood of alternative stable sites & increases the differences in alternative stable state values with increasing stray rates
-  %%% 3) decreases the PE (!!)
-{\bf The role of habitat heterogeneity} Increased differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater between-site differences in conditions that favor alternative physiologies, which we interpret here as increased habitat heterogeneity.
-If both populations were isolated, natural selection would direct the mean trait values of both populations towards their respective optima.
-%difference in optimal trait values between the local and remote populations, which we call .
-However when straying is allowed, we find that increasingly different trait optima generally lowers $N_T$ and exaggerates $\Delta N$, such that one population has the majority of the biomass (Appendix Fig.
-\ref{fig:thetadiff1},\ref{thetadiff2}).
-The impact of habitat heterogeneity on the portfolio effect is more complex, serving to emphasize the nonlinear relationship between the stray rate and the PE, regardless of heritability.
-%The rate of straying that gives rise to a large PE at the fold bifurcation is conditioned on the difference in optimal trait values between the local and remote populations, which we call $\Delta\theta = \sqrt{(\theta_i - \theta_j)^2}$.
-The extent of the nonlinearity between $m$ and the PE depends very much on $\Delta\theta$, particularly when trait heritability is low, such that $h^2<0.5$.
-As habitat heterogeneity increases, the PE spike generally occurs for lower values of stray rates, meaning that a smaller amount of straying can give rise to alternative stable states (Fig.
-\ref{fig:thetaPE}).
-In the region where alternative stable states are encountered (Fig.
-\ref{fig:thetaPE}), additional straying increases the portfolio effect to a local maximum before its negative effects serve to lower PE with $m\rightarrow 0.5$, and this effect is exaggerated with increasing habitat heterogeneity.
-
-% \begin{figure}[h!]
-% \centering
-% \includegraphics[width=0.4\textwidth]{figs/fig_MDPE_hm_cross358.pdf}
-% \caption{
-% The portfolio effect over the stray rate $m$ for trait heritability $h^2=(0.1,0.5)$ for systems with different habitat heterogeneities $\Delta\theta$. Light colors = high values.
-% } \label{fig:PEcross}
-% \end{figure}
-
-% \begin{figure*}[h!]
-% \centering
-% \includegraphics[width=0.8\textwidth]{figs/fig_MDPE_hm_ddm.pdf}
-% \caption{
-% The same simulations as presented in Figure \ref{fig:PE}, except with density dependent straying, where $m(t) = m_0\left(1-N(t)/(C+N(t)\right)$. Light colors = high values.
-% } \label{fig:ddm}
-% \end{figure*}
-
-% When m and thetadiff are linked
-%%Constant m
-  %%% Low stray rate (high theta1-theta2) leads to alternative stable states and relatively low PE ~ sharp transition
-  %%% High stray rate (low theta1-theta2) increases the steady state (and no A.S.S.)... 
-  %%% This suggests that space naturally buffers against the negative effects of mixing dissimilar populations, however it is NOT SMOOTH
-  %%% This is particularly true for low heritability. High heritability eliminates the A.S.S. region
-  %%% (!!!) Even low levels of straying from distant populations can result in highly divergent alternative steady states! THIS IS A COOL RESULT
 
@@
 
-{\bf Linking stray rate and habitat heterogeneity} We have so far treated $\Delta\theta$ and $m$ as independent parameters, however we may also assume that if environmental heterogeneity increases with distance -- in particular North-South difference if trait optimality is largely temperature-dependent -- the rate of straying may be expected to decline with distance.
-If we assume this interdependence of $m$ and $\Delta\theta$, low values of straying would correspond to mixing dissimilar (distant) populations, and high values of straying would correspond to mixing similar (nearby) populations.
+%% Increased habitat heterogeneity
+  %%% 1) exaggerates lowering of mean biomass with increasing stray rates
+  %%% 2) increases the likelihood of alternative stable sites & increases the differences in alternative stable state values with increasing stray rates
+  %%% 3) decreases the PE (!!)
+{\bf The role of habitat heterogeneity} Increased differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater between-site differences in conditions that favor alternative physiologies, interpreted here as increased habitat heterogeneity.
+At one end of the spectrum, if both populations are isolated, natural selection would direct the mean trait values of both populations towards their respective optima, such that $\mu_i(t) \rightarrow \theta_i$ as $t\rightarrow\infty$.
+%difference in optimal trait values between the local and remote populations, which we call .
+However when straying exists, we find that increasingly divergent trait optima generally lowers $N_T$ and exaggerates $\Delta N$, such that one population has the majority of the biomass (Appendix Fig.
+\ref{fig:thetadiff1},\ref{thetadiff2}).
+The impact of habitat heterogeneity on the portfolio effect is more complex, serving to emphasize the nonlinear relationship between the stray rate and the PE, regardless of heritability.
+%The rate of straying that gives rise to a large PE at the fold bifurcation is conditioned on the difference in optimal trait values between the local and remote populations, which we call $\Delta\theta = \sqrt{(\theta_i - \theta_j)^2}$.
+The extent of the nonlinearity between $m$ and the PE depends very much on $\Delta\theta$, particularly when trait heritability is low ($h^2<0.5$).
+As habitat heterogeneity increases, the fold bifurcation and associated spike in PE generally occurs for lower values of stray rates, meaning that a lower amount of straying can give rise to alternative stable states (Fig.
+\ref{fig:thetaPE}).
+In the region where alternative stable states are feasible (Fig.
+\ref{fig:thetaPE}), additional straying increases the portfolio effect to a local maximum before its negative effects serve to lower PE as $m\rightarrow 0.5$, and this effect is exaggerated with increasing habitat heterogeneity.
 
-In this case we find that alternative stable states now appear for low stray rates and low trait heritability, whereas increasing stray rates result in a single stable state with relatively high $N_T$.
-As before, there is a spike in the PE along the fold bifurcation separating the alternative stable state regime from the single stable state regime, but a sharp decline in the PE as stray rates become very low.
+% When m and thetadiff are linked
+%%Constant m
+  %%% Low stray rate (high theta1-theta2) leads to alternative stable states and relatively low PE ~ sharp transition
+  %%% High stray rate (low theta1-theta2) increases the steady state (and no A.S.S.)... 
+  %%% This suggests that space naturally buffers against the negative effects of mixing dissimilar populations, however it is NOT SMOOTH
+  %%% This is particularly true for low heritability. High heritability eliminates the A.S.S. region
+  %%% (!!!) Even low levels of straying from distant populations can result in highly divergent alternative steady states! THIS IS A COOL RESULT
+
+{\bf Linking stray rate and habitat heterogeneity} We have so far treated $\Delta\theta$ and $m$ as independent parameters, however we may also assume that if environmental heterogeneity increases with distance, the rate of straying may be expected to decline given that individuals are less likely to stray into distant habitats.
+%-- in particular North-South difference if trait optimality is largely temperature-dependent --
+If we incorporate this interdependence of $m$ and $\Delta\theta$, low rates of straying would correspond to mixing dissimilar (distant) populations, and high rates of straying would correspond to mixing similar (nearby) populations.
+
+Here we find that alternative stable states now appear for low stray rates and over any value of trait heritability $h^2<1$.
+As stray rates increase, a single stable state emerges with a correspondingly high $N_T$.
+As before, there is a spike in the PE at the fold bifurcation separating the alternative stable state regime from the single stable state regime, and a sharp decline in the PE as stray rates become very low.
 This is in accordance with intuition as increasing stray rates mean that two very similar populations are mixing, resulting in little negative effect of trait dissimilarity.
 
-However, that alternative stable states appear and that PE becomes severely depressed for very low values of PE is surprising: this means that even a small amount of straying of individuals from distant or dissimilar populations can qualitatively alter the dynamics of the metapopulation.
-(for the discussion: an example of this situation may be the salmon populations during the last glacial maximum, where any mixing would be from geographically distant populations)
+That alternative stable states appear and that PE becomes severely depressed for very low rates of straying is surprising: this means that even a small amount of mixing of populations from distant or dissimilar populations can qualitatively alter the dynamics of the metapopulation, regardless of trait heritability.
+%(for the discussion: an example of this situation may be the salmon populations during the last glacial maximum, where any mixing would be from geographically distant populations)
 
 \section*{Discussion}
+
+We have shown that the natural selection of mixed populations adapted to different environments can result in large effects on population dynamics and the degree to which the metapopulation is buffered against extinction, which we quantify as the portfolio effect.
+The inclusion of individuals with trait values far from the local optimum generally lowers the combined steady state biomass $N_T$ and increases the likelihood that the system gives way to alternative steady states, pushing one of the populations close to extinction.
+Although the emergence of alternative steady states can result in very large portfolio effects near the bifurcation, alternative steady state regimes tend to result in very low portfolio effects, indicating that the metapopulation is less likely to persist.
+
+The degree to which the PE is lowered depends to a large extent on \emph{i}) trait heritability $h^2$ and \emph{ii}) habitat heterogeneity $\Delta \theta$.
+Increased heritability means that the effects of natural selection on the metapopulation are larger, increasing the coupling between evolutionary and ecological dynamics. 
+Our minimal model of straying between two populations shows that greater $h^2$ buffers the metapopulation against the emergence of alternative stable states for lower stray rates, but magnifies the negative effects of alternative stable states (captured by increased $\Delta N$ and decreased PE; Fig.
+\ref{fig:PE}B,C) once this regime is encountered at higher stray rates.
+Although the rate of straying has been shown to be density dependent (REFS), the qualitative results of our model are relatively insensitive to this dynamic and we limit discussion to the simplified case of a constant stray rate $m$ with the understanding that our findings also apply to the case density dependent stray rates $m(t)$.
+
+%Our results in the context of Turing bifurcation
+The onset of alternative stable states in a spatial context describes the emergence of spatial pattern formation (REFS), which is more generally defined as a Turing instability (REFS).
+The mathematical conditions that lead to Turing instabilities are well-known in both continuous (REFS) and discrete spatial contexts (REFS)
+
+%North-South vs East-West
 

### 2017-05-12 Manuscript/msv1.tex
phenotypic divergence para

@@
 
+%Lowering Phenotypic diversity with increasing straying rate
+As expected, increasing rates of straying lowers phenotypic diversity because both local and remote populations become homogenized.
+The sensitivity of this phenotypic homogenization to straying rate is increased with lower heritability because the selective forces acting against trait means far from local optima is lessened for lower $h^2$.
+[A DISCRETE JUMP TO LOW DIVERSITY FOR LOW HERITABILTIY](Fig. \ref{fig:traitdiff}).
+
+\begin{figure}
+\centering
+\includegraphics[width=0.4\textwidth]{figs/fig_traitdiff.pdf}
+\caption{
+} \label{fig:traitdiff}
+\end{figure}
+
 % Density dependent m
@@
 %Because density dependent straying does not alter that qualitative results of our model, we limit additional analyses
+
+%% Increased habitat heterogeneity
+  %%% 1) exaggerates lowering of mean biomass with increasing stray rates
+  %%% 2) increases the likelihood of alternative stable sites & increases the differences in alternative stable state values with increasing stray rates
+  %%% 3) decreases the PE (!!)
+{\bf The role of habitat heterogeneity} Increased differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater between-site differences in conditions that favor alternative physiologies, interpreted here as increased habitat heterogeneity.
+At one end of the spectrum, if both populations are isolated, natural selection would direct the mean trait values of both populations towards their respective optima, such that $\mu_i(t) \rightarrow \theta_i$ as $t\rightarrow\infty$.
+%difference in optimal trait values between the local and remote populations, which we call .
+However when straying exists, we find that increasingly divergent trait optima generally lowers $N_T$ and exaggerates $\Delta N$, such that one population has the majority of the biomass (Appendix Fig.
+\ref{fig:thetadiff1},\ref{thetadiff2}).
+The impact of habitat heterogeneity on the portfolio effect is more complex, serving to emphasize the nonlinear relationship between the stray rate and the PE, regardless of heritability.
+%The rate of straying that gives rise to a large PE at the fold bifurcation is conditioned on the difference in optimal trait values between the local and remote populations, which we call $\Delta\theta = \sqrt{(\theta_i - \theta_j)^2}$.
+The extent of the nonlinearity between $m$ and the PE depends very much on $\Delta\theta$, particularly when trait heritability is low ($h^2<0.5$).
+As habitat heterogeneity increases, the fold bifurcation and associated spike in PE generally occurs for lower values of stray rates, meaning that a lower amount of straying can give rise to alternative stable states (Fig.
+\ref{fig:thetaPE}).
+In the region where alternative stable states are feasible (Fig.
+\ref{fig:thetaPE}), additional straying increases the portfolio effect to a local maximum before its negative effects serve to lower PE as $m\rightarrow 0.5$, and this effect is exaggerated with increasing habitat heterogeneity.
 
@@
 \end{figure*}
-
-%% Increased habitat heterogeneity
-  %%% 1) exaggerates lowering of mean biomass with increasing stray rates
-  %%% 2) increases the likelihood of alternative stable sites & increases the differences in alternative stable state values with increasing stray rates
-  %%% 3) decreases the PE (!!)
-{\bf The role of habitat heterogeneity} Increased differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater between-site differences in conditions that favor alternative physiologies, interpreted here as increased habitat heterogeneity.
-At one end of the spectrum, if both populations are isolated, natural selection would direct the mean trait values of both populations towards their respective optima, such that $\mu_i(t) \rightarrow \theta_i$ as $t\rightarrow\infty$.
-%difference in optimal trait values between the local and remote populations, which we call .
-However when straying exists, we find that increasingly divergent trait optima generally lowers $N_T$ and exaggerates $\Delta N$, such that one population has the majority of the biomass (Appendix Fig.
-\ref{fig:thetadiff1},\ref{thetadiff2}).
-The impact of habitat heterogeneity on the portfolio effect is more complex, serving to emphasize the nonlinear relationship between the stray rate and the PE, regardless of heritability.
-%The rate of straying that gives rise to a large PE at the fold bifurcation is conditioned on the difference in optimal trait values between the local and remote populations, which we call $\Delta\theta = \sqrt{(\theta_i - \theta_j)^2}$.
-The extent of the nonlinearity between $m$ and the PE depends very much on $\Delta\theta$, particularly when trait heritability is low ($h^2<0.5$).
-As habitat heterogeneity increases, the fold bifurcation and associated spike in PE generally occurs for lower values of stray rates, meaning that a lower amount of straying can give rise to alternative stable states (Fig.
-\ref{fig:thetaPE}).
-In the region where alternative stable states are feasible (Fig.
-\ref{fig:thetaPE}), additional straying increases the portfolio effect to a local maximum before its negative effects serve to lower PE as $m\rightarrow 0.5$, and this effect is exaggerated with increasing habitat heterogeneity.
 

### 2017-05-17 Manuscript/msv1.tex
ddm figure updates and manuscript edits

--- 2017-05-09 manuscript/msv1.tex
@@
 
-%\bibliographystyle{ieeetr}
+\bibliographystyle{ieeetr}
+
+\usepackage{bibunits}
 
@@
 %\title{Community assembly and dynamics by the deconstruction of species interactions}
-\title{Eco-evolutionary model of salmon straying in a metapopulation (but with a cool title)}
-%\author{Justin D. Yeakel${}^{1,2,*}$, Christopher P. Kempes${}^{2}$, \& Sidney Redner${}^{2,3}$ \\ \\
-%${}^1$School of Natural Science, University of California Merced, Merced, CA \\
-%${}^2$The Santa Fe Institute, Santa Fe, NM \\
-%${}^3$Department of Physics, Boston University, Boston MA \\
-%${}^*$To whom correspondence should be addressed: jdyeakel@gmail.com
-%}
+\title{Lost in space: the eco-evolutionary impacts of straying on metapopulation persistence}
+\author{Justin D. Yeakel${}^{1,2,*}$, Jean Phillipe Gibert${}^{1}$, Sean Anderson${}^{3}$, Peter Westley${}^{4}$, \& Jon W. Moore${}^{5}$ \\
+${}^1$School of Natural Sciences, University of California, Merced, Merced CA, USA \\
+${}^2$The Santa Fe Institute, Santa Fe NM, USA \\
+${}^3$Department of Biology, University of Washington, Seattle WA, USA \\
+${}^4$Institute for Arctic Biology, University of Alaska, Fairbanks, Fairbanks AK, USA \\
+${}^5$Earth${}_2$Oceans Research Group, Simon Fraser University, Vancouver BC, Canada \\
+${}^*$To whom correspondence should be addressed: jdyeakel@gmail.com
+}
 
@@
 
-\section*{Introduction}
+\section{Introduction}
 
-Salmony salmons
+%Motivation: straying
+Coordinated mass migrations are one of the great wonders of the natural world, and the ability of individuals and groups to navigate across great distances have long fascinated naturalists \cite{MilnerGulland:2011vm}.
+Anadromous salmonid fishes (genera \emph{Oncorhynchus} and \emph{Salmo}) embody these astonishing migrations by homing with high accuracy and precision to their natal streams for reproduction after years at sea \cite{Quinn:2011tf,Jonsson:2011kg,Keefer:2014gg}.
+Nothwithstanding their philopatric tendancy, not all individuals home, but rather `stray' to non-natal sites to spawn \cite{Quinn:1993ge,Hendry:2004wf,H:2013fs}.
+Although extensive work has been done to document the extent of straying from donor populations and into recipient populations \cite{Keefer:2014gg,Bett:2017ha}, only recently have the abiotic, biotic, and anthropogenic influences of `straying' behaviors been investigated systemically \cite{Keefer:2008bs,Westley:2015to,Bond:2016dz}.
+Most recently the role of social interactions and collective navigation has been hypothesized \cite[][; this volume]{Berdahl:2015kv,Berdahl:2016dx}.
+Although the straying of individuals into sites hosting other populations provides connections within the larger metapopulation, potentially promoting ecological and evolutionary rescue, it also serves to introduce maladapted individuals into habitats that are host to different environmental conditions, and this may lower the mean fitness of the local (mixed) population \cite{Muhlfeld:2014hs}.
 
-\section*{Model Description}
+%The tradeoff
+The dual nature of straying as both promoter of connections among metapopulation demes and potential eroder of locally adapted gene complexes highlights the interplay between ecological dynamics of connected populations as well as the evolutionary dynamics of mixed trait distributions that respond to alternative local conditions.
+Population-level biodiversity is recognized to increase species persistence (REFS), and the resilience and sustainability of metapopulations (general REF).
+For example, the long-term sustainability of the Bristol Bay sockeye salmon fishery is due in large part to intact habitat that in conjunction with fine-scale homing and reduced gene flow provides a template for the evolution of countless locally adapted populations \cite{Hilborn:2003gf,Schindler:2010he,Anonymous:2014ku,Satterthwaite:2015ge}.
+In contrast, areas such as California's Central Valley where habitats have been lost, degraded, and homogenized stand in marked contrast.
+Within species diversity (or lack thereof) can manifest itself by way of asynchronous population dynamics, where the drivers that give rise to changes in population size vary across a metapopulation, decreasing the potential for synchronization, and correspondingly increasing the potential for ecological rescue (REFS).
+This statistical buffer against extinction has traditionally been quantified as the Portfolio Effect (PE), which is the ratio of the population CV to the CV of the aggregated metapopulation.
+Weakened portfolio dynamics can also emerge from homogenization resulting from  reduced genetic variability among populations, where homogenous stocks are more likely to have similar life-history structures (REFS), be at greater risk disease-induced epidemics (REFS), and recruit sub-optimally in heterogeneous environments (REFS).
+Of course, these two measures of diversity often occur concurrently: for instance, lower genetic diversity increases the likelihood that two populations respond similarly to the same stressors, and this promotes synchronized dynamics, thus lowering portfolio effects.
 
-Here we consider two populations $N_1$ and $N_2$ that are separated in space, each with trait values $x_1$ and $x_2$ determining recruitment rates.
+%Eco-evolutionary dynamics in space
+%% From geographic mosaic of coevolution to...
+That evolutionary forces play out heterogeneously across geographic mosaics is now a foundational concept in ecology and evolutionary biology (REFS).
+These mosaics are in part driven by environmental differences between habitats that alter the selective forces acting on different phenotypes \cite{Endler:1986tz}, and a principle underlying assumption is that there is gene flow such that individuals from different habitats mix over space (REFS).
+Although the evolutionary outcomes of these spatial processes have been explored in depth (REFS), it is less well understood how selective mosaics and their consequent evolutionary forces impact population dynamics over contemporary timescales \cite{Hendry:2016un}.
+How the ecological dynamics of phenotypically diverse metapopulations, wherein constituent populations are subject to different selective regimes, are altered affected by evolutionary forces is less well known.
+% and the subject of this paper.
+
+%Movement through space (link back to straying)
+Migratory populations that return to a breeding ground or natal stream to reproduce are linked to each other by some proportion of the population that permanently disperse, or stray into the `wrong' site; we might say that there is at least one \emph{Kevin} in every school or flock in every school or flock (Fig.
+\ref{fig:xkcd}).
+The rate at which individuals stray, $m$, has been subject to a growing amount of both theoretical and empirical research \cite{H:2013fs,Keefer:2014gg,Bett:2017ha} and may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \cite{Berdahl:2015kv,Berdahl:2016dx}.
+Regardless of the mechanism's governing straying, the effect that it has on the dynamics of individual populations and the metapopulation as a whole is a topic of considerable interest that has tangible conservation implications \cite{Brenner:2012gl,Johnson:2012fe,Fullerton:2011ii}.
+Whether, and to what extent, the ecological consequences of straying depend on the evolutionary dynamics that emerge from populations distributed across a selection mosaic is unclear.
+How the assumed negative evolutionary effects of straying and subsequent gene flow is balanced by the positive effects of demographic rescue is the subject of this contribution.
+
+%Here we show: pattern formation, dependence on heritability, large effects of small amounts of straying
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.4\textwidth]{figs/fig_xkcd.png}
+\caption{
+\emph{Migrating Geese}, a comic from Randall Monroe's xkcd (https://xkcd.com/1729/). 
+A flock of geese travel in the direction of a shared destination, the lone stray named \emph{Kevin}.
+In this case, the rate of straying is $m=0.05$, which is not an uncommon rate for migrating populations of salmon \cite{Satterthwaite:2015ge}. 
+Reprinted under the Creative Commons Attribution-NonCommercial 2.5 License.
+} \label{fig:xkcd}
+\end{figure}
+
+%Alternative stable state
+Here we ask the overarching question of how collective behavior mediated dispersal and gene flow interact to influence the persistence of locally adapted populations and persistence of those populations.
+To address this question we construct a minimal eco-evolutionary model of two populations occupying different sites that are linked by straying individuals, each of which with an associated trait distribution subject to natural selection determined by local conditions.
+An important and relatively novel component of this model is the inclusion of a parameter that defines the difference in local conditions that favor different trait optima, increasing values of which correlates to populations that mix across increasingly heterogeneous environments \cite[though see ]{Berdahl:2015kv}.
+Although our proposed model is was constructed with the dynamics of salmon populations in mind, the framework is general and the conclusions are likely relevant to a diverse range of migratory organisms where locally adapted populations are linked by dispersal.
+We first show that specific rates of straying and trait heritability can have large effects on the qualitative dynamics of populations over time, in many cases giving rise to alternative steady states where one site is pushed towards very low biomass.
+The emergence of alternative steady states results in a nonlinear response of the portfolio effect, suggesting that metapopulation persistence can be quite sensitive to the combined influences of  collective straying across a selective landscape.
+
+%Habitat heterogeneity
+A second important finding of our minimal model reveals that systems with greater habitat heterogeneity (measured by an increased difference in the trait values that are optimal between sites) host an intermediate range of straying rates where the portfolio effect increases to a local maximum, signaling that under certain conditions moderate amounts of straying between populations can increase the likelihood of persistence despite the deleterious evolutionary effects overall.
+However, if we suppose that the rate of straying between two sites is correlated with distance (as indicated by empirical evidence), and that the difference in trait optima increases with distance as would be the case if the optima were associated with alternative temperature regimes (especially if sites are distributed latitudinally rather than longitudinally), even a very small amount of straying can drastically reduce the portfolio effect.
+Importantly, the qualitative nature of our results do not depend on whether the stray rate is density dependent or constant, suggesting a limited role of collective dispersal on the dynamics considered herein.
+
+%Concluding paragraph of the intro
+
+\section{Model Description}
+
+Here we consider two populations $N_1$ and $N_2$ that are separated in space in distinct habitats, each with trait values $x_1$ and $x_2$ determining recruitment rates.
 We assume that there is an optimum trait value $\theta_1$ and $\theta_2$ associated with each habitat, where recruitment is maximized if the trait value of the local population $x = \theta$.
@@
 As such, the recruitment rate for both populations is determined by the mean trait value of the local population, such that $r_1 = R_1[\mu_1(t)|\theta_1]$.
-Trait means for each population are subject to selection, the strength of which depends on the difference between the population mean and the local trait optimum at a given point in time.
+Trait means for each population are subject to selection, the strength of which depends on the difference between the population mean and the local trait optimum at a given point in time \cite{simpson1953major,Lande:1976ga}.
 
-The two populations are assumed to reproduce in spatially separate sites that are close enough such that a proportion of the population $m$ can stray into the wrong site, and where mortality occurs before individuals return to spawn.
+The two populations are assumed to reproduce in spatially separate sites that are close enough such that a proportion of the population $m$ can stray into the other site, and where mortality occurs before individuals return to reproduce.
 If there is no straying between these populations (such that they are independent), then the mean trait evolves towards the optimal value such that $x_1 \rightarrow \theta_1$, and the recruitment rate for that population will be maximized.
@@
 
-\begin{figure}[h]
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
 \centering
@@
 A) The steady state densities of $N_1$ and $N_2$ as a function of a constant stray rate $m$.
-B) The steady state trait values measured as $theta_i - x_i$, as a function of a constant stray rate $m$. 
+B) The steady state trait values measured as $\theta_i - x_i$, as a function of a constant stray rate $m$. 
 } \label{fig:traj}
@@
 
-We use the discrete Ricker population dynamic framework described by Shelton and Mangel \cite{} as the basis for our two-site model, with the added effect of the local population $N_i$ mixing with a set proportion $m$ of a remote population $N_j$ that is straying into it.
+We use the discrete Ricker population dynamic framework described by Shelton and Mangel \cite{Shelton:2011eq} as the basis for our two-site model, with the added effect of the local population $N_i$ mixing with a set proportion $m$ of a remote population $N_j$ that is straying into it.
 We first assume that the proportion ${\rm e}^{-Z}$ of both populations survive, and that the aggregated mix of the populations (local individuals in addition to the straying individuals) are subject to the same compensatory effects, determined by the parameter $\beta$.
@@
 
-\noindent As stated previously, it is the mismatch between the local trait mean $\mu_i(t)$ and the local optimum $\theta_i$ that determines the recruitment rate for the population.
+\noindent Again, it is the mismatch between the local trait mean $\mu_i(t)$ and the local optimum $\theta_i$ that determines the recruitment rate for the population.
 The parameter $\tau$ controls the sensitivity of recruitment to changes in the mean trait value away from the optimum, which we set as $\tau=1$ here and throughout.
@@
 \begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
 \centering
@@
 
-An increasing flow of incoming strays is thus expected to pull the mean trait value of the local population away from its optimum, which will decrease its rate of recruitment.
+An increasing flow of incoming strays is generally expected to pull the mean trait value of the local population away from its optimum over time, which will decrease its rate of recruitment.
 The mean trait value thus changes through time according to the difference equation
@@
 Specifically, the rate at which individuals stray has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tends to decrease the rate at which individuals stray, thus reducing the overall proportion of a population that strays.
-According to REF, given the probability that an individual strays $m_0$, the proportion of the local population $N_i(t)$ that strays is
+According to \citeauthor{Berdahl:2016dx}, given the probability that an individual strays $m_0$, the proportion of the local population $N_i(t)$ that strays is
 
@@
 %Linking m/m0 and thetadiff
-Stray rate is intrinsically linked to the distance between the local and straying population.
+The straying rate is intrinsically linked to the distance between the donor and recipient population.
 The greater the distance between two populations, the lower the expected rate of straying (REF).
-We can account for this interdependence in our model by assuming that $m$ (if the stray rate is constant) or $m_0$ (if the stray rate is density dependent) is a function of $theta_i-theta_j$, which can be assumed to be large if the remote site $j$ is a great distance away from the local site $i$.
-If sites $i$ and $j$ are very close, the stray rate is assumed to maximized at $m,m_0 = 0.5$.
+We can account for this interdependence in our model by assuming that $m$ (if the stray rate is constant) or $m_0$ (if the stray rate is density dependent) is a function of $\theta_i-\theta_j$, which can be assumed to be large if the remote site $j$ is a great distance away from the local site $i$.
+If sites $i$ and $j$ are very close, the stray rate is assumed to maximized at $m,m_0 = 0.5$, assuming both sites are equally attractive to the respective populations.
 Thus, we can integrate these two variables by setting $m,m_0 = (2 + \epsilon (\theta_i-\theta_j))^{-1}$, where $\epsilon$ sets the sensitivity of a declining $m$ to increasing distance (greater values of $\theta_i-\theta_j$).
 
-\section*{Results}
+\section{Results}
 
@@
 %Alternative stable states
-{\bf Nonlinear effects of straying on the Portfolio Effect} Straying generally lowers steady state densities for both local and remote populations.
+
+\noindent{\bf (a) Nonlinear effects of straying on the Portfolio Effect} \\
+\noindent Straying generally lowers steady state densities for both recipient and donor populations by \emph{i}) losing locally-adapted individuals due to straying and \emph{ii}) introducing maladapted individuals from the donor population.
 %, regardless of trait heritability and variance or habitat heterogeneity
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=0.4\textwidth]{figs/fig_thetaPE.pdf}
-\caption{
-Median portfolio effect as a function of stray rate $m$ for lower trait heritability ($h^2 < 0.5$)
-} \label{fig:thetaPE}
-\end{figure}
-
 Together these changes in steady state population densities in terms of $N_T$ and $\Delta N$ as a function of trait heritability and the rate of straying between populations give rise to highly nonlinear portfolio effects, quantified here as
@@
 \noindent We define $\overline{{\rm CV}}_{i,j}$ as the average coefficient of variation across the local population $i$ and the remote population $j$ at steady state densities, and ${\rm CV}_T$ as the coefficient of variation of the total or aggregated population at steady state densities.
-The minimum portfolio effect is by definition ${\rm PE_min}=1$, whereas portfolio effects greater than unity corresponds to a greater potential for ecological rescue among populations, thus buffering the system as a whole against extinction.
+The minimum portfolio effect is by definition ${\rm PE_{\rm min}}=1$, whereas portfolio effects greater than unity corresponds low synchronization \cite{Loreau:2008ju,Anderson:2014cx,Yeakel:2013vz} and a greater potential for demographic rescue among populations, thus buffering the system as a whole against extinction.
 
@@
 As the fold bifurcation is approached with greater rates of straying, the portfolio effect spikes due to a large increase in the standard deviation of both populations.
-This explosion in variance is a well-known phenomenon that occurs near a fold bifurcation and lies at the heart of early warning signal theory (REFS).
+This explosion in variance is due to critical slowing down that leads to a change in variance near fold bifurcations, and this marked change in dynamic behavior has been suggested to serve as an early warning signal of an approaching phase transition \cite{Scheffer:2009gg,Anonymous:2013br,Dakos:2014br}.
+The detectibility of such changes in dynamical behavior among salmonids appears to be idosynchratic across species, and the difficulty in measuring critical slowing down may - ironically - be masked by large portfolio effects \cite{Krkosek:2014ch} [this might be more appropriate for the discussion].
 For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}C), where there is a high and low steady state density among the sites, if heritability is low the portfolio effect becomes minimized, is elevated and then declines as $m\rightarrow 0.5$.
@@
 
-% \begin{figure*}[htp]
-% \centering
-% \begin{subfigure}[t]{0.55\textwidth}
-% \centering
-% \includegraphics[width=\textwidth]{figs/fig_MDPE_hm_theta3.pdf} 
-% \caption{Low habitat heterogeneity ($\Delta\theta=3$)} \label{fig:thetadiff1}
-% \end{subfigure}
-% \begin{subfigure}[t]{0.55\textwidth}
-% \centering
-% \includegraphics[width=\textwidth]{figs/fig_MDPE_hm_theta8.pdf} 
-% \caption{High habitat heterogeneity ($\Delta\theta=8$)} \label{fig:thetadiff2}
-% \end{subfigure}
-% \caption{Total means $N_t$, difference in means $\Delta N$, and the portfolio effect PE for different habitat heterogeneities $\Delta\theta$. Light colors = high values.
-% }
-% \end{figure*}
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.4\textwidth]{figs/fig_thetaPEmvm.pdf}
+\caption{
+(a) Median portfolio effect as a function of stray rate $m$ for lower trait heritability ($h^2 < 0.5$).
+Vertical dashed lines represent the fold bifurcation, which occurs at lower stray rates with increased habitat heterogeneity $\Delta \theta$.
+(b) Within the box shown in (a), PE as a function of the constant straying rate is shown as the solid line.
+Paired open circles show PE as a function of a density dependent straying rate.
+Straying rates are paired because the recipient and donor populations exist in alternative steady states.
+The lower straying rate of a pair is for the larger population; the higher straying rate is for the smaller population.
+The different steady state regimes that give rise to the range of straying rates were obtained by varying the individual straying probability $m_0$ (where $m_0$ and $m^*$ are linearly related).
+} \label{fig:thetaPE}
+\end{figure}
+
+%Lowering Phenotypic diversity with increasing straying rate
+As expected, an increased straying rate lowers phenotypic diversity ($\Delta \mu = |\mu_i(t)-\mu_j(t)|$, evaluated at the steady state) because both local and remote populations become homogenized.
+The sensitivity of this phenotypic homogenization to straying is increased with lower heritability because the selective forces acting against trait means far from the local optima is lessened for lower $h^2$.
+Less intuitively, we observe a discrete jump towards low phenotypic diversity at low-moderate ($0.05 \leq m < 0.2$) straying rates, which occurs as the consequence of crossing the fold bifurcation (Fig.
+\ref{fig:traitdiff}).
+\\
 
@@
 %% Higher rates of straying are tolerable (there is a not a decline in PE for very high m0 as there is for constant m)
-If we assume that the rate of straying is density dependent, the probability of straying at the individual level $m_0$ determines the rate of straying within the population, such that $m(t)$ becomes lower as $N(t)$ increases.
-At steady state values, $m(t)$ also becomes constant, though by definition is always less than $m_0$, such that $0 < m(t) < m_0$.
-We find that this dynamic has negligible impact on the qualitative results on our system (Fig. \ref{fig:diffddm}, Appendix Fig. \ref{fig:ddm}).
-Quantitatively, density dependent straying serves to lessen the effect of straying on $N_T$, $\Delta N$ and PE, given that the effective stray rate is effectively lessened by collective navigation.
-%Because density dependent straying does not alter that qualitative results of our model, we limit additional analyses
+
+\noindent{\bf (b) The effects of density dependent straying}\\
+\noindent If we assume that the rate of straying is density dependent, the probability of straying at the individual level $m_0$ determines the rate of straying within the population, such that $m(t)$ becomes lower as $N(t)$ increases (Eq.
+\ref{eq:ddm}).
+At the steady state $m(t)=m^*$ is also constant, though by definition, it is always less than $m_0$ such that $0 < m^* < m_0$.
+Because our analyses concern steady state conditions, the inclusion of density-dependent straying does not have a large impact on the qualitative nature of our findings (Appendix Fig. \ref{fig:ddm}).
+
+The assumption that straying is density dependent alters the strength of the portfolio effect for a given $m^*$.
+As shown in Figure \ref{fig:thetaPE}(b), in the alternative stable state regime, because the recipient and donor population exist at different steady state densities, they have different rates of straying such that a given value of the portfolio effect is now associated with a pair of straying rates $(m_i^*,m_j^*)$.
+A range of $(m_i^*,m_j^*)$ values are obtained by varying $m_0$, which is positively and linearly related to $(m_i^*,m_j^*)$.
+Importantly, we show that metapopulations settling at relatively low rates of straying have lower-than-expected portfolio effects, whereas metapopulations settling at relatively high rates of straying have higher-than expected portfolio effects.
+This suggests that there is a strong tradeoff in the role of density dependent straying on the persistence of metapopulations: as the individual probability of straying is lowered, the straying rates of both recipient and donor populations are similarly lowered, however this diminishes the system's ability to buffer itself against extinction as measured by the portfolio effect.
+Conversely, as the probability that an individual strays increases, so does rate of straying for both recipient and donor populations, and this serves to elevate the portfolio effect compared to a constant $m$.
+This effect remains but is somewhat diminished as habitats become increasingly heterogeneous.
+\\
 
 \begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
 \centering
-\includegraphics[width=0.4\textwidth]{figs/fig_diffddm.pdf}
+\includegraphics[width=0.4\textwidth]{figs/fig_traitdiff.pdf}
 \caption{
-.
-} \label{fig:diffddm}
+Phenotypic diversity ($\Delta \mu$) evaluated at the steady state as a function of straying rate $m$ and trait heritability $h^2$.
+The descrete jump occurs as the system crosses the fold bifurcation; lower phenoytic diversity emerges with higher straying rates and in the alternative steady state regime.
+} \label{fig:traitdiff}
 \end{figure}
@@
   %%% 3) decreases the PE (!!)
-{\bf The role of habitat heterogeneity} Increased differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater between-site differences in conditions that favor alternative physiologies, which we interpret here as increased habitat heterogeneity.
-If both populations were isolated, natural selection would direct the mean trait values of both populations towards their respective optima.
+
+\noindent{\bf (c) The role of habitat heterogeneity on the portfolio effect}\\ 
+\noindent Increased differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater between-site differences in conditions that favor alternative traits, interpreted here as increased habitat heterogeneity.
+At one end of the spectrum, if both populations are isolated, natural selection would direct the mean trait values of both populations towards their respective optima, such that $\mu_i(t) \rightarrow \theta_i$ as $t\rightarrow\infty$.
 %difference in optimal trait values between the local and remote populations, which we call .
-However when straying is allowed, we find that increasingly different trait optima generally lowers $N_T$ and exaggerates $\Delta N$, such that one population has the majority of the biomass (Appendix Fig.
+However when straying exists, we find that increasingly divergent trait optima generally lowers $N_T$ and exaggerates $\Delta N$, such that one population has the majority of the biomass (Appendix Fig.
 \ref{fig:thetadiff1},\ref{thetadiff2}).
+Which population becomes large or small depends on the initial conditions, which are randomized. 
 The impact of habitat heterogeneity on the portfolio effect is more complex, serving to emphasize the nonlinear relationship between the stray rate and the PE, regardless of heritability.
 %The rate of straying that gives rise to a large PE at the fold bifurcation is conditioned on the difference in optimal trait values between the local and remote populations, which we call $\Delta\theta = \sqrt{(\theta_i - \theta_j)^2}$.
-The extent of the nonlinearity between $m$ and the PE depends very much on $\Delta\theta$, particularly when trait heritability is low, such that $h^2<0.5$.
-As habitat heterogeneity increases, the PE spike generally occurs for lower values of stray rates, meaning that a smaller amount of straying can give rise to alternative stable states (Fig.
+The extent of the nonlinearity between $m$ and the PE depends very much on $\Delta\theta$, particularly when trait heritability is low ($h^2<0.5$).
+As habitat heterogeneity increases, the fold bifurcation and associated spike in PE generally occurs for lower values of straying rate $m$, meaning that a lower amount of straying can give rise to alternative stable states (Fig.
 \ref{fig:thetaPE}).
-In the region where alternative stable states are encountered (Fig.
-\ref{fig:thetaPE}), additional straying increases the portfolio effect to a local maximum before its negative effects serve to lower PE with $m\rightarrow 0.5$, and this effect is exaggerated with increasing habitat heterogeneity.
+In the region where alternative stable states are feasible (Fig.
+\ref{fig:thetaPE}), additional straying increases the portfolio effect to a local maximum before its negative effects serve to lower PE as $m\rightarrow 0.5$, and this effect is exaggerated with increasing habitat heterogeneity.
+\\
 
-% \begin{figure}[h!]
+% 
+% \begin{figure}
+%   \captionsetup{justification=raggedright,
+% singlelinecheck=false
+% }
 % \centering
-% \includegraphics[width=0.4\textwidth]{figs/fig_MDPE_hm_cross358.pdf}
+% \includegraphics[width=0.4\textwidth]{figs/fig_diffddm.pdf}
 % \caption{
-% The portfolio effect over the stray rate $m$ for trait heritability $h^2=(0.1,0.5)$ for systems with different habitat heterogeneities $\Delta\theta$. Light colors = high values.
-% } \label{fig:PEcross}
+% Difference in PE for simulations with a constant stray rate $m$ and PE for simulations with a density dependent stray rate $m(t)$.
+The differences are minimal (mean difference in PE = -0.05) and result in no discernible qualitative difference in dynamics, though density dependent stray rates generally amplify PE.
+% } \label{fig:diffddm}
 % \end{figure}
 
-% \begin{figure*}[h!]
-% \centering
-% \includegraphics[width=0.8\textwidth]{figs/fig_MDPE_hm_ddm.pdf}
-% \caption{
-% The same simulations as presented in Figure \ref{fig:PE}, except with density dependent straying, where $m(t) = m_0\left(1-N(t)/(C+N(t)\right)$. Light colors = high values.
-% } \label{fig:ddm}
-% \end{figure*}
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+  \centering
+  \includegraphics[width=0.8\textwidth]{figs/fig_MDPE_hm_mtheta.pdf}
+  \caption{
+  Assuming that the rate of straying is linked directly to habitat heterogeneity.
+A low stray rate corresponds to very different (or distant) habitats (high $\Delta\theta$), whereas a higher rate of straying corresponds to very similar (or nearby) habitats (low $\Delta\theta$).
+Light colors = high values.
+  } \label{fig:mtheta}
+\end{figure*}
 
@@
 
-  \begin{figure*}
-    \centering
-    \includegraphics[width=0.8\textwidth]{figs/fig_MDPE_hm_mtheta.pdf}
-    \caption{
-    Assuming that the rate of straying is linked directly to habitat heterogeneity.
-A low stray rate corresponds to very different (or distant) habitats (high $\Delta\theta$), whereas a higher rate of straying corresponds to very similar (or nearby) habitats (low $\Delta\theta$).
-Light colors = high values.
-    } \label{fig:mtheta}
-  \end{figure*}
+\noindent{\bf (d) Linking stray rate and habitat heterogeneity}\\ 
+\noindent We have so far treated $\Delta\theta$ and $m$ as independent parameters, however we may also assume that if environmental heterogeneity increases with distance, the rate of straying may be expected to decline given that individuals are less likely to stray into distant habitats.
+%-- in particular North-South difference if trait optimality is largely temperature-dependent --
+If we incorporate this interdependence of $m$ and $\Delta\theta$, low rates of straying would correspond to mixing dissimilar (distant) populations, and high rates of straying would correspond to mixing similar (nearby) populations.
 
-{\bf Linking stray rate and habitat heterogeneity} We have so far treated $\Delta\theta$ and $m$ as independent parameters, however we may also assume that if environmental heterogeneity increases with distance -- in particular North-South difference if trait optimality is largely temperature-dependent -- the rate of straying may be expected to decline with distance.
-If we assume this interdependence of $m$ and $\Delta\theta$, low values of straying would correspond to mixing dissimilar (distant) populations, and high values of straying would correspond to mixing similar (nearby) populations.
+Here we find that alternative stable states now appear for low straying rates ($m<0.15$) and over any value of trait heritability $h^2<1$.
+As the straying rate increases, a single stable state emerges with a correspondingly high $N_T$.
+As before, there is a spike in the PE at the fold bifurcation separating the alternative stable state regime from the single stable state regime, and a sharp decline in the PE as the straying rate becomes very low.
+This is in accordance with intuition as increasing straying rates mean that two very similar populations are mixing, such that the rate of recruitment is less influenced by gene flow.
 
-In this case we find that alternative stable states now appear for low stray rates and low trait heritability, whereas increasing stray rates result in a single stable state with relatively high $N_T$.
-As before, there is a spike in the PE along the fold bifurcation separating the alternative stable state regime from the single stable state regime, but a sharp decline in the PE as stray rates become very low.
-This is in accordance with intuition as increasing stray rates mean that two very similar populations are mixing, resulting in little negative effect of trait dissimilarity.
+That alternative stable states appear and that PE becomes severely depressed for very low rates of straying is surprising: this means that even a small amount of mixing of populations from distant or dissimilar populations can qualitatively alter the dynamics of the metapopulation, regardless of trait heritability.
+%(for the discussion: an example of this situation may be the salmon populations during the last glacial maximum, where any mixing would be from geographically distant populations)
 
-However, that alternative stable states appear and that PE becomes severely depressed for very low values of PE is surprising: this means that even a small amount of straying of individuals from distant or dissimilar populations can qualitatively alter the dynamics of the metapopulation.
-(for the discussion: an example of this situation may be the salmon populations during the last glacial maximum, where any mixing would be from geographically distant populations)
+\section{Discussion}
 
-\section*{Discussion}
+We have shown that the natural selection of mixed populations adapted to different environments can result in large effects on population dynamics and the degree to which the metapopulation is buffered against extinction, quantified here as the portfolio effect.
+The immigration of dispersing strays with trait values far from the local optimum generally lowers the combined steady state biomass $N_T$ and increases the likelihood that the system gives way to alternative steady states, pushing one of the populations close to extinction.
+Although the emergence of alternative steady states can result in very large portfolio effects near the bifurcation, alternative steady state regimes tend to result in very low portfolio effects, indicating that the metapopulation is less likely to persist in the long run.
 
-\clearpage
+The degree to which the PE is lowered depends to a large extent on \emph{i}) trait heritability $h^2$ and \emph{ii}) habitat heterogeneity $\Delta \theta$.
+Increased heritability results in greater response to natural selection on the metapopulation, increasing the coupling between evolutionary and ecological dynamics. 
+Our minimal model of straying between two populations shows that greater $h^2$ buffers the metapopulation against the emergence of alternative stable states, but magnifies the negative effects of alternative stable states (captured by increased $\Delta N$ and decreased PE; Fig.
+\ref{fig:PE}B,C) once this regime is encountered at higher straying rates.
+We note that although the rate of straying has been shown to be density dependent (REFS), the qualitative results of our model are relatively insensitive to this dynamic (Fig.
+\ref{fig:diffddm}) and we limit discussion to the simplified case of a constant stray rate $m$ with the understanding that our findings also apply to the case density dependent stray rates $m(t)$.
+
+Trait heritability determines the coupling between ecological and evolutionary dynamics, effectively setting the strength of natural selection.
+Among salmon species, recruitment among local populations is highly sensitive to local climatic conditions, with temperature thought to play a central role (REFS).
+Populations distributed across a temperature gradient are assumed to be locally adapted to different temperature regimes, such that migration between them should take into account differential environmental effects on the recruitment of mixed populations.
+\citeauthor{Carlson:2008hl} showed heritability 0.1-0.3.
+Our results show that trait heritability has large effects on both metapopulation persistance as well as phenotypic diversity, particularly in the regime of low-to-moderate straying rates.
+
+%Our results in the context of Turing bifurcation
+[JP - could use some help here]
+The onset of alternative stable states in a spatial context describes the emergence of spatial pattern formation (REFS), which is more generally defined as a Turing instability (REFS).
+The mathematical conditions that lead to Turing instabilities are well-known in both continuous (REFS) and discrete spatial contexts (REFS)
+
+%North-South vs East-West
+
+\bibliography{aa_kevin}
+
+%\newpage
 \setcounter{figure}{0}
@@
 
-\begin{figure*}[h!]
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
 \centering

### 2017-05-17 model/mvsddm.jl

@@ -1,0 +1,309 @@
+using Distributions
+using RCall
+using HDF5
+using JLD
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/movingaverage.jl")
+
+
+mvec=collect(0.0:0.001:0.45);
+sigmavec = collect(0.1:0.1:3.0);
+hvec = collect(0.0:0.01:1.0);
+#Import constant m version
+d_m = load(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data/data_sig_h_m.jld"));
+#This loads the dictionary
+n1mean_m = d_m["n1mean"];
+n2mean_m = d_m["n2mean"];
+x1mean_m = d_m["x1mean"];
+x2mean_m = d_m["x2mean"];
+pe_m = d_m["pe"];
+pena_m = pe_m;
+pena_m[find(x->x==true,isnan(pe_m))] = 1;
+
+d_m3 = load(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data/data_sig_h_m_theta3.jld"));
+#This loads the dictionary
+n1mean_m3 = d_m3["n1mean"];
+n2mean_m3 = d_m3["n2mean"];
+x1mean_m3 = d_m3["x1mean"];
+x2mean_m3 = d_m3["x2mean"];
+pe_m3 = d_m3["pe"];
+pena_m3 = pe_m3;
+pena_m3[find(x->x==true,isnan(pe_m3))] = 1;
+
+
+d_m8 = load(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data/data_sig_h_m_theta8.jld"));
+#This loads the dictionary
+n1mean_m8 = d_m8["n1mean"];
+n2mean_m8 = d_m8["n2mean"];
+x1mean_m8 = d_m8["x1mean"];
+x2mean_m8 = d_m8["x2mean"];
+pe_m8 = d_m8["pe"];
+pena_m8 = pe_m8;
+pena_m8[find(x->x==true,isnan(pe_m8))] = 1;
+
+
+
+#Analysis over m & theta divergence
+indmvec=collect(0.0:0.001:0.45);
+C=1000;
+d_ddm = load(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data/data_sig_h_ddm.jld"));
+#This loads the dictionary
[…]

### 2017-05-17 model/src/movingaverage.jl

@@ -1,0 +1,19 @@
+function movingaverage(vec, window)
+  l = size(vec);
+  avgvec = zeros(Float64,l);
+  avgvec = avgvec[1:l[1]-window,:];
+  if length(l) > 1
+    for i = 1:l[2] #Over different elements of the timeseries
+      for t = 1:(l[1]-window)
+        avgvec[t,i] = mean(vec[t:t+window,i]);
+      end
+    end
+  else
+    for t = 1:(l[1]-window)
+      avgvec[t] = mean(vec[t:t+window]);
+    end
+  end
+  
+  
+  return(avgvec)
+end

### 2017-05-18 manuscript/msv2.tex
new pe results and extinction sims

--- 2017-05-17 Manuscript/msv1.tex
@@
 \centering
-\includegraphics[width=0.4\textwidth]{figs/fig_xkcd.png}
+\includegraphics[width=0.4\textwidth]{figs2/fig_xkcd.png}
 \caption{
@@
 \centering
-\includegraphics[width=0.4\textwidth]{figs/fig_Density.pdf}
+\includegraphics[width=0.4\textwidth]{figs2/fig_Density.pdf}
 \caption{
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figs/fig_MDPE_hm.pdf}
+\includegraphics[width=0.8\textwidth]{figs2/fig_MDPE_hm.pdf}
 \caption{
@@
 \centering
-\includegraphics[width=0.4\textwidth]{figs/fig_thetaPEmvm.pdf}
+\includegraphics[width=0.4\textwidth]{figs2/fig_thetaPEmvm.pdf}
 \caption{
@@
 \centering
-\includegraphics[width=0.4\textwidth]{figs/fig_traitdiff.pdf}
+\includegraphics[width=0.4\textwidth]{figs2/fig_traitdiff.pdf}
 \caption{
@@
 % \centering
-% \includegraphics[width=0.4\textwidth]{figs/fig_diffddm.pdf}
+% \includegraphics[width=0.4\textwidth]{figs2/fig_diffddm.pdf}
 % \caption{
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{figs/fig_MDPE_hm_mtheta.pdf}
+  \includegraphics[width=0.8\textwidth]{figs2/fig_MDPE_hm_mtheta.pdf}
   \caption{
@@
 \centering
-\includegraphics[width=\textwidth]{figs/fig_MDPE_hm_theta3.pdf} 
+\includegraphics[width=\textwidth]{figs2/fig_MDPE_hm_theta3.pdf} 
 \caption{Low habitat heterogeneity ($\Delta\theta=3$)} \label{fig:thetadiff1}
@@
 \centering
-\includegraphics[width=\textwidth]{figs/fig_MDPE_hm_theta8.pdf} 
+\includegraphics[width=\textwidth]{figs2/fig_MDPE_hm_theta8.pdf} 
 \caption{High habitat heterogeneity ($\Delta\theta=8$)} \label{fig:thetadiff2}
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figs/fig_MDPE_hm_ddm.pdf}
+\includegraphics[width=0.8\textwidth]{figs2/fig_MDPE_hm_ddm.pdf}
 \caption{

### 2017-05-18 model/KevinExtinction.jl

@@ -1,0 +1,76 @@
+using Distributions
+using RCall
+using JLD
+using HDF5
+
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve.jl")
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve_ddm.jl")
+
+
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolveExtinct.jl")
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/timeSS.jl")
+
+
+
+
+tmax=10000;
+z=0.5;
+rmax=2.0;
+beta=0.001;
+theta1=5.0;
+thetadiff=5.0;
+tau=1.0;
+h=0.5;
+sigmaE=0;
+sigmaG=1;
+perror=0.01;
+
+m=0.2
+
+extpop = "large";
+t_ext = Int64(round(tmax/2));
+
+n1, n2, x1, x2, w1, w2 = 
+KevinEvolveExtinct(tmax, 
+z, 
+rmax,
+beta,
+theta1,
+thetadiff,
+tau,
+h,
+sigmaE,
+sigmaG,
+m,
+perror,
+extpop,
+t_ext
+);
+
+
+R"""
+library(RColorBrewer)
+pal = brewer.pal(3,'Set1')
+par(mfrow=c(2,1))
+plot($n1,type='l',col=pal[1])
+lines($n2,col=pal[2])
+plot($x1,type='l',col=pal[1],ylim=c(4,12))
+lines($x2,col=pal[2])
+"""
+
+
+t_ss, relaxtime = timeSS(n1,n2,t_ext);
+
+
+R"""
+library(RColorBrewer)
+pal = brewer.pal(3,'Set1')
+plot($n1,type='l',col=pal[1],xlim=c($t_ext,$t_ss+50))
+lines($n2,col=pal[2])
+lines(c($t_ss,$t_ss),c(0,10000))
+"""
+
+
+
+#Calculate relaxation time as a function of 
+

### 2017-05-18 model/KevinSimulations2.jl

@@ -596,7 +596,7 @@ for i=1:length(mvec)
 end
 
 
-namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/figs/fig_traitdiff_theta3.pdf");
+namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/figs2/fig_traitdiff_theta3.pdf");
 traitdiff = abs(-1*(x1mean[:,:]-theta1) - (-1*(x2mean[:,:]-(theta1+thetadiff))));
 R"""
 pdf($namespace,height=4,width=5)

### 2017-05-18 model/KevinSimulations_ddm2.jl

@@ -277,20 +277,20 @@ dev.off()
 @everywhere using HDF5
 @everywhere using JLD
 
-@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve_ddm.jl")
 @everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/qualsfunc.jl")
 
 
 #Analysis over m & theta divergence
-mvec=collect(0.0:0.001:0.45);
+indmvec=collect(0.0:0.001:0.45);
 hvec = collect(0.0:0.01:1.0);
 
 
-n1mean=SharedArray(Float64,length(hvec),length(mvec));
-n2mean=SharedArray(Float64,length(hvec),length(mvec));
-x1mean=SharedArray(Float64,length(hvec),length(mvec));
-x2mean=SharedArray(Float64,length(hvec),length(mvec));
-pe=SharedArray(Float64,length(hvec),length(mvec));
+n1mean=SharedArray(Float64,length(hvec),length(indmvec));
+n2mean=SharedArray(Float64,length(hvec),length(indmvec));
+x1mean=SharedArray(Float64,length(hvec),length(indmvec));
+x2mean=SharedArray(Float64,length(hvec),length(indmvec));
+pe=SharedArray(Float64,length(hvec),length(indmvec));
 
 tmax=10000;
 z=0.5;
@@ -360,17 +360,17 @@ library(RColorBrewer)
 pal = rev(brewer.pal(9,"Blues"))
 pdf($namespace,height=3,width=8)
 par(mfrow=c(1,3))
-image(x=$mvec,y=$hvec,z=t($(n1mean[:,:]))+t($(n2mean[:,:])),zlim=c(0,3000),col=pal,xlab='m',ylab='h',main='Total biomass')
-image(x=$mvec,y=$hvec,z=sqrt((t($(n1mean[:,:]))-t($(n2mean[:,:])))^2),zlim=c(0,1500),col=pal,xlab='m',ylab='h',main='Biomass difference')
-image(x=$mvec,y=$hvec,z=t($(pe[:,:])),zlim=c(1,2),col=pal,xlab='m',ylab='h',main='PE')
[…]

### 2017-05-18 model/KevinSimulations_mtheta2.jl

@@ -1,0 +1,310 @@
+using Distributions
+using RCall
+
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve_mtheta.jl")
+
+#Analysis over m
+tmax=10000;
+mvec = collect(0.0:0.001:0.5);
+n1ts = zeros(Float64,length(mvec),tmax);
+n2ts = zeros(Float64,length(mvec),tmax);
+n1mean=zeros(Float64,length(mvec));
+n2mean=zeros(Float64,length(mvec));
+n1sd=zeros(Float64,length(mvec));
+n2sd=zeros(Float64,length(mvec));
+aggmean=zeros(Float64,length(mvec));
+aggsd=zeros(Float64,length(mvec));
+x1mean=zeros(Float64,length(mvec));
+x2mean=zeros(Float64,length(mvec));
+pe=zeros(Float64,length(mvec));
+eigs = Array(Array{Complex{Float64}},length(mvec));
+maxeigs = Array{Float64}(length(mvec));
+maximeigs = Array{Float64}(length(mvec));
+
+z=0.5;
+rmax=2.0;
+beta=0.001;
+theta1=5.0;
+thetascale=2;
+tau=1.0;
+h=0.2;
+sigmaE=0;
+sigmaG=1;
+perror=0.01;
+
+burnin=0.99
+@time for i=1:length(mvec)
+  m=mvec[i];
+  
+  n1, n2, x1, x2, w1, w2, thetadiff = 
+  KevinEvolve_mtheta(
+    tmax,
+    z,
+    rmax,
+    beta,
+    theta1,
+    thetascale,
+    tau,
+    h,
+    sigmaE,
+    sigmaG,
+    m,
+    perror
+    );
+
+  n1trim = n1[Int64(floor(tmax*burnin)):tmax];
+  n2trim = n2[Int64(floor(tmax*burnin)):tmax];
+  x1trim = x1[Int64(floor(tmax*burnin)):tmax];
+  x2trim = x2[Int64(floor(tmax*burnin)):tmax]
+  
+  n1ts[i,:] = n1;
+  n2ts[i,:] = n2;
+  
+  n1mean[i] = mean(n1trim);
+  n2mean[i] = mean(n2trim);
+  n1sd[i] = std(n1trim);
+  n2sd[i] = std(n2trim);
+  
+  aggmean[i] = mean(n1trim+n2trim);
+  aggsd[i] = std(n1trim+n2trim);
+
[…]

### 2017-05-18 model/mvsddm2.jl

@@ -1,0 +1,333 @@
+using Distributions
+using RCall
+using HDF5
+using JLD
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/movingaverage.jl")
+
+
+mvec=collect(0.0:0.001:0.45);
+sigmavec = collect(0.1:0.1:3.0);
+hvec = collect(0.0:0.01:1.0);
+#Import constant m version
+d_m = load(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data2/data_sig_h_m.jld"));
+#This loads the dictionary
+n1mean_m = d_m["n1mean"];
+n2mean_m = d_m["n2mean"];
+x1mean_m = d_m["x1mean"];
+x2mean_m = d_m["x2mean"];
+pe_m = d_m["pe"];
+pena_m = pe_m;
+pena_m[find(x->x==true,isnan(pe_m))] = 1;
+
+d_m3 = load(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data2/data_sig_h_m_theta3.jld"));
+#This loads the dictionary
+n1mean_m3 = d_m3["n1mean"];
+n2mean_m3 = d_m3["n2mean"];
+x1mean_m3 = d_m3["x1mean"];
+x2mean_m3 = d_m3["x2mean"];
+pe_m3 = d_m3["pe"];
+pena_m3 = pe_m3;
+pena_m3[find(x->x==true,isnan(pe_m3))] = 1;
+
+
+d_m8 = load(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data2/data_sig_h_m_theta8.jld"));
+#This loads the dictionary
+n1mean_m8 = d_m8["n1mean"];
+n2mean_m8 = d_m8["n2mean"];
+x1mean_m8 = d_m8["x1mean"];
+x2mean_m8 = d_m8["x2mean"];
+pe_m8 = d_m8["pe"];
+pena_m8 = pe_m8;
+pena_m8[find(x->x==true,isnan(pe_m8))] = 1;
+
+
+
+#Analysis over m & theta divergence
+indmvec=collect(0.0:0.001:0.45);
+C=1000;
+d_ddm = load(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data2/data_sig_h_m_ddm.jld"));
+#This loads the dictionary
[…]

### 2017-05-20 manuscript/msv2.tex
4panel plot and syncPE

@@
 
+\noindent{\bf (c) Rates of recovery following catastrophic collapse}\\ 
+(Results here - constant m)
+
+(Results here - density dependent m)
+
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.9\textwidth]{figs2/fig_relax_comb.pdf}
+\caption{
+Top row (a, b)
+Bottom row (c, d)
+} \label{fig:relax}
+\end{figure*}
+
 %% Increased habitat heterogeneity
@@
 
-\noindent{\bf (c) The role of habitat heterogeneity on the portfolio effect}\\ 
+\noindent{\bf (d) The role of habitat heterogeneity on the portfolio effect}\\ 
 \noindent Increased differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater between-site differences in conditions that favor alternative traits, interpreted here as increased habitat heterogeneity.
@@
 
-\noindent{\bf (d) Linking stray rate and habitat heterogeneity}\\ 
+\noindent{\bf (e) Linking stray rate and habitat heterogeneity}\\ 
 \noindent We have so far treated $\Delta\theta$ and $m$ as independent parameters, however we may also assume that if environmental heterogeneity increases with distance, the rate of straying may be expected to decline given that individuals are less likely to stray into distant habitats.

### 2017-05-20 manuscript/syncPE.tex
4panel plot and syncPE

\documentclass{article}

\begin{document}

\subsection{Synchronization and the Portfolio Effect}

In essence, synchronization of subpopulation trajectories over time represents an absence of dynamical diversity within the aggregate.
In contrast, the diversity of the aggregate can be described in terms of the number of subpopulations from which it is composed, and this is more generally referred to as metapopulation diversity.
These two forms of diversity are in fact closely related.
Loreau defined synchronization in terms of the sizes of fluctuations among the components of an aggregate, ${\rm VAR}(N_i)$, to the aggregate itself, ${\rm VAR}(N_T)$, which varies between 0 (not synchronized) and 1 (synchronized), such that

\begin{equation}
\label{eq_sync}
\phi = \frac{{\rm VAR}(N_T)}{\left(\sum_{i=1}^N\sqrt{{\rm VAR}(N_i)}\right)^2}.
\end{equation}

The diversity of the aggregate can defined with respect to the stability that such diversity confers.
The stability conferred by aggregation, which is often called `the portfolio effect' (PE), is defined by a relative decrease in the magnitude of fluctuations for the aggregate, with respect to the magnitude of fluctuations for the components of the aggregate.
Averaged over each subpopulation, the expected portfolio effect can be expressed as

\begin{equation}
\label{eq_PE}
\langle{\rm PE}\rangle =\frac{1}{X}\sum_{i=1}^X \frac{\sqrt{{\rm VAR}(N_i)}}{{\rm E}(N_i)}\cdot \frac{{\rm E}(N_T)}{\sqrt{{\rm VAR}(N_T)}},
\end{equation}

\noindent which is equivalently the average CV of the subpopulations divided by the CV of the aggregate.
Accordingly, as the CV of the aggregate decreases relative to that of the subpopulations, the stability of the aggregate increases, and $\langle{\rm PE}\rangle > 1$.
By rearranging and substituting eq. (\ref{eq_sync}) into eq. (\ref{eq_PE}), we can observe that $\langle{\rm PE}\rangle \propto \sqrt{\phi}^{-1}$.
Thus, the dynamical diversity of an aggregate offers a mirror to the stability conferred by aggregation, where perfect synchrony ($\phi = 1$) reflects the absence of a portfolio effect, or $\langle{\rm PE}\rangle = 1$.
Moreover, if and only if $\phi \neq 1$, it is expected that as the diversity of the aggregate increases, so should the portfolio effect, and this central thesis is thought to underlie the stability of populations, species, communities and even financial markets.

\end{document}

### 2017-05-20 model/src/KevinEvolve_mtheta_ddm.jl

@@ -84,6 +84,8 @@ function KevinEvolve_mtheta_ddm(
   x2,
   w1,
   w2,
+  m1,
+  m2,
   thetadiff
   )
   

### 2017-05-20 manuscript/msv2.tex
revised environ hetero fig

@@
 The emergence of alternative steady states results in a nonlinear response of the portfolio effect, suggesting that metapopulation persistence can be quite sensitive to the combined influences of collective straying across a selective landscape.
+
+%Extinction and return times
 
@@
 %, regardless of trait heritability and variance or habitat heterogeneity
-The decline in steady state densities is not gradual: as straying increases, the system crosses a fold bifurcation whereby the single steady state among both sites becomes two alternative steady states: one at high biomass density, and one at low biomass density (Fig.
-\ref{fig:traj}A,B).
-This bifurcation occurs at lower values of the stray rate $m$ for low trait heritability $h^2$ (Fig \ref{fig:PE}A), indicating that greater coupling between ecological and evolutionary dynamics in addition to greater rates of straying results in alternative stable states among the two sites.
+The decline in steady state densities is not gradual: as straying increases, the system crosses a fold bifurcation whereby the single steady state across both sites becomes two alternative steady states: one at high biomass density, and one at low biomass density (Fig.
+\ref{fig:traj}a,b).
+This bifurcation occurs at lower values of the stray rate $m$ for lower trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that lower coupling between ecological and evolutionary dynamics in addition to greater rates of straying guarantees the existence of alternative stable states.
 
-Trait heritability also has a large impact on the sensitivity of both the total steady state density ($N^*_T=N^*_1+N^*_2$) as well as the difference between steady state densities ($\Delta N=\sqrt{(N^*_1-N^*_2)^2}$).
+Trait heritability has a large impact on the sensitivity of both the total steady state density ($N^*_T=N^*_1+N^*_2$) as well as the difference between steady state densities ($\Delta N=\sqrt{(N^*_1-N^*_2)^2}$).
 Greater trait heritability results in a larger decline in $N_T^*$ for increasing stray rates $m$, though results in only moderate changes to $\Delta N$ (Figs. \ref{fig:PE}A,B).

### 2017-05-20 model/mvsddm2.jl

@@ -239,63 +239,35 @@ library(RColorBrewer)
 pal = brewer.pal(3,'Set1')
 pdf($namespace,height=8,width=5)
 par(mfrow=c(2,1),mai = c(0.8, 0.8, 0.1, 0.1))
-plot($(ma_mvec[mlist]),$(ma_medpe3[mlist]),col=pal[1],pch=16,ylim=c(1,3),xlab='m',ylab='PE',cex=0.5)
-points($(ma_mvec[mlist]),$(ma_medpe5[mlist]),col=pal[2],pch=16,cex=0.5)
-points($(ma_mvec[mlist]),$(ma_medpe8[mlist]),col=pal[3],pch=16,cex=0.5)
-
-lines(cbind(rep($bif5,2),c($(medpe5[bifsite5-1]),$(medpe5[bifsite5+1]))),col=pal[2],lty=2,lwd=2)
-lines(cbind(rep($bif8,2),c($(medpe8[bifsite8-1]),$(medpe8[bifsite8+1]))),col=pal[3],lty=2,lwd=2)
-
-legend(x=0.42,y=3,legend=c(3,5,8),col=pal,pch=22,xpd=TRUE,pt.bg=pal,cex=0.8, bty="n",title=expression(paste(Delta,theta)))
-xleft<-0.0;xright<-0.45;ybottom<-1;ytop<-3;
-#rect(xleft, ybottom, xright, ytop)
-
-
-###
-#[bifsite5[1]+1:length(ma_medpe5smooth)[1]]
-#[bifsite5[1]+1:length(ma_medpe5smooth)[1]]
-#[bifsite8[1]+1:length(ma_medpe8smooth)[1]]
-#[bifsite8[1]+1:length(ma_medpe8smooth)[1]]
-
-plot($(ma_mvecsmooth),$(ma_medpe5smooth),col=pal[2],type='l',xlim=c(xleft,xright),ylim=c(ybottom,ytop),xlab='m, m*',ylab='PE',lwd=3)
+plot($(ma_mvecsmooth),$(ma_medpe5smooth),col=pal[2],type='l',ylim=c(1,max($([ma_medpe3[mlist] ma_medpe5[mlist] ma_medpe8[mlist]]))),xlab='m',ylab='PE',lwd=3)
 lines($(ma_mvecsmooth),$(ma_medpe8smooth),col=pal[3],lwd=3)
 lines($(ma_mvecsmooth),$(ma_medpe3smooth),col=pal[1],lwd=3)
[…]

### 2017-05-20 manuscript/msv2.tex
added return time panel to 3panel plots

@@
 \ref{fig:traj}a,b).
-This bifurcation occurs at lower values of the stray rate $m$ for lower trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that lower coupling between ecological and evolutionary dynamics in addition to greater rates of straying guarantees the existence of alternative stable states.
+This bifurcation occurs at lower values of the stray rate $m$ for lower trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that lower coupling between ecological and evolutionary dynamics in addition to higher rates of straying guarantees the existence of alternative stable states.
 
-Trait heritability has a large impact on the sensitivity of both the total steady state density ($N^*_T=N^*_1+N^*_2$) as well as the difference between steady state densities ($\Delta N=\sqrt{(N^*_1-N^*_2)^2}$).
+Trait heritability has a large impact on the sensitivity of both the aggregate population steady state density ($N^*_T=N^*_1+N^*_2$) as well as the difference between steady state densities ($\Delta N=|N^*_1-N^*_2|$).
 Greater trait heritability results in a larger decline in $N_T^*$ for increasing stray rates $m$, though results in only moderate changes to $\Delta N$ (Figs. \ref{fig:PE}A,B).
 Conversely, if the trait is less heritable, an increase in the stray rate has little impact on the total biomass density and contrastingly large effects on the difference in population densities between sites (Fig.
-\ref{fig:PE}A,B).
+\ref{fig:PE}a,b).
 
-Together these changes in steady state population densities in terms of $N_T$ and $\Delta N$ as a function of trait heritability and the rate of straying between populations give rise to highly nonlinear portfolio effects, quantified here as
+Together these changes in steady state population densities in terms of $N_T$ and $\Delta N$ as a function of trait heritability and the rate of straying between populations give rise to highly nonlinear portfolio effects, which we quantify here as
 
 \begin{equation}
-  {\rm PE} = \frac{\overline{{\rm CV}}_{i,j}}{{\rm CV}_T}.
+\langle{\rm PE}\rangle =\frac{1}{X}\sum_{i=1}^{X} \frac{\sqrt{{\rm VAR}(N_i)}}{{\rm E}(N_i)}\cdot \frac{{\rm E}(N_T)}{\sqrt{{\rm VAR}(N_T)}},
 \label{eq:pe}
@@
 
-\noindent We define $\overline{{\rm CV}}_{i,j}$ as the average coefficient of variation across the local population $i$ and the remote population $j$ at steady state densities, and ${\rm CV}_T$ as the coefficient of variation of the total or aggregated population at steady state densities.
-The minimum portfolio effect is by definition ${\rm PE_{\rm min}}=1$, whereas portfolio effects greater than unity corresponds low synchronization \cite{Loreau:2008ju,Anderson:2014cx,Yeakel:2013vz} and a greater potential for demographic rescue among populations, thus buffering the system as a whole against extinction.
+\noindent where in our minimal model the number of populations is $X=2$ and the expectations $\rm E(\cdot)$ and variances $\rm VAR(\cdot)$ are evaluated at the steady state.
+The PE is equivalently the average CV of the subpopulations divided by the CV of the aggregate.
+Accordingly, as the CV of the aggregate decreases relative to that of the subpopulations, the stability of the aggregate increases, and $\langle{\rm PE}\rangle > 1$.
+A similar measure of metapopulation persistence can be evaluated by measuring the synchronization $\phi$ between populations, where 
+\begin{equation}
+\label{eq_sync}
+\phi = \frac{{\rm VAR}(N_T)}{\left(\sum_{i=1}^N\sqrt{{\rm VAR}(N_i)}\right)^2}.
+\end{equation}
+By rearranging and substituting eq. (\ref{eq_sync}) into eq. (\ref{eq_PE}), we can observe that $\langle{\rm PE}\rangle \propto \sqrt{\phi}^{-1}$.
+Thus, the dynamical diversity of an aggregate offers a mirror to the stability conferred by aggregation, where perfect synchrony ($\phi = 1$) reflects the absence of a portfolio effect, or $\langle{\rm PE}\rangle = 1$.
+%Moreover, if and only if $\phi \neq 1$, it is expected that as the diversity of the aggregate increases, so should the portfolio effect, and this central thesis is thought to underlie the stability of populations, species, communities and even financial markets.
+Portfolio effects greater than unity corresponds to less synchronization ($\phi < 1$) \cite{Loreau:2008ju,Anderson:2014cx,Yeakel:2013vz} and thus a greater potential for demographic rescue among populations, thus buffering the system as a whole against extinction.
 
-In the region where there is a single steady state among both populations, we find a correspondingly high portfolio effect, primarily due to the elevated mean values of $N_T$.
-As the fold bifurcation is approached with greater rates of straying, the portfolio effect spikes due to a large increase in the standard deviation of both populations.
-This explosion in variance is due to critical slowing down that leads to a change in variance near fold bifurcations, and this marked change in dynamic behavior has been suggested to serve as an early warning signal of an approaching phase transition \cite{Scheffer:2009gg,Anonymous:2013br,Dakos:2014br}.
-The detectibility of such changes in dynamical behavior among salmonids appears to be idosynchratic across species, and the difficulty in measuring critical slowing down may - ironically - be masked by large portfolio effects \cite{Krkosek:2014ch} [this might be more appropriate for the discussion].
-For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}C), where there is a high and low steady state density among the sites, if heritability is low the portfolio effect becomes minimized, is elevated and then declines as $m\rightarrow 0.5$.
-If heritability is high, the portfolio effect declines steadily after the explosion in variance associated with the development of alternative stable states.
+In the region where there is a single steady state among both populations, we find a portfolio effect ca. ${\rm PE}=1.4$, primarily due to the elevated mean values of $N_T$.
+As the fold bifurcation is approached with greater rates of straying, the portfolio effect spikes due to a large increase in ${\rm VAR}(N_{i,j})$.
+This increase in variance is due to critical slowing down that promotes an increase in variance near fold bifurcations, and this marked change in dynamic behavior has been suggested to serve as an early warning signal of an approaching phase transition \cite{Scheffer:2009gg,Anonymous:2013br,Dakos:2014br}.
+The detectability of such changes in dynamical behavior among salmonids appears to be idiosyncratic across species, and the difficulty in measuring critical slowing down may - ironically - be masked by large portfolio effects \cite{Krkosek:2014ch} [this might be more appropriate for the discussion].
+For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}), where there is a high and low steady state density among the sites, the portfolio effect declines steadily after the explosion in variance associated with the development of alternative stable states.
+This decline is more gradual as the straying rate increases if trait heritability is low, and steeper if trait heritability is high.
+%if heritability is low the portfolio effect becomes minimized, is elevated and then declines as $m\rightarrow 0.5$.
 % As the stray rate increases to the right of the bifurcation, increased trait heritability results in a steady decrease in the portfolio effect.
@@
 \centering
-\includegraphics[width=0.4\textwidth]{figs2/fig_thetaPEmvm.pdf}
+\includegraphics[width=0.5\textwidth]{figs2/fig_traitdiff.pdf}
 \caption{
-(a) Median portfolio effect as a function of stray rate $m$ for lower trait heritability ($h^2 < 0.5$).
-Vertical dashed lines represent the fold bifurcation, which occurs at lower stray rates with increased habitat heterogeneity $\Delta \theta$.
-(b) Within the box shown in (a), PE as a function of the constant straying rate is shown as the solid line.
-Paired open circles show PE as a function of a density dependent straying rate.
-Straying rates are paired because the recipient and donor populations exist in alternative steady states.
-The lower straying rate of a pair is for the larger population; the higher straying rate is for the smaller population.
-The different steady state regimes that give rise to the range of straying rates were obtained by varying the individual straying probability $m_0$ (where $m_0$ and $m^*$ are linearly related).
-} \label{fig:thetaPE}
+Phenotypic diversity ($\Delta \mu$) evaluated at the steady state as a function of straying rate $m$ and trait heritability $h^2$.
+The descrete jump occurs as the system crosses the fold bifurcation; lower phenoytic diversity emerges with higher straying rates and in the alternative steady state regime.
+} \label{fig:traitdiff}
 \end{figure}
@@
 \centering
-\includegraphics[width=0.4\textwidth]{figs2/fig_traitdiff.pdf}
+\includegraphics[width=0.4\textwidth]{figs2/fig_thetaPEmvm.pdf}
 \caption{
-Phenotypic diversity ($\Delta \mu$) evaluated at the steady state as a function of straying rate $m$ and trait heritability $h^2$.
-The descrete jump occurs as the system crosses the fold bifurcation; lower phenoytic diversity emerges with higher straying rates and in the alternative steady state regime.
-} \label{fig:traitdiff}
+(a) Median portfolio effect as a function of stray rate $m$ for lower trait heritability ($h^2 < 0.5$).
+Vertical dashed lines represent the fold bifurcation, which occurs at lower stray rates with increased habitat heterogeneity $\Delta \theta$.
+(b) Within the box shown in (a), PE as a function of the constant straying rate is shown as the solid line.
+Paired open circles show PE as a function of a density dependent straying rate.
+Straying rates are paired because the recipient and donor populations exist in alternative steady states.
+The lower straying rate of a pair is for the larger population; the higher straying rate is for the smaller population.
+The different steady state regimes that give rise to the range of straying rates were obtained by varying the individual straying probability $m_0$ (where $m_0$ and $m^*$ are linearly related).
+} \label{fig:thetaPE}
 \end{figure}

### 2017-05-20 model/KevinSimulations2.jl

@@ -83,7 +83,7 @@ beta=0.001;
 theta1=5.0;
 thetadiff=5;
 tau=1.0;
-h=0.8;
+h=0.5;
 sigmaE=0;
 sigmaG=1;
 perror=0.01;
@@ -188,8 +188,36 @@ R"""
 plot($mvec,$pe)
 """
 
+R"""
+plot($n1mean[1:500],($n1sd[1:500])^2,log='xy')
+points($n2mean[1:500],($n2sd[1:500])^2)
+"""
 
 
+#Portfolio effect plot
+R"""
+plot($mvec,$n1sd/$n1mean)
+points($mvec,$n2sd/$n2mean,col='blue')
+points($mvec,$aggsd/$aggmean,pch=16)
+"""
+n1trim = n1mean[200:500];
+n1sdtrim = n1sd[200:500];
+lmean = n1trim[find(x->x<3.1,n1sdtrim)];
+lsd = n1sdtrim[find(x->x<3.1,n1sdtrim)];
+umean = n1trim[find(x->x>3.1,n1sdtrim)];
+usd = n1sdtrim[find(x->x>3.1,n1sdtrim)];
+
+R"""
+plot(log($(lmean)),($lsd))
+m1 = lm($lsd ~ log($lmean));
+summary(m1)
+"""
+
+R"""
+plot(log($(umean)),($usd))
+m1 = lm($usd ~ log($umean));
+summary(m1)
+"""
 
 R"""
 par(mfrow=c(2,1))
@@ -198,7 +226,7 @@ points($mvec,$n1mean,col='blue')
 points($mvec,$n2mean)
 """
 R"""
-plot($mvec,$n1sd)
+plot($mvec,$n1sd,ylim=c(0,7))
 points($mvec,$n2sd,col='blue')
 points($mvec,$aggsd,pch=16)
 """
@@ -455,7 +483,7 @@ end
 save(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data2/data_sig_h_m_theta8.jld"),"n1mean",n1mean,"n2mean",n2mean,"x1mean",x1mean,"x2mean",x2mean,"pe",pe);
 
 
-d = load(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data/data2_sig_h_m_theta8.jld"));
[…]

### 2017-05-20 model/src/timeSS.jl

@@ -3,12 +3,14 @@ function timeSS(n1,n2,tstart)
   #Who went extinct? Calculate time to steady state for this one.
   
   if (n1[tstart] > 0) && (n2[tstart] > 0)
-    n = n1 + n2;
+    n = n1 + n2; # when does the aggregate return to ss?
   else
     if n1[tstart] == 0;
-      n = n1;
+      # n = n1;
+      n = n1 + n2; # when does the aggregate return to ss?
     else
-      n = n2;
+      # n = n2;
+      n = n1 + n2; # when does the aggregate return to ss?
     end
   end
   

### 2017-05-21 Manuscript/msv2.tex
older stuff

--- 2017-05-20 manuscript/msv2.tex
@@
 
+{\bf Measuring of metapopulation persistence}
+We evaluated metapopulation persistence by measuring the average-CV portfolio effect as well as the time it took the system to return to a steady state following an induced disturbance to one or both of the populations.
+
+\begin{equation}
+\langle{\rm PE}\rangle =\frac{1}{X}\sum_{i=1}^{X} \frac{\sqrt{{\rm VAR}(N_i)}}{{\rm E}(N_i)}\cdot \frac{{\rm E}(N_T)}{\sqrt{{\rm VAR}(N_T)}},
+\label{eq:pe}
+\end{equation}
+
+\noindent where in our minimal model the number of populations is $X=2$ and the expectations $\rm E(\cdot)$ and variances $\rm VAR(\cdot)$ are evaluated at the steady state.
+The PE is equivalently the average CV of the subpopulations divided by the CV of the aggregate.
+Accordingly, as the CV of the aggregate decreases relative to that of the subpopulations, the stability of the aggregate increases, and $\langle{\rm PE}\rangle > 1$.
+A similar measure of metapopulation persistence can be evaluated by measuring the synchronization $\phi$ between populations, where 
+\begin{equation}
+\label{eq_sync}
+\phi = \frac{{\rm VAR}(N_T)}{\left(\sum_{i=1}^N\sqrt{{\rm VAR}(N_i)}\right)^2}.
+\end{equation}
+By rearranging and substituting eq. (\ref{eq_sync}) into eq. (\ref{eq_PE}), we can observe that $\langle{\rm PE}\rangle \propto \sqrt{\phi}^{-1}$.
+Thus, the dynamical diversity of an aggregate offers a mirror to the stability conferred by aggregation, where perfect synchrony ($\phi = 1$) reflects the absence of a portfolio effect, or $\langle{\rm PE}\rangle = 1$.
+%Moreover, if and only if $\phi \neq 1$, it is expected that as the diversity of the aggregate increases, so should the portfolio effect, and this central thesis is thought to underlie the stability of populations, species, communities and even financial markets.
+Portfolio effects greater than unity corresponds to less synchronization ($\phi < 1$) \cite{Loreau:2008ju,Anderson:2014cx,Yeakel:2013vz} and thus a greater potential for demographic rescue among populations, thus buffering the system as a whole against extinction.
+
 %Density dependent m
@@
 
-Together these changes in steady state population densities in terms of $N_T$ and $\Delta N$ as a function of trait heritability and the rate of straying between populations give rise to highly nonlinear portfolio effects, which we quantify here as
+%Together these changes in steady state population densities in terms of $N_T$ and $\Delta N$ as a function of trait heritability and the rate of straying between populations give rise to highly nonlinear portfolio effects, which we quantify here as
 
-\begin{equation}
-\langle{\rm PE}\rangle =\frac{1}{X}\sum_{i=1}^{X} \frac{\sqrt{{\rm VAR}(N_i)}}{{\rm E}(N_i)}\cdot \frac{{\rm E}(N_T)}{\sqrt{{\rm VAR}(N_T)}},
-\label{eq:pe}
-\end{equation}
-
-\noindent where in our minimal model the number of populations is $X=2$ and the expectations $\rm E(\cdot)$ and variances $\rm VAR(\cdot)$ are evaluated at the steady state.
-The PE is equivalently the average CV of the subpopulations divided by the CV of the aggregate.
-Accordingly, as the CV of the aggregate decreases relative to that of the subpopulations, the stability of the aggregate increases, and $\langle{\rm PE}\rangle > 1$.
-A similar measure of metapopulation persistence can be evaluated by measuring the synchronization $\phi$ between populations, where 
-\begin{equation}
-\label{eq_sync}
-\phi = \frac{{\rm VAR}(N_T)}{\left(\sum_{i=1}^N\sqrt{{\rm VAR}(N_i)}\right)^2}.
-\end{equation}
-By rearranging and substituting eq. (\ref{eq_sync}) into eq. (\ref{eq_PE}), we can observe that $\langle{\rm PE}\rangle \propto \sqrt{\phi}^{-1}$.
-Thus, the dynamical diversity of an aggregate offers a mirror to the stability conferred by aggregation, where perfect synchrony ($\phi = 1$) reflects the absence of a portfolio effect, or $\langle{\rm PE}\rangle = 1$.
-%Moreover, if and only if $\phi \neq 1$, it is expected that as the diversity of the aggregate increases, so should the portfolio effect, and this central thesis is thought to underlie the stability of populations, species, communities and even financial markets.
-Portfolio effects greater than unity corresponds to less synchronization ($\phi < 1$) \cite{Loreau:2008ju,Anderson:2014cx,Yeakel:2013vz} and thus a greater potential for demographic rescue among populations, thus buffering the system as a whole against extinction.
+Nonlinear PE(!!!)
 
@@
 
-\noindent{\bf (c) Rates of recovery following catastrophic collapse}\\ 
-(Results here - constant m)
-
-(Results here - density dependent m)
-
 \begin{figure*}
@@
 \end{figure*}
+
+\noindent{\bf (c) Rates of recovery following catastrophic collapse}\\ 
+% Return time as a measure of system persistence and relation to PE
+We measured the rate of return to a steady state following three types of induced disturbance: (\emph{i}) extinction of the low-density population, (\emph{ii}) extinction of the high-density population; (\emph{iii}) near-collapse of both populations.
+
+%(Results here - constant m)
+
+%(Results here - density dependent m)
 
@@
 
-%\newpage
+\clearpage
 \setcounter{figure}{0}
@@
 
-\begin{figure*}
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.5\textwidth]{figs2/fig_rtvspe.pdf}
+\caption{
+Return time vs. portfolio effect.
+} \label{fig:rtvspe}
+\end{figure}
+
+\begin{figure*}[h!]
   \captionsetup{justification=raggedright,
@@
 
-\begin{figure*}
+\begin{figure*}[h!]
   \captionsetup{justification=raggedright,

### 2017-05-22 Manuscript/msv2.tex
bifurcation detection

@@
 
-Here we consider two populations $N_1$ and $N_2$ that are separated in space in distinct habitats, each with trait values $x_1$ and $x_2$ determining recruitment rates.
+\noindent{\bf (a) Metapopulation framework}\\
+\noindent Here we consider two populations $N_1$ and $N_2$ that are separated in space in distinct habitats, each with trait values $x_1$ and $x_2$ determining recruitment rates.
 We assume that there is an optimum trait value $\theta_1$ and $\theta_2$ associated with each habitat, where recruitment is maximized if the trait value of the local population $x = \theta$.
@@
   &\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z} \\ \nonumber
-  &+ \left(R_i[\mu_i(t)|\theta_i] (1-m)N_i(t) + R_i[\mu_j(t)|\theta_i] m N_j(t)\right) \\ \nonumber
+  &+ \left(R_i[\mu_i(t)] (1-m)N_i(t) + R_i[\mu_j(t)] m N_j(t)\right) \\ \nonumber
   &\times {\rm e}^{-\beta ((1-m)N_i(t) + m N_j(t))},
@@
 \begin{align}
-  &R_i[\mu_i(t)|\theta_i] = \\ \nonumber
-  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i(t)-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i(t)|\mu_i,\sigma^2) {\rm d}x_i(t) \\ \nonumber
-  &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-\mu_i(t))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\}.
+  &R_i[\mu_i(t)] = \\ \nonumber
+  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i(t)-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i(t)|\mu_i,\sigma^2) {\rm d}x_i(t) +\tilde{P}\\ \nonumber
+  &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-\mu_i(t))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\} +\tilde{P},
   \label{eq:R}
@@
 
-\noindent Again, it is the mismatch between the local trait mean $\mu_i(t)$ and the local optimum $\theta_i$ that determines the recruitment rate for the population.
+\noindent where the mismatch between the local trait mean $\mu_i(t)$ and the local optimum $\theta_i$ determines the recruitment rate for the population, and $\tilde{P}$ introduces a small amount of demographic error $\tilde{P}\sim {\rm Normal}(0,0.01)$.
 The parameter $\tau$ controls the sensitivity of recruitment to changes in the mean trait value away from the optimum, which we set as $\tau=1$ here and throughout.
 % The compensatory effects are then determined by the exponential, following the Ricker stock-recruitment relationship.
+\\
 
@@
 
-Because individuals from the local population are mixed with individuals from the remote population via staying, the resulting trait distribution is a mixed normal with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and for the straying individuals, $1-w_i$, where
+\noindent{\bf (b) Recruitment over a selective landscape}
+\noindent Because individuals from the local population are mixed with individuals from the remote population via staying, the resulting trait distribution is a mixed normal with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and for the straying individuals, $1-w_i$, where
 \begin{equation}
@@
   \mu_i(t+1) &= w_i\mu_i(t) + (1-w_i)\mu_j(t) \\ \nonumber
-  &+ \frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)|\theta_i] + (1-w_i)R_i[\mu_j(t)|\theta_i]  \right),
+  &+ h^2\sigma\frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)|\theta_i] + (1-w_i)R_i[\mu_j(t)|\theta_i]  \right),
   \label{eq:mu}
@@
 \ref{eq:mu} determines how the mean trait changes through time due to natural selection (REF), which is proportional to the change in mean fitness with respect to $\mu_i$.
+\\
 
-{\bf Measuring of metapopulation persistence}
-We evaluated metapopulation persistence by measuring the average-CV portfolio effect as well as the time it took the system to return to a steady state following an induced disturbance to one or both of the populations.
+\noindent {\bf (c) Measuring of metapopulation persistence}
+\noindent We evaluated metapopulation persistence by measuring the average-CV portfolio effect (PE) (REF: Anderson,others) as well as the time required for the system to return to a steady state following an induced disturbance to one or both of the populations.
+The average-CV portfolio effect is the average CV across each population divided by the CV of the aggregate, such that
 
@@
 
-\noindent where in our minimal model the number of populations is $X=2$ and the expectations $\rm E(\cdot)$ and variances $\rm VAR(\cdot)$ are evaluated at the steady state.
-The PE is equivalently the average CV of the subpopulations divided by the CV of the aggregate.
-Accordingly, as the CV of the aggregate decreases relative to that of the subpopulations, the stability of the aggregate increases, and $\langle{\rm PE}\rangle > 1$.
+\noindent where in this case the number of populations is limited to $X=2$ and the expectations $\rm E(\cdot)$ and variances $\rm VAR(\cdot)$ are evaluated at the steady state.
+As the CV of $N_T$ decreases relative to that of the constituent populations, $\langle{\rm PE}\rangle > 1$, and the metapopulation is presumed to become more stable.
 A similar measure of metapopulation persistence can be evaluated by measuring the synchronization $\phi$ between populations, where 
@@
 By rearranging and substituting eq. (\ref{eq_sync}) into eq. (\ref{eq_PE}), we can observe that $\langle{\rm PE}\rangle \propto \sqrt{\phi}^{-1}$.
-Thus, the dynamical diversity of an aggregate offers a mirror to the stability conferred by aggregation, where perfect synchrony ($\phi = 1$) reflects the absence of a portfolio effect, or $\langle{\rm PE}\rangle = 1$.
+Thus, the dynamical diversity of the metapopulation offers a mirror to the portfolio effect, where perfect synchrony ($\phi = 1$) also means $\langle{\rm PE}\rangle = 1$.
 %Moreover, if and only if $\phi \neq 1$, it is expected that as the diversity of the aggregate increases, so should the portfolio effect, and this central thesis is thought to underlie the stability of populations, species, communities and even financial markets.
@@
 
+%Although there are more robust ways to measure the portfolio effect (in particular for comparing the PE across species with different mean-variance relationships; REF: Anderson), its only utlity here is to provide a simple statistical measure of metapopulation persistence.
+A more direct way to measure the likelihood that a system will persist is to measure the time it takes to recover following an induced disturbance: systems that recover quickly (lower recovery times) are more robust than those that recover more slowly (higher recovery times).
+Measuring the time it takes for a perturbed system to relax to a steady state also permits a more detailed examination of the potential fragility of the metapopulation.
+For example, if populations settle into alternative steady states (alternative steady states in our model requiring one population to be high-density and one low-density), comparing recovery times after a disturbance applied to the high, low, and/or both populations permits an understanding of which component of the metapopulation has a stronger influence on persistence.
+\\
+
+\noindent {\bf (d) Density dependent straying rates and the effects of distance}
 %Density dependent m
-We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is good evidence that at least in some species the stray rate is density dependent.
+\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is good evidence that at least in some species the stray rate is density dependent.
 Specifically, the rate at which individuals stray has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tends to decrease the rate at which individuals stray, thus reducing the overall proportion of a population that strays.
@@
 
-\noindent{\bf (a) Nonlinear effects of straying on the Portfolio Effect} \\
+\noindent{\bf (a) Nonlinear effects of straying on the portfolio effect and return time} \\
 \noindent Straying generally lowers steady state densities for both recipient and donor populations by \emph{i}) losing locally-adapted individuals due to straying and \emph{ii}) introducing maladapted individuals from the donor population.
 %, regardless of trait heritability and variance or habitat heterogeneity
-The decline in steady state densities is not gradual: as straying increases, the system crosses a fold bifurcation whereby the single steady state across both sites becomes two alternative steady states: one at high biomass density, and one at low biomass density (Fig.
+The decline in steady state densities is not gradual: as straying increases, the system crosses a fold bifurcation whereby the single steady state across both sites bifurcates into two alternative steady states: one at high biomass density, and one at low biomass density (Fig.
 \ref{fig:traj}a,b).
-This bifurcation occurs at lower values of the stray rate $m$ for lower trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that lower coupling between ecological and evolutionary dynamics in addition to higher rates of straying guarantees the existence of alternative stable states.
+This bifurcation occurs at lower values of the straying rate $m$ at lower values of trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the existence of alternative stable states.
 
-Trait heritability has a large impact on the sensitivity of both the aggregate population steady state density ($N^*_T=N^*_1+N^*_2$) as well as the difference between steady state densities ($\Delta N=|N^*_1-N^*_2|$).
-Greater trait heritability results in a larger decline in $N_T^*$ for increasing stray rates $m$, though results in only moderate changes to $\Delta N$ (Figs. \ref{fig:PE}A,B).
-Conversely, if the trait is less heritable, an increase in the stray rate has little impact on the total biomass density and contrastingly large effects on the difference in population densities between sites (Fig.
+Trait heritability has a large impact on the sensitivity of the straying rate on both the aggregate population steady state density ($N^*_T=N^*_1+N^*_2$) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N=|N^*_1-N^*_2|$).
+Greater trait heritability results in a larger decline in $N_T^*$ for increasing straying rates $m$, though results in only moderate changes to $\Delta N$ (Figs. \ref{fig:PE}A,B).
+Conversely, if the trait is has a lower heritability, an increase in the straying rate has little impact on the total biomass density and contrastingly large effects on $\Delta N$ (Fig.
 \ref{fig:PE}a,b).
@@
 
-Nonlinear PE(!!!)
+% Nonlinear PE(!!!)
 
-In the region where there is a single steady state among both populations, we find a portfolio effect ca. ${\rm PE}=1.4$, primarily due to the elevated mean values of $N_T$.
-As the fold bifurcation is approached with greater rates of straying, the portfolio effect spikes due to a large increase in ${\rm VAR}(N_{i,j})$.
-This increase in variance is due to critical slowing down that promotes an increase in variance near fold bifurcations, and this marked change in dynamic behavior has been suggested to serve as an early warning signal of an approaching phase transition \cite{Scheffer:2009gg,Anonymous:2013br,Dakos:2014br}.
-The detectability of such changes in dynamical behavior among salmonids appears to be idiosyncratic across species, and the difficulty in measuring critical slowing down may - ironically - be masked by large portfolio effects \cite{Krkosek:2014ch} [this might be more appropriate for the discussion].
-For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}), where there is a high and low steady state density among the sites, the portfolio effect declines steadily after the explosion in variance associated with the development of alternative stable states.
+In the region where there is a single steady state among both populations, we find an low-intermediate portfolio effect ca. ${\rm PE}=1.4$, primarily due to the elevated mean values of $N_T$.
+As the fold bifurcation is reached with greater rates of straying, the portfolio effect explodes due to a large increase in the variance of both donor and recipient populations ${\rm VAR}(N_{i,j})$.
+This increase in variance is the product of a dynamical process known as critical slowing down that occurs near fold bifurcations (among others) \cite{Scheffer:2009gg}, and has been suggested to serve as an early warning signal of an approaching phase transition \cite{Scheffer:2009gg,Anonymous:2013br,Dakos:2014br}.
+For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}), where there alternative steady states among the recipient and donor sites, the portfolio effect declines steadily after the explosion in variance associated with onset of the fold bifurcation.
 This decline is more gradual as the straying rate increases if trait heritability is low, and steeper if trait heritability is high.
-%if heritability is low the portfolio effect becomes minimized, is elevated and then declines as $m\rightarrow 0.5$.
-% As the stray rate increases to the right of the bifurcation, increased trait heritability results in a steady decrease in the portfolio effect.
-% If trait heritability is low, the portfolio effect is nonlinear over increasing stray rates: after the spike associated with the fold bifurcation, {\rm PE} first increases at intermediate stray rates and then declines as $m\rightarrow0.5$.
+
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.9\textwidth]{figs2/fig_relax_comb.pdf}
+\caption{
+Top row (a, b)
+Bottom row (c, d)
+} \label{fig:relax}
+\end{figure*}
+
+% \noindent{\bf (c) Rates of recovery following catastrophic collapse}\\ 
+% Return time as a measure of system persistence and relation to PE
+We measured the rate of recovery to a steady state as a function of the rate of straying $m$ following three types of induced disturbance: (\emph{i}) extinction of the low-density population, (\emph{ii}) extinction of the high-density population; (\emph{iii}) near-collapse of both populations where just 1.0\% of both populations survive.
+At low rates of straying both populations are subject to the same steady state, such that extinction of either population results in the same recovery time, which is generally less than that required after a near-collapse of both populations.
+For values of $m$ that give rise to alternative steady states, the recovery times associated with each disturbance type depends strongly on trait heritability.
+If both straying rate and heritability are low (Fig.
+\ref{fig:relax}a), extinction of the small population results in a lower recovery time, whereas as the rate of straying increases, the extinction of the larger population results is lower recovery times.
+This switch is in part due to a large burst in biomass of the surviving (large) population when the smaller of the two becomes extinct at high straying rates, which extends the time required for the smaller population to recover.
+Conversely, with increased trait heritability (Fig.
+\ref{fig:relax}b) and low straying rates in the single steady state regime, near collapse of both populations (rather than the extinction of either) results in the lowest recovery time.
+Near the onset of the fold bifurcations, recovery times increase drastically, regardless of disturbance.
+For straying rates in the alternative steady state regime, extinction of the smaller population consistently provides the least time to recovery.
 
@@
 %Lowering Phenotypic diversity with increasing straying rate
-As expected, an increased straying rate lowers phenotypic diversity ($\Delta \mu = |\mu_i(t)-\mu_j(t)|$, evaluated at the steady state) because both local and remote populations become homogenized.
-The sensitivity of this phenotypic homogenization to straying is increased with lower heritability because the selective forces acting against trait means far from the local optima is lessened for lower $h^2$.
+Increased rates of straying lowers phenotypic diversity ($\Delta \mu = |\mu_i(t)-\mu_j(t)|$, evaluated at the steady state) because both local and remote populations are increasingly homogenized.
+The sensitivity of this phenotypic homogenization to straying is increased with lower heritability because the selective forces acting against trait means far from the local optima are lessened for lower $h^2$.
 Less intuitively, we observe a discrete jump towards low phenotypic diversity at low-moderate ($0.05 \leq m < 0.2$) straying rates, which occurs as the consequence of crossing the fold bifurcation (Fig.
 \ref{fig:traitdiff}).
-\\
-
-% Density dependent m
-%% Because m[t] @ steady state 0<m[t]<m0, we observe the same dynamics for higher m0 as we do for lower constant m.
-%% Higher rates of straying are tolerable (there is a not a decline in PE for very high m0 as there is for constant m)
-
-\noindent{\bf (b) The effects of density dependent straying}\\
-\noindent If we assume that the rate of straying is density dependent, the probability of straying at the individual level $m_0$ determines the rate of straying within the population, such that $m(t)$ becomes lower as $N(t)$ increases (Eq.
-\ref{eq:ddm}).
-At the steady state $m(t)=m^*$ is also constant, though by definition, it is always less than $m_0$ such that $0 < m^* < m_0$.
-Because our analyses concern steady state conditions, the inclusion of density-dependent straying does not have a large impact on the qualitative nature of our findings (Appendix Fig. \ref{fig:ddm}).
-
-The assumption that straying is density dependent alters the strength of the portfolio effect for a given $m^*$.
-As shown in Figure \ref{fig:thetaPE}(b), in the alternative stable state regime, because the recipient and donor population exist at different steady state densities, they have different rates of straying such that a given value of the portfolio effect is now associated with a pair of straying rates $(m_i^*,m_j^*)$.
-A range of $(m_i^*,m_j^*)$ values are obtained by varying $m_0$, which is positively and linearly related to $(m_i^*,m_j^*)$.
-Importantly, we show that metapopulations settling at relatively low rates of straying have lower-than-expected portfolio effects, whereas metapopulations settling at relatively high rates of straying have higher-than expected portfolio effects.
-This suggests that there is a strong tradeoff in the role of density dependent straying on the persistence of metapopulations: as the individual probability of straying is lowered, the straying rates of both recipient and donor populations are similarly lowered, however this diminishes the system's ability to buffer itself against extinction as measured by the portfolio effect.
-Conversely, as the probability that an individual strays increases, so does rate of straying for both recipient and donor populations, and this serves to elevate the portfolio effect compared to a constant $m$.
-This effect remains but is somewhat diminished as habitats become increasingly heterogeneous.
+Although the development of alternative stable states elevates the portfolio effect due to the variance-dampening effects of the aggregate, this change in dynamic regimes is also associated with a substantial decline in phenotypic diversity.
 \\
@@
 
-\begin{figure*}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.9\textwidth]{figs2/fig_relax_comb.pdf}
-\caption{
-Top row (a, b)
-Bottom row (c, d)
-} \label{fig:relax}
-\end{figure*}
-
-\noindent{\bf (c) Rates of recovery following catastrophic collapse}\\ 
-% Return time as a measure of system persistence and relation to PE
-We measured the rate of return to a steady state following three types of induced disturbance: (\emph{i}) extinction of the low-density population, (\emph{ii}) extinction of the high-density population; (\emph{iii}) near-collapse of both populations.
-
-%(Results here - constant m)
-
-%(Results here - density dependent m)
-
 %% Increased habitat heterogeneity
@@
 
-\noindent{\bf (d) The role of habitat heterogeneity on the portfolio effect}\\ 
+\noindent{\bf (b) The role of habitat heterogeneity on the portfolio effect}\\ 
 \noindent Increased differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater between-site differences in conditions that favor alternative traits, interpreted here as increased habitat heterogeneity.
@@
 
-% 
-% \begin{figure}
-%   \captionsetup{justification=raggedright,
-% singlelinecheck=false
-% }
-% \centering
-% \includegraphics[width=0.4\textwidth]{figs2/fig_diffddm.pdf}
-% \caption{
-% Difference in PE for simulations with a constant stray rate $m$ and PE for simulations with a density dependent stray rate $m(t)$.
-The differences are minimal (mean difference in PE = -0.05) and result in no discernible qualitative difference in dynamics, though density dependent stray rates generally amplify PE.
-% } \label{fig:diffddm}
-% \end{figure}
+% Density dependent m
+%% Because m[t] @ steady state 0<m[t]<m0, we observe the same dynamics for higher m0 as we do for lower constant m.
+%% Higher rates of straying are tolerable (there is a not a decline in PE for very high m0 as there is for constant m)
+
+\noindent{\bf (c) The effects of density dependent straying}\\
+\noindent If we assume that the rate of straying is density dependent, the probability of straying at the individual level $m_0$ determines the rate of straying within the population, such that $m(t)$ becomes lower as $N(t)$ increases (Eq.
+\ref{eq:ddm}).
+At the steady state $m(t)=m^*$ is also constant, though by definition, it is always less than $m_0$ such that $0 < m^* < m_0$.
+Because our analyses concern steady state conditions, the inclusion of density-dependent straying does not have a large impact on the qualitative nature of our findings (Appendix Fig. \ref{fig:ddm}).
+
+The assumption that straying is density dependent alters the strength of the portfolio effect for a given $m^*$.
+As shown in Figure \ref{fig:thetaPE}(b), in the alternative stable state regime, because the recipient and donor population exist at different steady state densities, they have different rates of straying such that a given value of the portfolio effect is now associated with a pair of straying rates $(m_i^*,m_j^*)$.
+A range of $(m_i^*,m_j^*)$ values are obtained by varying $m_0$, which is positively and linearly related to $(m_i^*,m_j^*)$.
+Importantly, we show that metapopulations settling at relatively low rates of straying have lower-than-expected portfolio effects, whereas metapopulations settling at relatively high rates of straying have higher-than expected portfolio effects.
+This suggests that there is a strong tradeoff in the role of density dependent straying on the persistence of metapopulations: as the individual probability of straying is lowered, the straying rates of both recipient and donor populations are similarly lowered, however this diminishes the system's ability to buffer itself against extinction as measured by the portfolio effect.
+Conversely, as the probability that an individual strays increases, so does rate of straying for both recipient and donor populations, and this serves to elevate the portfolio effect compared to a constant $m$.
+This effect remains but is somewhat diminished as habitats become increasingly heterogeneous.
+
+Return times (Fig. \ref{fig:relax}c,d).
+\\
 
@@
 
-\noindent{\bf (e) Linking stray rate and habitat heterogeneity}\\ 
+\noindent{\bf (d) Distance dependent straying and habitat heterogeneity}\\ 
 \noindent We have so far treated $\Delta\theta$ and $m$ as independent parameters, however we may also assume that if environmental heterogeneity increases with distance, the rate of straying may be expected to decline given that individuals are less likely to stray into distant habitats.
@@
 Although the emergence of alternative steady states can result in very large portfolio effects near the bifurcation, alternative steady state regimes tend to result in very low portfolio effects, indicating that the metapopulation is less likely to persist in the long run.
+
+The detectability of such changes in dynamical behavior among salmonids appears to be idiosyncratic across species, and the difficulty in measuring critical slowing down may - ironically - be masked by large portfolio effects \cite{Krkosek:2014ch} [this might be more appropriate for the discussion].
 
@@
 
-\begin{figure*}[h!]
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\begin{subfigure}[t]{0.55\textwidth}
-\centering
-\includegraphics[width=\textwidth]{figs2/fig_MDPE_hm_theta3.pdf} 
-\caption{Low habitat heterogeneity ($\Delta\theta=3$)} \label{fig:thetadiff1}
-\end{subfigure}
-\begin{subfigure}[t]{0.55\textwidth}
-\centering
-\includegraphics[width=\textwidth]{figs2/fig_MDPE_hm_theta8.pdf} 
-\caption{High habitat heterogeneity ($\Delta\theta=8$)} \label{fig:thetadiff2}
-\end{subfigure}
-\caption{Total means $N_t$, difference in means $\Delta N$, and the portfolio effect PE for different habitat heterogeneities $\Delta\theta$. Light colors = high values.
-}
-\end{figure*}
+% \begin{figure*}[h!]
+%   \captionsetup{justification=raggedright,
+% singlelinecheck=false
+% }
+% \centering
+% \begin{subfigure}[t]{0.55\textwidth}
+% \centering
+% \includegraphics[width=\textwidth]{figs2/fig_MDPE_hm_theta3.pdf} 
+% \caption{Low habitat heterogeneity ($\Delta\theta=3$)} \label{fig:thetadiff1}
+% \end{subfigure}
+% \begin{subfigure}[t]{0.55\textwidth}
+% \centering
+% \includegraphics[width=\textwidth]{figs2/fig_MDPE_hm_theta8.pdf} 
+% \caption{High habitat heterogeneity ($\Delta\theta=8$)} \label{fig:thetadiff2}
+% \end{subfigure}
+% \caption{Total means $N_t$, difference in means $\Delta N$, and the portfolio effect PE for different habitat heterogeneities $\Delta\theta$. Light colors = high values.
+% }
+% \end{figure*}
 
-\begin{figure*}[h!]
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.8\textwidth]{figs2/fig_MDPE_hm_ddm.pdf}
-\caption{
-The same simulations as presented in Figure \ref{fig:PE}, except with density dependent straying, where $m(t) = m_0\left(1-N(t)/(C+N(t)\right)$. Light colors = high values.
-} \label{fig:ddm}
-\end{figure*}
+% \begin{figure*}[h!]
+%   \captionsetup{justification=raggedright,
+% singlelinecheck=false
+% }
+% \centering
+% \includegraphics[width=0.8\textwidth]{figs2/fig_MDPE_hm_ddm.pdf}
+% \caption{
+% The same simulations as presented in Figure \ref{fig:PE}, except with density dependent straying, where $m(t) = m_0\left(1-N(t)/(C+N(t)\right)$. Light colors = high values.
+% } \label{fig:ddm}
+% \end{figure*}
 

### 2017-05-22 model/KevinSimulations2.jl

@@ -329,6 +329,7 @@ dev.off()
 
 @everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve.jl")
 @everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/qualsfunc.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/bifdet.jl")
 
 
 #Analysis over m & theta divergence
@@ -402,6 +403,14 @@ x1mean = d["x1mean"];
 x2mean = d["x2mean"];
 pe = d["pe"];
 
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/bifdet.jl")
+bifvalue = bifdet(
+n1mean,
+n2mean,
+mvec,
+hvec
+);
+
 
 namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/figs2/fig_MDPE_hm.pdf");
 R"""
@@ -410,10 +419,30 @@ pal = rev(brewer.pal(9,"Blues"))
 pdf($namespace,height=3,width=8)
 par(mfrow=c(1,3))
 image(x=$mvec,y=$hvec,z=t($(n1mean[:,:]))+t($(n2mean[:,:])),zlim=c(0,3000),col=pal,xlab='m',ylab='h',main='Total biomass')
+points($(bifvalue),type='l',cex=1)
 image(x=$mvec,y=$hvec,z=sqrt((t($(n1mean[:,:]))-t($(n2mean[:,:])))^2),zlim=c(0,1500),col=pal,xlab='m',ylab='h',main='Biomass difference')
+points($(bifvalue),type='l',cex=1)
 image(x=$mvec,y=$hvec,z=t($(pe[:,:])),zlim=c(1,2),col=pal,xlab='m',ylab='h',main='PE')
+points($(bifvalue),type='l',cex=1)
 dev.off()
 """
+
+R"""
+x<- 1:10
+  y<- 1:15
+  z<- outer( x,y,"+") 
+  image.plot(x,y,z) 
+
+# or 
+  obj<- list( x=x,y=y,z=z)
+  image.plot(obj, legend.lab="Sverdrups")
+
+# add some points on diagonal using standard plot function
[…]

### 2017-05-22 model/KevinSimulations_ddm2.jl

@@ -354,6 +354,12 @@ x2mean = d["x2mean"];
 pe = d["pe"];
 
 
+bifvalue = bifdet(
+n1mean,
+n2mean,
+indmvec,
+hvec
+);
 namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/figs2/fig_MDPE_hm_ddm.pdf");
 R"""
 library(RColorBrewer)
@@ -361,8 +367,11 @@ pal = rev(brewer.pal(9,"Blues"))
 pdf($namespace,height=3,width=8)
 par(mfrow=c(1,3))
 image(x=$indmvec,y=$hvec,z=t($(n1mean[:,:]))+t($(n2mean[:,:])),zlim=c(0,3000),col=pal,xlab='m0',ylab='h',main='Total biomass')
+points($bifvalue,type='l')
 image(x=$indmvec,y=$hvec,z=sqrt((t($(n1mean[:,:]))-t($(n2mean[:,:])))^2),zlim=c(0,1500),col=pal,xlab='m0',ylab='h',main='Biomass difference')
+points($bifvalue,type='l')
 image(x=$indmvec,y=$hvec,z=t($(pe[:,:])),zlim=c(1,2),col=pal,xlab='m0',ylab='h',main='PE')
+points($bifvalue,type='l')
 dev.off()
 """
 
@@ -467,7 +476,7 @@ end
 save(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data2/data_sig_h_m_theta8_ddm.jld"),"n1mean",n1mean,"n2mean",n2mean,"x1mean",x1mean,"x2mean",x2mean,"pe",pe);
 
 
-d = load(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data2/data2_sig_h_m_theta8_ddm.jld"));
+d = load(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data2/data_sig_h_m_theta8_ddm.jld"));
 #This loads the dictionary
 n1mean = d["n1mean"];
 n2mean = d["n2mean"];
@@ -477,6 +486,12 @@ pe = d["pe"];
 
 
 
+bifvalue = bifdet(
+n1mean,
+n2mean,
+indmvec,
+hvec
+);
[…]

### 2017-05-22 model/KevinSimulations_mtheta2.jl

@@ -191,6 +191,13 @@ pe = d["pe"];
 
 
 
+bifvalue = bifdet(
+n1mean,
+n2mean,
+indmvec,
+hvec
+);
+
 namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/figs2/fig_MDPE_hm_mtheta.pdf");
 R"""
 library(RColorBrewer)
@@ -198,8 +205,11 @@ pal = rev(brewer.pal(9,"Blues"))
 pdf($namespace,height=3,width=8)
 par(mfrow=c(1,3))
 image(x=$mvec,y=$hvec,z=t($(n1mean[:,:]))+t($(n2mean[:,:])),zlim=c(0,3000),col=pal,xlab='m',ylab='h',main='Total biomass')
+lines($bifvalue)
 image(x=$mvec,y=$hvec,z=sqrt((t($(n1mean[:,:]))-t($(n2mean[:,:])))^2),zlim=c(0,1500),col=pal,xlab='m',ylab='h',main='Biomass difference')
+lines($bifvalue)
 image(x=$mvec,y=$hvec,z=t($(pe[:,:])),zlim=c(1,2),col=pal,xlab='m',ylab='h',main='PE')
+lines($bifvalue)
 dev.off()
 """
 
@@ -297,6 +307,13 @@ x2mean = d["x2mean"];
 pe = d["pe"];
 
 
+bifvalue = bifdet(
+n1mean,
+n2mean,
+indmvec,
+hvec
+);
+
 namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/figs2/fig_MDPE_hm_mtheta_ddm.pdf");
 R"""
 library(RColorBrewer)
@@ -304,7 +321,10 @@ pal = rev(brewer.pal(9,"Blues"))
 pdf($namespace,height=3,width=8)
 par(mfrow=c(1,3))
 image(x=$indmvec,y=$hvec,z=t($(n1mean[:,:]))+t($(n2mean[:,:])),zlim=c(0,3000),col=pal,xlab='m0',ylab='h',main='Total biomass')
+lines($(bifvalue[1:60,:]))
 image(x=$indmvec,y=$hvec,z=sqrt((t($(n1mean[:,:]))-t($(n2mean[:,:])))^2),zlim=c(0,1500),col=pal,xlab='m0',ylab='h',main='Biomass difference')
[…]

### 2017-05-22 Manuscript/msv2.tex
figure and text update

@@
 \caption{
-Total means $N_t$, difference in means $\Delta N$, and the portfolio effect PE as a function of heritability $h^2$ and a constant stray rate $m$. Light colors = high values.
+(a) Total means $N_t$, (b) difference in means $\Delta N$, and (c) the portfolio effect PE as a function of heritability $h^2$ and a constant stray rate $m$. Light colors = high values.
+The black line shows the fold bifurcation separating a single steady state (left) from alternative steady states (right).
+(d) The relationship between the time to recovery following a disturbance and the portfolio effect.
 } \label{fig:PE}
@@
   \mu_i(t+1) &= w_i\mu_i(t) + (1-w_i)\mu_j(t) \\ \nonumber
-  &+ h^2\sigma\frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)|\theta_i] + (1-w_i)R_i[\mu_j(t)|\theta_i]  \right),
+  &+ h^2\sigma^2\frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)|\theta_i] + (1-w_i)R_i[\mu_j(t)|\theta_i]  \right),
   \label{eq:mu}
@@
 
-\noindent {\bf (c) Measuring of metapopulation persistence}
+\noindent {\bf (c) Measuring metapopulation persistence}
 \noindent We evaluated metapopulation persistence by measuring the average-CV portfolio effect (PE) (REF: Anderson,others) as well as the time required for the system to return to a steady state following an induced disturbance to one or both of the populations.
-The average-CV portfolio effect is the average CV across each population divided by the CV of the aggregate, such that
+The average-CV portfolio effect is, as the name implies, the average CV across each population divided by the CV of the aggregate, such that
 
@@
 %Moreover, if and only if $\phi \neq 1$, it is expected that as the diversity of the aggregate increases, so should the portfolio effect, and this central thesis is thought to underlie the stability of populations, species, communities and even financial markets.
-Portfolio effects greater than unity corresponds to less synchronization ($\phi < 1$) \cite{Loreau:2008ju,Anderson:2014cx,Yeakel:2013vz} and thus a greater potential for demographic rescue among populations, thus buffering the system as a whole against extinction.
+Portfolio effects greater than unity corresponds to less synchronization ($\phi < 1$) \cite{Loreau:2008ju,Anderson:2014cx,Yeakel:2013vz} and thus a greater potential for demographic rescue among populations, buffering the system as a whole against extinction.
 
 %Although there are more robust ways to measure the portfolio effect (in particular for comparing the PE across species with different mean-variance relationships; REF: Anderson), its only utlity here is to provide a simple statistical measure of metapopulation persistence.
-A more direct way to measure the likelihood that a system will persist is to measure the time it takes to recover following an induced disturbance: systems that recover quickly (lower recovery times) are more robust than those that recover more slowly (higher recovery times).
-Measuring the time it takes for a perturbed system to relax to a steady state also permits a more detailed examination of the potential fragility of the metapopulation.
-For example, if populations settle into alternative steady states (alternative steady states in our model requiring one population to be high-density and one low-density), comparing recovery times after a disturbance applied to the high, low, and/or both populations permits an understanding of which component of the metapopulation has a stronger influence on persistence.
+A more direct way to measure system persistence is to measure the time that it takes it to relax back to its steady state following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
+Measuring the time that it takes for a perturbed system to relax also permits a more detailed perspective of the potential fragility of the metapopulation.
+For example, if populations settle to alternative steady states (alternative steady states in our model requiring one population to be high-density and one low-density), comparing recovery times after a disturbance applied to the high, low, and/or both populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery.
 \\
 
-\noindent {\bf (d) Density dependent straying rates and the effects of distance}
+\noindent {\bf (d) The effects of density and distance on the rate of straying}
 %Density dependent m
-\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is good evidence that at least in some species the stray rate is density dependent.
-Specifically, the rate at which individuals stray has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tends to decrease the rate at which individuals stray, thus reducing the overall proportion of a population that strays.
-According to \citeauthor{Berdahl:2016dx}, given the probability that an individual strays $m_0$, the proportion of the local population $N_i(t)$ that strays is
+\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is good evidence that at least in some species the straying rate is density dependent (REFS).
+Specifically, the rate at which individuals stray has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tend to decrease the rate at which individuals err, reducing the overall proportion of a population that strays.
+According to Berdahl et al. \cite{Berdahl:2016dx}, given the probability that an individual strays is $m_0$, the proportion of the local population $N_i(t)$ that strays is
 
@@
 
-\noindent where $C$ is the half-saturation value of $N_i$ where the density-dependent stray rate decreases sharply.
+\noindent where $C$ is a half-saturation constant.
 We note that at the limit $C\rightarrow \infty$, the density dependent stray rate becomes constant such that $m(t) \rightarrow m_0$, and this corresponds to the original formulation where $m=m_0$.
@@
 The greater the distance between two populations, the lower the expected rate of straying (REF).
-We can account for this interdependence in our model by assuming that $m$ (if the stray rate is constant) or $m_0$ (if the stray rate is density dependent) is a function of $\theta_i-\theta_j$, which can be assumed to be large if the remote site $j$ is a great distance away from the local site $i$.
-If sites $i$ and $j$ are very close, the stray rate is assumed to maximized at $m,m_0 = 0.5$, assuming both sites are equally attractive to the respective populations.
+We can account for this interdependence in our model by assuming that $m$ (if the stray rate is constant) or $m_0$ (if the stray rate is density dependent) is a function of the difference between optimal trait values between sites $\theta_i-\theta_j$, which can be assumed to be large if the remote site $j$ is a great distance away from the local site $i$.
+If sites $i$ and $j$ are very close, the stray rate is maximized at $m,m_0 = 0.5$, assuming both sites are equally attractive to the respective populations.
 Thus, we can integrate these two variables by setting $m,m_0 = (2 + \epsilon (\theta_i-\theta_j))^{-1}$, where $\epsilon$ sets the sensitivity of a declining $m$ to increasing distance (greater values of $\theta_i-\theta_j$).
-
-\section{Results}
-
-%The effects of trait evolution :: trait heritability, trait variance, and habitat heterogeneity
-%% Bifurcation to alternative stable state occurs at higher stray rates with increased trait heritability
-%% There is a peak in the portfolio effect directly before this bifurcation is crossed (early warning signal), and then a decline in PE with high m
-%% When trait heritability is high, there is a greater sensitivity of total biomass to increasing stray rates, though a LOWER sensitivity of the difference between alternative steady state values for increasing stray rates (and vice versa)
-%% For low heritability (h = 0.01 - 0.2 or so), moderate and insensitive changes in total biomass, but large differences between the values of alternative steady states leads to a local maximum for PE at low-moderate stray rates
-
-%Alternative stable states
-
-\noindent{\bf (a) Nonlinear effects of straying on the portfolio effect and return time} \\
-\noindent Straying generally lowers steady state densities for both recipient and donor populations by \emph{i}) losing locally-adapted individuals due to straying and \emph{ii}) introducing maladapted individuals from the donor population.
-%, regardless of trait heritability and variance or habitat heterogeneity
-The decline in steady state densities is not gradual: as straying increases, the system crosses a fold bifurcation whereby the single steady state across both sites bifurcates into two alternative steady states: one at high biomass density, and one at low biomass density (Fig.
-\ref{fig:traj}a,b).
-This bifurcation occurs at lower values of the straying rate $m$ at lower values of trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the existence of alternative stable states.
-
-Trait heritability has a large impact on the sensitivity of the straying rate on both the aggregate population steady state density ($N^*_T=N^*_1+N^*_2$) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N=|N^*_1-N^*_2|$).
-Greater trait heritability results in a larger decline in $N_T^*$ for increasing straying rates $m$, though results in only moderate changes to $\Delta N$ (Figs. \ref{fig:PE}A,B).
-Conversely, if the trait is has a lower heritability, an increase in the straying rate has little impact on the total biomass density and contrastingly large effects on $\Delta N$ (Fig.
-\ref{fig:PE}a,b).
-
-%Together these changes in steady state population densities in terms of $N_T$ and $\Delta N$ as a function of trait heritability and the rate of straying between populations give rise to highly nonlinear portfolio effects, which we quantify here as
-
-% Nonlinear PE(!!!)
-
-In the region where there is a single steady state among both populations, we find an low-intermediate portfolio effect ca. ${\rm PE}=1.4$, primarily due to the elevated mean values of $N_T$.
-As the fold bifurcation is reached with greater rates of straying, the portfolio effect explodes due to a large increase in the variance of both donor and recipient populations ${\rm VAR}(N_{i,j})$.
-This increase in variance is the product of a dynamical process known as critical slowing down that occurs near fold bifurcations (among others) \cite{Scheffer:2009gg}, and has been suggested to serve as an early warning signal of an approaching phase transition \cite{Scheffer:2009gg,Anonymous:2013br,Dakos:2014br}.
-For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}), where there alternative steady states among the recipient and donor sites, the portfolio effect declines steadily after the explosion in variance associated with onset of the fold bifurcation.
-This decline is more gradual as the straying rate increases if trait heritability is low, and steeper if trait heritability is high.
 
@@
 
+\section{Results}
+
+%The effects of trait evolution :: trait heritability, trait variance, and habitat heterogeneity
+%% Bifurcation to alternative stable state occurs at higher stray rates with increased trait heritability
+%% There is a peak in the portfolio effect directly before this bifurcation is crossed (early warning signal), and then a decline in PE with high m
+%% When trait heritability is high, there is a greater sensitivity of total biomass to increasing stray rates, though a LOWER sensitivity of the difference between alternative steady state values for increasing stray rates (and vice versa)
+%% For low heritability (h = 0.01 - 0.2 or so), moderate and insensitive changes in total biomass, but large differences between the values of alternative steady states leads to a local maximum for PE at low-moderate stray rates
+
+%Alternative stable states
+
+\noindent{\bf (a) Nonlinear effects of straying on the portfolio effect and recovery time} \\
+\noindent Straying generally lowers steady state densities for both recipient and donor populations by \emph{i}) losing locally-adapted individuals to the non-local population and \emph{ii}) gaining maladapted individuals from that same population.
+%, regardless of trait heritability and variance or habitat heterogeneity
+The decline in steady state densities is not gradual: as straying increases, the system crosses a fold bifurcation whereby the single steady state across both sites bifurcates into two alternative steady states: one at high biomass density, and one at low biomass density (Fig.
+\ref{fig:traj}a,b).
+This bifurcation  (shown by the black line in Fig\ref{fig:PE}(a-c)) occurs at lower values of the straying rate $m$ at lower values of trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
+
+Trait heritability has a large impact on the sensitivity of the straying rate to both the aggregate population steady state density ($N^*_T=N^*_1+N^*_2$; Fig.
+\ref{fig:PE}a) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N=|N^*_1-N^*_2|$; Fig.
+\ref{fig:PE}b).
+Greater trait heritability results in a larger decline in $N_T^*$ for increasing straying rates $m$, but leads to only moderate changes to $\Delta N$.
+Conversely, if the trait is less heritable, an increase in the straying rate has little impact on the total biomass density but contrastingly large effects on $\Delta N$.
+
+%Together these changes in steady state population densities in terms of $N_T$ and $\Delta N$ as a function of trait heritability and the rate of straying between populations give rise to highly nonlinear portfolio effects, which we quantify here as
+
+% Nonlinear PE(!!!)
+
+%In the region where there is a single steady state among both populations, we find an low-intermediate portfolio effect ca. ${\rm PE}=1.4$, primarily due to the elevated mean values of $N_T$.
+As the fold bifurcation is reached with greater rates of straying, the portfolio effect increases sharply due to an explosion in variance of both donor and recipient populations ${\rm VAR}(N_{i,j})$.
+This increase in variance is the product of a dynamical process known as critical slowing down that occurs near fold bifurcations \cite{Scheffer:2009gg}, and may serve as an early warning signal of an approaching phase transition \cite{Scheffer:2009gg,Anonymous:2013br,Dakos:2014br}.
+For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}(a-c)), where alternative steady states occur, the portfolio effect declines steadily as the CV of $N_T$ increases.
+The decline over $m$ is more gradual if trait heritability is low, and steeper if trait heritability is high.
+
 % \noindent{\bf (c) Rates of recovery following catastrophic collapse}\\ 
 % Return time as a measure of system persistence and relation to PE
-We measured the rate of recovery to a steady state as a function of the rate of straying $m$ following three types of induced disturbance: (\emph{i}) extinction of the low-density population, (\emph{ii}) extinction of the high-density population; (\emph{iii}) near-collapse of both populations where just 1.0\% of both populations survive.
-At low rates of straying both populations are subject to the same steady state, such that extinction of either population results in the same recovery time, which is generally less than that required after a near-collapse of both populations.
-For values of $m$ that give rise to alternative steady states, the recovery times associated with each disturbance type depends strongly on trait heritability.
-If both straying rate and heritability are low (Fig.
-\ref{fig:relax}a), extinction of the small population results in a lower recovery time, whereas as the rate of straying increases, the extinction of the larger population results is lower recovery times.
+We measured the time required for the system as measured by $N_T$ to relax to its steady state as a function of straying rate following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} amount to the same in the single steady state regime); (\emph{iii}) near-collapse of both populations where just 1.0\% of both populations survive.
+In general, the average-CV portfolio effect is negatively correlated with recovery time (Fig.
+\ref{fig:PE}d), suggesting that for our system both measures are valuable indicators of metapopulation persistence.
+Because we can assess the time to recovery in response to the various disturbance types described above, it allows us to gain an in-depth perspective on the fragility of the metapopulation as a function of straying rate.
+
+At low rates of straying both populations are subject to the same steady state, such that extinction of either population results in the same recovery time.
+At low rates of straying and low trait heritability, the time required to recover following the extinction of either population is generally less than or equal to the recovery time required after the near-collapse of both populations (Fig.
+\ref{fig:relax}a).
+In the alternative stable state regime, extinction of the smaller population results in the shortest recovery time, whereas as $m$ increases, extinction of the larger population results in the shortest recovery time.
+Increased trait heritability flips this picture around (Fig. \ref{fig:relax}b).
+
+If both straying rate and heritability are low, extinction of the small population results in shorter recovery times, whereas for higher straying rates, the extinction of the larger population results in shorter recovery times.
 This switch is in part due to a large burst in biomass of the surviving (large) population when the smaller of the two becomes extinct at high straying rates, which extends the time required for the smaller population to recover.
@@
 Near the onset of the fold bifurcations, recovery times increase drastically, regardless of disturbance.
-For straying rates in the alternative steady state regime, extinction of the smaller population consistently provides the least time to recovery.
+For straying rates in the alternative steady state regime, extinction of the smaller population consistently results in the shortest time to recovery.
 

### 2017-05-22 model/KevinSimulations2.jl

@@ -402,6 +402,9 @@ n2mean = d["n2mean"];
 x1mean = d["x1mean"];
 x2mean = d["x2mean"];
 pe = d["pe"];
+dExt = load(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data2Ext/data_sig_h_m.jld"));
+pe_ext = dExt["pe"];
+rt_ext = dExt["rt"];
 
 @everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/bifdet.jl")
 bifvalue = bifdet(
@@ -416,14 +419,15 @@ namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/fi
 R"""
 library(RColorBrewer)
 pal = rev(brewer.pal(9,"Blues"))
-pdf($namespace,height=3,width=8)
-par(mfrow=c(1,3))
+pdf($namespace,height=3,width=10)
+par(mfrow=c(1,4))
 image(x=$mvec,y=$hvec,z=t($(n1mean[:,:]))+t($(n2mean[:,:])),zlim=c(0,3000),col=pal,xlab='m',ylab='h',main='Total biomass')
 points($(bifvalue),type='l',cex=1)
 image(x=$mvec,y=$hvec,z=sqrt((t($(n1mean[:,:]))-t($(n2mean[:,:])))^2),zlim=c(0,1500),col=pal,xlab='m',ylab='h',main='Biomass difference')
 points($(bifvalue),type='l',cex=1)
 image(x=$mvec,y=$hvec,z=t($(pe[:,:])),zlim=c(1,2),col=pal,xlab='m',ylab='h',main='PE')
 points($(bifvalue),type='l',cex=1)
+plot($(rt_ext[1:50,:]),$(pe_ext[1:50,:]),log='x',ylim=c(1,3),xlim=c(18,300),pch='.',xlab='Recovery time',ylab='PE')
 dev.off()
 """
 

### 2017-05-22 Manuscript/msv2.tex
return time over habitat heterogeneity analyses and ms

@@
 \caption{
-Top row (a, b)
-Bottom row (c, d)
+(a, b) Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming constant straying rates $m$ and (c, d) density dependent straying rates (evaluated at the steady state $m^*$).
+If $m$ is density dependent, in the alternative steady state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
 } \label{fig:relax}
@@
 \ref{fig:traj}a,b).
-This bifurcation  (shown by the black line in Fig\ref{fig:PE}(a-c)) occurs at lower values of the straying rate $m$ at lower values of trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
+This bifurcation  (shown by the black line in Fig\ref{fig:PE}a-c) occurs at lower values of the straying rate $m$ at lower values of trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
 
@@
 This increase in variance is the product of a dynamical process known as critical slowing down that occurs near fold bifurcations \cite{Scheffer:2009gg}, and may serve as an early warning signal of an approaching phase transition \cite{Scheffer:2009gg,Anonymous:2013br,Dakos:2014br}.
-For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}(a-c)), where alternative steady states occur, the portfolio effect declines steadily as the CV of $N_T$ increases.
+For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}a-c), where alternative steady states occur, the portfolio effect declines steadily as the CV of $N_T$ increases.
 The decline over $m$ is more gradual if trait heritability is low, and steeper if trait heritability is high.
@@
 % Return time as a measure of system persistence and relation to PE
-We measured the time required for the system as measured by $N_T$ to relax to its steady state as a function of straying rate following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} amount to the same in the single steady state regime); (\emph{iii}) near-collapse of both populations where just 1.0\% of both populations survive.
+As the portfolio effect is highly sensitive to the rate of straying between populations, so is the time the system requires to recover following a large disturbance.
+We measured the time required for the system $N_T$ to relax to its steady state following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} are equivalent in the single steady state regime); (\emph{iii}) near-collapse of both populations where just 1.0\% of both populations survive.
 In general, the average-CV portfolio effect is negatively correlated with recovery time (Fig.
-\ref{fig:PE}d), suggesting that for our system both measures are valuable indicators of metapopulation persistence.
-Because we can assess the time to recovery in response to the various disturbance types described above, it allows us to gain an in-depth perspective on the fragility of the metapopulation as a function of straying rate.
+\ref{fig:PE}d), indicating that for our system both measures are valuable indicators of metapopulation persistence.
+Because we can assess the time to recovery in response to the various disturbance types described above, this allows us to gain an in-depth perspective into the fragility of the metapopulation as a function of straying rate.
 
@@
 In the alternative stable state regime, extinction of the smaller population results in the shortest recovery time, whereas as $m$ increases, extinction of the larger population results in the shortest recovery time.
-Increased trait heritability flips this picture around (Fig. \ref{fig:relax}b).
-
-If both straying rate and heritability are low, extinction of the small population results in shorter recovery times, whereas for higher straying rates, the extinction of the larger population results in shorter recovery times.
-This switch is in part due to a large burst in biomass of the surviving (large) population when the smaller of the two becomes extinct at high straying rates, which extends the time required for the smaller population to recover.
-Conversely, with increased trait heritability (Fig.
-\ref{fig:relax}b) and low straying rates in the single steady state regime, near collapse of both populations (rather than the extinction of either) results in the lowest recovery time.
-Near the onset of the fold bifurcations, recovery times increase drastically, regardless of disturbance.
+This switch is in part due to a large burst in biomass of the surviving (large) population when the smaller of the two becomes extinct at high straying rates, which extends the time required for the smaller population (and thus the system as a whole) to recover.
+Increased trait heritability flips this relationship (Fig.
+\ref{fig:relax}b), such that systems suffering near-collapse of both populations recover more quickly than if either population is lost at low straying rates.
+As straying rates increase, the system recovers more quickly if the small population is lost, whereas the time to recovery is maximized if both populations suffer near-collapse.
+Near the onset of the fold bifurcation, recovery times increase drastically regardless of straying rate, heritability, or disturbance type.
 For straying rates in the alternative steady state regime, extinction of the smaller population consistently results in the shortest time to recovery.
@@
 \caption{
-Phenotypic diversity ($\Delta \mu$) evaluated at the steady state as a function of straying rate $m$ and trait heritability $h^2$.
+Phenotypic diversity ($\Delta \mu^*$) evaluated at the steady state as a function of straying rate $m$ and trait heritability $h^2$.
 The descrete jump occurs as the system crosses the fold bifurcation; lower phenoytic diversity emerges with higher straying rates and in the alternative steady state regime.
@@
 %Lowering Phenotypic diversity with increasing straying rate
-Increased rates of straying lowers phenotypic diversity ($\Delta \mu = |\mu_i(t)-\mu_j(t)|$, evaluated at the steady state) because both local and remote populations are increasingly homogenized.
+Increased rates of straying lowers phenotypic diversity ($\Delta \mu^* = |\mu_i(t)-\mu_j(t)|$, evaluated at the steady state) because both local and remote populations are increasingly homogenized.
 The sensitivity of this phenotypic homogenization to straying is increased with lower heritability because the selective forces acting against trait means far from the local optima are lessened for lower $h^2$.
-Less intuitively, we observe a discrete jump towards low phenotypic diversity at low-moderate ($0.05 \leq m < 0.2$) straying rates, which occurs as the consequence of crossing the fold bifurcation (Fig.
-\ref{fig:traitdiff}).
-Although the development of alternative stable states elevates the portfolio effect due to the variance-dampening effects of the aggregate, this change in dynamic regimes is also associated with a substantial decline in phenotypic diversity.
+Less intuitively, we observe a discrete jump towards low phenotypic diversity as the fold bifurcation is crossed (Fig. \ref{fig:traitdiff}).
+Although the development of alternative stable states elevates the portfolio effect due to the variance-dampening effects of the aggregate, entering this dynamic regime also results in a substantial decline in phenotypic diversity, which may have less predictable adverse effects on the population.
 \\
@@
 \caption{
-(a) Median portfolio effect as a function of stray rate $m$ for lower trait heritability ($h^2 < 0.5$).
-Vertical dashed lines represent the fold bifurcation, which occurs at lower stray rates with increased habitat heterogeneity $\Delta \theta$.
-(b) Within the box shown in (a), PE as a function of the constant straying rate is shown as the solid line.
-Paired open circles show PE as a function of a density dependent straying rate.
-Straying rates are paired because the recipient and donor populations exist in alternative steady states.
+(a) Median portfolio effect as a function of stray rate $m$ given heritability is $h^2 < 0.5$ for habitats with increasing heterogeneity as measured by the difference in regional trait optima $\Delta \theta$.
+(b) Median portfolio effects as shown in (a) are denoted by the solid lines. 
+Point pairs connected by a horizontal line represent the PE as a function of density dependent straying rates, evaluated for both low- and high-density populations at equilibrium.
 The lower straying rate of a pair is for the larger population; the higher straying rate is for the smaller population.
@@
 
-\noindent{\bf (b) The role of habitat heterogeneity on the portfolio effect}\\ 
-\noindent Increased differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater between-site differences in conditions that favor alternative traits, interpreted here as increased habitat heterogeneity.
-At one end of the spectrum, if both populations are isolated, natural selection would direct the mean trait values of both populations towards their respective optima, such that $\mu_i(t) \rightarrow \theta_i$ as $t\rightarrow\infty$.
+\noindent{\bf (b) The role of habitat heterogeneity on metapopulation persistence}\\ 
+\noindent Increasing differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater regional differences in the conditions that favor alternative trait complexes, which we interpret here as greater habitat heterogeneity.
+If both populations are isolated, natural selection will direct the mean trait values of both populations towards their respective optima, such that $\mu_i(t) \rightarrow \theta_i$ as $t\rightarrow\infty$.
 %difference in optimal trait values between the local and remote populations, which we call .
-However when straying exists, we find that increasingly divergent trait optima generally lowers $N_T$ and exaggerates $\Delta N$, such that one population has the majority of the biomass (Appendix Fig.
-\ref{fig:thetadiff1},\ref{thetadiff2}).
-Which population becomes large or small depends on the initial conditions, which are randomized. 
-The impact of habitat heterogeneity on the portfolio effect is more complex, serving to emphasize the nonlinear relationship between the stray rate and the PE, regardless of heritability.
+With the onset of straying, we find that increasingly divergent trait optima generally lower $N_T$ and exaggerate $\Delta N$, such that the biomass distribution becomes increasingly uneven.
+The impact of habitat heterogeneity on the portfolio effect is more complex, serving to emphasize the nonlinear relationship between the stray rate and the PE. %regardless of heritability.
 %The rate of straying that gives rise to a large PE at the fold bifurcation is conditioned on the difference in optimal trait values between the local and remote populations, which we call $\Delta\theta = \sqrt{(\theta_i - \theta_j)^2}$.
-The extent of the nonlinearity between $m$ and the PE depends very much on $\Delta\theta$, particularly when trait heritability is low ($h^2<0.5$).
-As habitat heterogeneity increases, the fold bifurcation and associated spike in PE generally occurs for lower values of straying rate $m$, meaning that a lower amount of straying can give rise to alternative stable states (Fig.
-\ref{fig:thetaPE}).
-In the region where alternative stable states are feasible (Fig.
-\ref{fig:thetaPE}), additional straying increases the portfolio effect to a local maximum before its negative effects serve to lower PE as $m\rightarrow 0.5$, and this effect is exaggerated with increasing habitat heterogeneity.
+As habitat heterogeneity increases, the fold bifurcation occurs at lower straying rates whereas the associated jump in the PE is greater (Fig.
+\ref{fig:thetaPE}), decreasing the recovery time (Appendix Fig.
+\ref{fig:thetarelax}).
+For greater straying rates, habitat heterogeneity erodes the PE and slows recovery time (Fig. \ref{fig:thetaPE}, Appendix Fig. \ref{fig:thetarelax})
+Taken together, greater $\Delta\theta$ promotes persistence when straying rates are low, but erodes persistence when straying rates are high.
+%This means that alternative stable states can emerge with relatively little dispersal between sites (Fig. \ref{fig:thetaPE}), however the increased PE (and shorter recovery times).
 \\
@@
 \noindent{\bf (c) The effects of density dependent straying}\\
-\noindent If we assume that the rate of straying is density dependent, the probability of straying at the individual level $m_0$ determines the rate of straying within the population, such that $m(t)$ becomes lower as $N(t)$ increases (Eq.
+\noindent If we assume that the rate of straying is density dependent, the probability that an individual strays $m_0$ determines the rate of straying within the population, such that $m(t)$ becomes lower as $N(t)$ increases (Eq.
 \ref{eq:ddm}).
-At the steady state $m(t)=m^*$ is also constant, though by definition, it is always less than $m_0$ such that $0 < m^* < m_0$.
-Because our analyses concern steady state conditions, the inclusion of density-dependent straying does not have a large impact on the qualitative nature of our findings (Appendix Fig. \ref{fig:ddm}).
+%Because our analyses concern steady state conditions, the inclusion of density-dependent straying does not have a large impact on the qualitative nature of our findings (Appendix Fig. \ref{fig:ddm}).
+Density dependent straying alters the straying rate at the steady state because $0 < m^* < m_0$, and this serves to rescale both the strength of the PE as well as the recovery time following an induced disturbance but does not change the qualitative nature of our findings.
+Moreover, in the alternative stable state regime, because the recipient and donor population exist at different steady state densities, they have different rates of straying such that a given value of the portfolio effect is now associated with a pair of straying rates $(m_i^*,m_j^*)$.
+A range of $(m_i^*,m_j^*)$ values are obtained by varying $m_0$, which is positively and linearly related to $(m_i^*,m_j^*)$.
+For instance, the recovery times for systems with constant and density dependent straying both show similar trends for a given $m$ or $m^*$ (Fig.
+\ref{fig:relax}c,d), however there are some distinct quantitative differences.
+When straying rate and trait heritability are low, near-collapse of both populations result in longer than expected recovery times, whereas the recovery times in the alternative stable state regime (higher $m^*$) are similar (Fig.
+\ref{fig:relax}c).
+There is greater similarity in the recovery times for constant and density dependent $m$ systems when trait heritability is high.
 
-The assumption that straying is density dependent alters the strength of the portfolio effect for a given $m^*$.
-As shown in Figure \ref{fig:thetaPE}(b), in the alternative stable state regime, because the recipient and donor population exist at different steady state densities, they have different rates of straying such that a given value of the portfolio effect is now associated with a pair of straying rates $(m_i^*,m_j^*)$.
-A range of $(m_i^*,m_j^*)$ values are obtained by varying $m_0$, which is positively and linearly related to $(m_i^*,m_j^*)$.
-Importantly, we show that metapopulations settling at relatively low rates of straying have lower-than-expected portfolio effects, whereas metapopulations settling at relatively high rates of straying have higher-than expected portfolio effects.
-This suggests that there is a strong tradeoff in the role of density dependent straying on the persistence of metapopulations: as the individual probability of straying is lowered, the straying rates of both recipient and donor populations are similarly lowered, however this diminishes the system's ability to buffer itself against extinction as measured by the portfolio effect.
-Conversely, as the probability that an individual strays increases, so does rate of straying for both recipient and donor populations, and this serves to elevate the portfolio effect compared to a constant $m$.
-This effect remains but is somewhat diminished as habitats become increasingly heterogeneous.
-
-Return times (Fig. \ref{fig:relax}c,d).
+A metapopulation in the alternative stable state regime has two straying rates: a lower straying rate associated with the high-density population, and a higher straying rate associated with the low-density population.
+We find that the resultant portfolio effect approximates that generated by a system with a similar constant straying rate, such that ${\rm PE}(m) \approx {\rm PE}(m^*_{\rm high-density})$ (Fig.
+\ref{fig:thetaPE}b).
+As the steady state straying rate increases, the portfolio effect follows a similar trajectory as shown for metapopulations with constant straying rates (Fig.
+\ref{fig:thetaPE}a), though is generally higher, promoting persistence.
+RETURN TIMES
+This suggests that density dependent straying...
+% 
+% 
+% In a metapopulation with density dependent straying, the lower straying rate of the high-density population generates a portfolio effect that would be expected if straying were constant.
+% 
+% A metapopulation with density dependent straying has a portfolio effect 
+% 
+% equal to that of a metapopulation with constant straying
+% 
+% as would be expected for a metapopulation with a constant straying rate equalt to that of the 
+% 
+% We find that the by-definition lower straying rate of the high-density population (left-most of the paired points in Fig.
+\ref{fig:thetaPE}b) largely determines the portfolio effect for the metapopulation.
+% In other words, 
 \\
@@
 \clearpage
+
 \setcounter{figure}{0}
@@
 \centering
-\includegraphics[width=0.5\textwidth]{figs2/fig_rtvspe.pdf}
+\includegraphics[width=0.5\textwidth]{figs2/fig_relaxtheta.pdf}
 \caption{
-Return time vs. portfolio effect.
-} \label{fig:rtvspe}
+Recovery time as a function of straying rate $m$ and habitat heterogeneity $\Delta\theta$.
+} \label{fig:relaxtheta}
 \end{figure}

### 2017-05-22 model/KevinExtinction.jl

@@ -321,11 +321,12 @@ library(RColorBrewer)
 pdf($namespace,height=8,width=10)
 pal = brewer.pal(9,'Greys')
 palsub = pal[c(4,6,8)];
-par(mfrow=c(2,2),mai = c(0.8, 0.9, 0.2, 0.1))
+par(mfrow=c(2,2),mai = c(0.8, 0.9, 0.3, 0.1))
 plot($mvec,$(ma_rth2[:,1]),col=palsub[1],type='l',log='y',cex=0.5,lwd=2,xlab='m',ylab='Recovery time',ylim=c(min($rt),max($(ma_rth2))))
 lines($mvec,$(ma_rth2[:,2]),col=palsub[2],cex=0.5,lwd=2)
 lines($mvec,$(ma_rth2[:,3]),col=palsub[3],cex=0.5,lwd=2)
 text(0.25,max($(ma_rth2))+23,expression(paste(h^2,'=0.2')), xpd=TRUE)
+text(par('usr')[1]-0.1,max($(ma_rth2))+23,'(a)', xpd=TRUE)
 text(-0.11, ## x position
      41, ## position of the low axis
      srt=90, ## angle
@@ -338,8 +339,9 @@ plot($mvec,$(ma_rth8[:,1]),col=palsub[1],type='l',log='y',cex=0.5,lwd=2,xlab='m'
 lines($mvec,$(ma_rth8[:,2]),col=palsub[2],cex=0.5,lwd=2)
 lines($mvec,$(ma_rth8[:,3]),col=palsub[3],cex=0.5,lwd=2)
 text(0.25,max($(ma_rth8))+1600,expression(paste(h^2,'=0.8')), xpd=TRUE)
-text(par('usr')[1]-0.1,max($(ma_rth2))+23,'(a)', xpd=TRUE)
-legend(x=0.43,y=3500,legend=$pvec,col=palsub,pch=22,xpd=TRUE,pt.bg=palsub,cex=1, bty="n") #,title=expression(paste(Delta,theta))
+text(par('usr')[1]-0.085,max($(ma_rth8))+1600,'(b)', xpd=TRUE)
+types = c('low-density extinction','high-density extinction','near-collapse')
[…]

### 2017-05-22 model/habitathetero.jl

@@ -1,0 +1,252 @@
+
+
+@everywhere using Distributions
+@everywhere using RCall
+@everywhere using HDF5
+@everywhere using JLD
+
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/qualsfunc.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/bifdet.jl")
+
+
+#Analysis over m & theta divergence
+mvec=collect(0.0:0.001:0.3);
+thetadiffvec = collect(0.0:0.1:10.0);
+
+thetadiffarray=SharedArray(Float64,length(thetadiffvec),length(mvec));
+
+n1mean=SharedArray(Float64,length(thetadiffvec),length(mvec));
+n2mean=SharedArray(Float64,length(thetadiffvec),length(mvec));
+x1mean=SharedArray(Float64,length(thetadiffvec),length(mvec));
+x2mean=SharedArray(Float64,length(thetadiffvec),length(mvec));
+pe=SharedArray(Float64,length(thetadiffvec),length(mvec));
+
+tmax=10000;
+z=0.5;
+rmax=2.0;
+beta=0.001;
+theta1=5.0;
+
+tau=1.0;
+sigmaE=0;
+sigmaG=1;
+h=0.2
+perror=0.01;
+
+@sync @parallel for k=1:length(thetadiffvec)
+    
+    thetadiff=thetadiffvec[k];
+    
+    for j=1:length(mvec)
+      m = mvec[j];
+      
+      n1, n2, x1, x2, w1, w2 = 
+      KevinEvolve(
+        tmax,
+        z,
+        rmax,
+        beta,
+        theta1,
+        thetadiff,
+        tau,
+        h,
+        sigmaE,
+        sigmaG,
+        m,
+        perror
+        );
+      burnin=0.8
+      n1trim = n1[Int64(floor(tmax*burnin)):tmax];
+      n2trim = n2[Int64(floor(tmax*burnin)):tmax];
+      x1trim = x1[Int64(floor(tmax*burnin)):tmax];
[…]

### 2017-05-22 model/mvsddm2.jl

@@ -249,12 +249,12 @@ plot($(ma_mvecsmooth),$(ma_medpe5smooth),col=pal[2],type='l',ylim=c(1,max($([ma_
 lines($(ma_mvecsmooth),$(ma_medpe8smooth),col=pal[3],lwd=3)
 lines($(ma_mvecsmooth),$(ma_medpe3smooth),col=pal[1],lwd=3)
 #Or moving average
-points($(ma_medindm3max[mddmlist]),$(ma_medpe3_ddm[mddmlist]),col=pal[1],pch=1,lwd=2)
-points($(ma_medindm3min[mddmlist]),$(ma_medpe3_ddm[mddmlist]),col=pal[1],pch=1,lwd=2)
-points($(ma_medindm5max[mddmlist]),$(ma_medpe5_ddm[mddmlist]),col=pal[2],pch=1,lwd=2)
-points($(ma_medindm5min[mddmlist]),$(ma_medpe5_ddm[mddmlist]),col=pal[2],pch=1,lwd=2)
-points($(ma_medindm8max[mddmlist]),$(ma_medpe8_ddm[mddmlist]),col=pal[3],pch=1,lwd=2)
-points($(ma_medindm8min[mddmlist]),$(ma_medpe8_ddm[mddmlist]),col=pal[3],pch=1,lwd=2)
+points($(ma_medindm3max[mddmlist]),$(ma_medpe3_ddm[mddmlist]),col=pal[1],pch=16,cex=0.8)
+points($(ma_medindm3min[mddmlist]),$(ma_medpe3_ddm[mddmlist]),col=pal[1],pch=16,cex=0.8)
+points($(ma_medindm5max[mddmlist]),$(ma_medpe5_ddm[mddmlist]),col=pal[2],pch=16,cex=0.8)
+points($(ma_medindm5min[mddmlist]),$(ma_medpe5_ddm[mddmlist]),col=pal[2],pch=16,cex=0.8)
+points($(ma_medindm8max[mddmlist]),$(ma_medpe8_ddm[mddmlist]),col=pal[3],pch=16,cex=0.8)
+points($(ma_medindm8min[mddmlist]),$(ma_medpe8_ddm[mddmlist]),col=pal[3],pch=16,cex=0.8)
 l = length($(ma_medindm5max[mddmlist]))
 for (i in 1:l) {
[…]

### 2017-05-24 Manuscript/msv2.tex
ms changes plus supp figs

@@
 \usepackage{bibunits}
+
+\newcommand{\beginsupplement}{%
+        \clearpage
+        \setcounter{table}{0}
+        \renewcommand{\thetable}{S\arabic{table}}%
+        \setcounter{figure}{0}
+        \renewcommand{\thefigure}{S\arabic{figure}}%
+     }
 
@@
 %Movement through space (link back to straying)
-Migratory populations that return to a breeding ground or natal stream to reproduce are linked to each other by some proportion of the population that permanently disperse, or stray into the `wrong' site; we might say that there is at least one \emph{Kevin} in every school or flock in every school or flock (Fig.
-\ref{fig:xkcd}).
+Migratory populations that return to a breeding ground or natal stream to reproduce are linked to each other by some proportion of the population that permanently disperse, or stray into the `wrong' site; we might say that there is at least one \emph{Kevin} in every school or flock in every school or flock (figure \ref{fig:xkcd}).
 The rate at which individuals stray, $m$, has been subject to a growing amount of both theoretical and empirical research \cite{H:2013fs,Keefer:2014gg,Bett:2017ha} and may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \cite{Berdahl:2015kv,Berdahl:2016dx}.
@@
 
-\section{Model Description}
+\section{Model Description \& Analysis}
 
 \noindent{\bf (a) Metapopulation framework}\\
-\noindent Here we consider two populations $N_1$ and $N_2$ that are separated in space in distinct habitats, each with trait values $x_1$ and $x_2$ determining recruitment rates.
+\noindent We consider two populations $N_1$ and $N_2$ that are separated in space in distinct habitats, each with trait values $x_1$ and $x_2$ determining recruitment rates.
 We assume that there is an optimum trait value $\theta_1$ and $\theta_2$ associated with each habitat, where recruitment is maximized if the trait value of the local population $x = \theta$.
-Moreover, we assume that $x_{1,2}$ are normally distributed with means $\mu_1$ and $\mu_2$ and have a shared standard deviation $\sigma$.
-As such, the recruitment rate for both populations is determined by the mean trait value of the local population, such that $r_1 = R_1[\mu_1(t)|\theta_1]$.
+Moreover, we assume that $x_{1,2}$ are normally distributed with means $\mu_1$ and $\mu_2$ and have the same standard deviation $\sigma$.
+As such, the recruitment rate for both populations is determined by the mean trait value of the local population, such that $r_1 = R_1[\mu_1(t),\theta_1]$.
 Trait means for each population are subject to selection, the strength of which depends on the difference between the population mean and the local trait optimum at a given point in time \cite{simpson1953major,Lande:1976ga}.
@@
 \centering
-\includegraphics[width=0.4\textwidth]{figs2/fig_Density.pdf}
+\includegraphics[width=0.4\textwidth]{figs2/fig_traj.pdf}
 \caption{
@@
 We use the discrete Ricker population dynamic framework described by Shelton and Mangel \cite{Shelton:2011eq} as the basis for our two-site model, with the added effect of the local population $N_i$ mixing with a set proportion $m$ of a remote population $N_j$ that is straying into it.
-We first assume that the proportion ${\rm e}^{-Z}$ of both populations survive, and that the aggregated mix of the populations (local individuals in addition to the straying individuals) are subject to the same compensatory effects, determined by the parameter $\beta$.
-For a local site $i\in(1,2)$ that collects straying individuals from a remote site $j\in(1,2)$, if $N_i$ is the local site and $N_j$ is the remote site, the difference equation that determine changes in population size is
+In this sense, both populations serve as donor and recipient populations.
+We first assume that the proportion ${\rm e}^{-Z}$ of both populations survive such that the surviving aggregated population, composed of both local individuals (at site $i$) and incoming strays (from site $j$), is $\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z}$.
+Because local individuals will recruit differently than incoming strays, the recruitment of the aggregate must incorporate two recruitment functions, given by $\left(R_i[\mu_i(t)] (1-m)N_i(t) + R_i[\mu_j(t)] m N_j(t)\right)$.
+This mix of individuals is subject to the same compensatory effects, which is determined by the parameter $\beta$.
+Taken together, the difference equation that determine changes in population size is
 
@@
 
-\noindent where a small amount of demographic process error is added to the reproductive rate, and where the difference equation for $N_j$ mirrors that for $N_i$.
+\noindent where the difference equation for $N_j$ mirrors that for $N_i$.
 
-The combined recruitment of local individuals $(1-m)N_i(t)$ and incoming strays $mN_j(t)$, as a function of their mean trait value at time $t$ and given the local trait optimum $\theta_i$, is then
+The combined recruitment of local individuals $(1-m)N_i(t)$ and incoming strays $mN_j(t)$, as a function of their mean trait value at time $t$.
+Given the local trait optimum is $\theta_i$, recruitment is
 
@@
   &R_i[\mu_i(t)] = \\ \nonumber
-  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i(t)-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i(t)|\mu_i,\sigma^2) {\rm d}x_i(t) +\tilde{P}\\ \nonumber
+  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i(t)-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i(t),\mu_i,\sigma^2) {\rm d}x_i(t) +\tilde{P}\\ \nonumber
   &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-\mu_i(t))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\} +\tilde{P},
@@
 
-\noindent where the mismatch between the local trait mean $\mu_i(t)$ and the local optimum $\theta_i$ determines the recruitment rate for the population, and $\tilde{P}$ introduces a small amount of demographic error $\tilde{P}\sim {\rm Normal}(0,0.01)$.
-The parameter $\tau$ controls the sensitivity of recruitment to changes in the mean trait value away from the optimum, which we set as $\tau=1$ here and throughout.
+\noindent where the mismatch between the local trait mean $\mu_i(t)$ and the local optimum $\theta_i$ determines scales the recruitment rate for the population, and $\tilde{P}\sim {\rm Normal}(0,0.01)$ introduces a small amount of demographic error.
+The parameter $\tau$ is the strength of selection, and controls the sensitivity of recruitment to changes in the mean trait value away from the optimum, which we set as $\tau=1$ here and throughout.
 % The compensatory effects are then determined by the exponential, following the Ricker stock-recruitment relationship.
@@
 First, we assume that the distribution resulting from the mix of remote and local individuals, following reproduction, is also normal with a mean value being that of the mixed-normal.
-Second, we assume that changes in trait variance through time are minimal, such that $\sigma$ is assumed to be constant.
+Second, we assume that changes in trait variance through time are minimal, such that $\sigma^2$ is assumed to be constant.
 
@@
 \begin{align}
+  \label{eq:mu}
   \mu_i(t+1) &= w_i\mu_i(t) + (1-w_i)\mu_j(t) \\ \nonumber
   &+ h^2\sigma^2\frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)|\theta_i] + (1-w_i)R_i[\mu_j(t)|\theta_i]  \right),
-  \label{eq:mu}
 \end{align}
 
-\noindent where the first two factors determine the mixed normal average of the now-mixed local and remote populations.
-This mixed normal is weighted by the proportion of the population that is local and remote, respectively, which depends on the stray rate $m$.
+\noindent where the first two factors determine the mixed normal average of the aggregated local and remote populations.
 The partial derivative in the Eq.
-\ref{eq:mu} determines how the mean trait changes through time due to natural selection (REF), which is proportional to the change in mean fitness with respect to $\mu_i$.
+\ref{eq:mu} determines how the mean trait changes through time due to natural selection \cite{Lande:1976ga}, which is proportional to the change in mean fitness with respect to $\mu_i$.
 \\
@@
 \noindent {\bf (c) Measuring metapopulation persistence}
-\noindent We evaluated metapopulation persistence by measuring the average-CV portfolio effect (PE) (REF: Anderson,others) as well as the time required for the system to return to a steady state following an induced disturbance to one or both of the populations.
-The average-CV portfolio effect is, as the name implies, the average CV across each population divided by the CV of the aggregate, such that
+\noindent We evaluated metapopulation persistence by measuring the average-CV portfolio effect (PE) \cite{Anderson:2014cx,Schindler:2015gf} as well as the time required for the system to return to a steady state following an induced disturbance to one or both of the populations \cite{Ovaskainen:2002il}.
+The average-CV portfolio effect is, as the name implies, the average CV across each population divided by the CV of the aggregate \cite{Anderson:2013gb}, such that
 
@@
 As the CV of $N_T$ decreases relative to that of the constituent populations, $\langle{\rm PE}\rangle > 1$, and the metapopulation is presumed to become more stable.
-A similar measure of metapopulation persistence can be evaluated by measuring the synchronization $\phi$ between populations, where 
+{\color{red} A similar measure of metapopulation persistence can be evaluated by measuring the synchronization $\phi$ between populations, where 
 \begin{equation}
@@
 \end{equation}
-By rearranging and substituting eq. (\ref{eq_sync}) into eq. (\ref{eq_PE}), we can observe that $\langle{\rm PE}\rangle \propto \sqrt{\phi}^{-1}$.
-Thus, the dynamical diversity of the metapopulation offers a mirror to the portfolio effect, where perfect synchrony ($\phi = 1$) also means $\langle{\rm PE}\rangle = 1$.
+By rearranging and substituting eq. (\ref{eq_sync}) into eq. (\ref{eq:pe}), we can observe that $\langle{\rm PE}\rangle \propto \sqrt{\phi}^{-1}$.
+Thus, the dynamical diversity of the metapopulation offers a mirror to the portfolio effect, where perfect synchrony ($\phi = 1$) also means $\langle{\rm PE}\rangle = 1$.}
 %Moreover, if and only if $\phi \neq 1$, it is expected that as the diversity of the aggregate increases, so should the portfolio effect, and this central thesis is thought to underlie the stability of populations, species, communities and even financial markets.
@@
 %Although there are more robust ways to measure the portfolio effect (in particular for comparing the PE across species with different mean-variance relationships; REF: Anderson), its only utlity here is to provide a simple statistical measure of metapopulation persistence.
-A more direct way to measure system persistence is to measure the time that it takes it to relax back to its steady state following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
+A more direct way to measure system persistence is to measure the time that it takes the system (measured as the aggregate biomass $N_T$) to recover its steady state abundance following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
 Measuring the time that it takes for a perturbed system to relax also permits a more detailed perspective of the potential fragility of the metapopulation.
@@
 %Density dependent m
-\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is good evidence that at least in some species the straying rate is density dependent (REFS).
+\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is good evidence that at least in some species the straying rate is density dependent \cite{Berdahl:2016dx,Bett:2017ha}.
 Specifically, the rate at which individuals stray has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tend to decrease the rate at which individuals err, reducing the overall proportion of a population that strays.
@@
 \noindent where $C$ is a half-saturation constant.
-We note that at the limit $C\rightarrow \infty$, the density dependent stray rate becomes constant such that $m(t) \rightarrow m_0$, and this corresponds to the original formulation where $m=m_0$.
-A similar observation shows that when the population density is very high, $m(t) \rightarrow 0$, and when it is close to extinction, $m(t) \rightarrow m_0$.
+We note that at the limit $C\rightarrow \infty$, the density dependent straying rate becomes constant such that $m(t) \rightarrow m_0$, and this corresponds to the original formulation where $m=m_0$.
+A similar observation shows that when the population density is very high, $m(t) \rightarrow 0$, and when it is small, individuals operate without regard to collective behavior, meaning $m(t) \rightarrow m_0$.
 Thus, for realistic population densities, $m(t) < m_0$.
@@
 We can account for this interdependence in our model by assuming that $m$ (if the stray rate is constant) or $m_0$ (if the stray rate is density dependent) is a function of the difference between optimal trait values between sites $\theta_i-\theta_j$, which can be assumed to be large if the remote site $j$ is a great distance away from the local site $i$.
-If sites $i$ and $j$ are very close, the stray rate is maximized at $m,m_0 = 0.5$, assuming both sites are equally attractive to the respective populations.
-Thus, we can integrate these two variables by setting $m,m_0 = (2 + \epsilon (\theta_i-\theta_j))^{-1}$, where $\epsilon$ sets the sensitivity of a declining $m$ to increasing distance (greater values of $\theta_i-\theta_j$).
+If sites $i$ and $j$ are very close, the stray rate is maximized at $m_{\rm max} = 0.5$, assuming both sites are equally attractive to the respective populations.
+Thus, we can integrate these two variables by setting $m,m_0 = (1/m_{\rm max} + \epsilon (\theta_i-\theta_j))^{-1}$, where $\epsilon$ sets the sensitivity of a declining $m$ to increasing distance (greater values of $\theta_i-\theta_j$).
 
@@
 \noindent{\bf (a) Nonlinear effects of straying on the portfolio effect and recovery time} \\
-\noindent Straying generally lowers steady state densities for both recipient and donor populations by \emph{i}) losing locally-adapted individuals to the non-local population and \emph{ii}) gaining maladapted individuals from that same population.
-%, regardless of trait heritability and variance or habitat heterogeneity
-The decline in steady state densities is not gradual: as straying increases, the system crosses a fold bifurcation whereby the single steady state across both sites bifurcates into two alternative steady states: one at high biomass density, and one at low biomass density (Fig.
-\ref{fig:traj}a,b).
-This bifurcation  (shown by the black line in Fig\ref{fig:PE}a-c) occurs at lower values of the straying rate $m$ at lower values of trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
+\noindent Straying generally lowers steady state densities for both populations by \emph{i}) the donor population losing locally-adapted individuals to the recipient population and \emph{ii}) the introduction of maladapted individuals to the recipient population from the donor population, and this accords with observations from natural populations \cite{Bett:2017ha} %, regardless of trait heritability and variance or habitat heterogeneity
+The decline in steady state densities is not gradual: as straying increases, the system crosses a fold bifurcation whereby the single steady state for the metapopulation bifurcates into two alternative steady states: one at high biomass, and one at low biomass density (figure \ref{fig:traj}a, \ref{fig:PE}a).
+Mean trait values for both populations bifurcate similarly (figure \ref{fig:traj}b), depending on which population attains a low- vs.
+high-density, which in our system is random due to a small amount of introduced variation in the initial conditions.
+The fold bifurcation (the black line in Figs.
+\ref{fig:PE}a-c) occurs at lower values of the straying rate $m$ with decreased trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
 
-Trait heritability has a large impact on the sensitivity of the straying rate to both the aggregate population steady state density ($N^*_T=N^*_1+N^*_2$; Fig.
-\ref{fig:PE}a) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N=|N^*_1-N^*_2|$; Fig.
-\ref{fig:PE}b).
-Greater trait heritability results in a larger decline in $N_T^*$ for increasing straying rates $m$, but leads to only moderate changes to $\Delta N$.
+Trait heritability has a large impact on the sensitivity of the straying rate to both the aggregate population steady state density ($N^*_T=N^*_1+N^*_2$; figure \ref{fig:PE}a) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N=|N^*_1-N^*_2|$; figure \ref{fig:PE}b).
+Greater trait heritability results in a faster decline in $N_T^*$ with increasing straying rates $m$, but leads to only moderate changes to $\Delta N$.
 Conversely, if the trait is less heritable, an increase in the straying rate has little impact on the total biomass density but contrastingly large effects on $\Delta N$.
@@
 %In the region where there is a single steady state among both populations, we find an low-intermediate portfolio effect ca. ${\rm PE}=1.4$, primarily due to the elevated mean values of $N_T$.
-As the fold bifurcation is reached with greater rates of straying, the portfolio effect increases sharply due to an explosion in variance of both donor and recipient populations ${\rm VAR}(N_{i,j})$.
-This increase in variance is the product of a dynamical process known as critical slowing down that occurs near fold bifurcations \cite{Scheffer:2009gg}, and may serve as an early warning signal of an approaching phase transition \cite{Scheffer:2009gg,Anonymous:2013br,Dakos:2014br}.
+As the fold bifurcation is approached with increasing $m$, the portfolio effect increases sharply due to an amplification in variance within both donor and recipient populations ${\rm VAR}(N_{i,j})$.
+This variance increase is the product of a dynamical process known as critical slowing down that occurs near fold bifurcations \cite{Scheffer:2009gg}, a phenomenon that some have suggested may serve as an early warning indicator for approaching phase transitions \cite{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br}.
 For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}a-c), where alternative steady states occur, the portfolio effect declines steadily as the CV of $N_T$ increases.
-The decline over $m$ is more gradual if trait heritability is low, and steeper if trait heritability is high.
+The decline over $m$ is more gradual if trait heritability is low, and steeper if trait heritability is high (figure \ref{fig:PE}c).
 
@@
 % Return time as a measure of system persistence and relation to PE
-As the portfolio effect is highly sensitive to the rate of straying between populations, so is the time the system requires to recover following a large disturbance.
-We measured the time required for the system $N_T$ to relax to its steady state following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} are equivalent in the single steady state regime); (\emph{iii}) near-collapse of both populations where just 1.0\% of both populations survive.
-In general, the average-CV portfolio effect is negatively correlated with recovery time (Fig.
-\ref{fig:PE}d), indicating that for our system both measures are valuable indicators of metapopulation persistence.
+As the portfolio effect is highly sensitive to the rate of straying between populations, so is the time required for the system to recover to a steady state following a large disturbance.
+We measured the time required for the system $N_T$ to relax to its steady state following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} are equivalent if the system is in the single steady state regime); (\emph{iii}) near-collapse of both populations where just 1.0\% of each survives.
+In general, we find that the average-CV portfolio effect is negatively correlated with recovery time (figure \ref{fig:PE}d), indicating that for our system both measures are valuable indicators of metapopulation persistence.
 Because we can assess the time to recovery in response to the various disturbance types described above, this allows us to gain an in-depth perspective into the fragility of the metapopulation as a function of straying rate.
 
-At low rates of straying both populations are subject to the same steady state, such that extinction of either population results in the same recovery time.
-At low rates of straying and low trait heritability, the time required to recover following the extinction of either population is generally less than or equal to the recovery time required after the near-collapse of both populations (Fig.
-\ref{fig:relax}a).
-In the alternative stable state regime, extinction of the smaller population results in the shortest recovery time, whereas as $m$ increases, extinction of the larger population results in the shortest recovery time.
-This switch is in part due to a large burst in biomass of the surviving (large) population when the smaller of the two becomes extinct at high straying rates, which extends the time required for the smaller population (and thus the system as a whole) to recover.
-Increased trait heritability flips this relationship (Fig.
-\ref{fig:relax}b), such that systems suffering near-collapse of both populations recover more quickly than if either population is lost at low straying rates.
-As straying rates increase, the system recovers more quickly if the small population is lost, whereas the time to recovery is maximized if both populations suffer near-collapse.
-Near the onset of the fold bifurcation, recovery times increase drastically regardless of straying rate, heritability, or disturbance type.
-For straying rates in the alternative steady state regime, extinction of the smaller population consistently results in the shortest time to recovery.
+At low rates of straying both populations are subject to the same steady state, such that extinction of either population results in the same recovery time, and when trait heritability is low, the time required to recover following the extinction of either population is generally less than or equal to the recovery time after near-collapse of both (figure \ref{fig:relax}a).
+In the alternative stable state regime where straying rates are higher, extinction of the smaller population results in the shortest recovery time, whereas as $m$ increases, extinction of the larger population results in the shortest recovery time.
+Because the mean trait values of both populations are skewed towards those of the high-density site, when the low-density population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into the recovering extends the length of time required for it to return to its steady state (figure \ref{fig:relaxtraj_ldlh}).
+Alternatively, when the high-density population collapses under high straying rates, the low-density population suffers loss as well due to the absence of incoming strays such that both populations must recover (figure \ref{fig:relaxtraj_hdlh}).
+
+Increased trait heritability flips this relationship by increasing the response of trait evolution to changes in demography (figure \ref{fig:relax}b).
+When straying rates are low, metapopulations that suffer near-collapse of both populations recover more quickly than systems suffering extinctions of either population.
+When straying rates are higher such that the system is in the alternative stable state regime, the metapopulation always recovers more quickly if the small population is lost (figure \ref{fig:relaxtraj_ldhh}), whereas the time to recovery is maximized if both populations suffer near-collapse.
+When the high-density population goes extinct, the stronger response of selection permits the surviving low-density population to grow quickly and fill the high-density state (figure \ref{fig:relaxtraj_hdhh}).
+Near the onset of the fold bifurcation, recovery times increase explosively regardless of straying rate, heritability, or disturbance type.
+%For straying rates in the alternative steady state regime, extinction of the smaller population consistently results in the shortest time to recovery.
 
@@
 Increased rates of straying lowers phenotypic diversity ($\Delta \mu^* = |\mu_i(t)-\mu_j(t)|$, evaluated at the steady state) because both local and remote populations are increasingly homogenized.
-The sensitivity of this phenotypic homogenization to straying is increased with lower heritability because the selective forces acting against trait means far from the local optima are lessened for lower $h^2$.
-Less intuitively, we observe a discrete jump towards low phenotypic diversity as the fold bifurcation is crossed (Fig. \ref{fig:traitdiff}).
+The loss of phenotypic diversity with increased straying is greater if trait heritability is low because the selective forces acting against trait means far from the local optima are lessened.
+% with decreased $h^2$.
+Less intuitively, we observe a discrete jump towards low phenotypic diversity as the fold bifurcation is crossed (figure \ref{fig:traitdiff}).
 Although the development of alternative stable states elevates the portfolio effect due to the variance-dampening effects of the aggregate, entering this dynamic regime also results in a substantial decline in phenotypic diversity, which may have less predictable adverse effects on the population.
@@
 With the onset of straying, we find that increasingly divergent trait optima generally lower $N_T$ and exaggerate $\Delta N$, such that the biomass distribution becomes increasingly uneven.
-The impact of habitat heterogeneity on the portfolio effect is more complex, serving to emphasize the nonlinear relationship between the stray rate and the PE. %regardless of heritability.
+%I've checked this - it's a theshold type (logistic) response Supp fig?
+The impact of habitat heterogeneity on the portfolio effect is more complex, serving to emphasize the nonlinear relationship between the straying rate and the PE. %regardless of heritability.
 %The rate of straying that gives rise to a large PE at the fold bifurcation is conditioned on the difference in optimal trait values between the local and remote populations, which we call $\Delta\theta = \sqrt{(\theta_i - \theta_j)^2}$.
-As habitat heterogeneity increases, the fold bifurcation occurs at lower straying rates whereas the associated jump in the PE is greater (Fig.
-\ref{fig:thetaPE}), decreasing the recovery time (Appendix Fig.
-\ref{fig:thetarelax}).
-For greater straying rates, habitat heterogeneity erodes the PE and slows recovery time (Fig. \ref{fig:thetaPE}, Appendix Fig. \ref{fig:thetarelax})
+As habitat heterogeneity increases, the fold bifurcation occurs at lower straying rates whereas the magnitude of increase in the PE is greater (figure \ref{fig:thetaPE}), decreasing the recovery time (figure \ref{fig:relaxtheta}).
+For greater straying rates, habitat heterogeneity erodes the PE and slows recovery time (figure \ref{fig:thetaPE}a, figure \ref{fig:relaxtheta})
 Taken together, greater $\Delta\theta$ promotes persistence when straying rates are low, but erodes persistence when straying rates are high.
-%This means that alternative stable states can emerge with relatively little dispersal between sites (Fig. \ref{fig:thetaPE}), however the increased PE (and shorter recovery times).
+%This means that alternative stable states can emerge with relatively little dispersal between sites (figure \ref{fig:thetaPE}), however the increased PE (and shorter recovery times).
 \\
@@
 \ref{eq:ddm}).
-%Because our analyses concern steady state conditions, the inclusion of density-dependent straying does not have a large impact on the qualitative nature of our findings (Appendix Fig. \ref{fig:ddm}).
+%Because our analyses concern steady state conditions, the inclusion of density-dependent straying does not have a large impact on the qualitative nature of our findings (figure \ref{fig:ddm}).
 Density dependent straying alters the straying rate at the steady state because $0 < m^* < m_0$, and this serves to rescale both the strength of the PE as well as the recovery time following an induced disturbance but does not change the qualitative nature of our findings.
@@
 A range of $(m_i^*,m_j^*)$ values are obtained by varying $m_0$, which is positively and linearly related to $(m_i^*,m_j^*)$.
-For instance, the recovery times for systems with constant and density dependent straying both show similar trends for a given $m$ or $m^*$ (Fig.
-\ref{fig:relax}c,d), however there are some distinct quantitative differences.
-When straying rate and trait heritability are low, near-collapse of both populations result in longer than expected recovery times, whereas the recovery times in the alternative stable state regime (higher $m^*$) are similar (Fig.
-\ref{fig:relax}c).
+For instance, the recovery times for systems with constant and density dependent straying both show similar trends for a given $m$ or $m^*$ (figure \ref{fig:relax}c,d), however there are some distinct quantitative differences.
+When straying rate and trait heritability are low, near-collapse of both populations result in longer than expected recovery times, whereas the recovery times in the alternative stable state regime (higher $m^*$) are similar (figure \ref{fig:relax}c).
 There is greater similarity in the recovery times for constant and density dependent $m$ systems when trait heritability is high.
@@
 A metapopulation in the alternative stable state regime has two straying rates: a lower straying rate associated with the high-density population, and a higher straying rate associated with the low-density population.
-We find that the resultant portfolio effect approximates that generated by a system with a similar constant straying rate, such that ${\rm PE}(m) \approx {\rm PE}(m^*_{\rm high-density})$ (Fig.
-\ref{fig:thetaPE}b).
-As the steady state straying rate increases, the portfolio effect follows a similar trajectory as shown for metapopulations with constant straying rates (Fig.
-\ref{fig:thetaPE}a), though is generally higher, promoting persistence.
+We find that the resultant portfolio effect approximates that generated by a system with a similar constant straying rate, such that ${\rm PE}(m) \approx {\rm PE}(m^*_{\rm high-density})$ (figure \ref{fig:thetaPE}b).
+As the steady state straying rate increases, the portfolio effect follows a similar trajectory as shown for metapopulations with constant straying rates (figure \ref{fig:thetaPE}a), though is generally higher, promoting persistence.
 RETURN TIMES
@@
 % 
-% We find that the by-definition lower straying rate of the high-density population (left-most of the paired points in Fig.
-\ref{fig:thetaPE}b) largely determines the portfolio effect for the metapopulation.
+% We find that the by-definition lower straying rate of the high-density population (left-most of the paired points in figure \ref{fig:thetaPE}b) largely determines the portfolio effect for the metapopulation.
 % In other words, 
@@
 Increased heritability results in greater response to natural selection on the metapopulation, increasing the coupling between evolutionary and ecological dynamics. 
-Our minimal model of straying between two populations shows that greater $h^2$ buffers the metapopulation against the emergence of alternative stable states, but magnifies the negative effects of alternative stable states (captured by increased $\Delta N$ and decreased PE; Fig.
-\ref{fig:PE}B,C) once this regime is encountered at higher straying rates.
-We note that although the rate of straying has been shown to be density dependent (REFS), the qualitative results of our model are relatively insensitive to this dynamic (Fig.
-\ref{fig:diffddm}) and we limit discussion to the simplified case of a constant stray rate $m$ with the understanding that our findings also apply to the case density dependent stray rates $m(t)$.
+Our minimal model of straying between two populations shows that greater $h^2$ buffers the metapopulation against the emergence of alternative stable states, but magnifies the negative effects of alternative stable states (captured by increased $\Delta N$ and decreased PE; figure \ref{fig:PE}B,C) once this regime is encountered at higher straying rates.
+We note that although the rate of straying has been shown to be density dependent (REFS), the qualitative results of our model are relatively insensitive to this dynamic (figure \ref{fig:diffddm}) and we limit discussion to the simplified case of a constant stray rate $m$ with the understanding that our findings also apply to the case density dependent stray rates $m(t)$.
 
@@
 Our results show that trait heritability has large effects on both metapopulation persistance as well as phenotypic diversity, particularly in the regime of low-to-moderate straying rates.
+
+Recovery times: largest perturbations lead to the longest recovery times \cite{Ovaskainen:2002il}.
+This is not true in our systems when (\emph{i}) heritability and (\emph{ii}) straying are low. Why?
 
@@
 
-\clearpage
-
-\setcounter{figure}{0}
-\section*{Appendix}
+\beginsupplement
 
@@
 \centering
-\includegraphics[width=0.5\textwidth]{figs2/fig_relaxtheta.pdf}
+\includegraphics[width=0.35\textwidth]{figs2/fig_relaxtheta.pdf}
 \caption{
@@
 } \label{fig:relaxtheta}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.35\textwidth]{figs2/fig_relax_small.pdf}
+\caption{
+Extinction of low-density population with a high constant straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
+Black line marks the calculated point of recovery post-perturbation.
+Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
+} \label{fig:relaxtraj_ldlh}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.35\textwidth]{figs2/fig_relax_large.pdf}
+\caption{
+Extinction of high-density population with a high straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
+Black line marks the calculated point of recovery post-perturbation.
+Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
+} \label{fig:relaxtraj_hdlh}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.35\textwidth]{figs2/fig_relax_small_highh.pdf}
+\caption{
+Extinction of low-density population with a high constant straying rate $m=0.4$ and high trait heritability $h^2=0.8$ (see figure \ref{fig:relax}a).
+Black line marks the calculated point of recovery post-perturbation.
+Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
+} \label{fig:relaxtraj_ldhh}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.35\textwidth]{figs2/fig_relax_large_highh.pdf}
+\caption{
+Extinction of high-density population with a high straying rate $m=0.4$ and high trait heritability $h^2=0.8$ (see figure \ref{fig:relax}a).
+Black line marks the calculated point of recovery post-perturbation.
+Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
+} \label{fig:relaxtraj_hdhh}
 \end{figure}

### 2017-05-24 model/KevinExtinction.jl

@@ -1,80 +1,82 @@
-# using Distributions
-# using RCall
-# using JLD
-# using HDF5
-# 
-# include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve.jl")
-# include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve_ddm.jl")
-# 
-# 
-# include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolveExtinct.jl")
-# include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/timeSS.jl")
-# include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/movingaverage.jl")
-# 
-# 
-# 
-# tmax=10000;
-# z=0.5;
-# rmax=2.0;
-# beta=0.001;
-# theta1=5.0;
-# thetadiff=5.0;
-# tau=1.0;
-# h=0.5;
-# sigmaE=0;
-# sigmaG=1;
-# perror=0.01;
-# refuge=0.01;
-# m=0.25
-# 
-# extpop = "large";
-# t_ext = Int64(round(tmax/2));
-# 
-# n1, n2, x1, x2, w1, w2 = 
-# KevinEvolveExtinct(tmax, 
-# z, 
-# rmax,
-# beta,
-# theta1,
-# thetadiff,
-# tau,
-# h,
-# sigmaE,
-# sigmaG,
-# m,
-# perror,
-# extpop,
-# t_ext,
-# refuge
-# );
-# 
-# 
-# R"""
-# library(RColorBrewer)
-# pal = brewer.pal(3,'Set1')
-# par(mfrow=c(2,1))
-# plot($n1,type='l',col=pal[1],xlim=c())
-# lines($n2,col=pal[2])
-# plot($x1,type='l',col=pal[1],ylim=c(4,12))
-# lines($x2,col=pal[2])
-# """
-# 
-# 
-# t_ss, relaxtime = timeSS(n1,n2,t_ext);
-# 
-# 
-# R"""
-# library(RColorBrewer)
-# pal = brewer.pal(3,'Set1')
-# par(mfrow=c(2,1),mai = c(0.8, 0.8, 0.1, 0.1))
[…]

### 2017-05-24 model/habitathetero.jl

@@ -76,6 +76,14 @@ end
 
 save(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data2/data_habitathetero.jld"),"n1mean",n1mean,"n2mean",n2mean,"x1mean",x1mean,"x2mean",x2mean,"pe",pe);
 
+d = load(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data2/data_habitathetero_ext.jld"));
+#This loads the dictionary
+n1mean = d["n1mean"];
+n2mean = d["n2mean"];
+x1mean = d["x1mean"];
+x2mean = d["x2mean"];
+pe = d["pe"];
+
 mediandiff = zeros(Float64,length(thetadiffvec));
 for k=1:length(thetadiffvec)
   mediandiff[k] = median(abs(n1mean[k,:] - n2mean[k,:]));
@@ -117,13 +125,25 @@ plot($thetadiffarray,$aggarray,pch='.')
 #Analysis over m & theta divergence
 mvec = collect(0.0001:0.001:0.5);
 thetadiffvec = collect(0.0:0.1:10.0);
+hvec = collect(0.2:0.2:0.8);
 
-thetadiffarray=SharedArray(Float64,length(thetadiffvec),length(mvec));
+thetadiffarray=SharedArray(Float64,length(thetadiffvec),length(hvec),length(mvec));
 
-rt=SharedArray(Float64,length(thetadiffvec),length(mvec));
-rt_ddm=SharedArray(Float64,length(thetadiffvec),length(mvec));
-m1mean=SharedArray(Float64,length(thetadiffvec),length(mvec));
-m2mean=SharedArray(Float64,length(thetadiffvec),length(mvec));
+n1mean=SharedArray(Float64,length(thetadiffvec),length(hvec),length(mvec));
+n2mean=SharedArray(Float64,length(thetadiffvec),length(hvec),length(mvec));
+x1mean=SharedArray(Float64,length(thetadiffvec),length(hvec),length(mvec));
+x2mean=SharedArray(Float64,length(thetadiffvec),length(hvec),length(mvec));
+pe=SharedArray(Float64,length(thetadiffvec),length(hvec),length(mvec));
[…]

### 2017-05-25 Manuscript/msv2.tex
mtheta recovery analysis, ms

@@
 %\title{Community assembly and dynamics by the deconstruction of species interactions}
-\title{Lost in space: the eco-evolutionary impacts of straying on metapopulation persistence}
+\title{Lost in space: the eco-evolutionary impacts of straying on metapopulation robustness}
 \author{Justin D. Yeakel${}^{1,2,*}$, Jean Phillipe Gibert${}^{1}$, Sean Anderson${}^{3}$, Peter Westley${}^{4}$, \& Jon W. Moore${}^{5}$ \\
@@
 %Alternative stable state
-Here we ask the overarching question of how collective behavior mediated dispersal and gene flow interact to influence the persistence of locally adapted populations and persistence of those populations.
+Here we ask the overarching question: how does collective behavior mediated dispersal and gene flow interact to influence the robustness of locally adapted populations?
 To address this question we construct a minimal eco-evolutionary model of two populations occupying different sites that are linked by straying individuals, each of which with an associated trait distribution subject to natural selection determined by local conditions.
@@
 We first show that specific rates of straying and trait heritability can have large effects on the qualitative dynamics of populations over time, in many cases giving rise to alternative steady states where one site is pushed towards very low biomass.
-The emergence of alternative steady states results in a nonlinear response of the portfolio effect, suggesting that metapopulation persistence can be quite sensitive to the combined influences of collective straying across a selective landscape.
+The emergence of alternative steady states results in a nonlinear response of the portfolio effect as well as the time required for the metapopulation to recover after an induced disturbance, suggesting that metapopulation robustness can be quite sensitive to the combined influences of dispersal across a selective landscape.
 
@@
 A) The steady state densities of $N_1$ and $N_2$ as a function of a constant stray rate $m$.
+Which population attains the low- or high-density state is random due to small applied fluctuations in the initial conditions.
 B) The steady state trait values measured as $\theta_i - x_i$, as a function of a constant stray rate $m$. 
@@
 
-\noindent {\bf (c) Measuring metapopulation persistence}
-\noindent We evaluated metapopulation persistence by measuring the average-CV portfolio effect (PE) \cite{Anderson:2014cx,Schindler:2015gf} as well as the time required for the system to return to a steady state following an induced disturbance to one or both of the populations \cite{Ovaskainen:2002il}.
+\noindent {\bf (c) Measuring metapopulation robustness}
+\noindent We evaluated metapopulation robustness by measuring the average-CV portfolio effect (PE) \cite{Anderson:2014cx,Schindler:2015gf} as well as the time required for the system to return to a steady state following an induced disturbance to one or both of the populations \cite{Ovaskainen:2002il}.
 The average-CV portfolio effect is, as the name implies, the average CV across each population divided by the CV of the aggregate \cite{Anderson:2013gb}, such that
@@
 As the CV of $N_T$ decreases relative to that of the constituent populations, $\langle{\rm PE}\rangle > 1$, and the metapopulation is presumed to become more stable.
-{\color{red} A similar measure of metapopulation persistence can be evaluated by measuring the synchronization $\phi$ between populations, where 
+{\color{red} A similar measure of metapopulation robustness can be evaluated by measuring the synchronization $\phi$ between populations, where 
 \begin{equation}
@@
 %Although there are more robust ways to measure the portfolio effect (in particular for comparing the PE across species with different mean-variance relationships; REF: Anderson), its only utlity here is to provide a simple statistical measure of metapopulation persistence.
-A more direct way to measure system persistence is to measure the time that it takes the system (measured as the aggregate biomass $N_T$) to recover its steady state abundance following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
+A more direct way to measure system robustness is to measure the time that it takes the system (measured as the aggregate biomass $N_T$) to recover its steady state abundance following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
 Measuring the time that it takes for a perturbed system to relax also permits a more detailed perspective of the potential fragility of the metapopulation.
 For example, if populations settle to alternative steady states (alternative steady states in our model requiring one population to be high-density and one low-density), comparing recovery times after a disturbance applied to the high, low, and/or both populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery.
+Throughout, we will refer to an increase in the portfolio effects and/or reduction in recovery times as promoting metapopulation robustness, which is expected to have a positive effect on persistence.
 \\
@@
 We measured the time required for the system $N_T$ to relax to its steady state following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} are equivalent if the system is in the single steady state regime); (\emph{iii}) near-collapse of both populations where just 1.0\% of each survives.
-In general, we find that the average-CV portfolio effect is negatively correlated with recovery time (figure \ref{fig:PE}d), indicating that for our system both measures are valuable indicators of metapopulation persistence.
+In general, we find that the average-CV portfolio effect is negatively correlated with recovery time (figure \ref{fig:PE}d), indicating that for our system both measures are valuable indicators of metapopulation robustness.
 Because we can assess the time to recovery in response to the various disturbance types described above, this allows us to gain an in-depth perspective into the fragility of the metapopulation as a function of straying rate.
@@
 
-\noindent{\bf (b) The role of habitat heterogeneity on metapopulation persistence}\\ 
+\noindent{\bf (b) The influence of habitat heterogeneity on metapopulation robustness}\\ 
 \noindent Increasing differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater regional differences in the conditions that favor alternative trait complexes, which we interpret here as greater habitat heterogeneity.
+Although trait heritability among salmonids is variable, most life history traits show $h^2 <0.5$ \cite{Carlson:2008hl}, and we largely focus additional analyses on that range.
 If both populations are isolated, natural selection will direct the mean trait values of both populations towards their respective optima, such that $\mu_i(t) \rightarrow \theta_i$ as $t\rightarrow\infty$.
 %difference in optimal trait values between the local and remote populations, which we call .
-With the onset of straying, we find that increasingly divergent trait optima generally lower $N_T$ and exaggerate $\Delta N$, such that the biomass distribution becomes increasingly uneven.
+With the onset of straying, we find that increasingly divergent trait optima generally lower $N_T$ and exaggerate $\Delta N$ (figure \ref{fig:thetadiffN}), such that the biomass distribution becomes increasingly uneven.
 %I've checked this - it's a theshold type (logistic) response Supp fig?
-The impact of habitat heterogeneity on the portfolio effect is more complex, serving to emphasize the nonlinear relationship between the straying rate and the PE. %regardless of heritability.
+The impact of habitat heterogeneity on the portfolio effect and recovery time is more complex, serving to emphasize the nonlinear relationship between rates of straying and metapopulation robustness.
+%regardless of heritability.
 %The rate of straying that gives rise to a large PE at the fold bifurcation is conditioned on the difference in optimal trait values between the local and remote populations, which we call $\Delta\theta = \sqrt{(\theta_i - \theta_j)^2}$.
-As habitat heterogeneity increases, the fold bifurcation occurs at lower straying rates whereas the magnitude of increase in the PE is greater (figure \ref{fig:thetaPE}), decreasing the recovery time (figure \ref{fig:relaxtheta}).
-For greater straying rates, habitat heterogeneity erodes the PE and slows recovery time (figure \ref{fig:thetaPE}a, figure \ref{fig:relaxtheta})
-Taken together, greater $\Delta\theta$ promotes persistence when straying rates are low, but erodes persistence when straying rates are high.
+As habitat heterogeneity increases, alternative stable states appear for lower straying rates -- with the crossing of the fold bifurcation, accompanied by a peak in the PE -- whereas the magnitude of increase in the PE increases (figure \ref{fig:thetaPE}a), reducing recovery time (figure \ref{fig:relaxtheta}).
+For increased rates of straying, greater habitat heterogeneity erodes the PE and increases the recovery time (figure \ref{fig:thetaPE}a, figure \ref{fig:relaxtheta})
+Taken together, habitat heterogeneity, as measured as the differences in trait optima between two habitats, promotes robustness when straying rates are low, but erodes robustness when straying rates are high.
 %This means that alternative stable states can emerge with relatively little dispersal between sites (figure \ref{fig:thetaPE}), however the increased PE (and shorter recovery times).
@@
 \noindent{\bf (c) The effects of density dependent straying}\\
-\noindent If we assume that the rate of straying is density dependent, the probability that an individual strays $m_0$ determines the rate of straying within the population, such that $m(t)$ becomes lower as $N(t)$ increases (Eq.
+\noindent If we assume that the rate of straying is density dependent, the probability that an individual strays $m_0$ determines the rate of straying within the population, such that $m(t)$ becomes lower as $N(t)$ increases due to the effects of collective decision-making \cite{Berdahl:2016dx} (Eq.
 \ref{eq:ddm}).
 %Because our analyses concern steady state conditions, the inclusion of density-dependent straying does not have a large impact on the qualitative nature of our findings (figure \ref{fig:ddm}).
-Density dependent straying alters the straying rate at the steady state because $0 < m^* < m_0$, and this serves to rescale both the strength of the PE as well as the recovery time following an induced disturbance but does not change the qualitative nature of our findings.
-Moreover, in the alternative stable state regime, because the recipient and donor population exist at different steady state densities, they have different rates of straying such that a given value of the portfolio effect is now associated with a pair of straying rates $(m_i^*,m_j^*)$.
-A range of $(m_i^*,m_j^*)$ values are obtained by varying $m_0$, which is positively and linearly related to $(m_i^*,m_j^*)$.
-For instance, the recovery times for systems with constant and density dependent straying both show similar trends for a given $m$ or $m^*$ (figure \ref{fig:relax}c,d), however there are some distinct quantitative differences.
-When straying rate and trait heritability are low, near-collapse of both populations result in longer than expected recovery times, whereas the recovery times in the alternative stable state regime (higher $m^*$) are similar (figure \ref{fig:relax}c).
+Density dependence alters the straying rate at steady state population densities because $0 < m^* < m_0$, and this serves to rescale both the strength of the PE as well as the recovery time, but importantly does not change the qualitative nature of our findings for constant straying rates.
+In the alternative stable state regime, because the each population exists at different steady state densities, there are likewise two alternative straying rates $(m_i^*,m_j^*)$: the higher straying rate is associated with the low-density population, and the lower straying rate is associated with the high-density population.
+We assessed metapopulation robustness across a range of $(m_i^*,m_j^*)$ values by varying $m_0$, which is positively and linearly related to $(m_i^*,m_j^*)$.
+
+%Recovery times
+The recovery times for systems with constant and density dependent straying both show similar trends for a given $m$ and $m^*$ (figure \ref{fig:relax}c,d), however there are some distinct quantitative differences.
+When the straying rate and trait heritability are low, near-collapse of both populations result in longer than expected recovery times, whereas in the alternative stable state regime (higher $m^*$), the recovery times for different disturbance types are very similar to systems with a constant $m$ (figure \ref{fig:relax}c; note difference in x-axis scales).
 There is greater similarity in the recovery times for constant and density dependent $m$ systems when trait heritability is high.
 
-A metapopulation in the alternative stable state regime has two straying rates: a lower straying rate associated with the high-density population, and a higher straying rate associated with the low-density population.
-We find that the resultant portfolio effect approximates that generated by a system with a similar constant straying rate, such that ${\rm PE}(m) \approx {\rm PE}(m^*_{\rm high-density})$ (figure \ref{fig:thetaPE}b).
-As the steady state straying rate increases, the portfolio effect follows a similar trajectory as shown for metapopulations with constant straying rates (figure \ref{fig:thetaPE}a), though is generally higher, promoting persistence.
-RETURN TIMES
-This suggests that density dependent straying...
+As with recovery time, we find that the portfolio effects generated in systems with density dependent straying are qualitatively similar to systems with constant straying, however there are some important quantitative differences.
+First, the PE associated with the high-density (low $m^*$) population is the same as that for a system with a constant $m$ (figure \ref{fig:thetaPE}b).
+As such, the difference in $m^*$ among low- and high-density populations appears to matter less than the straying generated by the high-density population.
+However, as $m^*$ increases, we observe a higher PE than that generated by systems with constant $m$.
+This higher PE at increased $m^*$ is mirrors recovery times that are also lower than those observed for systems with constant $m$ (figure \ref{fig:relax}).
+Together this suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our minimal model, we do find evidence that it can promote robustness, particularly when populations are low and straying is correspondingly high.
 % 
@@
 
+That recovery times and PE are similar for constant $m$ vs. density dependent $m$ suggests that both transient dynamics as well as asymptotic dynamics are not largely impacted.
+
 %Our results in the context of Turing bifurcation
@@
 \beginsupplement
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.35\textwidth]{figs2/fig_relaxtheta.pdf}
-\caption{
-Recovery time as a function of straying rate $m$ and habitat heterogeneity $\Delta\theta$.
-} \label{fig:relaxtheta}
-\end{figure}
 
@@
 
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.35\textwidth]{figs2/fig_thetadiffN.pdf}
+\caption{
+Median difference in population densities taken over the straying rate as a function of habitat heterogeneity $\Delta\theta$.
+Solid lines are for constant $m$; dashed lines are for density dependent $m$} \label{fig:thetadiffN}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.35\textwidth]{figs2/fig_relaxtheta.pdf}
+\caption{
+Recovery time as a function of straying rate $m$ and habitat heterogeneity $\Delta\theta$.
+} \label{fig:relaxtheta}
+\end{figure}
+
 % \begin{figure*}[h!]

### 2017-05-25 model/KevinSimulations_mtheta2.jl

@@ -115,6 +115,12 @@ plot($mvec,$pe)
 
 @everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/qualsfunc.jl")
 
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolveExtinct.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolveExtinct_ddm.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/timeSS.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/movingaverage.jl")
+
+
 
 #Analysis over m & theta divergence
 mvec=collect(0.0:0.001:0.45);
@@ -198,6 +204,151 @@ indmvec,
 hvec
 );
 
+
+
+
+
+
+#Recovery time sims
+#Analysis over m
+tmax=10000;
+
+mvec = collect(0.0001:0.001:0.5);
+pvec = ["small","large","both"];
+
+rt = SharedArray{Float64}(reps,length(mvec),length(pvec));
+rt_ddm = SharedArray{Float64}(reps,length(mvec),length(pvec));
+m1mean = SharedArray{Float64}(reps,length(mvec),length(pvec));
+m2mean = SharedArray{Float64}(reps,length(mvec),length(pvec));
+
+reps = 100;
+z=0.5;
+rmax=2.0;
+beta=0.001;
+theta1=5.0;
+thetascale=2.0;
+tau=1.0;
+C=1000;
+h=0.2;
+sigmaE=0;
+sigmaG=1;
+perror=0.01;
+refuge=0.01;
+burnin = 0.8;
+t_ext = Int64(round(tmax/2));
+
+@sync @parallel for r=1:reps
+  for j=1:length(pvec)
+    extpop = pvec[j]
+    for i=1:length(mvec)
+      
+      m = mvec[i];
+      a0 = 1-m;
+      
+      thetadiff = (1-2*m)/(thetascale*m);
+      
+      n1, n2, x1, x2, w1, w2 = 
+      KevinEvolveExtinct(tmax, 
[…]

### 2017-05-25 model/habitathetero.jl

@@ -125,7 +125,7 @@ plot($thetadiffarray,$aggarray,pch='.')
 #Analysis over m & theta divergence
 mvec = collect(0.0001:0.001:0.5);
 thetadiffvec = collect(0.0:0.1:10.0);
-hvec = collect(0.2:0.2:0.8);
+hvec = collect(0.1:0.1:0.9);
 
 thetadiffarray=SharedArray(Float64,length(thetadiffvec),length(hvec),length(mvec));
 
@@ -153,7 +153,6 @@ theta1=5.0;
 tau=1.0;
 sigmaE=0;
 sigmaG=1;
-h=0.2
 perror=0.01;
 refuge=0.01;
 extpop="both";
@@ -234,15 +233,15 @@ t_ext = Int64(round(tmax/2));
         t_ext,
         refuge
         )
-        n1trim = n1[Int64(floor(tmax*burnin)):tmax];
-        n2trim = n2[Int64(floor(tmax*burnin)):tmax];
-        x1trim = x1[Int64(floor(tmax*burnin)):tmax];
-        x2trim = x2[Int64(floor(tmax*burnin)):tmax];
+        n1trim_ddm = n1_ddm[Int64(floor(tmax*burnin)):tmax];
+        n2trim_ddm = n2_ddm[Int64(floor(tmax*burnin)):tmax];
+        x1trim_ddm = x1_ddm[Int64(floor(tmax*burnin)):tmax];
+        x2trim_ddm = x2_ddm[Int64(floor(tmax*burnin)):tmax];
         
-        n1mean_ddm[k,i,j] = mean(n1trim);
-        n2mean_ddm[k,i,j] = mean(n2trim);
-        x1mean_ddm[k,i,j] = theta1-mean(x1trim);
-        x2mean_ddm[k,i,j] = (theta1+thetadiff)-mean(x2trim);
+        n1mean_ddm[k,i,j] = mean(n1trim_ddm);
+        n2mean_ddm[k,i,j] = mean(n2trim_ddm);
+        x1mean_ddm[k,i,j] = theta1-mean(x1trim_ddm);
+        x2mean_ddm[k,i,j] = (theta1+thetadiff)-mean(x2trim_ddm);
         
         pe_ddm[k,i,j] = mean([(std(n1trim)/mean(n1trim)),(std(n2trim)/mean(n2trim))])*
[…]

### 2017-05-25 Manuscript/msv2.tex
revised mtheta fig

@@
   \centering
-  \includegraphics[width=0.8\textwidth]{figs2/fig_MDPE_hm_mtheta.pdf}
+  \includegraphics[width=0.8\textwidth]{figs2/fig_MDPE_hm_mtheta_rt.pdf}
   \caption{
@@
 \noindent{\bf (d) Distance dependent straying and habitat heterogeneity}\\ 
-\noindent We have so far treated $\Delta\theta$ and $m$ as independent parameters, however we may also assume that if environmental heterogeneity increases with distance, the rate of straying may be expected to decline given that individuals are less likely to stray into distant habitats.
+\noindent We have so far treated $\Delta\theta$ and $m$ as independent parameters, however we may also assume that if environmental heterogeneity increases with distance, the rate of straying may be expected to decline given that individuals will stray less into distant habitats.
 %-- in particular North-South difference if trait optimality is largely temperature-dependent --
-If we incorporate this interdependence of $m$ and $\Delta\theta$, low rates of straying would correspond to mixing dissimilar (distant) populations, and high rates of straying would correspond to mixing similar (nearby) populations.
+If we incorporate this interdependence of $m$ and $\Delta\theta$, low rates of straying correspond to mixing dissimilar (distant) populations, and high rates of straying would correspond to mixing similar (nearby) populations.
 
-Here we find that alternative stable states now appear for low straying rates ($m<0.15$) and over any value of trait heritability $h^2<1$.
+Here we find that alternative stable states now appear for low straying rates ($m<0.15$).
 As the straying rate increases, a single stable state emerges with a correspondingly high $N_T$.

### 2017-05-25 model/KevinSimulations_mtheta2.jl

@@ -119,6 +119,7 @@ plot($mvec,$pe)
 @everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolveExtinct_ddm.jl")
 @everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/timeSS.jl")
 @everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/movingaverage.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/bifdet.jl")
 
 
 
@@ -200,7 +201,7 @@ pe = d["pe"];
 bifvalue = bifdet(
 n1mean,
 n2mean,
-indmvec,
+mvec,
 hvec
 );
 
@@ -213,13 +214,13 @@ hvec
 #Analysis over m
 tmax=10000;
 
-mvec = collect(0.0001:0.001:0.5);
+mvec2 = collect(0.0001:0.001:0.5);
 pvec = ["small","large","both"];
 
-rt = SharedArray{Float64}(reps,length(mvec),length(pvec));
-rt_ddm = SharedArray{Float64}(reps,length(mvec),length(pvec));
-m1mean = SharedArray{Float64}(reps,length(mvec),length(pvec));
-m2mean = SharedArray{Float64}(reps,length(mvec),length(pvec));
+rt = SharedArray{Float64}(reps,length(mvec2),length(pvec));
+rt_ddm = SharedArray{Float64}(reps,length(mvec2),length(pvec));
+m1mean = SharedArray{Float64}(reps,length(mvec2),length(pvec));
+m2mean = SharedArray{Float64}(reps,length(mvec2),length(pvec));
 
 reps = 100;
 z=0.5;
@@ -242,7 +243,7 @@ t_ext = Int64(round(tmax/2));
     extpop = pvec[j]
     for i=1:length(mvec)
       
-      m = mvec[i];
+      m = mvec2[i];
       a0 = 1-m;
       
       thetadiff = (1-2*m)/(thetascale*m);
[…]

### 2017-05-25 Manuscript/msv2.tex
first draft of discussion

@@
 \title{Lost in space: the eco-evolutionary impacts of straying on metapopulation robustness}
-\author{Justin D. Yeakel${}^{1,2,*}$, Jean Phillipe Gibert${}^{1}$, Sean Anderson${}^{3}$, Peter Westley${}^{4}$, \& Jon W. Moore${}^{5}$ \\
+\author{Justin D. Yeakel${}^{1,2,*}$, Jean Philippe Gibert${}^{1}$, Peter Westley${}^{3}$, \& Jonathan W. Moore${}^{4}$ \\
 ${}^1$School of Natural Sciences, University of California, Merced, Merced CA, USA \\
 ${}^2$The Santa Fe Institute, Santa Fe NM, USA \\
-${}^3$Department of Biology, University of Washington, Seattle WA, USA \\
-${}^4$Institute for Arctic Biology, University of Alaska, Fairbanks, Fairbanks AK, USA \\
-${}^5$Earth${}_2$Oceans Research Group, Simon Fraser University, Vancouver BC, Canada \\
+${}^3$College of Fisheries and Ocean Sciences, University of Alaska, Fairbanks, Fairbanks AK, USA \\
+${}^4$Earth${}_2$Oceans Research Group, Simon Fraser University, Vancouver BC, Canada \\
 ${}^*$To whom correspondence should be addressed: jdyeakel@gmail.com
@@
 A more direct way to measure system robustness is to measure the time that it takes the system (measured as the aggregate biomass $N_T$) to recover its steady state abundance following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
-Measuring the time that it takes for a perturbed system to relax also permits a more detailed perspective of the potential fragility of the metapopulation.
+Although there is a direct eignvalue relationship between the rate of return following a small pulse perturbation (REF), because we aimed to 1) assess the effects of a large perturbation, and 2) estimate the time required for all transient effects to decay (including dampened oscillations), we used a simulation-based numerical procedure.
+
+Numerically estimating the time that it takes for a perturbed system to relax also permits a more detailed perspective of the potential fragility of the metapopulation.
 For example, if populations settle to alternative steady states (alternative steady states in our model requiring one population to be high-density and one low-density), comparing recovery times after a disturbance applied to the high, low, and/or both populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery.
@@
 A low stray rate corresponds to very different (or distant) habitats (high $\Delta\theta$), whereas a higher rate of straying corresponds to very similar (or nearby) habitats (low $\Delta\theta$).
-Light colors = high values.
+  (a) Portfolio effect over constant straying rate $m$ and trait heritability $h^2$.
+  (b-c) Recovery times for three disturbance types for (b) constant $m$ and (c) density dependent straying rate at the steady state $m^*$.
   } \label{fig:mtheta}
@@
 
-Here we find that alternative stable states now appear for low straying rates ($m<0.15$).
-As the straying rate increases, a single stable state emerges with a correspondingly high $N_T$.
-As before, there is a spike in the PE at the fold bifurcation separating the alternative stable state regime from the single stable state regime, and a sharp decline in the PE as the straying rate becomes very low.
-This is in accordance with intuition as increasing straying rates mean that two very similar populations are mixing, such that the rate of recruitment is less influenced by gene flow.
+We find that alternative stable states now appear at low straying rates, which corresponds to greater $\Delta\theta$.
+As the straying rate increases, a single stable state emerges.
+As before, the portfolio effect is higher in the alternative stable state regime, with a sharp decline in the PE as the straying rate becomes larger (figure \ref{fig:mtheta}a).
+This is in accordance with intuition as increasing straying rates mean that two very similar populations are mixing, such that the strength of selection on recruitment is lessened.
 
-That alternative stable states appear and that PE becomes severely depressed for very low rates of straying is surprising: this means that even a small amount of mixing of populations from distant or dissimilar populations can qualitatively alter the dynamics of the metapopulation, regardless of trait heritability.
+This dynamic is more complex when we observe the recovery times for systems with either a constant (figure \ref{fig:mtheta}b) or density dependent (figure \ref{fig:mtheta}c) straying rate.
+Although recovery times are lowest in the alternatice stable state regime ($m<0.1$), this is only true if the low-density population suffers extinction.
+Alternatively, at low rates of straying between dissimilar (distant) populations, recovery times become maximized if the high-density population suffers extinction.
+As rates of straying -- and the similarity of trait optima -- increase, extinction of either population requires less time for recovery than near-collapse of both.
+
+% That alternative stable states appear and that PE becomes severely depressed for very low rates of straying is surprising: this means that even a small amount of mixing of populations from distant or dissimilar populations can qualitatively alter the dynamics of the metapopulation, regardless of trait heritability.
 %(for the discussion: an example of this situation may be the salmon populations during the last glacial maximum, where any mixing would be from geographically distant populations)
@@
 
-We have shown that the natural selection of mixed populations adapted to different environments can result in large effects on population dynamics and the degree to which the metapopulation is buffered against extinction, quantified here as the portfolio effect.
+We have shown that dispersal between populations, coupled with localized selection against donor phenotypes, has a large and nonlinear impact on dynamic properties that directly impacts metapopulation robustness, and by extension persistence.
+Although there are various ways to measure `robustness' (REF:TREE), we employ two measures: 
+1) the average-CV portfolio effect, a statistical metric commonly used to assess the fragility of salmon populations; 
+2) the recovery time, defined here as the time required for the aggregate metapopulation biomass $N_T$ to return to its steady state after different types of induced disturbances, which is mechanistically linked to the likelihood of persistence (REF).
+In our minimal model of dispersal and natural selection between two populations, we show that these phenomenological and mechanistic descriptors of metapopulation robustness are tightly coupled (figure \ref{fig:PE}d).
+
+%Alternative steady states & critical slowing
 The immigration of dispersing strays with trait values far from the local optimum generally lowers the combined steady state biomass $N_T$ and increases the likelihood that the system gives way to alternative steady states, pushing one of the populations close to extinction.
-Although the emergence of alternative steady states can result in very large portfolio effects near the bifurcation, alternative steady state regimes tend to result in very low portfolio effects, indicating that the metapopulation is less likely to persist in the long run.
+The formation of alternative stable states in our system is a relatively common example of spatial pattern formation, and occurs above a threshold straying rate.
+Pattern formation can occur as a consequence of myriad ecological processes, including habitat selection (REFS), aggregation of individuals (REFS), local environmental conditions (REFS), or interspecific interactions like competition (REFS).
+It is also well known to occur with the onset of spatially or diffusion-induced instabilities (Turing instabilities, REFS). 
+Here, we describe an interesting alternative mechanism through which pattern formation in local densities can occur, when the onset of movement mediated maladaptation leads to local differences in reproductive rates that yield differences in densities that would, in the absence of straying, be the same.
+This mechanism of pattern formation through the simultaneous interplay between ecological and evolutionary dynamics may open up new areas of research.
 
-The detectability of such changes in dynamical behavior among salmonids appears to be idiosyncratic across species, and the difficulty in measuring critical slowing down may - ironically - be masked by large portfolio effects \cite{Krkosek:2014ch} [this might be more appropriate for the discussion].
+To what extent the existence of alternative stable states promotes or diminishes the system as a whole is a matter of perspective.
+On the one hand, one population is pushed to a low-density state (figure \ref{fig:traj}) with a lower total biomass of the aggregate (figure ]ref{fig:PE}a), suggesting that the system is more fragile.
+On the other hand, the presence of \emph{just enough} straying to cause formation of alternative stable states both increases the portfolio effect (figure \ref{fig:PE}c) and reduces the rate of return -- particularly if straying is density dependent (figure \ref{fig:relax}c,d) as might be common in systems with collective navigation (REF).
 
-The degree to which the PE is lowered depends to a large extent on \emph{i}) trait heritability $h^2$ and \emph{ii}) habitat heterogeneity $\Delta \theta$.
-Increased heritability results in greater response to natural selection on the metapopulation, increasing the coupling between evolutionary and ecological dynamics. 
-Our minimal model of straying between two populations shows that greater $h^2$ buffers the metapopulation against the emergence of alternative stable states, but magnifies the negative effects of alternative stable states (captured by increased $\Delta N$ and decreased PE; figure \ref{fig:PE}B,C) once this regime is encountered at higher straying rates.
-We note that although the rate of straying has been shown to be density dependent (REFS), the qualitative results of our model are relatively insensitive to this dynamic (figure \ref{fig:diffddm}) and we limit discussion to the simplified case of a constant stray rate $m$ with the understanding that our findings also apply to the case density dependent stray rates $m(t)$.
+%The bifurcation and critical slowing down
+Although the portfolio effect is negatively correlated with recovery time, there is one exception: the point of transition from a single steady state to alternative steady states, which occurs at the fold bifurcation (REF).
+At this point, the portfolio effect is maximized, but so is the recovery time (figures \ref{fig:PE}, \ref{fig:relax}).
+This suggests that while the sharply increasing variance of the individual populations is dampened by aggregation, a large disturbance will have a much greater adverse effect due to exponentially longer recovery times.
+Whether the change in variance, also known as critical slowing down, as the alternative steady state regime is approached could be used as an early warning signal of an oncoming phase transition is a hotly debated topic (REFS).
+The detectability of such changes in dynamical behavior among salmonids appears to be idiosyncratic across species, though the difficulty in measuring critical slowing down may - ironically - be masked by large portfolio effects \cite{Krkosek:2014ch}.
+
+%Comparing recovery times
+Previous theoretical work has shown that the largest perturbations lead to the longest recovery times \cite{Ovaskainen:2002il}.
+In terms of relative biomass lost, the largest perturbation that we investigate is the near-collapse of both fisheries, where only 1\% of the preperturbation population densities survive, however whether this disturbance results in the longest recovery time depends largely on the rate of straying and trait heritability.
+For example, when straying and trait heritability are low (figure \ref{fig:relax}a,c) the extinction of the larger population maximizes return time, whereas if straying is high, near-collapse maximizes return time.
+The latter is always true in the alternative stable state regime is trait heritability is high (figure \ref{fig:relax}b,d)
+
+%Density dependent straying
+This issue is largely focused on the notion that straying in biological systems where movement is a function of collective navigation may be density dependent.
+Berdahl et al.
+(REF) provided a mechanistic theory for density dependent straying where the proportion of the population that strays is less than the probability that a single individual (as illustrated by \emph{Kevin}; figure \ref{fig:xkcd}) takes the wrong turn.
+Neither the transient nor asymptotic dynamics that we describe here qualitatively differ as a function of whether the rate of straying is constant or density dependent, however we observe important quantitative differences that suggest density dependent straying may play an important role in the persistence of metapopulations over evolutionary time.
+
+First, density dependent straying reduces the time to recovery following an induced disturbance, and this is particularly true if (\emph{i}) trait heritability is low, and (\emph{ii}) in the case of near-collapse of both populations (figure \ref{fig:relax}c).
+This dynamic emerges because although both populations are near-collapse, the mean trait values of both are scewed towards those of the high-density population.
+As the low-density population has also collapsed, and the mean trait values of the system favor the high-density population, it grows more quickly, recovering faster (figure \ref{fig:relaxtraj_bothlh}).
+Second, density dependent straying increases the effective portfolio effects when the steady state straying rate is higher (for both low- and high-density populations).
+Third, while habitat heterogeneity is interpreted here in elevating the difference in trait optima between sites, it both increases the portfolio effect and lowers recovery time at low rates of straying, but has the opposite effect at higher rates of straying.
+Surprisingly, we find that density dependent straying both exaggerates the negative effects when $m^*$ is low (marginally lower portfolio effects, but subsatially longer return times; figures \ref{fig:thetaPE}b, \ref{fig:relaxtheta}a,b) and exaggerates the positive effects when $m^*$ is high (higher portfolio effects, abd shorter return times).
+
+HAVENT EDITED FURTHER
 
@@
 
-Recovery times: largest perturbations lead to the longest recovery times \cite{Ovaskainen:2002il}.
-This is not true in our systems when (\emph{i}) heritability and (\emph{ii}) straying are low. Why?
-
-That recovery times and PE are similar for constant $m$ vs. density dependent $m$ suggests that both transient dynamics as well as asymptotic dynamics are not largely impacted.
-
-%Our results in the context of Turing bifurcation
-[JP - could use some help here]
-The onset of alternative stable states in a spatial context describes the emergence of spatial pattern formation (REFS), which is more generally defined as a Turing instability (REFS).
-The mathematical conditions that lead to Turing instabilities are well-known in both continuous (REFS) and discrete spatial contexts (REFS)
-
-%North-South vs East-West
+%Coupling distance: North-South vs East-West
 
@@
 
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.35\textwidth]{figs2/fig_relax_both_lowh.pdf}
+\caption{
+Near collapse of both populations with a low straying rate $m=0.1$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
+Black line marks the calculated point of recovery post-perturbation.
+Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
+} \label{fig:relaxtraj_bothlh}
+\end{figure}
+
 % \begin{figure*}[h!]

### 2017-05-25 model/KevinSimulations_mtheta2.jl

@@ -216,13 +216,14 @@ tmax=10000;
 
 mvec2 = collect(0.0001:0.001:0.5);
 pvec = ["small","large","both"];
+reps = 100;
 
 rt = SharedArray{Float64}(reps,length(mvec2),length(pvec));
 rt_ddm = SharedArray{Float64}(reps,length(mvec2),length(pvec));
 m1mean = SharedArray{Float64}(reps,length(mvec2),length(pvec));
 m2mean = SharedArray{Float64}(reps,length(mvec2),length(pvec));
 
-reps = 100;
+
 z=0.5;
 rmax=2.0;
 beta=0.001;
@@ -323,7 +324,7 @@ ma_rth_ddm = mean([rt_ddm[i,:,:] for i=1:reps]);
 ma1_m_ddm = mean([m1mean[i,:,:] for i=1:reps]);
 ma2_m_ddm = mean([m2mean[i,:,:] for i=1:reps]);
 
-namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/figs2/fig_relax_mtheta.pdf");
+namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/Manuscript/figs2/fig_relax_mtheta.pdf");
 R"""
 library(RColorBrewer)
 pal = brewer.pal(9,'Greys')
@@ -348,7 +349,7 @@ for (i in 1:length($mvec)) {
 dev.off()
 """
 
-
+ 
 
 namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/Manuscript/figs2/fig_MDPE_hm_mtheta_rt.pdf");
 R"""

### 2017-05-27 manuscript/msv3.tex
msv3 and fig:relax changes

--- 2017-05-25 Manuscript/msv2.tex
@@
 %\title{Community assembly and dynamics by the deconstruction of species interactions}
-\title{Lost in space: the eco-evolutionary impacts of straying on metapopulation robustness}
-\author{Justin D. Yeakel${}^{1,2,*}$, Jean Philippe Gibert${}^{1}$, Peter Westley${}^{3}$, \& Jonathan W. Moore${}^{4}$ \\
+\title{The eco-evolutionary impacts of collective straying on metapopulation dynamics}
+\author{Justin D. Yeakel${}^{1,2,*}$, Jean Philippe Gibert${}^{1}$, Peter A. H. Westley${}^{3}$, \& Jonathan W. Moore${}^{4}$ \\
 ${}^1$School of Natural Sciences, University of California, Merced, Merced CA, USA \\
@@
 
-%Motivation: straying
-Coordinated mass migrations are one of the great wonders of the natural world, and the ability of individuals and groups to navigate across great distances have long fascinated naturalists \cite{MilnerGulland:2011vm}.
-Anadromous salmonid fishes (genera \emph{Oncorhynchus} and \emph{Salmo}) embody these astonishing migrations by homing with high accuracy and precision to their natal streams for reproduction after years at sea \cite{Quinn:2011tf,Jonsson:2011kg,Keefer:2014gg}.
-Nothwithstanding their philopatric tendancy, not all individuals home, but rather `stray' to non-natal sites to spawn \cite{Quinn:1993ge,Hendry:2004wf,H:2013fs}.
-Although extensive work has been done to document the extent of straying from donor populations and into recipient populations \cite{Keefer:2014gg,Bett:2017ha}, only recently have the abiotic, biotic, and anthropogenic influences of `straying' behaviors been investigated systemically \cite{Keefer:2008bs,Westley:2015to,Bond:2016dz}.
-Most recently the role of social interactions and collective navigation has been hypothesized \cite[][; this volume]{Berdahl:2015kv,Berdahl:2016dx}.
-Although the straying of individuals into sites hosting other populations provides connections within the larger metapopulation, potentially promoting ecological and evolutionary rescue, it also serves to introduce maladapted individuals into habitats that are host to different environmental conditions, and this may lower the mean fitness of the local (mixed) population \cite{Muhlfeld:2014hs}.
+Intraspecific diversity can increase the resilience and stability of species or metapopulations. 
+This diversity-stability linkage can occur when there are asynchronous population dynamics, where the changes in population size vary across a metapopulation. 
+This asynchrony will increase the potential for demographic rescue \cite{Earn:2000fm} and also decrease the variability of processes that integrate across the metapopulation \cite{Anonymous:2015gf}. 
+For example, different responses to climate variability within populations of a rare plant reduced abundance fluctuations \cite{Abbott:2017hl}. 
+This statistical buffer has traditionally been quantified as the Portfolio Effect (PE), which is the ratio of the population CV to the CV of the aggregated metapopulation \cite{Thibaut:2012km}.
 
-%The tradeoff
-The dual nature of straying as both promoter of connections among metapopulation demes and potential eroder of locally adapted gene complexes highlights the interplay between ecological dynamics of connected populations as well as the evolutionary dynamics of mixed trait distributions that respond to alternative local conditions.
-Population-level biodiversity is recognized to increase species persistence (REFS), and the resilience and sustainability of metapopulations (general REF).
-For example, the long-term sustainability of the Bristol Bay sockeye salmon fishery is due in large part to intact habitat that in conjunction with fine-scale homing and reduced gene flow provides a template for the evolution of countless locally adapted populations \cite{Hilborn:2003gf,Schindler:2010he,Anonymous:2014ku,Satterthwaite:2015ge}.
-In contrast, areas such as California's Central Valley where habitats have been lost, degraded, and homogenized stand in marked contrast.
-Within species diversity (or lack thereof) can manifest itself by way of asynchronous population dynamics, where the drivers that give rise to changes in population size vary across a metapopulation, decreasing the potential for synchronization, and correspondingly increasing the potential for ecological rescue (REFS).
-This statistical buffer against extinction has traditionally been quantified as the Portfolio Effect (PE), which is the ratio of the population CV to the CV of the aggregated metapopulation.
-Weakened portfolio dynamics can also emerge from homogenization resulting from  reduced genetic variability among populations, where homogenous stocks are more likely to have similar life-history structures (REFS), be at greater risk disease-induced epidemics (REFS), and recruit sub-optimally in heterogeneous environments (REFS).
-Of course, these two measures of diversity often occur concurrently: for instance, lower genetic diversity increases the likelihood that two populations respond similarly to the same stressors, and this promotes synchronized dynamics, thus lowering portfolio effects.
+Movement of individuals across populations (i.e., straying or dispersal) can have a large influence on metapopulation dynamics  \cite{MilnerGulland:2011vm}. 
+First, some level of straying is necessary to enable the rescue effect, whereby dispersal among populations can rescue small populations from local extinction (REFS). 
+Second, high levels of straying may synchronize populations and actually increase the risk of extinction of the entire metapopulation \cite{Earn:2000fm}. 
+Third, straying will influence the evolutionary dynamics of the metapopulation, however evolutionary impacts are less well understood. 
+Although the straying of individuals into sites hosting other populations provides connections within the larger metapopulation, potentially promoting ecological and evolutionary rescue, it may also introduce maladapted individuals into habitats that are host to different environmental conditions, possibly lowering the mean fitness of the local (mixed) population \cite{Muhlfeld:2014hs}.
+Further, this straying may lead to genetic homogenization that erodes the asynchrony that underpins portfolio effects and metapopulation persistence. 
+Thus, the dual nature of straying as both promoter of connections among metapopulation demes and potential eroder of locally adapted gene complexes highlights the interplay between ecological dynamics of connected populations and the evolutionary dynamics of mixed trait distributions that respond to heterogeneous local conditions.
+
+% %The tradeoff
+% Population-level biodiversity is recognized to increase species persistence (REFS), and the resilience and sustainability of metapopulations (general REF).
+% For example, the long-term sustainability of the Bristol Bay sockeye salmon fishery is due in large part to intact habitat that in conjunction with fine-scale homing and reduced gene flow provides a template for the evolution of countless locally adapted populations \cite{Hilborn:2003gf,Schindler:2010he,Anonymous:2014ku,Satterthwaite:2015ge}.
+% In contrast, areas such as California's Central Valley where habitats have been lost, degraded, and homogenized stand in marked contrast.
+% Within species diversity (or lack thereof) can manifest itself by way of asynchronous population dynamics, where the drivers that give rise to changes in population size vary across a metapopulation, decreasing the potential for synchronization, and correspondingly increasing the potential for ecological rescue (REFS).
+% This statistical buffer against extinction has traditionally been quantified as the Portfolio Effect (PE), which is the ratio of the population CV to the CV of the aggregated metapopulation.
+% Weakened portfolio dynamics can also emerge from homogenization resulting from  reduced genetic variability among populations, where homogenous stocks are more likely to have similar life-history structures (REFS), be at greater risk disease-induced epidemics (REFS), and recruit sub-optimally in heterogeneous environments (REFS).
+% Of course, these two measures of diversity often occur concurrently: for instance, lower genetic diversity increases the likelihood that two populations respond similarly to the same stressors, and this promotes synchronized dynamics, thus lowering portfolio effects.
+
+%Movement through space (link back to straying)
+Migratory populations that return to a breeding ground or natal stream to reproduce are linked to each other by some proportion of the population that permanently disperse, or stray into the `wrong' site; we might say that there is at least one \emph{Kevin} in every school or flock (figure \ref{fig:xkcd}).
+There is growing appreciation that the rate of straying may be influenced by abiotic, biotic, and anthropogenic factors \cite{H:2013fs,Keefer:2014gg,Bett:2017ha}. 
+Recently the role of social interactions and collective navigation has been hypothesized [This issue].
+The rate at which individuals stray, $m$, may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \cite{Berdahl:2015kv,Berdahl:2016dx}.
+Regardless of the mechanism's governing straying, the effect that it has on the dynamics of individual populations and the metapopulation as a whole is a topic of considerable interest that has tangible conservation implications \cite{Brenner:2012gl,Johnson:2012fe,Fullerton:2011ii}.
+Whether, and to what extent, the ecological consequences of straying depend on the evolutionary dynamics that emerge from populations distributed across a selective gradient is unclear.
+How the assumed negative evolutionary effects of straying and subsequent gene flow is balanced by the positive effects of demographic rescue is the subject of this contribution.
 
@@
 %% From geographic mosaic of coevolution to...
-That evolutionary forces play out heterogeneously across geographic mosaics is now a foundational concept in ecology and evolutionary biology (REFS).
-These mosaics are in part driven by environmental differences between habitats that alter the selective forces acting on different phenotypes \cite{Endler:1986tz}, and a principle underlying assumption is that there is gene flow such that individuals from different habitats mix over space (REFS).
-Although the evolutionary outcomes of these spatial processes have been explored in depth (REFS), it is less well understood how selective mosaics and their consequent evolutionary forces impact population dynamics over contemporary timescales \cite{Hendry:2016un}.
+That evolutionary forces of selection and gene flow play out heterogeneously across geographic mosaics is now a foundational concept in ecology and evolutionary biology \cite{Nuismer:1999ko,Thompson:2015hq,Nuismer:2000bb,Thompson:2005wf}.
+These mosaics are in part driven by environmental differences between habitats that alter the selective forces acting on different phenotypes \cite{Endler:1986tz}, and a principle underlying assumption is that there is gene flow such that individuals from different habitats mix over space \cite{Gomulkiewicz:2000bz,Thompson:2002hr,Nuismer:2003eh,Nuismer:2006be,Forde:2008kc,Guimaraes:2011hu,Gibert:2013hp}.
+Although the evolutionary outcomes of these spatial processes have been explored in depth (REFS), it is less well understood how selective mosaics and their consequent evolutionary forces impact population dynamics as they unfold \cite{Hendry:2016un}.
 How the ecological dynamics of phenotypically diverse metapopulations, wherein constituent populations are subject to different selective regimes, are altered affected by evolutionary forces is less well known.
@@
 
-%Movement through space (link back to straying)
-Migratory populations that return to a breeding ground or natal stream to reproduce are linked to each other by some proportion of the population that permanently disperse, or stray into the `wrong' site; we might say that there is at least one \emph{Kevin} in every school or flock in every school or flock (figure \ref{fig:xkcd}).
-The rate at which individuals stray, $m$, has been subject to a growing amount of both theoretical and empirical research \cite{H:2013fs,Keefer:2014gg,Bett:2017ha} and may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \cite{Berdahl:2015kv,Berdahl:2016dx}.
-Regardless of the mechanism's governing straying, the effect that it has on the dynamics of individual populations and the metapopulation as a whole is a topic of considerable interest that has tangible conservation implications \cite{Brenner:2012gl,Johnson:2012fe,Fullerton:2011ii}.
-Whether, and to what extent, the ecological consequences of straying depend on the evolutionary dynamics that emerge from populations distributed across a selection mosaic is unclear.
-How the assumed negative evolutionary effects of straying and subsequent gene flow is balanced by the positive effects of demographic rescue is the subject of this contribution.
+%Motivation: straying
+The eco-evolutionary impacts of straying likely has important implications for conservation and management in key taxa such as in salmon.
+Salmon may operate as metapopulations, where populations are genetically distinct but linked by some level of straying \cite{Schtickzelle:2007wb,Anderson:2014cx}.
+Asynchrony among salmon populations also increases the stability of fisheries that integrate across this diversity \cite{Schindler:2010he,Anonymous:2016jc}.
+However, these portfolio effects may be eroded if asynchrony is influenced by genetic diversity and this genetic diversity is lost. 
+Braun et al. \cite{Braun:2016ib} found that genetic differences among populations of Chinook salmon was correlated with the degree of asynchrony in population dynamics. 
+Artificial propagation of salmon by hatcheries and their strays may erode local genetic diversity, asynchrony among populations, and portfolio effects \cite{Carlson:2008hl,Moore:2010gs,Anonymous:2014ku}.
+%Coordinated mass migrations are one of the great wonders of the natural world, and the ability of individuals and groups to navigate across great distances have long fascinated naturalists \cite{MilnerGulland:2011vm}.
+While anadromous salmonid fishes (genera \emph{Oncorhynchus} and \emph{Salmo}) are reknown for returning to their natal spawning habitats with high accuracy and precision after years at sea \cite{Quinn:2011tf,Jonsson:2011kg,Keefer:2014gg}, there are generally some individuals that `stray' to non-natal sites to spawn \cite{Quinn:1993ge,Hendry:2004wf}.
+Although extensive work has been done to document the extent of straying from donor populations and into recipient populations \cite{Keefer:2014gg,Bett:2017ha}, only recently have the abiotic, biotic, and anthropogenic influences of `straying' behaviors been investigated systemically \cite{Keefer:2008bs,Westley:2015to,Bond:2016dz}.
+Most recently the role of social interactions and collective navigation has been hypothesized \cite{Berdahl:2015kv,Berdahl:2016dx}.
+These strays can introduce new maladaptive genotypes into the recipient population.
+Straying amon salmon may be influenced by environmental factors such as water temperature, human activities such as hatchery practices, and population density as predicted by the collective migration hypothesis \cite{Peterson:2014gy}.
+Thus, there is an opportunity to consider the eco-evolutionary consequences of straying for metapopulations in species of conservation and management concern such as salmon \cite{Carlson:2011ce}. 
+%Although the straying of individuals into sites hosting other populations provides connections within the larger metapopulation, potentially promoting ecological and evolutionary rescue, it also serves to introduce maladapted individuals into habitats that are host to different environmental conditions, and this may lower the mean fitness of the local (mixed) population \cite{Muhlfeld:2014hs}.
 
@@
 %Alternative stable state
-Here we ask the overarching question: how does collective behavior mediated dispersal and gene flow interact to influence the robustness of locally adapted populations?
+Here we seek to explore how the potentially detrimental evolutionary effects of straying (erosion of local adaptation) and subsequent gene flow is balanced by the positive effects of demographic rescue is the subject of this contribution.
+% Here we ask the overarching question: how does collective behavior mediated dispersal and gene flow interact to influence the robustness of locally adapted populations?
 To address this question we construct a minimal eco-evolutionary model of two populations occupying different sites that are linked by straying individuals, each of which with an associated trait distribution subject to natural selection determined by local conditions.
-An important and relatively novel component of this model is the inclusion of a parameter that defines the difference in local conditions that favor different trait optima, increasing values of which correlates to populations that mix across increasingly heterogeneous environments \cite[though see ]{Berdahl:2015kv}.
-Although our proposed model is was constructed with the dynamics of salmon populations in mind, the framework is general and the conclusions are likely relevant to a diverse range of migratory organisms where locally adapted populations are linked by dispersal.
-We first show that specific rates of straying and trait heritability can have large effects on the qualitative dynamics of populations over time, in many cases giving rise to alternative steady states where one site is pushed towards very low biomass.
-The emergence of alternative steady states results in a nonlinear response of the portfolio effect as well as the time required for the metapopulation to recover after an induced disturbance, suggesting that metapopulation robustness can be quite sensitive to the combined influences of dispersal across a selective landscape.
+[We show that taking straying into account leads to alternative stable states in population densities and trait values, which has consequences for maladaptation, intraspecific trait variability, and long-term sustainability of salmon metapopulations.]
 
-%Extinction and return times
-
-%Habitat heterogeneity
-A second important finding of our minimal model reveals that systems with greater habitat heterogeneity (measured by an increased difference in the trait values that are optimal between sites) host an intermediate range of straying rates where the portfolio effect increases to a local maximum, signaling that under certain conditions moderate amounts of straying between populations can increase the likelihood of persistence despite the deleterious evolutionary effects overall.
-However, if we suppose that the rate of straying between two sites is correlated with distance (as indicated by empirical evidence), and that the difference in trait optima increases with distance as would be the case if the optima were associated with alternative temperature regimes (especially if sites are distributed latitudinally rather than longitudinally), even a very small amount of straying can drastically reduce the portfolio effect.
-Importantly, the qualitative nature of our results do not depend on whether the stray rate is density dependent or constant, suggesting a limited role of collective dispersal on the dynamics considered herein.
-
-%Concluding paragraph of the intro
+% An important and relatively novel component of this model is the inclusion of a parameter that defines the difference in local conditions that favor different trait optima, increasing values of which correlates to populations that mix across increasingly heterogeneous environments \cite[though see ]{Berdahl:2015kv}.
+% Although our proposed model is was constructed with the dynamics of salmon populations in mind, the framework is general and the conclusions are likely relevant to a diverse range of migratory organisms where locally adapted populations are linked by dispersal.
+% We first show that specific rates of straying and trait heritability can have large effects on the qualitative dynamics of populations over time, in many cases giving rise to alternative steady states where one site is pushed towards very low biomass.
+% The emergence of alternative steady states results in a nonlinear response of the portfolio effect as well as the time required for the metapopulation to recover after an induced disturbance, suggesting that metapopulation robustness can be quite sensitive to the combined influences of dispersal across a selective landscape.
+% 
+% %Extinction and return times
+% 
+% %Habitat heterogeneity
+% A second important finding of our minimal model reveals that systems with greater habitat heterogeneity (measured by an increased difference in the trait values that are optimal between sites) host an intermediate range of straying rates where the portfolio effect increases to a local maximum, signaling that under certain conditions moderate amounts of straying between populations can increase the likelihood of persistence despite the deleterious evolutionary effects overall.
+% However, if we suppose that the rate of straying between two sites is correlated with distance (as indicated by empirical evidence), and that the difference in trait optima increases with distance as would be the case if the optima were associated with alternative temperature regimes (especially if sites are distributed latitudinally rather than longitudinally), even a very small amount of straying can drastically reduce the portfolio effect.
+% Importantly, the qualitative nature of our results do not depend on whether the stray rate is density dependent or constant, suggesting a limited role of collective dispersal on the dynamics considered herein.
+% 
+% %Concluding paragraph of the intro
+% 
+%
 
@@
 \noindent{\bf (a) Metapopulation framework}\\
-\noindent We consider two populations $N_1$ and $N_2$ that are separated in space in distinct habitats, each with trait values $x_1$ and $x_2$ determining recruitment rates.
-We assume that there is an optimum trait value $\theta_1$ and $\theta_2$ associated with each habitat, where recruitment is maximized if the trait value of the local population $x = \theta$.
+\noindent We consider two populations $N_1$ and $N_2$ that inhabit two distinct habitats, each with trait values $x_1$ and $x_2$ determing recruitment rates.
+We assume that there is an optimum trait value $\theta_1$ and $\theta_2$ associated with each habitat, where recruitment is maximized if the trait value of the local population equals the optimum, such that $x = \theta$.
 Moreover, we assume that $x_{1,2}$ are normally distributed with means $\mu_1$ and $\mu_2$ and have the same standard deviation $\sigma$.
 As such, the recruitment rate for both populations is determined by the mean trait value of the local population, such that $r_1 = R_1[\mu_1(t),\theta_1]$.
-Trait means for each population are subject to selection, the strength of which depends on the difference between the population mean and the local trait optimum at a given point in time \cite{simpson1953major,Lande:1976ga}.
+Trait means for each population are subject to selection, the strength of which depends on the difference between the trait mean and the local trait optimum at a given point in time \cite{simpson1953major,Lande:1976ga}.
 
+%PW: Can we say anything about the effect of the recipient population size and that effect on evolutionary dynamics?
+Again, straying can be thought of as both from the donor or recipient population perspective.
 The two populations are assumed to reproduce in spatially separate sites that are close enough such that a proportion of the population $m$ can stray into the other site, and where mortality occurs before individuals return to reproduce.
@@
 
-The combined recruitment of local individuals $(1-m)N_i(t)$ and incoming strays $mN_j(t)$, as a function of their mean trait value at time $t$.
-Given the local trait optimum is $\theta_i$, recruitment is
+The recruitment of local individuals $(1-m)N_i(t)$ as a function of their mean trait value at time $t$ and the local trait optimum, is
 
@@
 
-\noindent where the mismatch between the local trait mean $\mu_i(t)$ and the local optimum $\theta_i$ determines scales the recruitment rate for the population, and $\tilde{P}\sim {\rm Normal}(0,0.01)$ introduces a small amount of demographic error.
-The parameter $\tau$ is the strength of selection, and controls the sensitivity of recruitment to changes in the mean trait value away from the optimum, which we set as $\tau=1$ here and throughout.
+\noindent where the mismatch between the local trait mean $\mu_i(t)$ and the local optimum $\theta_i$ scales the recruitment rate for the population, and $\tilde{P}\sim {\rm Normal}(0,0.01)$ introduces a small amount of demographic error.
+The parameter $\tau$ is the strength of selection, and controls the sensitivity of recruitment to changes in the mean trait value away from the optimum (the strength of selection increases with smaller values of $\tau$), which we set as $\tau=1$ here and throughout.
+Because straying individuals are emigrating from a population with a mean trait value farther from the local optimum, their rate of recruitment is diminished, and this matches what is observed in wild populations \cite{Peterson:2014gy}.
 % The compensatory effects are then determined by the exponential, following the Ricker stock-recruitment relationship.
@@
 \noindent{\bf (b) Recruitment over a selective landscape}
-\noindent Because individuals from the local population are mixed with individuals from the remote population via staying, the resulting trait distribution is a mixed normal with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and for the straying individuals, $1-w_i$, where
+\noindent Because individuals from the local population are mixed with individuals from the remote population via straying and subsequent reproduction, the resulting trait distribution is a mixed normal with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and for the straying individuals, $1-w_i$, where
 \begin{equation}
@@
 We make two simplifying assumptions.
-First, we assume that the distribution resulting from the mix of remote and local individuals, following reproduction, is also normal with a mean value being that of the mixed-normal.
-Second, we assume that changes in trait variance through time are minimal, such that $\sigma^2$ is assumed to be constant.
+First, we assume that the distribution resulting from the mix of remote and local individuals, following reproduction, is also normal with a mean value equal to that of the mean for the mixed-normal distribution.
+Second, we assume that changes in trait variance through time are minimal, such that $\sigma^2$ is assumed to be constant, and this is a common simplification in eco-evolutionary models of population dynamics \cite{Lande:1976ga,Schreiber:2011wx,Gibert:tz}.
 
 An increasing flow of incoming strays is generally expected to pull the mean trait value of the local population away from its optimum over time, which will decrease its rate of recruitment.
-The mean trait value thus changes through time according to the difference equation
+Following Lande \cite{Lande:1976ga}, the mean trait value thus changes through time according to the difference equation
 
@@
 A more direct way to measure system robustness is to measure the time that it takes the system (measured as the aggregate biomass $N_T$) to recover its steady state abundance following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
-Although there is a direct eignvalue relationship between the rate of return following a small pulse perturbation (REF), because we aimed to 1) assess the effects of a large perturbation, and 2) estimate the time required for all transient effects to decay (including dampened oscillations), we used a simulation-based numerical procedure.
+Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation \cite{GuckHolmes}, because we aimed to 1) assess the effects of a large perturbation, and 2) estimate the time required for all transient effects to decay (including dampened oscillations), we used a simulation-based numerical procedure.
 
-Numerically estimating the time that it takes for a perturbed system to relax also permits a more detailed perspective of the potential fragility of the metapopulation.
+Numerically estimating the time that it takes for a perturbed system to relax also permits a more detailed perspective of metapopulation fragility.
 For example, if populations settle to alternative steady states (alternative steady states in our model requiring one population to be high-density and one low-density), comparing recovery times after a disturbance applied to the high, low, and/or both populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery.
+We measured the time required for the system $N_T$ to recover to its steady state following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} are equivalent if the system is in the single steady state regime); (\emph{iii}) near-collapse of both populations where just 1.0\% of each survives.
 Throughout, we will refer to an increase in the portfolio effects and/or reduction in recovery times as promoting metapopulation robustness, which is expected to have a positive effect on persistence.
@@
 %Density dependent m
-\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is good evidence that at least in some species the straying rate is density dependent \cite{Berdahl:2016dx,Bett:2017ha}.
+\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is mounting evidence that at least in some species (including straying) the straying rate is density dependent with a signature of collective navigation \cite{Berdahl:2016dx,Bett:2017ha}.
 Specifically, the rate at which individuals stray has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tend to decrease the rate at which individuals err, reducing the overall proportion of a population that strays.
-According to Berdahl et al. \cite{Berdahl:2016dx}, given the probability that an individual strays is $m_0$, the proportion of the local population $N_i(t)$ that strays is
+According to Berdahl et al. \cite{Berdahl:2014bl,Berdahl:2016dx}, given the probability that an individual strays is $m_0$, the proportion of the local population $N_i(t)$ that strays is
 
@@
 %Linking m/m0 and thetadiff
-The straying rate is intrinsically linked to the distance between the donor and recipient population.
-The greater the distance between two populations, the lower the expected rate of straying (REF).
-We can account for this interdependence in our model by assuming that $m$ (if the stray rate is constant) or $m_0$ (if the stray rate is density dependent) is a function of the difference between optimal trait values between sites $\theta_i-\theta_j$, which can be assumed to be large if the remote site $j$ is a great distance away from the local site $i$.
+The straying rate is largely influenced by the distance between the donor and recipient population.
+The greater the distance between two populations, the lower the expected rate of straying \cite{H:2013fs,Westley:2015to}.
+We account for this interdependence in our model by assuming that $m$ (if the stray rate is constant) or $m_0$ (if the stray rate is density dependent) is a function of the difference between optimal trait values between sites $\theta_i-\theta_j$, which can be assumed to be large if the remote site $j$ is a great distance away from the local site $i$.
 If sites $i$ and $j$ are very close, the stray rate is maximized at $m_{\rm max} = 0.5$, assuming both sites are equally attractive to the respective populations.
@@
 \centering
-\includegraphics[width=0.9\textwidth]{figs2/fig_relax_comb.pdf}
+\includegraphics[width=0.9\textwidth]{figs2/fig_relax_lowh.pdf}
 \caption{
-(a, b) Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming constant straying rates $m$ and (c, d) density dependent straying rates (evaluated at the steady state $m^*$).
+Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.2$.
 If $m$ is density dependent, in the alternative steady state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
@@
 \noindent{\bf (a) Nonlinear effects of straying on the portfolio effect and recovery time} \\
-\noindent Straying generally lowers steady state densities for both populations by \emph{i}) the donor population losing locally-adapted individuals to the recipient population and \emph{ii}) the introduction of maladapted individuals to the recipient population from the donor population, and this accords with observations from natural populations \cite{Bett:2017ha} %, regardless of trait heritability and variance or habitat heterogeneity
+\noindent Regardless of density dependence, straying lowers steady state densities for both populations by \emph{i}) the donor population losing locally-adapted individuals to the recipient population and \emph{ii}) the introduction of maladapted individuals to the recipient population from the donor population (Fig.
+\ref{fig:traj}).
+This prediction is in accordance with observations from natural populations \cite{Bett:2017ha}. %, regardless of trait heritability and variance or habitat heterogeneity
 The decline in steady state densities is not gradual: as straying increases, the system crosses a fold bifurcation whereby the single steady state for the metapopulation bifurcates into two alternative steady states: one at high biomass, and one at low biomass density (figure \ref{fig:traj}a, \ref{fig:PE}a).
-Mean trait values for both populations bifurcate similarly (figure \ref{fig:traj}b), depending on which population attains a low- vs.
-high-density, which in our system is random due to a small amount of introduced variation in the initial conditions.
-The fold bifurcation (the black line in Figs.
-\ref{fig:PE}a-c) occurs at lower values of the straying rate $m$ with decreased trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
+Mean trait values for both populations bifurcate similarly (figure \ref{fig:traj}b), depending on which population attains a low- vs. high-density. 
+At high stray levels, there are two eco-evolutionary states: a population will either have a higher degree of local adaptation (low trait offset) and higher density or be more maladapted and have lower density.
+Whether the population goes to one state or the other in our model is random, which is due to a small amount of introduced variance in the initial conditions.
+Thus, we predict that high levels of straying in metapopulations will produce alternative eco-evolutionary states.
 
@@
 Greater trait heritability results in a faster decline in $N_T^*$ with increasing straying rates $m$, but leads to only moderate changes to $\Delta N$.
-Conversely, if the trait is less heritable, an increase in the straying rate has little impact on the total biomass density but contrastingly large effects on $\Delta N$.
+Conversely, in the context of lower trait heritability, an increase in the straying rate has little impact on the total biomass density but contrastingly large effects on $\Delta N$.
+The fold bifurcation (the black line in Figs.
+\ref{fig:PE}a-c) occurs at lower values of the straying rate $m$ with decreased trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
+Although trait heritability among salmonids is variable, most life history traits show $h^2 <0.5$ \cite{Carlson:2008hl}, and we largely focus additional analyses on that range.
 
@@
 
+%Density dependent straying
+As with recovery time, we find that the portfolio effects generated in systems with density dependent straying are qualitatively similar to systems with constant straying, however there are some important quantitative differences.
+First, the PE associated with the high-density (low $m^*$) population is the same as that for a system with a constant $m$ (figure \ref{fig:thetaPE}b).
+As such, the difference in $m^*$ among low- and high-density populations appears to matter less than the straying generated by the high-density population.
+However, as $m^*$ increases, we observe a higher PE than that generated by systems with constant $m$.
+This higher PE at increased $m^*$ is mirrors recovery times that are also lower than those observed for systems with constant $m$ (figure \ref{fig:relax}).
+Together this suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our minimal model, we do find evidence that it can promote robustness, particularly when populations are low and straying is correspondingly high.
+
 % \noindent{\bf (c) Rates of recovery following catastrophic collapse}\\ 
@@
 As the portfolio effect is highly sensitive to the rate of straying between populations, so is the time required for the system to recover to a steady state following a large disturbance.
-We measured the time required for the system $N_T$ to relax to its steady state following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} are equivalent if the system is in the single steady state regime); (\emph{iii}) near-collapse of both populations where just 1.0\% of each survives.
-In general, we find that the average-CV portfolio effect is negatively correlated with recovery time (figure \ref{fig:PE}d), indicating that for our system both measures are valuable indicators of metapopulation robustness.
+In general, we find that the average-CV portfolio effect is negatively correlated with recovery time (figure \ref{fig:PE}d), indicating that, for our system, both measures are valuable indicators of metapopulation robustness.
 Because we can assess the time to recovery in response to the various disturbance types described above, this allows us to gain an in-depth perspective into the fragility of the metapopulation as a function of straying rate.
 
-At low rates of straying both populations are subject to the same steady state, such that extinction of either population results in the same recovery time, and when trait heritability is low, the time required to recover following the extinction of either population is generally less than or equal to the recovery time after near-collapse of both (figure \ref{fig:relax}a).
-In the alternative stable state regime where straying rates are higher, extinction of the smaller population results in the shortest recovery time, whereas as $m$ increases, extinction of the larger population results in the shortest recovery time.
-Because the mean trait values of both populations are skewed towards those of the high-density site, when the low-density population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into the recovering extends the length of time required for it to return to its steady state (figure \ref{fig:relaxtraj_ldlh}).
-Alternatively, when the high-density population collapses under high straying rates, the low-density population suffers loss as well due to the absence of incoming strays such that both populations must recover (figure \ref{fig:relaxtraj_hdlh}).
+Straying had non-linear impacts on the recovery time of populations. 
+When the dominant state (well adapted and high density) is wiped out, high levels of straying allow it to recover quickly (figure \ref{fig:relax}a) because the surviving population has a mean trait value skewed towards the optimum of the recovering population (figure \ref{fig:relaxtraj_hdlh}).
+Yet, as straying decreases, recovery time for the disturbed dominant state population increases, in part because there is enough time for the trait distribution to move back towards the trait optimum of the subordinate state population.
+In contrast, when the subordinate state population (maladapted and low density) is wiped out, recovery rates are fastest at low to intermediate levels of straying.
+When the subordinate state population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into it extends the length of time required for it to return to its steady state (figure \ref{fig:relaxtraj_ldlh}).
+%At low rates of straying, 
+Near the onset of the fold bifurcation, recovery times increase explosively regardless of straying rate, heritability, or disturbance type.
 
-Increased trait heritability flips this relationship by increasing the response of trait evolution to changes in demography (figure \ref{fig:relax}b).
-When straying rates are low, metapopulations that suffer near-collapse of both populations recover more quickly than systems suffering extinctions of either population.
-When straying rates are higher such that the system is in the alternative stable state regime, the metapopulation always recovers more quickly if the small population is lost (figure \ref{fig:relaxtraj_ldhh}), whereas the time to recovery is maximized if both populations suffer near-collapse.
-When the high-density population goes extinct, the stronger response of selection permits the surviving low-density population to grow quickly and fill the high-density state (figure \ref{fig:relaxtraj_hdhh}).
-Near the onset of the fold bifurcation, recovery times increase explosively regardless of straying rate, heritability, or disturbance type.
-%For straying rates in the alternative steady state regime, extinction of the smaller population consistently results in the shortest time to recovery.
+Density-dependent straying modifies these recovery times (figure \ref{fig:relax}b). 
+First, in comparison with constant stray rates, density-dependent straying made recovery more rapid at elevated stray rates when both populations collapsed and when the subordinate population was extirpated.
+%Recovery times
+%The recovery times for systems with constant and density dependent straying both show similar trends for a given $m$ and $m^*$ (figure \ref{fig:relax}c,d), however there are some distinct quantitative differences.
+When the straying rate and trait heritability were low, near-collapse of both populations result in longer than expected recovery times, whereas in the alternative stable state regime (higher $m^*$), the recovery times for different disturbance types are very similar to systems with a constant $m$ (figure \ref{fig:relax}b; note difference in x-axis scales).
+As trait heritability increases, the metapopulation always recovers more quickly if the small population is lost (figure \ref{fig:relax_highh}).
+% There was greater similarity in the recovery times for constant and density dependent $m$ systems when trait heritability was high (figure \ref{fig:relax_highh}).
+
+% At low rates of straying both populations are subject to the same steady state, such that extinction of either population results in the same recovery time, and when trait heritability is low, the time required to recover following the extinction of either population is generally less than or equal to the recovery time after near-collapse of both (figure \ref{fig:relax}a).
+% In the alternative stable state regime where straying rates are higher, extinction of the smaller population results in the shortest recovery time, whereas as $m$ increases, extinction of the larger population results in the shortest recovery time.
+% Because the mean trait values of both populations are skewed towards those of the high-density site, when the low-density population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into the recovering extends the length of time required for it to return to its steady state (figure \ref{fig:relaxtraj_ldlh}).
+% Alternatively, when the high-density population collapses under high straying rates, the low-density population suffers loss as well due to the absence of incoming strays such that both populations must recover (figure \ref{fig:relaxtraj_hdlh}).
+% 
+% Increased trait heritability flips this relationship by increasing the response of trait evolution to changes in demography (figure \ref{fig:relax}b).
+% When straying rates are low, metapopulations that suffer near-collapse of both populations recover more quickly than systems suffering extinctions of either population.
+% When straying rates are higher such that the system is in the alternative stable state regime, the metapopulation always recovers more quickly if the small population is lost (figure \ref{fig:relaxtraj_ldhh}), whereas the time to recovery is maximized if both populations suffer near-collapse.
+% When the high-density population goes extinct, the stronger response of selection permits the surviving low-density population to grow quickly and fill the high-density state (figure \ref{fig:relaxtraj_hdhh}).
+% Near the onset of the fold bifurcation, recovery times increase explosively regardless of straying rate, heritability, or disturbance type.
+% %For straying rates in the alternative steady state regime, extinction of the smaller population consistently results in the shortest time to recovery.
 
@@
 \noindent Increasing differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater regional differences in the conditions that favor alternative trait complexes, which we interpret here as greater habitat heterogeneity.
-Although trait heritability among salmonids is variable, most life history traits show $h^2 <0.5$ \cite{Carlson:2008hl}, and we largely focus additional analyses on that range.
 If both populations are isolated, natural selection will direct the mean trait values of both populations towards their respective optima, such that $\mu_i(t) \rightarrow \theta_i$ as $t\rightarrow\infty$.
@@
 We assessed metapopulation robustness across a range of $(m_i^*,m_j^*)$ values by varying $m_0$, which is positively and linearly related to $(m_i^*,m_j^*)$.
+\\
 
-%Recovery times
-The recovery times for systems with constant and density dependent straying both show similar trends for a given $m$ and $m^*$ (figure \ref{fig:relax}c,d), however there are some distinct quantitative differences.
-When the straying rate and trait heritability are low, near-collapse of both populations result in longer than expected recovery times, whereas in the alternative stable state regime (higher $m^*$), the recovery times for different disturbance types are very similar to systems with a constant $m$ (figure \ref{fig:relax}c; note difference in x-axis scales).
-There is greater similarity in the recovery times for constant and density dependent $m$ systems when trait heritability is high.
-
-As with recovery time, we find that the portfolio effects generated in systems with density dependent straying are qualitatively similar to systems with constant straying, however there are some important quantitative differences.
-First, the PE associated with the high-density (low $m^*$) population is the same as that for a system with a constant $m$ (figure \ref{fig:thetaPE}b).
-As such, the difference in $m^*$ among low- and high-density populations appears to matter less than the straying generated by the high-density population.
-However, as $m^*$ increases, we observe a higher PE than that generated by systems with constant $m$.
-This higher PE at increased $m^*$ is mirrors recovery times that are also lower than those observed for systems with constant $m$ (figure \ref{fig:relax}).
-Together this suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our minimal model, we do find evidence that it can promote robustness, particularly when populations are low and straying is correspondingly high.
-% 
 % 
@@
 % In other words,
-\\
 
@@
 Although there are various ways to measure `robustness' (REF:TREE), we employ two measures: 
-1) the average-CV portfolio effect, a statistical metric commonly used to assess the fragility of salmon populations; 
+1) the average-CV portfolio effect, a statistical metric commonly used to assess the fragility of populations; 
 2) the recovery time, defined here as the time required for the aggregate metapopulation biomass $N_T$ to return to its steady state after different types of induced disturbances, which is mechanistically linked to the likelihood of persistence (REF).
@@
 %Alternative steady states & critical slowing
-The immigration of dispersing strays with trait values far from the local optimum generally lowers the combined steady state biomass $N_T$ and increases the likelihood that the system gives way to alternative steady states, pushing one of the populations close to extinction.
+The immigration of dispersing individuals with trait values far from the local optimum generally lowers the combined steady state biomass $N_T$ and increases the likelihood that the system gives way to alternative steady states, pushing one of the populations close to extinction.
 The formation of alternative stable states in our system is a relatively common example of spatial pattern formation, and occurs above a threshold straying rate.
@@
 Here, we describe an interesting alternative mechanism through which pattern formation in local densities can occur, when the onset of movement mediated maladaptation leads to local differences in reproductive rates that yield differences in densities that would, in the absence of straying, be the same.
-This mechanism of pattern formation through the simultaneous interplay between ecological and evolutionary dynamics may open up new areas of research.
+%This mechanism of pattern formation through the simultaneous interplay between ecological and evolutionary dynamics may open up new areas of research.
 
@@
 On the other hand, the presence of \emph{just enough} straying to cause formation of alternative stable states both increases the portfolio effect (figure \ref{fig:PE}c) and reduces the rate of return -- particularly if straying is density dependent (figure \ref{fig:relax}c,d) as might be common in systems with collective navigation (REF).
+Importantly, the asymmetry in population densities also leads an asymmetry in the mean trait values between sites: because the high-density population contains the majority of biomass, the proportion of the low-density population that is composed of incoming strays is generally large, and this leads to mean trait values in both sites skewed towards that of the high-density site.
+This has important ramifications for metapopulation robustness, as measured by both the portfolio effect as well as recovery time.
 
@@
 \end{figure}
+% 
+% \begin{figure}
+%   \captionsetup{justification=raggedright,
+% singlelinecheck=false
+% }
+% \centering
+% \includegraphics[width=0.35\textwidth]{figs2/fig_relax_small_highh.pdf}
+% \caption{
+% Extinction of low-density population with a high constant straying rate $m=0.4$ and high trait heritability $h^2=0.8$ (see figure \ref{fig:relax}a).
+% Black line marks the calculated point of recovery post-perturbation.
+% Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
+% } \label{fig:relaxtraj_ldhh}
+% \end{figure}
+% 
+% \begin{figure}
+%   \captionsetup{justification=raggedright,
+% singlelinecheck=false
+% }
+% \centering
+% \includegraphics[width=0.35\textwidth]{figs2/fig_relax_large_highh.pdf}
+% \caption{
+% Extinction of high-density population with a high straying rate $m=0.4$ and high trait heritability $h^2=0.8$ (see figure \ref{fig:relax}a).
+% Black line marks the calculated point of recovery post-perturbation.
+% Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
+% } \label{fig:relaxtraj_hdhh}
+% \end{figure}
 
-\begin{figure}
+\begin{figure*}
   \captionsetup{justification=raggedright,
@@
 \centering
-\includegraphics[width=0.35\textwidth]{figs2/fig_relax_small_highh.pdf}
+\includegraphics[width=0.9\textwidth]{figs2/fig_relax_highh.pdf}
 \caption{
-Extinction of low-density population with a high constant straying rate $m=0.4$ and high trait heritability $h^2=0.8$ (see figure \ref{fig:relax}a).
-Black line marks the calculated point of recovery post-perturbation.
-Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
-} \label{fig:relaxtraj_ldhh}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.35\textwidth]{figs2/fig_relax_large_highh.pdf}
-\caption{
-Extinction of high-density population with a high straying rate $m=0.4$ and high trait heritability $h^2=0.8$ (see figure \ref{fig:relax}a).
-Black line marks the calculated point of recovery post-perturbation.
-Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
-} \label{fig:relaxtraj_hdhh}
-\end{figure}
+Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.8$.
+If $m$ is density dependent, in the alternative steady state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
+} \label{fig:relax_highh}
+\end{figure*}
 

### 2017-05-28 manuscript/msv3.tex
comparing PE fig

@@
 \noindent{\bf (a) Nonlinear effects of straying on the portfolio effect and recovery time} \\
-\noindent Regardless of density dependence, straying lowers steady state densities for both populations by \emph{i}) the donor population losing locally-adapted individuals to the recipient population and \emph{ii}) the introduction of maladapted individuals to the recipient population from the donor population (Fig.
+\noindent Regardless of density dependence, straying lowers steady state densities for both populations by (\emph{i}) the donor population losing locally-adapted individuals to the recipient population and (\emph{ii}) the introduction of maladapted individuals to the recipient population from the donor population (Fig.
 \ref{fig:traj}).
@@
 Mean trait values for both populations bifurcate similarly (figure \ref{fig:traj}b), depending on which population attains a low- vs. high-density. 
-At high stray levels, there are two eco-evolutionary states: a population will either have a higher degree of local adaptation (low trait offset) and higher density or be more maladapted and have lower density.
-Whether the population goes to one state or the other in our model is random, which is due to a small amount of introduced variance in the initial conditions.
-Thus, we predict that high levels of straying in metapopulations will produce alternative eco-evolutionary states.
+Above the threshold straying rate defined by the fold bifurcation, there are two alternative eco-evolutionary states: the \emph{dominant state} population will have a higher a higher density and higher degree of local adaptation (lower trait offset from the local optimum), while the \emph{subordinate state} population will have lower density with maladapted trait values (higher trait offset from the local optimum).
+Whether a specific population goes to one state or the other in our model is random, which is due to a small amount of introduced variance in the initial conditions.
+% Thus, we predict that high levels of straying in metapopulations will produce alternative eco-evolutionary states.
 
-Trait heritability has a large impact on the sensitivity of the straying rate to both the aggregate population steady state density ($N^*_T=N^*_1+N^*_2$; figure \ref{fig:PE}a) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N=|N^*_1-N^*_2|$; figure \ref{fig:PE}b).
+Trait heritability has a large impact on the degree to which straying rate affects both the aggregate population steady state density ($N^*_T=N^*_1+N^*_2$; figure \ref{fig:PE}a) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N=|N^*_1-N^*_2|$; figure \ref{fig:PE}b).
 Greater trait heritability results in a faster decline in $N_T^*$ with increasing straying rates $m$, but leads to only moderate changes to $\Delta N$.
@@
 \ref{fig:PE}a-c) occurs at lower values of the straying rate $m$ with decreased trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
-Although trait heritability among salmonids is variable, most life history traits show $h^2 <0.5$ \cite{Carlson:2008hl}, and we largely focus additional analyses on that range.
+Although trait heritability among salmonids is variable, most life history traits have an $h^2 <0.5$ \cite{Carlson:2008hl}, and we largely focus additional analyses on that range.
 
-%Together these changes in steady state population densities in terms of $N_T$ and $\Delta N$ as a function of trait heritability and the rate of straying between populations give rise to highly nonlinear portfolio effects, which we quantify here as
-
-% Nonlinear PE(!!!)
-
-%In the region where there is a single steady state among both populations, we find an low-intermediate portfolio effect ca. ${\rm PE}=1.4$, primarily due to the elevated mean values of $N_T$.
+%Critical slowing down
 As the fold bifurcation is approached with increasing $m$, the portfolio effect increases sharply due to an amplification in variance within both donor and recipient populations ${\rm VAR}(N_{i,j})$.
-This variance increase is the product of a dynamical process known as critical slowing down that occurs near fold bifurcations \cite{Scheffer:2009gg}, a phenomenon that some have suggested may serve as an early warning indicator for approaching phase transitions \cite{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br}.
+This variance increase is the product of a dynamical process known as \emph{critical slowing down} that occurs near fold bifurcations \cite{Scheffer:2009gg}, a phenomenon that some have suggested may serve as an early warning indicator for approaching phase transitions \cite{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br}.
 For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}a-c), where alternative steady states occur, the portfolio effect declines steadily as the CV of $N_T$ increases.
@@
 %Density dependent straying
-As with recovery time, we find that the portfolio effects generated in systems with density dependent straying are qualitatively similar to systems with constant straying, however there are some important quantitative differences.
+In the alternative stable state regime, because the two populations exist in a dominant and subordinate state, there are also two alternative straying rates $(m_i^*,m_j^*)$: the higher of which associated with the subordinate, low-density population, and the lower of which is associated with the dominant, high-density population.
+We assessed metapopulation robustness across a range of $(m_i^*,m_j^*)$ values by varying the probability that an individual strays $m_0$, which is positively and linearly related to $(m_i^*,m_j^*)$.
+We find that the portfolio effects generated in systems with density dependent straying are qualitatively similar to systems with constant straying, however there are some important quantitative differences.
 First, the PE associated with the high-density (low $m^*$) population is the same as that for a system with a constant $m$ (figure \ref{fig:thetaPE}b).
-As such, the difference in $m^*$ among low- and high-density populations appears to matter less than the straying generated by the high-density population.
-However, as $m^*$ increases, we observe a higher PE than that generated by systems with constant $m$.
-This higher PE at increased $m^*$ is mirrors recovery times that are also lower than those observed for systems with constant $m$ (figure \ref{fig:relax}).
-Together this suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our minimal model, we do find evidence that it can promote robustness, particularly when populations are low and straying is correspondingly high.
+% As such, the difference in $m^*$ between subordinate and dominant populations appears to matter less than the lower straying rate generated by the dominant population.
+As $m^*$ increases, we observe a higher PE than systems with constant $m$.
 
@@
 %At low rates of straying, 
-Near the onset of the fold bifurcation, recovery times increase explosively regardless of straying rate, heritability, or disturbance type.
+Near the onset of the fold bifurcation, recovery time increases explosively regardless of straying rate, heritability, or disturbance type, and as the name implies, this is characteristic of \emph{slowing down} dynamics near critical transitions \cite{Kuehn:2010p2591}.
 
@@
 % There was greater similarity in the recovery times for constant and density dependent $m$ systems when trait heritability was high (figure \ref{fig:relax_highh}).
+The lower recovery time for systems with increased $m^*$ mirrors an elevated PE with higher density dependent straying rates (figure \ref{fig:PEtheta}).
+% This higher PE at increased $m^*$ mirrors recovery times that are also lower than those observed for systems with constant $m$ (figure \ref{fig:relax}).
+Together this suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our minimal model, we do find evidence that it can promote robustness, particularly when the aggregate biomass is low and straying is correspondingly high.
 

### 2017-05-28 manuscript/msv3.tex
added FB notation on figs and found some really cool distance dependent stuff

@@
 
-\noindent{\bf (b) Recruitment over a selective landscape}
+% \noindent{\bf (b) Recruitment over a selective landscape}
 \noindent Because individuals from the local population are mixed with individuals from the remote population via straying and subsequent reproduction, the resulting trait distribution is a mixed normal with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and for the straying individuals, $1-w_i$, where
@@
 
-\noindent {\bf (c) Measuring metapopulation robustness}
+\noindent {\bf (b) Measuring metapopulation robustness}
 \noindent We evaluated metapopulation robustness by measuring the average-CV portfolio effect (PE) \cite{Anderson:2014cx,Schindler:2015gf} as well as the time required for the system to return to a steady state following an induced disturbance to one or both of the populations \cite{Ovaskainen:2002il}.
@@
 As the CV of $N_T$ decreases relative to that of the constituent populations, $\langle{\rm PE}\rangle > 1$, and the metapopulation is presumed to become more stable.
-{\color{red} A similar measure of metapopulation robustness can be evaluated by measuring the synchronization $\phi$ between populations, where 
-\begin{equation}
-\label{eq_sync}
-\phi = \frac{{\rm VAR}(N_T)}{\left(\sum_{i=1}^N\sqrt{{\rm VAR}(N_i)}\right)^2}.
-\end{equation}
-By rearranging and substituting eq. (\ref{eq_sync}) into eq. (\ref{eq:pe}), we can observe that $\langle{\rm PE}\rangle \propto \sqrt{\phi}^{-1}$.
-Thus, the dynamical diversity of the metapopulation offers a mirror to the portfolio effect, where perfect synchrony ($\phi = 1$) also means $\langle{\rm PE}\rangle = 1$.}
+% {\color{red} A similar measure of metapopulation robustness can be evaluated by measuring the synchronization $\phi$ between populations, where 
+% \begin{equation}
+% \label{eq_sync}
+% \phi = \frac{{\rm VAR}(N_T)}{\left(\sum_{i=1}^N\sqrt{{\rm VAR}(N_i)}\right)^2}.
+% \end{equation}
+% By rearranging and substituting eq. (\ref{eq_sync}) into eq. (\ref{eq:pe}), we can observe that $\langle{\rm PE}\rangle \propto \sqrt{\phi}^{-1}$.
+% Thus, the dynamical diversity of the metapopulation offers a mirror to the portfolio effect, where perfect synchrony ($\phi = 1$) also means $\langle{\rm PE}\rangle = 1$.}
 %Moreover, if and only if $\phi \neq 1$, it is expected that as the diversity of the aggregate increases, so should the portfolio effect, and this central thesis is thought to underlie the stability of populations, species, communities and even financial markets.
-Portfolio effects greater than unity corresponds to less synchronization ($\phi < 1$) \cite{Loreau:2008ju,Anderson:2014cx,Yeakel:2013vz} and thus a greater potential for demographic rescue among populations, buffering the system as a whole against extinction.
+Portfolio effects greater than unity corresponds to less synchronization  \cite{Loreau:2008ju,Anderson:2014cx,Yeakel:2013vz} and thus a greater potential for demographic rescue among populations, buffering the system as a whole against extinction.
 
@@
 
-\noindent {\bf (d) The effects of density and distance on the rate of straying}
+\noindent {\bf (c) The effects of density and distance on the rate of straying}
 %Density dependent m
@@
 
-\begin{figure*}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.9\textwidth]{figs2/fig_relax_lowh.pdf}
-\caption{
-Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.2$.
-If $m$ is density dependent, in the alternative steady state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
-} \label{fig:relax}
-\end{figure*}
-
 \section{Results}
@@
 %% For low heritability (h = 0.01 - 0.2 or so), moderate and insensitive changes in total biomass, but large differences between the values of alternative steady states leads to a local maximum for PE at low-moderate stray rates
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.4\textwidth]{figs2/fig_thetaPEmvm.pdf}
+\caption{
+(a) Median portfolio effect as a function of a constant stray rate $m$ (solid line) and density dependent stray rate (point pairs) given heritability is $h^2 < 0.5$ and $\Delta\theta=5$.
+Point pairs connected by a horizontal line represent the PE as a function of density dependent straying rates, evaluated for both low- and high-density populations at equilibrium.
+The lower straying rate of a pair is for the larger population; the higher straying rate is for the smaller population.
+(b) Median portfolio effects for habitats with increasing heterogeneity as measured by the difference in regional trait optima $\Delta \theta$ for both constant and density dependent stray rates as shown in (a).
+} \label{fig:thetaPE}
+\end{figure}
 
@@
 % Thus, we predict that high levels of straying in metapopulations will produce alternative eco-evolutionary states.
+
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.9\textwidth]{figs2/fig_relax_lowh.pdf}
+\caption{
+Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.2$.
+If $m$ is density dependent, in the alternative steady state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
+} \label{fig:relax}
+\end{figure*}
 
@@
 
+If we assume that the rate of straying is density dependent, the probability that an individual strays $m_0$ determines the rate of straying within the population, such that $m(t)$ becomes lower as $N(t)$ increases, likely due to the effects of collective decision-making \cite{Berdahl:2016dx} (Eq.
+\ref{eq:ddm}).
+%Because our analyses concern steady state conditions, the inclusion of density-dependent straying does not have a large impact on the qualitative nature of our findings (figure \ref{fig:ddm}).
+Density dependence alters the straying rate at steady state population densities because $0 < m^* < m_0$, and this serves to rescale both the strength of the PE as well as the recovery time, but does not change the qualitative nature of our findings.
+In the alternative stable state regime, because each population exists at different steady state densities, there are likewise two alternative straying rates $(m_i^*,m_j^*)$: the higher straying rate is associated with the low-density population, and the lower straying rate is associated with the high-density population.
+% We assessed metapopulation robustness across a range of $(m_i^*,m_j^*)$ values by varying $m_0$, which is positively and linearly related to $(m_i^*,m_j^*)$.
 %Density dependent straying
-In the alternative stable state regime, because the two populations exist in a dominant and subordinate state, there are also two alternative straying rates $(m_i^*,m_j^*)$: the higher of which associated with the subordinate, low-density population, and the lower of which is associated with the dominant, high-density population.
+% In the alternative stable state regime, because the two populations exist in a dominant and subordinate state, there are also two alternative straying rates $(m_i^*,m_j^*)$: the higher of which associated with the subordinate, low-density population, and the lower of which is associated with the dominant, high-density population.
 We assessed metapopulation robustness across a range of $(m_i^*,m_j^*)$ values by varying the probability that an individual strays $m_0$, which is positively and linearly related to $(m_i^*,m_j^*)$.
 We find that the portfolio effects generated in systems with density dependent straying are qualitatively similar to systems with constant straying, however there are some important quantitative differences.
-First, the PE associated with the high-density (low $m^*$) population is the same as that for a system with a constant $m$ (figure \ref{fig:thetaPE}b).
+First, the PE associated with the high-density (low $m^*$) population is the same as that for a system with a constant $m$ (figure \ref{fig:thetaPE}a).
 % As such, the difference in $m^*$ between subordinate and dominant populations appears to matter less than the lower straying rate generated by the dominant population.
-As $m^*$ increases, we observe a higher PE than systems with constant $m$.
+As $m^*$ increases, we observe an increase in the PE than for systems with constant $m$.
 
@@
 In contrast, when the subordinate state population (maladapted and low density) is wiped out, recovery rates are fastest at low to intermediate levels of straying.
-When the subordinate state population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into it extends the length of time required for it to return to its steady state (figure \ref{fig:relaxtraj_ldlh}).
+Because the mean trait values of both populations are skewed towards those of the dominant population, when the subordinate population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into the recovering population extends the length of time required for it to return to its steady state (figure \ref{fig:relaxtraj_ldlh}).
+% When the subordinate state population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into it extends the length of time required for it to return to its steady state (figure \ref{fig:relaxtraj_ldlh}).
 %At low rates of straying, 
-Near the onset of the fold bifurcation, recovery time increases explosively regardless of straying rate, heritability, or disturbance type, and as the name implies, this is characteristic of \emph{slowing down} dynamics near critical transitions \cite{Kuehn:2010p2591}.
+When both populations are both dramatically reduced, recovery time is generally fastest at lower levels of straying, while near the onset of the fold bifurcation, recovery time increases explosively and this is -- as the name implies -- characteristic of \emph{slowing} dynamics that occur near critical transitions \cite{Scheffer:2009gg,Kuehn:2010p2591}.
 
-Density-dependent straying modifies these recovery times (figure \ref{fig:relax}b). 
+Density-dependent straying alters these recovery times (figure \ref{fig:relax}b). 
 First, in comparison with constant stray rates, density-dependent straying made recovery more rapid at elevated stray rates when both populations collapsed and when the subordinate population was extirpated.
@@
 %The recovery times for systems with constant and density dependent straying both show similar trends for a given $m$ and $m^*$ (figure \ref{fig:relax}c,d), however there are some distinct quantitative differences.
-When the straying rate and trait heritability were low, near-collapse of both populations result in longer than expected recovery times, whereas in the alternative stable state regime (higher $m^*$), the recovery times for different disturbance types are very similar to systems with a constant $m$ (figure \ref{fig:relax}b; note difference in x-axis scales).
-As trait heritability increases, the metapopulation always recovers more quickly if the small population is lost (figure \ref{fig:relax_highh}).
+At low straying rates, near-collapse of both populations resulted in longer than expected recovery times, whereas in the alternative stable state regime (higher $m^*$), the recovery times for different disturbance types were very similar to systems with a constant $m$ (figure \ref{fig:relax}b; note difference in x-axis scales).
+As trait heritability increased, the metapopulation always recovered more quickly if the small population was lost (figure \ref{fig:relax_highh}).
 % There was greater similarity in the recovery times for constant and density dependent $m$ systems when trait heritability was high (figure \ref{fig:relax_highh}).
@@
 % This higher PE at increased $m^*$ mirrors recovery times that are also lower than those observed for systems with constant $m$ (figure \ref{fig:relax}).
-Together this suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our minimal model, we do find evidence that it can promote robustness, particularly when the aggregate biomass is low and straying is correspondingly high.
+In tandem, analysis of both PE and recovery time suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our minimal model, it does appear to promote robustness, particularly when the aggregate biomass is low and straying is correspondingly high.
 
@@
 Increased rates of straying lowers phenotypic diversity ($\Delta \mu^* = |\mu_i(t)-\mu_j(t)|$, evaluated at the steady state) because both local and remote populations are increasingly homogenized.
-The loss of phenotypic diversity with increased straying is greater if trait heritability is low because the selective forces acting against trait means far from the local optima are lessened.
+The loss of phenotypic diversity with increased straying is greater if trait heritability is low because traits take longer to go back to their local optima than they do when heritability is large. 
+Hence straying counters the effect of diversifying local adaptation. 
+% The loss of phenotypic diversity with increased straying is greater if trait heritability is low because the selective forces acting against trait means far from the local optima are lessened.
 % with decreased $h^2$.
@@
 \\
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.4\textwidth]{figs2/fig_thetaPEmvm.pdf}
-\caption{
-(a) Median portfolio effect as a function of stray rate $m$ given heritability is $h^2 < 0.5$ for habitats with increasing heterogeneity as measured by the difference in regional trait optima $\Delta \theta$.
-(b) Median portfolio effects as shown in (a) are denoted by the solid lines. 
-Point pairs connected by a horizontal line represent the PE as a function of density dependent straying rates, evaluated for both low- and high-density populations at equilibrium.
-The lower straying rate of a pair is for the larger population; the higher straying rate is for the smaller population.
-The different steady state regimes that give rise to the range of straying rates were obtained by varying the individual straying probability $m_0$ (where $m_0$ and $m^*$ are linearly related).
-} \label{fig:thetaPE}
-\end{figure}
 
@@
 \noindent{\bf (b) The influence of habitat heterogeneity on metapopulation robustness}\\ 
-\noindent Increasing differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater regional differences in the conditions that favor alternative trait complexes, which we interpret here as greater habitat heterogeneity.
+\noindent Increasing differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater regional differences in the conditions that favor alternative trait complexes, which we interpret here as greater habitat heterogeneity (REFS).
 If both populations are isolated, natural selection will direct the mean trait values of both populations towards their respective optima, such that $\mu_i(t) \rightarrow \theta_i$ as $t\rightarrow\infty$.
@@
 %The rate of straying that gives rise to a large PE at the fold bifurcation is conditioned on the difference in optimal trait values between the local and remote populations, which we call $\Delta\theta = \sqrt{(\theta_i - \theta_j)^2}$.
-As habitat heterogeneity increases, alternative stable states appear for lower straying rates -- with the crossing of the fold bifurcation, accompanied by a peak in the PE -- whereas the magnitude of increase in the PE increases (figure \ref{fig:thetaPE}a), reducing recovery time (figure \ref{fig:relaxtheta}).
-For increased rates of straying, greater habitat heterogeneity erodes the PE and increases the recovery time (figure \ref{fig:thetaPE}a, figure \ref{fig:relaxtheta})
-Taken together, habitat heterogeneity, as measured as the differences in trait optima between two habitats, promotes robustness when straying rates are low, but erodes robustness when straying rates are high.
+As habitat heterogeneity increases, alternative stable states appear at lower straying rates -- with the crossing of the fold bifurcation, accompanied by a peak in the PE -- whereas the magnitude of increase in the PE also increases (figure \ref{fig:thetaPE}b), reducing recovery time (figure \ref{fig:relaxtheta}).
+For increased rates of straying, greater habitat heterogeneity erodes the PE (figure \ref{fig:thetaPE}b) and increases the recovery time (figure \ref{fig:relaxtheta}).
+These results together suggest that habitat heterogeneity, as measured as the differences in trait optima between two habitats $\Delta\theta$, promotes robustness when straying rates are low, and erodes robustness when straying rates are high.
 %This means that alternative stable states can emerge with relatively little dispersal between sites (figure \ref{fig:thetaPE}), however the increased PE (and shorter recovery times).
 \\
-
-% Density dependent m
-%% Because m[t] @ steady state 0<m[t]<m0, we observe the same dynamics for higher m0 as we do for lower constant m.
-%% Higher rates of straying are tolerable (there is a not a decline in PE for very high m0 as there is for constant m)
-
-\noindent{\bf (c) The effects of density dependent straying}\\
-\noindent If we assume that the rate of straying is density dependent, the probability that an individual strays $m_0$ determines the rate of straying within the population, such that $m(t)$ becomes lower as $N(t)$ increases due to the effects of collective decision-making \cite{Berdahl:2016dx} (Eq.
-\ref{eq:ddm}).
-%Because our analyses concern steady state conditions, the inclusion of density-dependent straying does not have a large impact on the qualitative nature of our findings (figure \ref{fig:ddm}).
-Density dependence alters the straying rate at steady state population densities because $0 < m^* < m_0$, and this serves to rescale both the strength of the PE as well as the recovery time, but importantly does not change the qualitative nature of our findings for constant straying rates.
-In the alternative stable state regime, because the each population exists at different steady state densities, there are likewise two alternative straying rates $(m_i^*,m_j^*)$: the higher straying rate is associated with the low-density population, and the lower straying rate is associated with the high-density population.
-We assessed metapopulation robustness across a range of $(m_i^*,m_j^*)$ values by varying $m_0$, which is positively and linearly related to $(m_i^*,m_j^*)$.
-\\
-
-% 
-% In a metapopulation with density dependent straying, the lower straying rate of the high-density population generates a portfolio effect that would be expected if straying were constant.
-% 
-% A metapopulation with density dependent straying has a portfolio effect 
-% 
-% equal to that of a metapopulation with constant straying
-% 
-% as would be expected for a metapopulation with a constant straying rate equalt to that of the 
-% 
-% We find that the by-definition lower straying rate of the high-density population (left-most of the paired points in figure \ref{fig:thetaPE}b) largely determines the portfolio effect for the metapopulation.
-% In other words,
-
-\begin{figure*}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-  \centering
-  \includegraphics[width=0.8\textwidth]{figs2/fig_MDPE_hm_mtheta_rt.pdf}
-  \caption{
-  Assuming that the rate of straying is linked directly to habitat heterogeneity.
-A low stray rate corresponds to very different (or distant) habitats (high $\Delta\theta$), whereas a higher rate of straying corresponds to very similar (or nearby) habitats (low $\Delta\theta$).
-  (a) Portfolio effect over constant straying rate $m$ and trait heritability $h^2$.
-  (b-c) Recovery times for three disturbance types for (b) constant $m$ and (c) density dependent straying rate at the steady state $m^*$.
-  } \label{fig:mtheta}
-\end{figure*}
 
@@
 
-\noindent{\bf (d) Distance dependent straying and habitat heterogeneity}\\ 
-\noindent We have so far treated $\Delta\theta$ and $m$ as independent parameters, however we may also assume that if environmental heterogeneity increases with distance, the rate of straying may be expected to decline given that individuals will stray less into distant habitats.
+\noindent{\bf (c) Distance dependent straying and habitat heterogeneity}\\ 
+\noindent We have so far treated $\Delta\theta$ and $m$ as independent parameters, however we may also assume that if environmental heterogeneity increases with distance between sites, the rate of straying may be expected to decline with habitat heterogeneity, given that individuals will stray less into distant habitats.
 %-- in particular North-South difference if trait optimality is largely temperature-dependent --
+Alternatively, individuals may be less likely to stray into very different habitats due to the infleunce of environmental cues on individual decision-making (REFS?).
 If we incorporate this interdependence of $m$ and $\Delta\theta$, low rates of straying correspond to mixing dissimilar (distant) populations, and high rates of straying would correspond to mixing similar (nearby) populations.
 
-We find that alternative stable states now appear at low straying rates, which corresponds to greater $\Delta\theta$.
-As the straying rate increases, a single stable state emerges.
-As before, the portfolio effect is higher in the alternative stable state regime, with a sharp decline in the PE as the straying rate becomes larger (figure \ref{fig:mtheta}a).
-This is in accordance with intuition as increasing straying rates mean that two very similar populations are mixing, such that the strength of selection on recruitment is lessened.
+We find that alternative stable states now appear for very low rates of straying $0 < m \leq 0.43$, corresponding to greater habitat heterogeneity $\Delta\theta$.
+As the straying rate increases, a single stable state emerges as the fold bifurcation is crossed.
+As straying increases (now the single steady state regime), the time required for recovery following extinction of either the population declines (to the right of figure \ref{fig:mtheta}).
+However, as straying decreases the system crosses the fold bifurcation, entering the alternative steady state regime (to the left in figure \ref{fig:mtheta}).
+At low rates of straying, there is a small amount of mixing between dissimilar (distant) populations.
+Here we find that if the subordinate population is wiped out, the time to recovery is minimized, whereas if the dominant population is wiped out, the time to recovery is maximized; near-collapse of both results in intermediate recovery times.
 
-This dynamic is more complex when we observe the recovery times for systems with either a constant (figure \ref{fig:mtheta}b) or density dependent (figure \ref{fig:mtheta}c) straying rate.
-Although recovery times are lowest in the alternatice stable state regime ($m<0.1$), this is only true if the low-density population suffers extinction.
-Alternatively, at low rates of straying between dissimilar (distant) populations, recovery times become maximized if the high-density population suffers extinction.
-As rates of straying -- and the similarity of trait optima -- increase, extinction of either population requires less time for recovery than near-collapse of both.
+When straying rates are low and the difference in trait optima are correspondingly high, we might presume that extinction of the dominant population would result in relatively faster recovery.
+This would be a logical assumption because the mean trait value of the subordinate population is skewed towards that of the dominant (larger) population, such that recruitment at the disturbed site would be high.
+Our results show, however, that this assumption is wrong: although the mean trait value of the subordinate population \emph{is} skewed towars the optimum of the dominant population, when the latter is wiped out and the rate of straying is low, there is enough time and isolation for the subordinate trait mean to shift towards its own optimum, and away from that of the recovering dominant population (figure \ref{fig:inertia}).
+This \emph{selective inertia} occurs until the dominant population grows large enough that the evolving subordinate phenotype is overwhelmed by incoming strays, shifting it back towards that of the now-recovered dominant population.
+Importantly, this effect can lead to much longer recovery times for the dominant population, as well as a switch in which population is dominant/subordinate (a switch in who occupies the alternative states), when the straying rate is very low (and $\Delta\theta$ is large).
+The threshold value of $m$, below which this behavior occurs, is marked by the asterisk in figure \ref{fig:mtheta}. % at m = 0.034
+These results hold for both constant and density dependenty straying rates (figure \ref{fig:mthetamvm})
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+  \centering
+  \includegraphics[width=0.5\textwidth]{figs2/fig_mtheta_rt.pdf}
+  \caption{
+  Distance dependent recovery times for three disturbance types. When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases.
+  } \label{fig:mtheta}
+\end{figure}
 
@@
   \centering
+  \includegraphics[width=0.35\textwidth]{figs2/fig_MDPE_hm_mtheta_rt.pdf}
+  \caption{
+  Distance dependent portfolio effects as a function of straying rate $m$ and trait heritability $h^2$. When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases.
+  } \label{fig:mthetaPE}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.35\textwidth]{figs2/fig_relax_inertia.pdf}
+\caption{
+Distance dependent straying, where increased differences in trait optima between sites $\Delta\theta$ corresponds to lower rates of straying $m$.
+At low rates of straying $m=0.02$ ($\Delta\theta=24$), extinction of the dominant population leads to slower-than-expected recovery times because the subordinate population is isolated enough to evolve towards its own trait optimum until growth of the dominant population overwhelms this local selection.
+In this case, $m$ is less than $m=0.034$ (denoted by the asterisk in figure \ref{fig:mtheta}), such that isolation allows the subdominant population to `run away' from the influence of the dominant population, leading to a switch in states.
+} \label{fig:inertia}
+\end{figure}
+
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+  \centering
+  \includegraphics[width=0.8\textwidth]{figs2/fig_relax_mtheta.pdf}
+  \caption{
+  Distance dependent recovery times for three disturbance types.
+When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases for constant (a) and density dependent (b) staying rates.
+  The fold bifurcation is not as clear in (b) because $\Delta\theta$ is a function of the individual straying rate $m_0$, whereas the x-axis in (b) is the straying rate at the steady state $m^*$.
+  Despite this difference, the general trends shown in (a) are also present in (b).
+  } \label{fig:mthetamvm}
+\end{figure*}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
 \includegraphics[width=0.35\textwidth]{figs2/fig_relax_both_lowh.pdf}

### 2017-05-28 model/KevinSimulations2.jl

@@ -162,8 +162,12 @@ pdf($namespace,height=8,width=5)
 par(mfrow=c(2,1),mai = c(0.8, 0.8, 0.1, 0.1))
 plot($mvec,$n1mean,pch='.',col=cols[4],xlab="Straying rate m",ylab="Steady state biomass",cex=0.5,ylim=c(0,max($n1mean)))
 points($mvec,$n2mean,pch='.',col=cols[4],cex=0.5)
+arrows($(mvec[indmax(pe)]),1200,$(mvec[indmax(pe)]),1100,length=0.05,angle=40,lwd=3)
+text($(mvec[indmax(pe)]),1275,'FB')
 plot($mvec,$x1mean,pch=".",col=cols[1],ylim=c(-5,5),xlab="Straying rate m",ylab="Trait offset")
 points($mvec,$x2mean,pch=".",col=cols[2])
+arrows($(mvec[indmax(pe)]),3.4,$(mvec[indmax(pe)]),2.5,length=0.05,angle=40,lwd=3)
+text($(mvec[indmax(pe)]),4,'FB')
 dev.off()
 """
 

### 2017-05-28 model/KevinSimulations_mtheta2.jl

@@ -3,6 +3,90 @@ using RCall
 
 include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve_mtheta.jl")
 
+
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve.jl")
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve_ddm.jl")
+
+
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolveExtinct.jl")
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/timeSS.jl")
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/movingaverage.jl")
+
+
+
+tmax=10000;
+z=0.5;
+rmax=2.0;
+beta=0.001;
+theta1=5.0;
+thetascale=2.0;
+tau=1.0;
+h=0.2;
+sigmaE=0;
+sigmaG=1;
+perror=0.01;
+refuge=0.01;
+m=0.02;
+
+thetadiff = (1-2*m)/(thetascale*m);
+
+extpop = "large";
+t_ext = Int64(round(tmax/2));
+
+n1, n2, x1, x2, w1, w2 = 
+KevinEvolveExtinct(tmax, 
+z, 
+rmax,
+beta,
+theta1,
+thetadiff,
+tau,
+h,
+sigmaE,
+sigmaG,
+m,
+perror,
+extpop,
+t_ext,
+refuge
+);
+
+t_ss, relaxtime = timeSS(n1,n2,t_ext);
+
+
+R"""
+library(RColorBrewer)
+pal = brewer.pal(3,'Set1')
+par(mfrow=c(2,1),mai = c(0.8, 0.8, 0.4, 0.1))
+plot($n1,type='l',col=pal[1],xlim=c($t_ext,$t_ss+50),ylim=c(0,max($n1,$n2)),ylab='Population density',xlab='Time',main=paste(c('RT=',$relaxtime)))
+lines($n2,col=pal[2])
+lines(c($t_ss,$t_ss),c(-10,10000))
+plot($x1,type='l',col=pal[1],ylim=c(2,100),xlim=c($t_ext,$t_ss+50),ylab='Trait means',xlab='Time')
[…]

### 2017-05-28 model/mvsddm2.jl

@@ -233,16 +233,25 @@ ma_medindm8max = movingaverage(medindm8max,window);
 mlist = collect(1:2:length(ma_mvec));
 mddmlist = collect(1:10:length(ma_medpe5_ddm));
 
-namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/figs2/fig_thetaPEmvm.pdf");
+#lines($(ma_mvecsmooth),$(ma_medpe8smooth),col=pal[3],lwd=3)
+#lines($(ma_mvecsmooth),$(ma_medpe3smooth),col=pal[1],lwd=3)
+
+namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/Manuscript/figs2/fig_thetaPEmvm.pdf");
 R"""
 library(RColorBrewer)
 pal = brewer.pal(3,'Set1')
 pdf($namespace,height=8,width=5)
 par(mfrow=c(2,1),mai = c(0.8, 0.8, 0.1, 0.1))
 plot($(ma_mvecsmooth),$(ma_medpe5smooth),col=pal[2],type='l',ylim=c(1,max($([ma_medpe3[mlist] ma_medpe5[mlist] ma_medpe8[mlist]]))),xlab='m',ylab='PE',lwd=3)
-lines($(ma_mvecsmooth),$(ma_medpe8smooth),col=pal[3],lwd=3)
-lines($(ma_mvecsmooth),$(ma_medpe3smooth),col=pal[1],lwd=3)
-legend(x=0.41,y=2.6,legend=c(3,5,8),col=pal,pch=22,xpd=TRUE,pt.bg=pal,cex=0.8, bty="n",title=expression(paste(Delta,theta)))
+arrows(0.053,2.48,0.053,2.35,length=0.05,angle=40,lwd=3)
+text(0.053,2.55,'FB')
+
+points($(ma_medindm5max[mddmlist]),$(ma_medpe5_ddm[mddmlist]),col=pal[2],pch=16,cex=0.8)
+points($(ma_medindm5min[mddmlist]),$(ma_medpe5_ddm[mddmlist]),col=pal[2],pch=16,cex=0.8)
+l = length($(ma_medindm5max[mddmlist]))
+for (i in 1:l) {
[…]

### 2017-05-31 Manuscript/CoverLetter/YGWM_cover.tex
reorganization of sections and minimization

\documentclass[ucm,12pt]{ucletter}

\usepackage{setspace}
\usepackage{graphicx}

\setlength{\parindent}{5ex}
\linespread{1.2}

%\name{Justin D. Yeakel}
\telephone{(209)~285-9571}
\email{jyeakel@ucmerced.edu}

\begin{document}
\begin{letter}{
    \\ Philosophical Transaction of the Royal Society B: Biological Sciences\\
    The Royal Society\\
    6-9 Carlton House Terrace\\
    London England SW1Y 5AG\\ \\
    \vspace{1mm}
    \centerline{\bf{Re: The eco-evolutionary impacts of collective straying on metapopulation robustness}} \\
}

\opening{To the Editorial Board at \emph{Philosophical Transactions},}

%Intro
Please find attached our manuscript entitled ``The eco-evolutionary impacts of collective straying on metapopulation robustness'' co-authored by Justin Yeakel, Jean Philippe Gibert, Peter Westley, and Jonathan Moore, which we would like to submit for publication in the Philosophical Transactions of the Royal Society B themed issue \emph{Collective movement in ecology: From emerging technologies to management implications}, under the section title `Linking collective movement with populations, ecology and evolution'.

A longstanding question in ecology and evolutionary biology concerns how selective gradients and dispersal influences the robustness and persistence of spatially connected metapopulations.
Of particular interest to this themed issue is the role of dispersal among populations that collectively navigate where such dispersal -- or straying for salmonid metapopulations -- may vary as a function of population density (Berdahl et al., \emph{Fish and Fisheries}, 2016).
Because populations interact between habitats that are distributed across a selective mosaic, changes in population size, and by extension the rate of straying, may be strongly influenced by the evolutionary dynamics of trait complexes that determine population-level fitness.

In the accompanying manuscript, we introduce an eco-evolutionary model of metapopulation dynamics that takes into account density dependent straying as well as local selection on a generalized trait complex determining the rate of recruitment.
With this framework, we investigate how increased rates of straying and trait heritability alter dynamic expectations of the metapopulation, and relate this to two measures of metapopulation robustness: the portfolio effect (PE) and the time required for a population to recover following an induced disturbance.
The qualitative results of our model are in accord with empirical observations, and we make a number of predictions that we believe have important implications for conservation.
We make several specific contributions:

\begin{itemize}
  \item Increased straying between two locally-adapted populations results in the emergence of alternative stable states: one \emph{dominant state} population that contains a larger proportion of biomass, and one \emph{subordinate state} population.
  The trait means of both populations are skewed towards that of the dominant population, and this has a large effect on the resultant dynamics.
  \item There is a nonlinear effect of straying rates (either constant or density dependent) on both the portfolio effect and the recovery time: portfolio effects are generally maximized and recovery rates are generally minimized with a low-intermediate rate of straying -- particularly in the alternative steady state regime -- and this promotes robustness.
  % \item We show that the portfolio effect and recovery time following a large disturbance, two independent measures of robustness that concern steady state and transient dynamics, respectively, are strongly correlated.
  \item Increased habitat heterogeneity tends to increase robustness when rates of straying are low, but decrease robustness when rates are straying become large. Importantly, metapopulation robustness is generally increased by density dependent straying, regardless of habitat heterogeneity.
  \item If highly dissimilar habitats are linked by a very low rate of straying (as might be expected if two sites are very distant from one another), we find that the extinction of the dominant population 
\end{itemize}

Our theory is analytic, relatively simple, and synthesizes a variety of fundamental ecological theories (energy equivalence hypothesis, resource competition theory, Cope’s rule, the fasting endurance hypothesis) and shows how many of these are the natural consequence of a simple dynamical framework that incorporates the energetic consequences of starvation and recovery combined with allometric timescales. 
We believe that our findings will be of general interest to ecologists, evolutionary biologists, and paleoecologists, such that consideration for publication in a broad interest journal such as \emph{Philosophical Transactions} is warranted.

We appreciate your consideration of our manuscript.

\vspace{5mm}

\singlespacing
\closing{Sincerely,\\
\fromsig{\includegraphics[scale=0.2]{signature.jpg}}\\
\fromname{
Justin D. Yeakel,\\
Jean Philippe Gibert,\\
Peter A. H. Westley,\\
Jonathan W. Moore}
}

\end{letter}
\end{document}

### 2017-05-31 manuscript/msv3.tex
reorganization of sections and minimization

@@
 %\title{Community assembly and dynamics by the deconstruction of species interactions}
-\title{The eco-evolutionary impacts of collective straying on metapopulation dynamics}
-\author{Justin D. Yeakel${}^{1,2,*}$, Jean Philippe Gibert${}^{1}$, Peter A. H. Westley${}^{3}$, \& Jonathan W. Moore${}^{4}$ \\
+\title{Eco-evolutionary dynamics and collective straying: implications for salmon metapopulation robustness}
+\author{Justin D. Yeakel${}^{1,2,*}$, Jean P. Gibert${}^{1}$, Peter A. H. Westley${}^{3}$, \& Jonathan W. Moore${}^{4}$ \\
 ${}^1$School of Natural Sciences, University of California, Merced, Merced CA, USA \\
@@
 
+\section*{Abstract}
+{\small
+We show that taking straying into account leads to alternative stable states in population densities and trait values, which has consequences for maladaptation, intraspecific trait variability, and long-term robustness of salmon metapopulations.
+}
 \section{Introduction}
@@
 Intraspecific diversity can increase the resilience and stability of species or metapopulations. 
-This diversity-stability linkage can occur when there are asynchronous population dynamics, where the changes in population size vary across a metapopulation. 
-This asynchrony will increase the potential for demographic rescue \cite{Earn:2000fm} and also decrease the variability of processes that integrate across the metapopulation \cite{Anonymous:2015gf}. 
-For example, different responses to climate variability within populations of a rare plant reduced abundance fluctuations \cite{Abbott:2017hl}. 
+This diversity-stability linkage can occur when there are asynchronous population dynamics, where the changes in population vary across a metapopulation. 
+This asynchrony will increase the potential for demographic rescue \cite{Brown:1977gk,Earn:2000fm} and also decrease the variability of processes that integrate across the metapopulation \cite{Anonymous:2015gf}.
+For example, different responses to climate variability within populations of a rare plant reduced fluctuations in abundance \cite{Abbott:2017hl}. 
 This statistical buffer has traditionally been quantified as the Portfolio Effect (PE), which is the ratio of the population CV to the CV of the aggregated metapopulation \cite{Thibaut:2012km}. 
+Strengthened portfolio effects are expected to increase the robustness of metapopulations to external disturbances, and by extension promote persistence \cite{Thibaut:2012km}.
+In contrast, homogenization of populations leading to greater synchronization and weakened PE may be a harbinger of metapopulation collapse and extinction.
 
-Movement of individuals across populations (i.e., straying or dispersal) can have a large influence on metapopulation dynamics  \cite{MilnerGulland:2011vm}. 
-First, some level of straying is necessary to enable the rescue effect, whereby dispersal among populations can rescue small populations from local extinction (REFS). 
-Second, high levels of straying may synchronize populations and actually increase the risk of extinction of the entire metapopulation \cite{Earn:2000fm}. 
-Third, straying will influence the evolutionary dynamics of the metapopulation, however evolutionary impacts are less well understood. 
-Although the straying of individuals into sites hosting other populations provides connections within the larger metapopulation, potentially promoting ecological and evolutionary rescue, it may also introduce maladapted individuals into habitats that are host to different environmental conditions, possibly lowering the mean fitness of the local (mixed) population \cite{Muhlfeld:2014hs}.
+It is clear that dispersal (i.e., movement with consequences for gene flow) can influence the evolutionary dynamics of the metapopulation, the interplay between selection, population dynamics, and persistence is not well understood.
+Movement of individuals among local populations (i.e. dispersal) can have a large influence on metapopulation persistence \cite{MilnerGulland:2011vm}. 
+Dispersal facilitates evolutionary rescue, whereby immigration of individuals with heritable adapative traits can rescue small populations from local extinction in the context of maladaptive environmental change \cite{Bell:2011ki,Carlson:2014is}.
+% First, some level of straying is necessary to enable the rescue effect, whereby dispersal among populations can rescue small populations from local extinction (REFS). 
+On the other hand, high rates of dispersal may synchronize populations and actually increase the risk of extinction of the entire metapopulation \cite{Earn:2000fm}. 
+Third, straying will influence the evolutionary dynamics of the metapopulation.
+% though how the interplay between selection and population dynamics might impact persistence is less well understood. 
+Although the dispersal of individuals into sites hosting other populations provides connections within the larger metapopulation, potentially promoting demographic and evolutionary rescue, it may also introduce maladapted individuals into habitats that are host to different environmental conditions, possibly lowering the mean fitness of the recipient population \cite{Muhlfeld:2014hs}.
+More broadly, dispersal can provide a mechanism by which phenotypes are sorted in space rather than time and facilitates the spread of maladaptive genes \cite{Lowe:2015ft}.
 Further, this straying may lead to genetic homogenization that erodes the asynchrony that underpins portfolio effects and metapopulation persistence. 
-Thus, the dual nature of straying as both promoter of connections among metapopulation demes and potential eroder of locally adapted gene complexes highlights the interplay between ecological dynamics of connected populations and the evolutionary dynamics of mixed trait distributions that respond to heterogeneous local conditions.
-
-% %The tradeoff
-% Population-level biodiversity is recognized to increase species persistence (REFS), and the resilience and sustainability of metapopulations (general REF).
-% For example, the long-term sustainability of the Bristol Bay sockeye salmon fishery is due in large part to intact habitat that in conjunction with fine-scale homing and reduced gene flow provides a template for the evolution of countless locally adapted populations \cite{Hilborn:2003gf,Schindler:2010he,Anonymous:2014ku,Satterthwaite:2015ge}.
-% In contrast, areas such as California's Central Valley where habitats have been lost, degraded, and homogenized stand in marked contrast.
-% Within species diversity (or lack thereof) can manifest itself by way of asynchronous population dynamics, where the drivers that give rise to changes in population size vary across a metapopulation, decreasing the potential for synchronization, and correspondingly increasing the potential for ecological rescue (REFS).
-% This statistical buffer against extinction has traditionally been quantified as the Portfolio Effect (PE), which is the ratio of the population CV to the CV of the aggregated metapopulation.
-% Weakened portfolio dynamics can also emerge from homogenization resulting from  reduced genetic variability among populations, where homogenous stocks are more likely to have similar life-history structures (REFS), be at greater risk disease-induced epidemics (REFS), and recruit sub-optimally in heterogeneous environments (REFS).
-% Of course, these two measures of diversity often occur concurrently: for instance, lower genetic diversity increases the likelihood that two populations respond similarly to the same stressors, and this promotes synchronized dynamics, thus lowering portfolio effects.
+Thus, straying can influence the resilience and robustness of metapopulations through both ecological and evolutionary processes.
+% Thus, the dual nature of straying as both promoter of connections among metapopulation demes and potential eroder of locally adapted gene complexes highlights the interplay between ecological dynamics of connected populations and the evolutionary dynamics of mixed trait distributions that respond to heterogeneous local conditions.
 
 %Movement through space (link back to straying)
-Migratory populations that return to a breeding ground or natal stream to reproduce are linked to each other by some proportion of the population that permanently disperse, or stray into the `wrong' site; we might say that there is at least one \emph{Kevin} in every school or flock (figure \ref{fig:xkcd}).
-There is growing appreciation that the rate of straying may be influenced by abiotic, biotic, and anthropogenic factors \cite{H:2013fs,Keefer:2014gg,Bett:2017ha}. 
-Recently the role of social interactions and collective navigation has been hypothesized [This issue].
-The rate at which individuals stray, $m$, may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \cite{Berdahl:2015kv,Berdahl:2016dx}.
-Regardless of the mechanism's governing straying, the effect that it has on the dynamics of individual populations and the metapopulation as a whole is a topic of considerable interest that has tangible conservation implications \cite{Brenner:2012gl,Johnson:2012fe,Fullerton:2011ii}.
-Whether, and to what extent, the ecological consequences of straying depend on the evolutionary dynamics that emerge from populations distributed across a selective gradient is unclear.
-How the assumed negative evolutionary effects of straying and subsequent gene flow is balanced by the positive effects of demographic rescue is the subject of this contribution.
+% Migratory populations that return to a breeding ground or natal stream to reproduce are linked to each other by some proportion of the population that permanently disperse, or stray into the `wrong' site.
+%; we might say that there is at least one \emph{Kevin} in every school or flock (figure \ref{fig:xkcd}).
+There is growing appreciation that a combination of abiotic, biotic, and anthropogenic factors can control the rate of dispersal among populations \cite{H:2013fs,Keefer:2014gg,Bett:2017ha}.
+Migratory populations that return to breeding sites for reproduction are linked to each other by some proportion of the population that permanently disperses into the `wrong' site. 
+Recently, the role of social interactions and collective navigation has been hypothesized [This issue].
+The rate at which individuals stray, $m$, may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \cite{Simons:2004jo,Berdahl:2015kv,Berdahl:2016dx}.
+The potential influence of collective dispersal on the dynamics of individual populations and the metapopulation as a whole is a topic of considerable interest that has tangible conservation implications \cite{Brenner:2012gl,Johnson:2012fe,Fullerton:2011ii}.
+% Whether and to what extent the ecological consequences of straying depend on the evolutionary dynamics that emerge from populations distributed across a selective gradient is unclear.
+% How the assumed negative evolutionary effects of straying and subsequent gene flow is balanced by the positive effects of demographic rescue is the subject of this contribution.
 
@@
 %% From geographic mosaic of coevolution to...
-That evolutionary forces of selection and gene flow play out heterogeneously across geographic mosaics is now a foundational concept in ecology and evolutionary biology \cite{Nuismer:1999ko,Thompson:2015hq,Nuismer:2000bb,Thompson:2005wf}.
-These mosaics are in part driven by environmental differences between habitats that alter the selective forces acting on different phenotypes \cite{Endler:1986tz}, and a principle underlying assumption is that there is gene flow such that individuals from different habitats mix over space \cite{Gomulkiewicz:2000bz,Thompson:2002hr,Nuismer:2003eh,Nuismer:2006be,Forde:2008kc,Guimaraes:2011hu,Gibert:2013hp}.
-Although the evolutionary outcomes of these spatial processes have been explored in depth (REFS), it is less well understood how selective mosaics and their consequent evolutionary forces impact population dynamics as they unfold \cite{Hendry:2016un}.
-How the ecological dynamics of phenotypically diverse metapopulations, wherein constituent populations are subject to different selective regimes, are altered affected by evolutionary forces is less well known.
-% and the subject of this paper.
+% That evolutionary forces of selection and gene flow play out heterogeneously across geographic mosaics is now a foundational concept in ecology and evolutionary biology \cite{Nuismer:1999ko,Thompson:2015hq,Nuismer:2000bb,Thompson:2005wf}.
+% These mosaics are in part driven by environmental differences between habitats that alter the selective forces acting on different phenotypes \cite{Endler:1986tz}, and a principle underlying assumption is that there is gene flow such that individuals from different habitats mix over space \cite{Gomulkiewicz:2000bz,Thompson:2002hr,Nuismer:2003eh,Nuismer:2006be,Forde:2008kc,Guimaraes:2011hu,Gibert:2013hp}.
+% Although the evolutionary outcomes of these spatial processes have been explored in depth (REFS), it is less well understood how selective mosaics and their consequent evolutionary forces impact population dynamics as they unfold \cite{Hendry:2016un}.
 
 %Motivation: straying
-The eco-evolutionary impacts of straying likely has important implications for conservation and management in key taxa such as in salmon.
+The eco-evolutionary impacts of dispersal likely has important implications for conservation and management in key taxa such as in migratory salmon.
+While anadromous salmonid fishes (genera \emph{Oncorhynchus} and \emph{Salmo}) are renown for returning to their natal spawning habitats with high accuracy and precision after years at sea \cite{Quinn:2011tf,Jonsson:2011kg,Keefer:2014gg}, there are generally some individuals that `stray' to non-natal sites to spawn \cite{Quinn:1993ge,Hendry:2004wf}.
 Salmon may operate as metapopulations, where populations are genetically distinct but linked by some level of straying \cite{Schtickzelle:2007wb,Anderson:2014cx}.
-Asynchrony among salmon populations also increases the stability of fisheries that integrate across this diversity \cite{Schindler:2010he,Anonymous:2016jc}.
-However, these portfolio effects may be eroded if asynchrony is influenced by genetic diversity and this genetic diversity is lost. 
-Braun et al. \cite{Braun:2016ib} found that genetic differences among populations of Chinook salmon was correlated with the degree of asynchrony in population dynamics. 
-Artificial propagation of salmon by hatcheries and their strays may erode local genetic diversity, asynchrony among populations, and portfolio effects \cite{Carlson:2008hl,Moore:2010gs,Anonymous:2014ku}.
+Although extensive work has been done to document the extent of straying from donor populations and into recipient populations \cite{Keefer:2014gg,Bett:2017ha}, only recently have the abiotic, biotic, and anthropogenic influences of `straying' behaviors been investigated systemically \cite{Keefer:2008bs,Westley:2015to,Bond:2016dz}.
+Straying among salmon may be influenced by environmental factors such as water temperature, human activities such as hatchery practices, and population density as predicted by the collective migration hypothesis \cite{Peterson:2014gy}.
+These strays can introduce new maladaptive genotypes into the recipient population.
+Further, straying and genetic homogenization could synchronize population dynamics and erode portfolio effects \cite{Moore:2010gs,Carlson:2011ce,Braun:2016ib}.
+Thus, there is opportunity and a need to consider the eco-evolutionary consequences of straying for metapopulations in species of conservation and management concern such as salmon.
+
+% Asynchrony among salmon populations also increases the stability of fisheries that integrate across this diversity \cite{Schindler:2010he,Anonymous:2016jc}.
+% However, these portfolio effects may be eroded if asynchrony is influenced by genetic diversity and this genetic diversity is lost. 
+% Braun et al. \cite{Braun:2016ib} found that genetic differences among populations of Chinook salmon was correlated with the degree of asynchrony in population dynamics. 
+% Artificial propagation of salmon by hatcheries and their strays may erode local genetic diversity, asynchrony among populations, and portfolio effects \cite{Carlson:2008hl,Moore:2010gs,Anonymous:2014ku}.
 %Coordinated mass migrations are one of the great wonders of the natural world, and the ability of individuals and groups to navigate across great distances have long fascinated naturalists \cite{MilnerGulland:2011vm}.
-While anadromous salmonid fishes (genera \emph{Oncorhynchus} and \emph{Salmo}) are reknown for returning to their natal spawning habitats with high accuracy and precision after years at sea \cite{Quinn:2011tf,Jonsson:2011kg,Keefer:2014gg}, there are generally some individuals that `stray' to non-natal sites to spawn \cite{Quinn:1993ge,Hendry:2004wf}.
-Although extensive work has been done to document the extent of straying from donor populations and into recipient populations \cite{Keefer:2014gg,Bett:2017ha}, only recently have the abiotic, biotic, and anthropogenic influences of `straying' behaviors been investigated systemically \cite{Keefer:2008bs,Westley:2015to,Bond:2016dz}.
-Most recently the role of social interactions and collective navigation has been hypothesized \cite{Berdahl:2015kv,Berdahl:2016dx}.
-These strays can introduce new maladaptive genotypes into the recipient population.
-Straying amon salmon may be influenced by environmental factors such as water temperature, human activities such as hatchery practices, and population density as predicted by the collective migration hypothesis \cite{Peterson:2014gy}.
-Thus, there is an opportunity to consider the eco-evolutionary consequences of straying for metapopulations in species of conservation and management concern such as salmon \cite{Carlson:2011ce}. 
-%Although the straying of individuals into sites hosting other populations provides connections within the larger metapopulation, potentially promoting ecological and evolutionary rescue, it also serves to introduce maladapted individuals into habitats that are host to different environmental conditions, and this may lower the mean fitness of the local (mixed) population \cite{Muhlfeld:2014hs}.
 
-%Here we show: pattern formation, dependence on heritability, large effects of small amounts of straying
+% Most recently the role of social interactions and collective navigation has been hypothesized \cite{Berdahl:2015kv,Berdahl:2016dx}.
+% These strays can introduce new maladaptive genotypes into the recipient population.
+% Straying amon salmon may be influenced by environmental factors such as water temperature, human activities such as hatchery practices, and population density as predicted by the collective migration hypothesis \cite{Peterson:2014gy}.
+% Thus, there is an opportunity to consider the eco-evolutionary consequences of straying for metapopulations in species of conservation and management concern such as salmon \cite{Carlson:2011ce}.
 
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.4\textwidth]{figs2/fig_xkcd.png}
-\caption{
-\emph{Migrating Geese}, a comic from Randall Monroe's xkcd (https://xkcd.com/1729/). 
-A flock of geese travel in the direction of a shared destination, the lone stray named \emph{Kevin}.
-In this case, the rate of straying is $m=0.05$, which is not an uncommon rate for migrating populations of salmon \cite{Satterthwaite:2015ge}. 
-Reprinted under the Creative Commons Attribution-NonCommercial 2.5 License.
-} \label{fig:xkcd}
-\end{figure}
+% 
+% \begin{figure}
+%   \captionsetup{justification=raggedright,
+% singlelinecheck=false
+% }
+% \centering
+% \includegraphics[width=0.4\textwidth]{figs2/fig_xkcd.png}
+% \caption{
+% \emph{Migrating Geese}, a comic from Randall Monroe's xkcd (https://xkcd.com/1729/). 
+% A flock of geese travel in the direction of a shared destination, the lone stray named \emph{Kevin}.
+% In this case, the rate of straying is $m=0.05$, which is not an uncommon rate for migrating populations of salmon \cite{Satterthwaite:2015ge}. 
+% Reprinted under the Creative Commons Attribution-NonCommercial 2.5 License.
+% } \label{fig:xkcd}
+% \end{figure}
 
 %Alternative stable state
-Here we seek to explore how the potentially detrimental evolutionary effects of straying (erosion of local adaptation) and subsequent gene flow is balanced by the positive effects of demographic rescue is the subject of this contribution.
+Here we seek to explore how collective density-dependent straying influences the stability and robustness of metapopulations through ecological and evolutionary processes.
+% the potentially detrimental evolutionary effects of straying (erosion of local adaptation) and subsequent gene flow is balanced by the positive effects of demographic rescue is the subject of this contribution.
 % Here we ask the overarching question: how does collective behavior mediated dispersal and gene flow interact to influence the robustness of locally adapted populations?
 To address this question we construct a minimal eco-evolutionary model of two populations occupying different sites that are linked by straying individuals, each of which with an associated trait distribution subject to natural selection determined by local conditions.
-[We show that taking straying into account leads to alternative stable states in population densities and trait values, which has consequences for maladaptation, intraspecific trait variability, and long-term sustainability of salmon metapopulations.]
-
-% An important and relatively novel component of this model is the inclusion of a parameter that defines the difference in local conditions that favor different trait optima, increasing values of which correlates to populations that mix across increasingly heterogeneous environments \cite[though see ]{Berdahl:2015kv}.
-% Although our proposed model is was constructed with the dynamics of salmon populations in mind, the framework is general and the conclusions are likely relevant to a diverse range of migratory organisms where locally adapted populations are linked by dispersal.
-% We first show that specific rates of straying and trait heritability can have large effects on the qualitative dynamics of populations over time, in many cases giving rise to alternative steady states where one site is pushed towards very low biomass.
-% The emergence of alternative steady states results in a nonlinear response of the portfolio effect as well as the time required for the metapopulation to recover after an induced disturbance, suggesting that metapopulation robustness can be quite sensitive to the combined influences of dispersal across a selective landscape.
-% 
-% %Extinction and return times
-% 
-% %Habitat heterogeneity
-% A second important finding of our minimal model reveals that systems with greater habitat heterogeneity (measured by an increased difference in the trait values that are optimal between sites) host an intermediate range of straying rates where the portfolio effect increases to a local maximum, signaling that under certain conditions moderate amounts of straying between populations can increase the likelihood of persistence despite the deleterious evolutionary effects overall.
-% However, if we suppose that the rate of straying between two sites is correlated with distance (as indicated by empirical evidence), and that the difference in trait optima increases with distance as would be the case if the optima were associated with alternative temperature regimes (especially if sites are distributed latitudinally rather than longitudinally), even a very small amount of straying can drastically reduce the portfolio effect.
-% Importantly, the qualitative nature of our results do not depend on whether the stray rate is density dependent or constant, suggesting a limited role of collective dispersal on the dynamics considered herein.
-% 
-% %Concluding paragraph of the intro
-% 
-%
+Specifically we compare (a) different rates of straying and (b) the influence of collective movement, across (c) increasing environmental heterogeneity, by assessing two measures of metapopulation robustness: the portfolio effect and the time required for a population(s) to recover following an induced disturbance.
+This model enables us to explore the multiple and potentially opposing pathways by which straying influences metapopulation robustness such as the potentially detrimental erosion of local adaptation vs.
+the positive effects of demographic and evolutionary rescue.
+% [We show that taking straying into account leads to alternative stable states in population densities and trait values, which has consequences for maladaptation, intraspecific trait variability, and long-term sustainability of salmon metapopulations.]
 
@@
 B) The steady state trait values measured as $\theta_i - x_i$, as a function of a constant stray rate $m$. 
+FB marks the fold bifurcation.
 } \label{fig:traj}
@@
 The parameter $\tau$ is the strength of selection, and controls the sensitivity of recruitment to changes in the mean trait value away from the optimum (the strength of selection increases with smaller values of $\tau$), which we set as $\tau=1$ here and throughout.
-Because straying individuals are emigrating from a population with a mean trait value farther from the local optimum, their rate of recruitment is diminished, and this matches what is observed in wild populations \cite{Peterson:2014gy}.
+Because straying individuals are emigrating from a population with a mean trait value farther from the local optimum, their rate of recruitment is diminished.
+Recent studies of wild sockeye have indeed found that straying individuals have lower life-time fitness than individuals that do not stray, although it unknown at what life-stage this selection occurs \cite{Peterson:2014gy}.
 % The compensatory effects are then determined by the exponential, following the Ricker stock-recruitment relationship.
@@
 (a) Total means $N_t$, (b) difference in means $\Delta N$, and (c) the portfolio effect PE as a function of heritability $h^2$ and a constant stray rate $m$. Light colors = high values.
-The black line shows the fold bifurcation separating a single steady state (left) from alternative steady states (right).
+The black line shows the fold bifurcation separating a single steady state (left) from alternative stable states (right).
 (d) The relationship between the time to recovery following a disturbance and the portfolio effect.
+The black line denotes the fold bifurcation.
 } \label{fig:PE}
@@
 First, we assume that the distribution resulting from the mix of remote and local individuals, following reproduction, is also normal with a mean value equal to that of the mean for the mixed-normal distribution.
-Second, we assume that changes in trait variance through time are minimal, such that $\sigma^2$ is assumed to be constant, and this is a common simplification in eco-evolutionary models of population dynamics \cite{Lande:1976ga,Schreiber:2011wx,Gibert:tz}.
+Thus, we assume that strays can successfully reproduce and introduce their genotypes into the recipient population, which is supported by observations in wild populations \cite{Jasper:2013cc}.
+Second, we assume that changes in trait variance through time are minimal, such that $\sigma^2$ is assumed to be constant, and this is a common simplification in eco-evolutionary models of population dynamics \cite{Lande:1976ga,Schreiber:2011wx,Gilbert:2014ee,Gibert:2015kc}.
 
@@
   \mu_i(t+1) &= w_i\mu_i(t) + (1-w_i)\mu_j(t) \\ \nonumber
-  &+ h^2\sigma^2\frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)|\theta_i] + (1-w_i)R_i[\mu_j(t)|\theta_i]  \right),
+  &+ h^2\sigma^2\frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)] + (1-w_i)R_i[\mu_j(t)]  \right),
 \end{align}
@@
 
-\noindent {\bf (b) Measuring metapopulation robustness}
+\noindent {\bf (b) Density-dependent straying}
+%Density dependent m
+\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is mounting evidence that at least in some species (including salmon) the straying rate is density-dependent with a signature of collective navigation \cite{Berdahl:2016dx,Bett:2017ha}.
+Specifically, the rate at which individuals stray has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tend to decrease the rate at which individuals err, reducing the overall proportion of a population that strays.
+According to Berdahl et al. \cite{Berdahl:2014bl,Berdahl:2016dx}, given the probability that an individual strays is $m_0$, the proportion of the local population $N_i(t)$ that strays is
+
+\begin{equation}
+  m(t) = m_0\left(1- \frac{N_i(t)}{C+N_i(t)}\right),
+  \label{eq:ddm}
+\end{equation}
+
+\noindent where $C$ is a half-saturation constant.
+We note that at the limit $C\rightarrow \infty$, the density-dependent straying rate becomes constant such that $m(t) \rightarrow m_0$, and this corresponds to the original formulation where $m=m_0$.
+A similar observation shows that when the population density is very high, $m(t) \rightarrow 0$, and when it is small, individuals operate without regard to collective behavior, meaning $m(t) \rightarrow m_0$.
+Thus, for realistic population densities, $m(t) < m_0$.\\
+
+\noindent {\bf (c) Habitat heterogeneity}
+
+%Linking m/m0 and thetadiff
+The straying rate is largely influenced by the spatial distance between the donor and recipient population.
+The greater the distance between two populations, the lower the expected rate of straying \cite{H:2013fs,Westley:2015to}.
+We account for this interdependence in our model by assuming that $m$ (if the stray rate is constant) or $m_0$ (if the stray rate is density-dependent) is a function of the difference between optimal trait values between sites $\theta_i-\theta_j$, which can be assumed to be large if the remote site $j$ is a great distance away from the local site $i$.
+If sites $i$ and $j$ are very close, the stray rate is maximized at $m_{\rm max} = 0.5$, assuming both sites are equally attractive to the respective populations.
+Thus, we can integrate these two variables by setting $m,m_0 = (1/m_{\rm max} + \epsilon (\theta_i-\theta_j))^{-1}$, where $\epsilon$ sets the sensitivity of a declining $m$ to increasing distance (greater values of $\theta_i-\theta_j$).\\
+
+\noindent {\bf (d) Measuring metapopulation robustness}
 \noindent We evaluated metapopulation robustness by measuring the average-CV portfolio effect (PE) \cite{Anderson:2014cx,Schindler:2015gf} as well as the time required for the system to return to a steady state following an induced disturbance to one or both of the populations \cite{Ovaskainen:2002il}.
@@
 As the CV of $N_T$ decreases relative to that of the constituent populations, $\langle{\rm PE}\rangle > 1$, and the metapopulation is presumed to become more stable.
-% {\color{red} A similar measure of metapopulation robustness can be evaluated by measuring the synchronization $\phi$ between populations, where 
-% \begin{equation}
-% \label{eq_sync}
-% \phi = \frac{{\rm VAR}(N_T)}{\left(\sum_{i=1}^N\sqrt{{\rm VAR}(N_i)}\right)^2}.
-% \end{equation}
-% By rearranging and substituting eq. (\ref{eq_sync}) into eq. (\ref{eq:pe}), we can observe that $\langle{\rm PE}\rangle \propto \sqrt{\phi}^{-1}$.
-% Thus, the dynamical diversity of the metapopulation offers a mirror to the portfolio effect, where perfect synchrony ($\phi = 1$) also means $\langle{\rm PE}\rangle = 1$.}
-%Moreover, if and only if $\phi \neq 1$, it is expected that as the diversity of the aggregate increases, so should the portfolio effect, and this central thesis is thought to underlie the stability of populations, species, communities and even financial markets.
 Portfolio effects greater than unity corresponds to less synchronization  \cite{Loreau:2008ju,Anderson:2014cx,Yeakel:2013vz} and thus a greater potential for demographic rescue among populations, buffering the system as a whole against extinction.
@@
 Numerically estimating the time that it takes for a perturbed system to relax also permits a more detailed perspective of metapopulation fragility.
-For example, if populations settle to alternative steady states (alternative steady states in our model requiring one population to be high-density and one low-density), comparing recovery times after a disturbance applied to the high, low, and/or both populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery.
+For example, if populations settle to alternative stable states (alternative stable states in our model requiring one population to be high-density and one low-density), comparing recovery times after a disturbance applied to the high, low, and/or both populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery.
 We measured the time required for the system $N_T$ to recover to its steady state following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} are equivalent if the system is in the single steady state regime); (\emph{iii}) near-collapse of both populations where just 1.0\% of each survives.
@@
 
-\noindent {\bf (c) The effects of density and distance on the rate of straying}
-%Density dependent m
-\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is mounting evidence that at least in some species (including straying) the straying rate is density dependent with a signature of collective navigation \cite{Berdahl:2016dx,Bett:2017ha}.
-Specifically, the rate at which individuals stray has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tend to decrease the rate at which individuals err, reducing the overall proportion of a population that strays.
-According to Berdahl et al. \cite{Berdahl:2014bl,Berdahl:2016dx}, given the probability that an individual strays is $m_0$, the proportion of the local population $N_i(t)$ that strays is
-
-\begin{equation}
-  m(t) = m_0\left(1- \frac{N_i(t)}{C+N_i(t)}\right),
-  \label{eq:ddm}
-\end{equation}
-
-\noindent where $C$ is a half-saturation constant.
-We note that at the limit $C\rightarrow \infty$, the density dependent straying rate becomes constant such that $m(t) \rightarrow m_0$, and this corresponds to the original formulation where $m=m_0$.
-A similar observation shows that when the population density is very high, $m(t) \rightarrow 0$, and when it is small, individuals operate without regard to collective behavior, meaning $m(t) \rightarrow m_0$.
-Thus, for realistic population densities, $m(t) < m_0$.
-
-%Linking m/m0 and thetadiff
-The straying rate is largely influenced by the distance between the donor and recipient population.
-The greater the distance between two populations, the lower the expected rate of straying \cite{H:2013fs,Westley:2015to}.
-We account for this interdependence in our model by assuming that $m$ (if the stray rate is constant) or $m_0$ (if the stray rate is density dependent) is a function of the difference between optimal trait values between sites $\theta_i-\theta_j$, which can be assumed to be large if the remote site $j$ is a great distance away from the local site $i$.
-If sites $i$ and $j$ are very close, the stray rate is maximized at $m_{\rm max} = 0.5$, assuming both sites are equally attractive to the respective populations.
-Thus, we can integrate these two variables by setting $m,m_0 = (1/m_{\rm max} + \epsilon (\theta_i-\theta_j))^{-1}$, where $\epsilon$ sets the sensitivity of a declining $m$ to increasing distance (greater values of $\theta_i-\theta_j$).
-
 \section{Results}
-
-%The effects of trait evolution :: trait heritability, trait variance, and habitat heterogeneity
-%% Bifurcation to alternative stable state occurs at higher stray rates with increased trait heritability
-%% There is a peak in the portfolio effect directly before this bifurcation is crossed (early warning signal), and then a decline in PE with high m
-%% When trait heritability is high, there is a greater sensitivity of total biomass to increasing stray rates, though a LOWER sensitivity of the difference between alternative steady state values for increasing stray rates (and vice versa)
-%% For low heritability (h = 0.01 - 0.2 or so), moderate and insensitive changes in total biomass, but large differences between the values of alternative steady states leads to a local maximum for PE at low-moderate stray rates
 
@@
 \caption{
-(a) Median portfolio effect as a function of a constant stray rate $m$ (solid line) and density dependent stray rate (point pairs) given heritability is $h^2 < 0.5$ and $\Delta\theta=5$.
-Point pairs connected by a horizontal line represent the PE as a function of density dependent straying rates, evaluated for both low- and high-density populations at equilibrium.
+(a) Median portfolio effect as a function of a constant stray rate $m$ (solid line) and density-dependent stray rate (point pairs) given heritability is $h^2 < 0.5$ and $\Delta\theta=5$.
+Point pairs connected by a horizontal line represent the PE as a function of density-dependent straying rates, evaluated for both low- and high-density populations at equilibrium.
 The lower straying rate of a pair is for the larger population; the higher straying rate is for the smaller population.
-(b) Median portfolio effects for habitats with increasing heterogeneity as measured by the difference in regional trait optima $\Delta \theta$ for both constant and density dependent stray rates as shown in (a).
+(b) Median portfolio effects for habitats with increasing heterogeneity as measured by the difference in regional trait optima $\Delta \theta$ for both constant and density-dependent stray rates as shown in (a).
+FB marks the fold bifurcation.
 } \label{fig:thetaPE}
@@
 
-\noindent{\bf (a) Nonlinear effects of straying on the portfolio effect and recovery time} \\
+\noindent{\bf (a) Nonlinear effects of straying on metapopulation robustness} \\
 \noindent Regardless of density dependence, straying lowers steady state densities for both populations by (\emph{i}) the donor population losing locally-adapted individuals to the recipient population and (\emph{ii}) the introduction of maladapted individuals to the recipient population from the donor population (Fig.
@@
 This prediction is in accordance with observations from natural populations \cite{Bett:2017ha}. %, regardless of trait heritability and variance or habitat heterogeneity
-The decline in steady state densities is not gradual: as straying increases, the system crosses a fold bifurcation whereby the single steady state for the metapopulation bifurcates into two alternative steady states: one at high biomass, and one at low biomass density (figure \ref{fig:traj}a, \ref{fig:PE}a).
+The decline in steady state densities is not gradual: as straying increases, the system crosses a fold bifurcation whereby the single steady state for the metapopulation bifurcates into two alternative stable states: one at high biomass, and one at low biomass density (figure \ref{fig:traj}a, \ref{fig:PE}a).
 Mean trait values for both populations bifurcate similarly (figure \ref{fig:traj}b), depending on which population attains a low- vs. high-density. 
@@
 \caption{
-Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.2$.
-If $m$ is density dependent, in the alternative steady state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
+Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.2$.
+If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
+FB marks the fold bifurcation.
 } \label{fig:relax}
@@
 This variance increase is the product of a dynamical process known as \emph{critical slowing down} that occurs near fold bifurcations \cite{Scheffer:2009gg}, a phenomenon that some have suggested may serve as an early warning indicator for approaching phase transitions \cite{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br}.
-For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}a-c), where alternative steady states occur, the portfolio effect declines steadily as the CV of $N_T$ increases.
+For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}a-c), where alternative stable states occur, the portfolio effect declines steadily as the CV of $N_T$ increases.
 The decline over $m$ is more gradual if trait heritability is low, and steeper if trait heritability is high (figure \ref{fig:PE}c).
-
-If we assume that the rate of straying is density dependent, the probability that an individual strays $m_0$ determines the rate of straying within the population, such that $m(t)$ becomes lower as $N(t)$ increases, likely due to the effects of collective decision-making \cite{Berdahl:2016dx} (Eq.
-\ref{eq:ddm}).
-%Because our analyses concern steady state conditions, the inclusion of density-dependent straying does not have a large impact on the qualitative nature of our findings (figure \ref{fig:ddm}).
-Density dependence alters the straying rate at steady state population densities because $0 < m^* < m_0$, and this serves to rescale both the strength of the PE as well as the recovery time, but does not change the qualitative nature of our findings.
-In the alternative stable state regime, because each population exists at different steady state densities, there are likewise two alternative straying rates $(m_i^*,m_j^*)$: the higher straying rate is associated with the low-density population, and the lower straying rate is associated with the high-density population.
-% We assessed metapopulation robustness across a range of $(m_i^*,m_j^*)$ values by varying $m_0$, which is positively and linearly related to $(m_i^*,m_j^*)$.
-%Density dependent straying
-% In the alternative stable state regime, because the two populations exist in a dominant and subordinate state, there are also two alternative straying rates $(m_i^*,m_j^*)$: the higher of which associated with the subordinate, low-density population, and the lower of which is associated with the dominant, high-density population.
-We assessed metapopulation robustness across a range of $(m_i^*,m_j^*)$ values by varying the probability that an individual strays $m_0$, which is positively and linearly related to $(m_i^*,m_j^*)$.
-We find that the portfolio effects generated in systems with density dependent straying are qualitatively similar to systems with constant straying, however there are some important quantitative differences.
-First, the PE associated with the high-density (low $m^*$) population is the same as that for a system with a constant $m$ (figure \ref{fig:thetaPE}a).
-% As such, the difference in $m^*$ between subordinate and dominant populations appears to matter less than the lower straying rate generated by the dominant population.
-As $m^*$ increases, we observe an increase in the PE than for systems with constant $m$.
 
@@
 Because the mean trait values of both populations are skewed towards those of the dominant population, when the subordinate population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into the recovering population extends the length of time required for it to return to its steady state (figure \ref{fig:relaxtraj_ldlh}).
-% When the subordinate state population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into it extends the length of time required for it to return to its steady state (figure \ref{fig:relaxtraj_ldlh}).
-%At low rates of straying, 
-When both populations are both dramatically reduced, recovery time is generally fastest at lower levels of straying, while near the onset of the fold bifurcation, recovery time increases explosively and this is -- as the name implies -- characteristic of \emph{slowing} dynamics that occur near critical transitions \cite{Scheffer:2009gg,Kuehn:2010p2591}.
+When both populations are both dramatically reduced, recovery time is generally fastest at lower levels of straying, while near the onset of the fold bifurcation, recovery time increases explosively and this is -- as the name implies -- characteristic of \emph{slowing} dynamics that occur near critical transitions \cite{Scheffer:2009gg,Kuehn:2010p2591}.\\
 
+\noindent{\bf (b) The effects of collective migration and density dependent straying} \\
+%Portfolio effects
+If we assume that the rate of straying is density-dependent, the probability that an individual strays $m_0$ determines the rate of straying within the population, such that $m(t)$ becomes lower as $N(t)$ increases, likely due to the effects of collective decision-making \cite{Berdahl:2016dx} (Eq.
+\ref{eq:ddm}).
+%Because our analyses concern steady state conditions, the inclusion of density-dependent straying does not have a large impact on the qualitative nature of our findings (figure \ref{fig:ddm}).
+Density dependence alters the straying rate at steady state population densities because $0 < m^* < m_0$, and this serves to rescale both the strength of the PE as well as the recovery time, but does not change the qualitative nature of our findings.
+In the alternative stable state regime, because each population exists at different steady state densities, there are likewise two alternative straying rates $(m_i^*,m_j^*)$: the higher straying rate is associated with the low-density population, and the lower straying rate is associated with the high-density population.
+We assessed metapopulation robustness across a range of $(m_i^*,m_j^*)$ values by varying the probability that an individual strays $m_0$, which is positively and linearly related to $(m_i^*,m_j^*)$.
+We find that the portfolio effects generated in systems with density-dependent straying are qualitatively similar to systems with constant straying, however there are some important quantitative differences.
+First, the PE associated with the high-density (low $m^*$) population is the same as that for a system with a constant $m$ (figure \ref{fig:thetaPE}a).
+% As such, the difference in $m^*$ between subordinate and dominant populations appears to matter less than the lower straying rate generated by the dominant population.
+As $m^*$ increases, we observe an increase in the PE than for systems with constant $m$.
+
+%Recovery times
 Density-dependent straying alters these recovery times (figure \ref{fig:relax}b). 
@@
 %Recovery times
-%The recovery times for systems with constant and density dependent straying both show similar trends for a given $m$ and $m^*$ (figure \ref{fig:relax}c,d), however there are some distinct quantitative differences.
+%The recovery times for systems with constant and density-dependent straying both show similar trends for a given $m$ and $m^*$ (figure \ref{fig:relax}c,d), however there are some distinct quantitative differences.
 At low straying rates, near-collapse of both populations resulted in longer than expected recovery times, whereas in the alternative stable state regime (higher $m^*$), the recovery times for different disturbance types were very similar to systems with a constant $m$ (figure \ref{fig:relax}b; note difference in x-axis scales).
 As trait heritability increased, the metapopulation always recovered more quickly if the small population was lost (figure \ref{fig:relax_highh}).
-% There was greater similarity in the recovery times for constant and density dependent $m$ systems when trait heritability was high (figure \ref{fig:relax_highh}).
-The lower recovery time for systems with increased $m^*$ mirrors an elevated PE with higher density dependent straying rates (figure \ref{fig:PEtheta}).
+% There was greater similarity in the recovery times for constant and density-dependent $m$ systems when trait heritability was high (figure \ref{fig:relax_highh}).
+The lower recovery time for systems with increased $m^*$ mirrors an elevated PE with higher density-dependent straying rates (figure \ref{fig:PEtheta}).
 % This higher PE at increased $m^*$ mirrors recovery times that are also lower than those observed for systems with constant $m$ (figure \ref{fig:relax}).
 In tandem, analysis of both PE and recovery time suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our minimal model, it does appear to promote robustness, particularly when the aggregate biomass is low and straying is correspondingly high.
-
-% At low rates of straying both populations are subject to the same steady state, such that extinction of either population results in the same recovery time, and when trait heritability is low, the time required to recover following the extinction of either population is generally less than or equal to the recovery time after near-collapse of both (figure \ref{fig:relax}a).
-% In the alternative stable state regime where straying rates are higher, extinction of the smaller population results in the shortest recovery time, whereas as $m$ increases, extinction of the larger population results in the shortest recovery time.
-% Because the mean trait values of both populations are skewed towards those of the high-density site, when the low-density population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into the recovering extends the length of time required for it to return to its steady state (figure \ref{fig:relaxtraj_ldlh}).
-% Alternatively, when the high-density population collapses under high straying rates, the low-density population suffers loss as well due to the absence of incoming strays such that both populations must recover (figure \ref{fig:relaxtraj_hdlh}).
-% 
-% Increased trait heritability flips this relationship by increasing the response of trait evolution to changes in demography (figure \ref{fig:relax}b).
-% When straying rates are low, metapopulations that suffer near-collapse of both populations recover more quickly than systems suffering extinctions of either population.
-% When straying rates are higher such that the system is in the alternative stable state regime, the metapopulation always recovers more quickly if the small population is lost (figure \ref{fig:relaxtraj_ldhh}), whereas the time to recovery is maximized if both populations suffer near-collapse.
-% When the high-density population goes extinct, the stronger response of selection permits the surviving low-density population to grow quickly and fill the high-density state (figure \ref{fig:relaxtraj_hdhh}).
-% Near the onset of the fold bifurcation, recovery times increase explosively regardless of straying rate, heritability, or disturbance type.
-% %For straying rates in the alternative steady state regime, extinction of the smaller population consistently results in the shortest time to recovery.
 
@@
 Phenotypic diversity ($\Delta \mu^*$) evaluated at the steady state as a function of straying rate $m$ and trait heritability $h^2$.
-The descrete jump occurs as the system crosses the fold bifurcation; lower phenoytic diversity emerges with higher straying rates and in the alternative steady state regime.
+The descrete jump occurs as the system crosses the fold bifurcation; lower phenoytic diversity emerges with higher straying rates and in the alternative stable state regime.
 } \label{fig:traitdiff}
@@
 
-%% Increased habitat heterogeneity
-  %%% 1) exaggerates lowering of mean biomass with increasing stray rates
-  %%% 2) increases the likelihood of alternative stable sites & increases the differences in alternative stable state values with increasing stray rates
-  %%% 3) decreases the PE (!!)
-
-\noindent{\bf (b) The influence of habitat heterogeneity on metapopulation robustness}\\ 
+\noindent{\bf (c) The role of habitat heterogeneity and changing selective landscapes}\\ 
 \noindent Increasing differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater regional differences in the conditions that favor alternative trait complexes, which we interpret here as greater habitat heterogeneity (REFS).
@@
 %This means that alternative stable states can emerge with relatively little dispersal between sites (figure \ref{fig:thetaPE}), however the increased PE (and shorter recovery times).
-\\
 
-% When m and thetadiff are linked
-%%Constant m
-  %%% Low stray rate (high theta1-theta2) leads to alternative stable states and relatively low PE ~ sharp transition
-  %%% High stray rate (low theta1-theta2) increases the steady state (and no A.S.S.)... 
-  %%% This suggests that space naturally buffers against the negative effects of mixing dissimilar populations, however it is NOT SMOOTH
-  %%% This is particularly true for low heritability. High heritability eliminates the A.S.S. region
-  %%% (!!!) Even low levels of straying from distant populations can result in highly divergent alternative steady states! THIS IS A COOL RESULT
+If we assume that environmental heterogeneity increases with distance between sites, or that individuals may have behaviors that minimize their chances of straying into very different habitats, the rate of straying will decline with habitat heterogeneity.
+Accordingly, $\Delta\theta$ will decrease as $m$ increases, such that there are low rates of straying between dissimilar populations and high rates of straying between similar populations.
+We find that under these conditions alternative stable states now appear for very low rates of straying $0 < m \leq 0.43$.
+As the straying rate increases and $\Delta\theta$ decreases, a single stable state emerges as the fold bifurcation is crossed, which is opposite the pattern observed when these parameters are independent.
+There are two notable dynamics that emerge following extinction of the dominant population at low rates of straying between dissimilar (high $\Delta\theta$) sites:
+(\emph{i}) above a threshold $m$ value, the dominant population recovers quickly enough that the evolving subordinate phenotype is overwhelmed by incoming strays, and it shifts back to its pre-disturbance (subordinate) state;
+(\emph{ii}) below a threshold $m$ value, there is enough time and isolation for the subordinate trait mean to shift towards its local optimum, and away from that of the recovering dominant population, such that the dominant population becomes subordinate, and the subordinate population becomes dominant (figure \ref{fig:inertia}).
+This threshold value of $m$, below which this behavior occurs, is marked by the asterisk in figure \ref{fig:mtheta}, and holds for both constant and density-dependent straying (figure \ref{fig:mthetamvm}).
 
-\noindent{\bf (c) Distance dependent straying and habitat heterogeneity}\\ 
-\noindent We have so far treated $\Delta\theta$ and $m$ as independent parameters, however we may also assume that if environmental heterogeneity increases with distance between sites, the rate of straying may be expected to decline with habitat heterogeneity, given that individuals will stray less into distant habitats.
-%-- in particular North-South difference if trait optimality is largely temperature-dependent --
-Alternatively, individuals may be less likely to stray into very different habitats due to the infleunce of environmental cues on individual decision-making (REFS?).
-If we incorporate this interdependence of $m$ and $\Delta\theta$, low rates of straying correspond to mixing dissimilar (distant) populations, and high rates of straying would correspond to mixing similar (nearby) populations.
-
-We find that alternative stable states now appear for very low rates of straying $0 < m \leq 0.43$, corresponding to greater habitat heterogeneity $\Delta\theta$.
-As the straying rate increases, a single stable state emerges as the fold bifurcation is crossed.
-As straying increases (now the single steady state regime), the time required for recovery following extinction of either the population declines (to the right of figure \ref{fig:mtheta}).
-However, as straying decreases the system crosses the fold bifurcation, entering the alternative steady state regime (to the left in figure \ref{fig:mtheta}).
-At low rates of straying, there is a small amount of mixing between dissimilar (distant) populations.
-Here we find that if the subordinate population is wiped out, the time to recovery is minimized, whereas if the dominant population is wiped out, the time to recovery is maximized; near-collapse of both results in intermediate recovery times.
-
-When straying rates are low and the difference in trait optima are correspondingly high, we might presume that extinction of the dominant population would result in relatively faster recovery.
-This would be a logical assumption because the mean trait value of the subordinate population is skewed towards that of the dominant (larger) population, such that recruitment at the disturbed site would be high.
-Our results show, however, that this assumption is wrong: although the mean trait value of the subordinate population \emph{is} skewed towars the optimum of the dominant population, when the latter is wiped out and the rate of straying is low, there is enough time and isolation for the subordinate trait mean to shift towards its own optimum, and away from that of the recovering dominant population (figure \ref{fig:inertia}).
-This \emph{selective inertia} occurs until the dominant population grows large enough that the evolving subordinate phenotype is overwhelmed by incoming strays, shifting it back towards that of the now-recovered dominant population.
-Importantly, this effect can lead to much longer recovery times for the dominant population, as well as a switch in which population is dominant/subordinate (a switch in who occupies the alternative states), when the straying rate is very low (and $\Delta\theta$ is large).
-The threshold value of $m$, below which this behavior occurs, is marked by the asterisk in figure \ref{fig:mtheta}. % at m = 0.034
-These results hold for both constant and density dependenty straying rates (figure \ref{fig:mthetamvm})
+% At low rates of straying, there is little mixing between dissimilar populations, and this leads to the formation of alternative stable states.
+% Here we find that if the subordinate population is wiped out, the time to recovery is minimized, whereas if the dominant population is wiped out, the time to recovery is maximized; near-collapse of both results in intermediate recovery times.
+% When straying rates are low and the difference in trait optima are correspondingly high, we might assume that extinction of the dominant population would result in relatively faster recovery.
+% This would be a logical assumption because the mean trait value of the subordinate population would be skewed towards that of the dominant (larger) population, such that recruitment at the disturbed site should be high and the population should recovery more quickly.
+% Our results show, however, that \emph{this assumption is wrong:} although the mean trait value of the subordinate population \emph{is} skewed towards the optimum of the dominant population, when the latter is wiped out and the rate of straying is low, there is enough time and isolation for the subordinate trait mean to shift towards its own optimum, and away from that of the recovering dominant population (figure \ref{fig:inertia}).
+% This \emph{selective inertia} occurs until the dominant population grows large enough that the evolving subordinate phenotype is overwhelmed by incoming strays, shifting it back to its pre-disturbance (subordinate) state.
+% This effect can lead to much longer recovery times for the dominant population.
+% Importantly, if the rate of straying is below a threshold value, it can also result in a state-switch, wherein increased isolation permits the subordinate population to \emph{escape} the selective pull of the dominant population, leading to a switching in which population is in the dominant/subordinate state.
 
@@
   Distance dependent recovery times for three disturbance types. When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases.
+  The `$*$' marks the value of $(m,\Delta\theta)$ below which there is a switch in subordinate/dominant states following extinction of the dominant population.
+  FB marks the fold bifurcation.
   } \label{fig:mtheta}
@@
 
-We have shown that dispersal between populations, coupled with localized selection against donor phenotypes, has a large and nonlinear impact on dynamic properties that directly impacts metapopulation robustness, and by extension persistence.
-Although there are various ways to measure `robustness' (REF:TREE), we employ two measures: 
-1) the average-CV portfolio effect, a statistical metric commonly used to assess the fragility of populations; 
-2) the recovery time, defined here as the time required for the aggregate metapopulation biomass $N_T$ to return to its steady state after different types of induced disturbances, which is mechanistically linked to the likelihood of persistence (REF).
-In our minimal model of dispersal and natural selection between two populations, we show that these phenomenological and mechanistic descriptors of metapopulation robustness are tightly coupled (figure \ref{fig:PE}d).
+We have shown that density-dependent straying between populations consistent with collective navigation, coupled with localized selection against donor phenotypes, has a large and nonlinear impact on dynamic properties directly impacting the robustness of metapopulations against external disturbances.
+Specifically, we observed robustness in terms of: 
+1) the average-CV portfolio effect \cite{Anderson:2013gb,Anonymous:2015gf}, a statistical metric commonly used to assess the buffering capacity of metapopulations, and
+2) the recovery time, defined here as the time required for the aggregate metapopulation biomass $N_T$ to return to its steady state following an induced disturbance, which is mechanistically linked to persistence \cite{Ovaskainen:2002il}.
+In our minimal eco-evolutionary model of dispersal and natural selection between two populations, we show that these statistical and mechanistic descriptors of metapopulation dynamics and robustness are tightly coupled (figure \ref{fig:PE}d), which is not uncommon among different metrics of stability \cite{Donohue:2013iu}.
+Taken as a whole, we interpret our results to suggest an important role of density-dependent straying in the colonization and recovery dynamics for populations in shifting river mosaics, while also underscoring the risk of straying by individuals with maladaptive traits to reduce the productivity of locally adapted stock complexes.
 
-%Alternative steady states & critical slowing
-The immigration of dispersing individuals with trait values far from the local optimum generally lowers the combined steady state biomass $N_T$ and increases the likelihood that the system gives way to alternative steady states, pushing one of the populations close to extinction.
-The formation of alternative stable states in our system is a relatively common example of spatial pattern formation, and occurs above a threshold straying rate.
-Pattern formation can occur as a consequence of myriad ecological processes, including habitat selection (REFS), aggregation of individuals (REFS), local environmental conditions (REFS), or interspecific interactions like competition (REFS).
-It is also well known to occur with the onset of spatially or diffusion-induced instabilities (Turing instabilities, REFS). 
-Here, we describe an interesting alternative mechanism through which pattern formation in local densities can occur, when the onset of movement mediated maladaptation leads to local differences in reproductive rates that yield differences in densities that would, in the absence of straying, be the same.
+% 
+% %The bifurcation and critical slowing down
+% [This paragraph is interesting to me but could be cut] Although the portfolio effect is negatively correlated with recovery time, there is one exception: the point of transition from a single steady state to alternative stable states, which occurs at the fold bifurcation (REF).
+% At this point, the portfolio effect is maximized, but so is the recovery time (figures \ref{fig:PE}, \ref{fig:relax}).
+% This suggests that while the sharply increasing variance of the individual populations is dampened by aggregation, a large disturbance will have a much greater adverse effect due to exponentially longer recovery times.
+% Whether the change in variance, also known as critical slowing down, as the alternative stable state regime is approached could be used as an early warning signal of an oncoming phase transition is a hotly debated topic (REFS).
+% The detectability of such changes in dynamical behavior among salmonids appears to be idiosyncratic across species, though the difficulty in measuring critical slowing down may - ironically - be masked by large portfolio effects \cite{Krkosek:2014ch}.
+%
+
+%alternative stable states & critical slowing
+A salient finding from our results was that straying can push metapopulations into alternative stable states, pushing one of the populations to a high density population (the \emph{dominant state}), and one to a low density population (the \emph{subordinate population}).
+This asymmetry in population densities also leads to an asymmetry in the mean trait values, both of which are skewed towards the local optimum of the dominant population. 
+The formation of alternative stable states in our system is a relatively common example of spatial pattern formation, which only occurs above a threshold straying rate, the value of which depends on trait heritability and the difference in trait optima between sites.
+Pattern formation can occur as a consequence of myriad ecological processes, including habitat selection, aggregation, local environmental conditions, and/or interspecific interactions such as competition or predation, and is inexorably linked to scale \cite{Levin:1992p1472}.
+Additionally, pattern formation is also well known to occur with the onset of spatially or diffusion-induced instabilities, or Turing instabilities \cite{Meinhardt:2012hw}. 
+Here, we describe an additional mechanism through which pattern formation in local densities can occur, when the onset of collective movement-mediated maladaptation leads to local differences in reproductive rates resulting in alternative stable states across sites.
 %This mechanism of pattern formation through the simultaneous interplay between ecological and evolutionary dynamics may open up new areas of research.
+While metapopulation approaches have widely recognized the importance of source and sink populations \cite{Hanski:1999vu} and have observed formation of alternative states \cite{Boughton:1999fa}, here we show that even similar populations can evolve into divergent states where one population is large and the other is smaller and maladapted.
 
-To what extent the existence of alternative stable states promotes or diminishes the system as a whole is a matter of perspective.
-On the one hand, one population is pushed to a low-density state (figure \ref{fig:traj}) with a lower total biomass of the aggregate (figure ]ref{fig:PE}a), suggesting that the system is more fragile.
-On the other hand, the presence of \emph{just enough} straying to cause formation of alternative stable states both increases the portfolio effect (figure \ref{fig:PE}c) and reduces the rate of return -- particularly if straying is density dependent (figure \ref{fig:relax}c,d) as might be common in systems with collective navigation (REF).
-Importantly, the asymmetry in population densities also leads an asymmetry in the mean trait values between sites: because the high-density population contains the majority of biomass, the proportion of the low-density population that is composed of incoming strays is generally large, and this leads to mean trait values in both sites skewed towards that of the high-density site.
-This has important ramifications for metapopulation robustness, as measured by both the portfolio effect as well as recovery time.
+% 
+% To what extent the existence of alternative stable states promotes or diminishes the system as a whole is a matter of perspective, though with clear conservation implications.
+% At face value, the subordinate population is pushed closer to extinction (figure \ref{fig:traj}), suggesting that the system may be more fragile.
+% On the other hand, the presence of \emph{just enough} straying to cause formation of alternative stable states both increases the portfolio effect (figure \ref{fig:PE}c) and reduces the rate of return -- particularly if straying is density-dependent (figures \ref{PEtheta}a, \ref{fig:relax}b).
+% This asymmetry in population densities also leads an asymmetry in the mean trait values, skewed towards the local optimum of the dominant population, one, and this has important -- and often non-intuitive -- ramifications for metapopulation robustness.
+%, as measured by both the portfolio effect as well as recovery time.
 
-%The bifurcation and critical slowing down
-Although the portfolio effect is negatively correlated with recovery time, there is one exception: the point of transition from a single steady state to alternative steady states, which occurs at the fold bifurcation (REF).
-At this point, the portfolio effect is maximized, but so is the recovery time (figures \ref{fig:PE}, \ref{fig:relax}).
-This suggests that while the sharply increasing variance of the individual populations is dampened by aggregation, a large disturbance will have a much greater adverse effect due to exponentially longer recovery times.
-Whether the change in variance, also known as critical slowing down, as the alternative steady state regime is approached could be used as an early warning signal of an oncoming phase transition is a hotly debated topic (REFS).
-The detectability of such changes in dynamical behavior among salmonids appears to be idiosyncratic across species, though the difficulty in measuring critical slowing down may - ironically - be masked by large portfolio effects \cite{Krkosek:2014ch}.
+There is a `sweet spot' for straying that maximizes metapopulation robustness. 
+Results from our model reveal that the presence of just enough straying to cause formation of alternative stable states increases the portfolio effect (figure \ref{fig:PE}a). 
+Previous theoretical work has shown that increased connectivity may erode portfolio effects in herring metapopulations, where straying is also thought to be density-dependent \cite{Secor:2009ena}.
+Although high levels of dispersal in our system shows a similar trend, the interplay between dispersal and PE is more subtle when selection for local adaptations is considered, such that low to intermediate levels of density-dependent straying result in an elevated PE, increasing the buffering capacity of the metapopulation.
+Although PE is measured at the steady state, low to intermediate rates of straying also appear to have a beneficial effect on transient dynamics.
+When there is just enough straying to cause alternative stable states, recovery times following induced disturbances decline, though -- as with the PE -- the time to recovery then grows if straying becomes too large (figure \ref{fig:relax}a).
 
-%Comparing recovery times
-Previous theoretical work has shown that the largest perturbations lead to the longest recovery times \cite{Ovaskainen:2002il}.
-In terms of relative biomass lost, the largest perturbation that we investigate is the near-collapse of both fisheries, where only 1\% of the preperturbation population densities survive, however whether this disturbance results in the longest recovery time depends largely on the rate of straying and trait heritability.
-For example, when straying and trait heritability are low (figure \ref{fig:relax}a,c) the extinction of the larger population maximizes return time, whereas if straying is high, near-collapse maximizes return time.
-The latter is always true in the alternative stable state regime is trait heritability is high (figure \ref{fig:relax}b,d)
+% %Density dependent straying
+% This themed issue is largely concerned with the notion that straying in biological systems where movement is a function of collective navigation may be density-dependent.
+% Berdahl et al.
+(REF) provided a mechanistic hypothesis for density-dependent straying where the proportion of the population that strays is less than the probability that a single individual takes the wrong turn.
+%(as illustrated by \emph{Kevin}; figure \ref{fig:xkcd}) takes the wrong turn.
+% Neither the transient nor asymptotic dynamics that we describe here qualitatively differ as a function of whether the rate of straying is constant or density-dependent, however we observe important quantitative differences that suggest density-dependent straying may play an important role in the persistence of metapopulations over evolutionary time.
 
-%Density dependent straying
-This issue is largely focused on the notion that straying in biological systems where movement is a function of collective navigation may be density dependent.
-Berdahl et al.
-(REF) provided a mechanistic theory for density dependent straying where the proportion of the population that strays is less than the probability that a single individual (as illustrated by \emph{Kevin}; figure \ref{fig:xkcd}) takes the wrong turn.
-Neither the transient nor asymptotic dynamics that we describe here qualitatively differ as a function of whether the rate of straying is constant or density dependent, however we observe important quantitative differences that suggest density dependent straying may play an important role in the persistence of metapopulations over evolutionary time.
+This themed issue formalizes the role of collective movement in the ecology of natural systems and illuminates a signature of collective navigation in animal populations on the move.
+We observed two salient results that contribute to our understanding of collective movement that combine to suggest that density dependent straying may play an important role in the persistence of metapopulations over evolutionary time.
 
-First, density dependent straying reduces the time to recovery following an induced disturbance, and this is particularly true if (\emph{i}) trait heritability is low, and (\emph{ii}) in the case of near-collapse of both populations (figure \ref{fig:relax}c).
-This dynamic emerges because although both populations are near-collapse, the mean trait values of both are scewed towards those of the high-density population.
-As the low-density population has also collapsed, and the mean trait values of the system favor the high-density population, it grows more quickly, recovering faster (figure \ref{fig:relaxtraj_bothlh}).
-Second, density dependent straying increases the effective portfolio effects when the steady state straying rate is higher (for both low- and high-density populations).
-Third, while habitat heterogeneity is interpreted here in elevating the difference in trait optima between sites, it both increases the portfolio effect and lowers recovery time at low rates of straying, but has the opposite effect at higher rates of straying.
-Surprisingly, we find that density dependent straying both exaggerates the negative effects when $m^*$ is low (marginally lower portfolio effects, but subsatially longer return times; figures \ref{fig:thetaPE}b, \ref{fig:relaxtheta}a,b) and exaggerates the positive effects when $m^*$ is high (higher portfolio effects, abd shorter return times).
+First, compared to systems with constant dispersal, density-dependent straying appears to increase the portfolio effect across a range of stray rates (figure \ref{fig:PEtheta}a). 
+% and that at low to intermediate levels of straying, PE is high.
+Second, density-dependent straying reduces the time to recovery following disturbances, and this is particularly true in the case of near-collapse of both populations (figure \ref{fig:relax}b).
+In this case, although both populations are near-collapse, the mean trait values of both are skewed towards the optimum for the dominant population.
 
-HAVENT EDITED FURTHER
+ the straying rates are also higher because of the low-densities after collapse, and this permits faster recovery (figure \ref{fig:relaxtraj_bothlh}).
+% Third, in habitats with greater heterogeneity between sites (measured by an increase in $\Delta\theta$), density-dependent straying diminished robustness at low rates of straying (marginally lower portfolio effects and substantially longer return times; figures \ref{fig:thetaPE}b, \ref{fig:relaxtheta}a,b).
+% Conversely, for increased straying rates, greater and exaggerates the positive effects when $m^*$ is high (higher portfolio effects, and shorter return times).
 
-Trait heritability determines the coupling between ecological and evolutionary dynamics, effectively setting the strength of natural selection.
-Among salmon species, recruitment among local populations is highly sensitive to local climatic conditions, with temperature thought to play a central role (REFS).
-Populations distributed across a temperature gradient are assumed to be locally adapted to different temperature regimes, such that migration between them should take into account differential environmental effects on the recruitment of mixed populations.
-\citeauthor{Carlson:2008hl} showed heritability 0.1-0.3.
-Our results show that trait heritability has large effects on both metapopulation persistance as well as phenotypic diversity, particularly in the regime of low-to-moderate straying rates.
+% %Comparing recovery times
+% Previous theoretical work has shown that the largest perturbations lead to the longest recovery times \cite{Ovaskainen:2002il}.
+% In terms of relative biomass lost, the largest perturbation that we investigate is the near-collapse of both populations, where only 1\% of the pre-perturbation densities survive, however whether this disturbance results in the longest recovery time depends largely on the rate of straying and trait heritability.
+% For example, when straying and trait heritability are low (figure \ref{fig:relax}) the extinction of the larger population maximizes return times, whereas if straying is high, near-collapse maximizes return time.
+% The latter is always true in the alternative stable state regime if trait heritability is high (figure \ref{fig:relax_highh}).
 
-%Coupling distance: North-South vs East-West
+%Habitats 
+Salmon are distributed and stray across a diverse range of habitats, and the rates of straying between geographically diverse sites can be plastic and idiosyncratic \cite{Westley:2015to}.
+Our surrogate measure for habitat heterogeneity is the difference in trait optima between sites $\Delta\theta$.
+In general, our findings indicate that increased habitat heterogeneity promotes robustness (higher PE, shorter time to recovery) when straying rates are low, but may erode robustness when straying rates are high (figure \ref{fig:thetaPE}b, solid lines).
+This may be particularly consequential for populations that are spatially adjacent but separated by  sharp environmental boundary, such that trait optima are divergent yet dispersal is relatively high (is this a thing?)...
+
+%mtheta
+However, habitats that are closer in space generally have greater similarity in environmental conditions than habitats that are geographically distant (REF).
+From a purely statistical standpoint, it is reasonable to expect a larger number of straying individuals between sites that are closer together.
+Alternatively, salmon that cue to specific environmental conditions are more likely to stray into sites that have structurally and physiognamically similar habitats (REFS).
+These considerations justify imposing a direct relationship between the rate of straying $m$ and habitat heterogeneity: as site dissimilarity increases, so too should the straying rate decrease.
+As site dissimilarity increases, so to should the optima in trait values for the resident populations.
+
+When habitat heterogeneity and the rate of straying are linked, we show that small amounts of either constant or density-dependent straying result in longer recovery times for the dominant population because there is enough isolation to allow the surviving subordinant population locally adapt towards its own optimum.
+As selection pushes the subordinate population away from the dominant trait optimum, the recovery of the extinct population is slowed (figure \ref{fig:mtheta}).
+This may have particular conservation implications...
+Moreover, local temperature regimes are known to play a central role in dictating local adapation of salmon populations (REFS).
+If temperature is the primary determinant of $\Delta\theta$ in terms of our model, this would suggest that populations spanning North-South gradients are more predisposed to distance dependent dynamics.
+In constrast, the relationship between the rate of straying and habitat heterogeneity may be less clear-cut for populations spanning an East-West gradient, where distant habitats may equally fitting for straying populations.
+(Bristal Bay vs. California???)
+
+%Concluding
+The importance of selection and its influence on the dynamics of populations is increasingly recognized as vital for assessing species' fragility.
+Of particular interest is how selection might influence spatially coupled metapopulations distributed across environmental gradients.
+We have shown that two measures of metapopulation robustness -- the portfolio effect and the time to recovery following a disturbance -- are strongly influenced by the rate at which individuals from a donor population stray into habitats occupied by a recipient population.
+Importantly, density-dependent straying, which may occur when individuals collectively navigate, can both increase the portfolio and lower the time to recovery following a disturbance, which is anticipated to promote persistence.
+We suggest that understanding the spatial complexity of metapopulations dispersing across heterogeneous environments, combined with the mosaic of selective forces acting on those environments, may be key to discovering those factors that promote persistence.
 
@@
 \caption{
-Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.8$.
-If $m$ is density dependent, in the alternative steady state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
+Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.8$.
+If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
 } \label{fig:relax_highh}
@@
 \centering
-\includegraphics[width=0.35\textwidth]{figs2/fig_thetadiffN.pdf}
+\includegraphics[width=0.4\textwidth]{figs2/fig_thetadiffN.pdf}
 \caption{
 Median difference in population densities taken over the straying rate as a function of habitat heterogeneity $\Delta\theta$.
-Solid lines are for constant $m$; dashed lines are for density dependent $m$} \label{fig:thetadiffN}
+Solid lines are for constant $m$; dashed lines are for density-dependent $m$} \label{fig:thetadiffN}
 \end{figure}
 
-\begin{figure}
+\begin{figure*}
   \captionsetup{justification=raggedright,
@@
 \centering
-\includegraphics[width=0.35\textwidth]{figs2/fig_relaxtheta.pdf}
+\includegraphics[width=0.8\textwidth]{figs2/fig_relaxtheta2.pdf}
 \caption{
-Recovery time as a function of straying rate $m$ and habitat heterogeneity $\Delta\theta$.
+(a) Recovery time after near collapse of both populations as a function of straying rate $m$ and habitat heterogeneity $\Delta\theta$.
+(b) The same as (a) but including recovery times when straying is density-dependent, shown by linked pairs of points.
+Recovery times for systems with density-dependent straying are longer at low straying rates and shorter at higher straying rates, mirroring the change in portfolio effects shown in figure \ref{fig:thetaPE}.
 } \label{fig:relaxtheta}
-\end{figure}
+\end{figure*}
 
@@
 Distance dependent straying, where increased differences in trait optima between sites $\Delta\theta$ corresponds to lower rates of straying $m$.
-At low rates of straying $m=0.02$ ($\Delta\theta=24$), extinction of the dominant population leads to slower-than-expected recovery times because the subordinate population is isolated enough to evolve towards its own trait optimum until growth of the dominant population overwhelms this local selection.
+At low rates of straying $m=0.02$ ($\Delta\theta=24$), extinction of the dominant population leads to slower-than-expected recovery times because the subordinate population is isolated enough to evolve towards its own trait optimum.
+%until growth of the dominant population overwhelms this local selection.
 In this case, $m$ is less than $m=0.034$ (denoted by the asterisk in figure \ref{fig:mtheta}), such that isolation allows the subdominant population to `run away' from the influence of the dominant population, leading to a switch in states.
+If $m$ is low but greater than $0.034$, isolation permits the subdominant population to `run away' from the influence of the dominant population, until it is overwhelmed by the recovering dominant population, and reverts back to its previous trait mean prior to the disturbance.
 } \label{fig:inertia}
@@
   Distance dependent recovery times for three disturbance types.
-When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases for constant (a) and density dependent (b) staying rates.
+When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases for constant (a) and density-dependent (b) staying rates.
   The fold bifurcation is not as clear in (b) because $\Delta\theta$ is a function of the individual straying rate $m_0$, whereas the x-axis in (b) is the straying rate at the steady state $m^*$.
@@
 % \caption{
-% The same simulations as presented in Figure \ref{fig:PE}, except with density dependent straying, where $m(t) = m_0\left(1-N(t)/(C+N(t)\right)$. Light colors = high values.
+% The same simulations as presented in Figure \ref{fig:PE}, except with density-dependent straying, where $m(t) = m_0\left(1-N(t)/(C+N(t)\right)$. Light colors = high values.
 % } \label{fig:ddm}

### 2017-06-01 manuscript/msv3.tex
up to last paragraph

@@
 % There was greater similarity in the recovery times for constant and density-dependent $m$ systems when trait heritability was high (figure \ref{fig:relax_highh}).
-The lower recovery time for systems with increased $m^*$ mirrors an elevated PE with higher density-dependent straying rates (figure \ref{fig:PEtheta}).
+The lower recovery time for systems with increased $m^*$ mirrors an elevated PE with higher density-dependent straying rates (figure \ref{fig:thetaPE}).
 % This higher PE at increased $m^*$ mirrors recovery times that are also lower than those observed for systems with constant $m$ (figure \ref{fig:relax}).
@@
 % At face value, the subordinate population is pushed closer to extinction (figure \ref{fig:traj}), suggesting that the system may be more fragile.
-% On the other hand, the presence of \emph{just enough} straying to cause formation of alternative stable states both increases the portfolio effect (figure \ref{fig:PE}c) and reduces the rate of return -- particularly if straying is density-dependent (figures \ref{PEtheta}a, \ref{fig:relax}b).
+% On the other hand, the presence of \emph{just enough} straying to cause formation of alternative stable states both increases the portfolio effect (figure \ref{fig:PE}c) and reduces the rate of return -- particularly if straying is density-dependent (figures \ref{thetaPE}a, \ref{fig:relax}b).
 % This asymmetry in population densities also leads an asymmetry in the mean trait values, skewed towards the local optimum of the dominant population, one, and this has important -- and often non-intuitive -- ramifications for metapopulation robustness.
@@
 This themed issue formalizes the role of collective movement in the ecology of natural systems and illuminates a signature of collective navigation in animal populations on the move.
-We observed two salient results that contribute to our understanding of collective movement that combine to suggest that density dependent straying may play an important role in the persistence of metapopulations over evolutionary time.
-
-First, compared to systems with constant dispersal, density-dependent straying appears to increase the portfolio effect across a range of stray rates (figure \ref{fig:PEtheta}a). 
+We observed three salient results that contribute to our understanding of collective movement that combine to suggest that density-dependent straying may play an important role in the persistence of metapopulations over evolutionary time.
+First, the inclusion of density-dependent straying does not qualitatively alter either (\emph{i}) steady state or (\emph{ii}) transient dynamics of our minimal eco-evolutionary model, but effectively rescales measures of robustness to the straying rates the emerge from the coupled dynamics.
+Second, compared to systems with constant dispersal, density-dependent straying appears to increase the portfolio effect across a range of straying rates (figure \ref{fig:thetaPE}a). 
 % and that at low to intermediate levels of straying, PE is high.
-Second, density-dependent straying reduces the time to recovery following disturbances, and this is particularly true in the case of near-collapse of both populations (figure \ref{fig:relax}b).
-In this case, although both populations are near-collapse, the mean trait values of both are skewed towards the optimum for the dominant population.
-
- the straying rates are also higher because of the low-densities after collapse, and this permits faster recovery (figure \ref{fig:relaxtraj_bothlh}).
+Third, density-dependent straying reduces the time to recovery following disturbance, and this is particularly true in the case of near-collapse of the metapopulation (figure \ref{fig:relax}b).
+In the case of near-collapse, although both populations inherit low population densities, the mean trait values of both are skewed towards the optimum of the dominant population.
+Due to density-dependent straying, low population densities evince greater dispersal, and while this increased connectivity primarily facilities the growth of the dominant population (because the trait means are closer to the dominant optimum), because the dominant population is the bulk of the aggregate biomass, the overall recovery time is significantly lessened (figures \ref{fig:relax}b, \ref{fig:relaxtraj_bothlh}).
 % Third, in habitats with greater heterogeneity between sites (measured by an increase in $\Delta\theta$), density-dependent straying diminished robustness at low rates of straying (marginally lower portfolio effects and substantially longer return times; figures \ref{fig:thetaPE}b, \ref{fig:relaxtheta}a,b).
@@
 In general, our findings indicate that increased habitat heterogeneity promotes robustness (higher PE, shorter time to recovery) when straying rates are low, but may erode robustness when straying rates are high (figure \ref{fig:thetaPE}b, solid lines).
-This may be particularly consequential for populations that are spatially adjacent but separated by  sharp environmental boundary, such that trait optima are divergent yet dispersal is relatively high (is this a thing?)...
+This may be particularly consequential for populations that are spatially adjacent but separated by sharp environmental boundaries, such that trait optima are divergent yet dispersal is relatively high.
+Such a scenario plays out repeatedly in the context of adjacent wild and hatchery-produce salmon.
+Although wild and hatchery populations may occur close on the landscape, and indeed often are sympatric within the same river network, the selective environments to which they are locally adapted differ dramatically \cite{Christie:2012bj}.
+Straying of domesticated hatchery-produced fish from release sites and spawning in the wild drastically reduces the productivity of wild populations through competition and outbreeding depression \cite{Chilcote:2003bb,Araki:2007cm}.
 
 %mtheta
-However, habitats that are closer in space generally have greater similarity in environmental conditions than habitats that are geographically distant (REF).
-From a purely statistical standpoint, it is reasonable to expect a larger number of straying individuals between sites that are closer together.
-Alternatively, salmon that cue to specific environmental conditions are more likely to stray into sites that have structurally and physiognamically similar habitats (REFS).
+All else being equal, habitats that are closer in space generally have greater similarity in environmental conditions than habitats that are geographically distant, and correspondingly phenotypes of more proximately located populations should be more similar to those that are distant \cite{Westley:2012ui}.
+Correspondingly, it is reasonable to expect a larger number of straying individuals between sites that are geographically proximate and indeed evidence corroborates this prediction \cite{Candy:2000hu,JPE:JPE1383}.
+Alternatively, salmon that cue to specific environmental conditions may be more likely to stray into sites that have structurally and physiognamically similar habitats \cite{Peterson:2014gy} even if other potential sites are closer.
 These considerations justify imposing a direct relationship between the rate of straying $m$ and habitat heterogeneity: as site dissimilarity increases, so too should the straying rate decrease.
@@
 
-When habitat heterogeneity and the rate of straying are linked, we show that small amounts of either constant or density-dependent straying result in longer recovery times for the dominant population because there is enough isolation to allow the surviving subordinant population locally adapt towards its own optimum.
+When habitat heterogeneity and the rate of straying are linked, we show that small amounts of either constant or density-dependent straying result in longer recovery times for the dominant population because there is enough isolation to allow the surviving subordinate population to slowly adapt towards its own optimum and therefore strays from that population will be maladaptive in the dominant site.
 As selection pushes the subordinate population away from the dominant trait optimum, the recovery of the extinct population is slowed (figure \ref{fig:mtheta}).
-This may have particular conservation implications...
-Moreover, local temperature regimes are known to play a central role in dictating local adapation of salmon populations (REFS).
+This may have particular conservation implications for considerations of aiding dispersal following disturbances or reconnections of lost habitats \cite{Anderson:2013bf,Pess:2014isa}.
+Moreover, local temperature regimes are known to play a central role in dictating local adaptation of salmon populations (REFS).
 If temperature is the primary determinant of $\Delta\theta$ in terms of our model, this would suggest that populations spanning North-South gradients are more predisposed to distance dependent dynamics.
-In constrast, the relationship between the rate of straying and habitat heterogeneity may be less clear-cut for populations spanning an East-West gradient, where distant habitats may equally fitting for straying populations.
-(Bristal Bay vs. California???)
+In contrast, the relationship between the rate of straying and habitat heterogeneity may be less clear-cut for populations spanning an East-West gradient, where distant habitats may equally fitting for straying populations.
+
+%I'm here
 

### 2017-06-02 Manuscript/CoverLetter/YGWM_cover.tex
final tweaks on figs

@@
 
-Our theory is analytic, relatively simple, and synthesizes a variety of fundamental ecological theories (energy equivalence hypothesis, resource competition theory, Cope’s rule, the fasting endurance hypothesis) and shows how many of these are the natural consequence of a simple dynamical framework that incorporates the energetic consequences of starvation and recovery combined with allometric timescales.
-We believe that our findings will be of general interest to ecologists, evolutionary biologists, and paleoecologists, such that consideration for publication in a broad interest journal such as \emph{Philosophical Transactions} is warranted.
+Our theory is general, relatively simple, and synthesizes a variety of fundamental ecological and evolutionary theories to show that our results are the natural consequence of a simple dynamical framework that incorporates the dispersal and local selection.
+We believe that our findings will be of general interest to ecologists, freshwater biologists, and evolutionary biologists, such that consideration for publication in a broad interest journal such as \emph{Philosophical Transactions} is warranted.
 

### 2017-06-02 manuscript/msv4.tex
final tweaks on figs

--- 2017-06-01 manuscript/msv3.tex
@@
-\documentclass[twocolumn,preprintnumbers,amsmath,amssymb,superscriptaddress]{revtex4}
+\RequirePackage{lineno}
+\documentclass{revtex4}
 %\usepackage[pdftex]{graphicx}
@@
 \captionsetup{compatibility=false}
-
+\usepackage{xr}
+\externaldocument[extfig-]{supp}
 %\usepackage{epstopdf}
@@
 %\usepackage{rotating}
-%\usepackage{lineno}
+% \usepackage{lineno}
+
 %\usepackage[left]{lineno}
@@
 
-\bibliographystyle{ieeetr}
+\linespread{1.5}
 
-\usepackage{bibunits}
+\usepackage{natbib}
+\bibpunct{(}{)}{,}{a}{}{;}
+
+\renewcommand{\thesection}{\arabic{section}}
+
+\bibliographystyle{prsb}
+
+% \usepackage{bibunits}
 
@@
 
-%\linenumbers
-%\setlength\linenumbersep{3pt}
+% \pagewiselinenumbers
+\linenumbers
+% \setlength\linenumbersep{3pt}
 
@@
 
+% \Blindtext
 %\title{Simple rules yield complex communities: deconstructed species interactions and the assembly of communities}
@@
 \title{Eco-evolutionary dynamics and collective straying: implications for salmon metapopulation robustness}
-\author{Justin D. Yeakel${}^{1,2,*}$, Jean P. Gibert${}^{1}$, Peter A. H. Westley${}^{3}$, \& Jonathan W. Moore${}^{4}$ \\
+\author{
+Justin D. Yeakel${}^{1,2,*}$, Jean P. Gibert${}^{1}$, Peter A. H. Westley${}^{3}$, \& Jonathan W. Moore${}^{4}$ \\
 ${}^1$School of Natural Sciences, University of California, Merced, Merced CA, USA \\
@@
 
+\begin{abstract} %250 words
+The spatial dispersal of individuals is known to play an important role in the dynamics of populations, and is central to metapopulation theory.
+At the same time, local adaptation to different environmental conditions creates a geographic mosaic of evolutionary forces, where the combined drivers of selection and gene flow interact.
+Although the dispersal of individuals from donor to recipient populations provides connections within the metapopulation, potentially promoting demographic and evolutionary rescue, it may also introduce maladapted individuals into habitats that are host to different environmental conditions, potentially lowering the mean fitness of the recipient population.
+Here we explore a minimal model of the eco-evolutionary dynamics between two populations connected by dispersal, where the productivity of each is defined by a trait complex that is subject to local selection.
+Although general in nature, our model is inspired by salmon metapopulations, where dispersal between populations is defined in terms of the \emph{straying rate}, which has been shown to be density dependent, possibly emerging as a consequence of collective migration.
+The results of our model reveal that increased straying between evolving populations leads to alternative stable states, which has large and nonlinear effects on two measures of metapopulation robustness: the portfolio effect and the time to recovery following an induced disturbance.
+We show that intermediate levels of straying result in large gains in robustness, and that increased differences between sites (increased habitat heterogeneity) promotes robustness when straying rates are low, and erodes robustness when straying rates are high.
+Finally, we show that density-dependent straying promotes robustness, particularly when the aggregate biomass is low and straying is correspondingly high, and this has important ramifications for the conservation of salmon metapopulations facing both natural and anthropogenic disturbances.
+\end{abstract} %250 words
+
 \maketitle
 
-\section*{Abstract}
-{\small
-We show that taking straying into account leads to alternative stable states in population densities and trait values, which has consequences for maladaptation, intraspecific trait variability, and long-term robustness of salmon metapopulations.
-}
 \section{Introduction}
@@
 Intraspecific diversity can increase the resilience and stability of species or metapopulations. 
-This diversity-stability linkage can occur when there are asynchronous population dynamics, where the changes in population vary across a metapopulation. 
-This asynchrony will increase the potential for demographic rescue \cite{Brown:1977gk,Earn:2000fm} and also decrease the variability of processes that integrate across the metapopulation \cite{Anonymous:2015gf}.
-For example, different responses to climate variability within populations of a rare plant reduced fluctuations in abundance \cite{Abbott:2017hl}. 
-This statistical buffer has traditionally been quantified as the Portfolio Effect (PE), which is the ratio of the population CV to the CV of the aggregated metapopulation \cite{Thibaut:2012km}. 
-Strengthened portfolio effects are expected to increase the robustness of metapopulations to external disturbances, and by extension promote persistence \cite{Thibaut:2012km}.
+This diversity-stability linkage can occur when there are asynchronous population dynamics, where the changes in population size varies across the metapopulation. 
+This asynchrony will increase the potential for demographic rescue \citep{Brown:1977gk,Earn:2000fm} and also decrease the variability of processes that integrate across the metapopulation \citep{Anonymous:2015gf}.
+For example, different responses to climate variability within populations of a rare plant reduced fluctuations in abundance \citep{Abbott:2017hl}. 
+This statistical buffer has traditionally been quantified as the Portfolio Effect (PE), which is the ratio of the population CV to the CV of the aggregated metapopulation \citep{Thibaut:2012km}. 
+Strengthened portfolio effects are expected to increase the robustness of metapopulations to external disturbances, and by extension promote persistence \citep{Thibaut:2012km}.
 In contrast, homogenization of populations leading to greater synchronization and weakened PE may be a harbinger of metapopulation collapse and extinction.
 
-It is clear that dispersal (i.e., movement with consequences for gene flow) can influence the evolutionary dynamics of the metapopulation, the interplay between selection, population dynamics, and persistence is not well understood.
-Movement of individuals among local populations (i.e. dispersal) can have a large influence on metapopulation persistence \cite{MilnerGulland:2011vm}. 
-Dispersal facilitates evolutionary rescue, whereby immigration of individuals with heritable adapative traits can rescue small populations from local extinction in the context of maladaptive environmental change \cite{Bell:2011ki,Carlson:2014is}.
+% Although dispersal and the consequent gene flow can influence the evolutionary dynamics of metapopulations, the interplay between selection, population dynamics, and persistence is not well understood.
+Movement of individuals among local populations (i.e. dispersal) can have a large influence on metapopulation persistence \citep{MilnerGulland:2011vm}. 
+Dispersal facilitates evolutionary rescue, whereby immigration of individuals with heritable adapative traits can rescue small populations from local extinction in the context of maladaptive environmental change \citep{Bell:2011ki,Carlson:2014is}.
 % First, some level of straying is necessary to enable the rescue effect, whereby dispersal among populations can rescue small populations from local extinction (REFS). 
-On the other hand, high rates of dispersal may synchronize populations and actually increase the risk of extinction of the entire metapopulation \cite{Earn:2000fm}. 
+On the other hand, high rates of dispersal may synchronize populations and actually increase the risk of extinction of the entire metapopulation \citep{Earn:2000fm}. 
 Third, straying will influence the evolutionary dynamics of the metapopulation.
 % though how the interplay between selection and population dynamics might impact persistence is less well understood. 
-Although the dispersal of individuals into sites hosting other populations provides connections within the larger metapopulation, potentially promoting demographic and evolutionary rescue, it may also introduce maladapted individuals into habitats that are host to different environmental conditions, possibly lowering the mean fitness of the recipient population \cite{Muhlfeld:2014hs}.
-More broadly, dispersal can provide a mechanism by which phenotypes are sorted in space rather than time and facilitates the spread of maladaptive genes \cite{Lowe:2015ft}.
+Although the dispersal of individuals into sites hosting other populations provides connections within the larger metapopulation, potentially promoting demographic and evolutionary rescue, it may also introduce maladapted individuals into habitats that are host to different environmental conditions, possibly lowering the mean fitness of the recipient population \citep{Muhlfeld:2014hs}.
+More broadly, dispersal can provide a mechanism by which phenotypes are sorted in space rather than time and facilitates the spread of maladaptive genes \citep{Lowe:2015ft}.
 Further, this straying may lead to genetic homogenization that erodes the asynchrony that underpins portfolio effects and metapopulation persistence. 
@@
 %; we might say that there is at least one \emph{Kevin} in every school or flock (figure \ref{fig:xkcd}).
-There is growing appreciation that a combination of abiotic, biotic, and anthropogenic factors can control the rate of dispersal among populations \cite{H:2013fs,Keefer:2014gg,Bett:2017ha}.
+There is growing appreciation that a combination of abiotic, biotic, and anthropogenic factors can control the rate of dispersal among populations \citep{H:2013fs,Keefer:2014gg,Bett:2017ha}.
 Migratory populations that return to breeding sites for reproduction are linked to each other by some proportion of the population that permanently disperses into the `wrong' site. 
 Recently, the role of social interactions and collective navigation has been hypothesized [This issue].
-The rate at which individuals stray, $m$, may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \cite{Simons:2004jo,Berdahl:2015kv,Berdahl:2016dx}.
-The potential influence of collective dispersal on the dynamics of individual populations and the metapopulation as a whole is a topic of considerable interest that has tangible conservation implications \cite{Brenner:2012gl,Johnson:2012fe,Fullerton:2011ii}.
+The rate at which individuals stray, $m$, may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \citep{Simons:2004jo,Berdahl:2015kv,Berdahl:2016dx}.
+The potential influence of collective dispersal on the dynamics of individual populations and the metapopulation as a whole is a topic of considerable interest that has tangible conservation implications \citep{Brenner:2012gl,Johnson:2012fe,Fullerton:2011ii}.
 % Whether and to what extent the ecological consequences of straying depend on the evolutionary dynamics that emerge from populations distributed across a selective gradient is unclear.
@@
 %% From geographic mosaic of coevolution to...
-% That evolutionary forces of selection and gene flow play out heterogeneously across geographic mosaics is now a foundational concept in ecology and evolutionary biology \cite{Nuismer:1999ko,Thompson:2015hq,Nuismer:2000bb,Thompson:2005wf}.
-% These mosaics are in part driven by environmental differences between habitats that alter the selective forces acting on different phenotypes \cite{Endler:1986tz}, and a principle underlying assumption is that there is gene flow such that individuals from different habitats mix over space \cite{Gomulkiewicz:2000bz,Thompson:2002hr,Nuismer:2003eh,Nuismer:2006be,Forde:2008kc,Guimaraes:2011hu,Gibert:2013hp}.
-% Although the evolutionary outcomes of these spatial processes have been explored in depth (REFS), it is less well understood how selective mosaics and their consequent evolutionary forces impact population dynamics as they unfold \cite{Hendry:2016un}.
+% That evolutionary forces of selection and gene flow play out heterogeneously across geographic mosaics is now a foundational concept in ecology and evolutionary biology \citep{Nuismer:1999ko,Thompson:2015hq,Nuismer:2000bb,Thompson:2005wf}.
+% These mosaics are in part driven by environmental differences between habitats that alter the selective forces acting on different phenotypes \citep{Endler:1986tz}, and a principle underlying assumption is that there is gene flow such that individuals from different habitats mix over space \citep{Gomulkiewicz:2000bz,Thompson:2002hr,Nuismer:2003eh,Nuismer:2006be,Forde:2008kc,Guimaraes:2011hu,Gibert:2013hp}.
+% Although the evolutionary outcomes of these spatial processes have been explored in depth (REFS), it is less well understood how selective mosaics and their consequent evolutionary forces impact population dynamics as they unfold \citep{Hendry:2016un}.
 
@@
 The eco-evolutionary impacts of dispersal likely has important implications for conservation and management in key taxa such as in migratory salmon.
-While anadromous salmonid fishes (genera \emph{Oncorhynchus} and \emph{Salmo}) are renown for returning to their natal spawning habitats with high accuracy and precision after years at sea \cite{Quinn:2011tf,Jonsson:2011kg,Keefer:2014gg}, there are generally some individuals that `stray' to non-natal sites to spawn \cite{Quinn:1993ge,Hendry:2004wf}.
-Salmon may operate as metapopulations, where populations are genetically distinct but linked by some level of straying \cite{Schtickzelle:2007wb,Anderson:2014cx}.
-Although extensive work has been done to document the extent of straying from donor populations and into recipient populations \cite{Keefer:2014gg,Bett:2017ha}, only recently have the abiotic, biotic, and anthropogenic influences of `straying' behaviors been investigated systemically \cite{Keefer:2008bs,Westley:2015to,Bond:2016dz}.
-Straying among salmon may be influenced by environmental factors such as water temperature, human activities such as hatchery practices, and population density as predicted by the collective migration hypothesis \cite{Peterson:2014gy}.
+While anadromous salmonid fishes (genera \emph{Oncorhynchus} and \emph{Salmo}) are renown for returning to their natal spawning habitats with high accuracy and precision after years at sea \citep{Quinn:2011tf,Jonsson:2011kg,Keefer:2014gg}, there are generally some individuals that `stray' to non-natal sites to spawn \citep{Quinn:1993ge,Hendry:2004wf}.
+Salmon may operate as metapopulations, where populations are genetically distinct but linked by some level of straying \citep{Schtickzelle:2007wb,Anderson:2014cx}.
+Although extensive work has been done to document the extent of straying from donor populations and into recipient populations \citep{Keefer:2014gg,Bett:2017ha}, only recently have the abiotic, biotic, and anthropogenic influences of `straying' behaviors been investigated systemically \citep{Keefer:2008bs,Westley:2015to,Bond:2016dz}.
+Straying among salmon may be influenced by environmental factors such as water temperature, human activities such as hatchery practices, and population density as predicted by the collective migration hypothesis \citep{Peterson:2014gy}.
 These strays can introduce new maladaptive genotypes into the recipient population.
-Further, straying and genetic homogenization could synchronize population dynamics and erode portfolio effects \cite{Moore:2010gs,Carlson:2011ce,Braun:2016ib}.
+Further, straying and genetic homogenization could synchronize population dynamics and erode portfolio effects \citep{Moore:2010gs,Carlson:2011ce,Braun:2016ib}.
 Thus, there is opportunity and a need to consider the eco-evolutionary consequences of straying for metapopulations in species of conservation and management concern such as salmon.
 
-% Asynchrony among salmon populations also increases the stability of fisheries that integrate across this diversity \cite{Schindler:2010he,Anonymous:2016jc}.
+% Asynchrony among salmon populations also increases the stability of fisheries that integrate across this diversity \citep{Schindler:2010he,Anonymous:2016jc}.
 % However, these portfolio effects may be eroded if asynchrony is influenced by genetic diversity and this genetic diversity is lost. 
-% Braun et al. \cite{Braun:2016ib} found that genetic differences among populations of Chinook salmon was correlated with the degree of asynchrony in population dynamics. 
-% Artificial propagation of salmon by hatcheries and their strays may erode local genetic diversity, asynchrony among populations, and portfolio effects \cite{Carlson:2008hl,Moore:2010gs,Anonymous:2014ku}.
-%Coordinated mass migrations are one of the great wonders of the natural world, and the ability of individuals and groups to navigate across great distances have long fascinated naturalists \cite{MilnerGulland:2011vm}.
+% Braun et al. \citep{Braun:2016ib} found that genetic differences among populations of Chinook salmon was correlated with the degree of asynchrony in population dynamics. 
+% Artificial propagation of salmon by hatcheries and their strays may erode local genetic diversity, asynchrony among populations, and portfolio effects \citep{Carlson:2008hl,Moore:2010gs,Anonymous:2014ku}.
+%Coordinated mass migrations are one of the great wonders of the natural world, and the ability of individuals and groups to navigate across great distances have long fascinated naturalists \citep{MilnerGulland:2011vm}.
 
-% Most recently the role of social interactions and collective navigation has been hypothesized \cite{Berdahl:2015kv,Berdahl:2016dx}.
+% Most recently the role of social interactions and collective navigation has been hypothesized \citep{Berdahl:2015kv,Berdahl:2016dx}.
 % These strays can introduce new maladaptive genotypes into the recipient population.
-% Straying amon salmon may be influenced by environmental factors such as water temperature, human activities such as hatchery practices, and population density as predicted by the collective migration hypothesis \cite{Peterson:2014gy}.
-% Thus, there is an opportunity to consider the eco-evolutionary consequences of straying for metapopulations in species of conservation and management concern such as salmon \cite{Carlson:2011ce}.
+% Straying amon salmon may be influenced by environmental factors such as water temperature, human activities such as hatchery practices, and population density as predicted by the collective migration hypothesis \citep{Peterson:2014gy}.
+% Thus, there is an opportunity to consider the eco-evolutionary consequences of straying for metapopulations in species of conservation and management concern such as salmon \citep{Carlson:2011ce}.
 
@@
 % A flock of geese travel in the direction of a shared destination, the lone stray named \emph{Kevin}.
-% In this case, the rate of straying is $m=0.05$, which is not an uncommon rate for migrating populations of salmon \cite{Satterthwaite:2015ge}. 
+% In this case, the rate of straying is $m=0.05$, which is not an uncommon rate for migrating populations of salmon \citep{Satterthwaite:2015ge}. 
 % Reprinted under the Creative Commons Attribution-NonCommercial 2.5 License.
@@
 As such, the recruitment rate for both populations is determined by the mean trait value of the local population, such that $r_1 = R_1[\mu_1(t),\theta_1]$.
-Trait means for each population are subject to selection, the strength of which depends on the difference between the trait mean and the local trait optimum at a given point in time \cite{simpson1953major,Lande:1976ga}.
+Trait means for each population are subject to selection, the strength of which depends on the difference between the trait mean and the local trait optimum at a given point in time \citep{simpson1953major,Lande:1976ga}.
 
@@
 Again, straying can be thought of as both from the donor or recipient population perspective.
-The two populations are assumed to reproduce in spatially separate sites that are close enough such that a proportion of the population $m$ can stray into the other site, and where mortality occurs before individuals return to reproduce.
-If there is no straying between these populations (such that they are independent), then the mean trait evolves towards the optimal value such that $x_1 \rightarrow \theta_1$, and the recruitment rate for that population will be maximized.
+The two populations occur in spatially separate sites that are close enough such that a proportion of the population $m$ can stray into the other site, and where mortality occurs before individuals return to reproduce.
+If there is no straying between these populations, then the mean trait evolves towards the optimal value for each site $x_1 \rightarrow \theta_1$, and the recruitment rate for each population is maximized.
 If there is straying between populations at rate $m$, then the traits in each respective location will be pulled away from the optimum, and recruitment rates will be lowered.
@@
 
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.4\textwidth]{figs2/fig_traj.pdf}
-\caption{
-A) The steady state densities of $N_1$ and $N_2$ as a function of a constant stray rate $m$.
-Which population attains the low- or high-density state is random due to small applied fluctuations in the initial conditions.
-B) The steady state trait values measured as $\theta_i - x_i$, as a function of a constant stray rate $m$. 
-FB marks the fold bifurcation.
-} \label{fig:traj}
-\end{figure}
-
-We use the discrete Ricker population dynamic framework described by Shelton and Mangel \cite{Shelton:2011eq} as the basis for our two-site model, with the added effect of the local population $N_i$ mixing with a set proportion $m$ of a remote population $N_j$ that is straying into it.
-In this sense, both populations serve as donor and recipient populations.
+We use discrete Ricker framework described by Shelton and Mangel \citep{Shelton:2011eq} as the basis for our two-site metapopulation model, with the added effect of the local population $N_i$ mixing with a set proportion $m$ of a remote population $N_j$ that is straying into it.
+In this sense, both populations serve simultaneously as donor and recipient populations.
 We first assume that the proportion ${\rm e}^{-Z}$ of both populations survive such that the surviving aggregated population, composed of both local individuals (at site $i$) and incoming strays (from site $j$), is $\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z}$.
@@
 Because straying individuals are emigrating from a population with a mean trait value farther from the local optimum, their rate of recruitment is diminished.
-Recent studies of wild sockeye have indeed found that straying individuals have lower life-time fitness than individuals that do not stray, although it unknown at what life-stage this selection occurs \cite{Peterson:2014gy}.
+Recent studies of wild sockeye salmon have indeed found that straying individuals have lower life-time fitness than individuals that do not stray, although it unknown at what life-stage this selection occurs \citep{Peterson:2014gy}.
 % The compensatory effects are then determined by the exponential, following the Ricker stock-recruitment relationship.
 \\
-
-\begin{figure*}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.8\textwidth]{figs2/fig_MDPE_hm.pdf}
-\caption{
-(a) Total means $N_t$, (b) difference in means $\Delta N$, and (c) the portfolio effect PE as a function of heritability $h^2$ and a constant stray rate $m$. Light colors = high values.
-The black line shows the fold bifurcation separating a single steady state (left) from alternative stable states (right).
-(d) The relationship between the time to recovery following a disturbance and the portfolio effect.
-The black line denotes the fold bifurcation.
-} \label{fig:PE}
-\end{figure*}
 
@@
 First, we assume that the distribution resulting from the mix of remote and local individuals, following reproduction, is also normal with a mean value equal to that of the mean for the mixed-normal distribution.
-Thus, we assume that strays can successfully reproduce and introduce their genotypes into the recipient population, which is supported by observations in wild populations \cite{Jasper:2013cc}.
-Second, we assume that changes in trait variance through time are minimal, such that $\sigma^2$ is assumed to be constant, and this is a common simplification in eco-evolutionary models of population dynamics \cite{Lande:1976ga,Schreiber:2011wx,Gilbert:2014ee,Gibert:2015kc}.
+Thus, we assume that strays can successfully reproduce and introduce their genotypes into the recipient population, which is supported by observations in wild populations \citep{Jasper:2013cc}.
+Second, we assume that changes in trait variance through time are minimal, such that $\sigma^2$ is assumed to be constant, which is a common simplification in eco-evolutionary models of population dynamics \citep{Lande:1976ga,Schreiber:2011wx,Gilbert:2014ee,Gibert:2015kc}.
 
-An increasing flow of incoming strays is generally expected to pull the mean trait value of the local population away from its optimum over time, which will decrease its rate of recruitment.
-Following Lande \cite{Lande:1976ga}, the mean trait value thus changes through time according to the difference equation
+% An increasing flow of incoming strays is generally expected to pull the mean trait value of the local population away from its optimum over time, which will decrease its rate of recruitment.
+Following Lande \citep{Lande:1976ga}, the mean trait value thus changes through time according to the difference equation
 
@@
 
-\noindent where the first two factors determine the mixed normal average of the aggregated local and remote populations.
+\noindent where the first two components determine the mixed normal average of the aggregated local and remote populations.
 The partial derivative in the Eq.
-\ref{eq:mu} determines how the mean trait changes through time due to natural selection \cite{Lande:1976ga}, which is proportional to the change in mean fitness with respect to $\mu_i$.
+\ref{eq:mu} determines how the mean trait changes through time due to natural selection \citep{Lande:1976ga}, which is proportional to the change in mean fitness with respect to $\mu_i$.
 \\
@@
 %Density dependent m
-\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is mounting evidence that at least in some species (including salmon) the straying rate is density-dependent with a signature of collective navigation \cite{Berdahl:2016dx,Bett:2017ha}.
+\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is mounting evidence that at least in some species (including salmon) the straying rate is density-dependent with a signature of collective navigation \citep{Berdahl:2016dx,Bett:2017ha}.
 Specifically, the rate at which individuals stray has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tend to decrease the rate at which individuals err, reducing the overall proportion of a population that strays.
-According to Berdahl et al. \cite{Berdahl:2014bl,Berdahl:2016dx}, given the probability that an individual strays is $m_0$, the proportion of the local population $N_i(t)$ that strays is
+According to Berdahl et al. \citep{Berdahl:2014bl,Berdahl:2016dx}, given the probability that an individual strays is $m_0$, the proportion of the local population $N_i(t)$ that strays is
 
@@
 \noindent {\bf (c) Habitat heterogeneity}
+\noindent Increasing differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater regional differences in the conditions that favor alternative trait complexes, which we interpret here as increased habitat heterogeneity.
+If both populations are isolated, natural selection will direct the mean trait values of both populations towards their respective optima, such that $\mu_i(t) \rightarrow \theta_i$ as $t\rightarrow\infty$.
+%difference in optimal trait values between the local and remote populations, which we call . 
+%Linking m/m0 and thetadiff
+Although we largely treat habitat heterogeneity and the rate of straying as independent parameters, we evaluate a case where we assume thgat increased habitat heterogeneity correlates with lower straying rates, and vice versa.
+Two scenarios may lead to this correlation: 
+(\emph{i}) sites may be distributed over greater spatial distances, where habitat dissimilarity is assumed to be larger between more distant sites, whereas the likelihood of straying over greater distances would likewise be lower \citep{Candy:2000hu,JPE:JPE1383};
+(\emph{ii}) individuals may have behaviors promoting dispersal between habitats with structural or physiognomic similarities \citep{Peterson:2014gy}.
 
-%Linking m/m0 and thetadiff
-The straying rate is largely influenced by the spatial distance between the donor and recipient population.
-The greater the distance between two populations, the lower the expected rate of straying \cite{H:2013fs,Westley:2015to}.
-We account for this interdependence in our model by assuming that $m$ (if the stray rate is constant) or $m_0$ (if the stray rate is density-dependent) is a function of the difference between optimal trait values between sites $\theta_i-\theta_j$, which can be assumed to be large if the remote site $j$ is a great distance away from the local site $i$.
-If sites $i$ and $j$ are very close, the stray rate is maximized at $m_{\rm max} = 0.5$, assuming both sites are equally attractive to the respective populations.
-Thus, we can integrate these two variables by setting $m,m_0 = (1/m_{\rm max} + \epsilon (\theta_i-\theta_j))^{-1}$, where $\epsilon$ sets the sensitivity of a declining $m$ to increasing distance (greater values of $\theta_i-\theta_j$).\\
+Where indicated, we account for correlation between the straying rate and the differnce between optimal trait values by assuming that $m$ (if the stray rate is constant) or $m_0$ (if the stray rate is density-dependent) is a function of the difference between optimal trait values between sites $\Delta\theta$.
+The straying rate is maximized at $m_{\rm max} = 0.5$, assuming both sites are equally attractive to the respective populations.
+Thus, we can integrate these two variables by setting $m,m_0 = (1/m_{\rm max} + \epsilon \Delta\theta)^{-1}$, where $\epsilon$ sets the sensitivity of a declining $m$ to increasing distance (greater values of $\Delta\theta$), which we set to $\epsilon=2$.\\
 
 \noindent {\bf (d) Measuring metapopulation robustness}
-\noindent We evaluated metapopulation robustness by measuring the average-CV portfolio effect (PE) \cite{Anderson:2014cx,Schindler:2015gf} as well as the time required for the system to return to a steady state following an induced disturbance to one or both of the populations \cite{Ovaskainen:2002il}.
-The average-CV portfolio effect is, as the name implies, the average CV across each population divided by the CV of the aggregate \cite{Anderson:2013gb}, such that
+\noindent We evaluated metapopulation robustness by measuring the average-CV portfolio effect (PE) \citep{Anderson:2014cx,Schindler:2015gf} as well as the time required for the system to return to a steady state following an induced disturbance to one or both of the populations \citep{Ovaskainen:2002il}.
+The average-CV portfolio effect is, as the name implies, the average CV across each population divided by the CV of the aggregate \citep{Anderson:2013gb}, such that
 
@@
 As the CV of $N_T$ decreases relative to that of the constituent populations, $\langle{\rm PE}\rangle > 1$, and the metapopulation is presumed to become more stable.
-Portfolio effects greater than unity corresponds to less synchronization  \cite{Loreau:2008ju,Anderson:2014cx,Yeakel:2013vz} and thus a greater potential for demographic rescue among populations, buffering the system as a whole against extinction.
+Portfolio effects greater than unity corresponds to less synchronization  \citep{Loreau:2008ju,Anderson:2014cx,Yeakel:2013vz} and thus a greater potential for demographic rescue among populations, buffering the system as a whole against extinction.
 
 %Although there are more robust ways to measure the portfolio effect (in particular for comparing the PE across species with different mean-variance relationships; REF: Anderson), its only utlity here is to provide a simple statistical measure of metapopulation persistence.
-A more direct way to measure system robustness is to measure the time that it takes the system (measured as the aggregate biomass $N_T$) to recover its steady state abundance following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
-Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation \cite{GuckHolmes}, because we aimed to 1) assess the effects of a large perturbation, and 2) estimate the time required for all transient effects to decay (including dampened oscillations), we used a simulation-based numerical procedure.
+A more direct way to measure system robustness is to measure the time that the system (measured as the aggregate biomass $N_T$) takes to recover its steady state abundance following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
+Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation \citep{GuckHolmes}, because we aimed to 1) assess the effects of a large perturbation, and 2) estimate the time required for all transient effects to decay (including dampened oscillations), we used a simulation-based numerical procedure.
 
@@
 For example, if populations settle to alternative stable states (alternative stable states in our model requiring one population to be high-density and one low-density), comparing recovery times after a disturbance applied to the high, low, and/or both populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery.
-We measured the time required for the system $N_T$ to recover to its steady state following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} are equivalent if the system is in the single steady state regime); (\emph{iii}) near-collapse of both populations where just 1.0\% of each survives.
+We measured recovery time following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} are equivalent if the system is in the single steady state regime); (\emph{iii}) near-collapse of both populations where just 1.0\% of each survives.
 Throughout, we will refer to an increase in the portfolio effects and/or reduction in recovery times as promoting metapopulation robustness, which is expected to have a positive effect on persistence.
@@
 \section{Results}
+
+%Alternative stable states
+
+\noindent{\bf (a) Nonlinear effects of straying on metapopulation robustness} \\
+\noindent Regardless of density dependence, straying lowers steady state densities for both populations by (\emph{i}) the donor population losing locally-adapted individuals to the recipient population and (\emph{ii}) the introduction of maladapted individuals to the recipient population from the donor population (Fig.
+\ref{fig:traj}).
+This prediction is in accordance with observations from natural populations \citep{Bett:2017ha}. %, regardless of trait heritability and variance or habitat heterogeneity
+The decline in steady state densities is not gradual: as straying increases, the system crosses a fold bifurcation whereby the single steady state for the metapopulation bifurcates into two alternative stable states: one at high biomass, and one at low biomass density (figure \ref{fig:traj}a, \ref{fig:PE}a).
+Mean trait values for both populations bifurcate similarly (figure \ref{fig:traj}b), depending on which population attains a low- vs. high-density. 
+Above the threshold straying rate defined by the fold bifurcation, there are two alternative eco-evolutionary states: the \emph{dominant state} population will have a higher a higher density and higher degree of local adaptation (lower trait offset from the local optimum), while the \emph{subordinate state} population will have lower density with maladapted trait values (higher trait offset from the local optimum).
+Whether a specific population goes to one state or the other in our model is random, which is due to a small amount of introduced variance in the initial conditions.
+% Thus, we predict that high levels of straying in metapopulations will produce alternative eco-evolutionary states.
+
+Trait heritability has a large impact on the degree to which straying rate affects both the aggregate population steady state density ($N^*_T=N^*_1+N^*_2$; figure \ref{fig:PE}a) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N=|N^*_1-N^*_2|$; figure \ref{fig:PE}b).
+Greater trait heritability results in a faster decline in $N_T^*$ with increasing straying rates $m$, but leads to only moderate changes to $\Delta N$.
+Conversely, in the context of lower trait heritability, an increase in the straying rate has little impact on the total biomass density but contrastingly large effects on $\Delta N$.
+The fold bifurcation (the black line in Figs.
+\ref{fig:PE}a-c) occurs at lower values of the straying rate $m$ with decreased trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
+Although trait heritability among salmonids is variable, most life history traits have an $h^2 <0.5$ \citep{Carlson:2008hl}, and we largely focus additional analyses on that range.
+
+%Critical slowing down
+As the fold bifurcation is approached with increasing $m$, the portfolio effect increases sharply due to an amplification in variance within both donor and recipient populations.
+This amplification in variance is the product of a dynamical process known as \emph{critical slowing down} that occurs near fold bifurcations \citep{Scheffer:2009gg}, a phenomenon that some have suggested may serve as an early warning indicator for approaching phase transitions \citep{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br,Krkosek:2014ch}.
+For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}a-c), where alternative stable states occur, the portfolio effect declines steadily as the CV of $N_T$ increases.
+The decline over $m$ is more gradual if trait heritability is low, and steeper if trait heritability is high (figure \ref{fig:PE}c).
+
+% \noindent{\bf (c) Rates of recovery following catastrophic collapse}\\ 
+% Return time as a measure of system persistence and relation to PE
+As the portfolio effect is highly sensitive to the rate of straying between populations, so is the time required for the system to recover to a steady state following a large disturbance.
+In general, we find that the average-CV portfolio effect is negatively correlated with recovery time (figure \ref{fig:PE}d), indicating that, for our system, both measures are valuable indicators of metapopulation robustness.
+Because we can assess the time to recovery in response to the various disturbance types described above, this allows us to gain an in-depth perspective into the fragility of the metapopulation as a function of straying rate.
+
+Straying had non-linear impacts on the recovery time of populations. 
+When the dominant state (well adapted and high density) is wiped out, high levels of straying allow it to recover quickly (figure \ref{fig:relax}a) because the surviving population has a mean trait value skewed towards the optimum of the recovering population (figure \ref{extfig-fig:relaxtraj_hdlh}).
+Yet, as straying decreases, recovery time for the disturbed dominant state population increases, in part because there is enough time for the trait distribution to move back towards the trait optimum of the subordinate state population.
+In contrast, when the subordinate state population (maladapted and low density) is wiped out, recovery rates are fastest at low to intermediate levels of straying.
+Because the mean trait values of both populations are skewed towards those of the dominant population, when the subordinate population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into the recovering population extends the length of time required for it to return to its steady state (figure \ref{extfig-fig:relaxtraj_ldlh}).
+When both populations are both dramatically reduced, recovery time is generally fastest at lower levels of straying, while near the onset of the fold bifurcation, recovery time increases explosively and this is -- as the name implies -- characteristic of \emph{slowing} dynamics that occur near critical transitions \citep{Scheffer:2009gg,Kuehn:2010p2591}.\\
+
+\noindent{\bf (b) The effects of collective migration and density dependent straying} \\
+%Portfolio effects
+If we assume that the rate of straying is density-dependent, the probability that an individual strays $m_0$ determines the rate of straying within the population, such that $m(t)$ becomes lower as $N(t)$ increases, likely due to the effects of collective decision-making \citep{Berdahl:2016dx} (Eq.
+\ref{eq:ddm}).
+%Because our analyses concern steady state conditions, the inclusion of density-dependent straying does not have a large impact on the qualitative nature of our findings (figure \ref{fig:ddm}).
+Density dependence alters the straying rate at steady state population densities because $0 < m^* < m_0$, and this serves to rescale both the strength of the PE as well as the recovery time, but does not change the qualitative nature of our findings.
+In the alternative stable state regime, because each population exists at different steady state densities, there are likewise two alternative straying rates $(m_i^*,m_j^*)$: the higher straying rate is associated with the low-density population, and the lower straying rate is associated with the high-density population.
+We assessed metapopulation robustness across a range of $(m_i^*,m_j^*)$ values by varying the probability that an individual strays $m_0$, which is positively and linearly related to $(m_i^*,m_j^*)$.
+We find that the portfolio effects generated in systems with density-dependent straying are qualitatively similar to systems with constant straying, however there are some important quantitative differences.
+First, the PE associated with the high-density (low $m^*$) population is the same as that for a system with a constant $m$ (figure \ref{fig:thetaPE}a).
+% As such, the difference in $m^*$ between subordinate and dominant populations appears to matter less than the lower straying rate generated by the dominant population.
+As $m^*$ increases, we observe an increase in the PE than for systems with constant $m$.
+
+%Recovery times
+Density-dependent straying alters these recovery times (figure \ref{fig:relax}b). 
+First, in comparison with constant stray rates, density-dependent straying made recovery more rapid at elevated stray rates when both populations collapsed and when the subordinate population was extirpated.
+%Recovery times
+%The recovery times for systems with constant and density-dependent straying both show similar trends for a given $m$ and $m^*$ (figure \ref{fig:relax}c,d), however there are some distinct quantitative differences.
+At low straying rates, near-collapse of both populations resulted in longer than expected recovery times, whereas in the alternative stable state regime (higher $m^*$), the recovery times for different disturbance types were very similar to systems with a constant $m$ (figure \ref{fig:relax}b; note difference in x-axis scales).
+As trait heritability increased, the metapopulation always recovered more quickly if the small population was lost (figure \ref{extfig-fig:relax_highh}).
+% There was greater similarity in the recovery times for constant and density-dependent $m$ systems when trait heritability was high (figure \ref{fig:relax_highh}).
+The lower recovery time for systems with increased $m^*$ mirrors an elevated PE with higher density-dependent straying rates (figure \ref{fig:thetaPE}).
+% This higher PE at increased $m^*$ mirrors recovery times that are also lower than those observed for systems with constant $m$ (figure \ref{fig:relax}).
+In tandem, analysis of both PE and recovery time suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our minimal model, it does appear to promote robustness, particularly when the aggregate biomass is low and straying is correspondingly high.
+
+%Lowering Phenotypic diversity with increasing straying rate
+Increased rates of straying lowers phenotypic diversity ($\Delta \mu^* = |\mu_i(t)-\mu_j(t)|$, evaluated at the steady state) because both local and remote populations are increasingly homogenized.
+The loss of phenotypic diversity with increased straying is greater if trait heritability is low because traits take longer to go back to their local optima than they do when heritability is large. 
+Hence straying counters the effect of diversifying local adaptation. 
+% The loss of phenotypic diversity with increased straying is greater if trait heritability is low because the selective forces acting against trait means far from the local optima are lessened.
+% with decreased $h^2$.
+Less intuitively, we observe a discrete jump towards low phenotypic diversity as the fold bifurcation is crossed (figure \ref{fig:traitdiff}).
+Although the development of alternative stable states elevates the portfolio effect due to the variance-dampening effects of the aggregate, entering this dynamic regime also results in a substantial decline in phenotypic diversity, which may have less predictable adverse effects on the population.
+\\
+
+\noindent{\bf (c) The role of habitat heterogeneity and changing selective landscapes}\\ 
+\noindent With the onset of straying, we find that increasingly divergent trait optima generally lower $N_T$ and exaggerate $\Delta N$ (figure \ref{extfig-fig:thetadiffN}), such that the biomass distribution becomes increasingly uneven.
+%I've checked this - it's a theshold type (logistic) response Supp fig?
+The impact of habitat heterogeneity on the portfolio effect and recovery time is more complex, serving to emphasize the nonlinear relationship between rates of straying and metapopulation robustness.
+%regardless of heritability.
+%The rate of straying that gives rise to a large PE at the fold bifurcation is conditioned on the difference in optimal trait values between the local and remote populations, which we call $\Delta\theta = \sqrt{(\theta_i - \theta_j)^2}$.
+As habitat heterogeneity increases, alternative stable states appear at lower straying rates -- with the crossing of the fold bifurcation, accompanied by a peak in the PE -- whereas the magnitude of increase in the PE also increases (figure \ref{fig:thetaPE}b), reducing recovery time (figure \ref{extfig-fig:relaxtheta}).
+For increased rates of straying, greater habitat heterogeneity erodes the PE (figure \ref{fig:thetaPE}b) and increases the recovery time (figure \ref{extfig-fig:relaxtheta}).
+These results together suggest that habitat heterogeneity, as measured as the differences in trait optima between two habitats $\Delta\theta$, promotes robustness when straying rates are low, and erodes robustness when straying rates are high.
+%This means that alternative stable states can emerge with relatively little dispersal between sites (figure \ref{fig:thetaPE}), however the increased PE (and shorter recovery times).
+
+If we assume that environmental heterogeneity increases with distance between sites, or that individuals may have behaviors that minimize their chances of straying into very different habitats, the rate of straying will decline with habitat heterogeneity.
+Accordingly, $\Delta\theta$ will decrease as $m$ increases, such that there are low rates of straying between dissimilar populations and high rates of straying between similar populations.
+We find that under these conditions alternative stable states now appear for very low rates of straying $0 < m \leq 0.43$.
+As the straying rate increases and $\Delta\theta$ decreases, a single stable state emerges as the fold bifurcation is crossed, which is opposite the pattern observed when these parameters are independent.
+There are two notable dynamics that emerge following extinction of the dominant population at low rates of straying between dissimilar (high $\Delta\theta$) sites:
+(\emph{i}) above a threshold $m$ value, the dominant population recovers quickly enough that the evolving subordinate phenotype is overwhelmed by incoming strays, and it shifts back to its pre-disturbance (subordinate) state;
+(\emph{ii}) below a threshold $m$ value, there is an \emph{inversion} between subordinate and dominant states: because there is enough time and isolation for the subordinate trait mean to shift towards its local optimum, and away from that of the recovering dominant population, the dominant population becomes subordinate, and the subordinate population becomes dominant (figure \ref{extfig-fig:inertia}).
+This threshold value of $m$, below which the inversion dynamic behavior occurs, is marked by the asterisk in figure \ref{fig:mtheta}, and holds for both constant and density-dependent straying (figure \ref{extfig-fig:mthetamvm}).
+
+% At low rates of straying, there is little mixing between dissimilar populations, and this leads to the formation of alternative stable states.
+% Here we find that if the subordinate population is wiped out, the time to recovery is minimized, whereas if the dominant population is wiped out, the time to recovery is maximized; near-collapse of both results in intermediate recovery times.
+% When straying rates are low and the difference in trait optima are correspondingly high, we might assume that extinction of the dominant population would result in relatively faster recovery.
+% This would be a logical assumption because the mean trait value of the subordinate population would be skewed towards that of the dominant (larger) population, such that recruitment at the disturbed site should be high and the population should recovery more quickly.
+% Our results show, however, that \emph{this assumption is wrong:} although the mean trait value of the subordinate population \emph{is} skewed towards the optimum of the dominant population, when the latter is wiped out and the rate of straying is low, there is enough time and isolation for the subordinate trait mean to shift towards its own optimum, and away from that of the recovering dominant population (figure \ref{fig:inertia}).
+% This \emph{selective inertia} occurs until the dominant population grows large enough that the evolving subordinate phenotype is overwhelmed by incoming strays, shifting it back to its pre-disturbance (subordinate) state.
+% This effect can lead to much longer recovery times for the dominant population.
+% Importantly, if the rate of straying is below a threshold value, it can also result in a state-switch, wherein increased isolation permits the subordinate population to \emph{escape} the selective pull of the dominant population, leading to a switching in which population is in the dominant/subordinate state.
+
+% That alternative stable states appear and that PE becomes severely depressed for very low rates of straying is surprising: this means that even a small amount of mixing of populations from distant or dissimilar populations can qualitatively alter the dynamics of the metapopulation, regardless of trait heritability.
+%(for the discussion: an example of this situation may be the salmon populations during the last glacial maximum, where any mixing would be from geographically distant populations)
+
+\section{Discussion}
+
+We have shown that density-dependent straying between populations consistent with collective navigation, coupled with localized selection against donor phenotypes, has a large and nonlinear impact on dynamic properties that affect metapopulation robustness.
+We measured robustness as:
+1) the average-CV portfolio effect \citep{Anderson:2013gb,Anonymous:2015gf}, a statistical metric commonly used to assess the buffering capacity of metapopulations, and
+2) the recovery time, defined here as the time required for the aggregate metapopulation biomass $N_T$ to return to its steady state following an induced disturbance, which is mechanistically linked to persistence \citep{Ovaskainen:2002il}.
+In our minimal eco-evolutionary model of dispersal and natural selection between two populations, we show that these statistical and mechanistic descriptors of metapopulation dynamics and robustness are tightly coupled (figure \ref{fig:PE}d), which is not uncommon for diverse metrics of stability \citep{Donohue:2013iu}.
+Taken as a whole, our results point to an important role of density-dependent straying in the colonization and recovery dynamics within metapopulations, while also underscoring the risk of straying by individuals with maladaptive traits to reduce the productivity of locally adapted stock complexes.
+
+% 
+% %The bifurcation and critical slowing down
+% [This paragraph is interesting to me but could be cut] Although the portfolio effect is negatively correlated with recovery time, there is one exception: the point of transition from a single steady state to alternative stable states, which occurs at the fold bifurcation (REF).
+% At this point, the portfolio effect is maximized, but so is the recovery time (figures \ref{fig:PE}, \ref{fig:relax}).
+% This suggests that while the sharply increasing variance of the individual populations is dampened by aggregation, a large disturbance will have a much greater adverse effect due to exponentially longer recovery times.
+% Whether the change in variance, also known as critical slowing down, as the alternative stable state regime is approached could be used as an early warning signal of an oncoming phase transition is a hotly debated topic (REFS).
+% The detectability of such changes in dynamical behavior among salmonids appears to be idiosyncratic across species, though the difficulty in measuring critical slowing down may - ironically - be masked by large portfolio effects \citep{Krkosek:2014ch}.
+%
+
+%alternative stable states & critical slowing
+A salient finding from our results was that straying can lead to the emergence of alternative stable states, pushing one of the populations to high density (the \emph{dominant state}), and one to low density (the \emph{subordinate population}).
+This asymmetry in population densities also leads to an asymmetry in the mean trait values, both of which are skewed towards the local optimum of the dominant population. 
+The formation of alternative stable states in our system is a relatively common example of spatial pattern formation.
+%, only occurring above a threshold straying rate that largely depends on trait heritability and the difference in trait optima between sites.
+Pattern formation can occur as a consequence of myriad ecological processes, including habitat selection, aggregation, local environmental conditions, and/or interspecific interactions such as competition or predation, and is inexorably linked to scale \citep{Levin:1992p1472}.
+Pattern formation is also known to occur with the onset of spatially or diffusion-induced instabilities, or Turing instabilities \citep{Meinhardt:2012hw}. 
+Here, we describe an additional mechanism by which pattern formation in local densities can occur, when the onset of collective movement-mediated maladaptation leads to local differences in reproductive rates resulting in alternative stable states across sites.
+%This mechanism of pattern formation through the simultaneous interplay between ecological and evolutionary dynamics may open up new areas of research.
+While metapopulation approaches have widely recognized the importance of source and sink populations \citep{Hanski:1999vu} as well as the occurrence of alternative states \citep{Boughton:1999fa}, here we show that the coupled forces of ecological and evolutionary processes can result in similar populations diverging into alternative states where one population is large and well-adapted, while the other is smaller and maladapted.
+The appearance of these alternative states is largely dependent on the dispersal, or rate of straying, between sites.
+
+% 
+% To what extent the existence of alternative stable states promotes or diminishes the system as a whole is a matter of perspective, though with clear conservation implications.
+% At face value, the subordinate population is pushed closer to extinction (figure \ref{fig:traj}), suggesting that the system may be more fragile.
+% On the other hand, the presence of \emph{just enough} straying to cause formation of alternative stable states both increases the portfolio effect (figure \ref{fig:PE}c) and reduces the rate of return -- particularly if straying is density-dependent (figures \ref{thetaPE}a, \ref{fig:relax}b).
+% This asymmetry in population densities also leads an asymmetry in the mean trait values, skewed towards the local optimum of the dominant population, one, and this has important -- and often non-intuitive -- ramifications for metapopulation robustness.
+%, as measured by both the portfolio effect as well as recovery time.
+
+There is a `sweet spot' for straying that maximizes metapopulation robustness. 
+Results from our model reveal that the presence of just enough straying to cause formation of alternative stable states increases the portfolio effect (figure \ref{fig:PE}a). 
+Previous theoretical work has shown that increased connectivity may erode portfolio effects in herring metapopulations, where straying is also thought to be density-dependent \citep{Secor:2009ena}.
+Although high levels of dispersal in our system generally supports this finding, the interplay between dispersal and PE is more subtle when selection for local adaptations is considered, such that low to intermediate levels of density-dependent straying result in an elevated PE, thus increasing the buffering capacity of the metapopulation.
+Although PE is measured at the steady state, low to intermediate rates of straying also appear to have a beneficial effect on transient dynamics.
+When there is just enough straying to cause alternative stable states, the time to recovery following an induced disturbance declines, though -- as with the PE -- recovery time then grows if straying becomes very large (figure \ref{fig:relax}a).
+
+% %Density dependent straying
+% This themed issue is largely concerned with the notion that straying in biological systems where movement is a function of collective navigation may be density-dependent.
+% Berdahl et al.
+(REF) provided a mechanistic hypothesis for density-dependent straying where the proportion of the population that strays is less than the probability that a single individual takes the wrong turn.
+%(as illustrated by \emph{Kevin}; figure \ref{fig:xkcd}) takes the wrong turn.
+% Neither the transient nor asymptotic dynamics that we describe here qualitatively differ as a function of whether the rate of straying is constant or density-dependent, however we observe important quantitative differences that suggest density-dependent straying may play an important role in the persistence of metapopulations over evolutionary time.
+
+This themed issue formalizes the role of collective movement in the ecology of natural systems and illuminates a signature of collective navigation in animal populations on the move.
+We observed three salient results that contribute to our understanding of collective movement that combine to suggest that density-dependent straying may play an important role in the persistence of metapopulations over evolutionary time.
+First, the inclusion of density-dependent straying does not qualitatively alter either (\emph{i}) steady state or (\emph{ii}) transient dynamics of our minimal eco-evolutionary model, but effectively rescales measures of robustness to the lower straying rates that emerge as a consequence of the coupled dynamics.
+Second, compared to systems with constant dispersal, density-dependent straying appears to increase the portfolio effect across a range of straying rates (figure \ref{fig:thetaPE}a). 
+% and that at low to intermediate levels of straying, PE is high.
+Third, density-dependent straying reduces the time to recovery following disturbance, and this is particularly true in the case of near-collapse of the metapopulation (figure \ref{fig:relax}b).
+In the case of near-collapse, although both populations inherit low population densities, the mean trait values of both are skewed towards the optimum of the dominant population.
+Due to density-dependent straying, low population densities evince greater dispersal, and while this increased connectivity primarily facilities the growth of the dominant population (because the trait means are closer to the dominant optimum), because the dominant population contains the bulk of the aggregate biomass, the overall recovery time is lessened significantly (figures \ref{fig:relax}b, \ref{extfig-fig:relaxtraj_bothlh}).
+% Third, in habitats with greater heterogeneity between sites (measured by an increase in $\Delta\theta$), density-dependent straying diminished robustness at low rates of straying (marginally lower portfolio effects and substantially longer return times; figures \ref{fig:thetaPE}b, \ref{fig:relaxtheta}a,b).
+% Conversely, for increased straying rates, greater and exaggerates the positive effects when $m^*$ is high (higher portfolio effects, and shorter return times).
+
+% %Comparing recovery times
+% Previous theoretical work has shown that the largest perturbations lead to the longest recovery times \citep{Ovaskainen:2002il}.
+% In terms of relative biomass lost, the largest perturbation that we investigate is the near-collapse of both populations, where only 1\% of the pre-perturbation densities survive, however whether this disturbance results in the longest recovery time depends largely on the rate of straying and trait heritability.
+% For example, when straying and trait heritability are low (figure \ref{fig:relax}) the extinction of the larger population maximizes return times, whereas if straying is high, near-collapse maximizes return time.
+% The latter is always true in the alternative stable state regime if trait heritability is high (figure \ref{fig:relax_highh}).
+
+%Habitats 
+Salmon are distributed and stray across a diverse range of habitats, and the rates of straying between geographically diverse sites can be plastic and idiosyncratic \citep{Westley:2015to}.
+Our surrogate measure for habitat heterogeneity is the difference in trait optima between sites $\Delta\theta$.
+In general, our findings indicate that increased habitat heterogeneity promotes robustness (higher PE, shorter time to recovery) when straying rates are low, but may erode robustness when straying rates are high (figure \ref{fig:thetaPE}b, solid lines).
+This may be particularly consequential for populations that are spatially adjacent but separated by sharp environmental boundaries, such that trait optima are divergent yet dispersal is relatively high.
+Such a scenario plays out repeatedly in the context of adjacent wild and hatchery-produced salmon. 
+Although wild and hatchery populations may occur close on the landscape, and indeed often are sympatric within the same river network, the selective environments to which they are locally adapted differ dramatically \citep{Christie:2012bj}.
+Straying of domesticated hatchery-produced fish from release sites and spawning in the wild drastically reduces the productivity of wild populations through competition and outbreeding depression \citep{Chilcote:2003bb,Araki:2007cm}.
+
+%mtheta
+All else being equal, habitats that are closer in space generally have greater similarity in environmental conditions than habitats that are geographically distant, and correspondingly phenotypes of more proximately located populations should be more similar to those that are distant \citep{Westley:2012ui}.
+It is thus reasonable to expect a larger number of straying individuals between sites that are geographically proximate and indeed evidence corroborates this prediction \citep{Candy:2000hu,JPE:JPE1383}.
+Alternatively, salmon that cue to specific environmental conditions may be more likely to stray into sites that have structurally and physiognamically similar habitats \citep{Peterson:2014gy} even if other potential sites are closer.
+These considerations justify imposing a direct relationship between the rate of straying $m$ and habitat heterogeneity: as site dissimilarity increases, so too should the straying rate decrease.
+% As site dissimilarity increases, so to should the optima in trait values for the resident populations.
+When habitat heterogeneity and the rate of straying are linked in this way, we show that very small amounts of either constant or density-dependent straying result in long recovery times for the dominant population because there is time for selection to push the subordinate trait mean away from the optimum of the dominant population (figure \ref{fig:mtheta}).
+%there is enough isolation to allow the surviving subordinate population to adapt towards its own optimum.
+If the straying rate is very low, there is an inversion in the alternative stable states following the disturbance, resulting in a state shift in dominance.
+This may have particular conservation implications for considerations of aiding dispersal following disturbances or reconnections of lost habitats \citep{Anderson:2013bf,Pess:2014isa}.
+% Moreover, local temperature regimes are known to play a central role in dictating local adaptation of salmon populations.
+% If temperature is the primary determinant of $\Delta\theta$ in terms of our model, this would suggest that populations spanning North-South gradients are more predisposed to distance dependent dynamics.
+% In contrast, the relationship between the rate of straying and habitat heterogeneity may be less clear-cut for populations spanning an East-West gradient, where distant habitats may equally fitting for straying populations.
+
+% Our model is a simple framework for an initial consideration of the eco-evolutionary dynamics of metapopulation and accordingly made assumptions that are important to clarify. 
+% First, our model did not consider the role of straying and dispersal in introducing new genetic material that can combat genetic challenges of small populations such as inbreeding. 
+% Second, our model did not incorporate the potential genetic underpinnings of response diversity and asynchrony. 
+% We anticipate that including a link between genetic divergence and population stochasticity would exacerbate the rate at which higher straying would synchronize populations. 
+% Third, our model assumed that strays were incorporated into the trait distribution of the recipient populations—in natural systems it is possible that strays may die prior to reproduction when the two habitats are highly divergent.
+% Fourth, natural systems are quite temporally? variable, which could push systems out of steady-state dynamics, which is what our model focused on. 
+% Fifth, we focused on an extremely simple system of two linked populations.
+While more linked populations would introduce additional complexity, we predict that our qualitative results would apply to more complicated systems.
+% It is clear that understanding the eco-evolutionary dynamics of metapopulations is a rich area for further research.
+
+Although our study was inspired by salmon metapopulations, the results have general implications for the conservation and management of other migratory metapopulations as well. 
+Because changes in straying rates can have large and nonlinear impacts on robustness, human activities that alter straying rates could have unintended consequences. 
+For example, salmon produced by hatcheries often stray into proximate wild populations \citep{Brenner:2012gl}, and these recipient populations can have lower fitness due in part to the introduction of maladapted genes \citep{Ford:2002ip}.
+We show that there is an intermediate straying rate where disturbed populations that are recovering by the introduction of maladapted strays recover fastest: if the straying rate is too low or too high, recovery times increase (figure \ref{fig:relax}).
+This finding suggests that salmon stocking efforts that aim to lower recovery times following dam removal could actually prolong recovery if the rate at which individuals are introduced is not taken into account.
+Ongoing examinations of experimental restocking in the recently re-opened Elwha River (Washington State) will provide empirical insight into the potential short- and long-term consequences of facilitated recovery \citep{Liermann:2017gj}.
+
+% More generally, the current attributes and resilience of salmon metapopulations are likely a function of their history of evolving in dynamic river systems that periodically experienced massive disturbances such as landslides, glaciers, and floods over different climatic regimes (Waples et al.
+2009; Waples et al.
+2008).
+% Population declines following these disturbances would also magnify the rate of straying, and our results suggest that this response in the straying rate would quicken the recovery of the small populations that survive.
+% Indeed, we found that density-dependent migration appears to increase the robustness of metapopulations faced with large-scale disturbance, particularly in regimes where populations have lower steady states and correspondingly higher rates of straying.
+% As population numbers increase and stray rates decrease, the metapopulation could achieve higher production due to fine-scale local adaptation unhindered by high rates of maladaptive strays.
+
+The portfolio effect and the time to recovery following a disturbance are independent and correlated measures of metapopulation robustness that take into account both steady state and transient dynamics.
+We show that these measures of robustness are strongly influenced by the rate at which individuals from donor populations stray into habitats occupied by recipient populations. 
+Importantly, density-dependent straying, which may occur when individuals collectively navigate, can both increase the portfolio effect and lower the time to recovery following a disturbance, which is anticipated to promote persistence.
+We suggest that understanding the spatial complexity of metapopulations dispersing across heterogeneous environments, in tandem with the mosaic of selective forces acting on those environments, may be key to uncovering those factors that promote persistence in the wild.
+
+% 
+% 
+% %Concluding
+% The importance of selection and its influence on the dynamics of populations is increasingly recognized as vital for assessing species' fragility.
+% Of particular interest is how selection might influence spatially coupled metapopulations distributed across environmental gradients.
+% We have shown that two measures of metapopulation robustness -- the portfolio effect and the time to recovery following a disturbance -- are strongly influenced by the rate at which individuals from a donor population stray into habitats occupied by a recipient population.
+% Importantly, density-dependent straying, which may occur when individuals collectively navigate, can both increase the portfolio and lower the time to recovery following a disturbance, which is anticipated to promote persistence.
+% We suggest that understanding the spatial complexity of metapopulations dispersing across heterogeneous environments, combined with the mosaic of selective forces acting on those environments, may be key to discovering those factors that promote persistence.
+
+\bibliography{aa_kevin}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.4\textwidth]{figs2/fig_traj.pdf}
+\caption{
+(a) The steady state densities of $N_1$ and $N_2$ as a function of a constant stray rate $m$.
+Which population attains the low- or high-density state is random due to small applied fluctuations in the initial conditions.
+(b) The steady state trait values measured as $\theta_i - x_i$, as a function of a constant stray rate $m$. 
+FB marks the fold bifurcation.
+} \label{fig:traj}
+\end{figure}
+
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.9\textwidth]{figs2/fig_MDPE_hm.pdf}
+\caption{
+(a) Total means $N_t$, (b) difference in means $\Delta N$, and (c) the portfolio effect PE as a function of heritability $h^2$ and a constant stray rate $m$. Light colors = high values.
+The black line shows the fold bifurcation separating a single steady state (left) from alternative stable states (right).
+(d) The relationship between the time to recovery following a disturbance and the portfolio effect.
+The black line denotes the fold bifurcation.
+} \label{fig:PE}
+\end{figure*}
 
@@
 
-%Alternative stable states
-
-\noindent{\bf (a) Nonlinear effects of straying on metapopulation robustness} \\
-\noindent Regardless of density dependence, straying lowers steady state densities for both populations by (\emph{i}) the donor population losing locally-adapted individuals to the recipient population and (\emph{ii}) the introduction of maladapted individuals to the recipient population from the donor population (Fig.
-\ref{fig:traj}).
-This prediction is in accordance with observations from natural populations \cite{Bett:2017ha}. %, regardless of trait heritability and variance or habitat heterogeneity
-The decline in steady state densities is not gradual: as straying increases, the system crosses a fold bifurcation whereby the single steady state for the metapopulation bifurcates into two alternative stable states: one at high biomass, and one at low biomass density (figure \ref{fig:traj}a, \ref{fig:PE}a).
-Mean trait values for both populations bifurcate similarly (figure \ref{fig:traj}b), depending on which population attains a low- vs. high-density. 
-Above the threshold straying rate defined by the fold bifurcation, there are two alternative eco-evolutionary states: the \emph{dominant state} population will have a higher a higher density and higher degree of local adaptation (lower trait offset from the local optimum), while the \emph{subordinate state} population will have lower density with maladapted trait values (higher trait offset from the local optimum).
-Whether a specific population goes to one state or the other in our model is random, which is due to a small amount of introduced variance in the initial conditions.
-% Thus, we predict that high levels of straying in metapopulations will produce alternative eco-evolutionary states.
-
 \begin{figure*}
@@
 
-Trait heritability has a large impact on the degree to which straying rate affects both the aggregate population steady state density ($N^*_T=N^*_1+N^*_2$; figure \ref{fig:PE}a) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N=|N^*_1-N^*_2|$; figure \ref{fig:PE}b).
-Greater trait heritability results in a faster decline in $N_T^*$ with increasing straying rates $m$, but leads to only moderate changes to $\Delta N$.
-Conversely, in the context of lower trait heritability, an increase in the straying rate has little impact on the total biomass density but contrastingly large effects on $\Delta N$.
-The fold bifurcation (the black line in Figs.
-\ref{fig:PE}a-c) occurs at lower values of the straying rate $m$ with decreased trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
-Although trait heritability among salmonids is variable, most life history traits have an $h^2 <0.5$ \cite{Carlson:2008hl}, and we largely focus additional analyses on that range.
-
-%Critical slowing down
-As the fold bifurcation is approached with increasing $m$, the portfolio effect increases sharply due to an amplification in variance within both donor and recipient populations ${\rm VAR}(N_{i,j})$.
-This variance increase is the product of a dynamical process known as \emph{critical slowing down} that occurs near fold bifurcations \cite{Scheffer:2009gg}, a phenomenon that some have suggested may serve as an early warning indicator for approaching phase transitions \cite{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br}.
-For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}a-c), where alternative stable states occur, the portfolio effect declines steadily as the CV of $N_T$ increases.
-The decline over $m$ is more gradual if trait heritability is low, and steeper if trait heritability is high (figure \ref{fig:PE}c).
-
-% \noindent{\bf (c) Rates of recovery following catastrophic collapse}\\ 
-% Return time as a measure of system persistence and relation to PE
-As the portfolio effect is highly sensitive to the rate of straying between populations, so is the time required for the system to recover to a steady state following a large disturbance.
-In general, we find that the average-CV portfolio effect is negatively correlated with recovery time (figure \ref{fig:PE}d), indicating that, for our system, both measures are valuable indicators of metapopulation robustness.
-Because we can assess the time to recovery in response to the various disturbance types described above, this allows us to gain an in-depth perspective into the fragility of the metapopulation as a function of straying rate.
-
-Straying had non-linear impacts on the recovery time of populations. 
-When the dominant state (well adapted and high density) is wiped out, high levels of straying allow it to recover quickly (figure \ref{fig:relax}a) because the surviving population has a mean trait value skewed towards the optimum of the recovering population (figure \ref{fig:relaxtraj_hdlh}).
-Yet, as straying decreases, recovery time for the disturbed dominant state population increases, in part because there is enough time for the trait distribution to move back towards the trait optimum of the subordinate state population.
-In contrast, when the subordinate state population (maladapted and low density) is wiped out, recovery rates are fastest at low to intermediate levels of straying.
-Because the mean trait values of both populations are skewed towards those of the dominant population, when the subordinate population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into the recovering population extends the length of time required for it to return to its steady state (figure \ref{fig:relaxtraj_ldlh}).
-When both populations are both dramatically reduced, recovery time is generally fastest at lower levels of straying, while near the onset of the fold bifurcation, recovery time increases explosively and this is -- as the name implies -- characteristic of \emph{slowing} dynamics that occur near critical transitions \cite{Scheffer:2009gg,Kuehn:2010p2591}.\\
-
-\noindent{\bf (b) The effects of collective migration and density dependent straying} \\
-%Portfolio effects
-If we assume that the rate of straying is density-dependent, the probability that an individual strays $m_0$ determines the rate of straying within the population, such that $m(t)$ becomes lower as $N(t)$ increases, likely due to the effects of collective decision-making \cite{Berdahl:2016dx} (Eq.
-\ref{eq:ddm}).
-%Because our analyses concern steady state conditions, the inclusion of density-dependent straying does not have a large impact on the qualitative nature of our findings (figure \ref{fig:ddm}).
-Density dependence alters the straying rate at steady state population densities because $0 < m^* < m_0$, and this serves to rescale both the strength of the PE as well as the recovery time, but does not change the qualitative nature of our findings.
-In the alternative stable state regime, because each population exists at different steady state densities, there are likewise two alternative straying rates $(m_i^*,m_j^*)$: the higher straying rate is associated with the low-density population, and the lower straying rate is associated with the high-density population.
-We assessed metapopulation robustness across a range of $(m_i^*,m_j^*)$ values by varying the probability that an individual strays $m_0$, which is positively and linearly related to $(m_i^*,m_j^*)$.
-We find that the portfolio effects generated in systems with density-dependent straying are qualitatively similar to systems with constant straying, however there are some important quantitative differences.
-First, the PE associated with the high-density (low $m^*$) population is the same as that for a system with a constant $m$ (figure \ref{fig:thetaPE}a).
-% As such, the difference in $m^*$ between subordinate and dominant populations appears to matter less than the lower straying rate generated by the dominant population.
-As $m^*$ increases, we observe an increase in the PE than for systems with constant $m$.
-
-%Recovery times
-Density-dependent straying alters these recovery times (figure \ref{fig:relax}b). 
-First, in comparison with constant stray rates, density-dependent straying made recovery more rapid at elevated stray rates when both populations collapsed and when the subordinate population was extirpated.
-%Recovery times
-%The recovery times for systems with constant and density-dependent straying both show similar trends for a given $m$ and $m^*$ (figure \ref{fig:relax}c,d), however there are some distinct quantitative differences.
-At low straying rates, near-collapse of both populations resulted in longer than expected recovery times, whereas in the alternative stable state regime (higher $m^*$), the recovery times for different disturbance types were very similar to systems with a constant $m$ (figure \ref{fig:relax}b; note difference in x-axis scales).
-As trait heritability increased, the metapopulation always recovered more quickly if the small population was lost (figure \ref{fig:relax_highh}).
-% There was greater similarity in the recovery times for constant and density-dependent $m$ systems when trait heritability was high (figure \ref{fig:relax_highh}).
-The lower recovery time for systems with increased $m^*$ mirrors an elevated PE with higher density-dependent straying rates (figure \ref{fig:thetaPE}).
-% This higher PE at increased $m^*$ mirrors recovery times that are also lower than those observed for systems with constant $m$ (figure \ref{fig:relax}).
-In tandem, analysis of both PE and recovery time suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our minimal model, it does appear to promote robustness, particularly when the aggregate biomass is low and straying is correspondingly high.
-
 \begin{figure}
@@
 \end{figure}
-
-%Lowering Phenotypic diversity with increasing straying rate
-Increased rates of straying lowers phenotypic diversity ($\Delta \mu^* = |\mu_i(t)-\mu_j(t)|$, evaluated at the steady state) because both local and remote populations are increasingly homogenized.
-The loss of phenotypic diversity with increased straying is greater if trait heritability is low because traits take longer to go back to their local optima than they do when heritability is large. 
-Hence straying counters the effect of diversifying local adaptation. 
-% The loss of phenotypic diversity with increased straying is greater if trait heritability is low because the selective forces acting against trait means far from the local optima are lessened.
-% with decreased $h^2$.
-Less intuitively, we observe a discrete jump towards low phenotypic diversity as the fold bifurcation is crossed (figure \ref{fig:traitdiff}).
-Although the development of alternative stable states elevates the portfolio effect due to the variance-dampening effects of the aggregate, entering this dynamic regime also results in a substantial decline in phenotypic diversity, which may have less predictable adverse effects on the population.
-\\
-
-\noindent{\bf (c) The role of habitat heterogeneity and changing selective landscapes}\\ 
-\noindent Increasing differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater regional differences in the conditions that favor alternative trait complexes, which we interpret here as greater habitat heterogeneity (REFS).
-If both populations are isolated, natural selection will direct the mean trait values of both populations towards their respective optima, such that $\mu_i(t) \rightarrow \theta_i$ as $t\rightarrow\infty$.
-%difference in optimal trait values between the local and remote populations, which we call .
-With the onset of straying, we find that increasingly divergent trait optima generally lower $N_T$ and exaggerate $\Delta N$ (figure \ref{fig:thetadiffN}), such that the biomass distribution becomes increasingly uneven.
-%I've checked this - it's a theshold type (logistic) response Supp fig?
-The impact of habitat heterogeneity on the portfolio effect and recovery time is more complex, serving to emphasize the nonlinear relationship between rates of straying and metapopulation robustness.
-%regardless of heritability.
-%The rate of straying that gives rise to a large PE at the fold bifurcation is conditioned on the difference in optimal trait values between the local and remote populations, which we call $\Delta\theta = \sqrt{(\theta_i - \theta_j)^2}$.
-As habitat heterogeneity increases, alternative stable states appear at lower straying rates -- with the crossing of the fold bifurcation, accompanied by a peak in the PE -- whereas the magnitude of increase in the PE also increases (figure \ref{fig:thetaPE}b), reducing recovery time (figure \ref{fig:relaxtheta}).
-For increased rates of straying, greater habitat heterogeneity erodes the PE (figure \ref{fig:thetaPE}b) and increases the recovery time (figure \ref{fig:relaxtheta}).
-These results together suggest that habitat heterogeneity, as measured as the differences in trait optima between two habitats $\Delta\theta$, promotes robustness when straying rates are low, and erodes robustness when straying rates are high.
-%This means that alternative stable states can emerge with relatively little dispersal between sites (figure \ref{fig:thetaPE}), however the increased PE (and shorter recovery times).
-
-If we assume that environmental heterogeneity increases with distance between sites, or that individuals may have behaviors that minimize their chances of straying into very different habitats, the rate of straying will decline with habitat heterogeneity.
-Accordingly, $\Delta\theta$ will decrease as $m$ increases, such that there are low rates of straying between dissimilar populations and high rates of straying between similar populations.
-We find that under these conditions alternative stable states now appear for very low rates of straying $0 < m \leq 0.43$.
-As the straying rate increases and $\Delta\theta$ decreases, a single stable state emerges as the fold bifurcation is crossed, which is opposite the pattern observed when these parameters are independent.
-There are two notable dynamics that emerge following extinction of the dominant population at low rates of straying between dissimilar (high $\Delta\theta$) sites:
-(\emph{i}) above a threshold $m$ value, the dominant population recovers quickly enough that the evolving subordinate phenotype is overwhelmed by incoming strays, and it shifts back to its pre-disturbance (subordinate) state;
-(\emph{ii}) below a threshold $m$ value, there is enough time and isolation for the subordinate trait mean to shift towards its local optimum, and away from that of the recovering dominant population, such that the dominant population becomes subordinate, and the subordinate population becomes dominant (figure \ref{fig:inertia}).
-This threshold value of $m$, below which this behavior occurs, is marked by the asterisk in figure \ref{fig:mtheta}, and holds for both constant and density-dependent straying (figure \ref{fig:mthetamvm}).
-
-% At low rates of straying, there is little mixing between dissimilar populations, and this leads to the formation of alternative stable states.
-% Here we find that if the subordinate population is wiped out, the time to recovery is minimized, whereas if the dominant population is wiped out, the time to recovery is maximized; near-collapse of both results in intermediate recovery times.
-% When straying rates are low and the difference in trait optima are correspondingly high, we might assume that extinction of the dominant population would result in relatively faster recovery.
-% This would be a logical assumption because the mean trait value of the subordinate population would be skewed towards that of the dominant (larger) population, such that recruitment at the disturbed site should be high and the population should recovery more quickly.
-% Our results show, however, that \emph{this assumption is wrong:} although the mean trait value of the subordinate population \emph{is} skewed towards the optimum of the dominant population, when the latter is wiped out and the rate of straying is low, there is enough time and isolation for the subordinate trait mean to shift towards its own optimum, and away from that of the recovering dominant population (figure \ref{fig:inertia}).
-% This \emph{selective inertia} occurs until the dominant population grows large enough that the evolving subordinate phenotype is overwhelmed by incoming strays, shifting it back to its pre-disturbance (subordinate) state.
-% This effect can lead to much longer recovery times for the dominant population.
-% Importantly, if the rate of straying is below a threshold value, it can also result in a state-switch, wherein increased isolation permits the subordinate population to \emph{escape} the selective pull of the dominant population, leading to a switching in which population is in the dominant/subordinate state.
 
@@
 
-% That alternative stable states appear and that PE becomes severely depressed for very low rates of straying is surprising: this means that even a small amount of mixing of populations from distant or dissimilar populations can qualitatively alter the dynamics of the metapopulation, regardless of trait heritability.
-%(for the discussion: an example of this situation may be the salmon populations during the last glacial maximum, where any mixing would be from geographically distant populations)
-
-\section{Discussion}
-
-We have shown that density-dependent straying between populations consistent with collective navigation, coupled with localized selection against donor phenotypes, has a large and nonlinear impact on dynamic properties directly impacting the robustness of metapopulations against external disturbances.
-Specifically, we observed robustness in terms of: 
-1) the average-CV portfolio effect \cite{Anderson:2013gb,Anonymous:2015gf}, a statistical metric commonly used to assess the buffering capacity of metapopulations, and
-2) the recovery time, defined here as the time required for the aggregate metapopulation biomass $N_T$ to return to its steady state following an induced disturbance, which is mechanistically linked to persistence \cite{Ovaskainen:2002il}.
-In our minimal eco-evolutionary model of dispersal and natural selection between two populations, we show that these statistical and mechanistic descriptors of metapopulation dynamics and robustness are tightly coupled (figure \ref{fig:PE}d), which is not uncommon among different metrics of stability \cite{Donohue:2013iu}.
-Taken as a whole, we interpret our results to suggest an important role of density-dependent straying in the colonization and recovery dynamics for populations in shifting river mosaics, while also underscoring the risk of straying by individuals with maladaptive traits to reduce the productivity of locally adapted stock complexes.
-
 % 
-% %The bifurcation and critical slowing down
-% [This paragraph is interesting to me but could be cut] Although the portfolio effect is negatively correlated with recovery time, there is one exception: the point of transition from a single steady state to alternative stable states, which occurs at the fold bifurcation (REF).
-% At this point, the portfolio effect is maximized, but so is the recovery time (figures \ref{fig:PE}, \ref{fig:relax}).
-% This suggests that while the sharply increasing variance of the individual populations is dampened by aggregation, a large disturbance will have a much greater adverse effect due to exponentially longer recovery times.
-% Whether the change in variance, also known as critical slowing down, as the alternative stable state regime is approached could be used as an early warning signal of an oncoming phase transition is a hotly debated topic (REFS).
-% The detectability of such changes in dynamical behavior among salmonids appears to be idiosyncratic across species, though the difficulty in measuring critical slowing down may - ironically - be masked by large portfolio effects \cite{Krkosek:2014ch}.
 % 
-
-%alternative stable states & critical slowing
-A salient finding from our results was that straying can push metapopulations into alternative stable states, pushing one of the populations to a high density population (the \emph{dominant state}), and one to a low density population (the \emph{subordinate population}).
-This asymmetry in population densities also leads to an asymmetry in the mean trait values, both of which are skewed towards the local optimum of the dominant population. 
-The formation of alternative stable states in our system is a relatively common example of spatial pattern formation, which only occurs above a threshold straying rate, the value of which depends on trait heritability and the difference in trait optima between sites.
-Pattern formation can occur as a consequence of myriad ecological processes, including habitat selection, aggregation, local environmental conditions, and/or interspecific interactions such as competition or predation, and is inexorably linked to scale \cite{Levin:1992p1472}.
-Additionally, pattern formation is also well known to occur with the onset of spatially or diffusion-induced instabilities, or Turing instabilities \cite{Meinhardt:2012hw}. 
-Here, we describe an additional mechanism through which pattern formation in local densities can occur, when the onset of collective movement-mediated maladaptation leads to local differences in reproductive rates resulting in alternative stable states across sites.
-%This mechanism of pattern formation through the simultaneous interplay between ecological and evolutionary dynamics may open up new areas of research.
-While metapopulation approaches have widely recognized the importance of source and sink populations \cite{Hanski:1999vu} and have observed formation of alternative states \cite{Boughton:1999fa}, here we show that even similar populations can evolve into divergent states where one population is large and the other is smaller and maladapted.
-
+% \beginsupplement
 % 
-% To what extent the existence of alternative stable states promotes or diminishes the system as a whole is a matter of perspective, though with clear conservation implications.
-% At face value, the subordinate population is pushed closer to extinction (figure \ref{fig:traj}), suggesting that the system may be more fragile.
-% On the other hand, the presence of \emph{just enough} straying to cause formation of alternative stable states both increases the portfolio effect (figure \ref{fig:PE}c) and reduces the rate of return -- particularly if straying is density-dependent (figures \ref{thetaPE}a, \ref{fig:relax}b).
-% This asymmetry in population densities also leads an asymmetry in the mean trait values, skewed towards the local optimum of the dominant population, one, and this has important -- and often non-intuitive -- ramifications for metapopulation robustness.
-%, as measured by both the portfolio effect as well as recovery time.
-
-There is a `sweet spot' for straying that maximizes metapopulation robustness. 
-Results from our model reveal that the presence of just enough straying to cause formation of alternative stable states increases the portfolio effect (figure \ref{fig:PE}a). 
-Previous theoretical work has shown that increased connectivity may erode portfolio effects in herring metapopulations, where straying is also thought to be density-dependent \cite{Secor:2009ena}.
-Although high levels of dispersal in our system shows a similar trend, the interplay between dispersal and PE is more subtle when selection for local adaptations is considered, such that low to intermediate levels of density-dependent straying result in an elevated PE, increasing the buffering capacity of the metapopulation.
-Although PE is measured at the steady state, low to intermediate rates of straying also appear to have a beneficial effect on transient dynamics.
-When there is just enough straying to cause alternative stable states, recovery times following induced disturbances decline, though -- as with the PE -- the time to recovery then grows if straying becomes too large (figure \ref{fig:relax}a).
-
-% %Density dependent straying
-% This themed issue is largely concerned with the notion that straying in biological systems where movement is a function of collective navigation may be density-dependent.
-% Berdahl et al.
-(REF) provided a mechanistic hypothesis for density-dependent straying where the proportion of the population that strays is less than the probability that a single individual takes the wrong turn.
-%(as illustrated by \emph{Kevin}; figure \ref{fig:xkcd}) takes the wrong turn.
-% Neither the transient nor asymptotic dynamics that we describe here qualitatively differ as a function of whether the rate of straying is constant or density-dependent, however we observe important quantitative differences that suggest density-dependent straying may play an important role in the persistence of metapopulations over evolutionary time.
-
-This themed issue formalizes the role of collective movement in the ecology of natural systems and illuminates a signature of collective navigation in animal populations on the move.
-We observed three salient results that contribute to our understanding of collective movement that combine to suggest that density-dependent straying may play an important role in the persistence of metapopulations over evolutionary time.
-First, the inclusion of density-dependent straying does not qualitatively alter either (\emph{i}) steady state or (\emph{ii}) transient dynamics of our minimal eco-evolutionary model, but effectively rescales measures of robustness to the straying rates the emerge from the coupled dynamics.
-Second, compared to systems with constant dispersal, density-dependent straying appears to increase the portfolio effect across a range of straying rates (figure \ref{fig:thetaPE}a). 
-% and that at low to intermediate levels of straying, PE is high.
-Third, density-dependent straying reduces the time to recovery following disturbance, and this is particularly true in the case of near-collapse of the metapopulation (figure \ref{fig:relax}b).
-In the case of near-collapse, although both populations inherit low population densities, the mean trait values of both are skewed towards the optimum of the dominant population.
-Due to density-dependent straying, low population densities evince greater dispersal, and while this increased connectivity primarily facilities the growth of the dominant population (because the trait means are closer to the dominant optimum), because the dominant population is the bulk of the aggregate biomass, the overall recovery time is significantly lessened (figures \ref{fig:relax}b, \ref{fig:relaxtraj_bothlh}).
-% Third, in habitats with greater heterogeneity between sites (measured by an increase in $\Delta\theta$), density-dependent straying diminished robustness at low rates of straying (marginally lower portfolio effects and substantially longer return times; figures \ref{fig:thetaPE}b, \ref{fig:relaxtheta}a,b).
-% Conversely, for increased straying rates, greater and exaggerates the positive effects when $m^*$ is high (higher portfolio effects, and shorter return times).
-
-% %Comparing recovery times
-% Previous theoretical work has shown that the largest perturbations lead to the longest recovery times \cite{Ovaskainen:2002il}.
-% In terms of relative biomass lost, the largest perturbation that we investigate is the near-collapse of both populations, where only 1\% of the pre-perturbation densities survive, however whether this disturbance results in the longest recovery time depends largely on the rate of straying and trait heritability.
-% For example, when straying and trait heritability are low (figure \ref{fig:relax}) the extinction of the larger population maximizes return times, whereas if straying is high, near-collapse maximizes return time.
-% The latter is always true in the alternative stable state regime if trait heritability is high (figure \ref{fig:relax_highh}).
-
-%Habitats 
-Salmon are distributed and stray across a diverse range of habitats, and the rates of straying between geographically diverse sites can be plastic and idiosyncratic \cite{Westley:2015to}.
-Our surrogate measure for habitat heterogeneity is the difference in trait optima between sites $\Delta\theta$.
-In general, our findings indicate that increased habitat heterogeneity promotes robustness (higher PE, shorter time to recovery) when straying rates are low, but may erode robustness when straying rates are high (figure \ref{fig:thetaPE}b, solid lines).
-This may be particularly consequential for populations that are spatially adjacent but separated by sharp environmental boundaries, such that trait optima are divergent yet dispersal is relatively high.
-Such a scenario plays out repeatedly in the context of adjacent wild and hatchery-produce salmon.
-Although wild and hatchery populations may occur close on the landscape, and indeed often are sympatric within the same river network, the selective environments to which they are locally adapted differ dramatically \cite{Christie:2012bj}.
-Straying of domesticated hatchery-produced fish from release sites and spawning in the wild drastically reduces the productivity of wild populations through competition and outbreeding depression \cite{Chilcote:2003bb,Araki:2007cm}.
-
-%mtheta
-All else being equal, habitats that are closer in space generally have greater similarity in environmental conditions than habitats that are geographically distant, and correspondingly phenotypes of more proximately located populations should be more similar to those that are distant \cite{Westley:2012ui}.
-Correspondingly, it is reasonable to expect a larger number of straying individuals between sites that are geographically proximate and indeed evidence corroborates this prediction \cite{Candy:2000hu,JPE:JPE1383}.
-Alternatively, salmon that cue to specific environmental conditions may be more likely to stray into sites that have structurally and physiognamically similar habitats \cite{Peterson:2014gy} even if other potential sites are closer.
-These considerations justify imposing a direct relationship between the rate of straying $m$ and habitat heterogeneity: as site dissimilarity increases, so too should the straying rate decrease.
-As site dissimilarity increases, so to should the optima in trait values for the resident populations.
-
-When habitat heterogeneity and the rate of straying are linked, we show that small amounts of either constant or density-dependent straying result in longer recovery times for the dominant population because there is enough isolation to allow the surviving subordinate population to slowly adapt towards its own optimum and therefore strays from that population will be maladaptive in the dominant site.
-As selection pushes the subordinate population away from the dominant trait optimum, the recovery of the extinct population is slowed (figure \ref{fig:mtheta}).
-This may have particular conservation implications for considerations of aiding dispersal following disturbances or reconnections of lost habitats \cite{Anderson:2013bf,Pess:2014isa}.
-Moreover, local temperature regimes are known to play a central role in dictating local adaptation of salmon populations (REFS).
-If temperature is the primary determinant of $\Delta\theta$ in terms of our model, this would suggest that populations spanning North-South gradients are more predisposed to distance dependent dynamics.
-In contrast, the relationship between the rate of straying and habitat heterogeneity may be less clear-cut for populations spanning an East-West gradient, where distant habitats may equally fitting for straying populations.
-
-%I'm here
-
-%Concluding
-The importance of selection and its influence on the dynamics of populations is increasingly recognized as vital for assessing species' fragility.
-Of particular interest is how selection might influence spatially coupled metapopulations distributed across environmental gradients.
-We have shown that two measures of metapopulation robustness -- the portfolio effect and the time to recovery following a disturbance -- are strongly influenced by the rate at which individuals from a donor population stray into habitats occupied by a recipient population.
-Importantly, density-dependent straying, which may occur when individuals collectively navigate, can both increase the portfolio and lower the time to recovery following a disturbance, which is anticipated to promote persistence.
-We suggest that understanding the spatial complexity of metapopulations dispersing across heterogeneous environments, combined with the mosaic of selective forces acting on those environments, may be key to discovering those factors that promote persistence.
-
-\bibliography{aa_kevin}
-
-\beginsupplement
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.35\textwidth]{figs2/fig_relax_small.pdf}
-\caption{
-Extinction of low-density population with a high constant straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
-Black line marks the calculated point of recovery post-perturbation.
-Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
-} \label{fig:relaxtraj_ldlh}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.35\textwidth]{figs2/fig_relax_large.pdf}
-\caption{
-Extinction of high-density population with a high straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
-Black line marks the calculated point of recovery post-perturbation.
-Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
-} \label{fig:relaxtraj_hdlh}
-\end{figure}
 % 
@@
 % \centering
-% \includegraphics[width=0.35\textwidth]{figs2/fig_relax_small_highh.pdf}
+% \includegraphics[width=0.35\textwidth]{figs2/fig_relax_small.pdf}
 % \caption{
-% Extinction of low-density population with a high constant straying rate $m=0.4$ and high trait heritability $h^2=0.8$ (see figure \ref{fig:relax}a).
+% Extinction of low-density population with a high constant straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
 % Black line marks the calculated point of recovery post-perturbation.
 % Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
-% } \label{fig:relaxtraj_ldhh}
+% } \label{fig:relaxtraj_ldlh}
 % \end{figure}
@@
 % \centering
-% \includegraphics[width=0.35\textwidth]{figs2/fig_relax_large_highh.pdf}
+% \includegraphics[width=0.35\textwidth]{figs2/fig_relax_large.pdf}
 % \caption{
-% Extinction of high-density population with a high straying rate $m=0.4$ and high trait heritability $h^2=0.8$ (see figure \ref{fig:relax}a).
+% Extinction of high-density population with a high straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
 % Black line marks the calculated point of recovery post-perturbation.
 % Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
-% } \label{fig:relaxtraj_hdhh}
+% } \label{fig:relaxtraj_hdlh}
 % \end{figure}
-
-\begin{figure*}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.9\textwidth]{figs2/fig_relax_highh.pdf}
-\caption{
-Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.8$.
-If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
-} \label{fig:relax_highh}
-\end{figure*}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.4\textwidth]{figs2/fig_thetadiffN.pdf}
-\caption{
-Median difference in population densities taken over the straying rate as a function of habitat heterogeneity $\Delta\theta$.
-Solid lines are for constant $m$; dashed lines are for density-dependent $m$} \label{fig:thetadiffN}
-\end{figure}
-
-\begin{figure*}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.8\textwidth]{figs2/fig_relaxtheta2.pdf}
-\caption{
-(a) Recovery time after near collapse of both populations as a function of straying rate $m$ and habitat heterogeneity $\Delta\theta$.
-(b) The same as (a) but including recovery times when straying is density-dependent, shown by linked pairs of points.
-Recovery times for systems with density-dependent straying are longer at low straying rates and shorter at higher straying rates, mirroring the change in portfolio effects shown in figure \ref{fig:thetaPE}.
-} \label{fig:relaxtheta}
-\end{figure*}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-  \centering
-  \includegraphics[width=0.35\textwidth]{figs2/fig_MDPE_hm_mtheta_rt.pdf}
-  \caption{
-  Distance dependent portfolio effects as a function of straying rate $m$ and trait heritability $h^2$. When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases.
-  } \label{fig:mthetaPE}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.35\textwidth]{figs2/fig_relax_inertia.pdf}
-\caption{
-Distance dependent straying, where increased differences in trait optima between sites $\Delta\theta$ corresponds to lower rates of straying $m$.
-At low rates of straying $m=0.02$ ($\Delta\theta=24$), extinction of the dominant population leads to slower-than-expected recovery times because the subordinate population is isolated enough to evolve towards its own trait optimum.
-%until growth of the dominant population overwhelms this local selection.
-In this case, $m$ is less than $m=0.034$ (denoted by the asterisk in figure \ref{fig:mtheta}), such that isolation allows the subdominant population to `run away' from the influence of the dominant population, leading to a switch in states.
-If $m$ is low but greater than $0.034$, isolation permits the subdominant population to `run away' from the influence of the dominant population, until it is overwhelmed by the recovering dominant population, and reverts back to its previous trait mean prior to the disturbance.
-} \label{fig:inertia}
-\end{figure}
-
-\begin{figure*}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-  \centering
-  \includegraphics[width=0.8\textwidth]{figs2/fig_relax_mtheta.pdf}
-  \caption{
-  Distance dependent recovery times for three disturbance types.
-When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases for constant (a) and density-dependent (b) staying rates.
-  The fold bifurcation is not as clear in (b) because $\Delta\theta$ is a function of the individual straying rate $m_0$, whereas the x-axis in (b) is the straying rate at the steady state $m^*$.
-  Despite this difference, the general trends shown in (a) are also present in (b).
-  } \label{fig:mthetamvm}
-\end{figure*}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.35\textwidth]{figs2/fig_relax_both_lowh.pdf}
-\caption{
-Near collapse of both populations with a low straying rate $m=0.1$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
-Black line marks the calculated point of recovery post-perturbation.
-Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
-} \label{fig:relaxtraj_bothlh}
-\end{figure}
-
-% \begin{figure*}[h!]
+% 
+% 
+% \begin{figure*}
 %   \captionsetup{justification=raggedright,
@@
 % \centering
-% \begin{subfigure}[t]{0.55\textwidth}
-% \centering
-% \includegraphics[width=\textwidth]{figs2/fig_MDPE_hm_theta3.pdf} 
-% \caption{Low habitat heterogeneity ($\Delta\theta=3$)} \label{fig:thetadiff1}
-% \end{subfigure}
-% \begin{subfigure}[t]{0.55\textwidth}
-% \centering
-% \includegraphics[width=\textwidth]{figs2/fig_MDPE_hm_theta8.pdf} 
-% \caption{High habitat heterogeneity ($\Delta\theta=8$)} \label{fig:thetadiff2}
-% \end{subfigure}
-% \caption{Total means $N_t$, difference in means $\Delta N$, and the portfolio effect PE for different habitat heterogeneities $\Delta\theta$. Light colors = high values.
-% }
+% \includegraphics[width=0.9\textwidth]{figs2/fig_relax_highh.pdf}
+% \caption{
+% Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.8$.
+% If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
+% } \label{fig:relax_highh}
 % \end{figure*}
-
-% \begin{figure*}[h!]
+% 
+% 
+% 
+% \begin{figure}
 %   \captionsetup{justification=raggedright,
@@
 % \centering
-% \includegraphics[width=0.8\textwidth]{figs2/fig_MDPE_hm_ddm.pdf}
+% \includegraphics[width=0.4\textwidth]{figs2/fig_thetadiffN.pdf}
 % \caption{
-% The same simulations as presented in Figure \ref{fig:PE}, except with density-dependent straying, where $m(t) = m_0\left(1-N(t)/(C+N(t)\right)$. Light colors = high values.
-% } \label{fig:ddm}
+% Median difference in population densities taken over the straying rate as a function of habitat heterogeneity $\Delta\theta$.
+% Solid lines are for constant $m$; dashed lines are for density-dependent $m$} \label{fig:thetadiffN}
+% \end{figure}
+% 
+% 
+% 
+% \begin{figure*}
+%   \captionsetup{justification=raggedright,
+% singlelinecheck=false
+% }
+% \centering
+% \includegraphics[width=0.8\textwidth]{figs2/fig_relaxtheta2.pdf}
+% \caption{
+% (a) Recovery time after near collapse of both populations as a function of straying rate $m$ and habitat heterogeneity $\Delta\theta$.
+% (b) The same as (a) but including recovery times when straying is density-dependent, shown by linked pairs of points.
+% Recovery times for systems with density-dependent straying are longer at low straying rates and shorter at higher straying rates, mirroring the change in portfolio effects shown in figure \ref{fig:thetaPE}.
+% } \label{fig:relaxtheta}
 % \end{figure*}
+% 
+% 
+% \begin{figure}
+%   \captionsetup{justification=raggedright,
+% singlelinecheck=false
+% }
+%   \centering
+%   \includegraphics[width=0.35\textwidth]{figs2/fig_MDPE_hm_mtheta_rt.pdf}
+%   \caption{
+%   Distance dependent portfolio effects as a function of straying rate $m$ and trait heritability $h^2$. When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases.
+%   } \label{fig:mthetaPE}
+% \end{figure}
+% 
+% \begin{figure}
+%   \captionsetup{justification=raggedright,
+% singlelinecheck=false
+% }
+% \centering
+% \includegraphics[width=0.35\textwidth]{figs2/fig_relax_inertia.pdf}
+% \caption{
+% Distance dependent straying, where increased differences in trait optima between sites $\Delta\theta$ corresponds to lower rates of straying $m$.
+% At low rates of straying $m=0.02$ ($\Delta\theta=24$), extinction of the dominant population leads to slower-than-expected recovery times because the subordinate population is isolated enough to evolve towards its own trait optimum.
+%until growth of the dominant population overwhelms this local selection.
+% In this case, $m$ is less than $m=0.034$ (denoted by the asterisk in figure \ref{fig:mtheta}), such that isolation allows the subdominant population to `run away' from the influence of the dominant population, leading to a switch in states.
+% If $m$ is low but greater than $0.034$, isolation permits the subdominant population to `run away' from the influence of the dominant population, until it is overwhelmed by the recovering dominant population, and reverts back to its previous trait mean prior to the disturbance.
+% } \label{fig:inertia}
+% \end{figure}
+% 
+% \begin{figure*}
+%   \captionsetup{justification=raggedright,
+% singlelinecheck=false
+% }
+%   \centering
+%   \includegraphics[width=0.8\textwidth]{figs2/fig_relax_mtheta.pdf}
+%   \caption{
+%   Distance dependent recovery times for three disturbance types.
+When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases for constant (a) and density-dependent (b) staying rates.
+%   The fold bifurcation is not as clear in (b) because $\Delta\theta$ is a function of the individual straying rate $m_0$, whereas the x-axis in (b) is the straying rate at the steady state $m^*$.
+%   Despite this difference, the general trends shown in (a) are also present in (b).
+%   } \label{fig:mthetamvm}
+% \end{figure*}
+% 
+% 
+% 
+% \begin{figure}
+%   \captionsetup{justification=raggedright,
+% singlelinecheck=false
+% }
+% \centering
+% \includegraphics[width=0.35\textwidth]{figs2/fig_relax_both_lowh.pdf}
+% \caption{
+% Near collapse of both populations with a low straying rate $m=0.1$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
+% Black line marks the calculated point of recovery post-perturbation.
+% Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
+% } \label{fig:relaxtraj_bothlh}
+% \end{figure}
+%
 

### 2017-06-02 manuscript/supp.tex
final tweaks on figs

--- 2017-05-31 manuscript/msv3.tex
@@
 \captionsetup{compatibility=false}
-
+\usepackage{xr}
+\externaldocument[extfig-]{msv4}
 %\usepackage{epstopdf}
@@
 
-\bibliographystyle{ieeetr}
+\linespread{1.5}
 
-\usepackage{bibunits}
+\usepackage{natbib}
+\bibpunct{(}{)}{,}{a}{}{;}
+
+\bibliographystyle{prsb}
+
+% \usepackage{bibunits}
 
@@
 
+% \Blindtext
 %\title{Simple rules yield complex communities: deconstructed species interactions and the assembly of communities}
 %\title{Community assembly and dynamics by the deconstruction of species interactions}
-\title{Eco-evolutionary dynamics and collective straying: implications for salmon metapopulation robustness}
-\author{Justin D. Yeakel${}^{1,2,*}$, Jean P. Gibert${}^{1}$, Peter A. H. Westley${}^{3}$, \& Jonathan W. Moore${}^{4}$ \\
+\title{Supplementary Figures: Eco-evolutionary dynamics and collective straying: implications for salmon metapopulation robustness}
+\author{
+Justin D. Yeakel${}^{1,2,*}$, Jean P. Gibert${}^{1}$, Peter A. H. Westley${}^{3}$, \& Jonathan W. Moore${}^{4}$ \\
 ${}^1$School of Natural Sciences, University of California, Merced, Merced CA, USA \\
@@
 \maketitle
-
-\section*{Abstract}
-{\small
-We show that taking straying into account leads to alternative stable states in population densities and trait values, which has consequences for maladaptation, intraspecific trait variability, and long-term robustness of salmon metapopulations.
-}
-\section{Introduction}
-
-Intraspecific diversity can increase the resilience and stability of species or metapopulations. 
-This diversity-stability linkage can occur when there are asynchronous population dynamics, where the changes in population vary across a metapopulation. 
-This asynchrony will increase the potential for demographic rescue \cite{Brown:1977gk,Earn:2000fm} and also decrease the variability of processes that integrate across the metapopulation \cite{Anonymous:2015gf}.
-For example, different responses to climate variability within populations of a rare plant reduced fluctuations in abundance \cite{Abbott:2017hl}. 
-This statistical buffer has traditionally been quantified as the Portfolio Effect (PE), which is the ratio of the population CV to the CV of the aggregated metapopulation \cite{Thibaut:2012km}. 
-Strengthened portfolio effects are expected to increase the robustness of metapopulations to external disturbances, and by extension promote persistence \cite{Thibaut:2012km}.
-In contrast, homogenization of populations leading to greater synchronization and weakened PE may be a harbinger of metapopulation collapse and extinction.
-
-It is clear that dispersal (i.e., movement with consequences for gene flow) can influence the evolutionary dynamics of the metapopulation, the interplay between selection, population dynamics, and persistence is not well understood.
-Movement of individuals among local populations (i.e. dispersal) can have a large influence on metapopulation persistence \cite{MilnerGulland:2011vm}. 
-Dispersal facilitates evolutionary rescue, whereby immigration of individuals with heritable adapative traits can rescue small populations from local extinction in the context of maladaptive environmental change \cite{Bell:2011ki,Carlson:2014is}.
-% First, some level of straying is necessary to enable the rescue effect, whereby dispersal among populations can rescue small populations from local extinction (REFS). 
-On the other hand, high rates of dispersal may synchronize populations and actually increase the risk of extinction of the entire metapopulation \cite{Earn:2000fm}. 
-Third, straying will influence the evolutionary dynamics of the metapopulation.
-% though how the interplay between selection and population dynamics might impact persistence is less well understood. 
-Although the dispersal of individuals into sites hosting other populations provides connections within the larger metapopulation, potentially promoting demographic and evolutionary rescue, it may also introduce maladapted individuals into habitats that are host to different environmental conditions, possibly lowering the mean fitness of the recipient population \cite{Muhlfeld:2014hs}.
-More broadly, dispersal can provide a mechanism by which phenotypes are sorted in space rather than time and facilitates the spread of maladaptive genes \cite{Lowe:2015ft}.
-Further, this straying may lead to genetic homogenization that erodes the asynchrony that underpins portfolio effects and metapopulation persistence. 
-Thus, straying can influence the resilience and robustness of metapopulations through both ecological and evolutionary processes.
-% Thus, the dual nature of straying as both promoter of connections among metapopulation demes and potential eroder of locally adapted gene complexes highlights the interplay between ecological dynamics of connected populations and the evolutionary dynamics of mixed trait distributions that respond to heterogeneous local conditions.
-
-%Movement through space (link back to straying)
-% Migratory populations that return to a breeding ground or natal stream to reproduce are linked to each other by some proportion of the population that permanently disperse, or stray into the `wrong' site.
-%; we might say that there is at least one \emph{Kevin} in every school or flock (figure \ref{fig:xkcd}).
-There is growing appreciation that a combination of abiotic, biotic, and anthropogenic factors can control the rate of dispersal among populations \cite{H:2013fs,Keefer:2014gg,Bett:2017ha}.
-Migratory populations that return to breeding sites for reproduction are linked to each other by some proportion of the population that permanently disperses into the `wrong' site. 
-Recently, the role of social interactions and collective navigation has been hypothesized [This issue].
-The rate at which individuals stray, $m$, may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \cite{Simons:2004jo,Berdahl:2015kv,Berdahl:2016dx}.
-The potential influence of collective dispersal on the dynamics of individual populations and the metapopulation as a whole is a topic of considerable interest that has tangible conservation implications \cite{Brenner:2012gl,Johnson:2012fe,Fullerton:2011ii}.
-% Whether and to what extent the ecological consequences of straying depend on the evolutionary dynamics that emerge from populations distributed across a selective gradient is unclear.
-% How the assumed negative evolutionary effects of straying and subsequent gene flow is balanced by the positive effects of demographic rescue is the subject of this contribution.
-
-%Eco-evolutionary dynamics in space
-%% From geographic mosaic of coevolution to...
-% That evolutionary forces of selection and gene flow play out heterogeneously across geographic mosaics is now a foundational concept in ecology and evolutionary biology \cite{Nuismer:1999ko,Thompson:2015hq,Nuismer:2000bb,Thompson:2005wf}.
-% These mosaics are in part driven by environmental differences between habitats that alter the selective forces acting on different phenotypes \cite{Endler:1986tz}, and a principle underlying assumption is that there is gene flow such that individuals from different habitats mix over space \cite{Gomulkiewicz:2000bz,Thompson:2002hr,Nuismer:2003eh,Nuismer:2006be,Forde:2008kc,Guimaraes:2011hu,Gibert:2013hp}.
-% Although the evolutionary outcomes of these spatial processes have been explored in depth (REFS), it is less well understood how selective mosaics and their consequent evolutionary forces impact population dynamics as they unfold \cite{Hendry:2016un}.
-
-%Motivation: straying
-The eco-evolutionary impacts of dispersal likely has important implications for conservation and management in key taxa such as in migratory salmon.
-While anadromous salmonid fishes (genera \emph{Oncorhynchus} and \emph{Salmo}) are renown for returning to their natal spawning habitats with high accuracy and precision after years at sea \cite{Quinn:2011tf,Jonsson:2011kg,Keefer:2014gg}, there are generally some individuals that `stray' to non-natal sites to spawn \cite{Quinn:1993ge,Hendry:2004wf}.
-Salmon may operate as metapopulations, where populations are genetically distinct but linked by some level of straying \cite{Schtickzelle:2007wb,Anderson:2014cx}.
-Although extensive work has been done to document the extent of straying from donor populations and into recipient populations \cite{Keefer:2014gg,Bett:2017ha}, only recently have the abiotic, biotic, and anthropogenic influences of `straying' behaviors been investigated systemically \cite{Keefer:2008bs,Westley:2015to,Bond:2016dz}.
-Straying among salmon may be influenced by environmental factors such as water temperature, human activities such as hatchery practices, and population density as predicted by the collective migration hypothesis \cite{Peterson:2014gy}.
-These strays can introduce new maladaptive genotypes into the recipient population.
-Further, straying and genetic homogenization could synchronize population dynamics and erode portfolio effects \cite{Moore:2010gs,Carlson:2011ce,Braun:2016ib}.
-Thus, there is opportunity and a need to consider the eco-evolutionary consequences of straying for metapopulations in species of conservation and management concern such as salmon.
-
-% Asynchrony among salmon populations also increases the stability of fisheries that integrate across this diversity \cite{Schindler:2010he,Anonymous:2016jc}.
-% However, these portfolio effects may be eroded if asynchrony is influenced by genetic diversity and this genetic diversity is lost. 
-% Braun et al. \cite{Braun:2016ib} found that genetic differences among populations of Chinook salmon was correlated with the degree of asynchrony in population dynamics. 
-% Artificial propagation of salmon by hatcheries and their strays may erode local genetic diversity, asynchrony among populations, and portfolio effects \cite{Carlson:2008hl,Moore:2010gs,Anonymous:2014ku}.
-%Coordinated mass migrations are one of the great wonders of the natural world, and the ability of individuals and groups to navigate across great distances have long fascinated naturalists \cite{MilnerGulland:2011vm}.
-
-% Most recently the role of social interactions and collective navigation has been hypothesized \cite{Berdahl:2015kv,Berdahl:2016dx}.
-% These strays can introduce new maladaptive genotypes into the recipient population.
-% Straying amon salmon may be influenced by environmental factors such as water temperature, human activities such as hatchery practices, and population density as predicted by the collective migration hypothesis \cite{Peterson:2014gy}.
-% Thus, there is an opportunity to consider the eco-evolutionary consequences of straying for metapopulations in species of conservation and management concern such as salmon \cite{Carlson:2011ce}.
-
-% 
-% \begin{figure}
-%   \captionsetup{justification=raggedright,
-% singlelinecheck=false
-% }
-% \centering
-% \includegraphics[width=0.4\textwidth]{figs2/fig_xkcd.png}
-% \caption{
-% \emph{Migrating Geese}, a comic from Randall Monroe's xkcd (https://xkcd.com/1729/). 
-% A flock of geese travel in the direction of a shared destination, the lone stray named \emph{Kevin}.
-% In this case, the rate of straying is $m=0.05$, which is not an uncommon rate for migrating populations of salmon \cite{Satterthwaite:2015ge}. 
-% Reprinted under the Creative Commons Attribution-NonCommercial 2.5 License.
-% } \label{fig:xkcd}
-% \end{figure}
-
-%Alternative stable state
-Here we seek to explore how collective density-dependent straying influences the stability and robustness of metapopulations through ecological and evolutionary processes.
-% the potentially detrimental evolutionary effects of straying (erosion of local adaptation) and subsequent gene flow is balanced by the positive effects of demographic rescue is the subject of this contribution.
-% Here we ask the overarching question: how does collective behavior mediated dispersal and gene flow interact to influence the robustness of locally adapted populations?
-To address this question we construct a minimal eco-evolutionary model of two populations occupying different sites that are linked by straying individuals, each of which with an associated trait distribution subject to natural selection determined by local conditions.
-Specifically we compare (a) different rates of straying and (b) the influence of collective movement, across (c) increasing environmental heterogeneity, by assessing two measures of metapopulation robustness: the portfolio effect and the time required for a population(s) to recover following an induced disturbance.
-This model enables us to explore the multiple and potentially opposing pathways by which straying influences metapopulation robustness such as the potentially detrimental erosion of local adaptation vs.
-the positive effects of demographic and evolutionary rescue.
-% [We show that taking straying into account leads to alternative stable states in population densities and trait values, which has consequences for maladaptation, intraspecific trait variability, and long-term sustainability of salmon metapopulations.]
-
-\section{Model Description \& Analysis}
-
-\noindent{\bf (a) Metapopulation framework}\\
-\noindent We consider two populations $N_1$ and $N_2$ that inhabit two distinct habitats, each with trait values $x_1$ and $x_2$ determing recruitment rates.
-We assume that there is an optimum trait value $\theta_1$ and $\theta_2$ associated with each habitat, where recruitment is maximized if the trait value of the local population equals the optimum, such that $x = \theta$.
-Moreover, we assume that $x_{1,2}$ are normally distributed with means $\mu_1$ and $\mu_2$ and have the same standard deviation $\sigma$.
-As such, the recruitment rate for both populations is determined by the mean trait value of the local population, such that $r_1 = R_1[\mu_1(t),\theta_1]$.
-Trait means for each population are subject to selection, the strength of which depends on the difference between the trait mean and the local trait optimum at a given point in time \cite{simpson1953major,Lande:1976ga}.
-
-%PW: Can we say anything about the effect of the recipient population size and that effect on evolutionary dynamics?
-Again, straying can be thought of as both from the donor or recipient population perspective.
-The two populations are assumed to reproduce in spatially separate sites that are close enough such that a proportion of the population $m$ can stray into the other site, and where mortality occurs before individuals return to reproduce.
-If there is no straying between these populations (such that they are independent), then the mean trait evolves towards the optimal value such that $x_1 \rightarrow \theta_1$, and the recruitment rate for that population will be maximized.
-If there is straying between populations at rate $m$, then the traits in each respective location will be pulled away from the optimum, and recruitment rates will be lowered.
-As $m \rightarrow 0.5$, the populations are perfectly mixed, acting as a single population.
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.4\textwidth]{figs2/fig_traj.pdf}
-\caption{
-A) The steady state densities of $N_1$ and $N_2$ as a function of a constant stray rate $m$.
-Which population attains the low- or high-density state is random due to small applied fluctuations in the initial conditions.
-B) The steady state trait values measured as $\theta_i - x_i$, as a function of a constant stray rate $m$. 
-FB marks the fold bifurcation.
-} \label{fig:traj}
-\end{figure}
-
-We use the discrete Ricker population dynamic framework described by Shelton and Mangel \cite{Shelton:2011eq} as the basis for our two-site model, with the added effect of the local population $N_i$ mixing with a set proportion $m$ of a remote population $N_j$ that is straying into it.
-In this sense, both populations serve as donor and recipient populations.
-We first assume that the proportion ${\rm e}^{-Z}$ of both populations survive such that the surviving aggregated population, composed of both local individuals (at site $i$) and incoming strays (from site $j$), is $\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z}$.
-Because local individuals will recruit differently than incoming strays, the recruitment of the aggregate must incorporate two recruitment functions, given by $\left(R_i[\mu_i(t)] (1-m)N_i(t) + R_i[\mu_j(t)] m N_j(t)\right)$.
-This mix of individuals is subject to the same compensatory effects, which is determined by the parameter $\beta$.
-Taken together, the difference equation that determine changes in population size is
-
-\begin{align}
-  &N_i(t+1) = \\ \nonumber
-  &\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z} \\ \nonumber
-  &+ \left(R_i[\mu_i(t)] (1-m)N_i(t) + R_i[\mu_j(t)] m N_j(t)\right) \\ \nonumber
-  &\times {\rm e}^{-\beta ((1-m)N_i(t) + m N_j(t))},
-  \label{eq:N}
-\end{align}
-
-\noindent where the difference equation for $N_j$ mirrors that for $N_i$.
-
-The recruitment of local individuals $(1-m)N_i(t)$ as a function of their mean trait value at time $t$ and the local trait optimum, is
-
-\begin{align}
-  &R_i[\mu_i(t)] = \\ \nonumber
-  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i(t)-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i(t),\mu_i,\sigma^2) {\rm d}x_i(t) +\tilde{P}\\ \nonumber
-  &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-\mu_i(t))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\} +\tilde{P},
-  \label{eq:R}
-\end{align}
-
-\noindent where the mismatch between the local trait mean $\mu_i(t)$ and the local optimum $\theta_i$ scales the recruitment rate for the population, and $\tilde{P}\sim {\rm Normal}(0,0.01)$ introduces a small amount of demographic error.
-The parameter $\tau$ is the strength of selection, and controls the sensitivity of recruitment to changes in the mean trait value away from the optimum (the strength of selection increases with smaller values of $\tau$), which we set as $\tau=1$ here and throughout.
-Because straying individuals are emigrating from a population with a mean trait value farther from the local optimum, their rate of recruitment is diminished.
-Recent studies of wild sockeye have indeed found that straying individuals have lower life-time fitness than individuals that do not stray, although it unknown at what life-stage this selection occurs \cite{Peterson:2014gy}.
-% The compensatory effects are then determined by the exponential, following the Ricker stock-recruitment relationship.
-\\
-
-\begin{figure*}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.8\textwidth]{figs2/fig_MDPE_hm.pdf}
-\caption{
-(a) Total means $N_t$, (b) difference in means $\Delta N$, and (c) the portfolio effect PE as a function of heritability $h^2$ and a constant stray rate $m$. Light colors = high values.
-The black line shows the fold bifurcation separating a single steady state (left) from alternative stable states (right).
-(d) The relationship between the time to recovery following a disturbance and the portfolio effect.
-The black line denotes the fold bifurcation.
-} \label{fig:PE}
-\end{figure*}
-
-% \noindent{\bf (b) Recruitment over a selective landscape}
-\noindent Because individuals from the local population are mixed with individuals from the remote population via straying and subsequent reproduction, the resulting trait distribution is a mixed normal with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and for the straying individuals, $1-w_i$, where
-\begin{equation}
-w_i=\frac{(1-m)N_i(t)}{(1-m) N_i(t) + m N_j(t)}.
-\end{equation}
-We make two simplifying assumptions.
-First, we assume that the distribution resulting from the mix of remote and local individuals, following reproduction, is also normal with a mean value equal to that of the mean for the mixed-normal distribution.
-Thus, we assume that strays can successfully reproduce and introduce their genotypes into the recipient population, which is supported by observations in wild populations \cite{Jasper:2013cc}.
-Second, we assume that changes in trait variance through time are minimal, such that $\sigma^2$ is assumed to be constant, and this is a common simplification in eco-evolutionary models of population dynamics \cite{Lande:1976ga,Schreiber:2011wx,Gilbert:2014ee,Gibert:2015kc}.
-
-An increasing flow of incoming strays is generally expected to pull the mean trait value of the local population away from its optimum over time, which will decrease its rate of recruitment.
-Following Lande \cite{Lande:1976ga}, the mean trait value thus changes through time according to the difference equation
-
-\begin{align}
-  \label{eq:mu}
-  \mu_i(t+1) &= w_i\mu_i(t) + (1-w_i)\mu_j(t) \\ \nonumber
-  &+ h^2\sigma^2\frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)] + (1-w_i)R_i[\mu_j(t)]  \right),
-\end{align}
-
-\noindent where the first two factors determine the mixed normal average of the aggregated local and remote populations.
-The partial derivative in the Eq.
-\ref{eq:mu} determines how the mean trait changes through time due to natural selection \cite{Lande:1976ga}, which is proportional to the change in mean fitness with respect to $\mu_i$.
-\\
-
-\noindent {\bf (b) Density-dependent straying}
-%Density dependent m
-\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is mounting evidence that at least in some species (including salmon) the straying rate is density-dependent with a signature of collective navigation \cite{Berdahl:2016dx,Bett:2017ha}.
-Specifically, the rate at which individuals stray has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tend to decrease the rate at which individuals err, reducing the overall proportion of a population that strays.
-According to Berdahl et al. \cite{Berdahl:2014bl,Berdahl:2016dx}, given the probability that an individual strays is $m_0$, the proportion of the local population $N_i(t)$ that strays is
-
-\begin{equation}
-  m(t) = m_0\left(1- \frac{N_i(t)}{C+N_i(t)}\right),
-  \label{eq:ddm}
-\end{equation}
-
-\noindent where $C$ is a half-saturation constant.
-We note that at the limit $C\rightarrow \infty$, the density-dependent straying rate becomes constant such that $m(t) \rightarrow m_0$, and this corresponds to the original formulation where $m=m_0$.
-A similar observation shows that when the population density is very high, $m(t) \rightarrow 0$, and when it is small, individuals operate without regard to collective behavior, meaning $m(t) \rightarrow m_0$.
-Thus, for realistic population densities, $m(t) < m_0$.\\
-
-\noindent {\bf (c) Habitat heterogeneity}
-
-%Linking m/m0 and thetadiff
-The straying rate is largely influenced by the spatial distance between the donor and recipient population.
-The greater the distance between two populations, the lower the expected rate of straying \cite{H:2013fs,Westley:2015to}.
-We account for this interdependence in our model by assuming that $m$ (if the stray rate is constant) or $m_0$ (if the stray rate is density-dependent) is a function of the difference between optimal trait values between sites $\theta_i-\theta_j$, which can be assumed to be large if the remote site $j$ is a great distance away from the local site $i$.
-If sites $i$ and $j$ are very close, the stray rate is maximized at $m_{\rm max} = 0.5$, assuming both sites are equally attractive to the respective populations.
-Thus, we can integrate these two variables by setting $m,m_0 = (1/m_{\rm max} + \epsilon (\theta_i-\theta_j))^{-1}$, where $\epsilon$ sets the sensitivity of a declining $m$ to increasing distance (greater values of $\theta_i-\theta_j$).\\
-
-\noindent {\bf (d) Measuring metapopulation robustness}
-\noindent We evaluated metapopulation robustness by measuring the average-CV portfolio effect (PE) \cite{Anderson:2014cx,Schindler:2015gf} as well as the time required for the system to return to a steady state following an induced disturbance to one or both of the populations \cite{Ovaskainen:2002il}.
-The average-CV portfolio effect is, as the name implies, the average CV across each population divided by the CV of the aggregate \cite{Anderson:2013gb}, such that
-
-\begin{equation}
-\langle{\rm PE}\rangle =\frac{1}{X}\sum_{i=1}^{X} \frac{\sqrt{{\rm VAR}(N_i)}}{{\rm E}(N_i)}\cdot \frac{{\rm E}(N_T)}{\sqrt{{\rm VAR}(N_T)}},
-\label{eq:pe}
-\end{equation}
-
-\noindent where in this case the number of populations is limited to $X=2$ and the expectations $\rm E(\cdot)$ and variances $\rm VAR(\cdot)$ are evaluated at the steady state.
-As the CV of $N_T$ decreases relative to that of the constituent populations, $\langle{\rm PE}\rangle > 1$, and the metapopulation is presumed to become more stable.
-Portfolio effects greater than unity corresponds to less synchronization  \cite{Loreau:2008ju,Anderson:2014cx,Yeakel:2013vz} and thus a greater potential for demographic rescue among populations, buffering the system as a whole against extinction.
-
-%Although there are more robust ways to measure the portfolio effect (in particular for comparing the PE across species with different mean-variance relationships; REF: Anderson), its only utlity here is to provide a simple statistical measure of metapopulation persistence.
-A more direct way to measure system robustness is to measure the time that it takes the system (measured as the aggregate biomass $N_T$) to recover its steady state abundance following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
-Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation \cite{GuckHolmes}, because we aimed to 1) assess the effects of a large perturbation, and 2) estimate the time required for all transient effects to decay (including dampened oscillations), we used a simulation-based numerical procedure.
-
-Numerically estimating the time that it takes for a perturbed system to relax also permits a more detailed perspective of metapopulation fragility.
-For example, if populations settle to alternative stable states (alternative stable states in our model requiring one population to be high-density and one low-density), comparing recovery times after a disturbance applied to the high, low, and/or both populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery.
-We measured the time required for the system $N_T$ to recover to its steady state following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} are equivalent if the system is in the single steady state regime); (\emph{iii}) near-collapse of both populations where just 1.0\% of each survives.
-Throughout, we will refer to an increase in the portfolio effects and/or reduction in recovery times as promoting metapopulation robustness, which is expected to have a positive effect on persistence.
-\\
-
-\section{Results}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.4\textwidth]{figs2/fig_thetaPEmvm.pdf}
-\caption{
-(a) Median portfolio effect as a function of a constant stray rate $m$ (solid line) and density-dependent stray rate (point pairs) given heritability is $h^2 < 0.5$ and $\Delta\theta=5$.
-Point pairs connected by a horizontal line represent the PE as a function of density-dependent straying rates, evaluated for both low- and high-density populations at equilibrium.
-The lower straying rate of a pair is for the larger population; the higher straying rate is for the smaller population.
-(b) Median portfolio effects for habitats with increasing heterogeneity as measured by the difference in regional trait optima $\Delta \theta$ for both constant and density-dependent stray rates as shown in (a).
-FB marks the fold bifurcation.
-} \label{fig:thetaPE}
-\end{figure}
-
-%Alternative stable states
-
-\noindent{\bf (a) Nonlinear effects of straying on metapopulation robustness} \\
-\noindent Regardless of density dependence, straying lowers steady state densities for both populations by (\emph{i}) the donor population losing locally-adapted individuals to the recipient population and (\emph{ii}) the introduction of maladapted individuals to the recipient population from the donor population (Fig.
-\ref{fig:traj}).
-This prediction is in accordance with observations from natural populations \cite{Bett:2017ha}. %, regardless of trait heritability and variance or habitat heterogeneity
-The decline in steady state densities is not gradual: as straying increases, the system crosses a fold bifurcation whereby the single steady state for the metapopulation bifurcates into two alternative stable states: one at high biomass, and one at low biomass density (figure \ref{fig:traj}a, \ref{fig:PE}a).
-Mean trait values for both populations bifurcate similarly (figure \ref{fig:traj}b), depending on which population attains a low- vs. high-density. 
-Above the threshold straying rate defined by the fold bifurcation, there are two alternative eco-evolutionary states: the \emph{dominant state} population will have a higher a higher density and higher degree of local adaptation (lower trait offset from the local optimum), while the \emph{subordinate state} population will have lower density with maladapted trait values (higher trait offset from the local optimum).
-Whether a specific population goes to one state or the other in our model is random, which is due to a small amount of introduced variance in the initial conditions.
-% Thus, we predict that high levels of straying in metapopulations will produce alternative eco-evolutionary states.
-
-\begin{figure*}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.9\textwidth]{figs2/fig_relax_lowh.pdf}
-\caption{
-Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.2$.
-If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
-FB marks the fold bifurcation.
-} \label{fig:relax}
-\end{figure*}
-
-Trait heritability has a large impact on the degree to which straying rate affects both the aggregate population steady state density ($N^*_T=N^*_1+N^*_2$; figure \ref{fig:PE}a) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N=|N^*_1-N^*_2|$; figure \ref{fig:PE}b).
-Greater trait heritability results in a faster decline in $N_T^*$ with increasing straying rates $m$, but leads to only moderate changes to $\Delta N$.
-Conversely, in the context of lower trait heritability, an increase in the straying rate has little impact on the total biomass density but contrastingly large effects on $\Delta N$.
-The fold bifurcation (the black line in Figs.
-\ref{fig:PE}a-c) occurs at lower values of the straying rate $m$ with decreased trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
-Although trait heritability among salmonids is variable, most life history traits have an $h^2 <0.5$ \cite{Carlson:2008hl}, and we largely focus additional analyses on that range.
-
-%Critical slowing down
-As the fold bifurcation is approached with increasing $m$, the portfolio effect increases sharply due to an amplification in variance within both donor and recipient populations ${\rm VAR}(N_{i,j})$.
-This variance increase is the product of a dynamical process known as \emph{critical slowing down} that occurs near fold bifurcations \cite{Scheffer:2009gg}, a phenomenon that some have suggested may serve as an early warning indicator for approaching phase transitions \cite{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br}.
-For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}a-c), where alternative stable states occur, the portfolio effect declines steadily as the CV of $N_T$ increases.
-The decline over $m$ is more gradual if trait heritability is low, and steeper if trait heritability is high (figure \ref{fig:PE}c).
-
-% \noindent{\bf (c) Rates of recovery following catastrophic collapse}\\ 
-% Return time as a measure of system persistence and relation to PE
-As the portfolio effect is highly sensitive to the rate of straying between populations, so is the time required for the system to recover to a steady state following a large disturbance.
-In general, we find that the average-CV portfolio effect is negatively correlated with recovery time (figure \ref{fig:PE}d), indicating that, for our system, both measures are valuable indicators of metapopulation robustness.
-Because we can assess the time to recovery in response to the various disturbance types described above, this allows us to gain an in-depth perspective into the fragility of the metapopulation as a function of straying rate.
-
-Straying had non-linear impacts on the recovery time of populations. 
-When the dominant state (well adapted and high density) is wiped out, high levels of straying allow it to recover quickly (figure \ref{fig:relax}a) because the surviving population has a mean trait value skewed towards the optimum of the recovering population (figure \ref{fig:relaxtraj_hdlh}).
-Yet, as straying decreases, recovery time for the disturbed dominant state population increases, in part because there is enough time for the trait distribution to move back towards the trait optimum of the subordinate state population.
-In contrast, when the subordinate state population (maladapted and low density) is wiped out, recovery rates are fastest at low to intermediate levels of straying.
-Because the mean trait values of both populations are skewed towards those of the dominant population, when the subordinate population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into the recovering population extends the length of time required for it to return to its steady state (figure \ref{fig:relaxtraj_ldlh}).
-When both populations are both dramatically reduced, recovery time is generally fastest at lower levels of straying, while near the onset of the fold bifurcation, recovery time increases explosively and this is -- as the name implies -- characteristic of \emph{slowing} dynamics that occur near critical transitions \cite{Scheffer:2009gg,Kuehn:2010p2591}.\\
-
-\noindent{\bf (b) The effects of collective migration and density dependent straying} \\
-%Portfolio effects
-If we assume that the rate of straying is density-dependent, the probability that an individual strays $m_0$ determines the rate of straying within the population, such that $m(t)$ becomes lower as $N(t)$ increases, likely due to the effects of collective decision-making \cite{Berdahl:2016dx} (Eq.
-\ref{eq:ddm}).
-%Because our analyses concern steady state conditions, the inclusion of density-dependent straying does not have a large impact on the qualitative nature of our findings (figure \ref{fig:ddm}).
-Density dependence alters the straying rate at steady state population densities because $0 < m^* < m_0$, and this serves to rescale both the strength of the PE as well as the recovery time, but does not change the qualitative nature of our findings.
-In the alternative stable state regime, because each population exists at different steady state densities, there are likewise two alternative straying rates $(m_i^*,m_j^*)$: the higher straying rate is associated with the low-density population, and the lower straying rate is associated with the high-density population.
-We assessed metapopulation robustness across a range of $(m_i^*,m_j^*)$ values by varying the probability that an individual strays $m_0$, which is positively and linearly related to $(m_i^*,m_j^*)$.
-We find that the portfolio effects generated in systems with density-dependent straying are qualitatively similar to systems with constant straying, however there are some important quantitative differences.
-First, the PE associated with the high-density (low $m^*$) population is the same as that for a system with a constant $m$ (figure \ref{fig:thetaPE}a).
-% As such, the difference in $m^*$ between subordinate and dominant populations appears to matter less than the lower straying rate generated by the dominant population.
-As $m^*$ increases, we observe an increase in the PE than for systems with constant $m$.
-
-%Recovery times
-Density-dependent straying alters these recovery times (figure \ref{fig:relax}b). 
-First, in comparison with constant stray rates, density-dependent straying made recovery more rapid at elevated stray rates when both populations collapsed and when the subordinate population was extirpated.
-%Recovery times
-%The recovery times for systems with constant and density-dependent straying both show similar trends for a given $m$ and $m^*$ (figure \ref{fig:relax}c,d), however there are some distinct quantitative differences.
-At low straying rates, near-collapse of both populations resulted in longer than expected recovery times, whereas in the alternative stable state regime (higher $m^*$), the recovery times for different disturbance types were very similar to systems with a constant $m$ (figure \ref{fig:relax}b; note difference in x-axis scales).
-As trait heritability increased, the metapopulation always recovered more quickly if the small population was lost (figure \ref{fig:relax_highh}).
-% There was greater similarity in the recovery times for constant and density-dependent $m$ systems when trait heritability was high (figure \ref{fig:relax_highh}).
-The lower recovery time for systems with increased $m^*$ mirrors an elevated PE with higher density-dependent straying rates (figure \ref{fig:PEtheta}).
-% This higher PE at increased $m^*$ mirrors recovery times that are also lower than those observed for systems with constant $m$ (figure \ref{fig:relax}).
-In tandem, analysis of both PE and recovery time suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our minimal model, it does appear to promote robustness, particularly when the aggregate biomass is low and straying is correspondingly high.
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.5\textwidth]{figs2/fig_traitdiff.pdf}
-\caption{
-Phenotypic diversity ($\Delta \mu^*$) evaluated at the steady state as a function of straying rate $m$ and trait heritability $h^2$.
-The descrete jump occurs as the system crosses the fold bifurcation; lower phenoytic diversity emerges with higher straying rates and in the alternative stable state regime.
-} \label{fig:traitdiff}
-\end{figure}
-
-%Lowering Phenotypic diversity with increasing straying rate
-Increased rates of straying lowers phenotypic diversity ($\Delta \mu^* = |\mu_i(t)-\mu_j(t)|$, evaluated at the steady state) because both local and remote populations are increasingly homogenized.
-The loss of phenotypic diversity with increased straying is greater if trait heritability is low because traits take longer to go back to their local optima than they do when heritability is large. 
-Hence straying counters the effect of diversifying local adaptation. 
-% The loss of phenotypic diversity with increased straying is greater if trait heritability is low because the selective forces acting against trait means far from the local optima are lessened.
-% with decreased $h^2$.
-Less intuitively, we observe a discrete jump towards low phenotypic diversity as the fold bifurcation is crossed (figure \ref{fig:traitdiff}).
-Although the development of alternative stable states elevates the portfolio effect due to the variance-dampening effects of the aggregate, entering this dynamic regime also results in a substantial decline in phenotypic diversity, which may have less predictable adverse effects on the population.
-\\
-
-\noindent{\bf (c) The role of habitat heterogeneity and changing selective landscapes}\\ 
-\noindent Increasing differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater regional differences in the conditions that favor alternative trait complexes, which we interpret here as greater habitat heterogeneity (REFS).
-If both populations are isolated, natural selection will direct the mean trait values of both populations towards their respective optima, such that $\mu_i(t) \rightarrow \theta_i$ as $t\rightarrow\infty$.
-%difference in optimal trait values between the local and remote populations, which we call .
-With the onset of straying, we find that increasingly divergent trait optima generally lower $N_T$ and exaggerate $\Delta N$ (figure \ref{fig:thetadiffN}), such that the biomass distribution becomes increasingly uneven.
-%I've checked this - it's a theshold type (logistic) response Supp fig?
-The impact of habitat heterogeneity on the portfolio effect and recovery time is more complex, serving to emphasize the nonlinear relationship between rates of straying and metapopulation robustness.
-%regardless of heritability.
-%The rate of straying that gives rise to a large PE at the fold bifurcation is conditioned on the difference in optimal trait values between the local and remote populations, which we call $\Delta\theta = \sqrt{(\theta_i - \theta_j)^2}$.
-As habitat heterogeneity increases, alternative stable states appear at lower straying rates -- with the crossing of the fold bifurcation, accompanied by a peak in the PE -- whereas the magnitude of increase in the PE also increases (figure \ref{fig:thetaPE}b), reducing recovery time (figure \ref{fig:relaxtheta}).
-For increased rates of straying, greater habitat heterogeneity erodes the PE (figure \ref{fig:thetaPE}b) and increases the recovery time (figure \ref{fig:relaxtheta}).
-These results together suggest that habitat heterogeneity, as measured as the differences in trait optima between two habitats $\Delta\theta$, promotes robustness when straying rates are low, and erodes robustness when straying rates are high.
-%This means that alternative stable states can emerge with relatively little dispersal between sites (figure \ref{fig:thetaPE}), however the increased PE (and shorter recovery times).
-
-If we assume that environmental heterogeneity increases with distance between sites, or that individuals may have behaviors that minimize their chances of straying into very different habitats, the rate of straying will decline with habitat heterogeneity.
-Accordingly, $\Delta\theta$ will decrease as $m$ increases, such that there are low rates of straying between dissimilar populations and high rates of straying between similar populations.
-We find that under these conditions alternative stable states now appear for very low rates of straying $0 < m \leq 0.43$.
-As the straying rate increases and $\Delta\theta$ decreases, a single stable state emerges as the fold bifurcation is crossed, which is opposite the pattern observed when these parameters are independent.
-There are two notable dynamics that emerge following extinction of the dominant population at low rates of straying between dissimilar (high $\Delta\theta$) sites:
-(\emph{i}) above a threshold $m$ value, the dominant population recovers quickly enough that the evolving subordinate phenotype is overwhelmed by incoming strays, and it shifts back to its pre-disturbance (subordinate) state;
-(\emph{ii}) below a threshold $m$ value, there is enough time and isolation for the subordinate trait mean to shift towards its local optimum, and away from that of the recovering dominant population, such that the dominant population becomes subordinate, and the subordinate population becomes dominant (figure \ref{fig:inertia}).
-This threshold value of $m$, below which this behavior occurs, is marked by the asterisk in figure \ref{fig:mtheta}, and holds for both constant and density-dependent straying (figure \ref{fig:mthetamvm}).
-
-% At low rates of straying, there is little mixing between dissimilar populations, and this leads to the formation of alternative stable states.
-% Here we find that if the subordinate population is wiped out, the time to recovery is minimized, whereas if the dominant population is wiped out, the time to recovery is maximized; near-collapse of both results in intermediate recovery times.
-% When straying rates are low and the difference in trait optima are correspondingly high, we might assume that extinction of the dominant population would result in relatively faster recovery.
-% This would be a logical assumption because the mean trait value of the subordinate population would be skewed towards that of the dominant (larger) population, such that recruitment at the disturbed site should be high and the population should recovery more quickly.
-% Our results show, however, that \emph{this assumption is wrong:} although the mean trait value of the subordinate population \emph{is} skewed towards the optimum of the dominant population, when the latter is wiped out and the rate of straying is low, there is enough time and isolation for the subordinate trait mean to shift towards its own optimum, and away from that of the recovering dominant population (figure \ref{fig:inertia}).
-% This \emph{selective inertia} occurs until the dominant population grows large enough that the evolving subordinate phenotype is overwhelmed by incoming strays, shifting it back to its pre-disturbance (subordinate) state.
-% This effect can lead to much longer recovery times for the dominant population.
-% Importantly, if the rate of straying is below a threshold value, it can also result in a state-switch, wherein increased isolation permits the subordinate population to \emph{escape} the selective pull of the dominant population, leading to a switching in which population is in the dominant/subordinate state.
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-  \centering
-  \includegraphics[width=0.5\textwidth]{figs2/fig_mtheta_rt.pdf}
-  \caption{
-  Distance dependent recovery times for three disturbance types. When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases.
-  The `$*$' marks the value of $(m,\Delta\theta)$ below which there is a switch in subordinate/dominant states following extinction of the dominant population.
-  FB marks the fold bifurcation.
-  } \label{fig:mtheta}
-\end{figure}
-
-% That alternative stable states appear and that PE becomes severely depressed for very low rates of straying is surprising: this means that even a small amount of mixing of populations from distant or dissimilar populations can qualitatively alter the dynamics of the metapopulation, regardless of trait heritability.
-%(for the discussion: an example of this situation may be the salmon populations during the last glacial maximum, where any mixing would be from geographically distant populations)
-
-\section{Discussion}
-
-We have shown that density-dependent straying between populations consistent with collective navigation, coupled with localized selection against donor phenotypes, has a large and nonlinear impact on dynamic properties directly impacting the robustness of metapopulations against external disturbances.
-Specifically, we observed robustness in terms of: 
-1) the average-CV portfolio effect \cite{Anderson:2013gb,Anonymous:2015gf}, a statistical metric commonly used to assess the buffering capacity of metapopulations, and
-2) the recovery time, defined here as the time required for the aggregate metapopulation biomass $N_T$ to return to its steady state following an induced disturbance, which is mechanistically linked to persistence \cite{Ovaskainen:2002il}.
-In our minimal eco-evolutionary model of dispersal and natural selection between two populations, we show that these statistical and mechanistic descriptors of metapopulation dynamics and robustness are tightly coupled (figure \ref{fig:PE}d), which is not uncommon among different metrics of stability \cite{Donohue:2013iu}.
-Taken as a whole, we interpret our results to suggest an important role of density-dependent straying in the colonization and recovery dynamics for populations in shifting river mosaics, while also underscoring the risk of straying by individuals with maladaptive traits to reduce the productivity of locally adapted stock complexes.
-
-% 
-% %The bifurcation and critical slowing down
-% [This paragraph is interesting to me but could be cut] Although the portfolio effect is negatively correlated with recovery time, there is one exception: the point of transition from a single steady state to alternative stable states, which occurs at the fold bifurcation (REF).
-% At this point, the portfolio effect is maximized, but so is the recovery time (figures \ref{fig:PE}, \ref{fig:relax}).
-% This suggests that while the sharply increasing variance of the individual populations is dampened by aggregation, a large disturbance will have a much greater adverse effect due to exponentially longer recovery times.
-% Whether the change in variance, also known as critical slowing down, as the alternative stable state regime is approached could be used as an early warning signal of an oncoming phase transition is a hotly debated topic (REFS).
-% The detectability of such changes in dynamical behavior among salmonids appears to be idiosyncratic across species, though the difficulty in measuring critical slowing down may - ironically - be masked by large portfolio effects \cite{Krkosek:2014ch}.
-%
-
-%alternative stable states & critical slowing
-A salient finding from our results was that straying can push metapopulations into alternative stable states, pushing one of the populations to a high density population (the \emph{dominant state}), and one to a low density population (the \emph{subordinate population}).
-This asymmetry in population densities also leads to an asymmetry in the mean trait values, both of which are skewed towards the local optimum of the dominant population. 
-The formation of alternative stable states in our system is a relatively common example of spatial pattern formation, which only occurs above a threshold straying rate, the value of which depends on trait heritability and the difference in trait optima between sites.
-Pattern formation can occur as a consequence of myriad ecological processes, including habitat selection, aggregation, local environmental conditions, and/or interspecific interactions such as competition or predation, and is inexorably linked to scale \cite{Levin:1992p1472}.
-Additionally, pattern formation is also well known to occur with the onset of spatially or diffusion-induced instabilities, or Turing instabilities \cite{Meinhardt:2012hw}. 
-Here, we describe an additional mechanism through which pattern formation in local densities can occur, when the onset of collective movement-mediated maladaptation leads to local differences in reproductive rates resulting in alternative stable states across sites.
-%This mechanism of pattern formation through the simultaneous interplay between ecological and evolutionary dynamics may open up new areas of research.
-While metapopulation approaches have widely recognized the importance of source and sink populations \cite{Hanski:1999vu} and have observed formation of alternative states \cite{Boughton:1999fa}, here we show that even similar populations can evolve into divergent states where one population is large and the other is smaller and maladapted.
-
-% 
-% To what extent the existence of alternative stable states promotes or diminishes the system as a whole is a matter of perspective, though with clear conservation implications.
-% At face value, the subordinate population is pushed closer to extinction (figure \ref{fig:traj}), suggesting that the system may be more fragile.
-% On the other hand, the presence of \emph{just enough} straying to cause formation of alternative stable states both increases the portfolio effect (figure \ref{fig:PE}c) and reduces the rate of return -- particularly if straying is density-dependent (figures \ref{PEtheta}a, \ref{fig:relax}b).
-% This asymmetry in population densities also leads an asymmetry in the mean trait values, skewed towards the local optimum of the dominant population, one, and this has important -- and often non-intuitive -- ramifications for metapopulation robustness.
-%, as measured by both the portfolio effect as well as recovery time.
-
-There is a `sweet spot' for straying that maximizes metapopulation robustness. 
-Results from our model reveal that the presence of just enough straying to cause formation of alternative stable states increases the portfolio effect (figure \ref{fig:PE}a). 
-Previous theoretical work has shown that increased connectivity may erode portfolio effects in herring metapopulations, where straying is also thought to be density-dependent \cite{Secor:2009ena}.
-Although high levels of dispersal in our system shows a similar trend, the interplay between dispersal and PE is more subtle when selection for local adaptations is considered, such that low to intermediate levels of density-dependent straying result in an elevated PE, increasing the buffering capacity of the metapopulation.
-Although PE is measured at the steady state, low to intermediate rates of straying also appear to have a beneficial effect on transient dynamics.
-When there is just enough straying to cause alternative stable states, recovery times following induced disturbances decline, though -- as with the PE -- the time to recovery then grows if straying becomes too large (figure \ref{fig:relax}a).
-
-% %Density dependent straying
-% This themed issue is largely concerned with the notion that straying in biological systems where movement is a function of collective navigation may be density-dependent.
-% Berdahl et al.
-(REF) provided a mechanistic hypothesis for density-dependent straying where the proportion of the population that strays is less than the probability that a single individual takes the wrong turn.
-%(as illustrated by \emph{Kevin}; figure \ref{fig:xkcd}) takes the wrong turn.
-% Neither the transient nor asymptotic dynamics that we describe here qualitatively differ as a function of whether the rate of straying is constant or density-dependent, however we observe important quantitative differences that suggest density-dependent straying may play an important role in the persistence of metapopulations over evolutionary time.
-
-This themed issue formalizes the role of collective movement in the ecology of natural systems and illuminates a signature of collective navigation in animal populations on the move.
-We observed two salient results that contribute to our understanding of collective movement that combine to suggest that density dependent straying may play an important role in the persistence of metapopulations over evolutionary time.
-
-First, compared to systems with constant dispersal, density-dependent straying appears to increase the portfolio effect across a range of stray rates (figure \ref{fig:PEtheta}a). 
-% and that at low to intermediate levels of straying, PE is high.
-Second, density-dependent straying reduces the time to recovery following disturbances, and this is particularly true in the case of near-collapse of both populations (figure \ref{fig:relax}b).
-In this case, although both populations are near-collapse, the mean trait values of both are skewed towards the optimum for the dominant population.
-
- the straying rates are also higher because of the low-densities after collapse, and this permits faster recovery (figure \ref{fig:relaxtraj_bothlh}).
-% Third, in habitats with greater heterogeneity between sites (measured by an increase in $\Delta\theta$), density-dependent straying diminished robustness at low rates of straying (marginally lower portfolio effects and substantially longer return times; figures \ref{fig:thetaPE}b, \ref{fig:relaxtheta}a,b).
-% Conversely, for increased straying rates, greater and exaggerates the positive effects when $m^*$ is high (higher portfolio effects, and shorter return times).
-
-% %Comparing recovery times
-% Previous theoretical work has shown that the largest perturbations lead to the longest recovery times \cite{Ovaskainen:2002il}.
-% In terms of relative biomass lost, the largest perturbation that we investigate is the near-collapse of both populations, where only 1\% of the pre-perturbation densities survive, however whether this disturbance results in the longest recovery time depends largely on the rate of straying and trait heritability.
-% For example, when straying and trait heritability are low (figure \ref{fig:relax}) the extinction of the larger population maximizes return times, whereas if straying is high, near-collapse maximizes return time.
-% The latter is always true in the alternative stable state regime if trait heritability is high (figure \ref{fig:relax_highh}).
-
-%Habitats 
-Salmon are distributed and stray across a diverse range of habitats, and the rates of straying between geographically diverse sites can be plastic and idiosyncratic \cite{Westley:2015to}.
-Our surrogate measure for habitat heterogeneity is the difference in trait optima between sites $\Delta\theta$.
-In general, our findings indicate that increased habitat heterogeneity promotes robustness (higher PE, shorter time to recovery) when straying rates are low, but may erode robustness when straying rates are high (figure \ref{fig:thetaPE}b, solid lines).
-This may be particularly consequential for populations that are spatially adjacent but separated by  sharp environmental boundary, such that trait optima are divergent yet dispersal is relatively high (is this a thing?)...
-
-%mtheta
-However, habitats that are closer in space generally have greater similarity in environmental conditions than habitats that are geographically distant (REF).
-From a purely statistical standpoint, it is reasonable to expect a larger number of straying individuals between sites that are closer together.
-Alternatively, salmon that cue to specific environmental conditions are more likely to stray into sites that have structurally and physiognamically similar habitats (REFS).
-These considerations justify imposing a direct relationship between the rate of straying $m$ and habitat heterogeneity: as site dissimilarity increases, so too should the straying rate decrease.
-As site dissimilarity increases, so to should the optima in trait values for the resident populations.
-
-When habitat heterogeneity and the rate of straying are linked, we show that small amounts of either constant or density-dependent straying result in longer recovery times for the dominant population because there is enough isolation to allow the surviving subordinant population locally adapt towards its own optimum.
-As selection pushes the subordinate population away from the dominant trait optimum, the recovery of the extinct population is slowed (figure \ref{fig:mtheta}).
-This may have particular conservation implications...
-Moreover, local temperature regimes are known to play a central role in dictating local adapation of salmon populations (REFS).
-If temperature is the primary determinant of $\Delta\theta$ in terms of our model, this would suggest that populations spanning North-South gradients are more predisposed to distance dependent dynamics.
-In constrast, the relationship between the rate of straying and habitat heterogeneity may be less clear-cut for populations spanning an East-West gradient, where distant habitats may equally fitting for straying populations.
-(Bristal Bay vs. California???)
-
-%Concluding
-The importance of selection and its influence on the dynamics of populations is increasingly recognized as vital for assessing species' fragility.
-Of particular interest is how selection might influence spatially coupled metapopulations distributed across environmental gradients.
-We have shown that two measures of metapopulation robustness -- the portfolio effect and the time to recovery following a disturbance -- are strongly influenced by the rate at which individuals from a donor population stray into habitats occupied by a recipient population.
-Importantly, density-dependent straying, which may occur when individuals collectively navigate, can both increase the portfolio and lower the time to recovery following a disturbance, which is anticipated to promote persistence.
-We suggest that understanding the spatial complexity of metapopulations dispersing across heterogeneous environments, combined with the mosaic of selective forces acting on those environments, may be key to discovering those factors that promote persistence.
-
-\bibliography{aa_kevin}
 
@@
 \caption{
-Extinction of low-density population with a high constant straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
+Extinction of low-density population with a high constant straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{extfig-fig:relax}a).
 Black line marks the calculated point of recovery post-perturbation.
@@
 \caption{
-Extinction of high-density population with a high straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
+Extinction of high-density population with a high straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{extfig-fig:relax}a).
 Black line marks the calculated point of recovery post-perturbation.
@@
 (b) The same as (a) but including recovery times when straying is density-dependent, shown by linked pairs of points.
-Recovery times for systems with density-dependent straying are longer at low straying rates and shorter at higher straying rates, mirroring the change in portfolio effects shown in figure \ref{fig:thetaPE}.
+Recovery times for systems with density-dependent straying are longer at low straying rates and shorter at higher straying rates, mirroring the change in portfolio effects shown in figure \ref{extfig-fig:thetaPE}.
 } \label{fig:relaxtheta}
@@
 %until growth of the dominant population overwhelms this local selection.
-In this case, $m$ is less than $m=0.034$ (denoted by the asterisk in figure \ref{fig:mtheta}), such that isolation allows the subdominant population to `run away' from the influence of the dominant population, leading to a switch in states.
+In this case, $m$ is less than $m=0.034$ (denoted by the asterisk in figure \ref{extfig-fig:mtheta}), such that isolation allows the subdominant population to `run away' from the influence of the dominant population, leading to a switch in states.
 If $m$ is low but greater than $0.034$, isolation permits the subdominant population to `run away' from the influence of the dominant population, until it is overwhelmed by the recovering dominant population, and reverts back to its previous trait mean prior to the disturbance.
@@
 \caption{
-Near collapse of both populations with a low straying rate $m=0.1$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
+Near collapse of both populations with a low straying rate $m=0.1$ and low trait heritability $h^2=0.2$ (see figure \ref{extfig-fig:relax}a).
 Black line marks the calculated point of recovery post-perturbation.

### 2017-06-02 model/KevinExtinction.jl

@@ -390,7 +390,7 @@ pdf($namespace,height=4,width=10)
 pal = brewer.pal(9,'Greys')
 palsub = pal[c(4,6,8)];
 par(mfrow=c(1,2),mai = c(0.8, 0.8, 0.1, 0.1))
-plot($mvec,$(ma_rth2[:,1]),col=palsub[1],type='l',log='y',cex=0.5,lwd=2,xlab='m',ylab='Recovery time',ylim=c(min($rt),max($(ma_rth2))))
+plot($mvec,$(ma_rth2[:,1]),col=palsub[1],type='l',log='y',cex=0.5,lwd=2,xlab='m',ylab='Recovery time',ylim=c(min($rt),max($(ma_rth2))),las=1)
 lines($mvec,$(ma_rth2[:,2]),col=palsub[2],cex=0.5,lwd=2)
 lines($mvec,$(ma_rth2[:,3]),col=palsub[3],cex=0.5,lwd=2)
 text(par('usr')[1]-0.09,max($(ma_rth2)),'(a)', xpd=TRUE)
@@ -398,7 +398,7 @@ arrows($(mvec[indmax(ma_rth2[1:100,1])]),10,$(mvec[indmax(ma_rth2[1:100,1])]),12
 text($(mvec[indmax(ma_rth2[1:100,1])]),9,'FB')
 
 
-plot($(ma1_m2_ddm[:,1]),$(ma_rth2_ddm[:,1]),col=palsub[1],log='y',cex=0.5,pch=16,xlab='m*',ylab='Recovery time',ylim=c(min($rt_ddm),max($(ma_rth2_ddm))))
+plot($(ma1_m2_ddm[:,1]),$(ma_rth2_ddm[:,1]),col=palsub[1],log='y',cex=0.5,pch=16,xlab='m*',ylab='Recovery time',ylim=c(min($rt_ddm),max($(ma_rth2_ddm))),las=1)
 points($(ma2_m2_ddm[:,1]),$(ma_rth2_ddm[:,1]),col=palsub[1],cex=0.5,pch=16)
 points($(ma1_m2_ddm[:,2]),$(ma_rth2_ddm[:,2]),col=palsub[2],cex=0.5,pch=16)
 points($(ma2_m2_ddm[:,2]),$(ma_rth2_ddm[:,2]),col=palsub[2],cex=0.5,pch=16)

### 2017-06-02 model/KevinSimulations2.jl

@@ -160,14 +160,16 @@ library(RColorBrewer)
 cols = brewer.pal(4,'Set1')
 pdf($namespace,height=8,width=5)
 par(mfrow=c(2,1),mai = c(0.8, 0.8, 0.1, 0.1))
-plot($mvec,$n1mean,pch='.',col=cols[4],xlab="Straying rate m",ylab="Steady state biomass",cex=0.5,ylim=c(0,max($n1mean)))
+plot($mvec,$n1mean,pch='.',col=cols[4],xlab="Straying rate m",ylab="Steady state biomass",cex=0.5,ylim=c(0,max($n1mean)),las=1)
 points($mvec,$n2mean,pch='.',col=cols[4],cex=0.5)
 arrows($(mvec[indmax(pe)]),1200,$(mvec[indmax(pe)]),1100,length=0.05,angle=40,lwd=3)
 text($(mvec[indmax(pe)]),1275,'FB')
-plot($mvec,$x1mean,pch=".",col=cols[1],ylim=c(-5,5),xlab="Straying rate m",ylab="Trait offset")
+text(par('usr')[1]-0.09,1300,'(a)', xpd=TRUE)
+plot($mvec,$x1mean,pch=".",col=cols[1],ylim=c(-5,5),xlab="Straying rate m",ylab="Trait offset",las=1)
 points($mvec,$x2mean,pch=".",col=cols[2])
 arrows($(mvec[indmax(pe)]),3.4,$(mvec[indmax(pe)]),2.5,length=0.05,angle=40,lwd=3)
 text($(mvec[indmax(pe)]),4,'FB')
+text(par('usr')[1]-0.09,5,'(b)', xpd=TRUE)
 dev.off()
 """
 
@@ -178,7 +180,7 @@ library(RColorBrewer)
 cols = brewer.pal(3,'Set1')
 pdf($namespace,height=5,width=6)
 plot($mvec,$x1mean,pch=".",col=cols[1],ylim=c(-5,5),xlab="Stray rate",ylab="Trait offset")
-points($mvec,$x2mean,pch=".",col=cols[2])
+points($mvec,$x2mean,pch=".",col=cols[2],las=1)
 dev.off()
 """
 #Portfolio effect plot
@@ -429,13 +431,17 @@ library(RColorBrewer)
 pal = rev(brewer.pal(9,"Blues"))
[…]

### 2017-06-02 model/mvsddm2.jl

@@ -242,7 +242,7 @@ library(RColorBrewer)
 pal = brewer.pal(3,'Set1')
 pdf($namespace,height=8,width=5)
 par(mfrow=c(2,1),mai = c(0.8, 0.8, 0.1, 0.1))
-plot($(ma_mvecsmooth),$(ma_medpe5smooth),col=pal[2],type='l',ylim=c(1,max($([ma_medpe3[mlist] ma_medpe5[mlist] ma_medpe8[mlist]]))),xlab='m',ylab='PE',lwd=3)
+plot($(ma_mvecsmooth),$(ma_medpe5smooth),col=pal[2],type='l',ylim=c(1,max($([ma_medpe3[mlist] ma_medpe5[mlist] ma_medpe8[mlist]]))),xlab='m, m*',ylab='PE',lwd=3,las=1)
 arrows(0.053,2.48,0.053,2.35,length=0.05,angle=40,lwd=3)
 text(0.053,2.55,'FB')
 
@@ -253,8 +253,9 @@ for (i in 1:l) {
   segments($(ma_medindm5max[mddmlist])[i],$(ma_medpe5_ddm[mddmlist])[i],$(ma_medindm5min[mddmlist])[i],$(ma_medpe5_ddm[mddmlist])[i],col=paste(pal[2],'60',sep=''))
 }
 xleft<-0.0;xright<-0.45;ybottom<-1;ytop<-3;
+text(-0.09,2.65,'(a)', xpd=TRUE)
 
-plot($(ma_mvecsmooth),$(ma_medpe5smooth),col=pal[2],type='l',ylim=c(1,max($([ma_medpe3[mlist] ma_medpe5[mlist] ma_medpe8[mlist]]))),xlab='m, m*',ylab='PE',lwd=3)
+plot($(ma_mvecsmooth),$(ma_medpe5smooth),col=pal[2],type='l',ylim=c(1,max($([ma_medpe3[mlist] ma_medpe5[mlist] ma_medpe8[mlist]]))),xlab='m, m*',ylab='PE',lwd=3,las=1)
 lines($(ma_mvecsmooth),$(ma_medpe8smooth),col=pal[3],lwd=3)
 lines($(ma_mvecsmooth),$(ma_medpe3smooth),col=pal[1],lwd=3)
 #Or moving average
@@ -270,6 +271,7 @@ for (i in 1:l) {
[…]

### 2017-06-03 Manuscript/CoverLetter/YGWM_cover.tex
finalish draft

@@
     \vspace{1mm}
-    \centerline{\bf{Re: The eco-evolutionary impacts of collective straying on metapopulation robustness}} \\
+    \centerline{\bf{Re: Eco-evolutionary dynamics and collective migration: implications for salmon metapopulation robustness}} \\
 }
@@
 %Intro
-Please find attached our manuscript entitled ``The eco-evolutionary impacts of collective straying on metapopulation robustness'' co-authored by Justin Yeakel, Jean Philippe Gibert, Peter Westley, and Jonathan Moore, which we would like to submit for publication in the Philosophical Transactions of the Royal Society B themed issue \emph{Collective movement in ecology: From emerging technologies to management implications}, under the section title `Linking collective movement with populations, ecology and evolution'.
+Please find attached our manuscript entitled ``Eco-evolutionary dynamics and collective migration: implications for salmon metapopulation robustness'' co-authored by Justin Yeakel, Jean Philippe Gibert, Peter Westley, and Jonathan Moore, which we would like to submit for publication in the Philosophical Transactions of the Royal Society B themed issue \emph{Collective movement in ecology: From emerging technologies to management implications}, under the section title `Linking collective movement with populations, ecology and evolution'.
 
@@
   % \item We show that the portfolio effect and recovery time following a large disturbance, two independent measures of robustness that concern steady state and transient dynamics, respectively, are strongly correlated.
-  \item Increased habitat heterogeneity tends to increase robustness when rates of straying are low, but decrease robustness when rates are straying become large.
-Importantly, metapopulation robustness is generally increased by density dependent straying, regardless of habitat heterogeneity.
-  \item If highly dissimilar habitats are linked by a very low rate of straying (as might be expected if two sites are very distant from one another), we find that the extinction of the dominant population
+  \item Increased habitat heterogeneity tends to increase robustness when rates of straying are low, but decrease robustness when rates are straying are high. 
+  \item A key result of our analysis is that density dependent straying generally increases metapopulation robustness.
 \end{itemize}
 
-Our theory is general, relatively simple, and synthesizes a variety of fundamental ecological and evolutionary theories to show that our results are the natural consequence of a simple dynamical framework that incorporates the dispersal and local selection.
-We believe that our findings will be of general interest to ecologists, freshwater biologists, and evolutionary biologists, such that consideration for publication in a broad interest journal such as \emph{Philosophical Transactions} is warranted.
+Our theory is general, relatively simple, and synthesizes a variety of fundamental ecological and evolutionary theories to show that our results are the natural consequence of a simple dynamical framework that incorporates the dispersal of individuals between sites and local selection for a trait complex that determines productivity.
+We believe that our findings will be of general interest to ecologists, freshwater biologists, and evolutionary biologists, such that consideration for publication in \emph{Philosophical Transactions} is warranted.
 

### 2017-06-03 manuscript/msv4.tex
finalish draft

@@
 
-\usepackage{natbib}
-\bibpunct{(}{)}{,}{a}{}{;}
+\usepackage[comma,sort&compress]{natbib}
+% \bibpunct{[}{]}{,}{a}{}{;}
 
@@
 %\title{Community assembly and dynamics by the deconstruction of species interactions}
-\title{Eco-evolutionary dynamics and collective straying: implications for salmon metapopulation robustness}
+\title{Eco-evolutionary dynamics and collective migration: implications for salmon metapopulation robustness}
 \author{
@@
 ${}^3$College of Fisheries and Ocean Sciences, University of Alaska, Fairbanks, Fairbanks AK, USA \\
-${}^4$Earth${}_2$Oceans Research Group, Simon Fraser University, Vancouver BC, Canada \\
+${}^4$Earth${}_2$Oceans Research Group, Simon Fraser University, Burnaby BC, Canada \\
 ${}^*$To whom correspondence should be addressed: jdyeakel@gmail.com
@@
 Here we explore a minimal model of the eco-evolutionary dynamics between two populations connected by dispersal, where the productivity of each is defined by a trait complex that is subject to local selection.
-Although general in nature, our model is inspired by salmon metapopulations, where dispersal between populations is defined in terms of the \emph{straying rate}, which has been shown to be density dependent, possibly emerging as a consequence of collective migration.
+Although general in nature, our model is inspired by salmon metapopulations, where dispersal between populations is defined in terms of the \emph{straying rate}, which has been shown to be density dependent, and recently proposed to be shaped by social interactions consistent with collective movement.
+%possibly emerging as a consequence of collective migration.
 The results of our model reveal that increased straying between evolving populations leads to alternative stable states, which has large and nonlinear effects on two measures of metapopulation robustness: the portfolio effect and the time to recovery following an induced disturbance.
@@
 Intraspecific diversity can increase the resilience and stability of species or metapopulations. 
-This diversity-stability linkage can occur when there are asynchronous population dynamics, where the changes in population size varies across the metapopulation. 
+This diversity-stability linkage can occur when there are asynchronous population dynamics, where the changes in population size varies temporally across the metapopulation. 
 This asynchrony will increase the potential for demographic rescue \citep{Brown:1977gk,Earn:2000fm} and also decrease the variability of processes that integrate across the metapopulation \citep{Anonymous:2015gf}.
@@
 Migratory populations that return to breeding sites for reproduction are linked to each other by some proportion of the population that permanently disperses into the `wrong' site. 
-Recently, the role of social interactions and collective navigation has been hypothesized [This issue].
+Recently, the role of social interactions and collective navigation has been hypothesized \citep[][This issue]{Berdahl:2015kv,Berdahl:2016dx}.
 The rate at which individuals stray, $m$, may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \citep{Simons:2004jo,Berdahl:2015kv,Berdahl:2016dx}.
@@
 %alternative stable states & critical slowing
-A salient finding from our results was that straying can lead to the emergence of alternative stable states, pushing one of the populations to high density (the \emph{dominant state}), and one to low density (the \emph{subordinate population}).
+A salient finding from our results was that straying can lead to the emergence of alternative stable states, pushing one of the populations to high density (the \emph{dominant state}), and one to low density (the \emph{subordinate state}).
 This asymmetry in population densities also leads to an asymmetry in the mean trait values, both of which are skewed towards the local optimum of the dominant population. 
@@
 
-There is a `sweet spot' for straying that maximizes metapopulation robustness. 
+An intermediate straying rate maximizes metapopulation robustness. 
 Results from our model reveal that the presence of just enough straying to cause formation of alternative stable states increases the portfolio effect (figure \ref{fig:PE}a). 
@@
 In the case of near-collapse, although both populations inherit low population densities, the mean trait values of both are skewed towards the optimum of the dominant population.
-Due to density-dependent straying, low population densities evince greater dispersal, and while this increased connectivity primarily facilities the growth of the dominant population (because the trait means are closer to the dominant optimum), because the dominant population contains the bulk of the aggregate biomass, the overall recovery time is lessened significantly (figures \ref{fig:relax}b, \ref{extfig-fig:relaxtraj_bothlh}).
+Due to density-dependent straying, low population densities lead to greater dispersal, and while this increased connectivity primarily facilities the growth of the dominant population (because the trait means are closer to the dominant optimum), because the dominant population contains the bulk of the aggregate biomass, the overall recovery time is lessened significantly (figures \ref{fig:relax}b, \ref{extfig-fig:relaxtraj_bothlh}).
 % Third, in habitats with greater heterogeneity between sites (measured by an increase in $\Delta\theta$), density-dependent straying diminished robustness at low rates of straying (marginally lower portfolio effects and substantially longer return times; figures \ref{fig:thetaPE}b, \ref{fig:relaxtheta}a,b).
@@
 \centering
-\includegraphics[width=0.9\textwidth]{figs2/fig_MDPE_hm.pdf}
+\includegraphics[width=1\textwidth]{figs2/fig_MDPE_hm.pdf}
 \caption{
-(a) Total means $N_t$, (b) difference in means $\Delta N$, and (c) the portfolio effect PE as a function of heritability $h^2$ and a constant stray rate $m$. Light colors = high values.
+(a) Total means $N_t$, 
+(b) difference in means $\Delta N$, and 
+(c) the portfolio effect PE as a function of heritability $h^2$ and a constant stray rate $m$. Light colors = high values.
 The black line shows the fold bifurcation separating a single steady state (left) from alternative stable states (right).
-(d) The relationship between the time to recovery following a disturbance and the portfolio effect.
+(d) The relationship between the time to recovery following a disturbance and the portfolio effect. Portfolio effects greater than unity corresponds to less synchronization.
 The black line denotes the fold bifurcation.
@@
 (b) Median portfolio effects for habitats with increasing heterogeneity as measured by the difference in regional trait optima $\Delta \theta$ for both constant and density-dependent stray rates as shown in (a).
+Portfolio effects greater than unity corresponds to less synchronization.
 FB marks the fold bifurcation.

### 2017-08-31 manuscript/FinalDraft_rev/ReviewResponses.tex
revision

\documentclass[ucm,12pt]{ucletter}

\setlength\parindent{1cm}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{xcolor}

%\name{Justin D. Yeakel}
\telephone{(209)~285-9571}
\email{jyeakel@ucmerced.edu}

\newcounter{section}
\newcounter{subsection}[section]
\setcounter{secnumdepth}{3}
\makeatletter
\renewcommand\section{\@startsection {section}{1}{\z@}%
                                   {-3.5ex \@plus -1ex \@minus -.2ex}%
                                   {2.3ex \@plus.2ex}%
                                   {\normalfont\Large\bfseries}}
\newcommand\subsection{\@startsection{subsection}{2}{\z@}%
                                     {-3.25ex\@plus -1ex \@minus -.2ex}%
                                     {1.5ex \@plus .2ex}%
                                     {\normalfont\large\bfseries}}
\renewcommand \thesection{\@arabic\c@section}
\renewcommand\thesubsection{\thesection.\@arabic\c@subsection}

\makeatother

\begin{document}
\begin{letter}{
    Philosophical Transactions of the Royal Society B\\
    Publishing Section\\
    The Royal Society\\
    6-9 Carlton House Terrace\\
    London SW1Y 5AG\\
    England\\
    \centerline{\bf{Re: Eco-evolutionary dynamics and collective migration:}}\\
     \centerline{\bf{implications for salmon metapopulation robustness}}
}

\opening{To the editorial committee,}

\setstretch{1.2}

\section*{Reviewer I}
\subsection*{Major comments:}
\textcolor{cyan}{Equation (4), which describes the change in the population mean trait under dispersal and selection, is incorrect. In the second line of Eq (4), the immigrant term $(1-w_i) R_i\[\mu_j\]$ is independent of $\mu_i$, so that the derivative of this term is zero. This is as if selection did not change the allele frequencies/mean trait value of the immigrant subpopulation, which is obviously untrue (there is strong selection on the locally maladapted immigrants, favouring those who happen to be closer to the local optimum). One cannot apply Lande's equation without adapting it to the immigration model.}

We appreciate the Reviewer's concern, however note that equation 4) is written as
\begin{align}
  \label{eq:mu}
  \mu_i(t+1) &= w_i\mu_i(t) + (1-w_i)\mu_j(t) \\ \nonumber
  &+ h^2\sigma^2\frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)] + (1-w_i)R_i[\mu_j(t)]  \right),
\end{align}
and because the third term on the right is the derivative of the log of the sum (as it must be for the discrete time version of the Lande equation; Lande \emph{Evolution}, 1976), the derivative of the second term within the logarithm is not zero, and both phenotypic means contribute to the strength and direction of selection.

\textcolor{cyan}{
The model with constant dispersal ($m$) is very much the same as the model of Ronce and Kirkpatrick (2001, Evolution), with the difference that the latter is fully deterministic and set in continuous time; but as long as the system attains an equilibrium rather than population cycles or chaos, continuous time should not make a difference. Ronce and Kirkpatrick focused on equilibria (such as the dominant vs subordinate state here ) and on local adaptation, not metapopulation resilience/stability, but did study the consequence of disturbances for switching between alternative equilibria (see the next point).
}

ff

\textcolor{cyan}{
The bifurcation the authors call a fold bifurcation is not a fold bifurcation. Figure 1 suggests a pitchfork bifurcation (which is degenerate but will appear in symmetric models). An alternative possibility is that there exists an asymmetric equilibrium with a "dominant state" and a "subordinate state" also at low m, next to the symmetric equilibrium shown at low m in figure 1, so that the system has two stable equilibria; here the "FB" point in figure 1 would be the point where the symmetric equilibrium is destabilized (this is the pattern found by Ronce and Kirkpatrick). This second alternative seems likely to me because the abrupt changes seen in figure 1 and also in figure 5 may well indicate an attractor switch. The nature of the bifurcation should be properly investigated (one can find Jacobians etc with numerical analysis for the deterministic system).
}

ff

\textcolor{cyan}{
The ms does not describe how recovery time was measured. The deterministic system returns to an equilibrium asymptotically, i.e., in infinite time. For practice, one defines a return ``close enough" for recovery. The details of this should be given, otherwise the results are not reproducible. Moreover, I think that near the bifurcation point total population size will equilibrate fast, but the distribution of individuals over the two habitats takes a longer time to equilibrate (the system is on the edge between a symmetric and an asymmetric equilibrium). If time to recovery includes the time that $N_1$ and $N_2$ individually reach their steady states, then this can explain the peak of recovery time at the bifurcation; but this does not mean that the total population size takes long to recover. If total population size was monitored, but recovery was interpreted as "return to original" whereas the system attains a different equilibrium (symmetric vs asymmetric, see above), then obviously recovery time will be infinite, but again this does not mean that the population does not recover from near extinction.
}

ff

\textcolor{cyan}{
A strong "portfolio effect" as measured by PE in equation 6 indicates resilience against metapopulation extinction if PE is high due to a low CV of total population size. But PE is also high if the local populations have high CV. As I argued above, near the bifurcation N_1 and N_2 may be much less stable than $N_T=N_1+N_2$, i.e., the peak of PE may have little to do with the resilience of total population size.
}

hehehe

\subsection*{Minor comments:}
\textcolor{cyan}{``straying" is not defined in the Introduction/main text. Since the model is not specific to salmon, the more general term "dispersal" could be used throughout.}

\textcolor{cyan}{The figure legends should specify the parameter values used, figure 2 should give a key translating colours to values, figure 3 should have numbers on the axes. Generally, the authors should pay attention to reproducibility.}

\section*{Reviewer II}
\subsection*{Major comments}
My major criticism on this paper is that in the analysis presented, the bifurcation point is incorrectly identified and the claims based on this misidentification are most likely unsupported. What is presented as a Fold bifurcation is in reality a Pitchfork bifurcation. Pitchfork bifurcations are dependent on perfect symmetry in systems, something that is also true in the current case: The model as presented and analyzed is formulated in an entirely symmetrical way, which means that a Pitchfork bifurcation is de facto built into it.

As it is, the occurrence of the bifurcation point as well as the ecological implications derived from it, are mostly a mathematical construction, brought about by the complete symmetry in the system. It is absolutely necessary to investigate the robustness of the results under scenarios where the symmetry is broken. The eco-evolutionary aspects of the model have currently no role in the observed dynamics, because they hinge on the symmetrical formulation.

I would be very curious to see this analysis extended under asymmetrical assumptions, for example in the productivity of the different locations or the intraspecific competition between populations. Another potentially highly interesting element that has not been studied here is the potential for alternative stable states based on the density-dependent straying rate (subsection b in the Model formulation). The study currently addresses a scenario where the potential for population collapse due to collective navigation is ignored. I see that as the interesting angle; how do the dynamics caused by density-dependent straying interact with the dynamics caused by selective forces as imposed by the different environments on recruitment?

This manuscript can be improved by correcting the bifurcation analysis and by extending beyond perfect symmetry. The authors take a reasonable approach, using numerical analysis, which is fine, and I like the idea to focus on population robustness.  I understand that recruitment and population growth are not explicitly linked to environmental resources for reasons of mathematical simplicity. At the same time, I would prefer to see resources as an explicit variable in this model. This suggestion could directly be used to introduce asymmetries between two different locations and may be a feasible way to address the issues I raised above.

I would be looking forward to see a revised version of this manuscript that addresses my points.

It would be good to include more technical details about how the bifurcation analysis was carried out (or is this present in the missing SI text?). As it is, I have no idea about the numerical techniques that were used or even whether the authors have studied both increasing and decreasing parameter values, as necessary to enable detection of alternative stable states and discontinuities.

Please include a table with the definitions and values of model parameters (including their units) and definitions of model variables.

I could not understand section c, Habitat heterogeneity in the Model formulation. In particular the paragraph on lines 139-144 should be explained and justified better, where the authors should make sure to explain the ‘integrate the two variables’ part more formally and explicitly.

As I write above, I appreciate the focus on robustness and through numerical analysis. At the same time this model would lend itself to some analysis too, even making some simplifying assumptions or investigating some boundary conditions. That would strengthen the insights from the results and the conclusions based on them. This would also possibly facilitate very straightforward validation of the kind of bifurcation point that is encountered (although the figures clearly show it cannot be a Fold bifurcation).

Please present formal definitions (for example in a table as mentioned above) of all model variables and parameters. Right now I never found an explicit definition of NT, for example.

\subsection*{Minor comments}
I encountered some sloppiness in various parts of the manuscript, for example figure 3 is missing axes labels and has inconsistencies with the main text and caption. I did not find the SI figures or text, only the figure captions. Figure 4 was referenced before figure 3, as was the case with figures S2 and S1. Please see below for a list of textual errors that should be addressed (preferable before review, through using a spell checker). Note that these are the ones I could quickly jot down, but there are more.
Typos/errors on lines:
\begin{itemize}
\item 67:   ‘This model’ does not link back to anything about a model
\item 72:   determining
\item 90:   ‘than’ used awkwardly
\item 92:   inconsistency in number between noun and verb (equation that determine…)
\item 103  ‘is’ missing
\item 133: thgat
\item 139: differnce
\item 153:  inconsistency in number between noun and verb
\end{itemize}

\vspace{5mm}

\singlespacing
\closing{Sincerely,\\
\fromsig{\includegraphics[scale=0.2]{signature.jpg}}\\
\fromname{
Justin D. Yeakel\\
Assistant Professor\\
University of California, Merced}
}

\end{letter}
\end{document}

### 2017-08-31 model/KevinExtinction.jl

@@ -20,12 +20,12 @@ beta=0.001;
 theta1=5.0;
 thetadiff=5.0;
 tau=1.0;
-h=0.2;
+h=0.8;
 sigmaE=0;
 sigmaG=1;
 perror=0.01;
 refuge=0.01;
-m=0.1;
+m=0.4;
 
 extpop = "both";
 t_ext = Int64(round(tmax/2));
@@ -50,17 +50,22 @@ refuge
 
 t_ss, relaxtime = timeSS(n1,n2,t_ext);
 
-
+namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/Manuscript/figs2/fig_recovery.pdf");
 R"""
 library(RColorBrewer)
 pal = brewer.pal(3,'Set1')
-par(mfrow=c(2,1),mai = c(0.8, 0.8, 0.4, 0.1))
-plot($n1,type='l',col=pal[1],xlim=c($t_ext,$t_ss+50),ylim=c(0,max($n1,$n2)),ylab='Population density',xlab='Time',main=paste(c('RT=',$relaxtime)))
+pdf($namespace,height=5,width=6)
+par(mfrow=c(1,1),mai = c(0.8, 0.8, 0.4, 0.1))
+plot($n1,type='l',col=pal[1],xlim=c($t_ext,$t_ss+50),ylim=c(0,max($n1+$n2)),ylab='Population density',xlab='Time')
 lines($n2,col=pal[2])
+lines($n1+$n2,col='black')
 lines(c($t_ss,$t_ss),c(-10,10000))
-plot($x1,type='l',col=pal[1],ylim=c(5,10),xlim=c($t_ext,$t_ss+50),ylab='Trait means',xlab='Time')
-lines($x2,col=pal[2])
-lines(c($t_ss,$t_ss),c(0,10000))
+points($t_ss,$(n1[t_ss])+$(n2[t_ss]),pch=16)
+text(x=$t_ss+18,y=1,paste(c('Recovery time=',$relaxtime),collapse=''))
+#plot($x1,type='l',col=pal[1],ylim=c(5,10),xlim=c($t_ext,$t_ss+50),ylab='Trait means',xlab='Time')
+#lines($x2,col=pal[2])
+#lines(c($t_ss,$t_ss),c(0,10000))
+dev.off()
 """
 
 namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/Manuscript/figs2/fig_relax_both_lowh.pdf");

### 2017-08-31 model/KevinSimulations2_asym.jl

@@ -1,0 +1,859 @@
+using Distributions
+using RCall
+
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve_asym.jl")
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinJacobian.jl")
+
+asym=0.0;
+tmax=10000;
+z=0.5;
+rmax=2.0;
+beta=0.001;
+theta1=5.0;
+thetadiff=5.0;
+tau=1.0;
+h=0.49;
+sigmaE=0;
+sigmaG=1.0;
+m=0.35;
+perror=0.01;
+n1, n2, x1, x2, w1, w2 = 
+KevinEvolve_asym(
+  tmax,
+  z,
+  rmax,
+  beta,
+  theta1,
+  thetadiff,
+  tau,
+  h,
+  sigmaE,
+  sigmaG,
+  m, #m
+  perror,
+  asym
+  );
+  
+  R"""
+  par(mfrow=c(1,2))
+  library(RColorBrewer)
+  cols = brewer.pal(3,'Set1')
+  plot($(n1),type="l",col=cols[1],ylim=c(0,max(c(as.vector($n1),as.vector($n2)))))
+  lines($(n2),col=cols[2])
+  """
+  
+  R"""
+  library(RColorBrewer)
+  cols = brewer.pal(3,'Set1')
+  plot($(x1),type="l",col=cols[1],ylim=c(4,12))
+  lines($(x2),col=cols[2])
+  """
+
+
+#CalculateJacobian
+Jac = KevinJacobian(last(n1),last(n2),last(x1),last(x2),
+z,rmax,beta,theta1,thetadiff,tau,sigma,m)
+eigvals(Jac)
+
+
+
+
+
+
+#Analysis over m
+tmax=10000;
+mvec = collect(0.0:0.001:0.5);
+n1ts = zeros(Float64,length(mvec),tmax);
+n2ts = zeros(Float64,length(mvec),tmax);
+n1mean=zeros(Float64,length(mvec));
+n2mean=zeros(Float64,length(mvec));
+n1sd=zeros(Float64,length(mvec));
+n2sd=zeros(Float64,length(mvec));
+aggmean=zeros(Float64,length(mvec));
+aggsd=zeros(Float64,length(mvec));
+x1mean=zeros(Float64,length(mvec));
+x2mean=zeros(Float64,length(mvec));
+pe=zeros(Float64,length(mvec));
[…]

### 2017-08-31 model/src/KevinEvolve_asym.jl

@@ -13,12 +13,14 @@ function KevinEvolve_asym(
   perror,
   asym #Ranges from 0 to 0.5 (at which point rmax will be 100% part)
   )
+  sgn = [-1,1];
+  sr = rand(sgn);
+  rmax1 = rmax*(1+sr*asym);
+  rmax2 = rmax1*(1-sr*asym);
   
-  rmax1 = rmax*(1-asym);
-  rmax2 = rmax1*(1+asym);
-  
-  beta1 = beta*(1-asym);
-  beta2 = beta*(1+asym);
+  sb = rand(sgn);
+  beta1 = beta*(1+sb*asym);
+  beta2 = beta*(1-sb*asym);
   
   #Initialize variables
   n1 = zeros(Float64,tmax); n1[1]=2;

### 2017-09-05 manuscript/FinalDraft_rev/ReviewResponses.tex
asymmetry and responses

@@
 \subsection*{Major comments:}
-\textcolor{cyan}{Equation (4), which describes the change in the population mean trait under dispersal and selection, is incorrect.
-In the second line of Eq (4), the immigrant term $(1-w_i) R_i\[\mu_j\]$ is independent of $\mu_i$, so that the derivative of this term is zero.
+\noindent \textcolor{cyan}{
+{\bf Comment 1} Equation (4), which describes the change in the population mean trait under dispersal and selection, is incorrect.
+In the second line of Eq (4), the immigrant term $(1-w_i) R_i [\mu_j]$ is independent of $\mu_i$, so that the derivative of this term is zero.
 This is as if selection did not change the allele frequencies/mean trait value of the immigrant subpopulation, which is obviously untrue (there is strong selection on the locally maladapted immigrants, favouring those who happen to be closer to the local optimum).
-One cannot apply Lande's equation without adapting it to the immigration model.}
+One cannot apply Lande's equation without adapting it to the immigration model.
+}
 
-We appreciate the Reviewer's concern, however note that equation 4) is written as
+\noindent {\bf Response 1} We appreciate the Reviewer's concern, however note that equation 4) is written as
 \begin{align}
@@
 
-\textcolor{cyan}{
-The model with constant dispersal ($m$) is very much the same as the model of Ronce and Kirkpatrick (2001, Evolution), with the difference that the latter is fully deterministic and set in continuous time; but as long as the system attains an equilibrium rather than population cycles or chaos, continuous time should not make a difference.
+\noindent \textcolor{cyan}{
+{\bf Comment 2} The model with constant dispersal ($m$) is very much the same as the model of Ronce and Kirkpatrick (2001, Evolution), with the difference that the latter is fully deterministic and set in continuous time; but as long as the system attains an equilibrium rather than population cycles or chaos, continuous time should not make a difference.
 Ronce and Kirkpatrick focused on equilibria (such as the dominant vs subordinate state here ) and on local adaptation, not metapopulation resilience/stability, but did study the consequence of disturbances for switching between alternative equilibria (see the next point).
@@
 
-ff
+\noindent {\bf Response 2} We are thankful for the Reviewer in pointing out this reference, which we are embarrassed to admit to having overlooked.
+The model indeed has many similarities to our own, and we now acknowledge its importance and priority within the manuscript.
+The similarities in the resulting dynamics suggest that what we are observing in the two-population salmon system, as well as the implications that we derive for metapopulation robustness and recovery, may be generally relevant and illustrative of a general phenomenon outside of the constraints introduced in the specific formulation of our model.
 
-\textcolor{cyan}{
-The bifurcation the authors call a fold bifurcation is not a fold bifurcation.
+\noindent \textcolor{cyan}{
+{\bf Comment 3} The bifurcation the authors call a fold bifurcation is not a fold bifurcation.
 Figure 1 suggests a pitchfork bifurcation (which is degenerate but will appear in symmetric models).
@@
 
-ff
+\noindent {\bf Response 3} We agree with the Reviewer that the bifurcation is indeed a Pitchfork bifurcation and thank the Reviewer for pointing to this error. We provide an analysis?
 
-\textcolor{cyan}{
-The ms does not describe how recovery time was measured.
+\noindent \textcolor{cyan}{
+{\bf Comment 4} The ms does not describe how recovery time was measured.
 The deterministic system returns to an equilibrium asymptotically, i.e., in infinite time.
@@
 
-ff
+\noindent {\bf Response 4} We thank the Reviewer and apologize for our lack of clarity.
+We emphasize that our measure of recovery time is a numerical measurement, as eigenvalue-based (asymptotic) measures are not appropriate for the scale of disturbance that we introduce (extinction in some cases - far from the non-trivial fixed point(s) of the system).
+The procedure is now described in more detail within the text, and illustrated graphically in an additional supplemental figure (Fig.
+SXX).
+In short, we track the recovering aggregate population $N_T=N_1+N_2$ until it is within a small band of variation around the \emph{new} fixed point (thus accounting for fixed points that are altered by the disturbance).
+If it stays within this narrow range, it is counted as recovered.
+Accordingly, trajectories that cycle around the fixed point are not counted as recovered until the cycles decay.
+To ensure reproducibility, all code is provided online in the Github repository: https://github.com/jdyeakel/SalmonStrays.
 
-\textcolor{cyan}{
-A strong "portfolio effect" as measured by PE in equation 6 indicates resilience against metapopulation extinction if PE is high due to a low CV of total population size.
+\noindent \textcolor{cyan}{
+{\bf Comment 5} A strong "portfolio effect" as measured by PE in equation 6 indicates resilience against metapopulation extinction if PE is high due to a low CV of total population size.
 But PE is also high if the local populations have high CV.
-As I argued above, near the bifurcation N_1 and N_2 may be much less stable than $N_T=N_1+N_2$, i.e., the peak of PE may have little to do with the resilience of total population size.
+As I argued above, near the bifurcation $N_1$ and $N_2$ may be much less stable than $N_T=N_1+N_2$, i.e., the peak of PE may have little to do with the resilience of total population size.
 }
 
-hehehe
+\noindent {\bf Response 5} As measured here, PE is tracking the potential for the aggregate to buffer variance observed in the constituent populations.
+Near the bifurcation, if $N_1$ and $N_2$ are more variable relative to $N_T$, then PE should be high, and this is what we are trying to measure.
+However we agree that PE can bounce around in response to changes to any of the four attributes from which it is composed, so it is sometimes hard to evaluate its relationship to more descriptive measures of resilience such as recovery time.
+This was one of the primary motivations for incorporating both measures in our analysis.
+
+% While we agree with the Reviewer that the spike in PE at the bifurcation is not necessarily descriptive of a resilient population, the attributes that lead to an increase in the PE to the right of the bifurcation (higher $m$) also results in shorter recovery times, and is measuring the potential for the aggregate to buffer variance.
+We do not put much weight in the PE spike at the bifurcation as an indicator of metapopulation health
 
 \subsection*{Minor comments:}
-\textcolor{cyan}{``straying" is not defined in the Introduction/main text. Since the model is not specific to salmon, the more general term "dispersal" could be used throughout.}
+\noindent \textcolor{cyan}{
+{\bf Minor Comment 1} ``straying" is not defined in the Introduction/main text. Since the model is not specific to salmon, the more general term "dispersal" could be used throughout.}
 
-\textcolor{cyan}{The figure legends should specify the parameter values used, figure 2 should give a key translating colours to values, figure 3 should have numbers on the axes.
+\noindent {\bf Minor Response 1} We now equate straying with dispersal early in the text (and note that it is also defined in the abstract).
+
+\noindent \textcolor{cyan}{
+{\bf Minor Comment 2} The figure legends should specify the parameter values used, figure 2 should give a key translating colours to values, figure 3 should have numbers on the axes.
 Generally, the authors should pay attention to reproducibility.}
+
+\noindent {\bf Minor Response 2} Parameter values that we used are now provided, though some issues with the figures appear to have been caused by an issue with the upload.
+Moreover, to ensure reproducibility, all code that was used is provided in an online, publicly available, Github repository https://github.com/jdyeakel/SalmonStrays.
 
@@
 \subsection*{Major comments}
-My major criticism on this paper is that in the analysis presented, the bifurcation point is incorrectly identified and the claims based on this misidentification are most likely unsupported.
+\noindent \textcolor{cyan}{
+{\bf Comment 1} My major criticism on this paper is that in the analysis presented, the bifurcation point is incorrectly identified and the claims based on this misidentification are most likely unsupported.
 What is presented as a Fold bifurcation is in reality a Pitchfork bifurcation.
 Pitchfork bifurcations are dependent on perfect symmetry in systems, something that is also true in the current case: The model as presented and analyzed is formulated in an entirely symmetrical way, which means that a Pitchfork bifurcation is de facto built into it.
+}
 
-As it is, the occurrence of the bifurcation point as well as the ecological implications derived from it, are mostly a mathematical construction, brought about by the complete symmetry in the system.
+\noindent \textcolor{cyan}{
+{\bf Comment 1} As it is, the occurrence of the bifurcation point as well as the ecological implications derived from it, are mostly a mathematical construction, brought about by the complete symmetry in the system.
 It is absolutely necessary to investigate the robustness of the results under scenarios where the symmetry is broken.
 The eco-evolutionary aspects of the model have currently no role in the observed dynamics, because they hinge on the symmetrical formulation.
+}
 
-I would be very curious to see this analysis extended under asymmetrical assumptions, for example in the productivity of the different locations or the intraspecific competition between populations.
+\noindent \textcolor{cyan}{
+{\bf Comment 1} I would be very curious to see this analysis extended under asymmetrical assumptions, for example in the productivity of the different locations or the intraspecific competition between populations.
 Another potentially highly interesting element that has not been studied here is the potential for alternative stable states based on the density-dependent straying rate (subsection b in the Model formulation).
@@
 I see that as the interesting angle; how do the dynamics caused by density-dependent straying interact with the dynamics caused by selective forces as imposed by the different environments on recruitment?
+}
 
-This manuscript can be improved by correcting the bifurcation analysis and by extending beyond perfect symmetry.
+\noindent \textcolor{cyan}{
+{\bf Comment 1} This manuscript can be improved by correcting the bifurcation analysis and by extending beyond perfect symmetry.
 The authors take a reasonable approach, using numerical analysis, which is fine, and I like the idea to focus on population robustness.
@@
 This suggestion could directly be used to introduce asymmetries between two different locations and may be a feasible way to address the issues I raised above.
+}
 
-I would be looking forward to see a revised version of this manuscript that addresses my points.
+\noindent \textcolor{cyan}{
+{\bf Comment 1} It would be good to include more technical details about how the bifurcation analysis was carried out (or is this present in the missing SI text?).
+As it is, I have no idea about the numerical techniques that were used or even whether the authors have studied both increasing and decreasing parameter values, as necessary to enable detection of alternative stable states and discontinuities.
+}
 
-It would be good to include more technical details about how the bifurcation analysis was carried out (or is this present in the missing SI text?).
-As it is, I have no idea about the numerical techniques that were used or even whether the authors have studied both increasing and decreasing parameter values, as necessary to enable detection of alternative stable states and discontinuities.
+\noindent \textcolor{cyan}{
+{\bf Comment 1} Please include a table with the definitions and values of model parameters (including their units) and definitions of model variables.
+}
 
-Please include a table with the definitions and values of model parameters (including their units) and definitions of model variables.
+\noindent \textcolor{cyan}{
+{\bf Comment 1} I could not understand section c, Habitat heterogeneity in the Model formulation.
+In particular the paragraph on lines 139-144 should be explained and justified better, where the authors should make sure to explain the ‘integrate the two variables’ part more formally and explicitly.
+}
 
-I could not understand section c, Habitat heterogeneity in the Model formulation.
-In particular the paragraph on lines 139-144 should be explained and justified better, where the authors should make sure to explain the ‘integrate the two variables’ part more formally and explicitly.
-
-As I write above, I appreciate the focus on robustness and through numerical analysis.
+\noindent \textcolor{cyan}{
+{\bf Comment 1} As I write above, I appreciate the focus on robustness and through numerical analysis.
 At the same time this model would lend itself to some analysis too, even making some simplifying assumptions or investigating some boundary conditions.
@@
 This would also possibly facilitate very straightforward validation of the kind of bifurcation point that is encountered (although the figures clearly show it cannot be a Fold bifurcation).
+}
 
-Please present formal definitions (for example in a table as mentioned above) of all model variables and parameters. Right now I never found an explicit definition of NT, for example.
+Point to Jacobian analysis.
+
+\noindent \textcolor{cyan}{
+{\bf Comment 1} Please present formal definitions (for example in a table as mentioned above) of all model variables and parameters.
+Right now I never found an explicit definition of $N_T$, for example.
+}
 
 \subsection*{Minor comments}
-I encountered some sloppiness in various parts of the manuscript, for example figure 3 is missing axes labels and has inconsistencies with the main text and caption.
+\noindent \textcolor{cyan}{
+{\bf Comment 1} I encountered some sloppiness in various parts of the manuscript, for example figure 3 is missing axes labels and has inconsistencies with the main text and caption.
 I did not find the SI figures or text, only the figure captions.
@@
 \end{itemize}
+}
+
+We thank the Reviewer for being so meticulous and pointing out these errors.
+Although the noted issues with the figures (missing axes, labels) and missing supplemental materials appears to have been a problem with the upload, we acknowledge that the other issues were due to a lack of thoroughness on our part and have been corrected.
 

### 2017-09-05 model/JacobianAnalysis.jl

@@ -1,0 +1,139 @@
+using Distributions
+using RCall
+
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve_asym.jl")
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinJacobian.jl")
+
+
+
+#Analysis over m
+tmax=10000;
+mvec = collect(0.0:0.001:0.5);
+n1ts = zeros(Float64,length(mvec),tmax);
+n2ts = zeros(Float64,length(mvec),tmax);
+n1mean=zeros(Float64,length(mvec));
+n2mean=zeros(Float64,length(mvec));
+n1sd=zeros(Float64,length(mvec));
+n2sd=zeros(Float64,length(mvec));
+aggmean=zeros(Float64,length(mvec));
+aggsd=zeros(Float64,length(mvec));
+x1mean=zeros(Float64,length(mvec));
+x2mean=zeros(Float64,length(mvec));
+pe=zeros(Float64,length(mvec));
+eigs = Array(Array{Complex{Float64}},length(mvec));
+maxeigs = Array{Float64}(length(mvec));
+mineigs = Array{Float64}(length(mvec));
+maximeigs = Array{Float64}(length(mvec));
+minimeigs = Array{Float64}(length(mvec));
+
+z=0.5;
+rmax=2.0;
+beta=0.001;
+theta1=5.0;
+thetadiff=4;
+tau=1.0;
+h=0.5;
+sigmaE=0;
+sigmaG=1;
+perror=0.01;
+
+burnin=0.80
+@time for i=1:length(mvec)
+  m=mvec[i];
+  
+  n1, n2, x1, x2, w1, w2 = 
+  KevinEvolve(
+    tmax,
+    z,
+    rmax,
+    beta,
+    theta1,
+    thetadiff,
+    tau,
+    h,
+    sigmaE,
+    sigmaG,
+    m,
+    perror
+    );
+
+  n1trim = n1[Int64(floor(tmax*burnin)):tmax];
+  n2trim = n2[Int64(floor(tmax*burnin)):tmax];
+  x1trim = x1[Int64(floor(tmax*burnin)):tmax];
+  x2trim = x2[Int64(floor(tmax*burnin)):tmax]
+  
+  n1ts[i,:] = n1;
+  n2ts[i,:] = n2;
+  
[…]

### 2017-09-05 model/src/KevinEvolve_asym.jl

@@ -13,14 +13,19 @@ function KevinEvolve_asym(
   perror,
   asym #Ranges from 0 to 0.5 (at which point rmax will be 100% part)
   )
-  sgn = [-1,1];
-  sr = rand(sgn);
-  rmax1 = rmax*(1+sr*asym);
-  rmax2 = rmax1*(1-sr*asym);
+  # sgn = [-1,1];
+  rv1 = 0.0;
+  rv2 = 0.0;
+  if asym > 0.0
+      rv1 = rand(Normal(0,asym));
+      rv2 = rand(Normal(0,asym));
+  end
+  rmax1 = rmax*(1+rv1);
+  rmax2 = rmax1*(1+rv2);
   
-  sb = rand(sgn);
-  beta1 = beta*(1+sb*asym);
-  beta2 = beta*(1-sb*asym);
+  # sb = rand(sgn);
+  beta1 = beta*(1+rv1);
+  beta2 = beta*(1+rv2);
   
   #Initialize variables
   n1 = zeros(Float64,tmax); n1[1]=2;

### 2017-09-06 Manuscript/FinalDraft_rev/ReviewResponses.tex
?3

--- 2017-09-05 manuscript/FinalDraft_rev/ReviewResponses.tex
@@
 
-\noindent {\bf Response 3} We agree with the Reviewer that the bifurcation is indeed a Pitchfork bifurcation and thank the Reviewer for pointing to this error. We provide an analysis?
+\noindent {\bf Response 3} Again we thank the Reviewer for pointing out an error in that we incompletely identified the bifurcation as a `Fold bifurcation'.
+However, we do not think that the bifurcation is a Pitchfork bifurcation, as in discrete time systems, a discrete pitchfork is synonymous with a flip (or period-doubling) bifurcation.
+Instead, we believe
 

### 2017-09-06 Manuscript/FinalDraft_rev/msv6.tex
?3

--- 2017-06-03 manuscript/msv4.tex
@@
 \captionsetup{compatibility=false}
-\usepackage{xr}
-\externaldocument[extfig-]{supp}
+% \usepackage{xr}
+% \externaldocument[extfig-]{supp}
 %\usepackage{epstopdf}
@@
 
-\bibliographystyle{prsb}
+% \bibliographystyle{prsb}
 
@@
 The spatial dispersal of individuals is known to play an important role in the dynamics of populations, and is central to metapopulation theory.
-At the same time, local adaptation to different environmental conditions creates a geographic mosaic of evolutionary forces, where the combined drivers of selection and gene flow interact.
-Although the dispersal of individuals from donor to recipient populations provides connections within the metapopulation, potentially promoting demographic and evolutionary rescue, it may also introduce maladapted individuals into habitats that are host to different environmental conditions, potentially lowering the mean fitness of the recipient population.
+At the same time, local adaptation to environmental conditions creates a geographic mosaic of evolutionary forces, where the combined drivers of selection and gene flow interact.
+Although the dispersal of individuals from donor to recipient populations provides connections within the metapopulation, promoting demographic and evolutionary rescue, it may also introduce maladapted individuals into habitats host to different environmental conditions, potentially lowering the fitness of the recipient population.
 Here we explore a minimal model of the eco-evolutionary dynamics between two populations connected by dispersal, where the productivity of each is defined by a trait complex that is subject to local selection.
@@
 The results of our model reveal that increased straying between evolving populations leads to alternative stable states, which has large and nonlinear effects on two measures of metapopulation robustness: the portfolio effect and the time to recovery following an induced disturbance.
-We show that intermediate levels of straying result in large gains in robustness, and that increased differences between sites (increased habitat heterogeneity) promotes robustness when straying rates are low, and erodes robustness when straying rates are high.
+We show that intermediate levels of straying result in large gains in robustness, and that increased habitat heterogeneity promotes robustness when straying rates are low, and erodes robustness when straying rates are high.
 Finally, we show that density-dependent straying promotes robustness, particularly when the aggregate biomass is low and straying is correspondingly high, and this has important ramifications for the conservation of salmon metapopulations facing both natural and anthropogenic disturbances.
-\end{abstract} %250 words
+\end{abstract}
 
 \maketitle
+
+\centerline{Salmon metapopulations, Straying, Dispersal, Eco-evolutionary dynamics, Alternative stable states}
+\vspace{2mm}
+\noindent {\bf Media Summary} The rate at which individuals move within a metapopulation has large impacts on extinction risk.
+Among migratory species such as salmon, the rate of dispersal, or straying, may be density dependent due to collective decision-making.
+We explore an eco-evolutionary model of two populations connected by dispersal, where the productivity of each is defined by a trait complex subject to local selection.
+We show that intermediate straying rates result in large gains in robustness, and that density-dependent straying promotes robustness.
+These results have important ramifications for the conservation of salmon metapopulations facing both natural and anthropogenic disturbances.\\
 
@@
 \begin{equation}
-\langle{\rm PE}\rangle =\frac{1}{X}\sum_{i=1}^{X} \frac{\sqrt{{\rm VAR}(N_i)}}{{\rm E}(N_i)}\cdot \frac{{\rm E}(N_T)}{\sqrt{{\rm VAR}(N_T)}},
+\langle{\rm PE}\rangle =\frac{1}{X}\sum_{i=1}^{X} \frac{\sqrt{{\rm VAR}(N_i^*)}}{{\rm E}(N_i^*)}\cdot \frac{{\rm E}(N_T^*)}{\sqrt{{\rm VAR}(N_T^*)}},
 \label{eq:pe}
@@
 \noindent where in this case the number of populations is limited to $X=2$ and the expectations $\rm E(\cdot)$ and variances $\rm VAR(\cdot)$ are evaluated at the steady state.
-As the CV of $N_T$ decreases relative to that of the constituent populations, $\langle{\rm PE}\rangle > 1$, and the metapopulation is presumed to become more stable.
+The steady state condition is denoted by `$*$'.
+As the CV of $N_T^*$ decreases relative to that of the constituent populations, $\langle{\rm PE}\rangle > 1$, and the metapopulation is presumed to become more stable.
 Portfolio effects greater than unity corresponds to less synchronization  \citep{Loreau:2008ju,Anderson:2014cx,Yeakel:2013vz} and thus a greater potential for demographic rescue among populations, buffering the system as a whole against extinction.
@@
 %Although there are more robust ways to measure the portfolio effect (in particular for comparing the PE across species with different mean-variance relationships; REF: Anderson), its only utlity here is to provide a simple statistical measure of metapopulation persistence.
-A more direct way to measure system robustness is to measure the time that the system (measured as the aggregate biomass $N_T$) takes to recover its steady state abundance following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
+A more direct way to measure system robustness is to measure the time that the system (measured as the aggregate steady state biomass $N_T^*$) takes to recover its steady state abundance following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
 Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation \citep{GuckHolmes}, because we aimed to 1) assess the effects of a large perturbation, and 2) estimate the time required for all transient effects to decay (including dampened oscillations), we used a simulation-based numerical procedure.
@@
 
-Trait heritability has a large impact on the degree to which straying rate affects both the aggregate population steady state density ($N^*_T=N^*_1+N^*_2$; figure \ref{fig:PE}a) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N=|N^*_1-N^*_2|$; figure \ref{fig:PE}b).
-Greater trait heritability results in a faster decline in $N_T^*$ with increasing straying rates $m$, but leads to only moderate changes to $\Delta N$.
-Conversely, in the context of lower trait heritability, an increase in the straying rate has little impact on the total biomass density but contrastingly large effects on $\Delta N$.
+Trait heritability has a large impact on the degree to which straying rate affects both the aggregate population steady state density ($N^*_T=N^*_1+N^*_2$; figure \ref{fig:PE}a) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N^*=|N^*_1-N^*_2|$; figure \ref{fig:PE}b).
+Greater trait heritability results in a faster decline in $N_T^*$ with increasing straying rates $m$, but leads to only moderate changes to $\Delta N^*$.
+Conversely, in the context of lower trait heritability, an increase in the straying rate has little impact on the total biomass density but contrastingly large effects on $\Delta N^*$.
 The fold bifurcation (the black line in Figs.
@@
 This amplification in variance is the product of a dynamical process known as \emph{critical slowing down} that occurs near fold bifurcations \citep{Scheffer:2009gg}, a phenomenon that some have suggested may serve as an early warning indicator for approaching phase transitions \citep{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br,Krkosek:2014ch}.
-For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}a-c), where alternative stable states occur, the portfolio effect declines steadily as the CV of $N_T$ increases.
+For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}a-c), where alternative stable states occur, the portfolio effect declines steadily as the CV of $N_T^*$ increases.
 The decline over $m$ is more gradual if trait heritability is low, and steeper if trait heritability is high (figure \ref{fig:PE}c).
@@
 Straying had non-linear impacts on the recovery time of populations. 
-When the dominant state (well adapted and high density) is wiped out, high levels of straying allow it to recover quickly (figure \ref{fig:relax}a) because the surviving population has a mean trait value skewed towards the optimum of the recovering population (figure \ref{extfig-fig:relaxtraj_hdlh}).
+When the dominant state (well adapted and high density) is wiped out, high levels of straying allow it to recover quickly (figure \ref{fig:relax}a) because the surviving population has a mean trait value skewed towards the optimum of the recovering population (figure \ref{fig:relaxtraj_hdlh}).
 Yet, as straying decreases, recovery time for the disturbed dominant state population increases, in part because there is enough time for the trait distribution to move back towards the trait optimum of the subordinate state population.
 In contrast, when the subordinate state population (maladapted and low density) is wiped out, recovery rates are fastest at low to intermediate levels of straying.
-Because the mean trait values of both populations are skewed towards those of the dominant population, when the subordinate population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into the recovering population extends the length of time required for it to return to its steady state (figure \ref{extfig-fig:relaxtraj_ldlh}).
+Because the mean trait values of both populations are skewed towards those of the dominant population, when the subordinate population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into the recovering population extends the length of time required for it to return to its steady state (figure \ref{fig:relaxtraj_ldlh}).
 When both populations are both dramatically reduced, recovery time is generally fastest at lower levels of straying, while near the onset of the fold bifurcation, recovery time increases explosively and this is -- as the name implies -- characteristic of \emph{slowing} dynamics that occur near critical transitions \citep{Scheffer:2009gg,Kuehn:2010p2591}.\\
@@
 At low straying rates, near-collapse of both populations resulted in longer than expected recovery times, whereas in the alternative stable state regime (higher $m^*$), the recovery times for different disturbance types were very similar to systems with a constant $m$ (figure \ref{fig:relax}b; note difference in x-axis scales).
-As trait heritability increased, the metapopulation always recovered more quickly if the small population was lost (figure \ref{extfig-fig:relax_highh}).
+As trait heritability increased, the metapopulation always recovered more quickly if the small population was lost (figure \ref{fig:relax_highh}).
 % There was greater similarity in the recovery times for constant and density-dependent $m$ systems when trait heritability was high (figure \ref{fig:relax_highh}).
@@
 %Lowering Phenotypic diversity with increasing straying rate
-Increased rates of straying lowers phenotypic diversity ($\Delta \mu^* = |\mu_i(t)-\mu_j(t)|$, evaluated at the steady state) because both local and remote populations are increasingly homogenized.
+Increased rates of straying lowers phenotypic diversity ($\Delta \mu^* = |\mu_i^*-\mu_j^*|$, evaluated at the steady state) because both local and remote populations are increasingly homogenized.
 The loss of phenotypic diversity with increased straying is greater if trait heritability is low because traits take longer to go back to their local optima than they do when heritability is large. 
@@
 \noindent{\bf (c) The role of habitat heterogeneity and changing selective landscapes}\\ 
-\noindent With the onset of straying, we find that increasingly divergent trait optima generally lower $N_T$ and exaggerate $\Delta N$ (figure \ref{extfig-fig:thetadiffN}), such that the biomass distribution becomes increasingly uneven.
+\noindent With the onset of straying, we find that increasingly divergent trait optima generally lower $N_T^*$ and exaggerate $\Delta N^*$ (figure \ref{fig:thetadiffN}), such that the biomass distribution becomes increasingly uneven.
 %I've checked this - it's a theshold type (logistic) response Supp fig?
@@
 %The rate of straying that gives rise to a large PE at the fold bifurcation is conditioned on the difference in optimal trait values between the local and remote populations, which we call $\Delta\theta = \sqrt{(\theta_i - \theta_j)^2}$.
-As habitat heterogeneity increases, alternative stable states appear at lower straying rates -- with the crossing of the fold bifurcation, accompanied by a peak in the PE -- whereas the magnitude of increase in the PE also increases (figure \ref{fig:thetaPE}b), reducing recovery time (figure \ref{extfig-fig:relaxtheta}).
-For increased rates of straying, greater habitat heterogeneity erodes the PE (figure \ref{fig:thetaPE}b) and increases the recovery time (figure \ref{extfig-fig:relaxtheta}).
+As habitat heterogeneity increases, alternative stable states appear at lower straying rates -- with the crossing of the fold bifurcation, accompanied by a peak in the PE -- whereas the magnitude of increase in the PE also increases (figure \ref{fig:thetaPE}b), reducing recovery time (figure \ref{fig:relaxtheta}).
+For increased rates of straying, greater habitat heterogeneity erodes the PE (figure \ref{fig:thetaPE}b) and increases the recovery time (figure \ref{fig:relaxtheta}).
 These results together suggest that habitat heterogeneity, as measured as the differences in trait optima between two habitats $\Delta\theta$, promotes robustness when straying rates are low, and erodes robustness when straying rates are high.
@@
 (\emph{i}) above a threshold $m$ value, the dominant population recovers quickly enough that the evolving subordinate phenotype is overwhelmed by incoming strays, and it shifts back to its pre-disturbance (subordinate) state;
-(\emph{ii}) below a threshold $m$ value, there is an \emph{inversion} between subordinate and dominant states: because there is enough time and isolation for the subordinate trait mean to shift towards its local optimum, and away from that of the recovering dominant population, the dominant population becomes subordinate, and the subordinate population becomes dominant (figure \ref{extfig-fig:inertia}).
-This threshold value of $m$, below which the inversion dynamic behavior occurs, is marked by the asterisk in figure \ref{fig:mtheta}, and holds for both constant and density-dependent straying (figure \ref{extfig-fig:mthetamvm}).
+(\emph{ii}) below a threshold $m$ value, there is an \emph{inversion} between subordinate and dominant states: because there is enough time and isolation for the subordinate trait mean to shift towards its local optimum, and away from that of the recovering dominant population, the dominant population becomes subordinate, and the subordinate population becomes dominant (figure \ref{fig:inertia}).
+This threshold value of $m$, below which the inversion dynamic behavior occurs, is marked by the asterisk in figure \ref{fig:mtheta}, and holds for both constant and density-dependent straying (figure \ref{fig:mthetamvm}).
 
@@
 1) the average-CV portfolio effect \citep{Anderson:2013gb,Anonymous:2015gf}, a statistical metric commonly used to assess the buffering capacity of metapopulations, and
-2) the recovery time, defined here as the time required for the aggregate metapopulation biomass $N_T$ to return to its steady state following an induced disturbance, which is mechanistically linked to persistence \citep{Ovaskainen:2002il}.
+2) the recovery time, defined here as the time required for the aggregate metapopulation biomass $N_T^*$ to return to its steady state following an induced disturbance, which is mechanistically linked to persistence \citep{Ovaskainen:2002il}.
 In our minimal eco-evolutionary model of dispersal and natural selection between two populations, we show that these statistical and mechanistic descriptors of metapopulation dynamics and robustness are tightly coupled (figure \ref{fig:PE}d), which is not uncommon for diverse metrics of stability \citep{Donohue:2013iu}.
@@
 In the case of near-collapse, although both populations inherit low population densities, the mean trait values of both are skewed towards the optimum of the dominant population.
-Due to density-dependent straying, low population densities lead to greater dispersal, and while this increased connectivity primarily facilities the growth of the dominant population (because the trait means are closer to the dominant optimum), because the dominant population contains the bulk of the aggregate biomass, the overall recovery time is lessened significantly (figures \ref{fig:relax}b, \ref{extfig-fig:relaxtraj_bothlh}).
+Due to density-dependent straying, low population densities lead to greater dispersal, and while this increased connectivity primarily facilities the growth of the dominant population (because the trait means are closer to the dominant optimum), because the dominant population contains the bulk of the aggregate biomass, the overall recovery time is lessened significantly (figures \ref{fig:relax}b, \ref{fig:relaxtraj_bothlh}).
 % Third, in habitats with greater heterogeneity between sites (measured by an increase in $\Delta\theta$), density-dependent straying diminished robustness at low rates of straying (marginally lower portfolio effects and substantially longer return times; figures \ref{fig:thetaPE}b, \ref{fig:relaxtheta}a,b).
@@
 We suggest that understanding the spatial complexity of metapopulations dispersing across heterogeneous environments, in tandem with the mosaic of selective forces acting on those environments, may be key to uncovering those factors that promote persistence in the wild.
+\\ \\
 
@@
 
-\bibliography{aa_kevin}
+\noindent {\bf Competing interests:} The authors declare no competing interests
+\\
+\noindent {\bf Author contributions:} JDY and JWM conceived of the initial project design.
+JDY and JPG designed the modeling framework and conducted the analyses.
+JDY, JPG, PAHW, and JWM interpreted the results, and drafted and wrote the manuscript.
+\\
+\noindent {\bf Data Accessibility:} Code is made available at https://github.com/jdyeakel/SalmonStrays
+\\
+\noindent {\bf Acknowledgements:} We thank Sean Anderson for helpful discussions and comments on the manuscript.
+We also thank the guest editors Andrew Berdahl, Dora Biro, and Colin Torney, for inviting us to contribute to this themed issue.
+J.D.Y.
+was supported by startup funds at the University of California, Merced, and an Omidyar Postdoctoral Fellowship at the Santa Fe Institute.
+J.P.G.
+was supported by a James S.
+McDonnell Foundation Postdoctoral Fellowship in Complex Systems at the University of California, Merced.
+P.A.H.W.
+was supported by the UA Foundation at the University of Alaska Fairbanks.
+J.W.M.
+was supported by the Liber Ero Research Chair in Coastal Science and Management at Simon Fraser University.
+
+% \bibliography{aa_kevin}
+
+\begin{thebibliography}{10}
+\expandafter\ifx\csname urlstyle\endcsname\relax
+  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
+  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
+  \urlstyle{rm}\Url}\fi
+
+\bibitem{Brown:1977gk}
+Brown JH, Kodric-Brown A, 1977 {Turnover rates in insular biogeography: effect
+  of immigration on extinction}.
+\newblock \emph{Ecology} \textbf{58}, 445--449
+
+\bibitem{Earn:2000fm}
+Earn DJD, Levin SA, Rohani P, 2000 {Coherence and conservation}.
+\newblock \emph{Science} \textbf{290}, 1360--1364
+
+\bibitem{Anonymous:2015gf}
+Schindler DE, Armstrong JB, Reed TE, 2015 {The portfolio concept in ecology and
+  evolution}.
+\newblock \emph{Front. Ecol. Environ.} \textbf{13}, 257--263
+
+\bibitem{Abbott:2017hl}
+Abbott RE, Doak DF, Peterson ML, 2017 {Portfolio effects, climate change, and
+  the persistence of small populations: analyses on the rare plant Saussurea
+  weberi.}
+\newblock \emph{Ecology} \textbf{98}, 1071--1081
+
+\bibitem{Thibaut:2012km}
+Thibaut LM, Connolly SR, 2013 {Understanding diversity-stability relationships:
+  towards a unified model of portfolio effects.}
+\newblock \emph{Ecol. Lett.} \textbf{16}, 140--150
+
+\bibitem{MilnerGulland:2011vm}
+Milner-Gulland EJ, Fryxell JM, Sinclair ARE, 2011 \emph{{Animal Migration}}.
+\newblock Oxford: Oxford University Press
+
+\bibitem{Bell:2011ki}
+Bell G, Gonzalez A, 2011 {Adaptation and evolutionary rescue in metapopulations
+  experiencing environmental deterioration}.
+\newblock \emph{Science} \textbf{332}, 1327--1330
+
+\bibitem{Carlson:2014is}
+Carlson SM, Cunningham CJ, Westley PAH, 2014 {Evolutionary rescue in a changing
+  world}.
+\newblock \emph{Trends Ecol. Evol.} \textbf{29}, 521--530
+
+\bibitem{Muhlfeld:2014hs}
+Muhlfeld CC, Kovach RP, Jones LA, Al-Chokhachy R, Boyer MC, Leary RF, Lowe WH,
+  Luikart G, Allendorf FW, 2014 {Invasive hybridization in a threatened species
+  is accelerated by climate change}.
+\newblock \emph{Nature Climate Change} \textbf{4}, 620--624
+
+\bibitem{Lowe:2015ft}
+Lowe WH, Muhlfeld CC, Allendorf FW, 2015 {Spatial sorting promotes the spread
+  of maladaptive hybridization}.
+\newblock \emph{Trends Ecol. Evol.} \textbf{30}, 456--462
+
+\bibitem{H:2013fs}
+Westley PAH, Quinn TP, 2013 {Rates of straying by hatchery-produced Pacific
+  salmon (\emph{Oncorhynchus} spp.) and steelhead (\emph{Oncorhynchus mykiss})
+  differ among species, life history types, and populations}.
+\newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{70}, 735--746
+
+\bibitem{Keefer:2014gg}
+Keefer ML, Caudill CC, 2014 {Homing and straying by anadromous salmonids: a
+  review of mechanisms and rates}.
+\newblock \emph{Rev Fish Biol Fisher} \textbf{24}, 333--368
+
+\bibitem{Bett:2017ha}
+Bett NN, Hinch SG, Burnett NJ, Donaldson MR, Naman SM, 2017 {Causes and
+  consequences of straying into small populations of Pacific salmon}.
+\newblock \emph{Fisheries} \textbf{42}, 220--230
+
+\bibitem{Berdahl:2015kv}
+Berdahl A, Torney CJ, Schertzer E, Levin SA, 2015 {On the evolutionary
+  interplay between dispersal and local adaptation in heterogeneous
+  environments}.
+\newblock \emph{Evolution} \textbf{69}, 1390--1405
+
+\bibitem{Berdahl:2016dx}
+Berdahl A, 2016 {Collective behavior as a driver of critical transitions in
+  migratory populations}.
+\newblock \emph{Movement Ecology} \textbf{4}, 1--12
+
+\bibitem{Simons:2004jo}
+Simons AM, 2004 {Many wrongs: The advantage of group navigation}.
+\newblock \emph{Trends Ecol. Evol.} \textbf{19}, 453--455
+
+\bibitem{Brenner:2012gl}
+Brenner RE, Moffitt SD, Grant WS, 2012 {Straying of hatchery salmon in Prince
+  William Sound, Alaska}.
+\newblock \emph{Environ Biol Fish} \textbf{94}, 179--195
+
+\bibitem{Johnson:2012fe}
+Johnson RC, Weber PK, Wikert JD, Workman ML, MacFarlane RB, Grove MJ, Schmitt
+  AK, 2012 {Managed metapopulations: do salmon hatchery
+  {\textquoteleft}sources{\textquoteright} lead to in-river
+  {\textquoteleft}sinks{\textquoteright} in conservation?}
+\newblock \emph{PLoS ONE} \textbf{7}, e28880--11
+
+\bibitem{Fullerton:2011ii}
+Fullerton AH, Lindley ST, Pess GR, Feist BE, Steel EA, Mcelhany P, 2011 {Human
+  influence on the spatial structure of threatened Pacific salmon
+  metapopulations}.
+\newblock \emph{Conserv Biol} \textbf{25}, 932--944
+
+\bibitem{Quinn:2011tf}
+Quinn TP, 2011 \emph{{The Behavior and Ecology of Pacific Salmon and Trout}}.
+\newblock Vancouver: UBC Press
+
+\bibitem{Jonsson:2011kg}
+Jonsson B, Jonsson N, 2011 \emph{{Ecology of Atlantic Salmon and Brown Trout}}.
+\newblock Dordrecht: Springer Netherlands
+
+\bibitem{Quinn:1993ge}
+Quinn TP, 1993 {A review of homing and straying of wild and hatchery-produced
+  salmon}.
+\newblock \emph{Fish Res} \textbf{18}, 29--44
+
+\bibitem{Hendry:2004wf}
+Hendry AP, Bohlin T, Jonsson B, Berg OK, 2004 {The evolution of philopatry and
+  dispersal: homing versus straying in salmonids}.
+\newblock In AP~Hendry, SC~Stearns, eds., \emph{Evolution Illuminated}. Oxford:
+  Oxford University Press on Demand
+
+\bibitem{Schtickzelle:2007wb}
+Schtickzelle N, Quinn TP, 2007 {A metapopulation perspective for salmon and
+  other anadromous fish}.
+\newblock \emph{Fish Fish.} \textbf{8}, 297--314
+
+\bibitem{Anderson:2014cx}
+Anderson SC, Moore JW, McClure MM, Dulvy NK, Cooper AB, 2015 {Portfolio
+  conservation of metapopulations under climate change}.
+\newblock \emph{Ecol. Appl.} \textbf{25}, 559--572
+
+\bibitem{Keefer:2008bs}
+Keefer ML, Caudill CC, Peery CA, Lee SR, 2008 {Transporting juvenile salmonids
+  around dams impairs adult migration}.
+\newblock \emph{Ecol. Appl.} \textbf{18}, 1888--1900
+
+\bibitem{Westley:2015to}
+Westley PAH, Dittman AH, Ward EJ, Quinn TP, 2015 {Signals of climate,
+  conspecific density, and watershed features in patterns of homing and
+  dispersal by Pacific salmon.}
+\newblock \emph{Ecology} \textbf{96}, 2823--2833
+
+\bibitem{Bond:2016dz}
+Bond MH, Westley PAH, Dittman AH, Holecek D, Marsh T, Quinn TP, 2016 {Combined
+  effects of barge transportation, river environment, and rearing location on
+  straying and migration of adult snake river fall-run Chinook salmon}.
+\newblock \emph{T Am Fish Soc} \textbf{146}, 60--73
+
+\bibitem{Peterson:2014gy}
+Peterson DA, Hilborn R, Hauser L, 2014 {Local adaptation limits lifetime
+  reproductive success of dispersers in a wild salmon metapopulation}.
+\newblock \emph{Nat Commun} \textbf{5}, 3696
+
+\bibitem{Moore:2010gs}
+Moore JW, McClure M, Rogers LA, Schindler DE, 2010 {Synchronization and
+  portfolio performance of threatened salmon}.
+\newblock \emph{Conserv. Lett.} \textbf{3}, 340--348
+
+\bibitem{Carlson:2011ce}
+Carlson SM, Satterthwaite WH, Fleming IA, 2011 {Weakened portfolio effect in a
+  collapsed salmon population complex}.
+\newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{68}, 1579--1589
+
+\bibitem{Braun:2016ib}
+Braun DC, Moore JW, Candy J, Bailey RE, 2016 {Population diversity in salmon:
+  linkages among response, genetic and life history diversity}.
+\newblock \emph{Ecography} \textbf{39}, 317--328
+
+\bibitem{simpson1953major}
+Simpson GG, 1953 \emph{{The Major Features of Evolution}}.
+\newblock New York: Simon and Schuster
+
+\bibitem{Lande:1976ga}
+Lande R, 1976 {Natural selection and random genetic drift in phenotypic
+  evolution}.
+\newblock \emph{Evolution} \textbf{30}, 314--334
+
+\bibitem{Shelton:2011eq}
+Shelton AO, Mangel M, 2011 {Fluctuations of fish populations and the magnifying
+  effects of fishing.}
+\newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{108}, 7075--7080
+
+\bibitem{Jasper:2013cc}
+Jasper JR, Habicht C, Moffitt S, Brenner R, Marsh J, Lewis B, Fox EC, Grauvogel
+  Z, Olive SDR, Grant WS, 2013 {Source-sink estimates of genetic introgression
+  show influence of hatchery strays on wild chum salmon populations in Prince
+  William Sound, Alaska}.
+\newblock \emph{PLoS ONE} \textbf{8}, e81916
+
+\bibitem{Schreiber:2011wx}
+Schreiber SJ, B{\"u}rger R, Bolnick DI, 2011 {The community effects of
+  phenotypic and genetic variation within a predator population.}
+\newblock \emph{Ecology} \textbf{92}, 1582--1593
+
+\bibitem{Gilbert:2014ee}
+Gibert JP, Brassil CE, 2014 {Individual phenotypic variation reduces
+  interaction strengths in a consumer{\textendash}resource system}.
+\newblock \emph{Ecol Evol} \textbf{4}, 3703--3713
+
+\bibitem{Gibert:2015kc}
+Gibert JP, DeLong JP, 2015 {Individual variation decreases interference
+  competition but increases species persistence}.
+\newblock \emph{Adv Ecol Res} \textbf{52}, 45--64
+
+\bibitem{Berdahl:2014bl}
+Berdahl A, Westley PAH, Levin SA, Couzin ID, Quinn TP, 2014 {A collective
+  navigation hypothesis for homeward migration in anadromous salmonids}.
+\newblock \emph{Fish Fish.} \textbf{17}, 525--542
+
+\bibitem{Candy:2000hu}
+Candy JR, Beacham TD, 2000 {Patterns of homing and straying in southern British
+  Columbia coded-wire tagged Chinook salmon (\emph{Oncorhynchus tshawytscha})
+  populations}.
+\newblock \emph{Fish Res} \textbf{47}, 41--56
+
+\bibitem{JPE:JPE1383}
+Schick RS, Lindley ST, 2007 {Directed connectivity among fish populations in a
+  riverine network}.
+\newblock \emph{J. Appl. Ecol.} \textbf{44}, 1116--1126
+
+\bibitem{Schindler:2015gf}
+Schindler DE, Armstrong JB, Reed TE, 2015 {The portfolio concept in ecology and
+  evolution}.
+\newblock \emph{Front. Ecol. Environ.} \textbf{13}, 257--263
+
+\bibitem{Ovaskainen:2002il}
+Ovaskainen O, Hanski I, 2002 {Transient dynamics in metapopulation response to
+  perturbation}.
+\newblock \emph{Theor Popul Biol} \textbf{61}, 285--295
+
+\bibitem{Anderson:2013gb}
+Anderson SC, Cooper AB, Dulvy NK, 2013 {Ecological prophets: quantifying
+  metapopulation portfolio effects}.
+\newblock \emph{Methods Ecol Evol} \textbf{4}, 971--981
+
+\bibitem{Loreau:2008ju}
+Loreau M, de~Mazancourt C, 2008 {Species synchrony and its drivers: neutral and
+  nonneutral community dynamics in fluctuating environments}.
+\newblock \emph{Am. Nat.} \textbf{172}, E48--E66
+
+\bibitem{Yeakel:2013vz}
+Yeakel JD, Moore JW, Guimar{\~a}es~Jr PR, de~Aguiar MAM, 2014 {Synchronisation
+  and stability in river metapopulation networks}.
+\newblock \emph{Ecol. Lett.} \textbf{17}, 273--283
+
+\bibitem{GuckHolmes}
+Guckenheimer J, Holmes P, 1983 \emph{{Nonlinear Oscillations, Dynamical
+  Systems, and Bifurcations of Vector Fields}}.
+\newblock New York: Springer
+
+\bibitem{Carlson:2008hl}
+Carlson SM, Seamons TR, 2008 {A review of quantitative genetic components of
+  fitness in salmonids: implications for adaptation to future change}.
+\newblock \emph{Evol Appl} \textbf{1}, 222--238
+
+\bibitem{Scheffer:2009gg}
+Scheffer M, Bascompte J, Brock WA, Brovkin V, Carpenter SR, Dakos V, Held H,
+  van Nes EH, Rietkerk M, Sugihara G, 2009 {Early-warning signals for critical
+  transitions}.
+\newblock \emph{Nature} \textbf{461}, 53--59
+
+\bibitem{Lade:2012eu}
+Lade SJ, Gross T, 2012 {Early warning signals for critical transitions: A
+  generalized modeling approach}.
+\newblock \emph{PLoS Comp. Biol.} \textbf{8}, e1002360
+
+\bibitem{Anonymous:2013br}
+Boettiger C, Ross N, Hastings A, 2013 {Early warning signals: the charted and
+  uncharted territories}.
+\newblock \emph{Theor. Ecol.} \textbf{6}, 255--264
+
+\bibitem{Dakos:2014br}
+Dakos V, Bascompte J, 2014 {Critical slowing down as early warning for the
+  onset of collapse in mutualistic communities}.
+\newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{111}, 17546--17551
+
+\bibitem{Krkosek:2014ch}
+Krko{\v s}ek M, Drake JM, 2014 {On signals of phase transitions in salmon
+  population dynamics}.
+\newblock \emph{Proc. Roy. Soc. B} \textbf{281}, 20133221
+
+\bibitem{Kuehn:2010p2591}
+Kuehn C, 2011 {A mathematical framework for critical transitions: bifurcations,
+  fast-slow systems and stochastic dynamics}.
+\newblock \emph{Physica D} \textbf{240}, 1020--1035
+
+\bibitem{Donohue:2013iu}
+Donohue I, Petchey OL, Montoya JM, Jackson AL, McNally L, Viana M, Healy K,
+  Lurgi M, O'Connor NE, Emmerson MC, 2013 {On the dimensionality of ecological
+  stability.}
+\newblock \emph{Ecol. Lett.} \textbf{16}, 421--429
+
+\bibitem{Levin:1992p1472}
+Levin SA, 1992 {The problem of pattern and scale in ecology}.
+\newblock \emph{Ecology} \textbf{73}, 1943--1967
+
+\bibitem{Meinhardt:2012hw}
+Meinhardt H, 2012 {Turing's theory of morphogenesis of 1952 and the subsequent
+  discovery of the crucial role of local self-enhancement and long-range
+  inhibition}.
+\newblock \emph{Interface Focus} \textbf{2}, 407--416
+
+\bibitem{Hanski:1999vu}
+Hanski I, 1999 \emph{{Metapopulation Ecology}}.
+\newblock Oxford: Oxford University Press
+
+\bibitem{Boughton:1999fa}
+Boughton DA, 1999 {Empirical evidence for complex source{\textendash}sink
+  dynamics with alternative states in a butterfly metapopulation}.
+\newblock \emph{Ecology} \textbf{80}, 2727--2739
+
+\bibitem{Secor:2009ena}
+Secor DH, Kerr LA, Cadrin SX, 2009 {Connectivity effects on productivity,
+  stability, and persistence in a herring metapopulation model}.
+\newblock \emph{ICES J Mar Sci} \textbf{66}, 1726--1732
+
+\bibitem{Christie:2012bj}
+Christie MR, Marine ML, French RA, Blouin MS, 2012 {Genetic adaptation to
+  captivity can occur in a single generation.}
+\newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{109}, 238--242
+
+\bibitem{Chilcote:2003bb}
+Chilcote MW, 2003 {Relationship between natural productivity and the frequency
+  of wild fish in mixed spawning populations of wild and hatchery steelhead
+  (\emph{Oncorhynchus mykiss})}.
+\newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{60}, 1057--1067
+
+\bibitem{Araki:2007cm}
+Araki H, Cooper B, Blouin MS, 2007 {Genetic effects of captive breeding cause a
+  rapid, cumulative fitness decline in the wild}.
+\newblock \emph{Science} \textbf{318}, 100--103
+
+\bibitem{Westley:2012ui}
+Westley PAH, { Corinne M Conway}, Fleming IA, 2012 {Phenotypic divergence of
+  exotic fish populations is shaped by spatial proximity and habitat
+  differences across an invaded landscape}.
+\newblock \emph{Evol. Ecol. Res.} \textbf{14}, 147--167
+
+\bibitem{Anderson:2013bf}
+Anderson JH, Faulds PL, Atlas WI, Quinn TP, 2013 {Reproductive success of
+  captively bred and naturally spawned Chinook salmon colonizing newly
+  accessible habitat}.
+\newblock \emph{Evol Appl} \textbf{6}, 165--179
+
+\bibitem{Pess:2014isa}
+Pess GR, Quinn TP, Gephard SR, Saunders R, 2014 {Re-colonization of Atlantic
+  and Pacific rivers by anadromous fishes: linkages between life history and
+  the benefits of barrier removal}.
+\newblock \emph{Rev Fish Biol Fisher} \textbf{24}, 881--900
+
+\bibitem{Ford:2002ip}
+Ford MJ, 2002 {Selection in captivity during supportive breeding may reduce
+  fitness in the wild}.
+\newblock \emph{Conserv Biol} \textbf{16}, 815--825
+
+\bibitem{Liermann:2017gj}
+Liermann M, PESS G, McHenry M, McMillan J, Elofson M, Bennett T, Moses R, 2017
+  {Relocation and recolonization of Coho salmon \emph{Oncorhynchus kisutch} in
+  two tributaries to the elwha river: Implications for management and
+  monitoring}.
+\newblock \emph{T Am Fish Soc} \textbf{17}, 1--39
+
+\end{thebibliography}
+
+\clearpage
 
@@
 \centering
-\includegraphics[width=0.4\textwidth]{figs2/fig_traj.pdf}
+\includegraphics[width=0.4\textwidth]{fig_traj.pdf}
 \caption{
@@
 \centering
-\includegraphics[width=1\textwidth]{figs2/fig_MDPE_hm.pdf}
+\includegraphics[width=1\textwidth]{fig_MDPE_hm.pdf}
 \caption{
-(a) Total means $N_t$, 
-(b) difference in means $\Delta N$, and 
+(a) Total means $N_t^*$, 
+(b) difference in means $\Delta N^*$, and 
 (c) the portfolio effect PE as a function of heritability $h^2$ and a constant stray rate $m$. Light colors = high values.
@@
 \centering
-\includegraphics[width=0.4\textwidth]{figs2/fig_thetaPEmvm.pdf}
+\includegraphics[width=0.4\textwidth]{fig_thetaPEmvm.pdf}
 \caption{
@@
 \centering
-\includegraphics[width=0.9\textwidth]{figs2/fig_relax_lowh.pdf}
+\includegraphics[width=0.9\textwidth]{fig_relax_lowh.pdf}
 \caption{
-Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.2$.
+Recovery time of $N_T^*$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.2$.
 If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
@@
 \centering
-\includegraphics[width=0.5\textwidth]{figs2/fig_traitdiff.pdf}
+\includegraphics[width=0.5\textwidth]{fig_traitdiff.pdf}
 \caption{
@@
   \centering
-  \includegraphics[width=0.5\textwidth]{figs2/fig_mtheta_rt.pdf}
+  \includegraphics[width=0.5\textwidth]{fig_mtheta_rt.pdf}
   \caption{
@@
 
-% 
-% 
-% \beginsupplement
-% 
-% 
-% \begin{figure}
-%   \captionsetup{justification=raggedright,
-% singlelinecheck=false
-% }
-% \centering
-% \includegraphics[width=0.35\textwidth]{figs2/fig_relax_small.pdf}
-% \caption{
-% Extinction of low-density population with a high constant straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
-% Black line marks the calculated point of recovery post-perturbation.
-% Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
-% } \label{fig:relaxtraj_ldlh}
-% \end{figure}
-% 
-% \begin{figure}
-%   \captionsetup{justification=raggedright,
-% singlelinecheck=false
-% }
-% \centering
-% \includegraphics[width=0.35\textwidth]{figs2/fig_relax_large.pdf}
-% \caption{
-% Extinction of high-density population with a high straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
-% Black line marks the calculated point of recovery post-perturbation.
-% Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
-% } \label{fig:relaxtraj_hdlh}
-% \end{figure}
-% 
-% 
-% \begin{figure*}
-%   \captionsetup{justification=raggedright,
-% singlelinecheck=false
-% }
-% \centering
-% \includegraphics[width=0.9\textwidth]{figs2/fig_relax_highh.pdf}
-% \caption{
-% Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.8$.
-% If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
-% } \label{fig:relax_highh}
-% \end{figure*}
-% 
-% 
-% 
-% \begin{figure}
-%   \captionsetup{justification=raggedright,
-% singlelinecheck=false
-% }
-% \centering
-% \includegraphics[width=0.4\textwidth]{figs2/fig_thetadiffN.pdf}
-% \caption{
-% Median difference in population densities taken over the straying rate as a function of habitat heterogeneity $\Delta\theta$.
-% Solid lines are for constant $m$; dashed lines are for density-dependent $m$} \label{fig:thetadiffN}
-% \end{figure}
-% 
-% 
-% 
-% \begin{figure*}
-%   \captionsetup{justification=raggedright,
-% singlelinecheck=false
-% }
-% \centering
-% \includegraphics[width=0.8\textwidth]{figs2/fig_relaxtheta2.pdf}
-% \caption{
-% (a) Recovery time after near collapse of both populations as a function of straying rate $m$ and habitat heterogeneity $\Delta\theta$.
-% (b) The same as (a) but including recovery times when straying is density-dependent, shown by linked pairs of points.
-% Recovery times for systems with density-dependent straying are longer at low straying rates and shorter at higher straying rates, mirroring the change in portfolio effects shown in figure \ref{fig:thetaPE}.
-% } \label{fig:relaxtheta}
-% \end{figure*}
-% 
-% 
-% \begin{figure}
-%   \captionsetup{justification=raggedright,
-% singlelinecheck=false
-% }
-%   \centering
-%   \includegraphics[width=0.35\textwidth]{figs2/fig_MDPE_hm_mtheta_rt.pdf}
-%   \caption{
-%   Distance dependent portfolio effects as a function of straying rate $m$ and trait heritability $h^2$. When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases.
-%   } \label{fig:mthetaPE}
-% \end{figure}
-% 
-% \begin{figure}
-%   \captionsetup{justification=raggedright,
-% singlelinecheck=false
-% }
-% \centering
-% \includegraphics[width=0.35\textwidth]{figs2/fig_relax_inertia.pdf}
-% \caption{
-% Distance dependent straying, where increased differences in trait optima between sites $\Delta\theta$ corresponds to lower rates of straying $m$.
-% At low rates of straying $m=0.02$ ($\Delta\theta=24$), extinction of the dominant population leads to slower-than-expected recovery times because the subordinate population is isolated enough to evolve towards its own trait optimum.
+\beginsupplement
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+% \includegraphics[width=0.35\textwidth]{fig_relax_small.pdf}
+\caption{
+Extinction of low-density population with a high constant straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
+Black line marks the calculated point of recovery post-perturbation.
+Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
+} \label{fig:relaxtraj_ldlh}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+% \includegraphics[width=0.35\textwidth]{fig_relax_large.pdf}
+\caption{
+Extinction of high-density population with a high straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
+Black line marks the calculated point of recovery post-perturbation.
+Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
+} \label{fig:relaxtraj_hdlh}
+\end{figure}
+
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+% \includegraphics[width=0.9\textwidth]{fig_relax_highh.pdf}
+\caption{
+Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.8$.
+If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
+} \label{fig:relax_highh}
+\end{figure*}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+% \includegraphics[width=0.4\textwidth]{fig_thetadiffN.pdf}
+\caption{
+Median difference in population densities taken over the straying rate as a function of habitat heterogeneity $\Delta\theta$.
+Solid lines are for constant $m$; dashed lines are for density-dependent $m$} \label{fig:thetadiffN}
+\end{figure}
+
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+% \includegraphics[width=0.8\textwidth]{fig_relaxtheta2.pdf}
+\caption{
+(a) Recovery time after near collapse of both populations as a function of straying rate $m$ and habitat heterogeneity $\Delta\theta$.
+(b) The same as (a) but including recovery times when straying is density-dependent, shown by linked pairs of points.
+Recovery times for systems with density-dependent straying are longer at low straying rates and shorter at higher straying rates, mirroring the change in portfolio effects shown in figure \ref{fig:thetaPE}.
+} \label{fig:relaxtheta}
+\end{figure*}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+  \centering
+  % \includegraphics[width=0.35\textwidth]{fig_MDPE_hm_mtheta_rt.pdf}
+  \caption{
+  Distance dependent portfolio effects as a function of straying rate $m$ and trait heritability $h^2$. When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases.
+  } \label{fig:mthetaPE}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+% \includegraphics[width=0.35\textwidth]{fig_relax_inertia.pdf}
+\caption{
+Distance dependent straying, where increased differences in trait optima between sites $\Delta\theta$ corresponds to lower rates of straying $m$.
+At low rates of straying $m=0.02$ ($\Delta\theta=24$), extinction of the dominant population leads to slower-than-expected recovery times because the subordinate population is isolated enough to evolve towards its own trait optimum.
 %until growth of the dominant population overwhelms this local selection.
-% In this case, $m$ is less than $m=0.034$ (denoted by the asterisk in figure \ref{fig:mtheta}), such that isolation allows the subdominant population to `run away' from the influence of the dominant population, leading to a switch in states.
-% If $m$ is low but greater than $0.034$, isolation permits the subdominant population to `run away' from the influence of the dominant population, until it is overwhelmed by the recovering dominant population, and reverts back to its previous trait mean prior to the disturbance.
-% } \label{fig:inertia}
-% \end{figure}
-% 
-% \begin{figure*}
-%   \captionsetup{justification=raggedright,
-% singlelinecheck=false
-% }
-%   \centering
-%   \includegraphics[width=0.8\textwidth]{figs2/fig_relax_mtheta.pdf}
-%   \caption{
-%   Distance dependent recovery times for three disturbance types.
+In this case, $m$ is less than $m=0.034$ (denoted by the asterisk in figure \ref{fig:mtheta}), such that isolation allows the subdominant population to `run away' from the influence of the dominant population, leading to a switch in states.
+If $m$ is low but greater than $0.034$, isolation permits the subdominant population to `run away' from the influence of the dominant population, until it is overwhelmed by the recovering dominant population, and reverts back to its previous trait mean prior to the disturbance.
+} \label{fig:inertia}
+\end{figure}
+
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+  \centering
+  % \includegraphics[width=0.8\textwidth]{fig_relax_mtheta.pdf}
+  \caption{
+  Distance dependent recovery times for three disturbance types.
 When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases for constant (a) and density-dependent (b) staying rates.
-%   The fold bifurcation is not as clear in (b) because $\Delta\theta$ is a function of the individual straying rate $m_0$, whereas the x-axis in (b) is the straying rate at the steady state $m^*$.
-%   Despite this difference, the general trends shown in (a) are also present in (b).
-%   } \label{fig:mthetamvm}
-% \end{figure*}
-% 
-% 
-% 
-% \begin{figure}
-%   \captionsetup{justification=raggedright,
-% singlelinecheck=false
-% }
-% \centering
-% \includegraphics[width=0.35\textwidth]{figs2/fig_relax_both_lowh.pdf}
-% \caption{
-% Near collapse of both populations with a low straying rate $m=0.1$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
-% Black line marks the calculated point of recovery post-perturbation.
-% Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
-% } \label{fig:relaxtraj_bothlh}
-% \end{figure}
-%
+  The fold bifurcation is not as clear in (b) because $\Delta\theta$ is a function of the individual straying rate $m_0$, whereas the x-axis in (b) is the straying rate at the steady state $m^*$.
+  Despite this difference, the general trends shown in (a) are also present in (b).
+  } \label{fig:mthetamvm}
+\end{figure*}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+% \includegraphics[width=0.35\textwidth]{fig_relax_both_lowh.pdf}
+\caption{
+Near collapse of both populations with a low straying rate $m=0.1$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
+Black line marks the calculated point of recovery post-perturbation.
+Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
+} \label{fig:relaxtraj_bothlh}
+\end{figure}
 

### 2017-09-06 Manuscript/FinalDraft_rev/supp.tex
?3

--- 2017-06-02 manuscript/supp.tex
@@
 \usepackage{xr}
-\externaldocument[extfig-]{msv4}
+\externaldocument[extfig-]{msv5}
 %\usepackage{epstopdf}
@@
 %\title{Community assembly and dynamics by the deconstruction of species interactions}
-\title{Supplementary Figures: Eco-evolutionary dynamics and collective straying: implications for salmon metapopulation robustness}
+\title{Supplementary Figures: Eco-evolutionary dynamics and collective migration: implications for salmon metapopulation robustness}
 \author{
@@
 \centering
-\includegraphics[width=0.35\textwidth]{figs2/fig_relax_small.pdf}
+\includegraphics[width=0.35\textwidth]{fig_relax_small.pdf}
 \caption{
@@
 \centering
-\includegraphics[width=0.35\textwidth]{figs2/fig_relax_large.pdf}
+\includegraphics[width=0.35\textwidth]{fig_relax_large.pdf}
 \caption{
@@
 \centering
-\includegraphics[width=0.9\textwidth]{figs2/fig_relax_highh.pdf}
+\includegraphics[width=0.9\textwidth]{fig_relax_highh.pdf}
 \caption{
@@
 \centering
-\includegraphics[width=0.4\textwidth]{figs2/fig_thetadiffN.pdf}
+\includegraphics[width=0.4\textwidth]{fig_thetadiffN.pdf}
 \caption{
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figs2/fig_relaxtheta2.pdf}
+\includegraphics[width=0.8\textwidth]{fig_relaxtheta2.pdf}
 \caption{
@@
   \centering
-  \includegraphics[width=0.35\textwidth]{figs2/fig_MDPE_hm_mtheta_rt.pdf}
+  \includegraphics[width=0.35\textwidth]{fig_MDPE_hm_mtheta_rt.pdf}
   \caption{
@@
 \centering
-\includegraphics[width=0.35\textwidth]{figs2/fig_relax_inertia.pdf}
+\includegraphics[width=0.35\textwidth]{fig_relax_inertia.pdf}
 \caption{
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{figs2/fig_relax_mtheta.pdf}
+  \includegraphics[width=0.8\textwidth]{fig_relax_mtheta.pdf}
   \caption{
@@
 \centering
-\includegraphics[width=0.35\textwidth]{figs2/fig_relax_both_lowh.pdf}
+\includegraphics[width=0.35\textwidth]{fig_relax_both_lowh.pdf}
 \caption{

### 2017-09-12 Manuscript/FinalDraft_rev/ReviewResponses.tex
edits

--- 2017-08-31 manuscript/FinalDraft_rev/ReviewResponses.tex
@@
 \subsection*{Major comments:}
-\textcolor{cyan}{Equation (4), which describes the change in the population mean trait under dispersal and selection, is incorrect.
-In the second line of Eq (4), the immigrant term $(1-w_i) R_i\[\mu_j\]$ is independent of $\mu_i$, so that the derivative of this term is zero.
+\noindent \textcolor{cyan}{
+{\bf Comment 1.1} Equation (4), which describes the change in the population mean trait under dispersal and selection, is incorrect.
+In the second line of Eq (4), the immigrant term $(1-w_i) R_i [\mu_j]$ is independent of $\mu_i$, so that the derivative of this term is zero.
 This is as if selection did not change the allele frequencies/mean trait value of the immigrant subpopulation, which is obviously untrue (there is strong selection on the locally maladapted immigrants, favouring those who happen to be closer to the local optimum).
-One cannot apply Lande's equation without adapting it to the immigration model.}
+One cannot apply Lande's equation without adapting it to the immigration model.
+}
 
-We appreciate the Reviewer's concern, however note that equation 4) is written as
+\noindent {\bf Response 1.1} We appreciate the Reviewer's concern, however note that equation 4) is written as
 \begin{align}
@@
   \mu_i(t+1) &= w_i\mu_i(t) + (1-w_i)\mu_j(t) \\ \nonumber
-  &+ h^2\sigma^2\frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)] + (1-w_i)R_i[\mu_j(t)]  \right),
+  &+ h^2\sigma^2\frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)] + (1-w_i)R_i[\mu_j(t)]  \right).
 \end{align}
-and because the third term on the right is the derivative of the log of the sum (as it must be for the discrete time version of the Lande equation; Lande \emph{Evolution}, 1976), the derivative of the second term within the logarithm is not zero, and both phenotypic means contribute to the strength and direction of selection.
+Because the third term on the right is the derivative of the log of the sum (as it must be for the discrete time version of the Lande equation; Lande \emph{Evolution}, 1976), the derivative of the second term within the logarithm is not zero, and both phenotypic means contribute to the strength and direction of selection, such that
+\begin{equation}
+  \frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i] + (1-w_i)R_i[\mu_j]  \right) = \frac{\theta_i-\mu_i}{\left(\sigma ^2+\tau ^2\right) \left(1+\left(\frac{1}{w_i}-1\right) {\rm exp}\left[\frac{(\mu_i-\mu_j) (-2 \theta_i+\mu_i+\mu_j)}{2 \left(\sigma ^2+\tau ^2\right)}\right]\right)}.
+\end{equation}
 
-\textcolor{cyan}{
-The model with constant dispersal ($m$) is very much the same as the model of Ronce and Kirkpatrick (2001, Evolution), with the difference that the latter is fully deterministic and set in continuous time; but as long as the system attains an equilibrium rather than population cycles or chaos, continuous time should not make a difference.
+\noindent \textcolor{cyan}{
+{\bf Comment 1.2} The model with constant dispersal ($m$) is very much the same as the model of Ronce and Kirkpatrick (2001, Evolution), with the difference that the latter is fully deterministic and set in continuous time; but as long as the system attains an equilibrium rather than population cycles or chaos, continuous time should not make a difference.
 Ronce and Kirkpatrick focused on equilibria (such as the dominant vs subordinate state here ) and on local adaptation, not metapopulation resilience/stability, but did study the consequence of disturbances for switching between alternative equilibria (see the next point).
@@
 
-ff
+\noindent {\bf Response 1.2} We are thankful for the Reviewer in pointing out this reference, which we are embarrassed to admit to having overlooked.
+The model indeed has many similarities to our own, and we now acknowledge its importance and priority within the manuscript.
+The similarities in the resulting dynamics suggest that what we are observing in the two-population salmon system, as well as the implications that we derive for metapopulation robustness and recovery, may be generally relevant and illustrative of a general phenomenon outside of the constraints introduced in the specific formulation of our model.
 
-\textcolor{cyan}{
-The bifurcation the authors call a fold bifurcation is not a fold bifurcation.
+\noindent We have made the following edits to the text:\\
+\noindent {\bf Line XXX:} \emph{This model formulation has parallels to that proposed by Ronce and Kirkpatrick [ref]), where habitat specialization evolves between two populations as a function of dispersal.
+The largest difference between these approaches is that our framework treats trait evolution mechanistically (at some cost to analytical tractability).
+Importantly, we show that the resulting dynamics are qualitatively similar, suggesting that the dynamical features present in both of these approaches have potentially widespread ramifications for the evolutionary dynamics of connected populations.}\\
+\noindent {\bf Line XXX} \emph{This dynamic is also observed in the Ronce and Kirkpatrick model, which they describe as a transition from symmetric to asymmetric states [ref].}
+
+\noindent \textcolor{cyan}{
+{\bf Comment 1.3} The bifurcation the authors call a fold bifurcation is not a fold bifurcation.
 Figure 1 suggests a pitchfork bifurcation (which is degenerate but will appear in symmetric models).
@@
 
-ff
+\noindent {\bf Response 1.3} Both Reviewers indicated that our labeling of the Fold Bifurcation was incorrect, and that the identity of the transition of the system from symmetric to asymmetric states was in fact a Pitchfork bifurcation.
+We are happy to modify the identification of the bifurcation as the Reviewers suggest as the Pitchfork label will be more familiar to readers and is somewhat self-evident in terms of the bifurcation diagram that is displayed.
+At the same time, we would like to explain our reasoning, and maintain our position that the bifurcation is technically (and more generally) a fold bifurcation as well.
 
-\textcolor{cyan}{
-The ms does not describe how recovery time was measured.
+According to Kuznetsov (1998), there are two principle bifurcations in discrete maps, from which others are composed: the Fold and Flip (period-doubling).
+The Fold bifurcation is generally defined by the dominant eigenvalue $\lambda$ intersecting the unit circle in Real-Imaginary space at ${\rm Re}[\lambda]=+1$.
+We now include a new Supplemental Figure (SXX) that shows this is indeed the case for our system, and was the basis on which we initially identified the bifurcation.
+Although we are not specialists in bifurcation theory, our understanding is that the fold bifurcation is a general attribute based on this eigenvalue behavior, but as the Reviewers indicate, it is not a particularly clear or complete description of what is going on.
+A familiar, and more complicated bifurcation that occurs in systems with $\geq2$ parameters is the Cusp bifurcation, which occurs when two fold bifurcations interact at a cusp.
+A one-dimensional transect through the cusp results in what is commonly referred to as a Pitchfork Bifurcation (topological normal form: $x \mapsto x + \beta x - x^3$).
+At the cusp ($\beta=0$), the single fixed point crosses the combined fold bifurcations, and here it branches into 3 fixed points; two stable fixed points separated by an unstable fixed point.
+Identifying the bifurcation as a Pitchfork bifurcation increases clarity and specificity, and we have made the appropriate edits in an effort to do so.
+
+% Interestingly, because the system passes through a Pitchfork, or the Cusp at the cusp point, rather than one fold bifurcation at a time (which would result in similar stable-unstable-stable fixed points), the dependence of whether the system will veer to both states
+
+As the Reviewers suggest, we investigated whether or not our system displays hysteresis, as does the system in the Ronce and Kirkpatrick model.
+We now show that it does (Supplementary Figure SXX), a feature characteristic of Cusp bifurcations in general.
+This is more evidence that the dynamics described here and in Ronce and Kirkpatrick are perhaps general phenomena of spatially-linked eco-evolutionary systems.
+This perspective is reinforced by the fact that altering parameters between the two sites (creating asymmetry) does not alter the qualitative nature of the system.
+
+% we believe that our 4-dimensional discrete time system crosses a discrete cusp bifurcation at intermediate values of $m$.
+The discrete cusp bifurcation is characterized by the intersection of two branches of a fold bifurcation, of which the topological normal form is $x \mapsto f(x) = x + \beta_1 + \beta_2 x - x^3$.
+The eigenvalue evaluated at the steady state $\partial f(x)/\partial x|_{x=x^*}$ increases to unity on the unit circle at the cusp bifurcation $(\beta_1 = 0, \beta_2 = 0)$.
+
+\noindent \textcolor{cyan}{
+{\bf Comment 1.4} The ms does not describe how recovery time was measured.
 The deterministic system returns to an equilibrium asymptotically, i.e., in infinite time.
@@
 
-ff
+\noindent {\bf Response 1.4} We thank the Reviewer and apologize for our lack of clarity.
+We emphasize that our measure of recovery time is a numerical measurement, as eigenvalue-based (asymptotic) measures are not appropriate for the scale of disturbance that we introduce (extinction in some cases - far from the non-trivial fixed point(s) of the system).
+The procedure is now described in more detail within the text, and illustrated graphically in an additional supplemental figure (Supplementary Figure SXX).
+In short, we track the recovering aggregate population $N_T=N_1+N_2$ until it is within a standard deviation around the \emph{updated} fixed point (thus accounting for fixed points that are altered by the disturbance).
+Note that this recovery time is specifically for the aggregate population, such that if a single population takes a long time to recovery relative to the other, the aggregate takes longer to recover as well.
+If $N_T$ remains within a standard deviation of the final fixed point for some period of time, it is counted as recovered.
+Accordingly, trajectories that cycle around the fixed point are not counted as recovered until the cycles decay.
+To ensure reproducibility, all code is provided online in the Github repository: https://github.com/jdyeakel/SalmonStrays.
 
-\textcolor{cyan}{
-A strong "portfolio effect" as measured by PE in equation 6 indicates resilience against metapopulation extinction if PE is high due to a low CV of total population size.
+\noindent We have made the following edits to the text:\\
+\noindent {\bf Line XXX:} \emph{Recovery time was calculated by initiating a disturbance at $t=t_d$, and monitoring $N_T(t_d+t)$ as $t\rightarrow T$, where $T$ is large. 
+The aggregate was deemed recovered at $t_r$, such that recovery time was calculated as $t_r-t_d$, and recovery at $t=t_r$ was measured as the initial $t$ where $N_T(t) < {\rm SD}\left( N_T^* \right)$ for $t\in(t_r,T)$, where $\rm{SD}(\cdot)$ is standard deviation (illustrated in figure [ref]).}\\
+
+\noindent \textcolor{cyan}{
+{\bf Comment 1.5} A strong "portfolio effect" as measured by PE in equation 6 indicates resilience against metapopulation extinction if PE is high due to a low CV of total population size.
 But PE is also high if the local populations have high CV.
-As I argued above, near the bifurcation N_1 and N_2 may be much less stable than $N_T=N_1+N_2$, i.e., the peak of PE may have little to do with the resilience of total population size.
+As I argued above, near the bifurcation $N_1$ and $N_2$ may be much less stable than $N_T=N_1+N_2$, i.e., the peak of PE may have little to do with the resilience of total population size.
 }
 
-hehehe
+\noindent {\bf Response 1.5} We utilize the portfolio effect as a measurement to track the potential for the aggregate $N_T$ to buffer variance observed in the constituent populations $(N_1,N_2)$.
+Near the bifurcation, if $N_1$ and $N_2$ are more variable relative to $N_T$, then PE should be high, which is the principle goal of the measurement.
+However we agree that PE can change in response to changes to any of the four attributes from which it is composed (means of the aggregate, components; standard deviations of the aggregate, components), so it is sometimes hard to evaluate what is changing and what those changes mean relative to a single, descriptive measures of resilience such as recovery time.
+This is the primary motivation for incorporating both PE and recovery time in our analysis.
+
+% While we agree with the Reviewer that the spike in PE at the bifurcation is not necessarily descriptive of a resilient population, the attributes that lead to an increase in the PE to the right of the bifurcation (higher $m$) also results in shorter recovery times, and is measuring the potential for the aggregate to buffer variance.
+We do not put much weight in the PE spike at the bifurcation as an indicator of metapopulation health
 
 \subsection*{Minor comments:}
-\textcolor{cyan}{``straying" is not defined in the Introduction/main text. Since the model is not specific to salmon, the more general term "dispersal" could be used throughout.}
+\noindent \textcolor{cyan}{
+{\bf Minor Comment 1.1} ``straying" is not defined in the Introduction/main text. Since the model is not specific to salmon, the more general term "dispersal" could be used throughout.}
 
-\textcolor{cyan}{The figure legends should specify the parameter values used, figure 2 should give a key translating colours to values, figure 3 should have numbers on the axes.
+\noindent {\bf Minor Response 1.1} We now equate straying with dispersal early in the text (and note that it is also defined in the abstract).
+
+\noindent We have made the following edits to the text:\\
+\noindent {\bf Line XXX:} \emph{The rate at which individuals stray, $m$, is in this case synonymous with dispersal and may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices [refs].}\\
+
+\noindent \textcolor{cyan}{
+{\bf Minor Comment 1.2} The figure legends should specify the parameter values used, figure 2 should give a key translating colours to values, figure 3 should have numbers on the axes.
 Generally, the authors should pay attention to reproducibility.}
+
+\noindent {\bf Minor Response 1.2} We agree and thank the Reviewer for point this out. We have made the following edits to the text:\\
+\noindent {\bf Figure 1 caption} \emph{Unless otherwise indicated, the default parameter values used are: $r_{\rm max}=2$; $Z=0.5$; $\beta=0.001$; $\theta_1=5$; $\Delta\theta=5$; $\tau=1$; $\sigma=1$; $T=1\times10^5$.}\\
 
@@
 \subsection*{Major comments}
-My major criticism on this paper is that in the analysis presented, the bifurcation point is incorrectly identified and the claims based on this misidentification are most likely unsupported.
+\noindent \textcolor{cyan}{
+{\bf Comment 2.1} My major criticism on this paper is that in the analysis presented, the bifurcation point is incorrectly identified and the claims based on this misidentification are most likely unsupported.
 What is presented as a Fold bifurcation is in reality a Pitchfork bifurcation.
 Pitchfork bifurcations are dependent on perfect symmetry in systems, something that is also true in the current case: The model as presented and analyzed is formulated in an entirely symmetrical way, which means that a Pitchfork bifurcation is de facto built into it.
+}
 
-As it is, the occurrence of the bifurcation point as well as the ecological implications derived from it, are mostly a mathematical construction, brought about by the complete symmetry in the system.
+\noindent {\bf Response 2.1} We thank the Reviewer for their comment and have now clarified the manuscript. Please see Response 1.3.
+
+\noindent \textcolor{cyan}{
+{\bf Comment 2.2} As it is, the occurrence of the bifurcation point as well as the ecological implications derived from it, are mostly a mathematical construction, brought about by the complete symmetry in the system.
 It is absolutely necessary to investigate the robustness of the results under scenarios where the symmetry is broken.
 The eco-evolutionary aspects of the model have currently no role in the observed dynamics, because they hinge on the symmetrical formulation.
-
 I would be very curious to see this analysis extended under asymmetrical assumptions, for example in the productivity of the different locations or the intraspecific competition between populations.
@@
 I see that as the interesting angle; how do the dynamics caused by density-dependent straying interact with the dynamics caused by selective forces as imposed by the different environments on recruitment?
-
+\\
 This manuscript can be improved by correcting the bifurcation analysis and by extending beyond perfect symmetry.
@@
 This suggestion could directly be used to introduce asymmetries between two different locations and may be a feasible way to address the issues I raised above.
+}
 
-I would be looking forward to see a revised version of this manuscript that addresses my points.
+\noindent {\bf Response 2} We thank the Reviewer for their perspective and comments, and agree that we need to understand to what extent our results are generalizable.
+From our perspective, there are two ways in which our model incorporates symmetry: 1) the architectural symmetry of the populations between sites and 2) the symmetry in parameter values between sites.
+Regarding the first, both populations have the same growth and mortality functions, and both populations have the same functional forms that control straying.
+This architectural symmetry lies at the heart of our question, as we are exploring the dynamics of two similar populations that have similar constraints.
+We cannot alter this type of symmetry without vastly changing the questions that we are addressing.
 
-It would be good to include more technical details about how the bifurcation analysis was carried out (or is this present in the missing SI text?).
+We assume that the Reviewer is referring to the second type of symmetry: that many of the parameters of the model controlling vital rates are symmetric between sites, primary examples being: $r_{\rm max}=2;~Z=0.5;~\beta=0.001$.
+These parameters were held constant primarily to minimize the parameter space that we investigate (making an already long paper a bit shorter), but we acknowledge that this invites the question: are we investigating a mathematical knife-edge?
+
+We suggest that our analysis is not merely a mathematical construction (interpreted here to mean dynamic behaviors that only occur within a thin slice of parameter space, unlikely to be realized in biological systems) for 3 reasons:
+
+1) We already inject process error into the dynamics.
+This process error (independent for both sites) serves to push the system away from the symmetrical vital rates $(r_{\rm max}, Z, \beta)$, and if there is a knife-edge of parameter space that results in the dynamics we observe, the process error should knock the system away from it.
+Despite these independent sources of error, their means are the same, and this warrants additional investigation as the Reviewer suggests.
+
+2) We have included a modified analysis of the system to assess increasing asymmetry in the vital rates between sites.
+Asymmetry in parameter values is introduced as a parameter $\alpha$, where maximal growth at sites 1 and 2 are now  $r_{\rm max}(1)=r_{\rm max}(1+\tilde{rv}_1)$ and $r_{\rm max}(2)=r_{\rm max}(1+\tilde{rv}_2)$ where $rv_{1,2}$ are independently drawn from $\rm{Normal}(0,\alpha)$ and $r_{\rm max}=2$.
+Similarly the strength of density dependence is calculated at sites 1 and 2 as $\beta(1)=\beta(1+\tilde{rv}_1)$ and $\beta(2)=\beta(1+\tilde{rv}_2)$ where $\tilde{rv}_{1,2}$ are independently drawn from $\rm{Normal}(0,\alpha)$ and $\beta=0.001$.
+Thus as asymmetry increases, so do the differences in vital rates between sites.
+
+To investigate the effects of increasing asymmetry on the main dynamical feature of our model, we assessed how increasing $\alpha$ altered the steady state portrait as a function of the straying rate $m$ shown in figure 1a, and this analysis is now included as a supplementary figure (figure Sxx).
+As can be seen, greater asymmetry (warm colors in the supplementary figure) does not alter the qualitative nature of the dynamics, and does not significantly change the position of the bifurcation.
+We take this as strong evidence that the qualitative nature of the system is not a mathematical construction relegated to a small slice of parameter space.
+
+3) Reviewer 1 pointed to an embarrassing oversight on our part: that a similar model had been explored by Ronce \& Kirkpatrick (Evolution, 2001).
+This model differed substantially in its formulation: it is continuous (rather than discrete), evolution occurs in the mortality term (rather than the recruitment rate), the strength of selection is determined by the linear difference of trait values and their optima (rather than by mechanistically incorporating the fitness landscape a la Lande 1976), etc.
+Still, the underlying dynamics appear to accord with those that we describe.
+This suggests that the dynamics are potentially of a general nature, though we do not investigate how general (that would be an interesting pursuit for another time).
+
+Finally, although we like the idea of modeling changes in resource abundance in response to changes in populations and trait means, we feel that this is beyond the scope of the contribution outlined here, and suggest that it would be something very interesting to explore in a future work.
+
+\noindent \textcolor{cyan}{
+{\bf Comment 2.3} It would be good to include more technical details about how the bifurcation analysis was carried out (or is this present in the missing SI text?).
 As it is, I have no idea about the numerical techniques that were used or even whether the authors have studied both increasing and decreasing parameter values, as necessary to enable detection of alternative stable states and discontinuities.
+}
 
-Please include a table with the definitions and values of model parameters (including their units) and definitions of model variables.
+\noindent {\bf Response 2.3} We thank the Reviewer, and now briefly describe our analysis of the bifurcation (also see Response 1.3).
+We include a plot of the Jacobian eigenvalues across $m$ (figure SXX).
+Moreover, we note that hysteresis is observed at the bifurcation (evaluated by increasing and decreasing $m$), and this is shown in figure SXX.
 
-I could not understand section c, Habitat heterogeneity in the Model formulation.
+\noindent \textcolor{cyan}{
+{\bf Comment 2.4} Please include a table with the definitions and values of model parameters (including their units) and definitions of model variables.
+}
+
+\noindent {\bf Response 2.4} A table is now included.
+
+\noindent \textcolor{cyan}{
+{\bf Comment 2.5} I could not understand section c, Habitat heterogeneity in the Model formulation.
 In particular the paragraph on lines 139-144 should be explained and justified better, where the authors should make sure to explain the ‘integrate the two variables’ part more formally and explicitly.
+}
 
-As I write above, I appreciate the focus on robustness and through numerical analysis.
+\noindent {\bf Response 2.5} We thank the Reviewer for their comment and apologize for our lack of clarity.
+We have moved the text and attempt to explain our reasonings and justifications more clearly.
+We have made a figure that illustrates the relationship between straying and $\Delta\theta$ if it is assumed that individuals are less likely to stray into sites with very different trait optima (greater habitat heterogeneity).
+
+\noindent We have made the following edits to the text:\\
+\noindent {\bf Line XXX:} \emph{Although we largely treat habitat heterogeneity and the rate of straying as independent parameters, we evaluate a case where we assume that increased habitat heterogeneity correlates with lower straying rates, and vice versa.
+Two scenarios may lead to this correlation: 
+(\emph{i}) sites may be distributed over greater spatial distances, where habitat differences are assumed to be exaggerated and the likelihood of straying over greater distances would be lower [ref];
+(\emph{ii}) individuals may have behaviors promoting dispersal between habitats with structural or physiognomic similarities [ref].
+In this case, the rate of straying would be greater between habitats with smaller differences in trait optima (lower $\Delta\theta$) and lesser between habitats with greater differences in trait optima (higher $\Delta\theta$).}
+
+\noindent {\bf Line XXX:} \emph{Sites that are farther apart are likely to be more heterogeneous, while simultaneously facilitating lower rates of straying between populations.
+% If we assume that environmental heterogeneity increases with distance between sites, or that individuals may have behaviors that minimize their chances of straying into very different habitats, the rate of straying will decline with habitat heterogeneity.
+% We can implement this dependence by allowing $\Delta\theta$ to decrease as $m$ increases, such that there are low rates of straying between dissimilar populations and high rates of straying between similar populations.
+We implemented this dependence as set $m,m_0 = 0.5(1 + \Delta\theta)^{-1}$ where maximum straying is assumed to occur at $m=0.5$ (perfect mixing; figure [ref]).
+This assumes that $m$ or $m_0$ increases in situations where $\Delta\theta$ is lower, such that there are low rates of straying between dissimilar sites and high straying rates between similar sites.
+We find that under these conditions alternative stable states now appear for very low rates of straying $0 < m \leq 0.43$.
+As the straying rate increases and $\Delta\theta$ decreases, a single stable state emerges as the pitchfork bifurcation is crossed, which is opposite the pattern observed when these parameters are independent.}
+
+\noindent \textcolor{cyan}{
+{\bf Comment 2.6} As I write above, I appreciate the focus on robustness and through numerical analysis.
 At the same time this model would lend itself to some analysis too, even making some simplifying assumptions or investigating some boundary conditions.
@@
 This would also possibly facilitate very straightforward validation of the kind of bifurcation point that is encountered (although the figures clearly show it cannot be a Fold bifurcation).
+}
 
-Please present formal definitions (for example in a table as mentioned above) of all model variables and parameters. Right now I never found an explicit definition of NT, for example.
+\noindent {\bf Response 2.6} Please see Responses 1.3 and 2.3.
+
+\noindent \textcolor{cyan}{
+{\bf Comment 2.7} Please present formal definitions (for example in a table as mentioned above) of all model variables and parameters.
+Right now I never found an explicit definition of $N_T$, for example.
+}
+
+\noindent {\bf Response 2.7} A table is now included.
 
 \subsection*{Minor comments}
-I encountered some sloppiness in various parts of the manuscript, for example figure 3 is missing axes labels and has inconsistencies with the main text and caption.
+\noindent \textcolor{cyan}{
+{\bf Minor Comment 2.1} I encountered some sloppiness in various parts of the manuscript, for example figure 3 is missing axes labels and has inconsistencies with the main text and caption.
 I did not find the SI figures or text, only the figure captions.
@@
 \end{itemize}
+}
+
+\noindent {\bf Minor Response 2.1} We thank the Reviewer for being so meticulous and pointing out these errors.
+Although the noted issues with the figures (missing axes, labels) and missing supplemental materials appears to have been a problem with the upload, we acknowledge that the other issues were due to a lack of thoroughness on our part and have been corrected.
 
@@
 Justin D. Yeakel\\
-Assistant Professor\\
-University of California, Merced}
+Jean Phillipe Gibert\\
+Peter Westley\\
+Jonathan Moore}
 }

### 2017-09-12 Manuscript/FinalDraft_rev/msv6.tex
edits

@@
 
+% \usepackage{longtable}
+
 \linespread{1.5}
@@
 Recently, the role of social interactions and collective navigation has been hypothesized \citep[][This issue]{Berdahl:2015kv,Berdahl:2016dx}.
-The rate at which individuals stray, $m$, may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \citep{Simons:2004jo,Berdahl:2015kv,Berdahl:2016dx}.
+The rate at which individuals stray, $m$, is in this case synonymous with dispersal and may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \citep{Simons:2004jo,Berdahl:2015kv,Berdahl:2016dx}.
 The potential influence of collective dispersal on the dynamics of individual populations and the metapopulation as a whole is a topic of considerable interest that has tangible conservation implications \citep{Brenner:2012gl,Johnson:2012fe,Fullerton:2011ii}.
@@
 % These mosaics are in part driven by environmental differences between habitats that alter the selective forces acting on different phenotypes \citep{Endler:1986tz}, and a principle underlying assumption is that there is gene flow such that individuals from different habitats mix over space \citep{Gomulkiewicz:2000bz,Thompson:2002hr,Nuismer:2003eh,Nuismer:2006be,Forde:2008kc,Guimaraes:2011hu,Gibert:2013hp}.
-% Although the evolutionary outcomes of these spatial processes have been explored in depth (REFS), it is less well understood how selective mosaics and their consequent evolutionary forces impact population dynamics as they unfold \citep{Hendry:2016un}.
+% Although the evolutionary outcomes of these spatial processes have been explored in depth (REFS), it is less well understood how selective mosaics and their consequent evolutionary forces impact population dynamics as they unpitchfork \citep{Hendry:2016un}.
 
@@
 \noindent{\bf (a) Metapopulation framework}\\
-\noindent We consider two populations $N_1$ and $N_2$ that inhabit two distinct habitats, each with trait values $x_1$ and $x_2$ determing recruitment rates.
+\noindent We consider two populations $N_1$ and $N_2$ that inhabit two distinct habitats, each with trait values $x_1$ and $x_2$ determining recruitment rates.
 We assume that there is an optimum trait value $\theta_1$ and $\theta_2$ associated with each habitat, where recruitment is maximized if the trait value of the local population equals the optimum, such that $x = \theta$.
@@
 % \noindent{\bf (b) Recruitment over a selective landscape}
-\noindent Because individuals from the local population are mixed with individuals from the remote population via straying and subsequent reproduction, the resulting trait distribution is a mixed normal with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and for the straying individuals, $1-w_i$, where
+\noindent Because individuals from the local population are mixed with individuals from the remote population via straying and subsequent reproduction, the resulting trait distribution is a mixed normal with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and straying individuals, $1-w_i$, where
 \begin{equation}
@@
 \ref{eq:mu} determines how the mean trait changes through time due to natural selection \citep{Lande:1976ga}, which is proportional to the change in mean fitness with respect to $\mu_i$.
+%Comparison to the Ronce and Kirkpatrick paper
+This model formulation has parallels to that proposed by Ronce and Kirkpatrick \citep{Ronce:2001dp}, where habitat specialization evolves between two populations as a function of dispersal.
+The largest difference between these approaches is that our framework treats trait evolution mechanistically at some cost to analytical tractability.
+Importantly, we show that the resulting dynamics are qualitatively similar, suggesting that the dynamical features present in both of these approaches have potentially widespread ramifications for the evolutionary dynamics of spatially connected populations.
 \\
@@
 %Linking m/m0 and thetadiff
-Although we largely treat habitat heterogeneity and the rate of straying as independent parameters, we evaluate a case where we assume thgat increased habitat heterogeneity correlates with lower straying rates, and vice versa.
+Although we largely treat habitat heterogeneity and the rate of straying as independent parameters, we evaluate a case where we assume that increased habitat heterogeneity correlates with lower straying rates, and vice versa (illustrated in figure \ref{fig:mthetarelation}).
 Two scenarios may lead to this correlation: 
-(\emph{i}) sites may be distributed over greater spatial distances, where habitat dissimilarity is assumed to be larger between more distant sites, whereas the likelihood of straying over greater distances would likewise be lower \citep{Candy:2000hu,JPE:JPE1383};
+(\emph{i}) sites may be distributed over greater spatial distances, where habitat differences are assumed to be exaggerated and the likelihood of straying over greater distances would be lower \citep{Candy:2000hu,JPE:JPE1383};
 (\emph{ii}) individuals may have behaviors promoting dispersal between habitats with structural or physiognomic similarities \citep{Peterson:2014gy}.
-
-Where indicated, we account for correlation between the straying rate and the differnce between optimal trait values by assuming that $m$ (if the stray rate is constant) or $m_0$ (if the stray rate is density-dependent) is a function of the difference between optimal trait values between sites $\Delta\theta$.
-The straying rate is maximized at $m_{\rm max} = 0.5$, assuming both sites are equally attractive to the respective populations.
-Thus, we can integrate these two variables by setting $m,m_0 = (1/m_{\rm max} + \epsilon \Delta\theta)^{-1}$, where $\epsilon$ sets the sensitivity of a declining $m$ to increasing distance (greater values of $\Delta\theta$), which we set to $\epsilon=2$.\\
+In this case, the rate of straying would be greater between habitats with smaller differences in trait optima (lower $\Delta\theta$) and lesser between habitats with greater differences in trait optima (higher $\Delta\theta$).
+
+% Where indicated, we account for correlation between the straying rate and the difference between optimal trait values by assuming that $m$ (if the stray rate is constant) or $m_0$ (if the stray rate is density-dependent) is a function of the difference between optimal trait values between sites $\Delta\theta$.
+% The straying rate is maximized at $m_{\rm max} = 0.5$, assuming both sites are equally attractive to the respective populations.
+% Thus, we can integrate these two variables by setting $m,m_0 = (1/m_{\rm max} + \epsilon \Delta\theta)^{-1}$, where $\epsilon$ sets the sensitivity of a declining $m$ to increasing distance (greater values of $\Delta\theta$), which we set to $\epsilon=2$.\\
 
@@
 Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation \citep{GuckHolmes}, because we aimed to 1) assess the effects of a large perturbation, and 2) estimate the time required for all transient effects to decay (including dampened oscillations), we used a simulation-based numerical procedure.
-
-Numerically estimating the time that it takes for a perturbed system to relax also permits a more detailed perspective of metapopulation fragility.
-For example, if populations settle to alternative stable states (alternative stable states in our model requiring one population to be high-density and one low-density), comparing recovery times after a disturbance applied to the high, low, and/or both populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery.
-We measured recovery time following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} are equivalent if the system is in the single steady state regime); (\emph{iii}) near-collapse of both populations where just 1.0\% of each survives.
-Throughout, we will refer to an increase in the portfolio effects and/or reduction in recovery times as promoting metapopulation robustness, which is expected to have a positive effect on persistence.
+Recovery time was calculated by initiating a disturbance at $t=t_d$, and monitoring $N_T(t_d+t)$ as $t\rightarrow T$, where $T$ is large. 
+The aggregate was deemed recovered at $t_r$, such that recovery time was calculated as $t_r-t_d$, and recovery at $t=t_r$ was measured as the initial $t$ where $N_T(t) < {\rm SD}\left( N_T^* \right)$ for $t\in(t_r,T)$, where $\rm{SD}(\cdot)$ is standard deviation (illustrated in figure \ref{fig:recovery}).
+% The system was considered `recovered' at time $t=t_r$ where $N_T(t_d+t_r) < \pm {\rm SD}[N_T(t=T)]$ for $t = (t_r,T)$, assuming SD is standard deviation and $T$ is large.
+
+Numerically estimating the time that it takes for a perturbed system to recover also permits a more detailed perspective of metapopulation fragility.
+For example, if populations settle to alternative stable states, comparing recovery times after a disturbance applied to individual populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery.
+% (alternative stable states in our model requiring one population to be high-density and one low-density)
+We measured recovery time following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} are equivalent if populations have the same density); (\emph{iii}) near-collapse of both populations where just 1.0\% of each survives.
+Throughout, we will refer to an increase in the portfolio effects and/or reduction in recovery times as promoting metapopulation robustness, which promotes persistence.
 \\
@@
 This prediction is in accordance with observations from natural populations \citep{Bett:2017ha}. %, regardless of trait heritability and variance or habitat heterogeneity
-The decline in steady state densities is not gradual: as straying increases, the system crosses a fold bifurcation whereby the single steady state for the metapopulation bifurcates into two alternative stable states: one at high biomass, and one at low biomass density (figure \ref{fig:traj}a, \ref{fig:PE}a).
+The decline in steady state densities is not gradual: as straying increases, the system crosses a pitchfork bifurcation (PFB) \citep{AleksandrovichKuznetsov:1995p2580} whereby the single steady state for the metapopulation bifurcates into two basins of attraction: one at high biomass, and one at low biomass density (figure \ref{fig:traj}a, \ref{fig:PE}a).
+In discrete systems, a pitchfork bifurcation is a specific case of the cusp bifurcation (which occurs when two fold bifurcations intersect at a cusp), which occurs when the real part of the dominant eigenvalue of the Jacobian matrix crosses the unit circle at +1, which we show in figure \ref{fig:eigs}.
+
 Mean trait values for both populations bifurcate similarly (figure \ref{fig:traj}b), depending on which population attains a low- vs. high-density. 
-Above the threshold straying rate defined by the fold bifurcation, there are two alternative eco-evolutionary states: the \emph{dominant state} population will have a higher a higher density and higher degree of local adaptation (lower trait offset from the local optimum), while the \emph{subordinate state} population will have lower density with maladapted trait values (higher trait offset from the local optimum).
-Whether a specific population goes to one state or the other in our model is random, which is due to a small amount of introduced variance in the initial conditions.
-% Thus, we predict that high levels of straying in metapopulations will produce alternative eco-evolutionary states.
-
-Trait heritability has a large impact on the degree to which straying rate affects both the aggregate population steady state density ($N^*_T=N^*_1+N^*_2$; figure \ref{fig:PE}a) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N^*=|N^*_1-N^*_2|$; figure \ref{fig:PE}b).
+Above the threshold straying rate defined by the PFB, there are two alternative eco-evolutionary states: the \emph{dominant state} population will have a higher density and higher degree of local adaptation (smaller trait offset from the local optimum), while the \emph{subordinate state} population will have lower density with maladapted trait values (larger trait offset from the local optimum).
+Hysteresis is observed to occur at this transition, such that the single steady state regime cannot be easily recovered by lowering $m$ after the system attains alternative steady states (figure \ref{fig:hysteresis}).
+These dynamics are also observed in the Ronce and Kirkpatrick model, where populations are described as transitioning between symmetric to asymmetric states \citep{Ronce:2001dp}.
+Whether a specific population goes to one state or the other in our model is random, due to a small amount of introduced variance in the initial conditions.
+
+Trait heritability has a large impact on the degree to which straying affects both the aggregate population steady state density ($N^*_T=N^*_1+N^*_2$; figure \ref{fig:PE}a) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N^*=|N^*_1-N^*_2|$; figure \ref{fig:PE}b).
 Greater trait heritability results in a faster decline in $N_T^*$ with increasing straying rates $m$, but leads to only moderate changes to $\Delta N^*$.
 Conversely, in the context of lower trait heritability, an increase in the straying rate has little impact on the total biomass density but contrastingly large effects on $\Delta N^*$.
-The fold bifurcation (the black line in Figs.
+The pitchfork bifurcation (the black line in Figs.
 \ref{fig:PE}a-c) occurs at lower values of the straying rate $m$ with decreased trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
@@
 %Critical slowing down
-As the fold bifurcation is approached with increasing $m$, the portfolio effect increases sharply due to an amplification in variance within both donor and recipient populations.
-This amplification in variance is the product of a dynamical process known as \emph{critical slowing down} that occurs near fold bifurcations \citep{Scheffer:2009gg}, a phenomenon that some have suggested may serve as an early warning indicator for approaching phase transitions \citep{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br,Krkosek:2014ch}.
-For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}a-c), where alternative stable states occur, the portfolio effect declines steadily as the CV of $N_T^*$ increases.
+As the pitchfork bifurcation is approached with increasing $m$, the portfolio effect increases sharply due to an amplification in variance within both donor and recipient populations.
+This amplification in variance is the product of a dynamical process known as \emph{critical slowing down} that occurs near pitchfork bifurcations \citep{Scheffer:2009gg}, a phenomenon that some have suggested may serve as an early warning indicator for approaching phase transitions \citep{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br,Krkosek:2014ch}.
+For larger values of $m$ (to the right of the pitchfork bifurcation in Fig \ref{fig:PE}a-c), where alternative stable states occur, the portfolio effect declines steadily as the CV of $N_T^*$ increases.
 The decline over $m$ is more gradual if trait heritability is low, and steeper if trait heritability is high (figure \ref{fig:PE}c).
@@
 Because the mean trait values of both populations are skewed towards those of the dominant population, when the subordinate population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into the recovering population extends the length of time required for it to return to its steady state (figure \ref{fig:relaxtraj_ldlh}).
-When both populations are both dramatically reduced, recovery time is generally fastest at lower levels of straying, while near the onset of the fold bifurcation, recovery time increases explosively and this is -- as the name implies -- characteristic of \emph{slowing} dynamics that occur near critical transitions \citep{Scheffer:2009gg,Kuehn:2010p2591}.\\
+When both populations are both dramatically reduced, recovery time is generally fastest at lower levels of straying, while near the onset of the pitchfork bifurcation, recovery time increases explosively and this is -- as the name implies -- characteristic of \emph{slowing} dynamics that occur near critical transitions \citep{Scheffer:2009gg,Kuehn:2010p2591}.\\
 
@@
 % with decreased $h^2$.
-Less intuitively, we observe a discrete jump towards low phenotypic diversity as the fold bifurcation is crossed (figure \ref{fig:traitdiff}).
+Less intuitively, we observe a discrete jump towards low phenotypic diversity as the pitchfork bifurcation is crossed (figure \ref{fig:traitdiff}).
 Although the development of alternative stable states elevates the portfolio effect due to the variance-dampening effects of the aggregate, entering this dynamic regime also results in a substantial decline in phenotypic diversity, which may have less predictable adverse effects on the population.
@@
 %regardless of heritability.
-%The rate of straying that gives rise to a large PE at the fold bifurcation is conditioned on the difference in optimal trait values between the local and remote populations, which we call $\Delta\theta = \sqrt{(\theta_i - \theta_j)^2}$.
-As habitat heterogeneity increases, alternative stable states appear at lower straying rates -- with the crossing of the fold bifurcation, accompanied by a peak in the PE -- whereas the magnitude of increase in the PE also increases (figure \ref{fig:thetaPE}b), reducing recovery time (figure \ref{fig:relaxtheta}).
+%The rate of straying that gives rise to a large PE at the pitchfork bifurcation is conditioned on the difference in optimal trait values between the local and remote populations, which we call $\Delta\theta = \sqrt{(\theta_i - \theta_j)^2}$.
+As habitat heterogeneity increases, alternative stable states appear at lower straying rates -- with the crossing of the pitchfork bifurcation, accompanied by a peak in the PE -- whereas the magnitude of increase in the PE also increases (figure \ref{fig:thetaPE}b), reducing recovery time (figure \ref{fig:relaxtheta}).
 For increased rates of straying, greater habitat heterogeneity erodes the PE (figure \ref{fig:thetaPE}b) and increases the recovery time (figure \ref{fig:relaxtheta}).
@@
 
-If we assume that environmental heterogeneity increases with distance between sites, or that individuals may have behaviors that minimize their chances of straying into very different habitats, the rate of straying will decline with habitat heterogeneity.
-Accordingly, $\Delta\theta$ will decrease as $m$ increases, such that there are low rates of straying between dissimilar populations and high rates of straying between similar populations.
+% We next investigate the dynamics across different types of sites that host 
+Sites that are farther apart are likely to be more heterogeneous, while simultaneously facilitating lower rates of straying between populations.
+% If we assume that environmental heterogeneity increases with distance between sites, or that individuals may have behaviors that minimize their chances of straying into very different habitats, the rate of straying will decline with habitat heterogeneity.
+% We can implement this dependence by allowing $\Delta\theta$ to decrease as $m$ increases, such that there are low rates of straying between dissimilar populations and high rates of straying between similar populations.
+We implemented this dependence as set $m,m_0 = 0.5(1 + \Delta\theta)^{-1}$ where maximum straying is assumed to occur at $m=0.5$ (perfect mixing; figure \ref{fig:mthetarelation}).
+This assumes that $m$ or $m_0$ increases in situations where $\Delta\theta$ is lower, such that there are low rates of straying between dissimilar sites and high straying rates between similar sites.
 We find that under these conditions alternative stable states now appear for very low rates of straying $0 < m \leq 0.43$.
-As the straying rate increases and $\Delta\theta$ decreases, a single stable state emerges as the fold bifurcation is crossed, which is opposite the pattern observed when these parameters are independent.
+As the straying rate increases and $\Delta\theta$ decreases, a single stable state emerges as the pitchfork bifurcation is crossed, which is opposite the pattern observed when these parameters are independent.
+
 There are two notable dynamics that emerge following extinction of the dominant population at low rates of straying between dissimilar (high $\Delta\theta$) sites:
@@
 % %The bifurcation and critical slowing down
-% [This paragraph is interesting to me but could be cut] Although the portfolio effect is negatively correlated with recovery time, there is one exception: the point of transition from a single steady state to alternative stable states, which occurs at the fold bifurcation (REF).
+% [This paragraph is interesting to me but could be cut] Although the portfolio effect is negatively correlated with recovery time, there is one exception: the point of transition from a single steady state to alternative stable states, which occurs at the pitchfork bifurcation (REF).
 % At this point, the portfolio effect is maximized, but so is the recovery time (figures \ref{fig:PE}, \ref{fig:relax}).
@@
 %alternative stable states & critical slowing
-A salient finding from our results was that straying can lead to the emergence of alternative stable states, pushing one of the populations to high density (the \emph{dominant state}), and one to low density (the \emph{subordinate state}).
-This asymmetry in population densities also leads to an asymmetry in the mean trait values, both of which are skewed towards the local optimum of the dominant population. 
-The formation of alternative stable states in our system is a relatively common example of spatial pattern formation.
-%, only occurring above a threshold straying rate that largely depends on trait heritability and the difference in trait optima between sites.
-Pattern formation can occur as a consequence of myriad ecological processes, including habitat selection, aggregation, local environmental conditions, and/or interspecific interactions such as competition or predation, and is inexorably linked to scale \citep{Levin:1992p1472}.
-Pattern formation is also known to occur with the onset of spatially or diffusion-induced instabilities, or Turing instabilities \citep{Meinhardt:2012hw}. 
-Here, we describe an additional mechanism by which pattern formation in local densities can occur, when the onset of collective movement-mediated maladaptation leads to local differences in reproductive rates resulting in alternative stable states across sites.
-%This mechanism of pattern formation through the simultaneous interplay between ecological and evolutionary dynamics may open up new areas of research.
-While metapopulation approaches have widely recognized the importance of source and sink populations \citep{Hanski:1999vu} as well as the occurrence of alternative states \citep{Boughton:1999fa}, here we show that the coupled forces of ecological and evolutionary processes can result in similar populations diverging into alternative states where one population is large and well-adapted, while the other is smaller and maladapted.
-The appearance of these alternative states is largely dependent on the dispersal, or rate of straying, between sites.
+% A salient finding from our results was that straying can lead to the emergence of alternative stable states, pushing one of the populations to high density (the \emph{dominant state}), and one to low density (the \emph{subordinate state}).
+% This asymmetry in population densities also leads to an asymmetry in the mean trait values, both of which are skewed towards the local optimum of the dominant population. 
+% The formation of alternative stable states in our system is a relatively common example of spatial pattern formation. 
+% Pattern formation can occur as a consequence of myriad ecological processes, including habitat selection, aggregation, local environmental conditions, and/or interspecific interactions such as competition or predation, and is inexorably linked to scale \citep{Levin:1992p1472}.
+% Pattern formation is also known to occur with the onset of spatially or diffusion-induced instabilities, or Turing instabilities \citep{Meinhardt:2012hw}. 
+% Here, we describe an additional mechanism by which pattern formation in local densities can occur, when the onset of collective movement-mediated maladaptation leads to local differences in reproductive rates resulting in alternative stable states across sites.
+% While metapopulation approaches have widely recognized the importance of source and sink populations \citep{Hanski:1999vu} as well as the occurrence of alternative states \citep{Boughton:1999fa}, here we show that the coupled forces of ecological and evolutionary processes can result in similar populations diverging into alternative states where one population is large and well-adapted, while the other is smaller and maladapted.
+% The appearance of these alternative states is largely dependent on the dispersal, or rate of straying, between sites.
+
+A salient finding from our results is that straying can lead to the emergence of alternative stable states, pushing one of the populations to high density (the \emph{dominant state}), and one to low density (the \emph{subordinate state}).
+This pattern has been observed in other models of eco-evolutionary dynamics with explicit space \citep{Ronce:2001dp}, suggesting that it may be a general feature of spatially-linked traits that evolve toward local optima while being hindered by dispersal.
+An important quality of our framework is that there are similar forces that dictate interactions within and between sites, and this naturally results in a symmetry that could preclude the relevance of our finding for natural (and inherently less symmetric) systems.
+Although the emergence of alternative stable states via a pitchfork bifurcation is characteristic of symmetrical systems, we find that increasing the asymmetry in the vital rates of populations across sites does not alter the presence or position of the PFB (figure \ref{fig:symmetry}).
+That these patterns arise in alternative formulations and are insensitive to parameter asymmetry suggests that the emergence of dominant and subordinate states directly from eco-evolutionary dynamics may be a generally relevant phenomenon in ecological systems.
+%, though this requires additional investigation and scrutiny.
 
@@
 
-% \bibliography{aa_kevin}
-
-\begin{thebibliography}{10}
-\expandafter\ifx\csname urlstyle\endcsname\relax
-  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
-  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
-  \urlstyle{rm}\Url}\fi
-
-\bibitem{Brown:1977gk}
-Brown JH, Kodric-Brown A, 1977 {Turnover rates in insular biogeography: effect
-  of immigration on extinction}.
-\newblock \emph{Ecology} \textbf{58}, 445--449
-
-\bibitem{Earn:2000fm}
-Earn DJD, Levin SA, Rohani P, 2000 {Coherence and conservation}.
-\newblock \emph{Science} \textbf{290}, 1360--1364
-
-\bibitem{Anonymous:2015gf}
-Schindler DE, Armstrong JB, Reed TE, 2015 {The portfolio concept in ecology and
-  evolution}.
-\newblock \emph{Front. Ecol. Environ.} \textbf{13}, 257--263
-
-\bibitem{Abbott:2017hl}
-Abbott RE, Doak DF, Peterson ML, 2017 {Portfolio effects, climate change, and
-  the persistence of small populations: analyses on the rare plant Saussurea
-  weberi.}
-\newblock \emph{Ecology} \textbf{98}, 1071--1081
-
-\bibitem{Thibaut:2012km}
-Thibaut LM, Connolly SR, 2013 {Understanding diversity-stability relationships:
-  towards a unified model of portfolio effects.}
-\newblock \emph{Ecol. Lett.} \textbf{16}, 140--150
-
-\bibitem{MilnerGulland:2011vm}
-Milner-Gulland EJ, Fryxell JM, Sinclair ARE, 2011 \emph{{Animal Migration}}.
-\newblock Oxford: Oxford University Press
-
-\bibitem{Bell:2011ki}
-Bell G, Gonzalez A, 2011 {Adaptation and evolutionary rescue in metapopulations
-  experiencing environmental deterioration}.
-\newblock \emph{Science} \textbf{332}, 1327--1330
-
-\bibitem{Carlson:2014is}
-Carlson SM, Cunningham CJ, Westley PAH, 2014 {Evolutionary rescue in a changing
-  world}.
-\newblock \emph{Trends Ecol. Evol.} \textbf{29}, 521--530
-
-\bibitem{Muhlfeld:2014hs}
-Muhlfeld CC, Kovach RP, Jones LA, Al-Chokhachy R, Boyer MC, Leary RF, Lowe WH,
-  Luikart G, Allendorf FW, 2014 {Invasive hybridization in a threatened species
-  is accelerated by climate change}.
-\newblock \emph{Nature Climate Change} \textbf{4}, 620--624
-
-\bibitem{Lowe:2015ft}
-Lowe WH, Muhlfeld CC, Allendorf FW, 2015 {Spatial sorting promotes the spread
-  of maladaptive hybridization}.
-\newblock \emph{Trends Ecol. Evol.} \textbf{30}, 456--462
-
-\bibitem{H:2013fs}
-Westley PAH, Quinn TP, 2013 {Rates of straying by hatchery-produced Pacific
-  salmon (\emph{Oncorhynchus} spp.) and steelhead (\emph{Oncorhynchus mykiss})
-  differ among species, life history types, and populations}.
-\newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{70}, 735--746
-
-\bibitem{Keefer:2014gg}
-Keefer ML, Caudill CC, 2014 {Homing and straying by anadromous salmonids: a
-  review of mechanisms and rates}.
-\newblock \emph{Rev Fish Biol Fisher} \textbf{24}, 333--368
-
-\bibitem{Bett:2017ha}
-Bett NN, Hinch SG, Burnett NJ, Donaldson MR, Naman SM, 2017 {Causes and
-  consequences of straying into small populations of Pacific salmon}.
-\newblock \emph{Fisheries} \textbf{42}, 220--230
-
-\bibitem{Berdahl:2015kv}
-Berdahl A, Torney CJ, Schertzer E, Levin SA, 2015 {On the evolutionary
-  interplay between dispersal and local adaptation in heterogeneous
-  environments}.
-\newblock \emph{Evolution} \textbf{69}, 1390--1405
-
-\bibitem{Berdahl:2016dx}
-Berdahl A, 2016 {Collective behavior as a driver of critical transitions in
-  migratory populations}.
-\newblock \emph{Movement Ecology} \textbf{4}, 1--12
-
-\bibitem{Simons:2004jo}
-Simons AM, 2004 {Many wrongs: The advantage of group navigation}.
-\newblock \emph{Trends Ecol. Evol.} \textbf{19}, 453--455
-
-\bibitem{Brenner:2012gl}
-Brenner RE, Moffitt SD, Grant WS, 2012 {Straying of hatchery salmon in Prince
-  William Sound, Alaska}.
-\newblock \emph{Environ Biol Fish} \textbf{94}, 179--195
-
-\bibitem{Johnson:2012fe}
-Johnson RC, Weber PK, Wikert JD, Workman ML, MacFarlane RB, Grove MJ, Schmitt
-  AK, 2012 {Managed metapopulations: do salmon hatchery
-  {\textquoteleft}sources{\textquoteright} lead to in-river
-  {\textquoteleft}sinks{\textquoteright} in conservation?}
-\newblock \emph{PLoS ONE} \textbf{7}, e28880--11
-
-\bibitem{Fullerton:2011ii}
-Fullerton AH, Lindley ST, Pess GR, Feist BE, Steel EA, Mcelhany P, 2011 {Human
-  influence on the spatial structure of threatened Pacific salmon
-  metapopulations}.
-\newblock \emph{Conserv Biol} \textbf{25}, 932--944
-
-\bibitem{Quinn:2011tf}
-Quinn TP, 2011 \emph{{The Behavior and Ecology of Pacific Salmon and Trout}}.
-\newblock Vancouver: UBC Press
-
-\bibitem{Jonsson:2011kg}
-Jonsson B, Jonsson N, 2011 \emph{{Ecology of Atlantic Salmon and Brown Trout}}.
-\newblock Dordrecht: Springer Netherlands
-
-\bibitem{Quinn:1993ge}
-Quinn TP, 1993 {A review of homing and straying of wild and hatchery-produced
-  salmon}.
-\newblock \emph{Fish Res} \textbf{18}, 29--44
-
-\bibitem{Hendry:2004wf}
-Hendry AP, Bohlin T, Jonsson B, Berg OK, 2004 {The evolution of philopatry and
-  dispersal: homing versus straying in salmonids}.
-\newblock In AP~Hendry, SC~Stearns, eds., \emph{Evolution Illuminated}. Oxford:
-  Oxford University Press on Demand
-
-\bibitem{Schtickzelle:2007wb}
-Schtickzelle N, Quinn TP, 2007 {A metapopulation perspective for salmon and
-  other anadromous fish}.
-\newblock \emph{Fish Fish.} \textbf{8}, 297--314
-
-\bibitem{Anderson:2014cx}
-Anderson SC, Moore JW, McClure MM, Dulvy NK, Cooper AB, 2015 {Portfolio
-  conservation of metapopulations under climate change}.
-\newblock \emph{Ecol. Appl.} \textbf{25}, 559--572
-
-\bibitem{Keefer:2008bs}
-Keefer ML, Caudill CC, Peery CA, Lee SR, 2008 {Transporting juvenile salmonids
-  around dams impairs adult migration}.
-\newblock \emph{Ecol. Appl.} \textbf{18}, 1888--1900
-
-\bibitem{Westley:2015to}
-Westley PAH, Dittman AH, Ward EJ, Quinn TP, 2015 {Signals of climate,
-  conspecific density, and watershed features in patterns of homing and
-  dispersal by Pacific salmon.}
-\newblock \emph{Ecology} \textbf{96}, 2823--2833
-
-\bibitem{Bond:2016dz}
-Bond MH, Westley PAH, Dittman AH, Holecek D, Marsh T, Quinn TP, 2016 {Combined
-  effects of barge transportation, river environment, and rearing location on
-  straying and migration of adult snake river fall-run Chinook salmon}.
-\newblock \emph{T Am Fish Soc} \textbf{146}, 60--73
-
-\bibitem{Peterson:2014gy}
-Peterson DA, Hilborn R, Hauser L, 2014 {Local adaptation limits lifetime
-  reproductive success of dispersers in a wild salmon metapopulation}.
-\newblock \emph{Nat Commun} \textbf{5}, 3696
-
-\bibitem{Moore:2010gs}
-Moore JW, McClure M, Rogers LA, Schindler DE, 2010 {Synchronization and
-  portfolio performance of threatened salmon}.
-\newblock \emph{Conserv. Lett.} \textbf{3}, 340--348
-
-\bibitem{Carlson:2011ce}
-Carlson SM, Satterthwaite WH, Fleming IA, 2011 {Weakened portfolio effect in a
-  collapsed salmon population complex}.
-\newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{68}, 1579--1589
-
-\bibitem{Braun:2016ib}
-Braun DC, Moore JW, Candy J, Bailey RE, 2016 {Population diversity in salmon:
-  linkages among response, genetic and life history diversity}.
-\newblock \emph{Ecography} \textbf{39}, 317--328
-
-\bibitem{simpson1953major}
-Simpson GG, 1953 \emph{{The Major Features of Evolution}}.
-\newblock New York: Simon and Schuster
-
-\bibitem{Lande:1976ga}
-Lande R, 1976 {Natural selection and random genetic drift in phenotypic
-  evolution}.
-\newblock \emph{Evolution} \textbf{30}, 314--334
-
-\bibitem{Shelton:2011eq}
-Shelton AO, Mangel M, 2011 {Fluctuations of fish populations and the magnifying
-  effects of fishing.}
-\newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{108}, 7075--7080
-
-\bibitem{Jasper:2013cc}
-Jasper JR, Habicht C, Moffitt S, Brenner R, Marsh J, Lewis B, Fox EC, Grauvogel
-  Z, Olive SDR, Grant WS, 2013 {Source-sink estimates of genetic introgression
-  show influence of hatchery strays on wild chum salmon populations in Prince
-  William Sound, Alaska}.
-\newblock \emph{PLoS ONE} \textbf{8}, e81916
-
-\bibitem{Schreiber:2011wx}
-Schreiber SJ, B{\"u}rger R, Bolnick DI, 2011 {The community effects of
-  phenotypic and genetic variation within a predator population.}
-\newblock \emph{Ecology} \textbf{92}, 1582--1593
-
-\bibitem{Gilbert:2014ee}
-Gibert JP, Brassil CE, 2014 {Individual phenotypic variation reduces
-  interaction strengths in a consumer{\textendash}resource system}.
-\newblock \emph{Ecol Evol} \textbf{4}, 3703--3713
-
-\bibitem{Gibert:2015kc}
-Gibert JP, DeLong JP, 2015 {Individual variation decreases interference
-  competition but increases species persistence}.
-\newblock \emph{Adv Ecol Res} \textbf{52}, 45--64
-
-\bibitem{Berdahl:2014bl}
-Berdahl A, Westley PAH, Levin SA, Couzin ID, Quinn TP, 2014 {A collective
-  navigation hypothesis for homeward migration in anadromous salmonids}.
-\newblock \emph{Fish Fish.} \textbf{17}, 525--542
-
-\bibitem{Candy:2000hu}
-Candy JR, Beacham TD, 2000 {Patterns of homing and straying in southern British
-  Columbia coded-wire tagged Chinook salmon (\emph{Oncorhynchus tshawytscha})
-  populations}.
-\newblock \emph{Fish Res} \textbf{47}, 41--56
-
-\bibitem{JPE:JPE1383}
-Schick RS, Lindley ST, 2007 {Directed connectivity among fish populations in a
-  riverine network}.
-\newblock \emph{J. Appl. Ecol.} \textbf{44}, 1116--1126
-
-\bibitem{Schindler:2015gf}
-Schindler DE, Armstrong JB, Reed TE, 2015 {The portfolio concept in ecology and
-  evolution}.
-\newblock \emph{Front. Ecol. Environ.} \textbf{13}, 257--263
-
-\bibitem{Ovaskainen:2002il}
-Ovaskainen O, Hanski I, 2002 {Transient dynamics in metapopulation response to
-  perturbation}.
-\newblock \emph{Theor Popul Biol} \textbf{61}, 285--295
-
-\bibitem{Anderson:2013gb}
-Anderson SC, Cooper AB, Dulvy NK, 2013 {Ecological prophets: quantifying
-  metapopulation portfolio effects}.
-\newblock \emph{Methods Ecol Evol} \textbf{4}, 971--981
-
-\bibitem{Loreau:2008ju}
-Loreau M, de~Mazancourt C, 2008 {Species synchrony and its drivers: neutral and
-  nonneutral community dynamics in fluctuating environments}.
-\newblock \emph{Am. Nat.} \textbf{172}, E48--E66
-
-\bibitem{Yeakel:2013vz}
-Yeakel JD, Moore JW, Guimar{\~a}es~Jr PR, de~Aguiar MAM, 2014 {Synchronisation
-  and stability in river metapopulation networks}.
-\newblock \emph{Ecol. Lett.} \textbf{17}, 273--283
-
-\bibitem{GuckHolmes}
-Guckenheimer J, Holmes P, 1983 \emph{{Nonlinear Oscillations, Dynamical
-  Systems, and Bifurcations of Vector Fields}}.
-\newblock New York: Springer
-
-\bibitem{Carlson:2008hl}
-Carlson SM, Seamons TR, 2008 {A review of quantitative genetic components of
-  fitness in salmonids: implications for adaptation to future change}.
-\newblock \emph{Evol Appl} \textbf{1}, 222--238
-
-\bibitem{Scheffer:2009gg}
-Scheffer M, Bascompte J, Brock WA, Brovkin V, Carpenter SR, Dakos V, Held H,
-  van Nes EH, Rietkerk M, Sugihara G, 2009 {Early-warning signals for critical
-  transitions}.
-\newblock \emph{Nature} \textbf{461}, 53--59
-
-\bibitem{Lade:2012eu}
-Lade SJ, Gross T, 2012 {Early warning signals for critical transitions: A
-  generalized modeling approach}.
-\newblock \emph{PLoS Comp. Biol.} \textbf{8}, e1002360
-
-\bibitem{Anonymous:2013br}
-Boettiger C, Ross N, Hastings A, 2013 {Early warning signals: the charted and
-  uncharted territories}.
-\newblock \emph{Theor. Ecol.} \textbf{6}, 255--264
-
-\bibitem{Dakos:2014br}
-Dakos V, Bascompte J, 2014 {Critical slowing down as early warning for the
-  onset of collapse in mutualistic communities}.
-\newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{111}, 17546--17551
-
-\bibitem{Krkosek:2014ch}
-Krko{\v s}ek M, Drake JM, 2014 {On signals of phase transitions in salmon
-  population dynamics}.
-\newblock \emph{Proc. Roy. Soc. B} \textbf{281}, 20133221
-
-\bibitem{Kuehn:2010p2591}
-Kuehn C, 2011 {A mathematical framework for critical transitions: bifurcations,
-  fast-slow systems and stochastic dynamics}.
-\newblock \emph{Physica D} \textbf{240}, 1020--1035
-
-\bibitem{Donohue:2013iu}
-Donohue I, Petchey OL, Montoya JM, Jackson AL, McNally L, Viana M, Healy K,
-  Lurgi M, O'Connor NE, Emmerson MC, 2013 {On the dimensionality of ecological
-  stability.}
-\newblock \emph{Ecol. Lett.} \textbf{16}, 421--429
-
-\bibitem{Levin:1992p1472}
-Levin SA, 1992 {The problem of pattern and scale in ecology}.
-\newblock \emph{Ecology} \textbf{73}, 1943--1967
-
-\bibitem{Meinhardt:2012hw}
-Meinhardt H, 2012 {Turing's theory of morphogenesis of 1952 and the subsequent
-  discovery of the crucial role of local self-enhancement and long-range
-  inhibition}.
-\newblock \emph{Interface Focus} \textbf{2}, 407--416
-
-\bibitem{Hanski:1999vu}
-Hanski I, 1999 \emph{{Metapopulation Ecology}}.
-\newblock Oxford: Oxford University Press
-
-\bibitem{Boughton:1999fa}
-Boughton DA, 1999 {Empirical evidence for complex source{\textendash}sink
-  dynamics with alternative states in a butterfly metapopulation}.
-\newblock \emph{Ecology} \textbf{80}, 2727--2739
-
-\bibitem{Secor:2009ena}
-Secor DH, Kerr LA, Cadrin SX, 2009 {Connectivity effects on productivity,
-  stability, and persistence in a herring metapopulation model}.
-\newblock \emph{ICES J Mar Sci} \textbf{66}, 1726--1732
-
-\bibitem{Christie:2012bj}
-Christie MR, Marine ML, French RA, Blouin MS, 2012 {Genetic adaptation to
-  captivity can occur in a single generation.}
-\newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{109}, 238--242
-
-\bibitem{Chilcote:2003bb}
-Chilcote MW, 2003 {Relationship between natural productivity and the frequency
-  of wild fish in mixed spawning populations of wild and hatchery steelhead
-  (\emph{Oncorhynchus mykiss})}.
-\newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{60}, 1057--1067
-
-\bibitem{Araki:2007cm}
-Araki H, Cooper B, Blouin MS, 2007 {Genetic effects of captive breeding cause a
-  rapid, cumulative fitness decline in the wild}.
-\newblock \emph{Science} \textbf{318}, 100--103
-
-\bibitem{Westley:2012ui}
-Westley PAH, { Corinne M Conway}, Fleming IA, 2012 {Phenotypic divergence of
-  exotic fish populations is shaped by spatial proximity and habitat
-  differences across an invaded landscape}.
-\newblock \emph{Evol. Ecol. Res.} \textbf{14}, 147--167
-
-\bibitem{Anderson:2013bf}
-Anderson JH, Faulds PL, Atlas WI, Quinn TP, 2013 {Reproductive success of
-  captively bred and naturally spawned Chinook salmon colonizing newly
-  accessible habitat}.
-\newblock \emph{Evol Appl} \textbf{6}, 165--179
-
-\bibitem{Pess:2014isa}
-Pess GR, Quinn TP, Gephard SR, Saunders R, 2014 {Re-colonization of Atlantic
-  and Pacific rivers by anadromous fishes: linkages between life history and
-  the benefits of barrier removal}.
-\newblock \emph{Rev Fish Biol Fisher} \textbf{24}, 881--900
-
-\bibitem{Ford:2002ip}
-Ford MJ, 2002 {Selection in captivity during supportive breeding may reduce
-  fitness in the wild}.
-\newblock \emph{Conserv Biol} \textbf{16}, 815--825
-
-\bibitem{Liermann:2017gj}
-Liermann M, PESS G, McHenry M, McMillan J, Elofson M, Bennett T, Moses R, 2017
-  {Relocation and recolonization of Coho salmon \emph{Oncorhynchus kisutch} in
-  two tributaries to the elwha river: Implications for management and
-  monitoring}.
-\newblock \emph{T Am Fish Soc} \textbf{17}, 1--39
-
-\end{thebibliography}
+\bibliography{aa_kevin}
+% 
+% \begin{thebibliography}{10}
+% \expandafter\ifx\csname urlstyle\endcsname\relax
+%   \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
+%   \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
+%   \urlstyle{rm}\Url}\fi
+% 
+% \bibitem{Brown:1977gk}
+% Brown JH, Kodric-Brown A, 1977 {Turnover rates in insular biogeography: effect
+%   of immigration on extinction}.
+% \newblock \emph{Ecology} \textbf{58}, 445--449
+% 
+% \bibitem{Earn:2000fm}
+% Earn DJD, Levin SA, Rohani P, 2000 {Coherence and conservation}.
+% \newblock \emph{Science} \textbf{290}, 1360--1364
+% 
+% \bibitem{Anonymous:2015gf}
+% Schindler DE, Armstrong JB, Reed TE, 2015 {The portfolio concept in ecology and
+%   evolution}.
+% \newblock \emph{Front. Ecol. Environ.} \textbf{13}, 257--263
+% 
+% \bibitem{Abbott:2017hl}
+% Abbott RE, Doak DF, Peterson ML, 2017 {Portfolio effects, climate change, and
+%   the persistence of small populations: analyses on the rare plant Saussurea
+%   weberi.}
+% \newblock \emph{Ecology} \textbf{98}, 1071--1081
+% 
+% \bibitem{Thibaut:2012km}
+% Thibaut LM, Connolly SR, 2013 {Understanding diversity-stability relationships:
+%   towards a unified model of portfolio effects.}
+% \newblock \emph{Ecol. Lett.} \textbf{16}, 140--150
+% 
+% \bibitem{MilnerGulland:2011vm}
+% Milner-Gulland EJ, Fryxell JM, Sinclair ARE, 2011 \emph{{Animal Migration}}.
+% \newblock Oxford: Oxford University Press
+% 
+% \bibitem{Bell:2011ki}
+% Bell G, Gonzalez A, 2011 {Adaptation and evolutionary rescue in metapopulations
+%   experiencing environmental deterioration}.
+% \newblock \emph{Science} \textbf{332}, 1327--1330
+% 
+% \bibitem{Carlson:2014is}
+% Carlson SM, Cunningham CJ, Westley PAH, 2014 {Evolutionary rescue in a changing
+%   world}.
+% \newblock \emph{Trends Ecol. Evol.} \textbf{29}, 521--530
+% 
+% \bibitem{Muhlfeld:2014hs}
+% Muhlfeld CC, Kovach RP, Jones LA, Al-Chokhachy R, Boyer MC, Leary RF, Lowe WH,
+%   Luikart G, Allendorf FW, 2014 {Invasive hybridization in a threatened species
+%   is accelerated by climate change}.
+% \newblock \emph{Nature Climate Change} \textbf{4}, 620--624
+% 
+% \bibitem{Lowe:2015ft}
+% Lowe WH, Muhlfeld CC, Allendorf FW, 2015 {Spatial sorting promotes the spread
+%   of maladaptive hybridization}.
+% \newblock \emph{Trends Ecol. Evol.} \textbf{30}, 456--462
+% 
+% \bibitem{H:2013fs}
+% Westley PAH, Quinn TP, 2013 {Rates of straying by hatchery-produced Pacific
+%   salmon (\emph{Oncorhynchus} spp.) and steelhead (\emph{Oncorhynchus mykiss})
+%   differ among species, life history types, and populations}.
+% \newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{70}, 735--746
+% 
+% \bibitem{Keefer:2014gg}
+% Keefer ML, Caudill CC, 2014 {Homing and straying by anadromous salmonids: a
+%   review of mechanisms and rates}.
+% \newblock \emph{Rev Fish Biol Fisher} \textbf{24}, 333--368
+% 
+% \bibitem{Bett:2017ha}
+% Bett NN, Hinch SG, Burnett NJ, Donaldson MR, Naman SM, 2017 {Causes and
+%   consequences of straying into small populations of Pacific salmon}.
+% \newblock \emph{Fisheries} \textbf{42}, 220--230
+% 
+% \bibitem{Berdahl:2015kv}
+% Berdahl A, Torney CJ, Schertzer E, Levin SA, 2015 {On the evolutionary
+%   interplay between dispersal and local adaptation in heterogeneous
+%   environments}.
+% \newblock \emph{Evolution} \textbf{69}, 1390--1405
+% 
+% \bibitem{Berdahl:2016dx}
+% Berdahl A, 2016 {Collective behavior as a driver of critical transitions in
+%   migratory populations}.
+% \newblock \emph{Movement Ecology} \textbf{4}, 1--12
+% 
+% \bibitem{Simons:2004jo}
+% Simons AM, 2004 {Many wrongs: The advantage of group navigation}.
+% \newblock \emph{Trends Ecol. Evol.} \textbf{19}, 453--455
+% 
+% \bibitem{Brenner:2012gl}
+% Brenner RE, Moffitt SD, Grant WS, 2012 {Straying of hatchery salmon in Prince
+%   William Sound, Alaska}.
+% \newblock \emph{Environ Biol Fish} \textbf{94}, 179--195
+% 
+% \bibitem{Johnson:2012fe}
+% Johnson RC, Weber PK, Wikert JD, Workman ML, MacFarlane RB, Grove MJ, Schmitt
+%   AK, 2012 {Managed metapopulations: do salmon hatchery
+%   {\textquoteleft}sources{\textquoteright} lead to in-river
+%   {\textquoteleft}sinks{\textquoteright} in conservation?}
+% \newblock \emph{PLoS ONE} \textbf{7}, e28880--11
+% 
+% \bibitem{Fullerton:2011ii}
+% Fullerton AH, Lindley ST, Pess GR, Feist BE, Steel EA, Mcelhany P, 2011 {Human
+%   influence on the spatial structure of threatened Pacific salmon
+%   metapopulations}.
+% \newblock \emph{Conserv Biol} \textbf{25}, 932--944
+% 
+% \bibitem{Quinn:2011tf}
+% Quinn TP, 2011 \emph{{The Behavior and Ecology of Pacific Salmon and Trout}}.
+% \newblock Vancouver: UBC Press
+% 
+% \bibitem{Jonsson:2011kg}
+% Jonsson B, Jonsson N, 2011 \emph{{Ecology of Atlantic Salmon and Brown Trout}}.
+% \newblock Dordrecht: Springer Netherlands
+% 
+% \bibitem{Quinn:1993ge}
+% Quinn TP, 1993 {A review of homing and straying of wild and hatchery-produced
+%   salmon}.
+% \newblock \emph{Fish Res} \textbf{18}, 29--44
+% 
+% \bibitem{Hendry:2004wf}
+% Hendry AP, Bohlin T, Jonsson B, Berg OK, 2004 {The evolution of philopatry and
+%   dispersal: homing versus straying in salmonids}.
+% \newblock In AP~Hendry, SC~Stearns, eds., \emph{Evolution Illuminated}. Oxford:
+%   Oxford University Press on Demand
+% 
+% \bibitem{Schtickzelle:2007wb}
+% Schtickzelle N, Quinn TP, 2007 {A metapopulation perspective for salmon and
+%   other anadromous fish}.
+% \newblock \emph{Fish Fish.} \textbf{8}, 297--314
+% 
+% \bibitem{Anderson:2014cx}
+% Anderson SC, Moore JW, McClure MM, Dulvy NK, Cooper AB, 2015 {Portfolio
+%   conservation of metapopulations under climate change}.
+% \newblock \emph{Ecol. Appl.} \textbf{25}, 559--572
+% 
+% \bibitem{Keefer:2008bs}
+% Keefer ML, Caudill CC, Peery CA, Lee SR, 2008 {Transporting juvenile salmonids
+%   around dams impairs adult migration}.
+% \newblock \emph{Ecol. Appl.} \textbf{18}, 1888--1900
+% 
+% \bibitem{Westley:2015to}
+% Westley PAH, Dittman AH, Ward EJ, Quinn TP, 2015 {Signals of climate,
+%   conspecific density, and watershed features in patterns of homing and
+%   dispersal by Pacific salmon.}
+% \newblock \emph{Ecology} \textbf{96}, 2823--2833
+% 
+% \bibitem{Bond:2016dz}
+% Bond MH, Westley PAH, Dittman AH, Holecek D, Marsh T, Quinn TP, 2016 {Combined
+%   effects of barge transportation, river environment, and rearing location on
+%   straying and migration of adult snake river fall-run Chinook salmon}.
+% \newblock \emph{T Am Fish Soc} \textbf{146}, 60--73
+% 
+% \bibitem{Peterson:2014gy}
+% Peterson DA, Hilborn R, Hauser L, 2014 {Local adaptation limits lifetime
+%   reproductive success of dispersers in a wild salmon metapopulation}.
+% \newblock \emph{Nat Commun} \textbf{5}, 3696
+% 
+% \bibitem{Moore:2010gs}
+% Moore JW, McClure M, Rogers LA, Schindler DE, 2010 {Synchronization and
+%   portfolio performance of threatened salmon}.
+% \newblock \emph{Conserv. Lett.} \textbf{3}, 340--348
+% 
+% \bibitem{Carlson:2011ce}
+% Carlson SM, Satterthwaite WH, Fleming IA, 2011 {Weakened portfolio effect in a
+%   collapsed salmon population complex}.
+% \newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{68}, 1579--1589
+% 
+% \bibitem{Braun:2016ib}
+% Braun DC, Moore JW, Candy J, Bailey RE, 2016 {Population diversity in salmon:
+%   linkages among response, genetic and life history diversity}.
+% \newblock \emph{Ecography} \textbf{39}, 317--328
+% 
+% \bibitem{simpson1953major}
+% Simpson GG, 1953 \emph{{The Major Features of Evolution}}.
+% \newblock New York: Simon and Schuster
+% 
+% \bibitem{Lande:1976ga}
+% Lande R, 1976 {Natural selection and random genetic drift in phenotypic
+%   evolution}.
+% \newblock \emph{Evolution} \textbf{30}, 314--334
+% 
+% \bibitem{Shelton:2011eq}
+% Shelton AO, Mangel M, 2011 {Fluctuations of fish populations and the magnifying
+%   effects of fishing.}
+% \newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{108}, 7075--7080
+% 
+% \bibitem{Jasper:2013cc}
+% Jasper JR, Habicht C, Moffitt S, Brenner R, Marsh J, Lewis B, Fox EC, Grauvogel
+%   Z, Olive SDR, Grant WS, 2013 {Source-sink estimates of genetic introgression
+%   show influence of hatchery strays on wild chum salmon populations in Prince
+%   William Sound, Alaska}.
+% \newblock \emph{PLoS ONE} \textbf{8}, e81916
+% 
+% \bibitem{Schreiber:2011wx}
+% Schreiber SJ, B{\"u}rger R, Bolnick DI, 2011 {The community effects of
+%   phenotypic and genetic variation within a predator population.}
+% \newblock \emph{Ecology} \textbf{92}, 1582--1593
+% 
+% \bibitem{Gilbert:2014ee}
+% Gibert JP, Brassil CE, 2014 {Individual phenotypic variation reduces
+%   interaction strengths in a consumer{\textendash}resource system}.
+% \newblock \emph{Ecol Evol} \textbf{4}, 3703--3713
+% 
+% \bibitem{Gibert:2015kc}
+% Gibert JP, DeLong JP, 2015 {Individual variation decreases interference
+%   competition but increases species persistence}.
+% \newblock \emph{Adv Ecol Res} \textbf{52}, 45--64
+% 
+% \bibitem{Berdahl:2014bl}
+% Berdahl A, Westley PAH, Levin SA, Couzin ID, Quinn TP, 2014 {A collective
+%   navigation hypothesis for homeward migration in anadromous salmonids}.
+% \newblock \emph{Fish Fish.} \textbf{17}, 525--542
+% 
+% \bibitem{Candy:2000hu}
+% Candy JR, Beacham TD, 2000 {Patterns of homing and straying in southern British
+%   Columbia coded-wire tagged Chinook salmon (\emph{Oncorhynchus tshawytscha})
+%   populations}.
+% \newblock \emph{Fish Res} \textbf{47}, 41--56
+% 
+% \bibitem{JPE:JPE1383}
+% Schick RS, Lindley ST, 2007 {Directed connectivity among fish populations in a
+%   riverine network}.
+% \newblock \emph{J. Appl. Ecol.} \textbf{44}, 1116--1126
+% 
+% \bibitem{Schindler:2015gf}
+% Schindler DE, Armstrong JB, Reed TE, 2015 {The portfolio concept in ecology and
+%   evolution}.
+% \newblock \emph{Front. Ecol. Environ.} \textbf{13}, 257--263
+% 
+% \bibitem{Ovaskainen:2002il}
+% Ovaskainen O, Hanski I, 2002 {Transient dynamics in metapopulation response to
+%   perturbation}.
+% \newblock \emph{Theor Popul Biol} \textbf{61}, 285--295
+% 
+% \bibitem{Anderson:2013gb}
+% Anderson SC, Cooper AB, Dulvy NK, 2013 {Ecological prophets: quantifying
+%   metapopulation portfolio effects}.
+% \newblock \emph{Methods Ecol Evol} \textbf{4}, 971--981
+% 
+% \bibitem{Loreau:2008ju}
+% Loreau M, de~Mazancourt C, 2008 {Species synchrony and its drivers: neutral and
+%   nonneutral community dynamics in fluctuating environments}.
+% \newblock \emph{Am. Nat.} \textbf{172}, E48--E66
+% 
+% \bibitem{Yeakel:2013vz}
+% Yeakel JD, Moore JW, Guimar{\~a}es~Jr PR, de~Aguiar MAM, 2014 {Synchronisation
+%   and stability in river metapopulation networks}.
+% \newblock \emph{Ecol. Lett.} \textbf{17}, 273--283
+% 
+% \bibitem{GuckHolmes}
+% Guckenheimer J, Holmes P, 1983 \emph{{Nonlinear Oscillations, Dynamical
+%   Systems, and Bifurcations of Vector Fields}}.
+% \newblock New York: Springer
+% 
+% \bibitem{Carlson:2008hl}
+% Carlson SM, Seamons TR, 2008 {A review of quantitative genetic components of
+%   fitness in salmonids: implications for adaptation to future change}.
+% \newblock \emph{Evol Appl} \textbf{1}, 222--238
+% 
+% \bibitem{Scheffer:2009gg}
+% Scheffer M, Bascompte J, Brock WA, Brovkin V, Carpenter SR, Dakos V, Held H,
+%   van Nes EH, Rietkerk M, Sugihara G, 2009 {Early-warning signals for critical
+%   transitions}.
+% \newblock \emph{Nature} \textbf{461}, 53--59
+% 
+% \bibitem{Lade:2012eu}
+% Lade SJ, Gross T, 2012 {Early warning signals for critical transitions: A
+%   generalized modeling approach}.
+% \newblock \emph{PLoS Comp. Biol.} \textbf{8}, e1002360
+% 
+% \bibitem{Anonymous:2013br}
+% Boettiger C, Ross N, Hastings A, 2013 {Early warning signals: the charted and
+%   uncharted territories}.
+% \newblock \emph{Theor. Ecol.} \textbf{6}, 255--264
+% 
+% \bibitem{Dakos:2014br}
+% Dakos V, Bascompte J, 2014 {Critical slowing down as early warning for the
+%   onset of collapse in mutualistic communities}.
+% \newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{111}, 17546--17551
+% 
+% \bibitem{Krkosek:2014ch}
+% Krko{\v s}ek M, Drake JM, 2014 {On signals of phase transitions in salmon
+%   population dynamics}.
+% \newblock \emph{Proc. Roy. Soc. B} \textbf{281}, 20133221
+% 
+% \bibitem{Kuehn:2010p2591}
+% Kuehn C, 2011 {A mathematical framework for critical transitions: bifurcations,
+%   fast-slow systems and stochastic dynamics}.
+% \newblock \emph{Physica D} \textbf{240}, 1020--1035
+% 
+% \bibitem{Donohue:2013iu}
+% Donohue I, Petchey OL, Montoya JM, Jackson AL, McNally L, Viana M, Healy K,
+%   Lurgi M, O'Connor NE, Emmerson MC, 2013 {On the dimensionality of ecological
+%   stability.}
+% \newblock \emph{Ecol. Lett.} \textbf{16}, 421--429
+% 
+% \bibitem{Levin:1992p1472}
+% Levin SA, 1992 {The problem of pattern and scale in ecology}.
+% \newblock \emph{Ecology} \textbf{73}, 1943--1967
+% 
+% \bibitem{Meinhardt:2012hw}
+% Meinhardt H, 2012 {Turing's theory of morphogenesis of 1952 and the subsequent
+%   discovery of the crucial role of local self-enhancement and long-range
+%   inhibition}.
+% \newblock \emph{Interface Focus} \textbf{2}, 407--416
+% 
+% \bibitem{Hanski:1999vu}
+% Hanski I, 1999 \emph{{Metapopulation Ecology}}.
+% \newblock Oxford: Oxford University Press
+% 
+% \bibitem{Boughton:1999fa}
+% Boughton DA, 1999 {Empirical evidence for complex source{\textendash}sink
+%   dynamics with alternative states in a butterfly metapopulation}.
+% \newblock \emph{Ecology} \textbf{80}, 2727--2739
+% 
+% \bibitem{Secor:2009ena}
+% Secor DH, Kerr LA, Cadrin SX, 2009 {Connectivity effects on productivity,
+%   stability, and persistence in a herring metapopulation model}.
+% \newblock \emph{ICES J Mar Sci} \textbf{66}, 1726--1732
+% 
+% \bibitem{Christie:2012bj}
+% Christie MR, Marine ML, French RA, Blouin MS, 2012 {Genetic adaptation to
+%   captivity can occur in a single generation.}
+% \newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{109}, 238--242
+% 
+% \bibitem{Chilcote:2003bb}
+% Chilcote MW, 2003 {Relationship between natural productivity and the frequency
+%   of wild fish in mixed spawning populations of wild and hatchery steelhead
+%   (\emph{Oncorhynchus mykiss})}.
+% \newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{60}, 1057--1067
+% 
+% \bibitem{Araki:2007cm}
+% Araki H, Cooper B, Blouin MS, 2007 {Genetic effects of captive breeding cause a
+%   rapid, cumulative fitness decline in the wild}.
+% \newblock \emph{Science} \textbf{318}, 100--103
+% 
+% \bibitem{Westley:2012ui}
+% Westley PAH, { Corinne M Conway}, Fleming IA, 2012 {Phenotypic divergence of
+%   exotic fish populations is shaped by spatial proximity and habitat
+%   differences across an invaded landscape}.
+% \newblock \emph{Evol. Ecol. Res.} \textbf{14}, 147--167
+% 
+% \bibitem{Anderson:2013bf}
+% Anderson JH, Faulds PL, Atlas WI, Quinn TP, 2013 {Reproductive success of
+%   captively bred and naturally spawned Chinook salmon colonizing newly
+%   accessible habitat}.
+% \newblock \emph{Evol Appl} \textbf{6}, 165--179
+% 
+% \bibitem{Pess:2014isa}
+% Pess GR, Quinn TP, Gephard SR, Saunders R, 2014 {Re-colonization of Atlantic
+%   and Pacific rivers by anadromous fishes: linkages between life history and
+%   the benefits of barrier removal}.
+% \newblock \emph{Rev Fish Biol Fisher} \textbf{24}, 881--900
+% 
+% \bibitem{Ford:2002ip}
+% Ford MJ, 2002 {Selection in captivity during supportive breeding may reduce
+%   fitness in the wild}.
+% \newblock \emph{Conserv Biol} \textbf{16}, 815--825
+% 
+% \bibitem{Liermann:2017gj}
+% Liermann M, PESS G, McHenry M, McMillan J, Elofson M, Bennett T, Moses R, 2017
+%   {Relocation and recolonization of Coho salmon \emph{Oncorhynchus kisutch} in
+%   two tributaries to the elwha river: Implications for management and
+%   monitoring}.
+% \newblock \emph{T Am Fish Soc} \textbf{17}, 1--39
+% 
+% \end{thebibliography}
+
+\clearpage
+
+% \begin{table}[!t]
+\begin{center}
+\begin{tabular}{ l|l }
+\hline
+Parameter & Definition \\
+\hline
+$N_i,N_T$ & Individual, aggregate population density\\
+$\mu_i$ & Mean trait value of population $i$\\
+$m$ & Straying rate\\
+$m_0$ & Straying rate of an individual\\
+$R_i[\mu]$ & Recruitment rate of population $i$\\
+$r_{\rm max}$ & Maximum recruitment rate\\
+${\rm e}^{-Z}$ & Survival rate\\
+$\beta$ & Strength of density dependence\\
+$\theta_i$ & Optimal trait value at habitat $i$\\
+$\Delta\theta$ & Difference in habitat optima\\
+$\tau$ & Strength of selection\\
+$\sigma^2$ & Genetic variance $i$\\
+$h^2$ & Heritability\\
+$C$ & Half saturation constant for density dependent straying\\
+${\rm PE}$ & Portfolio Effect\\
+\hline
+\end{tabular}
+\end{center}
+% \end{table}
 
@@
 (b) The steady state trait values measured as $\theta_i - x_i$, as a function of a constant stray rate $m$. 
-FB marks the fold bifurcation.
+PFB marks the pitchfork bifurcation.
+Unless otherwise indicated, the default parameter values used are: $r_{\rm max}=2$; $Z=0.5$; $\beta=0.001$; $\theta_1=5$; $\Delta\theta=5$; $\tau=1$; $\sigma=1$; $T=1\times10^5$.
 } \label{fig:traj}
 \end{figure}
+% tmax=10000;
+% z=0.5;
+% rmax=2.0;
+% beta=0.001;
+% theta1=5.0;
+% thetadiff=5.0;
+% tau=1.0;
+% h=0.49;
+% sigmaE=0;
+% sigmaG=1.0;
+% m=0.35;
+% perror=0.01;
 
@@
 (c) the portfolio effect PE as a function of heritability $h^2$ and a constant stray rate $m$. Light colors = high values.
-The black line shows the fold bifurcation separating a single steady state (left) from alternative stable states (right).
+The black line shows the pitchfork bifurcation separating a single steady state (left) from alternative stable states (right).
 (d) The relationship between the time to recovery following a disturbance and the portfolio effect. Portfolio effects greater than unity corresponds to less synchronization.
-The black line denotes the fold bifurcation.
+The black line denotes the pitchfork bifurcation.
 } \label{fig:PE}
@@
 Portfolio effects greater than unity corresponds to less synchronization.
-FB marks the fold bifurcation.
+PFB marks the pitchfork bifurcation.
 } \label{fig:thetaPE}
@@
 If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
-FB marks the fold bifurcation.
+PFB marks the pitchfork bifurcation.
 } \label{fig:relax}
@@
 Phenotypic diversity ($\Delta \mu^*$) evaluated at the steady state as a function of straying rate $m$ and trait heritability $h^2$.
-The descrete jump occurs as the system crosses the fold bifurcation; lower phenoytic diversity emerges with higher straying rates and in the alternative stable state regime.
+The descrete jump occurs as the system crosses the pitchfork bifurcation; lower phenoytic diversity emerges with higher straying rates and in the alternative stable state regime.
 } \label{fig:traitdiff}
@@
   The `$*$' marks the value of $(m,\Delta\theta)$ below which there is a switch in subordinate/dominant states following extinction of the dominant population.
-  FB marks the fold bifurcation.
+  PFB marks the pitchfork bifurcation.
   } \label{fig:mtheta}
@@
 
+%%%%%%%%%%%%%
+% SUPPLEMENTARY FIGURES
+%%%%%%%%%%%%%
+
 \beginsupplement
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+% \includegraphics[width=0.35\textwidth]{fig_recovery.pdf}
+\caption{
+Example of the numerical procedure used to estimate recovery time.
+After a disturbance is introduced, the recovery time is calculated by measuring the point in time where $N_T$ (in black), which is the aggregate of both populations (blue, red) settles to within one standard deviation of the new equilibrium $N_T^*$.
+} \label{fig:recovery}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+% \includegraphics[width=0.35\textwidth]{fig_mthetarelation.pdf}
+\caption{
+In some cases, habitat heterogeneity may be assumed to determine the rate of straying, if for example:
+1) sites may be distributed over greater spatial distances, where habitat differences are assumed to be greater between more distant sites, or 2) individuals have behaviors promoting dispersal between habitats that are more similar.
+To examine such cases, we use the relationship $m= 0.5(1 + \Delta\theta)^{-1}$ where maximum straying is assumed to occur at $m=0.5$ (perfect mixing).
+} \label{fig:mthetarelation}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+% \includegraphics[width=0.35\textwidth]{fig_mthetarelation.pdf}
+\caption{
+The real parts of the four eigenvalues for the Jacobian matrix of the 4-dimensional system.
+The pitchfork bifurcation occurs when the dominant eigenvalue crosses the unit circle at $+1$. 
+} \label{fig:eigs}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+% \includegraphics[width=0.35\textwidth]{fig_mthetarelation.pdf}
+\caption{
+Increasing the straying rate results in the transition from a single steady state for both populations to a dominant and subordinate states.
+If the straying rate is subsequently lowered, the single steady state is not easily obtained, which is the hallmark of hysteresis.
+} \label{fig:hysteresis}
+\end{figure}
 
@@
 
-\begin{figure*}
+\begin{figure}
   \captionsetup{justification=raggedright,
@@
 } \label{fig:relax_highh}
-\end{figure*}
+\end{figure}
+
+\clearpage
 
@@
 } \label{fig:relaxtheta}
-\end{figure*}
+\end{figure}
 
@@
 \end{figure}
+
+% \clearpage
 
@@
 When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases for constant (a) and density-dependent (b) staying rates.
-  The fold bifurcation is not as clear in (b) because $\Delta\theta$ is a function of the individual straying rate $m_0$, whereas the x-axis in (b) is the straying rate at the steady state $m^*$.
+  The pitchfork bifurcation is not as clear in (b) because $\Delta\theta$ is a function of the individual straying rate $m_0$, whereas the x-axis in (b) is the straying rate at the steady state $m^*$.
   Despite this difference, the general trends shown in (a) are also present in (b).
   } \label{fig:mthetamvm}
-\end{figure*}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+% \includegraphics[width=0.35\textwidth]{fig_asymdensity.pdf}
+\caption{
+Steady state densities of both populations as a function of $m$, where a pitchfork bifurcation indicates the emergence of alternative steady states: one in a dominant state and one in a subordinate state.
+Steady states for populations with symmetrical values ($\alpha=0$) in the vital rates $r_{\rm max}$ and $\beta$ are shown with cool tones.
+As the asymmetry among populations between sites increases ($\alpha>0$), their vital rates diverge, such that the maximal growth at sites 1 and 2 is now $r_{\rm max}(1)=r_{\rm max}(1+\tilde{rv}_1)$ and $r_{\rm max}(2)=r_{\rm max}(1+\tilde{rv}_2)$ where $rv_{1,2}$ are independently drawn from $\rm{Normal}(0,\alpha)$ and $r_{\rm max}=2$.
+Similarly the strength of density dependence is calculated at sites 1 and 2 as $\beta(1)=\beta(1+\tilde{rv}_1)$ and $\beta(2)=\beta(1+\tilde{rv}_2)$ where $\tilde{rv}_{1,2}$ are independently drawn from $\rm{Normal}(0,\alpha)$ and $\beta=0.001$.
+Steady states for populations with increasingly asymmetric values ($\alpha>0$) in the vital rates $r_{\rm max}$ and $\beta$ are shown in cool tones.
+} \label{fig:symmetry}
+\end{figure}
 

### 2017-09-12 Manuscript/FinalDraft_rev/supp.tex
edits

@@
 \usepackage{xr}
-\externaldocument[extfig-]{msv5}
+\externaldocument[extfig-]{msv6}
 %\usepackage{epstopdf}
@@
 \beginsupplement
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.35\textwidth]{fig_recovery.pdf}
+\caption{
+Example of the numerical procedure used to estimate recovery time.
+After a disturbance is introduced, the recovery time is calculated by measuring the point in time where $N_T$ (in black), which is the aggregate of both populations (blue, red) settles to within one standard deviation of the new equilibrium $N_T^*$.
+} \label{fig:recovery}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.35\textwidth]{fig_mthetarelation.pdf}
+\caption{
+In some cases, habitat heterogeneity may be assumed to determine the rate of straying, if for example:
+1) sites may be distributed over greater spatial distances, where habitat differences are assumed to be greater between more distant sites, or 2) individuals have behaviors promoting dispersal between habitats that are more similar.
+To examine such cases, we use the relationship $m= 0.5(1 + \Delta\theta)^{-1}$ where maximum straying is assumed to occur at $m=0.5$ (perfect mixing).
+} \label{fig:mthetarelation}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.35\textwidth]{fig_eigs.pdf}
+\caption{
+The real parts of the four eigenvalues for the Jacobian matrix of the 4-dimensional system.
+The pitchfork bifurcation occurs when the dominant eigenvalue crosses the unit circle at $+1$. 
+} \label{fig:eigs}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.35\textwidth]{fig_hysteresis.pdf}
+\caption{
+Increasing the straying rate results in the transition from a single steady state for both populations to a dominant and subordinate states.
+If the straying rate is subsequently lowered, the single steady state is not easily obtained, which is the hallmark of hysteresis.
+} \label{fig:hysteresis}
+\end{figure}
 
@@
 When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases for constant (a) and density-dependent (b) staying rates.
-  The fold bifurcation is not as clear in (b) because $\Delta\theta$ is a function of the individual straying rate $m_0$, whereas the x-axis in (b) is the straying rate at the steady state $m^*$.
+  The pitchfork bifurcation is not as clear in (b) because $\Delta\theta$ is a function of the individual straying rate $m_0$, whereas the x-axis in (b) is the straying rate at the steady state $m^*$.
   Despite this difference, the general trends shown in (a) are also present in (b).
@@
 \end{figure*}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.35\textwidth]{fig_asymdensity.pdf}
+\caption{
+Steady state densities of both populations as a function of $m$, where a pitchfork bifurcation indicates the emergence of alternative steady states: one in a dominant state and one in a subordinate state.
+Steady states for populations with symmetrical values ($\alpha=0$) in the vital rates $r_{\rm max}$ and $\beta$ are shown with cool tones.
+As the asymmetry among populations between sites increases ($\alpha>0$), their vital rates diverge, such that the maximal growth at sites 1 and 2 is now $r_{\rm max,1}=r_{\rm max}(1+\tilde{rv}_1)$ and $r_{\rm max,2}=r_{\rm max}(1+\tilde{rv}_2)$ where $rv_{1,2}$ are independently drawn from $\rm{Normal}(0,\alpha)$ and $r_{\rm max}=2$.
+Similarly the strength of density dependence is calculated at sites 1 and 2 as $\beta_1=\beta(1+\tilde{rv}_1)$ and $\beta_2=\beta(1+\tilde{rv}_2)$ where $\tilde{rv}_{1,2}$ are independently drawn from $\rm{Normal}(0,\alpha)$ and $\beta=0.001$.
+Steady states for populations with increasingly asymmetric values ($\alpha>0$) in the vital rates $r_{\rm max}$ and $\beta$ are shown in warm tones.
+} \label{fig:symmetry}
+\end{figure}
 

### 2017-09-13 Manuscript/FinalDraft_rev/ReviewResponses.tex
updated figures

@@
 The similarities in the resulting dynamics suggest that what we are observing in the two-population salmon system, as well as the implications that we derive for metapopulation robustness and recovery, may be generally relevant and illustrative of a general phenomenon outside of the constraints introduced in the specific formulation of our model.
+Though as Reviewer I notes, our model while complementary to Ronce and Kirkpatrick 2001, focuses instead on the metapopulation effects (portfolio effects, resilience, and overall robustness) of the observed dynamics rather than a detailed analytical treatment of the dynamics themselves.
+We highlight areas of complementarity and extension in the revised manuscript.
 
@@
 The Fold bifurcation is generally defined by the dominant eigenvalue $\lambda$ intersecting the unit circle in Real-Imaginary space at ${\rm Re}[\lambda]=+1$.
-We now include a new Supplemental Figure (SXX) that shows this is indeed the case for our system, and was the basis on which we initially identified the bifurcation.
+We now include a new figure S3 that shows this is indeed the case for our system, and was the basis on which we initially identified the bifurcation.
 Although we are not specialists in bifurcation theory, our understanding is that the fold bifurcation is a general attribute based on this eigenvalue behavior, but as the Reviewers indicate, it is not a particularly clear or complete description of what is going on.
@@
 As the Reviewers suggest, we investigated whether or not our system displays hysteresis, as does the system in the Ronce and Kirkpatrick model.
-We now show that it does (Supplementary Figure SXX), a feature characteristic of Cusp bifurcations in general.
+We now show that it does (figure S4), a feature characteristic of Cusp bifurcations in general.
 This is more evidence that the dynamics described here and in Ronce and Kirkpatrick are perhaps general phenomena of spatially-linked eco-evolutionary systems.
@@
 We emphasize that our measure of recovery time is a numerical measurement, as eigenvalue-based (asymptotic) measures are not appropriate for the scale of disturbance that we introduce (extinction in some cases - far from the non-trivial fixed point(s) of the system).
-The procedure is now described in more detail within the text, and illustrated graphically in an additional supplemental figure (Supplementary Figure SXX).
+The procedure is now described in more detail within the text, and illustrated graphically in an additional supplemental figure (figure S1).
 In short, we track the recovering aggregate population $N_T=N_1+N_2$ until it is within a standard deviation around the \emph{updated} fixed point (thus accounting for fixed points that are altered by the disturbance).
@@
 
-To investigate the effects of increasing asymmetry on the main dynamical feature of our model, we assessed how increasing $\alpha$ altered the steady state portrait as a function of the straying rate $m$ shown in figure 1a, and this analysis is now included as a supplementary figure (figure Sxx).
+To investigate the effects of increasing asymmetry on the main dynamical feature of our model, we assessed how increasing $\alpha$ altered the steady state portrait as a function of the straying rate $m$ shown in figure 1a, and this analysis is now included as a supplementary figure (figure S13).
 As can be seen, greater asymmetry (warm colors in the supplementary figure) does not alter the qualitative nature of the dynamics, and does not significantly change the position of the bifurcation.
@@
 \noindent {\bf Response 2.3} We thank the Reviewer, and now briefly describe our analysis of the bifurcation (also see Response 1.3).
-We include a plot of the Jacobian eigenvalues across $m$ (figure SXX).
-Moreover, we note that hysteresis is observed at the bifurcation (evaluated by increasing and decreasing $m$), and this is shown in figure SXX.
+We include a plot of the Jacobian eigenvalues across $m$ (figure S3).
+Moreover, we note that hysteresis is observed at the bifurcation (evaluated by increasing and decreasing $m$), and this is shown in figure S4.
 

### 2017-09-13 Manuscript/FinalDraft_rev/msv6.tex
updated figures

@@
 On the other hand, high rates of dispersal may synchronize populations and actually increase the risk of extinction of the entire metapopulation \citep{Earn:2000fm}. 
-Third, straying will influence the evolutionary dynamics of the metapopulation.
+Third, dispersal will influence the evolutionary dynamics of the metapopulation.
 % though how the interplay between selection and population dynamics might impact persistence is less well understood. 
@@
 More broadly, dispersal can provide a mechanism by which phenotypes are sorted in space rather than time and facilitates the spread of maladaptive genes \citep{Lowe:2015ft}.
-Further, this straying may lead to genetic homogenization that erodes the asynchrony that underpins portfolio effects and metapopulation persistence. 
-Thus, straying can influence the resilience and robustness of metapopulations through both ecological and evolutionary processes.
+Dispersal in this case may lead to genetic homogenization that erodes the asynchrony that underpins portfolio effects and metapopulation persistence. 
+% Thus, straying can influence the resilience and robustness of metapopulations through both ecological and evolutionary processes.
 % Thus, the dual nature of straying as both promoter of connections among metapopulation demes and potential eroder of locally adapted gene complexes highlights the interplay between ecological dynamics of connected populations and the evolutionary dynamics of mixed trait distributions that respond to heterogeneous local conditions.
@@
 
-\noindent where $C$ is a half-saturation constant.
+\noindent where $C$ is a half-saturation constant and is set to $C=1000$ throughout.
 We note that at the limit $C\rightarrow \infty$, the density-dependent straying rate becomes constant such that $m(t) \rightarrow m_0$, and this corresponds to the original formulation where $m=m_0$.
@@
 \hline
-$N_i,N_T$ & Individual, aggregate population density\\
-$\mu_i$ & Mean trait value of population $i$\\
-$m$ & Straying rate\\
+$N_i(t)$, $N_T(t)$ & Individual, aggregate population over time\\
+$\mu_i(t)$ & Mean trait value of population $i$ over time\\
+$m$, $m(t)$ & Constant, density dependent straying rate\\
 $m_0$ & Straying rate of an individual\\
-$R_i[\mu]$ & Recruitment rate of population $i$\\
+$R_i[\mu(t)]$ & Recruitment rate of population $i$\\
 $r_{\rm max}$ & Maximum recruitment rate\\
@@
 $\beta$ & Strength of density dependence\\
-$\theta_i$ & Optimal trait value at habitat $i$\\
-$\Delta\theta$ & Difference in habitat optima\\
+$\theta_i$ & Optimal trait value for habitat $i$\\
+$\Delta\theta$ & Habitat heterogeneity\\
 $\tau$ & Strength of selection\\
-$\sigma^2$ & Genetic variance $i$\\
+$\sigma^2$ & Genetic variance\\
 $h^2$ & Heritability\\
-$C$ & Half saturation constant for density dependent straying\\
+$C$ & Half saturation constant for $m(t)$\\
 ${\rm PE}$ & Portfolio Effect\\
@@
 Similarly the strength of density dependence is calculated at sites 1 and 2 as $\beta(1)=\beta(1+\tilde{rv}_1)$ and $\beta(2)=\beta(1+\tilde{rv}_2)$ where $\tilde{rv}_{1,2}$ are independently drawn from $\rm{Normal}(0,\alpha)$ and $\beta=0.001$.
-Steady states for populations with increasingly asymmetric values ($\alpha>0$) in the vital rates $r_{\rm max}$ and $\beta$ are shown in cool tones.
+Steady states for populations with increasingly asymmetric values ($\alpha\rightarrow 0.1$) for vital rates $r_{\rm max}$ and $\beta$ are shown in warmer tones.
 } \label{fig:symmetry}

### 2017-09-13 Manuscript/FinalDraft_rev/supp.tex
updated figures

@@
 Similarly the strength of density dependence is calculated at sites 1 and 2 as $\beta_1=\beta(1+\tilde{rv}_1)$ and $\beta_2=\beta(1+\tilde{rv}_2)$ where $\tilde{rv}_{1,2}$ are independently drawn from $\rm{Normal}(0,\alpha)$ and $\beta=0.001$.
-Steady states for populations with increasingly asymmetric values ($\alpha>0$) in the vital rates $r_{\rm max}$ and $\beta$ are shown in warm tones.
+Steady states for populations with increasingly asymmetric values ($\alpha\rightarrow 0.1$) for vital rates $r_{\rm max}$ and $\beta$ are shown in warmer tones.
 } \label{fig:symmetry}

### 2017-09-13 model/KevinExtinction.jl

@@ -388,7 +388,7 @@ dev.off()
 """
 
 
-namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/figs2/fig_relax_lowh.pdf");
+namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/FinalDraft_rev/fig_relax_lowh.pdf");
 R"""
 library(RColorBrewer)
 pdf($namespace,height=4,width=10)
@@ -400,7 +400,7 @@ lines($mvec,$(ma_rth2[:,2]),col=palsub[2],cex=0.5,lwd=2)
 lines($mvec,$(ma_rth2[:,3]),col=palsub[3],cex=0.5,lwd=2)
 text(par('usr')[1]-0.09,max($(ma_rth2)),'(a)', xpd=TRUE)
 arrows($(mvec[indmax(ma_rth2[1:100,1])]),10,$(mvec[indmax(ma_rth2[1:100,1])]),12,length=0.05,angle=40,lwd=3)
-text($(mvec[indmax(ma_rth2[1:100,1])]),9,'FB')
+text($(mvec[indmax(ma_rth2[1:100,1])]),9,'PFB')
 
 
 plot($(ma1_m2_ddm[:,1]),$(ma_rth2_ddm[:,1]),col=palsub[1],log='y',cex=0.5,pch=16,xlab='m*',ylab='Recovery time',ylim=c(min($rt_ddm),max($(ma_rth2_ddm))),las=1)
@@ -419,7 +419,7 @@ text(-0.072,max($(ma_rth2_ddm)),'(b)', xpd=TRUE)
 types = c('subordinate extinct','dominant extinct','near-collapse')
 legend(x=0.18,y=130,legend=types,col=palsub,pch=22,xpd=TRUE,pt.bg=palsub,cex=1, bty="n") #,title=expression(paste(Delta,theta))
 arrows($(ma1_m2_ddm[indmax(ma_rth2_ddm[1:100,3]),3]),10,$(ma1_m2_ddm[indmax(ma_rth2_ddm[1:100,3]),3]),12,length=0.05,angle=40,lwd=3)
-text($(ma1_m2_ddm[indmax(ma_rth2_ddm[1:100,3]),3]),9,'FB')
+text($(ma1_m2_ddm[indmax(ma_rth2_ddm[1:100,3]),3]),9,'PFB')
 
 
 dev.off()

### 2017-09-13 model/KevinSimulations2.jl

@@ -61,7 +61,7 @@ eigvals(Jac)
 
 #Analysis over m
 tmax=10000;
-mvec = collect(0.0:0.001:0.5);
+mvec = collect(0.0:0.0001:0.5);
 n1ts = zeros(Float64,length(mvec),tmax);
 n2ts = zeros(Float64,length(mvec),tmax);
 n1mean=zeros(Float64,length(mvec));
@@ -154,7 +154,7 @@ points($mvec,$n2mean,pch='.',col=cols[2],cex=0.5)
 
 
 #Steady state plot
-namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/figs2/fig_traj.pdf");
+namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/FinalDraft_rev/fig_traj.pdf");
 R"""
 library(RColorBrewer)
 cols = brewer.pal(4,'Set1')
@@ -163,12 +163,12 @@ par(mfrow=c(2,1),mai = c(0.8, 0.8, 0.1, 0.1))
 plot($mvec,$n1mean,pch='.',col=cols[4],xlab="Straying rate m",ylab="Steady state biomass",cex=0.5,ylim=c(0,max($n1mean)),las=1)
 points($mvec,$n2mean,pch='.',col=cols[4],cex=0.5)
 arrows($(mvec[indmax(pe)]),1200,$(mvec[indmax(pe)]),1100,length=0.05,angle=40,lwd=3)
-text($(mvec[indmax(pe)]),1275,'FB')
+text($(mvec[indmax(pe)]),1275,'PFB')
 text(par('usr')[1]-0.09,1300,'(a)', xpd=TRUE)
 plot($mvec,$x1mean,pch=".",col=cols[1],ylim=c(-5,5),xlab="Straying rate m",ylab="Trait offset",las=1)
 points($mvec,$x2mean,pch=".",col=cols[2])
 arrows($(mvec[indmax(pe)]),3.4,$(mvec[indmax(pe)]),2.5,length=0.05,angle=40,lwd=3)
-text($(mvec[indmax(pe)]),4,'FB')
+text($(mvec[indmax(pe)]),4,'PFB')
 text(par('usr')[1]-0.09,5,'(b)', xpd=TRUE)
 dev.off()
 """

### 2017-09-13 model/KevinSimulations_mtheta2.jl

@@ -474,7 +474,7 @@ dev.off()
 #  segments($(ma1_m_ddm[:,3])[i],$(ma_rth_ddm[:,3])[i],$(ma2_m_ddm[:,3])[i],$(ma_rth_ddm[:,3])[i],col=palsub[3])
 # }
 
-namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/Manuscript/figs2/fig_mtheta_rt.pdf");
+namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/Manuscript/FinalDraft_rev/fig_mtheta_rt.pdf");
 R"""
 library(RColorBrewer)
 library(shape)
@@ -488,7 +488,7 @@ lines($mvec2,$(ma_rth[:,3]),col=palsub[3],cex=0.5,lwd=2)
 arrows(0.122,2800,0.122,1800,length=0.05,angle=40,lwd=3)
 types = c('subordinate extinct','dominant extinct','near-collapse')
 legend(x=0.14,y=8000,legend=types,col=palsub,pch=22,xpd=TRUE,pt.bg=palsub,cex=0.9, bty="n") #,title=expression(paste(Delta,theta))
-text(0.122,4000,'FB')
+text(0.122,4000,'PFB')
 text(0.034,90,'*',cex=2)
 dev.off()
 """

### 2017-09-13 Manuscript/FinalDraft_rev/ReviewResponses.tex
some more edits to rev

@@
 \end{align}
-Because the third term on the right is the derivative of the log of the sum (as it must be for the discrete time version of the Lande equation; Lande \emph{Evolution}, 1976), the derivative of the second term within the logarithm is not zero, and both phenotypic means contribute to the strength and direction of selection, such that
+Because the third term on the right is the derivative of the log of the sum (as it must be for the discrete time version of the Lande equation; Lande \emph{Evolution}, 1976), the derivative of the second term within the logarithm is not zero, and both phenotypic means contribute to selection, such that
 \begin{equation}
@@
 \end{equation}
+
+\noindent We have made the following edits to the text:\\
+\noindent {\bf Line XXX:} \emph{The partial derivative in the Eq.
+4 determines how the mean trait changes through time due to natural selection [ref], which is proportional to the change in mean fitness with respect to $\mu_i$.
+We note that the derivative is dependent on both $\mu_i$ and $\mu_j$ due to the influence of the logarithm on the sum.}
 

### 2017-09-13 Manuscript/FinalDraft_rev/msv6.tex
some more edits to rev

@@
 On the other hand, high rates of dispersal may synchronize populations and actually increase the risk of extinction of the entire metapopulation \citep{Earn:2000fm}. 
-Third, dispersal will influence the evolutionary dynamics of the metapopulation.
+Dispersal will also influence the evolutionary dynamics of the metapopulation.
 % though how the interplay between selection and population dynamics might impact persistence is less well understood. 
@@
 More broadly, dispersal can provide a mechanism by which phenotypes are sorted in space rather than time and facilitates the spread of maladaptive genes \citep{Lowe:2015ft}.
-Dispersal in this case may lead to genetic homogenization that erodes the asynchrony that underpins portfolio effects and metapopulation persistence. 
+Dispersal in this case may lead to genetic homogenization that erodes the asynchrony underpinning portfolio effects and metapopulation persistence. 
 % Thus, straying can influence the resilience and robustness of metapopulations through both ecological and evolutionary processes.
@@
 %Motivation: straying
-The eco-evolutionary impacts of dispersal likely has important implications for conservation and management in key taxa such as in migratory salmon.
+The eco-evolutionary impacts of dispersal likely have important implications for conservation and management in key taxa such as in migratory salmon.
 While anadromous salmonid fishes (genera \emph{Oncorhynchus} and \emph{Salmo}) are renown for returning to their natal spawning habitats with high accuracy and precision after years at sea \citep{Quinn:2011tf,Jonsson:2011kg,Keefer:2014gg}, there are generally some individuals that `stray' to non-natal sites to spawn \citep{Quinn:1993ge,Hendry:2004wf}.
 Salmon may operate as metapopulations, where populations are genetically distinct but linked by some level of straying \citep{Schtickzelle:2007wb,Anderson:2014cx}.
-Although extensive work has been done to document the extent of straying from donor populations and into recipient populations \citep{Keefer:2014gg,Bett:2017ha}, only recently have the abiotic, biotic, and anthropogenic influences of `straying' behaviors been investigated systemically \citep{Keefer:2008bs,Westley:2015to,Bond:2016dz}.
+Although extensive work has been done to document the extent of straying from donor populations and into recipient populations \citep{Keefer:2014gg,Bett:2017ha}, only recently have the abiotic, biotic, and anthropogenic influences of straying behaviors been investigated systemically \citep{Keefer:2008bs,Westley:2015to,Bond:2016dz}.
 Straying among salmon may be influenced by environmental factors such as water temperature, human activities such as hatchery practices, and population density as predicted by the collective migration hypothesis \citep{Peterson:2014gy}.
-These strays can introduce new maladaptive genotypes into the recipient population.
-Further, straying and genetic homogenization could synchronize population dynamics and erode portfolio effects \citep{Moore:2010gs,Carlson:2011ce,Braun:2016ib}.
-Thus, there is opportunity and a need to consider the eco-evolutionary consequences of straying for metapopulations in species of conservation and management concern such as salmon.
+Straying can introduce new maladaptive genotypes into the recipient population, while the ensuing genetic homogenization could synchronize population dynamics and erode portfolio effects \citep{Moore:2010gs,Carlson:2011ce,Braun:2016ib}.
+Thus, there is an opportunity and need to consider the eco-evolutionary consequences of straying for metapopulations in species of conservation and management concern such as salmon.
 
@@
 % Here we ask the overarching question: how does collective behavior mediated dispersal and gene flow interact to influence the robustness of locally adapted populations?
-To address this question we construct a minimal eco-evolutionary model of two populations occupying different sites that are linked by straying individuals, each of which with an associated trait distribution subject to natural selection determined by local conditions.
+To address this question we construct a minimal eco-evolutionary model of two populations occupying different sites that are linked by straying individuals, each with an associated trait distribution subject to natural selection determined by local conditions.
 Specifically we compare (a) different rates of straying and (b) the influence of collective movement, across (c) increasing environmental heterogeneity, by assessing two measures of metapopulation robustness: the portfolio effect and the time required for a population(s) to recover following an induced disturbance.
-This model enables us to explore the multiple and potentially opposing pathways by which straying influences metapopulation robustness such as the potentially detrimental erosion of local adaptation vs.
+% This model enables us to explore the multiple and potentially opposing pathways by which straying influences metapopulation robustness such as the potentially detrimental erosion of local adaptation vs.
 the positive effects of demographic and evolutionary rescue.
+This model enables us to explore the tradeoff between the potentially detrimental erosion of local adaptation vs.
+the positive effects of demographic and evolutionary rescue, both of which are facilitated by straying.
+%such as the potentially detrimental erosion of local adaptation vs.
+the positive effects of demographic and evolutionary rescue.
+
 % [We show that taking straying into account leads to alternative stable states in population densities and trait values, which has consequences for maladaptation, intraspecific trait variability, and long-term sustainability of salmon metapopulations.]
@@
 Moreover, we assume that $x_{1,2}$ are normally distributed with means $\mu_1$ and $\mu_2$ and have the same standard deviation $\sigma$.
-As such, the recruitment rate for both populations is determined by the mean trait value of the local population, such that $r_1 = R_1[\mu_1(t),\theta_1]$.
-Trait means for each population are subject to selection, the strength of which depends on the difference between the trait mean and the local trait optimum at a given point in time \citep{simpson1953major,Lande:1976ga}.
+As such, the recruitment rate $R_1[\mu_1(t),\theta_1]$ for both populations is determined by the mean trait value of the local population relative to optimal value at that site.
+Trait means for each population are subject to selection, the strength of which is proportional to the difference between the trait mean and the local trait optimum at a given point in time \citep{simpson1953major,Lande:1976ga}.
 
@@
 Again, straying can be thought of as both from the donor or recipient population perspective.
-The two populations occur in spatially separate sites that are close enough such that a proportion of the population $m$ can stray into the other site, and where mortality occurs before individuals return to reproduce.
+The two populations occur in spatially separate sites that are close enough that a proportion of the population $m$ can stray into the other site, and where mortality occurs before individuals return to reproduce.
 If there is no straying between these populations, then the mean trait evolves towards the optimal value for each site $x_1 \rightarrow \theta_1$, and the recruitment rate for each population is maximized.
@@
 
-We use discrete Ricker framework described by Shelton and Mangel \citep{Shelton:2011eq} as the basis for our two-site metapopulation model, with the added effect of the local population $N_i$ mixing with a set proportion $m$ of a remote population $N_j$ that is straying into it.
+We use the discrete Ricker framework described by Shelton and Mangel \citep{Shelton:2011eq} as the basis for our two-site metapopulation model, with the added effect of the local population $N_i$ mixing with a set proportion $m$ of a remote population $N_j$ that is straying into it.
 In this sense, both populations serve simultaneously as donor and recipient populations.
-We first assume that the proportion ${\rm e}^{-Z}$ of both populations survive such that the surviving aggregated population, composed of both local individuals (at site $i$) and incoming strays (from site $j$), is $\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z}$.
+We first assume that a proportion ${\rm e}^{-Z}$ of both populations survive such that the surviving aggregated population, composed of both local individuals (at site $i$) and incoming strays (from site $j$), is $\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z}$.
 Because local individuals will recruit differently than incoming strays, the recruitment of the aggregate must incorporate two recruitment functions, given by $\left(R_i[\mu_i(t)] (1-m)N_i(t) + R_i[\mu_j(t)] m N_j(t)\right)$.
-This mix of individuals is subject to the same compensatory effects, which is determined by the parameter $\beta$.
-Taken together, the difference equation that determine changes in population size is
+This mix of individuals is subject to identical compensatory effects, which is determined by the parameter $\beta$.
+Taken together, the difference equation that determines changes in population size is
 
@@
 
-\noindent where the difference equation for $N_j$ mirrors that for $N_i$.
+\noindent where the equation for $N_j$ mirrors that for $N_i$.
 
@@
   &R_i[\mu_i(t)] = \\ \nonumber
-  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i(t)-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i(t),\mu_i,\sigma^2) {\rm d}x_i(t) +\tilde{P}\\ \nonumber
-  &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-\mu_i(t))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\} +\tilde{P},
+  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i(t)-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i(t),\mu_i,\sigma^2) {\rm d}x_i(t) +\tilde{P}_i\\ \nonumber
+  &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-\mu_i(t))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\} +\tilde{P}_i,
   \label{eq:R}
@@
 
-\noindent where the mismatch between the local trait mean $\mu_i(t)$ and the local optimum $\theta_i$ scales the recruitment rate for the population, and $\tilde{P}\sim {\rm Normal}(0,0.01)$ introduces a small amount of demographic error.
+\noindent where the mismatch between the local trait mean $\mu_i(t)$ and the local optimum $\theta_i$ scales the recruitment rate for the population, and $\tilde{P}_i\sim {\rm Normal}(0,0.01)$ introduces a small amount of demographic error.
 The parameter $\tau$ is the strength of selection, and controls the sensitivity of recruitment to changes in the mean trait value away from the optimum (the strength of selection increases with smaller values of $\tau$), which we set as $\tau=1$ here and throughout.
 Because straying individuals are emigrating from a population with a mean trait value farther from the local optimum, their rate of recruitment is diminished.
-Recent studies of wild sockeye salmon have indeed found that straying individuals have lower life-time fitness than individuals that do not stray, although it unknown at what life-stage this selection occurs \citep{Peterson:2014gy}.
+Recent studies of wild sockeye salmon have indeed found that straying individuals have lower life-time fitness than individuals that do not stray, although it is unknown at what life-stage this selection occurs \citep{Peterson:2014gy}.
 % The compensatory effects are then determined by the exponential, following the Ricker stock-recruitment relationship.
@@
 \ref{eq:mu} determines how the mean trait changes through time due to natural selection \citep{Lande:1976ga}, which is proportional to the change in mean fitness with respect to $\mu_i$.
+We note that the derivative is dependent on both $\mu_i$ and $\mu_j$ due to the influence of the logarithm on the sum.
 %Comparison to the Ronce and Kirkpatrick paper
@@
 The largest difference between these approaches is that our framework treats trait evolution mechanistically at some cost to analytical tractability.
-Importantly, we show that the resulting dynamics are qualitatively similar, suggesting that the dynamical features present in both of these approaches have potentially widespread ramifications for the evolutionary dynamics of spatially connected populations.
+Importantly, we show that the resulting dynamics are qualitatively similar, suggesting that the dynamical features present in both of these approaches have potentially widespread ramifications for the evolutionary dynamics of spatially connected populations, regardless of the specific formulation proposed.
 \\
@@
 %Density dependent m
-\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is mounting evidence that at least in some species (including salmon) the straying rate is density-dependent with a signature of collective navigation \citep{Berdahl:2016dx,Bett:2017ha}.
+\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is mounting evidence that at least in some species (including salmon) the straying rate is density-dependent, a signature of collective navigation \citep{Berdahl:2016dx,Bett:2017ha}.
 Specifically, the rate at which individuals stray has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tend to decrease the rate at which individuals err, reducing the overall proportion of a population that strays.
@@
 \noindent where $C$ is a half-saturation constant and is set to $C=1000$ throughout.
-We note that at the limit $C\rightarrow \infty$, the density-dependent straying rate becomes constant such that $m(t) \rightarrow m_0$, and this corresponds to the original formulation where $m=m_0$.
-A similar observation shows that when the population density is very high, $m(t) \rightarrow 0$, and when it is small, individuals operate without regard to collective behavior, meaning $m(t) \rightarrow m_0$.
-Thus, for realistic population densities, $m(t) < m_0$.\\
+% We note that at the limit $C\rightarrow \infty$, the density-dependent straying rate becomes constant such that $m(t) \rightarrow m_0$, and this corresponds to the original formulation where $m=m_0$.
+When the population density is very high, $m(t) \rightarrow 0$, and when the population is small, individuals operate without regard to collective behavior, such that $m(t) \rightarrow m_0$.
+For realistic population densities, $m(t) < m_0$.\\
 
 \noindent {\bf (c) Habitat heterogeneity}
-\noindent Increasing differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater regional differences in the conditions that favor alternative trait complexes, which we interpret here as increased habitat heterogeneity.
+\noindent Increasing differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater regional differences in the conditions that favor alternative trait complexes, which can be interpreted as increased habitat heterogeneity.
 If both populations are isolated, natural selection will direct the mean trait values of both populations towards their respective optima, such that $\mu_i(t) \rightarrow \theta_i$ as $t\rightarrow\infty$.
@@
 %Linking m/m0 and thetadiff
-Although we largely treat habitat heterogeneity and the rate of straying as independent parameters, we evaluate a case where we assume that increased habitat heterogeneity correlates with lower straying rates, and vice versa (illustrated in figure \ref{fig:mthetarelation}).
+Habitat heterogeneity and the rate of straying are treated both independently, and as parameters that covary.
+In the latter instance, we evaluate a case where it is assumed that increased habitat heterogeneity correlates with lower straying rates, and vice versa (illustrated in figure \ref{fig:mthetarelation}).
 Two scenarios may lead to this correlation: 
-(\emph{i}) sites may be distributed over greater spatial distances, where habitat differences are assumed to be exaggerated and the likelihood of straying over greater distances would be lower \citep{Candy:2000hu,JPE:JPE1383};
+(\emph{i}) sites may be distributed over greater spatial distances, where habitat differences are assumed to be exaggerated and the likelihood of straying over greater distances is lower \citep{Candy:2000hu,JPE:JPE1383};
 (\emph{ii}) individuals may have behaviors promoting dispersal between habitats with structural or physiognomic similarities \citep{Peterson:2014gy}.
 In this case, the rate of straying would be greater between habitats with smaller differences in trait optima (lower $\Delta\theta$) and lesser between habitats with greater differences in trait optima (higher $\Delta\theta$).
-
+\\
 % Where indicated, we account for correlation between the straying rate and the difference between optimal trait values by assuming that $m$ (if the stray rate is constant) or $m_0$ (if the stray rate is density-dependent) is a function of the difference between optimal trait values between sites $\Delta\theta$.
@@
 \noindent We evaluated metapopulation robustness by measuring the average-CV portfolio effect (PE) \citep{Anderson:2014cx,Schindler:2015gf} as well as the time required for the system to return to a steady state following an induced disturbance to one or both of the populations \citep{Ovaskainen:2002il}.
-The average-CV portfolio effect is, as the name implies, the average CV across each population divided by the CV of the aggregate \citep{Anderson:2013gb}, such that
+The average-CV portfolio effect is, as the name implies, the average CV across each population $N_i$ divided by the CV of the aggregate $N_T=\sum_i N_i$ \citep{Anderson:2013gb}, such that
 
@@
 The steady state condition is denoted by `$*$'.
-As the CV of $N_T^*$ decreases relative to that of the constituent populations, $\langle{\rm PE}\rangle > 1$, and the metapopulation is presumed to become more stable.
-Portfolio effects greater than unity corresponds to less synchronization  \citep{Loreau:2008ju,Anderson:2014cx,Yeakel:2013vz} and thus a greater potential for demographic rescue among populations, buffering the system as a whole against extinction.
+As the CV of $N_T^*$ decreases relative to that of the constituent populations, $\langle{\rm PE}\rangle > 1$, and the metapopulation is presumed to be more stable because the aggregate has functioned to dampen population-level variance.
+Moreover, portfolio effects greater than unity correspond to less synchronization  \citep{Loreau:2008ju,Anderson:2014cx,Yeakel:2013vz} and thus a greater potential for demographic rescue among populations, buffering the system as a whole against extinction.
 
 %Although there are more robust ways to measure the portfolio effect (in particular for comparing the PE across species with different mean-variance relationships; REF: Anderson), its only utlity here is to provide a simple statistical measure of metapopulation persistence.
-A more direct way to measure system robustness is to measure the time that the system (measured as the aggregate steady state biomass $N_T^*$) takes to recover its steady state abundance following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
+A more direct way to measure system robustness is to measure the time that the system (measured as the aggregate steady state biomass $N_T^*$) takes to return to a steady state following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
 Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation \citep{GuckHolmes}, because we aimed to 1) assess the effects of a large perturbation, and 2) estimate the time required for all transient effects to decay (including dampened oscillations), we used a simulation-based numerical procedure.
@@
 We measured recovery time following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} are equivalent if populations have the same density); (\emph{iii}) near-collapse of both populations where just 1.0\% of each survives.
-Throughout, we will refer to an increase in the portfolio effects and/or reduction in recovery times as promoting metapopulation robustness, which promotes persistence.
+Throughout, we will refer to an increase in the portfolio effects and/or reduction in recovery times as promoting metapopulation robustness and by extension persistence.
 \\
@@
 The decline in steady state densities is not gradual: as straying increases, the system crosses a pitchfork bifurcation (PFB) \citep{AleksandrovichKuznetsov:1995p2580} whereby the single steady state for the metapopulation bifurcates into two basins of attraction: one at high biomass, and one at low biomass density (figure \ref{fig:traj}a, \ref{fig:PE}a).
-In discrete systems, a pitchfork bifurcation is a specific case of the cusp bifurcation (which occurs when two fold bifurcations intersect at a cusp), which occurs when the real part of the dominant eigenvalue of the Jacobian matrix crosses the unit circle at +1, which we show in figure \ref{fig:eigs}.
-
-Mean trait values for both populations bifurcate similarly (figure \ref{fig:traj}b), depending on which population attains a low- vs. high-density. 
-Above the threshold straying rate defined by the PFB, there are two alternative eco-evolutionary states: the \emph{dominant state} population will have a higher density and higher degree of local adaptation (smaller trait offset from the local optimum), while the \emph{subordinate state} population will have lower density with maladapted trait values (larger trait offset from the local optimum).
-Hysteresis is observed to occur at this transition, such that the single steady state regime cannot be easily recovered by lowering $m$ after the system attains alternative steady states (figure \ref{fig:hysteresis}).
+In discrete systems, a pitchfork bifurcation is a specific case of the cusp bifurcation (where two fold bifurcations intersect at a cusp) \citep{AleksandrovichKuznetsov:1995p2580}, and occurs when the real part of the dominant eigenvalue of the Jacobian matrix crosses the unit circle at +1 (figure \ref{fig:eigs}).
+Mean trait values for both populations bifurcate similarly (figure \ref{fig:traj}b).
+
+Above the threshold straying rate defined by the PFB, there are two alternative eco-evolutionary states: the \emph{dominant state} population will have a higher density and greater degree of local adaptation (smaller trait offset from the local optimum), while the \emph{subordinate state} population will have lower density with maladapted trait values (larger trait offset from the local optimum).
+Hysteresis is observed to occur at this transition, such that the single steady state regime cannot easily be recovered by reducing straying after the system attains alternative steady states (figure \ref{fig:hysteresis}).
 These dynamics are also observed in the Ronce and Kirkpatrick model, where populations are described as transitioning between symmetric to asymmetric states \citep{Ronce:2001dp}.
@@
 
-Trait heritability has a large impact on the degree to which straying affects both the aggregate population steady state density ($N^*_T=N^*_1+N^*_2$; figure \ref{fig:PE}a) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N^*=|N^*_1-N^*_2|$; figure \ref{fig:PE}b).
-Greater trait heritability results in a faster decline in $N_T^*$ with increasing straying rates $m$, but leads to only moderate changes to $\Delta N^*$.
-Conversely, in the context of lower trait heritability, an increase in the straying rate has little impact on the total biomass density but contrastingly large effects on $\Delta N^*$.
+Trait heritability $h^2$ has a large effect on the degree to which straying affects both the aggregate population steady state density ($N^*_T$; figure \ref{fig:PE}a) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N^*=|N^*_1-N^*_2|$; figure \ref{fig:PE}b).
+Greater trait heritability results in a steeper decline in $N_T^*$ with increasing straying rates $m$, but leads to only moderate changes to $\Delta N^*$.
+Lower trait heritability has little effect on the total biomass density but contrastingly large effects on $\Delta N^*$.
 The pitchfork bifurcation (the black line in Figs.
-\ref{fig:PE}a-c) occurs at lower values of the straying rate $m$ with decreased trait heritability $h^2$ (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
-Although trait heritability among salmonids is variable, most life history traits have an $h^2 <0.5$ \citep{Carlson:2008hl}, and we largely focus additional analyses on that range.
+\ref{fig:PE}a-c) occurs at lower values of the straying rate $m$ with lower heritability (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
+Although trait heritability among salmonids is variable, most life history traits have an $h^2 <0.5$ \citep{Carlson:2008hl}, and we largely focus our efforts on that range.
 
@@
 As the pitchfork bifurcation is approached with increasing $m$, the portfolio effect increases sharply due to an amplification in variance within both donor and recipient populations.
-This amplification in variance is the product of a dynamical process known as \emph{critical slowing down} that occurs near pitchfork bifurcations \citep{Scheffer:2009gg}, a phenomenon that some have suggested may serve as an early warning indicator for approaching phase transitions \citep{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br,Krkosek:2014ch}.
+This amplification in variance is the product of a dynamical process known as \emph{critical slowing down} that can occur near bifurcations \citep{Scheffer:2009gg}, a phenomenon that some have suggested may serve as an early warning indicator for approaching phase transitions \citep{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br,Krkosek:2014ch}.
 For larger values of $m$ (to the right of the pitchfork bifurcation in Fig \ref{fig:PE}a-c), where alternative stable states occur, the portfolio effect declines steadily as the CV of $N_T^*$ increases.
@@
 $N_i(t)$, $N_T(t)$ & Individual, aggregate population over time\\
-$\mu_i(t)$ & Mean trait value of population $i$ over time\\
+$x_i$ & Trait value for an individual in population $i$\\
+$\mu_i(t)$ & Mean of $x$ for population $i$ over time\\
+$\sigma^2$ & Genetic variance of trait $x$\\
 $m$, $m(t)$ & Constant, density dependent straying rate\\
@@
 $\tau$ & Strength of selection\\
-$\sigma^2$ & Genetic variance\\
 $h^2$ & Heritability\\
@@
 
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.9\textwidth]{fig_relax_lowh.pdf}
+\caption{
+Recovery time of $N_T^*$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.2$.
+If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
+PFB marks the pitchfork bifurcation.
+} \label{fig:relax}
+\end{figure*}
+
 \begin{figure}
@@
 
-\begin{figure*}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.9\textwidth]{fig_relax_lowh.pdf}
-\caption{
-Recovery time of $N_T^*$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.2$.
-If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
-PFB marks the pitchfork bifurcation.
-} \label{fig:relax}
-\end{figure*}
-
 \begin{figure}
@@
 \centering
+% \includegraphics[width=0.35\textwidth]{fig_mthetarelation.pdf}
+\caption{
+In some cases, habitat heterogeneity may be assumed to determine the rate of straying, if for example:
+1) sites are distributed over greater spatial distances, where habitat differences are assumed to be greater between more distant sites, or 2) individuals have behaviors promoting dispersal between habitats that are more similar.
+To examine such cases, we use the relationship $m= 0.5(1 + \Delta\theta)^{-1}$ where maximum straying is assumed to occur at $m=0.5$ (perfect mixing).
+} \label{fig:mthetarelation}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
 % \includegraphics[width=0.35\textwidth]{fig_recovery.pdf}
@@
 } \label{fig:recovery}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-% \includegraphics[width=0.35\textwidth]{fig_mthetarelation.pdf}
-\caption{
-In some cases, habitat heterogeneity may be assumed to determine the rate of straying, if for example:
-1) sites may be distributed over greater spatial distances, where habitat differences are assumed to be greater between more distant sites, or 2) individuals have behaviors promoting dispersal between habitats that are more similar.
-To examine such cases, we use the relationship $m= 0.5(1 + \Delta\theta)^{-1}$ where maximum straying is assumed to occur at $m=0.5$ (perfect mixing).
-} \label{fig:mthetarelation}
 \end{figure}

### 2017-09-13 Manuscript/FinalDraft_rev/supp.tex
some more edits to rev

@@
 \centering
-\includegraphics[width=0.35\textwidth]{fig_recovery.pdf}
+\includegraphics[width=0.35\textwidth]{fig_mthetarelation.pdf}
 \caption{
-Example of the numerical procedure used to estimate recovery time.
-After a disturbance is introduced, the recovery time is calculated by measuring the point in time where $N_T$ (in black), which is the aggregate of both populations (blue, red) settles to within one standard deviation of the new equilibrium $N_T^*$.
-} \label{fig:recovery}
+In some cases, habitat heterogeneity may be assumed to determine the rate of straying, if for example:
+1) sites are distributed over greater spatial distances, where habitat differences are assumed to be greater between more distant sites, or 2) individuals have behaviors promoting dispersal between habitats that are more similar.
+To examine such cases, we use the relationship $m= 0.5(1 + \Delta\theta)^{-1}$ where maximum straying is assumed to occur at $m=0.5$ (perfect mixing).
+} \label{fig:mthetarelation}
 \end{figure}
@@
 \centering
-\includegraphics[width=0.35\textwidth]{fig_mthetarelation.pdf}
+\includegraphics[width=0.35\textwidth]{fig_recovery.pdf}
 \caption{
-In some cases, habitat heterogeneity may be assumed to determine the rate of straying, if for example:
-1) sites may be distributed over greater spatial distances, where habitat differences are assumed to be greater between more distant sites, or 2) individuals have behaviors promoting dispersal between habitats that are more similar.
-To examine such cases, we use the relationship $m= 0.5(1 + \Delta\theta)^{-1}$ where maximum straying is assumed to occur at $m=0.5$ (perfect mixing).
-} \label{fig:mthetarelation}
+Example of the numerical procedure used to estimate recovery time.
+After a disturbance is introduced, the recovery time is calculated by measuring the point in time where $N_T$ (in black), which is the aggregate of both populations (blue, red) settles to within one standard deviation of the new equilibrium $N_T^*$.
+} \label{fig:recovery}
 \end{figure}

### 2017-09-19 Manuscript/FinalDraft_rev/ReviewResponses.tex
final review response

@@
 
-\noindent {\bf Response 1.1} We appreciate the Reviewer's concern, however note that equation 4) is written as
+\noindent {\bf Response 1.1} We appreciate the Reviewer's concern, however note that equation 4 is written as
 \begin{align}
@@
 
-\noindent We have made the following edits to the text:\\
-\noindent {\bf Line XXX:} \emph{The partial derivative in the Eq.
+\noindent To clarify, we have made the following edits to the text:\\
+\noindent {\bf Line 113:} \emph{The partial derivative in Eq.
 4 determines how the mean trait changes through time due to natural selection [ref], which is proportional to the change in mean fitness with respect to $\mu_i$.
@@
 The model indeed has many similarities to our own, and we now acknowledge its importance and priority within the manuscript.
-The similarities in the resulting dynamics suggest that what we are observing in the two-population salmon system, as well as the implications that we derive for metapopulation robustness and recovery, may be generally relevant and illustrative of a general phenomenon outside of the constraints introduced in the specific formulation of our model.
+The similarities in the resulting dynamics suggest that what we are observing in the two-population salmon system, as well as the implications that we derive for metapopulation robustness and recovery, may be  relevant and illustrative of a general phenomenon outside of the constraints introduced in the specific formulation of our model.
 Though as Reviewer I notes, our model while complementary to Ronce and Kirkpatrick 2001, focuses instead on the metapopulation effects (portfolio effects, resilience, and overall robustness) of the observed dynamics rather than a detailed analytical treatment of the dynamics themselves.
@@
 \noindent We have made the following edits to the text:\\
-\noindent {\bf Line XXX:} \emph{This model formulation has parallels to that proposed by Ronce and Kirkpatrick [ref]), where habitat specialization evolves between two populations as a function of dispersal.
+\noindent {\bf Line 115:} \emph{This model formulation has parallels to that proposed by Ronce and Kirkpatrick [ref]), where habitat specialization evolves between two populations as a function of dispersal.
 The largest difference between these approaches is that our framework treats trait evolution mechanistically (at some cost to analytical tractability).
 Importantly, we show that the resulting dynamics are qualitatively similar, suggesting that the dynamical features present in both of these approaches have potentially widespread ramifications for the evolutionary dynamics of connected populations.}\\
-\noindent {\bf Line XXX} \emph{This dynamic is also observed in the Ronce and Kirkpatrick model, which they describe as a transition from symmetric to asymmetric states [ref].}
+\noindent {\bf Line 191} \emph{These dynamics are also observed in the Ronce and Kirkpatrick model, which they describe as a transition from symmetric to asymmetric states [ref].}
 
@@
 \noindent {\bf Response 1.3} Both Reviewers indicated that our labeling of the Fold Bifurcation was incorrect, and that the identity of the transition of the system from symmetric to asymmetric states was in fact a Pitchfork bifurcation.
-We are happy to modify the identification of the bifurcation as the Reviewers suggest as the Pitchfork label will be more familiar to readers and is somewhat self-evident in terms of the bifurcation diagram that is displayed.
-At the same time, we would like to explain our reasoning, and maintain our position that the bifurcation is technically (and more generally) a fold bifurcation as well.
+To our knowledge, the Pitchfork Bifurcation applies only to continuous time systems, and not to discrete maps.
+However, there is a qualitative (visual) analogue to the Pitchfork in discrete maps, and we think that we can be more precise in our description.
 
-According to Kuznetsov (1998), there are two principle bifurcations in discrete maps, from which others are composed: the Fold and Flip (period-doubling).
+According to Kuznetsov (1998), there are three principle bifurcations in discrete maps, from which others are composed: the Fold, Flip (period-doubling), and Neimark-Sacker.
 The Fold bifurcation is generally defined by the dominant eigenvalue $\lambda$ intersecting the unit circle in Real-Imaginary space at ${\rm Re}[\lambda]=+1$.
-We now include a new figure S3 that shows this is indeed the case for our system, and was the basis on which we initially identified the bifurcation.
-Although we are not specialists in bifurcation theory, our understanding is that the fold bifurcation is a general attribute based on this eigenvalue behavior, but as the Reviewers indicate, it is not a particularly clear or complete description of what is going on.
-A familiar, and more complicated bifurcation that occurs in systems with $\geq2$ parameters is the Cusp bifurcation, which occurs when two fold bifurcations interact at a cusp.
-A one-dimensional transect through the cusp results in what is commonly referred to as a Pitchfork Bifurcation (topological normal form: $x \mapsto x + \beta x - x^3$).
-At the cusp ($\beta=0$), the single fixed point crosses the combined fold bifurcations, and here it branches into 3 fixed points; two stable fixed points separated by an unstable fixed point.
-Identifying the bifurcation as a Pitchfork bifurcation increases clarity and specificity, and we have made the appropriate edits in an effort to do so.
+We now include a new figure S3 that shows that this is indeed the case for our system, and this was the basis on which we initially identified the bifurcation.
+Although we are not mathematicians, our understanding is that the fold bifurcation is a general attribute based on this eigenvalue condition, but as the Reviewers indicate, it is not a particularly clear or complete description of what is going on.
+A familiar, and more complicated bifurcation that occurs in maps with $\geq2$ parameters is the Discrete Cusp Bifurcation, which occurs when two fold bifurcations intersect at a cusp.
+A one-dimensional transect through the cusp results in behavior similar to that observed for Pitchfork Bifurcations in vector fields (topological normal form for the discrete map: $x \mapsto x + \beta x - x^3$).
+At the cusp ($\beta=0$), the single fixed point crosses the intersecting fold bifurcations, and here it branches into 3 fixed points; two stable fixed points separated by an unstable fixed point.
+We now identify the bifurcation as a Discrete Cusp bifurcation and mention that it displays dynamics that are visually similar to Pitchfork bifurcations in continuous time systems, which we hope will increase clarity and specificity.
 
@@
 The eigenvalue evaluated at the steady state $\partial f(x)/\partial x|_{x=x^*}$ increases to unity on the unit circle at the cusp bifurcation $(\beta_1 = 0, \beta_2 = 0)$.
+
+\noindent We have made the following edits to the text:\\
+\noindent {\bf Line 115:} \emph{The decline in steady state densities is not gradual: as straying increases, the system crosses a discrete cusp bifurcation (DCB) [ref] whereby the single steady state for the metapopulation bifurcates into two basins of attraction: one at high biomass, and one at low biomass density (figure 1a, 2a).
+Mean trait values for both populations bifurcate similarly (figure 1b). 
+In discrete systems, the cusp bifurcation is defined by two fold bifurcations intersecting at a cusp point [ref], and is observed when the real part of the dominant eigenvalue of the Jacobian matrix crosses the unit circle at +1 (figure S3).
+Visually, the dynamics are similar to those observed at a pitchfork bifurcation in continuous systems, where a single steady state gives rise to two steady states separated by an unstable fixed point.
+}
 
@@
 \noindent We have made the following edits to the text:\\
-\noindent {\bf Line XXX:} \emph{Recovery time was calculated by initiating a disturbance at $t=t_d$, and monitoring $N_T(t_d+t)$ as $t\rightarrow T$, where $T$ is large. 
-The aggregate was deemed recovered at $t_r$, such that recovery time was calculated as $t_r-t_d$, and recovery at $t=t_r$ was measured as the initial $t$ where $N_T(t) < {\rm SD}\left( N_T^* \right)$ for $t\in(t_r,T)$, where $\rm{SD}(\cdot)$ is standard deviation (illustrated in figure [ref]).}\\
+\noindent {\bf Line 160:} \emph{Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation [ref], because we aimed to 1) assess the effects of a large perturbation, and 2) estimate the time required for all transient effects to decay (including dampened oscillations), we used a simulation-based numerical procedure.
+Recovery time was calculated by initiating a disturbance at $t=t_d$, and monitoring $N_T(t_d+t)$ as $t\rightarrow T$, where $T$ is large. 
+The aggregate was deemed recovered at $t_r$, such that recovery time was calculated as $t_r-t_d$, and recovery at $t=t_r$ was measured as the initial $t$ where $N_T(t) < {\rm SD}\left( N_T^* \right)$ for $t\in(t_r,T)$, where $\rm{SD}(\cdot)$ is standard deviation (illustrated in figure S2)}\\
 
@@
 
-\noindent {\bf Minor Response 1.1} We now equate straying with dispersal early in the text (and note that it is also defined in the abstract).
+\noindent {\bf Minor Response 1.1} We now equate straying with dispersal early in the text (it is now also defined in the abstract).
 
 \noindent We have made the following edits to the text:\\
-\noindent {\bf Line XXX:} \emph{The rate at which individuals stray, $m$, is in this case synonymous with dispersal and may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices [refs].}\\
+\noindent {\bf Line 42:} \emph{The rate at which individuals stray, $m$, is in this case synonymous with dispersal and may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices [refs].}\\
 
@@
 
-\noindent {\bf Minor Response 1.2} We agree and thank the Reviewer for point this out. We have made the following edits to the text:\\
+\noindent {\bf Minor Response 1.2} We agree and thank the Reviewer for pointing this out. We have made the following edits to the text:\\
 \noindent {\bf Figure 1 caption} \emph{Unless otherwise indicated, the default parameter values used are: $r_{\rm max}=2$; $Z=0.5$; $\beta=0.001$; $\theta_1=5$; $\Delta\theta=5$; $\tau=1$; $\sigma=1$; $T=1\times10^5$.}\\
@@
 
-\noindent {\bf Response 2.1} We thank the Reviewer for their comment and have now clarified the manuscript. Please see Response 1.3.
+\noindent {\bf Response 2.1} We thank the Reviewer for their comments and have now clarified the manuscript. Please see Responses 1.3 and 2.2, which address these concerns.
 
@@
 I would be very curious to see this analysis extended under asymmetrical assumptions, for example in the productivity of the different locations or the intraspecific competition between populations.
-Another potentially highly interesting element that has not been studied here is the potential for alternative stable states based on the density-dependent straying rate (subsection b in the Model formulation).
-The study currently addresses a scenario where the potential for population collapse due to collective navigation is ignored.
-I see that as the interesting angle; how do the dynamics caused by density-dependent straying interact with the dynamics caused by selective forces as imposed by the different environments on recruitment?
 \\
@@
 
-\noindent {\bf Response 2} We thank the Reviewer for their perspective and comments, and agree that we need to understand to what extent our results are generalizable.
-From our perspective, there are two ways in which our model incorporates symmetry: 1) the architectural symmetry of the populations between sites and 2) the symmetry in parameter values between sites.
+\noindent {\bf Response 2.2} We thank the Reviewer for their perspective and comments, and agree that we need to understand to what extent our results are generalizable.
+% From our perspective, there are two ways in which our model incorporates symmetry: 1) the architectural symmetry of the populations between sites and 2) the symmetry in parameter values between sites.
 Regarding the first, both populations have the same growth and mortality functions, and both populations have the same functional forms that control straying.
@@
 We cannot alter this type of symmetry without vastly changing the questions that we are addressing.
+In general, although discrete cusp bifurcations with pitchfork-like behaviors are a consequence of symmetry (defined by a system where $N_1\rightarrow-N_1$, $N_2\rightarrow-N_2$, $x_1\rightarrow-x_1$, $x_2\rightarrow-x_2$ without altering the equations), the general features of the bifurcation need not be altered as symmetry is broken.
+Consider the topological normal form $\dot{x}=h + \beta x - x^3$, which is a continuous system with a perfect pitchfork bifurcation at $\beta = 0,~h=0$.
+It is symmetric in the sense that $x\rightarrow-x$ without altering anything if and only if $h=0$.
+If $h\neq0$, the system is asymmetric, however the pitchfork-like dynamic does not disappear but becomes an imperfect pitchfork where a saddle node bifurcation appears alongside a non-varying steady state.
+This creates a parameter region where alternative steady states exist, separated by an unstable steady state.
+As such, symmetry exists only in a perfect sense (and we describe below why we think our system already violates perfect symmetry), and the observed dynamics that we observe need not require perfect symmetry.
+However, we agree with the Reviewer that this begs investigation.
 
-We assume that the Reviewer is referring to the second type of symmetry: that many of the parameters of the model controlling vital rates are symmetric between sites, primary examples being: $r_{\rm max}=2;~Z=0.5;~\beta=0.001$.
+Our model treats many of the parameters of the model controlling vital rates to be the same between sites, primary examples being: $r_{\rm max}=2;~Z=0.5;~\beta=0.001$.
 These parameters were held constant primarily to minimize the parameter space that we investigate (making an already long paper a bit shorter), but we acknowledge that this invites the question: are we investigating a mathematical knife-edge?
-
 We suggest that our analysis is not merely a mathematical construction (interpreted here to mean dynamic behaviors that only occur within a thin slice of parameter space, unlikely to be realized in biological systems) for 3 reasons:
@@
 Asymmetry in parameter values is introduced as a parameter $\alpha$, where maximal growth at sites 1 and 2 are now  $r_{\rm max}(1)=r_{\rm max}(1+\tilde{rv}_1)$ and $r_{\rm max}(2)=r_{\rm max}(1+\tilde{rv}_2)$ where $rv_{1,2}$ are independently drawn from $\rm{Normal}(0,\alpha)$ and $r_{\rm max}=2$.
+% Increasing $\alpha$ results in more widely diverging vital rates between sites.
 Similarly the strength of density dependence is calculated at sites 1 and 2 as $\beta(1)=\beta(1+\tilde{rv}_1)$ and $\beta(2)=\beta(1+\tilde{rv}_2)$ where $\tilde{rv}_{1,2}$ are independently drawn from $\rm{Normal}(0,\alpha)$ and $\beta=0.001$.
-Thus as asymmetry increases, so do the differences in vital rates between sites.
+Thus as asymmetry ($\alpha$) increases, so do the differences in vital rates between sites.
 
@@
 \noindent \textcolor{cyan}{
-{\bf Comment 2.3} It would be good to include more technical details about how the bifurcation analysis was carried out (or is this present in the missing SI text?).
+{\bf Comment 2.3} Another potentially highly interesting element that has not been studied here is the potential for alternative stable states based on the density-dependent straying rate (subsection b in the Model formulation).
+The study currently addresses a scenario where the potential for population collapse due to collective navigation is ignored.
+I see that as the interesting angle; how do the dynamics caused by density-dependent straying interact with the dynamics caused by selective forces as imposed by the different environments on recruitment?
+}
+
+\noindent {\bf Response 2.3} We agree that this is an interesting line of inquiry, but do investigate it in the manuscript.
+Specifically we analyze scenarios where a) individual populations that collectively stray collapse, and b) both populations that collectively stray suffer near-collapse.
+If this is in line with what the Reviewer is describing, we suggest that the uploading issues, which prevented delivery of the Supplemental information (and many of the pertinent figures) made the manuscript somewhat opaque with regard to these analyses.
+We examine the effects of density dependent straying on all aspects of the model, and strive to highlight those results in our revised version.
+
+\noindent \textcolor{cyan}{
+{\bf Comment 2.4} It would be good to include more technical details about how the bifurcation analysis was carried out (or is this present in the missing SI text?).
 As it is, I have no idea about the numerical techniques that were used or even whether the authors have studied both increasing and decreasing parameter values, as necessary to enable detection of alternative stable states and discontinuities.
@@
 
-\noindent {\bf Response 2.3} We thank the Reviewer, and now briefly describe our analysis of the bifurcation (also see Response 1.3).
+\noindent {\bf Response 2.4} We thank the Reviewer, and now briefly describe our analysis of the bifurcation (also see Response 1.3).
 We include a plot of the Jacobian eigenvalues across $m$ (figure S3).
@@
 \noindent \textcolor{cyan}{
-{\bf Comment 2.4} Please include a table with the definitions and values of model parameters (including their units) and definitions of model variables.
+{\bf Comment 2.5} Please include a table with the definitions and values of model parameters (including their units) and definitions of model variables.
 }
 
-\noindent {\bf Response 2.4} A table is now included.
+\noindent {\bf Response 2.5} We thank the Reviewer for their suggestion. A table is now included.
 
 \noindent \textcolor{cyan}{
-{\bf Comment 2.5} I could not understand section c, Habitat heterogeneity in the Model formulation.
+{\bf Comment 2.6} I could not understand section c, Habitat heterogeneity in the Model formulation.
 In particular the paragraph on lines 139-144 should be explained and justified better, where the authors should make sure to explain the ‘integrate the two variables’ part more formally and explicitly.
@@
 
-\noindent {\bf Response 2.5} We thank the Reviewer for their comment and apologize for our lack of clarity.
+\noindent {\bf Response 2.6} We thank the Reviewer for their comment and apologize for our lack of clarity.
 We have moved the text and attempt to explain our reasonings and justifications more clearly.
@@
 \noindent We have made the following edits to the text:\\
-\noindent {\bf Line XXX:} \emph{Although we largely treat habitat heterogeneity and the rate of straying as independent parameters, we evaluate a case where we assume that increased habitat heterogeneity correlates with lower straying rates, and vice versa.
+\noindent {\bf Line 136:} \emph{Habitat heterogeneity and the rate of straying are treated both independently, and as parameters that covary.
+In the latter instance, we evaluate a case where it is assumed that increased habitat heterogeneity correlates with lower straying rates, and vice versa (illustrated in figure S1).
 Two scenarios may lead to this correlation: 
-(\emph{i}) sites may be distributed over greater spatial distances, where habitat differences are assumed to be exaggerated and the likelihood of straying over greater distances would be lower [ref];
+(\emph{i}) sites may be distributed over greater spatial distances, where habitat differences are assumed to be exaggerated and the likelihood of straying over greater distances is lower [ref];
 (\emph{ii}) individuals may have behaviors promoting dispersal between habitats with structural or physiognomic similarities [ref].
@@
 
-\noindent {\bf Line XXX:} \emph{Sites that are farther apart are likely to be more heterogeneous, while simultaneously facilitating lower rates of straying between populations.
-% If we assume that environmental heterogeneity increases with distance between sites, or that individuals may have behaviors that minimize their chances of straying into very different habitats, the rate of straying will decline with habitat heterogeneity.
-% We can implement this dependence by allowing $\Delta\theta$ to decrease as $m$ increases, such that there are low rates of straying between dissimilar populations and high rates of straying between similar populations.
-We implemented this dependence as set $m,m_0 = 0.5(1 + \Delta\theta)^{-1}$ where maximum straying is assumed to occur at $m=0.5$ (perfect mixing; figure [ref]).
-This assumes that $m$ or $m_0$ increases in situations where $\Delta\theta$ is lower, such that there are low rates of straying between dissimilar sites and high straying rates between similar sites.
-We find that under these conditions alternative stable states now appear for very low rates of straying $0 < m \leq 0.43$.
-As the straying rate increases and $\Delta\theta$ decreases, a single stable state emerges as the pitchfork bifurcation is crossed, which is opposite the pattern observed when these parameters are independent.}
+\noindent {\bf Line 272:} \emph{Until now, we have treated the rate of straying and habitat heterogeneity as independent parameters, however they may also be assumed to covary.
+For instance, if sites are separated by greater distance, they may be assumed to have increased habitat heterogeneity as well as lower rates of straying.
+Alternatively, individuals may be genetically predisposed to stray into sites that are more similar, such that greater between-site heterogeneity will correspond to lower straying rates.
+We implemented this inverse relationship by setting $m = 0.5(1 + \Delta\theta)^{-1}$ where maximum straying is assumed to occur at $m=0.5$ (perfect mixing; figure S1).
+This assumes that $m$ is greater for lower $\Delta\theta$, such that there are low rates of straying between dissimilar (distant) sites and high straying rates between similar (close) sites.
+Under these conditions we find that alternative stable states appear for very low rates of straying (figure S10). % $0 < m \leq 0.43$.
+As the straying rate increases and $\Delta\theta$ decreases, a single stable state emerges as the cusp bifurcation is crossed, which is opposite the pattern observed when straying is independent of habitat heterogeneity.}
 
 \noindent \textcolor{cyan}{
-{\bf Comment 2.6} As I write above, I appreciate the focus on robustness and through numerical analysis.
+{\bf Comment 2.7} As I write above, I appreciate the focus on robustness and through numerical analysis.
 At the same time this model would lend itself to some analysis too, even making some simplifying assumptions or investigating some boundary conditions.
@@
 
-\noindent {\bf Response 2.6} Please see Responses 1.3 and 2.3.
+\noindent {\bf Response 2.7} Please see Responses 1.3 and 2.3.
 
 \noindent \textcolor{cyan}{
-{\bf Comment 2.7} Please present formal definitions (for example in a table as mentioned above) of all model variables and parameters.
+{\bf Comment 2.8} Please present formal definitions (for example in a table as mentioned above) of all model variables and parameters.
 Right now I never found an explicit definition of $N_T$, for example.
@@
 
-\noindent {\bf Response 2.7} A table is now included.
+\noindent {\bf Response 2.8} A table is now included.
 

### 2017-09-19 Manuscript/FinalDraft_rev/msv6.tex
final review response

@@
 
-% \bibliographystyle{prsb}
+\bibliographystyle{prsb}
 
@@
 \noindent where the first two components determine the mixed normal average of the aggregated local and remote populations.
-The partial derivative in the Eq.
+The partial derivative in Eq.
 \ref{eq:mu} determines how the mean trait changes through time due to natural selection \citep{Lande:1976ga}, which is proportional to the change in mean fitness with respect to $\mu_i$.
@@
 Numerically estimating the time that it takes for a perturbed system to recover also permits a more detailed perspective of metapopulation fragility.
-For example, if populations settle to alternative stable states, comparing recovery times after a disturbance applied to individual populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery.
+For example, if populations settle to alternative stable states (one at low- and one at high-density), comparing recovery times after a disturbance applied to individual populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery.
 % (alternative stable states in our model requiring one population to be high-density and one low-density)
 We measured recovery time following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} are equivalent if populations have the same density); (\emph{iii}) near-collapse of both populations where just 1.0\% of each survives.
-Throughout, we will refer to an increase in the portfolio effects and/or reduction in recovery times as promoting metapopulation robustness and by extension persistence.
+Throughout, we will refer to an increase in the portfolio effects and/or reduction in recovery times as promoting metapopulation robustness.
 \\
@@
 This prediction is in accordance with observations from natural populations \citep{Bett:2017ha}. %, regardless of trait heritability and variance or habitat heterogeneity
-The decline in steady state densities is not gradual: as straying increases, the system crosses a pitchfork bifurcation (PFB) \citep{AleksandrovichKuznetsov:1995p2580} whereby the single steady state for the metapopulation bifurcates into two basins of attraction: one at high biomass, and one at low biomass density (figure \ref{fig:traj}a, \ref{fig:PE}a).
-In discrete systems, a pitchfork bifurcation is a specific case of the cusp bifurcation (where two fold bifurcations intersect at a cusp) \citep{AleksandrovichKuznetsov:1995p2580}, and occurs when the real part of the dominant eigenvalue of the Jacobian matrix crosses the unit circle at +1 (figure \ref{fig:eigs}).
+The decline in steady state densities is not gradual: as straying increases, the system crosses a discrete cusp bifurcation (DCB) \citep{AleksandrovichKuznetsov:1995p2580} whereby the single steady state for the metapopulation bifurcates into two basins of attraction: one at high biomass, and one at low biomass density (figure \ref{fig:traj}a, \ref{fig:PE}a).
 Mean trait values for both populations bifurcate similarly (figure \ref{fig:traj}b). 
-
-Above the threshold straying rate defined by the PFB, there are two alternative eco-evolutionary states: the \emph{dominant state} population will have a higher density and greater degree of local adaptation (smaller trait offset from the local optimum), while the \emph{subordinate state} population will have lower density with maladapted trait values (larger trait offset from the local optimum).
+In discrete systems, the cusp bifurcation is defined by two fold bifurcations intersecting at a cusp point \citep{AleksandrovichKuznetsov:1995p2580}, and is observed when the real part of the dominant eigenvalue of the Jacobian matrix crosses the unit circle at +1 (figure \ref{fig:eigs}).
+Visually, the dynamics are similar to those observed at a pitchfork bifurcation in continuous systems, where a single steady state gives rise to two steady states separated by an unstable fixed point.
+
+Above the threshold straying rate defined by the DCB, there are two alternative eco-evolutionary states: the \emph{dominant state} population will have a higher density and greater degree of local adaptation (smaller trait offset from the local optimum), while the \emph{subordinate state} population will have lower density with maladapted trait values (larger trait offset from the local optimum).
 Hysteresis is observed to occur at this transition, such that the single steady state regime cannot easily be recovered by reducing straying after the system attains alternative steady states (figure \ref{fig:hysteresis}).
@@
 Lower trait heritability has little effect on the total biomass density but contrastingly large effects on $\Delta N^*$.
-The pitchfork bifurcation (the black line in Figs.
+The cusp bifurcation (the black line in Figs.
 \ref{fig:PE}a-c) occurs at lower values of the straying rate $m$ with lower heritability (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
@@
 %Critical slowing down
-As the pitchfork bifurcation is approached with increasing $m$, the portfolio effect increases sharply due to an amplification in variance within both donor and recipient populations.
+As the cusp bifurcation is approached with increasing $m$, the portfolio effect increases sharply due to an amplification in variance within both donor and recipient populations.
 This amplification in variance is the product of a dynamical process known as \emph{critical slowing down} that can occur near bifurcations \citep{Scheffer:2009gg}, a phenomenon that some have suggested may serve as an early warning indicator for approaching phase transitions \citep{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br,Krkosek:2014ch}.
-For larger values of $m$ (to the right of the pitchfork bifurcation in Fig \ref{fig:PE}a-c), where alternative stable states occur, the portfolio effect declines steadily as the CV of $N_T^*$ increases.
+For larger values of $m$ (to the right of the cusp bifurcation in Fig \ref{fig:PE}a-c), where alternative stable states occur, the portfolio effect declines steadily as the CV of $N_T^*$ increases.
 The decline over $m$ is more gradual if trait heritability is low, and steeper if trait heritability is high (figure \ref{fig:PE}c).
@@
 Straying had non-linear impacts on the recovery time of populations. 
-When the dominant state (well adapted and high density) is wiped out, high levels of straying allow it to recover quickly (figure \ref{fig:relax}a) because the surviving population has a mean trait value skewed towards the optimum of the recovering population (figure \ref{fig:relaxtraj_hdlh}).
-Yet, as straying decreases, recovery time for the disturbed dominant state population increases, in part because there is enough time for the trait distribution to move back towards the trait optimum of the subordinate state population.
+When the dominant state (well adapted and high density) is wiped out, high rates of straying allow it to recover quickly (figure \ref{fig:relax}a) because the surviving population has a mean trait value skewed towards the optimum of the recovering population (figure \ref{fig:relaxtraj_hdlh}).
+Yet, as straying decreases, recovery time for the disturbed dominant state population increases, in part because there is time for the trait distribution to move back towards the trait optimum of the subordinate state population.
 In contrast, when the subordinate state population (maladapted and low density) is wiped out, recovery rates are fastest at low to intermediate levels of straying.
 Because the mean trait values of both populations are skewed towards those of the dominant population, when the subordinate population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into the recovering population extends the length of time required for it to return to its steady state (figure \ref{fig:relaxtraj_ldlh}).
-When both populations are both dramatically reduced, recovery time is generally fastest at lower levels of straying, while near the onset of the pitchfork bifurcation, recovery time increases explosively and this is -- as the name implies -- characteristic of \emph{slowing} dynamics that occur near critical transitions \citep{Scheffer:2009gg,Kuehn:2010p2591}.\\
+When both populations are both dramatically reduced, recovery time is generally fastest at lower levels of straying due to reduced mixing of maladaptive phenotypes.
+Near the onset of the cusp bifurcation, recovery time increases explosively, however this is -- as the name implies -- characteristic of \emph{slow} dynamics occurring near critical transitions \citep{Scheffer:2009gg,Kuehn:2010p2591}.
+
+%Lowering Phenotypic diversity with increasing straying rate
+Increased rates of straying lowers phenotypic diversity ($\Delta \mu^* = |\mu_i^*-\mu_j^*|$, evaluated at the steady state) because both local and remote populations are increasingly homogenized.
+The loss of phenotypic diversity with increased straying is greater if trait heritability is low because traits take longer to go back to their local optima than they do when heritability is high. 
+Hence straying counters the effect of diversifying local adaptation. 
+% The loss of phenotypic diversity with increased straying is greater if trait heritability is low because the selective forces acting against trait means far from the local optima are lessened.
+% with decreased $h^2$.
+Less intuitively, we observe a discrete jump towards low phenotypic diversity as the cusp bifurcation is crossed (figure \ref{fig:traitdiff}).
+Although the development of alternative stable states elevates the portfolio effect due to the variance-dampening effects of the aggregate, entering this dynamic regime also results in a substantial decline in phenotypic diversity, which may have less predictable adverse effects on the population.\\
 
@@
 %Recovery times
-Density-dependent straying alters these recovery times (figure \ref{fig:relax}b). 
-First, in comparison with constant stray rates, density-dependent straying made recovery more rapid at elevated stray rates when both populations collapsed and when the subordinate population was extirpated.
+Density-dependent straying alters these recovery times (figure \ref{fig:relax}b; note difference in x-axis scale). 
+First, in comparison with constant stray rates, density-dependent straying lowered recovery times at elevated straying rates for near-collapse of both populations and extirpation of the subordinate population.
 %Recovery times
 %The recovery times for systems with constant and density-dependent straying both show similar trends for a given $m$ and $m^*$ (figure \ref{fig:relax}c,d), however there are some distinct quantitative differences.
-At low straying rates, near-collapse of both populations resulted in longer than expected recovery times, whereas in the alternative stable state regime (higher $m^*$), the recovery times for different disturbance types were very similar to systems with a constant $m$ (figure \ref{fig:relax}b; note difference in x-axis scales).
-As trait heritability increased, the metapopulation always recovered more quickly if the small population was lost (figure \ref{fig:relax_highh}).
+In contrast, at low straying rates, near-collapse of both populations required more time to recover.
+%whereas in the alternative stable state regime (higher $m^*$), the recovery times for different disturbance types were very similar to systems with a constant $m$ (figure \ref{fig:relax}b; note difference in x-axis scales).
+Trait heritability had a large effect on recovery times, with near-collapse requiring a more protracted time to recover in the alternative steady state regime (figure \ref{fig:relax_highh}).
+% As trait heritability increased, the metapopulation always recovered more quickly if the subordinate population was lost (figure \ref{fig:relax_highh}).
 % There was greater similarity in the recovery times for constant and density-dependent $m$ systems when trait heritability was high (figure \ref{fig:relax_highh}).
-The lower recovery time for systems with increased $m^*$ mirrors an elevated PE with higher density-dependent straying rates (figure \ref{fig:thetaPE}).
+In general, the lower recovery time for systems with increased $m^*$ mirrors an elevated PE with higher density-dependent straying rates (figure \ref{fig:thetaPE}a).
 % This higher PE at increased $m^*$ mirrors recovery times that are also lower than those observed for systems with constant $m$ (figure \ref{fig:relax}).
-In tandem, analysis of both PE and recovery time suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our minimal model, it does appear to promote robustness, particularly when the aggregate biomass is low and straying is correspondingly high.
-
-%Lowering Phenotypic diversity with increasing straying rate
-Increased rates of straying lowers phenotypic diversity ($\Delta \mu^* = |\mu_i^*-\mu_j^*|$, evaluated at the steady state) because both local and remote populations are increasingly homogenized.
-The loss of phenotypic diversity with increased straying is greater if trait heritability is low because traits take longer to go back to their local optima than they do when heritability is large. 
-Hence straying counters the effect of diversifying local adaptation. 
-% The loss of phenotypic diversity with increased straying is greater if trait heritability is low because the selective forces acting against trait means far from the local optima are lessened.
-% with decreased $h^2$.
-Less intuitively, we observe a discrete jump towards low phenotypic diversity as the pitchfork bifurcation is crossed (figure \ref{fig:traitdiff}).
-Although the development of alternative stable states elevates the portfolio effect due to the variance-dampening effects of the aggregate, entering this dynamic regime also results in a substantial decline in phenotypic diversity, which may have less predictable adverse effects on the population.
+Together, analysis of both PE and recovery time suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our minimal model, it does appear to promote robustness, particularly in the case of near-collapse of both populations when straying is high.
 \\
@@
 \noindent{\bf (c) The role of habitat heterogeneity and changing selective landscapes}\\ 
-\noindent With the onset of straying, we find that increasingly divergent trait optima generally lower $N_T^*$ and exaggerate $\Delta N^*$ (figure \ref{fig:thetadiffN}), such that the biomass distribution becomes increasingly uneven.
+\noindent With the onset of straying, we find that increasingly divergent trait optima generally lower $N_T^*$ and exaggerate $\Delta N^*$, and this is particularly pronounced for density dependent straying (figure \ref{fig:thetadiffN}).
+%, such that the biomass distribution becomes increasingly uneven.
 %I've checked this - it's a theshold type (logistic) response Supp fig?
@@
 %regardless of heritability.
-%The rate of straying that gives rise to a large PE at the pitchfork bifurcation is conditioned on the difference in optimal trait values between the local and remote populations, which we call $\Delta\theta = \sqrt{(\theta_i - \theta_j)^2}$.
-As habitat heterogeneity increases, alternative stable states appear at lower straying rates -- with the crossing of the pitchfork bifurcation, accompanied by a peak in the PE -- whereas the magnitude of increase in the PE also increases (figure \ref{fig:thetaPE}b), reducing recovery time (figure \ref{fig:relaxtheta}).
-For increased rates of straying, greater habitat heterogeneity erodes the PE (figure \ref{fig:thetaPE}b) and increases the recovery time (figure \ref{fig:relaxtheta}).
-These results together suggest that habitat heterogeneity, as measured as the differences in trait optima between two habitats $\Delta\theta$, promotes robustness when straying rates are low, and erodes robustness when straying rates are high.
+%The rate of straying that gives rise to a large PE at the cusp bifurcation is conditioned on the difference in optimal trait values between the local and remote populations, which we call $\Delta\theta = \sqrt{(\theta_i - \theta_j)^2}$.
+As habitat heterogeneity increases, alternative steady states appear at lower straying rates -- with the crossing of the cusp bifurcation, accompanied by a peak in the PE -- whereas the magnitude of increase in the PE also increases (figure \ref{fig:thetaPE}b), lowering recovery time (figure \ref{fig:relaxtheta}).
+For increased rates of straying, greater habitat heterogeneity erodes the PE (figure \ref{fig:thetaPE}b) and increases the recovery times (figure \ref{fig:relaxtheta}).
+In tandem, these results suggest that habitat heterogeneity promotes robustness when straying rates are low, and erodes robustness when straying rates are high.
 %This means that alternative stable states can emerge with relatively little dispersal between sites (figure \ref{fig:thetaPE}), however the increased PE (and shorter recovery times).
 
+Until now, we have treated the rate of straying and habitat heterogeneity as independent parameters, however they may also be assumed to covary.
+For instance, if sites are separated by greater distance, they may be assumed to have increased habitat heterogeneity as well as lower rates of straying.
+Alternatively, individuals may be genetically predisposed to stray into sites that are more similar, such that greater between-site heterogeneity will correspond to lower straying rates.
 % We next investigate the dynamics across different types of sites that host 
-Sites that are farther apart are likely to be more heterogeneous, while simultaneously facilitating lower rates of straying between populations.
+% Sites that are farther apart are likely to be more heterogeneous, while simultaneously facilitating lower rates of straying between populations.
 % If we assume that environmental heterogeneity increases with distance between sites, or that individuals may have behaviors that minimize their chances of straying into very different habitats, the rate of straying will decline with habitat heterogeneity.
 % We can implement this dependence by allowing $\Delta\theta$ to decrease as $m$ increases, such that there are low rates of straying between dissimilar populations and high rates of straying between similar populations.
-We implemented this dependence as set $m,m_0 = 0.5(1 + \Delta\theta)^{-1}$ where maximum straying is assumed to occur at $m=0.5$ (perfect mixing; figure \ref{fig:mthetarelation}).
-This assumes that $m$ or $m_0$ increases in situations where $\Delta\theta$ is lower, such that there are low rates of straying between dissimilar sites and high straying rates between similar sites.
-We find that under these conditions alternative stable states now appear for very low rates of straying $0 < m \leq 0.43$.
-As the straying rate increases and $\Delta\theta$ decreases, a single stable state emerges as the pitchfork bifurcation is crossed, which is opposite the pattern observed when these parameters are independent.
-
-There are two notable dynamics that emerge following extinction of the dominant population at low rates of straying between dissimilar (high $\Delta\theta$) sites:
-(\emph{i}) above a threshold $m$ value, the dominant population recovers quickly enough that the evolving subordinate phenotype is overwhelmed by incoming strays, and it shifts back to its pre-disturbance (subordinate) state;
+We implemented this inverse relationship by setting $m = 0.5(1 + \Delta\theta)^{-1}$ where maximum straying is assumed to occur at $m=0.5$ (perfect mixing; figure \ref{fig:mthetarelation}).
+This assumes that $m$ is greater for lower $\Delta\theta$, such that there are low rates of straying between dissimilar (distant) sites and high straying rates between similar (close) sites.
+Under these conditions we find that alternative stable states appear for very low rates of straying (figure \ref{fig:mthetaPE}). % $0 < m \leq 0.43$.
+As the straying rate increases and $\Delta\theta$ decreases, a single stable state emerges as the cusp bifurcation is crossed, which is opposite the pattern observed when straying is independent of habitat heterogeneity.
+
+There are two notable dynamics that emerge following extinction of the dominant population at low rates of straying between dissimilar (high $\Delta\theta$) sites (figure \ref{fig:mtheta}).
+(\emph{i}) Above a threshold $m$ value, the dominant population recovers quickly enough that the evolving subordinate phenotype is overwhelmed by incoming strays, and it shifts back to its pre-disturbance subordinate state;
 (\emph{ii}) below a threshold $m$ value, there is an \emph{inversion} between subordinate and dominant states: because there is enough time and isolation for the subordinate trait mean to shift towards its local optimum, and away from that of the recovering dominant population, the dominant population becomes subordinate, and the subordinate population becomes dominant (figure \ref{fig:inertia}).
@@
 A salient finding from our results is that straying can lead to the emergence of alternative stable states, pushing one of the populations to high density (the \emph{dominant state}), and one to low density (the \emph{subordinate state}).
-This pattern has been observed in other models of eco-evolutionary dynamics with explicit space \citep{Ronce:2001dp}, suggesting that it may be a general feature of spatially-linked traits that evolve toward local optima while being hindered by dispersal.
-An important quality of our framework is that there are similar forces that dictate interactions within and between sites, and this naturally results in a symmetry that could preclude the relevance of our finding for natural (and inherently less symmetric) systems.
-Although the emergence of alternative stable states via a pitchfork bifurcation is characteristic of symmetrical systems, we find that increasing the asymmetry in the vital rates of populations across sites does not alter the presence or position of the PFB (figure \ref{fig:symmetry}).
+This pattern has been observed in other models of eco-evolutionary dynamics with explicit space \citep{Ronce:2001dp}, suggesting that it may be a general feature of spatially-linked populations that evolve toward local optima while being hindered by dispersal.
+An important aspect of our framework is that there are similar forces that dictate interactions within and between sites, and this naturally results in a symmetry that could be perceived as limiting the relevance of our findings for natural (and inherently less symmetric) systems.
+Although the emergence of alternative stable states via a cusp bifurcation is characteristic of symmetrical systems, we find that increasing the asymmetry in the vital rates of populations across sites does not significantly alter the presence or position of the DCB (figure \ref{fig:symmetry}).
 That these patterns arise in alternative formulations and are insensitive to parameter asymmetry suggests that the emergence of dominant and subordinate states directly from eco-evolutionary dynamics may be a generally relevant phenomenon in ecological systems.
@@
 
-An intermediate straying rate maximizes metapopulation robustness. 
+An intermediate straying rate increases metapopulation robustness. 
 Results from our model reveal that the presence of just enough straying to cause formation of alternative stable states increases the portfolio effect (figure \ref{fig:PE}a). 
+We note that we do not consider the extremely high PE at the DCB, matched by an extremely long recovery time, to be an indicator of robustness, as the dynamics exactly at or very close to a bifurcation are unlikely to be realized in nature.
 Previous theoretical work has shown that increased connectivity may erode portfolio effects in herring metapopulations, where straying is also thought to be density-dependent \citep{Secor:2009ena}.
-Although high levels of dispersal in our system generally supports this finding, the interplay between dispersal and PE is more subtle when selection for local adaptations is considered, such that low to intermediate levels of density-dependent straying result in an elevated PE, thus increasing the buffering capacity of the metapopulation.
+Although high levels of dispersal in our system generally supports this finding, the interplay between dispersal and PE is more subtle when selection for local adaptations is considered.
+Low to intermediate levels of density-dependent straying result in an elevated PE, increasing the buffering capacity of the metapopulation.
 Although PE is measured at the steady state, low to intermediate rates of straying also appear to have a beneficial effect on transient dynamics.
-When there is just enough straying to cause alternative stable states, the time to recovery following an induced disturbance declines, though -- as with the PE -- recovery time then grows if straying becomes very large (figure \ref{fig:relax}a).
+When there is just enough straying to cause alternative stable states, the time to recovery following an induced disturbance declines, though -- as with the PE -- it then grows if straying becomes large (figure \ref{fig:relax}a).
 
@@
 This themed issue formalizes the role of collective movement in the ecology of natural systems and illuminates a signature of collective navigation in animal populations on the move.
-We observed three salient results that contribute to our understanding of collective movement that combine to suggest that density-dependent straying may play an important role in the persistence of metapopulations over evolutionary time.
+We highlight three important findings that contribute to our understanding of collective movement suggesting that density-dependent straying may play an important role in the persistence of metapopulations over evolutionary time.
 First, the inclusion of density-dependent straying does not qualitatively alter either (\emph{i}) steady state or (\emph{ii}) transient dynamics of our minimal eco-evolutionary model, but effectively rescales measures of robustness to the lower straying rates that emerge as a consequence of the coupled dynamics.
@@
 This may be particularly consequential for populations that are spatially adjacent but separated by sharp environmental boundaries, such that trait optima are divergent yet dispersal is relatively high.
-Such a scenario plays out repeatedly in the context of adjacent wild and hatchery-produced salmon. 
+Such a scenario plays out repeatedly in the context of wild and hatchery-produced salmon. 
 Although wild and hatchery populations may occur close on the landscape, and indeed often are sympatric within the same river network, the selective environments to which they are locally adapted differ dramatically \citep{Christie:2012bj}.
@@
 %mtheta
-All else being equal, habitats that are closer in space generally have greater similarity in environmental conditions than habitats that are geographically distant, and correspondingly phenotypes of more proximately located populations should be more similar to those that are distant \citep{Westley:2012ui}.
+In other cases, habitats that are closer in space can be assumed to have greater similarity in environmental conditions than those that are geographically distant, and phenotypes of more proximately located populations should be more similar \citep{Westley:2012ui}.
 It is thus reasonable to expect a larger number of straying individuals between sites that are geographically proximate and indeed evidence corroborates this prediction \citep{Candy:2000hu,JPE:JPE1383}.
-Alternatively, salmon that cue to specific environmental conditions may be more likely to stray into sites that have structurally and physiognamically similar habitats \citep{Peterson:2014gy} even if other potential sites are closer.
-These considerations justify imposing a direct relationship between the rate of straying $m$ and habitat heterogeneity: as site dissimilarity increases, so too should the straying rate decrease.
+Alternatively, salmon that cue to specific environmental conditions may be more likely to stray into sites that are structurally and physiognamically more similar \citep{Peterson:2014gy}.
+These considerations justify imposing a direct relationship between the rate of straying $m$ and habitat heterogeneity: as site dissimilarity increases, so too should the straying rate decrease (figure \ref{fig:mthetarelation}).
 % As site dissimilarity increases, so to should the optima in trait values for the resident populations.
@@
 %there is enough isolation to allow the surviving subordinate population to adapt towards its own optimum.
-If the straying rate is very low, there is an inversion in the alternative stable states following the disturbance, resulting in a state shift in dominance.
-This may have particular conservation implications for considerations of aiding dispersal following disturbances or reconnections of lost habitats \citep{Anderson:2013bf,Pess:2014isa}.
+Morever, there is an inversion in the alternative stable states following the disturbance, resulting in a state shift in dominance.
+This may have particular conservation implications when aiding dispersal following disturbances or reconnections of lost habitats \citep{Anderson:2013bf,Pess:2014isa}.
 % Moreover, local temperature regimes are known to play a central role in dictating local adaptation of salmon populations.
@@
 \noindent {\bf Acknowledgements:} We thank Sean Anderson for helpful discussions and comments on the manuscript.
-We also thank the guest editors Andrew Berdahl, Dora Biro, and Colin Torney, for inviting us to contribute to this themed issue.
+We also thank the guest editors Andrew Berdahl, Dora Biro, and Colin Torney, for inviting us to contribute to this themed issue, and two anonymous reviewers for their insightful comments and critiques.
 J.D.Y.
-was supported by startup funds at the University of California, Merced, and an Omidyar Postdoctoral Fellowship at the Santa Fe Institute.
+was supported by startup funds at the University of California, Merced and an Omidyar Postdoctoral Fellowship at the Santa Fe Institute.
 J.P.G.
@@
 
-\bibliography{aa_kevin}
+% \bibliography{aa_kevin}
 % 
-% \begin{thebibliography}{10}
-% \expandafter\ifx\csname urlstyle\endcsname\relax
-%   \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
-%   \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
-%   \urlstyle{rm}\Url}\fi
-% 
-% \bibitem{Brown:1977gk}
-% Brown JH, Kodric-Brown A, 1977 {Turnover rates in insular biogeography: effect
-%   of immigration on extinction}.
-% \newblock \emph{Ecology} \textbf{58}, 445--449
-% 
-% \bibitem{Earn:2000fm}
-% Earn DJD, Levin SA, Rohani P, 2000 {Coherence and conservation}.
-% \newblock \emph{Science} \textbf{290}, 1360--1364
-% 
-% \bibitem{Anonymous:2015gf}
-% Schindler DE, Armstrong JB, Reed TE, 2015 {The portfolio concept in ecology and
-%   evolution}.
-% \newblock \emph{Front. Ecol. Environ.} \textbf{13}, 257--263
-% 
-% \bibitem{Abbott:2017hl}
-% Abbott RE, Doak DF, Peterson ML, 2017 {Portfolio effects, climate change, and
-%   the persistence of small populations: analyses on the rare plant Saussurea
-%   weberi.}
-% \newblock \emph{Ecology} \textbf{98}, 1071--1081
-% 
-% \bibitem{Thibaut:2012km}
-% Thibaut LM, Connolly SR, 2013 {Understanding diversity-stability relationships:
-%   towards a unified model of portfolio effects.}
-% \newblock \emph{Ecol. Lett.} \textbf{16}, 140--150
-% 
-% \bibitem{MilnerGulland:2011vm}
-% Milner-Gulland EJ, Fryxell JM, Sinclair ARE, 2011 \emph{{Animal Migration}}.
-% \newblock Oxford: Oxford University Press
-% 
-% \bibitem{Bell:2011ki}
-% Bell G, Gonzalez A, 2011 {Adaptation and evolutionary rescue in metapopulations
-%   experiencing environmental deterioration}.
-% \newblock \emph{Science} \textbf{332}, 1327--1330
-% 
-% \bibitem{Carlson:2014is}
-% Carlson SM, Cunningham CJ, Westley PAH, 2014 {Evolutionary rescue in a changing
-%   world}.
-% \newblock \emph{Trends Ecol. Evol.} \textbf{29}, 521--530
-% 
-% \bibitem{Muhlfeld:2014hs}
-% Muhlfeld CC, Kovach RP, Jones LA, Al-Chokhachy R, Boyer MC, Leary RF, Lowe WH,
-%   Luikart G, Allendorf FW, 2014 {Invasive hybridization in a threatened species
-%   is accelerated by climate change}.
-% \newblock \emph{Nature Climate Change} \textbf{4}, 620--624
-% 
-% \bibitem{Lowe:2015ft}
-% Lowe WH, Muhlfeld CC, Allendorf FW, 2015 {Spatial sorting promotes the spread
-%   of maladaptive hybridization}.
-% \newblock \emph{Trends Ecol. Evol.} \textbf{30}, 456--462
-% 
-% \bibitem{H:2013fs}
-% Westley PAH, Quinn TP, 2013 {Rates of straying by hatchery-produced Pacific
-%   salmon (\emph{Oncorhynchus} spp.) and steelhead (\emph{Oncorhynchus mykiss})
-%   differ among species, life history types, and populations}.
-% \newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{70}, 735--746
-% 
-% \bibitem{Keefer:2014gg}
-% Keefer ML, Caudill CC, 2014 {Homing and straying by anadromous salmonids: a
-%   review of mechanisms and rates}.
-% \newblock \emph{Rev Fish Biol Fisher} \textbf{24}, 333--368
-% 
-% \bibitem{Bett:2017ha}
-% Bett NN, Hinch SG, Burnett NJ, Donaldson MR, Naman SM, 2017 {Causes and
-%   consequences of straying into small populations of Pacific salmon}.
-% \newblock \emph{Fisheries} \textbf{42}, 220--230
-% 
-% \bibitem{Berdahl:2015kv}
-% Berdahl A, Torney CJ, Schertzer E, Levin SA, 2015 {On the evolutionary
-%   interplay between dispersal and local adaptation in heterogeneous
-%   environments}.
-% \newblock \emph{Evolution} \textbf{69}, 1390--1405
-% 
-% \bibitem{Berdahl:2016dx}
-% Berdahl A, 2016 {Collective behavior as a driver of critical transitions in
-%   migratory populations}.
-% \newblock \emph{Movement Ecology} \textbf{4}, 1--12
-% 
-% \bibitem{Simons:2004jo}
-% Simons AM, 2004 {Many wrongs: The advantage of group navigation}.
-% \newblock \emph{Trends Ecol. Evol.} \textbf{19}, 453--455
-% 
-% \bibitem{Brenner:2012gl}
-% Brenner RE, Moffitt SD, Grant WS, 2012 {Straying of hatchery salmon in Prince
-%   William Sound, Alaska}.
-% \newblock \emph{Environ Biol Fish} \textbf{94}, 179--195
-% 
-% \bibitem{Johnson:2012fe}
-% Johnson RC, Weber PK, Wikert JD, Workman ML, MacFarlane RB, Grove MJ, Schmitt
-%   AK, 2012 {Managed metapopulations: do salmon hatchery
-%   {\textquoteleft}sources{\textquoteright} lead to in-river
-%   {\textquoteleft}sinks{\textquoteright} in conservation?}
-% \newblock \emph{PLoS ONE} \textbf{7}, e28880--11
-% 
-% \bibitem{Fullerton:2011ii}
-% Fullerton AH, Lindley ST, Pess GR, Feist BE, Steel EA, Mcelhany P, 2011 {Human
-%   influence on the spatial structure of threatened Pacific salmon
-%   metapopulations}.
-% \newblock \emph{Conserv Biol} \textbf{25}, 932--944
-% 
-% \bibitem{Quinn:2011tf}
-% Quinn TP, 2011 \emph{{The Behavior and Ecology of Pacific Salmon and Trout}}.
-% \newblock Vancouver: UBC Press
-% 
-% \bibitem{Jonsson:2011kg}
-% Jonsson B, Jonsson N, 2011 \emph{{Ecology of Atlantic Salmon and Brown Trout}}.
-% \newblock Dordrecht: Springer Netherlands
-% 
-% \bibitem{Quinn:1993ge}
-% Quinn TP, 1993 {A review of homing and straying of wild and hatchery-produced
-%   salmon}.
-% \newblock \emph{Fish Res} \textbf{18}, 29--44
-% 
-% \bibitem{Hendry:2004wf}
-% Hendry AP, Bohlin T, Jonsson B, Berg OK, 2004 {The evolution of philopatry and
-%   dispersal: homing versus straying in salmonids}.
-% \newblock In AP~Hendry, SC~Stearns, eds., \emph{Evolution Illuminated}. Oxford:
-%   Oxford University Press on Demand
-% 
-% \bibitem{Schtickzelle:2007wb}
-% Schtickzelle N, Quinn TP, 2007 {A metapopulation perspective for salmon and
-%   other anadromous fish}.
-% \newblock \emph{Fish Fish.} \textbf{8}, 297--314
-% 
-% \bibitem{Anderson:2014cx}
-% Anderson SC, Moore JW, McClure MM, Dulvy NK, Cooper AB, 2015 {Portfolio
-%   conservation of metapopulations under climate change}.
-% \newblock \emph{Ecol. Appl.} \textbf{25}, 559--572
-% 
-% \bibitem{Keefer:2008bs}
-% Keefer ML, Caudill CC, Peery CA, Lee SR, 2008 {Transporting juvenile salmonids
-%   around dams impairs adult migration}.
-% \newblock \emph{Ecol. Appl.} \textbf{18}, 1888--1900
-% 
-% \bibitem{Westley:2015to}
-% Westley PAH, Dittman AH, Ward EJ, Quinn TP, 2015 {Signals of climate,
-%   conspecific density, and watershed features in patterns of homing and
-%   dispersal by Pacific salmon.}
-% \newblock \emph{Ecology} \textbf{96}, 2823--2833
-% 
-% \bibitem{Bond:2016dz}
-% Bond MH, Westley PAH, Dittman AH, Holecek D, Marsh T, Quinn TP, 2016 {Combined
-%   effects of barge transportation, river environment, and rearing location on
-%   straying and migration of adult snake river fall-run Chinook salmon}.
-% \newblock \emph{T Am Fish Soc} \textbf{146}, 60--73
-% 
-% \bibitem{Peterson:2014gy}
-% Peterson DA, Hilborn R, Hauser L, 2014 {Local adaptation limits lifetime
-%   reproductive success of dispersers in a wild salmon metapopulation}.
-% \newblock \emph{Nat Commun} \textbf{5}, 3696
-% 
-% \bibitem{Moore:2010gs}
-% Moore JW, McClure M, Rogers LA, Schindler DE, 2010 {Synchronization and
-%   portfolio performance of threatened salmon}.
-% \newblock \emph{Conserv. Lett.} \textbf{3}, 340--348
-% 
-% \bibitem{Carlson:2011ce}
-% Carlson SM, Satterthwaite WH, Fleming IA, 2011 {Weakened portfolio effect in a
-%   collapsed salmon population complex}.
-% \newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{68}, 1579--1589
-% 
-% \bibitem{Braun:2016ib}
-% Braun DC, Moore JW, Candy J, Bailey RE, 2016 {Population diversity in salmon:
-%   linkages among response, genetic and life history diversity}.
-% \newblock \emph{Ecography} \textbf{39}, 317--328
-% 
-% \bibitem{simpson1953major}
-% Simpson GG, 1953 \emph{{The Major Features of Evolution}}.
-% \newblock New York: Simon and Schuster
-% 
-% \bibitem{Lande:1976ga}
-% Lande R, 1976 {Natural selection and random genetic drift in phenotypic
-%   evolution}.
-% \newblock \emph{Evolution} \textbf{30}, 314--334
-% 
-% \bibitem{Shelton:2011eq}
-% Shelton AO, Mangel M, 2011 {Fluctuations of fish populations and the magnifying
-%   effects of fishing.}
-% \newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{108}, 7075--7080
-% 
-% \bibitem{Jasper:2013cc}
-% Jasper JR, Habicht C, Moffitt S, Brenner R, Marsh J, Lewis B, Fox EC, Grauvogel
-%   Z, Olive SDR, Grant WS, 2013 {Source-sink estimates of genetic introgression
-%   show influence of hatchery strays on wild chum salmon populations in Prince
-%   William Sound, Alaska}.
-% \newblock \emph{PLoS ONE} \textbf{8}, e81916
-% 
-% \bibitem{Schreiber:2011wx}
-% Schreiber SJ, B{\"u}rger R, Bolnick DI, 2011 {The community effects of
-%   phenotypic and genetic variation within a predator population.}
-% \newblock \emph{Ecology} \textbf{92}, 1582--1593
-% 
-% \bibitem{Gilbert:2014ee}
-% Gibert JP, Brassil CE, 2014 {Individual phenotypic variation reduces
-%   interaction strengths in a consumer{\textendash}resource system}.
-% \newblock \emph{Ecol Evol} \textbf{4}, 3703--3713
-% 
-% \bibitem{Gibert:2015kc}
-% Gibert JP, DeLong JP, 2015 {Individual variation decreases interference
-%   competition but increases species persistence}.
-% \newblock \emph{Adv Ecol Res} \textbf{52}, 45--64
-% 
-% \bibitem{Berdahl:2014bl}
-% Berdahl A, Westley PAH, Levin SA, Couzin ID, Quinn TP, 2014 {A collective
-%   navigation hypothesis for homeward migration in anadromous salmonids}.
-% \newblock \emph{Fish Fish.} \textbf{17}, 525--542
-% 
-% \bibitem{Candy:2000hu}
-% Candy JR, Beacham TD, 2000 {Patterns of homing and straying in southern British
-%   Columbia coded-wire tagged Chinook salmon (\emph{Oncorhynchus tshawytscha})
-%   populations}.
-% \newblock \emph{Fish Res} \textbf{47}, 41--56
-% 
-% \bibitem{JPE:JPE1383}
-% Schick RS, Lindley ST, 2007 {Directed connectivity among fish populations in a
-%   riverine network}.
-% \newblock \emph{J. Appl. Ecol.} \textbf{44}, 1116--1126
-% 
-% \bibitem{Schindler:2015gf}
-% Schindler DE, Armstrong JB, Reed TE, 2015 {The portfolio concept in ecology and
-%   evolution}.
-% \newblock \emph{Front. Ecol. Environ.} \textbf{13}, 257--263
-% 
-% \bibitem{Ovaskainen:2002il}
-% Ovaskainen O, Hanski I, 2002 {Transient dynamics in metapopulation response to
-%   perturbation}.
-% \newblock \emph{Theor Popul Biol} \textbf{61}, 285--295
-% 
-% \bibitem{Anderson:2013gb}
-% Anderson SC, Cooper AB, Dulvy NK, 2013 {Ecological prophets: quantifying
-%   metapopulation portfolio effects}.
-% \newblock \emph{Methods Ecol Evol} \textbf{4}, 971--981
-% 
-% \bibitem{Loreau:2008ju}
-% Loreau M, de~Mazancourt C, 2008 {Species synchrony and its drivers: neutral and
-%   nonneutral community dynamics in fluctuating environments}.
-% \newblock \emph{Am. Nat.} \textbf{172}, E48--E66
-% 
-% \bibitem{Yeakel:2013vz}
-% Yeakel JD, Moore JW, Guimar{\~a}es~Jr PR, de~Aguiar MAM, 2014 {Synchronisation
-%   and stability in river metapopulation networks}.
-% \newblock \emph{Ecol. Lett.} \textbf{17}, 273--283
-% 
-% \bibitem{GuckHolmes}
-% Guckenheimer J, Holmes P, 1983 \emph{{Nonlinear Oscillations, Dynamical
-%   Systems, and Bifurcations of Vector Fields}}.
-% \newblock New York: Springer
-% 
-% \bibitem{Carlson:2008hl}
-% Carlson SM, Seamons TR, 2008 {A review of quantitative genetic components of
-%   fitness in salmonids: implications for adaptation to future change}.
-% \newblock \emph{Evol Appl} \textbf{1}, 222--238
-% 
-% \bibitem{Scheffer:2009gg}
-% Scheffer M, Bascompte J, Brock WA, Brovkin V, Carpenter SR, Dakos V, Held H,
-%   van Nes EH, Rietkerk M, Sugihara G, 2009 {Early-warning signals for critical
-%   transitions}.
-% \newblock \emph{Nature} \textbf{461}, 53--59
-% 
-% \bibitem{Lade:2012eu}
-% Lade SJ, Gross T, 2012 {Early warning signals for critical transitions: A
-%   generalized modeling approach}.
-% \newblock \emph{PLoS Comp. Biol.} \textbf{8}, e1002360
-% 
-% \bibitem{Anonymous:2013br}
-% Boettiger C, Ross N, Hastings A, 2013 {Early warning signals: the charted and
-%   uncharted territories}.
-% \newblock \emph{Theor. Ecol.} \textbf{6}, 255--264
-% 
-% \bibitem{Dakos:2014br}
-% Dakos V, Bascompte J, 2014 {Critical slowing down as early warning for the
-%   onset of collapse in mutualistic communities}.
-% \newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{111}, 17546--17551
-% 
-% \bibitem{Krkosek:2014ch}
-% Krko{\v s}ek M, Drake JM, 2014 {On signals of phase transitions in salmon
-%   population dynamics}.
-% \newblock \emph{Proc. Roy. Soc. B} \textbf{281}, 20133221
-% 
-% \bibitem{Kuehn:2010p2591}
-% Kuehn C, 2011 {A mathematical framework for critical transitions: bifurcations,
-%   fast-slow systems and stochastic dynamics}.
-% \newblock \emph{Physica D} \textbf{240}, 1020--1035
-% 
-% \bibitem{Donohue:2013iu}
-% Donohue I, Petchey OL, Montoya JM, Jackson AL, McNally L, Viana M, Healy K,
-%   Lurgi M, O'Connor NE, Emmerson MC, 2013 {On the dimensionality of ecological
-%   stability.}
-% \newblock \emph{Ecol. Lett.} \textbf{16}, 421--429
-% 
-% \bibitem{Levin:1992p1472}
-% Levin SA, 1992 {The problem of pattern and scale in ecology}.
-% \newblock \emph{Ecology} \textbf{73}, 1943--1967
-% 
-% \bibitem{Meinhardt:2012hw}
-% Meinhardt H, 2012 {Turing's theory of morphogenesis of 1952 and the subsequent
-%   discovery of the crucial role of local self-enhancement and long-range
-%   inhibition}.
-% \newblock \emph{Interface Focus} \textbf{2}, 407--416
-% 
-% \bibitem{Hanski:1999vu}
-% Hanski I, 1999 \emph{{Metapopulation Ecology}}.
-% \newblock Oxford: Oxford University Press
-% 
-% \bibitem{Boughton:1999fa}
-% Boughton DA, 1999 {Empirical evidence for complex source{\textendash}sink
-%   dynamics with alternative states in a butterfly metapopulation}.
-% \newblock \emph{Ecology} \textbf{80}, 2727--2739
-% 
-% \bibitem{Secor:2009ena}
-% Secor DH, Kerr LA, Cadrin SX, 2009 {Connectivity effects on productivity,
-%   stability, and persistence in a herring metapopulation model}.
-% \newblock \emph{ICES J Mar Sci} \textbf{66}, 1726--1732
-% 
-% \bibitem{Christie:2012bj}
-% Christie MR, Marine ML, French RA, Blouin MS, 2012 {Genetic adaptation to
-%   captivity can occur in a single generation.}
-% \newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{109}, 238--242
-% 
-% \bibitem{Chilcote:2003bb}
-% Chilcote MW, 2003 {Relationship between natural productivity and the frequency
-%   of wild fish in mixed spawning populations of wild and hatchery steelhead
-%   (\emph{Oncorhynchus mykiss})}.
-% \newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{60}, 1057--1067
-% 
-% \bibitem{Araki:2007cm}
-% Araki H, Cooper B, Blouin MS, 2007 {Genetic effects of captive breeding cause a
-%   rapid, cumulative fitness decline in the wild}.
-% \newblock \emph{Science} \textbf{318}, 100--103
-% 
-% \bibitem{Westley:2012ui}
-% Westley PAH, { Corinne M Conway}, Fleming IA, 2012 {Phenotypic divergence of
-%   exotic fish populations is shaped by spatial proximity and habitat
-%   differences across an invaded landscape}.
-% \newblock \emph{Evol. Ecol. Res.} \textbf{14}, 147--167
-% 
-% \bibitem{Anderson:2013bf}
-% Anderson JH, Faulds PL, Atlas WI, Quinn TP, 2013 {Reproductive success of
-%   captively bred and naturally spawned Chinook salmon colonizing newly
-%   accessible habitat}.
-% \newblock \emph{Evol Appl} \textbf{6}, 165--179
-% 
-% \bibitem{Pess:2014isa}
-% Pess GR, Quinn TP, Gephard SR, Saunders R, 2014 {Re-colonization of Atlantic
-%   and Pacific rivers by anadromous fishes: linkages between life history and
-%   the benefits of barrier removal}.
-% \newblock \emph{Rev Fish Biol Fisher} \textbf{24}, 881--900
-% 
-% \bibitem{Ford:2002ip}
-% Ford MJ, 2002 {Selection in captivity during supportive breeding may reduce
-%   fitness in the wild}.
-% \newblock \emph{Conserv Biol} \textbf{16}, 815--825
-% 
-% \bibitem{Liermann:2017gj}
-% Liermann M, PESS G, McHenry M, McMillan J, Elofson M, Bennett T, Moses R, 2017
-%   {Relocation and recolonization of Coho salmon \emph{Oncorhynchus kisutch} in
-%   two tributaries to the elwha river: Implications for management and
-%   monitoring}.
-% \newblock \emph{T Am Fish Soc} \textbf{17}, 1--39
-% 
-% \end{thebibliography}
+\begin{thebibliography}{10}
+\expandafter\ifx\csname urlstyle\endcsname\relax
+  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
+  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
+  \urlstyle{rm}\Url}\fi
+
+\bibitem{Brown:1977gk}
+Brown JH, Kodric-Brown A, 1977 {Turnover rates in insular biogeography: effect
+  of immigration on extinction}.
+\newblock \emph{Ecology} \textbf{58}, 445--449
+
+\bibitem{Earn:2000fm}
+Earn DJD, Levin SA, Rohani P, 2000 {Coherence and conservation}.
+\newblock \emph{Science} \textbf{290}, 1360--1364
+
+\bibitem{Anonymous:2015gf}
+Schindler DE, Armstrong JB, Reed TE, 2015 {The portfolio concept in ecology and
+  evolution}.
+\newblock \emph{Front. Ecol. Environ.} \textbf{13}, 257--263
+
+\bibitem{Abbott:2017hl}
+Abbott RE, Doak DF, Peterson ML, 2017 {Portfolio effects, climate change, and
+  the persistence of small populations: analyses on the rare plant Saussurea
+  weberi.}
+\newblock \emph{Ecology} \textbf{98}, 1071--1081
+
+\bibitem{Thibaut:2012km}
+Thibaut LM, Connolly SR, 2013 {Understanding diversity-stability relationships:
+  towards a unified model of portfolio effects.}
+\newblock \emph{Ecol. Lett.} \textbf{16}, 140--150
+
+\bibitem{MilnerGulland:2011vm}
+Milner-Gulland EJ, Fryxell JM, Sinclair ARE, 2011 \emph{{Animal Migration}}.
+\newblock Oxford: Oxford University Press
+
+\bibitem{Bell:2011ki}
+Bell G, Gonzalez A, 2011 {Adaptation and evolutionary rescue in metapopulations
+  experiencing environmental deterioration}.
+\newblock \emph{Science} \textbf{332}, 1327--1330
+
+\bibitem{Carlson:2014is}
+Carlson SM, Cunningham CJ, Westley PAH, 2014 {Evolutionary rescue in a changing
+  world}.
+\newblock \emph{Trends Ecol. Evol.} \textbf{29}, 521--530
+
+\bibitem{Muhlfeld:2014hs}
+Muhlfeld CC, Kovach RP, Jones LA, Al-Chokhachy R, Boyer MC, Leary RF, Lowe WH,
+  Luikart G, Allendorf FW, 2014 {Invasive hybridization in a threatened species
+  is accelerated by climate change}.
+\newblock \emph{Nature Climate Change} \textbf{4}, 620--624
+
+\bibitem{Lowe:2015ft}
+Lowe WH, Muhlfeld CC, Allendorf FW, 2015 {Spatial sorting promotes the spread
+  of maladaptive hybridization}.
+\newblock \emph{Trends Ecol. Evol.} \textbf{30}, 456--462
+
+\bibitem{H:2013fs}
+Westley PAH, Quinn TP, 2013 {Rates of straying by hatchery-produced Pacific
+  salmon (\emph{Oncorhynchus} spp.) and steelhead (\emph{Oncorhynchus mykiss})
+  differ among species, life history types, and populations}.
+\newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{70}, 735--746
+
+\bibitem{Keefer:2014gg}
+Keefer ML, Caudill CC, 2014 {Homing and straying by anadromous salmonids: a
+  review of mechanisms and rates}.
+\newblock \emph{Rev Fish Biol Fisher} \textbf{24}, 333--368
+
+\bibitem{Bett:2017ha}
+Bett NN, Hinch SG, Burnett NJ, Donaldson MR, Naman SM, 2017 {Causes and
+  consequences of straying into small populations of Pacific salmon}.
+\newblock \emph{Fisheries} \textbf{42}, 220--230
+
+\bibitem{Berdahl:2015kv}
+Berdahl A, Torney CJ, Schertzer E, Levin SA, 2015 {On the evolutionary
+  interplay between dispersal and local adaptation in heterogeneous
+  environments}.
+\newblock \emph{Evolution} \textbf{69}, 1390--1405
+
+\bibitem{Berdahl:2016dx}
+Berdahl A, 2016 {Collective behavior as a driver of critical transitions in
+  migratory populations}.
+\newblock \emph{Movement Ecology} \textbf{4}, 1--12
+
+\bibitem{Simons:2004jo}
+Simons AM, 2004 {Many wrongs: The advantage of group navigation}.
+\newblock \emph{Trends Ecol. Evol.} \textbf{19}, 453--455
+
+\bibitem{Brenner:2012gl}
+Brenner RE, Moffitt SD, Grant WS, 2012 {Straying of hatchery salmon in Prince
+  William Sound, Alaska}.
+\newblock \emph{Environ Biol Fish} \textbf{94}, 179--195
+
+\bibitem{Johnson:2012fe}
+Johnson RC, Weber PK, Wikert JD, Workman ML, MacFarlane RB, Grove MJ, Schmitt
+  AK, 2012 {Managed metapopulations: do salmon hatchery
+  {\textquoteleft}sources{\textquoteright} lead to in-river
+  {\textquoteleft}sinks{\textquoteright} in conservation?}
+\newblock \emph{PLoS ONE} \textbf{7}, e28880--11
+
+\bibitem{Fullerton:2011ii}
+Fullerton AH, Lindley ST, Pess GR, Feist BE, Steel EA, Mcelhany P, 2011 {Human
+  influence on the spatial structure of threatened Pacific salmon
+  metapopulations}.
+\newblock \emph{Conserv Biol} \textbf{25}, 932--944
+
+\bibitem{Quinn:2011tf}
+Quinn TP, 2011 \emph{{The Behavior and Ecology of Pacific Salmon and Trout}}.
+\newblock Vancouver: UBC Press
+
+\bibitem{Jonsson:2011kg}
+Jonsson B, Jonsson N, 2011 \emph{{Ecology of Atlantic Salmon and Brown Trout}}.
+\newblock Dordrecht: Springer Netherlands
+
+\bibitem{Quinn:1993ge}
+Quinn TP, 1993 {A review of homing and straying of wild and hatchery-produced
+  salmon}.
+\newblock \emph{Fish Res} \textbf{18}, 29--44
+
+\bibitem{Hendry:2004wf}
+Hendry AP, Bohlin T, Jonsson B, Berg OK, 2004 {The evolution of philopatry and
+  dispersal: homing versus straying in salmonids}.
+\newblock In AP~Hendry, SC~Stearns, eds., \emph{Evolution Illuminated}. Oxford:
+  Oxford University Press on Demand
+
+\bibitem{Schtickzelle:2007wb}
+Schtickzelle N, Quinn TP, 2007 {A metapopulation perspective for salmon and
+  other anadromous fish}.
+\newblock \emph{Fish Fish.} \textbf{8}, 297--314
+
+\bibitem{Anderson:2014cx}
+Anderson SC, Moore JW, McClure MM, Dulvy NK, Cooper AB, 2015 {Portfolio
+  conservation of metapopulations under climate change}.
+\newblock \emph{Ecol. Appl.} \textbf{25}, 559--572
+
+\bibitem{Keefer:2008bs}
+Keefer ML, Caudill CC, Peery CA, Lee SR, 2008 {Transporting juvenile salmonids
+  around dams impairs adult migration}.
+\newblock \emph{Ecol. Appl.} \textbf{18}, 1888--1900
+
+\bibitem{Westley:2015to}
+Westley PAH, Dittman AH, Ward EJ, Quinn TP, 2015 {Signals of climate,
+  conspecific density, and watershed features in patterns of homing and
+  dispersal by Pacific salmon.}
+\newblock \emph{Ecology} \textbf{96}, 2823--2833
+
+\bibitem{Bond:2016dz}
+Bond MH, Westley PAH, Dittman AH, Holecek D, Marsh T, Quinn TP, 2016 {Combined
+  effects of barge transportation, river environment, and rearing location on
+  straying and migration of adult snake river fall-run Chinook salmon}.
+\newblock \emph{T Am Fish Soc} \textbf{146}, 60--73
+
+\bibitem{Peterson:2014gy}
+Peterson DA, Hilborn R, Hauser L, 2014 {Local adaptation limits lifetime
+  reproductive success of dispersers in a wild salmon metapopulation}.
+\newblock \emph{Nat Commun} \textbf{5}, 3696
+
+\bibitem{Moore:2010gs}
+Moore JW, McClure M, Rogers LA, Schindler DE, 2010 {Synchronization and
+  portfolio performance of threatened salmon}.
+\newblock \emph{Conserv. Lett.} \textbf{3}, 340--348
+
+\bibitem{Carlson:2011ce}
+Carlson SM, Satterthwaite WH, Fleming IA, 2011 {Weakened portfolio effect in a
+  collapsed salmon population complex}.
+\newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{68}, 1579--1589
+
+\bibitem{Braun:2016ib}
+Braun DC, Moore JW, Candy J, Bailey RE, 2016 {Population diversity in salmon:
+  linkages among response, genetic and life history diversity}.
+\newblock \emph{Ecography} \textbf{39}, 317--328
+
+\bibitem{simpson1953major}
+Simpson GG, 1953 \emph{{The Major Features of Evolution}}.
+\newblock New York: Simon and Schuster
+
+\bibitem{Lande:1976ga}
+Lande R, 1976 {Natural selection and random genetic drift in phenotypic
+  evolution}.
+\newblock \emph{Evolution} \textbf{30}, 314--334
+
+\bibitem{Shelton:2011eq}
+Shelton AO, Mangel M, 2011 {Fluctuations of fish populations and the magnifying
+  effects of fishing.}
+\newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{108}, 7075--7080
+
+\bibitem{Jasper:2013cc}
+Jasper JR, Habicht C, Moffitt S, Brenner R, Marsh J, Lewis B, Fox EC, Grauvogel
+  Z, Olive SDR, Grant WS, 2013 {Source-sink estimates of genetic introgression
+  show influence of hatchery strays on wild chum salmon populations in Prince
+  William Sound, Alaska}.
+\newblock \emph{PLoS ONE} \textbf{8}, e81916
+
+\bibitem{Schreiber:2011wx}
+Schreiber SJ, B{\"u}rger R, Bolnick DI, 2011 {The community effects of
+  phenotypic and genetic variation within a predator population.}
+\newblock \emph{Ecology} \textbf{92}, 1582--1593
+
+\bibitem{Gilbert:2014ee}
+Gibert JP, Brassil CE, 2014 {Individual phenotypic variation reduces
+  interaction strengths in a consumer{\textendash}resource system}.
+\newblock \emph{Ecol Evol} \textbf{4}, 3703--3713
+
+\bibitem{Gibert:2015kc}
+Gibert JP, DeLong JP, 2015 {Individual variation decreases interference
+  competition but increases species persistence}.
+\newblock \emph{Adv Ecol Res} \textbf{52}, 45--64
+
+\bibitem{Ronce:2001dp}
+Ronce O, Kirkpatrick M, 2001 {When sources become sinks: migrational meltdown
+  in heterogeneous habitats}.
+\newblock \emph{Evolution} \textbf{55}, 1520
+
+\bibitem{Berdahl:2014bl}
+Berdahl A, Westley PAH, Levin SA, Couzin ID, Quinn TP, 2014 {A collective
+  navigation hypothesis for homeward migration in anadromous salmonids}.
+\newblock \emph{Fish Fish.} \textbf{17}, 525--542
+
+\bibitem{Candy:2000hu}
+Candy JR, Beacham TD, 2000 {Patterns of homing and straying in southern British
+  Columbia coded-wire tagged Chinook salmon (\emph{Oncorhynchus tshawytscha})
+  populations}.
+\newblock \emph{Fish Res} \textbf{47}, 41--56
+
+\bibitem{JPE:JPE1383}
+Schick RS, Lindley ST, 2007 {Directed connectivity among fish populations in a
+  riverine network}.
+\newblock \emph{J. Appl. Ecol.} \textbf{44}, 1116--1126
+
+\bibitem{Schindler:2015gf}
+Schindler DE, Armstrong JB, Reed TE, 2015 {The portfolio concept in ecology and
+  evolution}.
+\newblock \emph{Front. Ecol. Environ.} \textbf{13}, 257--263
+
+\bibitem{Ovaskainen:2002il}
+Ovaskainen O, Hanski I, 2002 {Transient dynamics in metapopulation response to
+  perturbation}.
+\newblock \emph{Theor Popul Biol} \textbf{61}, 285--295
+
+\bibitem{Anderson:2013gb}
+Anderson SC, Cooper AB, Dulvy NK, 2013 {Ecological prophets: quantifying
+  metapopulation portfolio effects}.
+\newblock \emph{Methods Ecol Evol} \textbf{4}, 971--981
+
+\bibitem{Loreau:2008ju}
+Loreau M, de~Mazancourt C, 2008 {Species synchrony and its drivers: neutral and
+  nonneutral community dynamics in fluctuating environments}.
+\newblock \emph{Am. Nat.} \textbf{172}, E48--E66
+
+\bibitem{Yeakel:2013vz}
+Yeakel JD, Moore JW, Guimar{\~a}es~Jr PR, de~Aguiar MAM, 2014 {Synchronisation
+  and stability in river metapopulation networks}.
+\newblock \emph{Ecol. Lett.} \textbf{17}, 273--283
+
+\bibitem{GuckHolmes}
+Guckenheimer J, Holmes P, 1983 \emph{{Nonlinear Oscillations, Dynamical
+  Systems, and Bifurcations of Vector Fields}}.
+\newblock New York: Springer
+
+\bibitem{AleksandrovichKuznetsov:1995p2580}
+Kuznetsov Y, 1998 \emph{{Elements of Applied Bifurcation Theory}}.
+\newblock New York: Springer
+
+\bibitem{Carlson:2008hl}
+Carlson SM, Seamons TR, 2008 {A review of quantitative genetic components of
+  fitness in salmonids: implications for adaptation to future change}.
+\newblock \emph{Evol Appl} \textbf{1}, 222--238
+
+\bibitem{Scheffer:2009gg}
+Scheffer M, Bascompte J, Brock WA, Brovkin V, Carpenter SR, Dakos V, Held H,
+  van Nes EH, Rietkerk M, Sugihara G, 2009 {Early-warning signals for critical
+  transitions}.
+\newblock \emph{Nature} \textbf{461}, 53--59
+
+\bibitem{Lade:2012eu}
+Lade SJ, Gross T, 2012 {Early warning signals for critical transitions: A
+  generalized modeling approach}.
+\newblock \emph{PLoS Comp. Biol.} \textbf{8}, e1002360
+
+\bibitem{Anonymous:2013br}
+Boettiger C, Ross N, Hastings A, 2013 {Early warning signals: the charted and
+  uncharted territories}.
+\newblock \emph{Theor. Ecol.} \textbf{6}, 255--264
+
+\bibitem{Dakos:2014br}
+Dakos V, Bascompte J, 2014 {Critical slowing down as early warning for the
+  onset of collapse in mutualistic communities}.
+\newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{111}, 17546--17551
+
+\bibitem{Krkosek:2014ch}
+Krko{\v s}ek M, Drake JM, 2014 {On signals of phase transitions in salmon
+  population dynamics}.
+\newblock \emph{Proc. Roy. Soc. B} \textbf{281}, 20133221
+
+\bibitem{Kuehn:2010p2591}
+Kuehn C, 2011 {A mathematical framework for critical transitions: bifurcations,
+  fast-slow systems and stochastic dynamics}.
+\newblock \emph{Physica D} \textbf{240}, 1020--1035
+
+\bibitem{Donohue:2013iu}
+Donohue I, Petchey OL, Montoya JM, Jackson AL, McNally L, Viana M, Healy K,
+  Lurgi M, O'Connor NE, Emmerson MC, 2013 {On the dimensionality of ecological
+  stability.}
+\newblock \emph{Ecol. Lett.} \textbf{16}, 421--429
+
+\bibitem{Secor:2009ena}
+Secor DH, Kerr LA, Cadrin SX, 2009 {Connectivity effects on productivity,
+  stability, and persistence in a herring metapopulation model}.
+\newblock \emph{ICES J Mar Sci} \textbf{66}, 1726--1732
+
+\bibitem{Christie:2012bj}
+Christie MR, Marine ML, French RA, Blouin MS, 2012 {Genetic adaptation to
+  captivity can occur in a single generation.}
+\newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{109}, 238--242
+
+\bibitem{Chilcote:2003bb}
+Chilcote MW, 2003 {Relationship between natural productivity and the frequency
+  of wild fish in mixed spawning populations of wild and hatchery steelhead
+  (\emph{Oncorhynchus mykiss})}.
+\newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{60}, 1057--1067
+
+\bibitem{Araki:2007cm}
+Araki H, Cooper B, Blouin MS, 2007 {Genetic effects of captive breeding cause a
+  rapid, cumulative fitness decline in the wild}.
+\newblock \emph{Science} \textbf{318}, 100--103
+
+\bibitem{Westley:2012ui}
+Westley PAH, { Corinne M Conway}, Fleming IA, 2012 {Phenotypic divergence of
+  exotic fish populations is shaped by spatial proximity and habitat
+  differences across an invaded landscape}.
+\newblock \emph{Evol. Ecol. Res.} \textbf{14}, 147--167
+
+\bibitem{Anderson:2013bf}
+Anderson JH, Faulds PL, Atlas WI, Quinn TP, 2013 {Reproductive success of
+  captively bred and naturally spawned Chinook salmon colonizing newly
+  accessible habitat}.
+\newblock \emph{Evol Appl} \textbf{6}, 165--179
+
+\bibitem{Pess:2014isa}
+Pess GR, Quinn TP, Gephard SR, Saunders R, 2014 {Re-colonization of Atlantic
+  and Pacific rivers by anadromous fishes: linkages between life history and
+  the benefits of barrier removal}.
+\newblock \emph{Rev Fish Biol Fisher} \textbf{24}, 881--900
+
+\bibitem{Ford:2002ip}
+Ford MJ, 2002 {Selection in captivity during supportive breeding may reduce
+  fitness in the wild}.
+\newblock \emph{Conserv Biol} \textbf{16}, 815--825
+
+\bibitem{Liermann:2017gj}
+Liermann M, PESS G, McHenry M, McMillan J, Elofson M, Bennett T, Moses R, 2017
+  {Relocation and recolonization of Coho salmon \emph{Oncorhynchus kisutch} in
+  two tributaries to the elwha river: Implications for management and
+  monitoring}.
+\newblock \emph{T Am Fish Soc} \textbf{17}, 1--39
+
+\end{thebibliography}
 
@@
 (b) The steady state trait values measured as $\theta_i - x_i$, as a function of a constant stray rate $m$. 
-PFB marks the pitchfork bifurcation.
+DCB marks the discrete cusp bifurcation.
 Unless otherwise indicated, the default parameter values used are: $r_{\rm max}=2$; $Z=0.5$; $\beta=0.001$; $\theta_1=5$; $\Delta\theta=5$; $\tau=1$; $\sigma=1$; $T=1\times10^5$.
@@
 (c) the portfolio effect PE as a function of heritability $h^2$ and a constant stray rate $m$. Light colors = high values.
-The black line shows the pitchfork bifurcation separating a single steady state (left) from alternative stable states (right).
+The black line shows the cusp bifurcation separating a single steady state (left) from alternative stable states (right).
 (d) The relationship between the time to recovery following a disturbance and the portfolio effect. Portfolio effects greater than unity corresponds to less synchronization.
-The black line denotes the pitchfork bifurcation.
 } \label{fig:PE}
@@
 If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
-PFB marks the pitchfork bifurcation.
+DCB marks the discrete cusp bifurcation.
 } \label{fig:relax}
 \end{figure*}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.5\textwidth]{fig_traitdiff.pdf}
+\caption{
+Phenotypic diversity ($\Delta \mu^*$) evaluated at the steady state as a function of straying rate $m$ and trait heritability $h^2$.
+The jump occurs as the system crosses the cusp bifurcation; lower phenoytic diversity emerges with higher straying rates and in the alternative stable state regime.
+} \label{fig:traitdiff}
+\end{figure}
 
@@
 Portfolio effects greater than unity corresponds to less synchronization.
-PFB marks the pitchfork bifurcation.
+DCB marks the discrete cusp bifurcation.
 } \label{fig:thetaPE}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.5\textwidth]{fig_traitdiff.pdf}
-\caption{
-Phenotypic diversity ($\Delta \mu^*$) evaluated at the steady state as a function of straying rate $m$ and trait heritability $h^2$.
-The descrete jump occurs as the system crosses the pitchfork bifurcation; lower phenoytic diversity emerges with higher straying rates and in the alternative stable state regime.
-} \label{fig:traitdiff}
 \end{figure}
@@
   The `$*$' marks the value of $(m,\Delta\theta)$ below which there is a switch in subordinate/dominant states following extinction of the dominant population.
-  PFB marks the pitchfork bifurcation.
+  DCB marks the discrete cusp bifurcation.
   } \label{fig:mtheta}
@@
 \centering
-% \includegraphics[width=0.35\textwidth]{fig_mthetarelation.pdf}
+% \includegraphics[width=0.35\textwidth]{fig_eigs.pdf}
 \caption{
 The real parts of the four eigenvalues for the Jacobian matrix of the 4-dimensional system.
-The pitchfork bifurcation occurs when the dominant eigenvalue crosses the unit circle at $+1$. 
+The cusp bifurcation occurs when the dominant eigenvalue crosses the unit circle at $+1$. 
 } \label{fig:eigs}
@@
 \centering
-% \includegraphics[width=0.35\textwidth]{fig_mthetarelation.pdf}
+% \includegraphics[width=0.35\textwidth]{fig_hysteresis.pdf}
 \caption{
@@
 } \label{fig:hysteresis}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+% \includegraphics[width=0.35\textwidth]{fig_relax_large.pdf}
+\caption{
+Extinction of high-density population with a high straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
+Black line marks the calculated point of recovery post-perturbation.
+Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
+} \label{fig:relaxtraj_hdlh}
 \end{figure}
@@
 \centering
-% \includegraphics[width=0.35\textwidth]{fig_relax_large.pdf}
-\caption{
-Extinction of high-density population with a high straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
-Black line marks the calculated point of recovery post-perturbation.
-Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
-} \label{fig:relaxtraj_hdlh}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
 % \includegraphics[width=0.9\textwidth]{fig_relax_highh.pdf}
@@
 Median difference in population densities taken over the straying rate as a function of habitat heterogeneity $\Delta\theta$.
-Solid lines are for constant $m$; dashed lines are for density-dependent $m$} \label{fig:thetadiffN}
+Solid lines are for constant $m$; dashed lines are for density-dependent $m$.} \label{fig:thetadiffN}
 \end{figure}
@@
 (a) Recovery time after near collapse of both populations as a function of straying rate $m$ and habitat heterogeneity $\Delta\theta$.
-(b) The same as (a) but including recovery times when straying is density-dependent, shown by linked pairs of points.
-Recovery times for systems with density-dependent straying are longer at low straying rates and shorter at higher straying rates, mirroring the change in portfolio effects shown in figure \ref{fig:thetaPE}.
+(b) The same as (a) but including recovery times when straying is density-dependent evaluated at $m^*$, shown by linked point pairs.
+Recovery times for systems with density-dependent straying are longer when straying is low and shorter when straying is high, mirroring the change in portfolio effects with respect to density dependent straying shown in figure \ref{fig:thetaPE}.
 } \label{fig:relaxtheta}
@@
   \caption{
-  Distance dependent portfolio effects as a function of straying rate $m$ and trait heritability $h^2$. When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases.
+  Portfolio effects as a function of straying rate $m$ and trait heritability $h^2$ when the rate of straying is $m = 0.5(1 + \Delta\theta)^{-1}$.
+Alternative steady states emerge for low values of $m$ (left of the cusp bifurcation, denoted by the black line), whereas a single steady state exists for high $m$.
   } \label{fig:mthetaPE}
@@
 \caption{
-Distance dependent straying, where increased differences in trait optima between sites $\Delta\theta$ corresponds to lower rates of straying $m$.
+Population inversion where increased differences in trait optima between sites $\Delta\theta$ corresponds to lower rates of straying $m$.
 At low rates of straying $m=0.02$ ($\Delta\theta=24$), extinction of the dominant population leads to slower-than-expected recovery times because the subordinate population is isolated enough to evolve towards its own trait optimum.
 %until growth of the dominant population overwhelms this local selection.
-In this case, $m$ is less than $m=0.034$ (denoted by the asterisk in figure \ref{fig:mtheta}), such that isolation allows the subdominant population to `run away' from the influence of the dominant population, leading to a switch in states.
-If $m$ is low but greater than $0.034$, isolation permits the subdominant population to `run away' from the influence of the dominant population, until it is overwhelmed by the recovering dominant population, and reverts back to its previous trait mean prior to the disturbance.
+In this case, $m$ is less than $m=0.034$ (denoted by the asterisk in figure \ref{fig:mtheta}), such that isolation allows the subordinate population to `run away' from the influence of the dominant population.
+This leads to a switch in subordinate/dominant states for the two populations.
+If $m$ is low but greater than $0.034$, isolation permits the subordinate population to `run away' from the influence of the dominant population, until it is overwhelmed by the recovering dominant population, and reverts back to its previous trait mean prior to disturbance.
 } \label{fig:inertia}
@@
   \caption{
-  Distance dependent recovery times for three disturbance types.
-When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases for constant (a) and density-dependent (b) staying rates.
-  The pitchfork bifurcation is not as clear in (b) because $\Delta\theta$ is a function of the individual straying rate $m_0$, whereas the x-axis in (b) is the straying rate at the steady state $m^*$.
-  Despite this difference, the general trends shown in (a) are also present in (b).
+  Recovery times for three disturbance types when the straying rate covaries with habitat heterogeneity as $m = 0.5(1 + \Delta\theta)^{-1}$ for constant (a) and density-dependent (b) straying rates.
+  The cusp bifurcation is not as clear in (b) because $\Delta\theta$ is a function of the individual straying rate $m_0$, whereas the x-axis in (b) is the straying rate at the steady state $m^*$.
+  Despite this difference, the general behavior shown in (a) are the same in (b).
   } \label{fig:mthetamvm}
@@
 \caption{
-Steady state densities of both populations as a function of $m$, where a pitchfork bifurcation indicates the emergence of alternative steady states: one in a dominant state and one in a subordinate state.
+Steady state densities of both populations as a function of $m$, where a cusp bifurcation indicates the emergence of alternative steady states: one in a dominant state and one in a subordinate state.
 Steady states for populations with symmetrical values ($\alpha=0$) in the vital rates $r_{\rm max}$ and $\beta$ are shown with cool tones.

### 2017-09-19 Manuscript/FinalDraft_rev/supp.tex
final review response

@@
 The real parts of the four eigenvalues for the Jacobian matrix of the 4-dimensional system.
-The pitchfork bifurcation occurs when the dominant eigenvalue crosses the unit circle at $+1$. 
+The cusp bifurcation occurs when the dominant eigenvalue crosses the unit circle at $+1$. 
 } \label{fig:eigs}
@@
 \centering
+\includegraphics[width=0.35\textwidth]{fig_relax_large.pdf}
+\caption{
+Extinction of high-density population with a high straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
+Black line marks the calculated point of recovery post-perturbation.
+Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
+} \label{fig:relaxtraj_hdlh}
+\end{figure}
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
 \includegraphics[width=0.35\textwidth]{fig_relax_small.pdf}
 \caption{
-Extinction of low-density population with a high constant straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{extfig-fig:relax}a).
+Extinction of low-density population with a high constant straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
 Black line marks the calculated point of recovery post-perturbation.
@@
 \centering
-\includegraphics[width=0.35\textwidth]{fig_relax_large.pdf}
-\caption{
-Extinction of high-density population with a high straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{extfig-fig:relax}a).
-Black line marks the calculated point of recovery post-perturbation.
-Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
-} \label{fig:relaxtraj_hdlh}
-\end{figure}
-% 
-% \begin{figure}
-%   \captionsetup{justification=raggedright,
-% singlelinecheck=false
-% }
-% \centering
-% \includegraphics[width=0.35\textwidth]{figs2/fig_relax_small_highh.pdf}
-% \caption{
-% Extinction of low-density population with a high constant straying rate $m=0.4$ and high trait heritability $h^2=0.8$ (see figure \ref{fig:relax}a).
-% Black line marks the calculated point of recovery post-perturbation.
-% Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
-% } \label{fig:relaxtraj_ldhh}
-% \end{figure}
-% 
-% \begin{figure}
-%   \captionsetup{justification=raggedright,
-% singlelinecheck=false
-% }
-% \centering
-% \includegraphics[width=0.35\textwidth]{figs2/fig_relax_large_highh.pdf}
-% \caption{
-% Extinction of high-density population with a high straying rate $m=0.4$ and high trait heritability $h^2=0.8$ (see figure \ref{fig:relax}a).
-% Black line marks the calculated point of recovery post-perturbation.
-% Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
-% } \label{fig:relaxtraj_hdhh}
-% \end{figure}
-
-\begin{figure*}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
 \includegraphics[width=0.9\textwidth]{fig_relax_highh.pdf}
@@
 } \label{fig:relax_highh}
-\end{figure*}
+\end{figure}
 
@@
 Median difference in population densities taken over the straying rate as a function of habitat heterogeneity $\Delta\theta$.
-Solid lines are for constant $m$; dashed lines are for density-dependent $m$} \label{fig:thetadiffN}
+Solid lines are for constant $m$; dashed lines are for density-dependent $m$.} \label{fig:thetadiffN}
 \end{figure}
 
-\begin{figure*}
+\begin{figure}
   \captionsetup{justification=raggedright,
@@
 (a) Recovery time after near collapse of both populations as a function of straying rate $m$ and habitat heterogeneity $\Delta\theta$.
-(b) The same as (a) but including recovery times when straying is density-dependent, shown by linked pairs of points.
-Recovery times for systems with density-dependent straying are longer at low straying rates and shorter at higher straying rates, mirroring the change in portfolio effects shown in figure \ref{extfig-fig:thetaPE}.
+(b) The same as (a) but including recovery times when straying is density-dependent evaluated at $m^*$, shown by linked point pairs.
+Recovery times for systems with density-dependent straying are longer when straying is low and shorter when straying is high, mirroring the change in portfolio effects with respect to density dependent straying shown in figure \ref{fig:thetaPE}.
 } \label{fig:relaxtheta}
-\end{figure*}
+\end{figure}
 
@@
   \caption{
-  Distance dependent portfolio effects as a function of straying rate $m$ and trait heritability $h^2$. When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases.
+  Portfolio effects as a function of straying rate $m$ and trait heritability $h^2$ when the rate of straying is $m = 0.5(1 + \Delta\theta)^{-1}$.
+Alternative steady states emerge for low values of $m$ (left of the cusp bifurcation, denoted by the black line), whereas a single steady state exists for high $m$.
   } \label{fig:mthetaPE}
 \end{figure}
+
+% \clearpage
 
@@
 \caption{
-Distance dependent straying, where increased differences in trait optima between sites $\Delta\theta$ corresponds to lower rates of straying $m$.
+Population inversion where increased differences in trait optima between sites $\Delta\theta$ corresponds to lower rates of straying $m$.
 At low rates of straying $m=0.02$ ($\Delta\theta=24$), extinction of the dominant population leads to slower-than-expected recovery times because the subordinate population is isolated enough to evolve towards its own trait optimum.
 %until growth of the dominant population overwhelms this local selection.
-In this case, $m$ is less than $m=0.034$ (denoted by the asterisk in figure \ref{extfig-fig:mtheta}), such that isolation allows the subdominant population to `run away' from the influence of the dominant population, leading to a switch in states.
-If $m$ is low but greater than $0.034$, isolation permits the subdominant population to `run away' from the influence of the dominant population, until it is overwhelmed by the recovering dominant population, and reverts back to its previous trait mean prior to the disturbance.
+In this case, $m$ is less than $m=0.034$ (denoted by the asterisk in figure \ref{fig:mtheta}), such that isolation allows the subordinate population to `run away' from the influence of the dominant population.
+This leads to a switch in subordinate/dominant states for the two populations.
+If $m$ is low but greater than $0.034$, isolation permits the subordinate population to `run away' from the influence of the dominant population, until it is overwhelmed by the recovering dominant population, and reverts back to its previous trait mean prior to disturbance.
 } \label{fig:inertia}
@@
   \caption{
-  Distance dependent recovery times for three disturbance types.
-When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases for constant (a) and density-dependent (b) staying rates.
-  The pitchfork bifurcation is not as clear in (b) because $\Delta\theta$ is a function of the individual straying rate $m_0$, whereas the x-axis in (b) is the straying rate at the steady state $m^*$.
-  Despite this difference, the general trends shown in (a) are also present in (b).
+  Recovery times for three disturbance types when the straying rate covaries with habitat heterogeneity as $m = 0.5(1 + \Delta\theta)^{-1}$ for constant (a) and density-dependent (b) straying rates.
+  The cusp bifurcation is not as clear in (b) because $\Delta\theta$ is a function of the individual straying rate $m_0$, whereas the x-axis in (b) is the straying rate at the steady state $m^*$.
+  Despite this difference, the general behavior shown in (a) are the same in (b).
   } \label{fig:mthetamvm}
-\end{figure*}
+\end{figure}
 
@@
 \caption{
-Steady state densities of both populations as a function of $m$, where a pitchfork bifurcation indicates the emergence of alternative steady states: one in a dominant state and one in a subordinate state.
+Steady state densities of both populations as a function of $m$, where a cusp bifurcation indicates the emergence of alternative steady states: one in a dominant state and one in a subordinate state.
 Steady states for populations with symmetrical values ($\alpha=0$) in the vital rates $r_{\rm max}$ and $\beta$ are shown with cool tones.
-As the asymmetry among populations between sites increases ($\alpha>0$), their vital rates diverge, such that the maximal growth at sites 1 and 2 is now $r_{\rm max,1}=r_{\rm max}(1+\tilde{rv}_1)$ and $r_{\rm max,2}=r_{\rm max}(1+\tilde{rv}_2)$ where $rv_{1,2}$ are independently drawn from $\rm{Normal}(0,\alpha)$ and $r_{\rm max}=2$.
-Similarly the strength of density dependence is calculated at sites 1 and 2 as $\beta_1=\beta(1+\tilde{rv}_1)$ and $\beta_2=\beta(1+\tilde{rv}_2)$ where $\tilde{rv}_{1,2}$ are independently drawn from $\rm{Normal}(0,\alpha)$ and $\beta=0.001$.
+As the asymmetry among populations between sites increases ($\alpha>0$), their vital rates diverge, such that the maximal growth at sites 1 and 2 is now $r_{\rm max}(1)=r_{\rm max}(1+\tilde{rv}_1)$ and $r_{\rm max}(2)=r_{\rm max}(1+\tilde{rv}_2)$ where $rv_{1,2}$ are independently drawn from $\rm{Normal}(0,\alpha)$ and $r_{\rm max}=2$.
+Similarly the strength of density dependence is calculated at sites 1 and 2 as $\beta(1)=\beta(1+\tilde{rv}_1)$ and $\beta(2)=\beta(1+\tilde{rv}_2)$ where $\tilde{rv}_{1,2}$ are independently drawn from $\rm{Normal}(0,\alpha)$ and $\beta=0.001$.
 Steady states for populations with increasingly asymmetric values ($\alpha\rightarrow 0.1$) for vital rates $r_{\rm max}$ and $\beta$ are shown in warmer tones.
@@
 \caption{
-Near collapse of both populations with a low straying rate $m=0.1$ and low trait heritability $h^2=0.2$ (see figure \ref{extfig-fig:relax}a).
+Near collapse of both populations with a low straying rate $m=0.1$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
 Black line marks the calculated point of recovery post-perturbation.
@@
 
-% \begin{figure*}[h!]
-%   \captionsetup{justification=raggedright,
-% singlelinecheck=false
-% }
-% \centering
-% \begin{subfigure}[t]{0.55\textwidth}
-% \centering
-% \includegraphics[width=\textwidth]{figs2/fig_MDPE_hm_theta3.pdf} 
-% \caption{Low habitat heterogeneity ($\Delta\theta=3$)} \label{fig:thetadiff1}
-% \end{subfigure}
-% \begin{subfigure}[t]{0.55\textwidth}
-% \centering
-% \includegraphics[width=\textwidth]{figs2/fig_MDPE_hm_theta8.pdf} 
-% \caption{High habitat heterogeneity ($\Delta\theta=8$)} \label{fig:thetadiff2}
-% \end{subfigure}
-% \caption{Total means $N_t$, difference in means $\Delta N$, and the portfolio effect PE for different habitat heterogeneities $\Delta\theta$. Light colors = high values.
-% }
-% \end{figure*}
-
-% \begin{figure*}[h!]
-%   \captionsetup{justification=raggedright,
-% singlelinecheck=false
-% }
-% \centering
-% \includegraphics[width=0.8\textwidth]{figs2/fig_MDPE_hm_ddm.pdf}
-% \caption{
-% The same simulations as presented in Figure \ref{fig:PE}, except with density-dependent straying, where $m(t) = m_0\left(1-N(t)/(C+N(t)\right)$. Light colors = high values.
-% } \label{fig:ddm}
-% \end{figure*}
-
 \end{document}

### 2017-09-19 model/src/KevinEvolveSS.jl

@@ -1,5 +1,6 @@
 function KevinEvolveSS(
   n0,
+  x0,
   tmax, 
   z, 
   rmax,
@@ -18,8 +19,10 @@ function KevinEvolveSS(
   n1 = zeros(Float64,tmax); n1[1]=n0[1];
   n2 = zeros(Float64,tmax); n2[1]=n0[2];
   
-  x1 = zeros(Float64,tmax); x1[1]=theta1 + rand(Normal(0,0.0001));
-  x2 = zeros(Float64,tmax); x2[1]=(theta1 + thetadiff)  + rand(Normal(0,0.0001));
+  # x1 = zeros(Float64,tmax); x1[1]=theta1 + rand(Normal(0,0.0001));
+  # x2 = zeros(Float64,tmax); x2[1]=(theta1 + thetadiff)  + rand(Normal(0,0.0001));
+  x1 = zeros(Float64,tmax); x1[1] = x0[1];
+  x2 = zeros(Float64,tmax); x2[1] = x0[1];
   
   w1 = zeros(Float64,tmax-1);
   w2 = zeros(Float64,tmax-1);

### 2017-09-20 Manuscript/FinalDraft_rev/ReviewResponses.tex
supplemental ext-fig

@@
 \opening{To the editorial committee,}
+
+We thank both the Reviewers for their insightful comments and criticisms.
+Below, we address each one, point by point.
+We believe that the edits that we have made have increased the clarity of our contribution, in no small part due to the thought and detail exercised by the Reviewers.
+We hope that our responses satisfy the concerns listed below.
+Please do not hesitate to contact us if there are additional clarifications required.
 

### 2017-09-20 Manuscript/FinalDraft_rev/supp.tex
supplemental ext-fig

@@
 (b) The same as (a) but including recovery times when straying is density-dependent evaluated at $m^*$, shown by linked point pairs.
-Recovery times for systems with density-dependent straying are longer when straying is low and shorter when straying is high, mirroring the change in portfolio effects with respect to density dependent straying shown in figure \ref{fig:thetaPE}.
+Recovery times for systems with density-dependent straying are longer when straying is low and shorter when straying is high, mirroring the change in portfolio effects with respect to density dependent straying shown in figure \ref{extfig-fig:thetaPE}.
 } \label{fig:relaxtheta}
@@
 %until growth of the dominant population overwhelms this local selection.
-In this case, $m$ is less than $m=0.034$ (denoted by the asterisk in figure \ref{fig:mtheta}), such that isolation allows the subordinate population to `run away' from the influence of the dominant population.
+In this case, $m$ is less than $m=0.034$ (denoted by the asterisk in figure \ref{extfig-fig:mtheta}), such that isolation allows the subordinate population to `run away' from the influence of the dominant population.
 This leads to a switch in subordinate/dominant states for the two populations.

### 2017-09-20 Manuscript/FinalDraft_rev/ReviewResponses.tex
peter edits integrated

@@
 \noindent We have made the following edits to the text:\\
-\noindent {\bf Line 115:} \emph{This model formulation has parallels to that proposed by Ronce and Kirkpatrick [ref]), where habitat specialization evolves between two populations as a function of dispersal.
-The largest difference between these approaches is that our framework treats trait evolution mechanistically (at some cost to analytical tractability).
-Importantly, we show that the resulting dynamics are qualitatively similar, suggesting that the dynamical features present in both of these approaches have potentially widespread ramifications for the evolutionary dynamics of connected populations.}\\
+\noindent {\bf Line 115:} \emph{This model formulation has parallels to that proposed by Ronce and Kirkpatrick [ref], where habitat specialization evolves between two populations as a function of dispersal, yet differs in that we treat trait evolution mechanistically at some cost to analytical tractability.}\\
 \noindent {\bf Line 191} \emph{These dynamics are also observed in the Ronce and Kirkpatrick model, which they describe as a transition from symmetric to asymmetric states [ref].}
+\noindent {\bf Line XXX} \emph{A salient finding from our results is that straying can lead to the emergence of alternative stable states, pushing one of the populations to high density (the \emph{dominant state}), and one to low density (the \emph{subordinate state}).
+This pattern has been observed in other models of eco-evolutionary dynamics with explicit space [ref], suggesting that it may be a general feature of spatially-linked populations that evolve toward local optima while being hindered by dispersal.}
 
@@
 We cannot alter this type of symmetry without vastly changing the questions that we are addressing.
-In general, although discrete cusp bifurcations with pitchfork-like behaviors are a consequence of symmetry (defined by a system where $N_1\rightarrow-N_1$, $N_2\rightarrow-N_2$, $x_1\rightarrow-x_1$, $x_2\rightarrow-x_2$ without altering the equations), the general features of the bifurcation need not be altered as symmetry is broken.
+In general, although discrete cusp bifurcations with pitchfork-like behaviors are a consequence of symmetry (defined by a system where $N_1\rightarrow-N_1$, $N_2\rightarrow-N_2$, $\mu_1\rightarrow-\mu_1$, $\mu_2\rightarrow-\mu_2$ without altering the equations), the general features of the bifurcation need not be altered as symmetry is broken.
 Consider the topological normal form $\dot{x}=h + \beta x - x^3$, which is a continuous system with a perfect pitchfork bifurcation at $\beta = 0,~h=0$.

### 2017-09-20 Manuscript/FinalDraft_rev/msv6.tex
peter edits integrated

--- 2017-09-13 Manuscript/FinalDraft_rev/msv6.tex
@@
 %\title{Community assembly and dynamics by the deconstruction of species interactions}
-\title{Eco-evolutionary dynamics and collective migration: implications for salmon metapopulation robustness}
+\title{Eco-evolutionary dynamics and collective dispersal: implications for salmon metapopulation robustness}
 \author{
@@
 Although the dispersal of individuals from donor to recipient populations provides connections within the metapopulation, promoting demographic and evolutionary rescue, it may also introduce maladapted individuals into habitats host to different environmental conditions, potentially lowering the fitness of the recipient population.
+Thus, dispersal plays a dual role in both promoting and inhibiting local adaptation.
 Here we explore a minimal model of the eco-evolutionary dynamics between two populations connected by dispersal, where the productivity of each is defined by a trait complex that is subject to local selection.
@@
 We show that intermediate levels of straying result in large gains in robustness, and that increased habitat heterogeneity promotes robustness when straying rates are low, and erodes robustness when straying rates are high.
-Finally, we show that density-dependent straying promotes robustness, particularly when the aggregate biomass is low and straying is correspondingly high, and this has important ramifications for the conservation of salmon metapopulations facing both natural and anthropogenic disturbances.
+Finally, we show that density-dependent straying promotes robustness, particularly when the aggregate biomass is low and straying is correspondingly high, which has important ramifications for the conservation of salmon metapopulations facing both natural and anthropogenic disturbances.
 \end{abstract}
@@
 \noindent {\bf Media Summary} The rate at which individuals move within a metapopulation has large impacts on extinction risk.
-Among migratory species such as salmon, the rate of dispersal, or straying, may be density dependent due to collective decision-making.
+Among migratory species such as salmon, the rate of dispersal, also referred to as straying, may be density dependent due to collective decision-making.
 We explore an eco-evolutionary model of two populations connected by dispersal, where the productivity of each is defined by a trait complex subject to local selection.
-We show that intermediate straying rates result in large gains in robustness, and that density-dependent straying promotes robustness.
+We show that intermediate straying rates result in faster recovery times and more stable population dynamics, and that density-dependent straying tends to promote this robustness.
 These results have important ramifications for the conservation of salmon metapopulations facing both natural and anthropogenic disturbances.\\
@@
 Intraspecific diversity can increase the resilience and stability of species or metapopulations. 
-This diversity-stability linkage can occur when there are asynchronous population dynamics, where the changes in population size varies temporally across the metapopulation. 
+This diversity-stability linkage occurs when there are asynchronous population dynamics, where the changes in population size varies temporally across the metapopulation. 
 This asynchrony will increase the potential for demographic rescue \citep{Brown:1977gk,Earn:2000fm} and also decrease the variability of processes that integrate across the metapopulation \citep{Anonymous:2015gf}.
@@
 Migratory populations that return to breeding sites for reproduction are linked to each other by some proportion of the population that permanently disperses into the `wrong' site. 
-Recently, the role of social interactions and collective navigation has been hypothesized \citep[][This issue]{Berdahl:2015kv,Berdahl:2016dx}.
-The rate at which individuals stray, $m$, is in this case synonymous with dispersal and may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \citep{Simons:2004jo,Berdahl:2015kv,Berdahl:2016dx}.
+Recently, the role of social interactions and collective navigation has been hypothesized \citep[][this volume]{Berdahl:2015kv,Berdahl:2016dx,HardestyMoore:wg}.
+The rate at which individuals disperse may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \citep{Simons:2004jo,Berdahl:2015kv,Berdahl:2016dx}.
 The potential influence of collective dispersal on the dynamics of individual populations and the metapopulation as a whole is a topic of considerable interest that has tangible conservation implications \citep{Brenner:2012gl,Johnson:2012fe,Fullerton:2011ii}.
@@
 The eco-evolutionary impacts of dispersal likely have important implications for conservation and management in key taxa such as in migratory salmon.
-While anadromous salmonid fishes (genera \emph{Oncorhynchus} and \emph{Salmo}) are renown for returning to their natal spawning habitats with high accuracy and precision after years at sea \citep{Quinn:2011tf,Jonsson:2011kg,Keefer:2014gg}, there are generally some individuals that `stray' to non-natal sites to spawn \citep{Quinn:1993ge,Hendry:2004wf}.
+While anadromous salmonid fishes (genera \emph{Oncorhynchus} and \emph{Salmo}) are renown for returning to their natal spawning habitats with high accuracy and precision after years at sea \citep{Quinn:2011tf,Jonsson:2011kg,Keefer:2014gg}, there are generally some individuals that `stray' (synonymously used hereafter to refer to dispersal) to non-natal sites to spawn \citep{Quinn:1993ge,Hendry:2004wf}.
 Salmon may operate as metapopulations, where populations are genetically distinct but linked by some level of straying \citep{Schtickzelle:2007wb,Anderson:2014cx}.
-Although extensive work has been done to document the extent of straying from donor populations and into recipient populations \citep{Keefer:2014gg,Bett:2017ha}, only recently have the abiotic, biotic, and anthropogenic influences of straying behaviors been investigated systemically \citep{Keefer:2008bs,Westley:2015to,Bond:2016dz}.
-Straying among salmon may be influenced by environmental factors such as water temperature, human activities such as hatchery practices, and population density as predicted by the collective migration hypothesis \citep{Peterson:2014gy}.
+Although extensive work has been done to document the extent of straying from donor populations into recipient populations \citep{Keefer:2014gg,Bett:2017ha}, only recently have the abiotic, biotic, and anthropogenic influences of straying behaviors been investigated systemically \citep{Keefer:2008bs,Westley:2015to,Bond:2016dz}.
+Straying among salmon may be influenced by environmental factors such as water temperature, human activities such as hatchery practices, and population density as predicted by the collective navigation hypothesis \citep{Peterson:2014gy,Berdahl:2017uu}.
 Straying can introduce new maladaptive genotypes into the recipient population, while the ensuing genetic homogenization could synchronize population dynamics and erode portfolio effects \citep{Moore:2010gs,Carlson:2011ce,Braun:2016ib}.
@@
 % Here we ask the overarching question: how does collective behavior mediated dispersal and gene flow interact to influence the robustness of locally adapted populations?
-To address this question we construct a minimal eco-evolutionary model of two populations occupying different sites that are linked by straying individuals, each with an associated trait distribution subject to natural selection determined by local conditions.
-Specifically we compare (a) different rates of straying and (b) the influence of collective movement, across (c) increasing environmental heterogeneity, by assessing two measures of metapopulation robustness: the portfolio effect and the time required for a population(s) to recover following an induced disturbance.
+To address this question we constructed a minimal eco-evolutionary model of two populations occupying different sites that are linked by straying individuals, each with an associated trait distribution subject to natural selection determined by local conditions.
+Specifically we compared (a) different rates of straying and (b) the influence of collective movement, across (c) increasing environmental heterogeneity, by assessing two measures of metapopulation robustness: the portfolio effect and the time required for a population(s) to recover following an induced disturbance.
 % This model enables us to explore the multiple and potentially opposing pathways by which straying influences metapopulation robustness such as the potentially detrimental erosion of local adaptation vs.
@@
 \noindent{\bf (a) Metapopulation framework}\\
-\noindent We consider two populations $N_1$ and $N_2$ that inhabit two distinct habitats, each with trait values $x_1$ and $x_2$ determining recruitment rates.
-We assume that there is an optimum trait value $\theta_1$ and $\theta_2$ associated with each habitat, where recruitment is maximized if the trait value of the local population equals the optimum, such that $x = \theta$.
-Moreover, we assume that $x_{1,2}$ are normally distributed with means $\mu_1$ and $\mu_2$ and have the same standard deviation $\sigma$.
+\noindent We considered two populations $N_1$ and $N_2$ that inhabit two distinct habitats, each with trait values $x_1$ and $x_2$ determining recruitment rates.
+We assumed that there is an optimum trait value $\theta_1$ and $\theta_2$ associated with each habitat, where recruitment is maximized if the trait value of the local population equals the optimum, such that $x = \theta$.
+Moreover, we assumed that $x_{1,2}$ are normally distributed with means $\mu_1$ and $\mu_2$ and have the same standard deviation $\sigma$.
 As such, the recruitment rate $R_1[\mu_1(t),\theta_1]$ for both populations is determined by the mean trait value of the local population relative to optimal value at that site.
@@
 The two populations occur in spatially separate sites that are close enough that a proportion of the population $m$ can stray into the other site, and where mortality occurs before individuals return to reproduce.
-If there is no straying between these populations, then the mean trait evolves towards the optimal value for each site $x_1 \rightarrow \theta_1$, and the recruitment rate for each population is maximized.
+If there is no straying between these populations, then the mean trait evolves towards the optimal value for each site $\mu_1 \rightarrow \theta_1$, and the recruitment rate for each population is maximized.
 If there is straying between populations at rate $m$, then the traits in each respective location will be pulled away from the optimum, and recruitment rates will be lowered.
@@
 
-We use the discrete Ricker framework described by Shelton and Mangel \citep{Shelton:2011eq} as the basis for our two-site metapopulation model, with the added effect of the local population $N_i$ mixing with a set proportion $m$ of a remote population $N_j$ that is straying into it.
+We used the discrete Ricker framework described by Shelton and Mangel \citep{Shelton:2011eq} as the basis for our two-site metapopulation model, with the added effect of the local population $N_i$ mixing with a set proportion $m$ of a remote population $N_j$ that is straying into it.
 In this sense, both populations serve simultaneously as donor and recipient populations.
-We first assume that a proportion ${\rm e}^{-Z}$ of both populations survive such that the surviving aggregated population, composed of both local individuals (at site $i$) and incoming strays (from site $j$), is $\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z}$.
+We first assumed that a proportion ${\rm e}^{-Z}$ of both populations survive such that the surviving aggregated population, composed of both local individuals (at site $i$) and incoming strays (from site $j$), is $\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z}$.
 Because local individuals will recruit differently than incoming strays, the recruitment of the aggregate must incorporate two recruitment functions, given by $\left(R_i[\mu_i(t)] (1-m)N_i(t) + R_i[\mu_j(t)] m N_j(t)\right)$.
@@
 % \noindent{\bf (b) Recruitment over a selective landscape}
-\noindent Because individuals from the local population are mixed with individuals from the remote population via straying and subsequent reproduction, the resulting trait distribution is a mixed normal with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and straying individuals, $1-w_i$, where
+Because individuals from the local population are mixed with individuals from the remote population via straying and subsequent reproduction, the resulting trait distribution is a mixed normal with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and straying individuals, $1-w_i$, where
 \begin{equation}
@@
 \end{equation}
-We make two simplifying assumptions.
-First, we assume that the distribution resulting from the mix of remote and local individuals, following reproduction, is also normal with a mean value equal to that of the mean for the mixed-normal distribution.
-Thus, we assume that strays can successfully reproduce and introduce their genotypes into the recipient population, which is supported by observations in wild populations \citep{Jasper:2013cc}.
-Second, we assume that changes in trait variance through time are minimal, such that $\sigma^2$ is assumed to be constant, which is a common simplification in eco-evolutionary models of population dynamics \citep{Lande:1976ga,Schreiber:2011wx,Gilbert:2014ee,Gibert:2015kc}.
+We made two simplifying assumptions.
+First, we assumed that the distribution resulting from the mix of remote and local individuals, following reproduction, is also normal with a mean value equal to that of the mean for the mixed-normal distribution.
+Thus, strays can successfully reproduce and introduce their genotypes into the recipient population, which is supported by observations in wild populations \citep{Jasper:2013cc}.
+Second, we assumed that changes in trait variance through time are minimal, such that $\sigma^2$ is constant over time, which is a common simplification in eco-evolutionary models of population dynamics \citep{Lande:1976ga,Schreiber:2011wx,Gilbert:2014ee,Gibert:2015kc}.
 
@@
 %Comparison to the Ronce and Kirkpatrick paper
-This model formulation has parallels to that proposed by Ronce and Kirkpatrick \citep{Ronce:2001dp}, where habitat specialization evolves between two populations as a function of dispersal.
-The largest difference between these approaches is that our framework treats trait evolution mechanistically at some cost to analytical tractability.
-Importantly, we show that the resulting dynamics are qualitatively similar, suggesting that the dynamical features present in both of these approaches have potentially widespread ramifications for the evolutionary dynamics of spatially connected populations, regardless of the specific formulation proposed.
+This model formulation has parallels to that proposed by Ronce and Kirkpatrick \citep{Ronce:2001dp}, where habitat specialization evolves between two populations as a function of dispersal, yet differs in that we treat trait evolution mechanistically at some cost to analytical tractability.
 \\
@@
 %Density dependent m
-\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is mounting evidence that at least in some species (including salmon) the straying rate is density-dependent, a signature of collective navigation \citep{Berdahl:2016dx,Bett:2017ha}.
+\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is mounting evidence that at least in some species (including salmon) the straying rate is density-dependent, a signature of collective navigation \citep{Berdahl:2014bl,Berdahl:2017uu}.
 Specifically, the rate at which individuals stray has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tend to decrease the rate at which individuals err, reducing the overall proportion of a population that strays.
-According to Berdahl et al. \citep{Berdahl:2014bl,Berdahl:2016dx}, given the probability that an individual strays is $m_0$, the proportion of the local population $N_i(t)$ that strays is
+According to Berdahl et al. \citep{Berdahl:2016dx}, given the probability that an individual strays is $m_0$, the proportion of the local population $N_i(t)$ that strays is
 
@@
 Habitat heterogeneity and the rate of straying are treated both independently, and as parameters that covary.
-In the latter instance, we evaluate a case where it is assumed that increased habitat heterogeneity correlates with lower straying rates, and vice versa (illustrated in figure \ref{fig:mthetarelation}).
+In the latter instance, we examined a case where it is assumed that increased habitat heterogeneity correlates with lower straying rates, and vice versa (illustrated in figure \ref{fig:mthetarelation}).
 Two scenarios may lead to this correlation: 
@@
 \noindent {\bf (d) Measuring metapopulation robustness}
-\noindent We evaluated metapopulation robustness by measuring the average-CV portfolio effect (PE) \citep{Anderson:2014cx,Schindler:2015gf} as well as the time required for the system to return to a steady state following an induced disturbance to one or both of the populations \citep{Ovaskainen:2002il}.
+\noindent We evaluated metapopulation robustness by measuring the average-CV portfolio effect (PE) \citep{Anderson:2014cx,Schindler:2015gf} as well as the recovery time, which is the time required for the system to return to a steady state following an induced disturbance to one or both of the populations \citep{Ovaskainen:2002il}.
+Throughout, we refer to an increase in portfolio effects and/or reduction in recovery time as promoting metapopulation robustness.
+
 The average-CV portfolio effect is, as the name implies, the average CV across each population $N_i$ divided by the CV of the aggregate $N_T=\sum_i N_i$ \citep{Anderson:2013gb}, such that
@@
 A more direct way to measure system robustness is to measure the time that the system (measured as the aggregate steady state biomass $N_T^*$) takes to return to a steady state following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
-Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation \citep{GuckHolmes}, because we aimed to 1) assess the effects of a large perturbation, and 2) estimate the time required for all transient effects to decay (including dampened oscillations), we used a simulation-based numerical procedure.
+Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation \citep{GuckHolmes}, because we aimed to 
+1) assess the effects of a large perturbation far from the steady state, and 
+2) estimate the time required for all transient effects to decay following this perturbation (including dampened oscillations), we used a simulation-based numerical procedure.
 Recovery time was calculated by initiating a disturbance at $t=t_d$, and monitoring $N_T(t_d+t)$ as $t\rightarrow T$, where $T$ is large. 
@@
 
-Numerically estimating the time that it takes for a perturbed system to recover also permits a more detailed perspective of metapopulation fragility.
-For example, if populations settle to alternative stable states, comparing recovery times after a disturbance applied to individual populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery.
+Numerically estimating the time that it takes for a perturbed system to recover also permits a more detailed perspective of metapopulation robustness.
+For example, if populations settle to alternative stable states (one at low- and one at high-density), comparing recovery times after a disturbance applied to individual populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery.
 % (alternative stable states in our model requiring one population to be high-density and one low-density)
 We measured recovery time following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} are equivalent if populations have the same density); (\emph{iii}) near-collapse of both populations where just 1.0\% of each survives.
-Throughout, we will refer to an increase in the portfolio effects and/or reduction in recovery times as promoting metapopulation robustness and by extension persistence.
 \\
@@
 Straying had non-linear impacts on the recovery time of populations. 
-When the dominant state (well adapted and high density) is wiped out, high levels of straying allow it to recover quickly (figure \ref{fig:relax}a) because the surviving population has a mean trait value skewed towards the optimum of the recovering population (figure \ref{fig:relaxtraj_hdlh}).
-Yet, as straying decreases, recovery time for the disturbed dominant state population increases, in part because there is enough time for the trait distribution to move back towards the trait optimum of the subordinate state population.
+When the dominant state population (well adapted and high density) goes extinct, high rates of straying allow recolonization of the extirpated habitat and quick recovery (figure \ref{fig:relax}a) because the surviving population has a mean trait value skewed towards the optimum of the affected habitat (figure \ref{fig:relaxtraj_hdlh}).
+Yet, as straying decreases, recovery time for the disturbed dominant state population increases, in part because there is time for the trait distribution to move back towards the trait optimum of the subordinate state population.
 In contrast, when the subordinate state population (maladapted and low density) is wiped out, recovery rates are fastest at low to intermediate levels of straying.
 Because the mean trait values of both populations are skewed towards those of the dominant population, when the subordinate population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into the recovering population extends the length of time required for it to return to its steady state (figure \ref{fig:relaxtraj_ldlh}).
-When both populations are both dramatically reduced, recovery time is generally fastest at lower levels of straying, while near the onset of the pitchfork bifurcation, recovery time increases explosively and this is -- as the name implies -- characteristic of \emph{slowing} dynamics that occur near critical transitions \citep{Scheffer:2009gg,Kuehn:2010p2591}.\\
-
-\noindent{\bf (b) The effects of collective migration and density dependent straying} \\
+When both populations are both dramatically reduced, recovery time is generally fastest at lower levels of straying due to reduced mixing of maladaptive phenotypes.
+Near the onset of the cusp bifurcation, recovery time increases explosively, however this is -- as the name implies -- characteristic of \emph{slow} dynamics occurring near critical transitions \citep{Scheffer:2009gg,Kuehn:2010p2591}.
+
+%Lowering Phenotypic diversity with increasing straying rate
+Increased rates of straying lowers phenotypic diversity ($\Delta \mu^* = |\mu_i^*-\mu_j^*|$, evaluated at the steady state) because both local and remote populations are increasingly homogenized.
+The loss of phenotypic diversity is greater with increased straying when trait heritability is low because traits take longer to go back to their local optima than they do when heritability is high. 
+Hence straying counters the effect of diversifying local adaptation. 
+% The loss of phenotypic diversity with increased straying is greater if trait heritability is low because the selective forces acting against trait means far from the local optima are lessened.
+% with decreased $h^2$.
+Less intuitively, we observe a discrete jump towards low phenotypic diversity as the cusp bifurcation is crossed (figure \ref{fig:traitdiff}).
+Although the development of alternative stable states elevates the portfolio effect due to the variance-dampening effects of the aggregate, entering this dynamic regime also results in a substantial decline in phenotypic diversity, which may have less predictable adverse effects on the population.\\
+
+\noindent{\bf (b) The effects of collective navigation and density dependent straying} \\
 %Portfolio effects
@@
 % As such, the difference in $m^*$ between subordinate and dominant populations appears to matter less than the lower straying rate generated by the dominant population.
-As $m^*$ increases, we observe an increase in the PE than for systems with constant $m$.
+As $m^*$ increases, we observe an increase in the PE relative to systems with constant $m$.
 
 %Recovery times
-Density-dependent straying alters these recovery times (figure \ref{fig:relax}b). 
-First, in comparison with constant stray rates, density-dependent straying made recovery more rapid at elevated stray rates when both populations collapsed and when the subordinate population was extirpated.
+Density-dependent straying alters these recovery times (figure \ref{fig:relax}b; note difference in x-axis scale). 
+First, in comparison with constant stray rates, density-dependent straying lowered recovery times at elevated straying rates for near-collapse of both populations and extirpation of the subordinate population.
 %Recovery times
 %The recovery times for systems with constant and density-dependent straying both show similar trends for a given $m$ and $m^*$ (figure \ref{fig:relax}c,d), however there are some distinct quantitative differences.
-At low straying rates, near-collapse of both populations resulted in longer than expected recovery times, whereas in the alternative stable state regime (higher $m^*$), the recovery times for different disturbance types were very similar to systems with a constant $m$ (figure \ref{fig:relax}b; note difference in x-axis scales).
-As trait heritability increased, the metapopulation always recovered more quickly if the small population was lost (figure \ref{fig:relax_highh}).
+In contrast, at low straying rates, near-collapse of both populations required more time to recover.
+%whereas in the alternative stable state regime (higher $m^*$), the recovery times for different disturbance types were very similar to systems with a constant $m$ (figure \ref{fig:relax}b; note difference in x-axis scales).
+Trait heritability had a large effect on recovery times, with near-collapse requiring a more protracted time to recover in the alternative steady state regime (figure \ref{fig:relax_highh}).
+% As trait heritability increased, the metapopulation always recovered more quickly if the subordinate population was lost (figure \ref{fig:relax_highh}).
 % There was greater similarity in the recovery times for constant and density-dependent $m$ systems when trait heritability was high (figure \ref{fig:relax_highh}).
-The lower recovery time for systems with increased $m^*$ mirrors an elevated PE with higher density-dependent straying rates (figure \ref{fig:thetaPE}).
+In general, the lower recovery time for systems with increased $m^*$ mirrors an elevated PE with higher density-dependent straying rates (figure \ref{fig:thetaPE}a).
 % This higher PE at increased $m^*$ mirrors recovery times that are also lower than those observed for systems with constant $m$ (figure \ref{fig:relax}).
-In tandem, analysis of both PE and recovery time suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our minimal model, it does appear to promote robustness, particularly when the aggregate biomass is low and straying is correspondingly high.
-
-%Lowering Phenotypic diversity with increasing straying rate
-Increased rates of straying lowers phenotypic diversity ($\Delta \mu^* = |\mu_i^*-\mu_j^*|$, evaluated at the steady state) because both local and remote populations are increasingly homogenized.
-The loss of phenotypic diversity with increased straying is greater if trait heritability is low because traits take longer to go back to their local optima than they do when heritability is large. 
-Hence straying counters the effect of diversifying local adaptation. 
-% The loss of phenotypic diversity with increased straying is greater if trait heritability is low because the selective forces acting against trait means far from the local optima are lessened.
-% with decreased $h^2$.
-Less intuitively, we observe a discrete jump towards low phenotypic diversity as the pitchfork bifurcation is crossed (figure \ref{fig:traitdiff}).
-Although the development of alternative stable states elevates the portfolio effect due to the variance-dampening effects of the aggregate, entering this dynamic regime also results in a substantial decline in phenotypic diversity, which may have less predictable adverse effects on the population.
+Together, analysis of both PE and recovery time suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our minimal model, it does appear to promote robustness, particularly in the case of near-collapse of both populations when straying is high.
 \\
@@
 A salient finding from our results is that straying can lead to the emergence of alternative stable states, pushing one of the populations to high density (the \emph{dominant state}), and one to low density (the \emph{subordinate state}).
-This pattern has been observed in other models of eco-evolutionary dynamics with explicit space \citep{Ronce:2001dp}, suggesting that it may be a general feature of spatially-linked traits that evolve toward local optima while being hindered by dispersal.
-An important quality of our framework is that there are similar forces that dictate interactions within and between sites, and this naturally results in a symmetry that could preclude the relevance of our finding for natural (and inherently less symmetric) systems.
-Although the emergence of alternative stable states via a pitchfork bifurcation is characteristic of symmetrical systems, we find that increasing the asymmetry in the vital rates of populations across sites does not alter the presence or position of the PFB (figure \ref{fig:symmetry}).
-That these patterns arise in alternative formulations and are insensitive to parameter asymmetry suggests that the emergence of dominant and subordinate states directly from eco-evolutionary dynamics may be a generally relevant phenomenon in ecological systems.
-%, though this requires additional investigation and scrutiny.
+This pattern has been observed in other models of eco-evolutionary dynamics with explicit space \citep{Ronce:2001dp}, suggesting that it may be a general feature of spatially-linked populations that evolve toward local optima while being hindered by dispersal.
+An important aspect of our framework is that there are similar forces that dictate interactions within and between sites, and this naturally results in a symmetry that could be perceived as limiting the relevance of our findings for natural (and inherently less symmetric) systems.
+Although the emergence of alternative stable states via a cusp bifurcation is characteristic of symmetrical systems, we find that increasing the asymmetry in the vital rates of populations across sites does not significantly alter the presence or position of the DCB (figure \ref{fig:symmetry}).
+That these patterns arise in alternative formulations and are relatively insensitive to parameter asymmetry suggests that the dynamical features observed have potentially widespread ramifications for the evolutionary dynamics of spatially connected populations in general.
+% suggests that the emergence of dominant and subordinate states directly from eco-evolutionary dynamics may be a generally relevant phenomenon in ecological systems.
+% Importantly, we show that the resulting dynamics are qualitatively similar, suggesting that the dynamical features present in both of these approaches have potentially widespread ramifications for the evolutionary dynamics of spatially connected populations, regardless of the specific formulation implemented.
+<%, though this requires additional investigation and scrutiny.
 
@@
 
-An intermediate straying rate maximizes metapopulation robustness. 
+An intermediate straying rate increases metapopulation robustness. 
 Results from our model reveal that the presence of just enough straying to cause formation of alternative stable states increases the portfolio effect (figure \ref{fig:PE}a). 
+We note that we do not consider the extremely high PE at the DCB, matched by an extremely long recovery time, to be an indicator of robustness, as the dynamics exactly at or very close to a bifurcation are unlikely to be realized in nature.
 Previous theoretical work has shown that increased connectivity may erode portfolio effects in herring metapopulations, where straying is also thought to be density-dependent \citep{Secor:2009ena}.
-Although high levels of dispersal in our system generally supports this finding, the interplay between dispersal and PE is more subtle when selection for local adaptations is considered, such that low to intermediate levels of density-dependent straying result in an elevated PE, thus increasing the buffering capacity of the metapopulation.
+Although high levels of dispersal in our system generally supports this finding, the interplay between dispersal and PE is more subtle when selection for local adaptations is considered.
+Low to intermediate levels of straying result in an elevated PE, increasing the buffering capacity of the metapopulation.
+
 Although PE is measured at the steady state, low to intermediate rates of straying also appear to have a beneficial effect on transient dynamics.
-When there is just enough straying to cause alternative stable states, the time to recovery following an induced disturbance declines, though -- as with the PE -- recovery time then grows if straying becomes very large (figure \ref{fig:relax}a).
+When there is just enough straying to cause alternative stable states, the time to recovery following an induced disturbance declines, though -- as with the PE -- it then grows if straying becomes large (figure \ref{fig:relax}a).
+In the alternative steady state regime, a lower rate of straying inhibits admixture of maladapted individuals.
+Following a large disturbance, such as the near-collapse of both dominant and subordinate populations, this limited mixing increases the growth rates of both populations, permitting faster recovery times.
+If the rate of straying becomes too high, an influx of maladapted individuals into both populations inhibits local growth rates, and recovery slows.
 
@@
 %mtheta
-All else being equal, habitats that are closer in space generally have greater similarity in environmental conditions than habitats that are geographically distant, and correspondingly phenotypes of more proximately located populations should be more similar to those that are distant \citep{Westley:2012ui}.
+In other cases, habitats that are closer in space can be assumed to have greater similarity in environmental conditions than those that are geographically distant, and phenotypes of more proximately located populations should be more similar \citep{Reisenbichler:1988ex,Fraser:2011co,Westley:2012ui}.
 It is thus reasonable to expect a larger number of straying individuals between sites that are geographically proximate and indeed evidence corroborates this prediction \citep{Candy:2000hu,JPE:JPE1383}.
-Alternatively, salmon that cue to specific environmental conditions may be more likely to stray into sites that have structurally and physiognamically similar habitats \citep{Peterson:2014gy} even if other potential sites are closer.
-These considerations justify imposing a direct relationship between the rate of straying $m$ and habitat heterogeneity: as site dissimilarity increases, so too should the straying rate decrease.
+Alternatively, salmon that cue to specific environmental conditions may be more likely to stray into sites that are structurally and physiognamically more similar \citep{Peterson:2014gy}.
+These considerations justify imposing a direct relationship between the rate of straying $m$ and habitat heterogeneity: as site dissimilarity increases, so too should the straying rate decrease (figure \ref{fig:mthetarelation}).
 % As site dissimilarity increases, so to should the optima in trait values for the resident populations.
@@
 We show that there is an intermediate straying rate where disturbed populations that are recovering by the introduction of maladapted strays recover fastest: if the straying rate is too low or too high, recovery times increase (figure \ref{fig:relax}).
-This finding suggests that salmon stocking efforts that aim to lower recovery times following dam removal could actually prolong recovery if the rate at which individuals are introduced is not taken into account.
+This finding suggests that salmon stocking efforts that aim to lower recovery times following dam removal could actually prolong recovery if the rate at which individuals are introduced and the suitability of those fish in that habitat (i.e.
+their measure of pre-adaptation) is not taken into account.
 Ongoing examinations of experimental restocking in the recently re-opened Elwha River (Washington State) will provide empirical insight into the potential short- and long-term consequences of facilitated recovery \citep{Liermann:2017gj}.
@@
 Importantly, density-dependent straying, which may occur when individuals collectively navigate, can both increase the portfolio effect and lower the time to recovery following a disturbance, which is anticipated to promote persistence.
+Therefore, preserving the biological processes that facilitate this collective behavior may be an important conservation target in its own right, echoing the sentiments of  Hardesty-Moore et al.
+\citep{HardestyMoore:wg}.
 We suggest that understanding the spatial complexity of metapopulations dispersing across heterogeneous environments, in tandem with the mosaic of selective forces acting on those environments, may be key to uncovering those factors that promote persistence in the wild.
@@
 % 
+% \bibitem{Ronce:2001dp}
+% Ronce O, Kirkpatrick M, 2001 {When sources become sinks: migrational meltdown
+%   in heterogeneous habitats}.
+% \newblock \emph{Evolution} \textbf{55}, 1520
+% 
 % \bibitem{Berdahl:2014bl}
@@
 % 
+% \bibitem{AleksandrovichKuznetsov:1995p2580}
+% Kuznetsov Y, 1998 \emph{{Elements of Applied Bifurcation Theory}}.
+% \newblock New York: Springer
+% 
 % \bibitem{Carlson:2008hl}
@@
 % \newblock \emph{Ecol. Lett.} \textbf{16}, 421--429
-% 
-% \bibitem{Levin:1992p1472}
-% Levin SA, 1992 {The problem of pattern and scale in ecology}.
-% \newblock \emph{Ecology} \textbf{73}, 1943--1967
-% 
-% \bibitem{Meinhardt:2012hw}
-% Meinhardt H, 2012 {Turing's theory of morphogenesis of 1952 and the subsequent
-%   discovery of the crucial role of local self-enhancement and long-range
-%   inhibition}.
-% \newblock \emph{Interface Focus} \textbf{2}, 407--416
-% 
-% \bibitem{Hanski:1999vu}
-% Hanski I, 1999 \emph{{Metapopulation Ecology}}.
-% \newblock Oxford: Oxford University Press
-% 
-% \bibitem{Boughton:1999fa}
-% Boughton DA, 1999 {Empirical evidence for complex source{\textendash}sink
-%   dynamics with alternative states in a butterfly metapopulation}.
-% \newblock \emph{Ecology} \textbf{80}, 2727--2739
 % 
@@
 % \end{thebibliography}
+%
 
@@
 \caption{
-(a) The steady state densities of $N_1$ and $N_2$ as a function of a constant stray rate $m$.
-Which population attains the low- or high-density state is random due to small applied fluctuations in the initial conditions.
-(b) The steady state trait values measured as $\theta_i - x_i$, as a function of a constant stray rate $m$. 
-PFB marks the pitchfork bifurcation.
+(a) The steady state densities of $N_1$ and $N_2$ vs. the stray rate $m$. Which population attains the low- or high-density state is random due to small applied fluctuations in the initial conditions.
+(b) The steady state trait values measured as the offset from the local optimum $\theta_i - mu_i$, vs. the stray rate $m$. 
+DCB marks the discrete cusp bifurcation.
 Unless otherwise indicated, the default parameter values used are: $r_{\rm max}=2$; $Z=0.5$; $\beta=0.001$; $\theta_1=5$; $\Delta\theta=5$; $\tau=1$; $\sigma=1$; $T=1\times10^5$.

### 2017-09-20 model/Hysteresis.jl

@@ -7,7 +7,7 @@ include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinJacobian.
 
 #Analysis over m
 tmax=10000;
-mvec1 = collect(0.0:0.0001:0.1);
+mvec1 = collect(0.0:0.0001:0.15);
 mvec = [mvec1 ; reverse(mvec1)];
 n1ts = zeros(Float64,length(mvec),tmax);
 n2ts = zeros(Float64,length(mvec),tmax);
@@ -109,18 +109,18 @@ burnin=0.80
   
 end
 midpoint = Int64(floor(length(n1mean)/2));
-namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/figs2/fig_hysteresis.pdf");
+namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/FinalDraft_rev/fig_hysteresis.pdf");
 R"""
 library(RColorBrewer)
 cols = brewer.pal(3,'Set1')
-#pdf($namespace,height=5,width=6)
+pdf($namespace,height=5,width=6)
 plot($(mvec[1:midpoint]),$(n1mean[1:midpoint]),col=cols[1],xlab="Straying rate m",ylab="Steady state biomass",ylim=c(1,1500),type='l')
 lines($(mvec[1:midpoint]),$(n2mean[1:midpoint]),col=cols[2])
 lines($(mvec[midpoint+1:length(mvec)]),$(n1mean[midpoint+1:length(n1mean)]),col=cols[1],lty=2)
 lines($(mvec[midpoint+1:length(mvec)]),$(n2mean[midpoint+1:length(n2mean)]),col=cols[2],lty=2)
 types = c('Increasing m','Decreasing m')
-legend(x=0.15,y=1500,legend=types,col='black',lty=c(1,2),xpd=TRUE,cex=0.9, bty="n")
-#dev.off()
+legend(x=0.1,y=1500,legend=types,col='black',lty=c(1,2),xpd=TRUE,cex=0.9, bty="n")
+dev.off()
 """
 
 

### 2017-09-23 Manuscript/FinalDraft_rev/ReviewResponses.tex
final edits to first round of reviews

@@
     England\\
-    \centerline{\bf{Re: Eco-evolutionary dynamics and collective migration:}}\\
+    \centerline{\bf{Re: Eco-evolutionary dynamics and collective dispersal:}}\\
      \centerline{\bf{implications for salmon metapopulation robustness}}
@@
 \noindent To clarify, we have made the following edits to the text:\\
-\noindent {\bf Line 113:} \emph{The partial derivative in Eq.
+\noindent {\bf Line 111:} \emph{The partial derivative in Eq.
 4 determines how the mean trait changes through time due to natural selection [ref], which is proportional to the change in mean fitness with respect to $\mu_i$.
@@
 \noindent We have made the following edits to the text:\\
-\noindent {\bf Line 115:} \emph{This model formulation has parallels to that proposed by Ronce and Kirkpatrick [ref], where habitat specialization evolves between two populations as a function of dispersal, yet differs in that we treat trait evolution mechanistically at some cost to analytical tractability.}\\
-\noindent {\bf Line 191} \emph{These dynamics are also observed in the Ronce and Kirkpatrick model, which they describe as a transition from symmetric to asymmetric states [ref].}
-\noindent {\bf Line XXX} \emph{A salient finding from our results is that straying can lead to the emergence of alternative stable states, pushing one of the populations to high density (the \emph{dominant state}), and one to low density (the \emph{subordinate state}).
+\noindent {\bf Line 113:} \emph{This model formulation has parallels to that proposed by Ronce and Kirkpatrick [ref], where habitat specialization evolves between two populations as a function of dispersal, yet differs in that we treat trait evolution mechanistically at some cost to analytical tractability.}\\
+\noindent {\bf Line 187} \emph{These dynamics are also observed in the Ronce and Kirkpatrick model, which they describe as a transition from symmetric to asymmetric states [ref].}
+\noindent {\bf Line 299} \emph{A salient finding from our results is that straying can lead to the emergence of alternative stable states, pushing one of the populations to high density (the \emph{dominant state}), and one to low density (the \emph{subordinate state}).
 This pattern has been observed in other models of eco-evolutionary dynamics with explicit space [ref], suggesting that it may be a general feature of spatially-linked populations that evolve toward local optima while being hindered by dispersal.}
@@
 We now include a new figure S3 that shows that this is indeed the case for our system, and this was the basis on which we initially identified the bifurcation. 
-Although we are not mathematicians, our understanding is that the fold bifurcation is a general attribute based on this eigenvalue condition, but as the Reviewers indicate, it is not a particularly clear or complete description of what is going on.
+Our understanding is that the fold bifurcation is a general attribute based on this eigenvalue condition, but as the Reviewers indicate, it is not a particularly clear or complete description of what is going on.
 A familiar, and more complicated bifurcation that occurs in maps with $\geq2$ parameters is the Discrete Cusp Bifurcation, which occurs when two fold bifurcations intersect at a cusp.
@@
 The discrete cusp bifurcation is characterized by the intersection of two branches of a fold bifurcation, of which the topological normal form is $x \mapsto f(x) = x + \beta_1 + \beta_2 x - x^3$.
-The eigenvalue evaluated at the steady state $\partial f(x)/\partial x|_{x=x^*}$ increases to unity on the unit circle at the cusp bifurcation $(\beta_1 = 0, \beta_2 = 0)$.
+The eigenvalue evaluated at the steady-state $\partial f(x)/\partial x|_{x=x^*}$ increases to unity on the unit circle at the cusp bifurcation $(\beta_1 = 0, \beta_2 = 0)$.
 
 \noindent We have made the following edits to the text:\\
-\noindent {\bf Line 115:} \emph{The decline in steady state densities is not gradual: as straying increases, the system crosses a discrete cusp bifurcation (DCB) [ref] whereby the single steady state for the metapopulation bifurcates into two basins of attraction: one at high biomass, and one at low biomass density (figure 1a, 2a).
+\noindent {\bf Line 176:} \emph{The decline in steady-state densities is not gradual: as straying increases, the system crosses a discrete cusp bifurcation (DCB) [ref] whereby the single steady-state for the metapopulation bifurcates into two basins of attraction: one at high biomass, and one at low biomass density (figure 1a, 2a).
 Mean trait values for both populations bifurcate similarly (figure 1b). 
 In discrete systems, the cusp bifurcation is defined by two fold bifurcations intersecting at a cusp point [ref], and is observed when the real part of the dominant eigenvalue of the Jacobian matrix crosses the unit circle at +1 (figure S3).
-Visually, the dynamics are similar to those observed at a pitchfork bifurcation in continuous systems, where a single steady state gives rise to two steady states separated by an unstable fixed point.
+Visually, the dynamics are similar to those observed at a pitchfork bifurcation in continuous systems, where a single steady-state gives rise to two alternative steady-states.
 }
@@
 Moreover, I think that near the bifurcation point total population size will equilibrate fast, but the distribution of individuals over the two habitats takes a longer time to equilibrate (the system is on the edge between a symmetric and an asymmetric equilibrium).
-If time to recovery includes the time that $N_1$ and $N_2$ individually reach their steady states, then this can explain the peak of recovery time at the bifurcation; but this does not mean that the total population size takes long to recover.
+If time to recovery includes the time that $N_1$ and $N_2$ individually reach their steady-states, then this can explain the peak of recovery time at the bifurcation; but this does not mean that the total population size takes long to recover.
 If total population size was monitored, but recovery was interpreted as "return to original" whereas the system attains a different equilibrium (symmetric vs asymmetric, see above), then obviously recovery time will be infinite, but again this does not mean that the population does not recover from near extinction.
@@
 \noindent We have made the following edits to the text:\\
-\noindent {\bf Line 160:} \emph{Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation [ref], because we aimed to 1) assess the effects of a large perturbation, and 2) estimate the time required for all transient effects to decay (including dampened oscillations), we used a simulation-based numerical procedure.
+\noindent {\bf Line 154:} \emph{Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation [ref], because we aimed to 1) assess the effects of a large perturbation, and 2) estimate the time required for all transient effects to decay (including dampened oscillations), we used a simulation-based numerical procedure.
 Recovery time was calculated by initiating a disturbance at $t=t_d$, and monitoring $N_T(t_d+t)$ as $t\rightarrow T$, where $T$ is large. 
@@
 \noindent We have made the following edits to the text:\\
-\noindent {\bf Line 42:} \emph{The rate at which individuals stray, $m$, is in this case synonymous with dispersal and may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices [refs].}\\
+\noindent {\bf Line 45:} \emph{While anadromous salmonid fishes (genera \emph{Oncorhynchus} and \emph{Salmo}) are renown for returning to their natal spawning habitats with high accuracy and precision after years at sea [refs], there are generally some individuals that `stray' (synonymously used hereafter to refer to dispersal) to non-natal sites to spawn [refs].}\\
 
@@
 It is symmetric in the sense that $x\rightarrow-x$ without altering anything if and only if $h=0$.
-If $h\neq0$, the system is asymmetric, however the pitchfork-like dynamic does not disappear but becomes an imperfect pitchfork where a saddle node bifurcation appears alongside a non-varying steady state.
-This creates a parameter region where alternative steady states exist, separated by an unstable steady state.
+If $h\neq0$, the system is asymmetric, however the pitchfork-like dynamic does not disappear but becomes an imperfect pitchfork where a saddle node bifurcation appears alongside a non-varying steady-state.
+This creates a parameter region where alternative steady-states exist, separated by an unstable steady-state.
 As such, symmetry exists only in a perfect sense (and we describe below why we think our system already violates perfect symmetry), and the observed dynamics that we observe need not require perfect symmetry.
@@
 
-To investigate the effects of increasing asymmetry on the main dynamical feature of our model, we assessed how increasing $\alpha$ altered the steady state portrait as a function of the straying rate $m$ shown in figure 1a, and this analysis is now included as a supplementary figure (figure S13).
+To investigate the effects of increasing asymmetry on the main dynamical feature of our model, we assessed how increasing $\alpha$ altered the steady-state portrait as a function of the straying rate $m$ shown in figure 1a, and this analysis is now included as a supplementary figure (figure S13).
 As can be seen, greater asymmetry (warm colors in the supplementary figure) does not alter the qualitative nature of the dynamics, and does not significantly change the position of the bifurcation.
@@
 If this is in line with what the Reviewer is describing, we suggest that the uploading issues, which prevented delivery of the Supplemental information (and many of the pertinent figures) made the manuscript somewhat opaque with regard to these analyses.
-We examine the effects of density dependent straying on all aspects of the model, and strive to highlight those results in our revised version.
+We examine the effects of density-dependent straying on all aspects of the model, and strive to highlight those results in our revised version.
 
@@
 \noindent We have made the following edits to the text:\\
-\noindent {\bf Line 136:} \emph{Habitat heterogeneity and the rate of straying are treated both independently, and as parameters that covary.
+\noindent {\bf Line 131:} \emph{Habitat heterogeneity and the rate of straying are treated both independently, and as parameters that covary.
 In the latter instance, we evaluate a case where it is assumed that increased habitat heterogeneity correlates with lower straying rates, and vice versa (illustrated in figure S1).
@@
 
-\noindent {\bf Line 272:} \emph{Until now, we have treated the rate of straying and habitat heterogeneity as independent parameters, however they may also be assumed to covary.
+\noindent {\bf Line 268:} \emph{Until now, we have treated the rate of straying and habitat heterogeneity as independent parameters, however they may also be assumed to covary.
 For instance, if sites are separated by greater distance, they may be assumed to have increased habitat heterogeneity as well as lower rates of straying.
@@
 
-\noindent {\bf Minor Response 2.1} We thank the Reviewer for being so meticulous and pointing out these errors.
-Although the noted issues with the figures (missing axes, labels) and missing supplemental materials appears to have been a problem with the upload, we acknowledge that the other issues were due to a lack of thoroughness on our part and have been corrected.
+\noindent {\bf Minor Response 2.1} We thank the Reviewer for pointing out these errors.
+Although the noted issues with the figures (missing axes, labels) and missing supplemental materials appears to have been a problem generated during the upload, we acknowledge that the other issues were due to a lack of thoroughness on our part and have since been corrected.
 

### 2017-09-23 Manuscript/FinalDraft_rev/supp.tex
peter edits integrated
final edits to first round of reviews

@@
 %\title{Community assembly and dynamics by the deconstruction of species interactions}
-\title{Supplementary Figures: Eco-evolutionary dynamics and collective migration: implications for salmon metapopulation robustness}
+\title{Supplementary Figures: Eco-evolutionary dynamics and collective dispersal: implications for salmon metapopulation robustness}
 \author{
@@
 \caption{
-Increasing the straying rate results in the transition from a single steady state for both populations to a dominant and subordinate states.
-If the straying rate is subsequently lowered, the single steady state is not easily obtained, which is the hallmark of hysteresis.
+Increasing the straying rate results in the transition from a single steady-state for both populations to a dominant and subordinate states.
+If the straying rate is subsequently lowered, the single steady-state is not easily obtained, which is the hallmark of hysteresis.
 } \label{fig:hysteresis}
@@
 \caption{
-Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.8$.
+Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady-state $m^*$) with trait heritability $h^2=0.8$.
 If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
@@
 (b) The same as (a) but including recovery times when straying is density-dependent evaluated at $m^*$, shown by linked point pairs.
-Recovery times for systems with density-dependent straying are longer when straying is low and shorter when straying is high, mirroring the change in portfolio effects with respect to density dependent straying shown in figure \ref{extfig-fig:thetaPE}.
+Recovery times for systems with density-dependent straying are longer when straying is low and shorter when straying is high, mirroring the change in portfolio effects with respect to density-dependent straying shown in figure \ref{extfig-fig:thetaPE}.
 } \label{fig:relaxtheta}
@@
   Portfolio effects as a function of straying rate $m$ and trait heritability $h^2$ when the rate of straying is $m = 0.5(1 + \Delta\theta)^{-1}$.
-Alternative steady states emerge for low values of $m$ (left of the cusp bifurcation, denoted by the black line), whereas a single steady state exists for high $m$.
+Alternative steady-states emerge for low values of $m$ (left of the cusp bifurcation, denoted by the black line), whereas a single steady-state exists for high $m$.
   } \label{fig:mthetaPE}
@@
   Recovery times for three disturbance types when the straying rate covaries with habitat heterogeneity as $m = 0.5(1 + \Delta\theta)^{-1}$ for constant (a) and density-dependent (b) straying rates.
-  The cusp bifurcation is not as clear in (b) because $\Delta\theta$ is a function of the individual straying rate $m_0$, whereas the x-axis in (b) is the straying rate at the steady state $m^*$.
+  The cusp bifurcation is not as clear in (b) because $\Delta\theta$ is a function of the individual straying rate $m_0$, whereas the x-axis in (b) is the straying rate at the steady-state $m^*$.
   Despite this difference, the general behavior shown in (a) are the same in (b).
@@
 \caption{
-Steady state densities of both populations as a function of $m$, where a cusp bifurcation indicates the emergence of alternative steady states: one in a dominant state and one in a subordinate state.
-Steady states for populations with symmetrical values ($\alpha=0$) in the vital rates $r_{\rm max}$ and $\beta$ are shown with cool tones.
+Steady-state densities of both populations as a function of $m$, where a cusp bifurcation indicates the emergence of alternative steady-states: one in a dominant state and one in a subordinate state.
+Steady-states for populations with symmetrical values ($\alpha=0$) in the vital rates $r_{\rm max}$ and $\beta$ are shown with cool tones.
 As the asymmetry among populations between sites increases ($\alpha>0$), their vital rates diverge, such that the maximal growth at sites 1 and 2 is now $r_{\rm max}(1)=r_{\rm max}(1+\tilde{rv}_1)$ and $r_{\rm max}(2)=r_{\rm max}(1+\tilde{rv}_2)$ where $rv_{1,2}$ are independently drawn from $\rm{Normal}(0,\alpha)$ and $r_{\rm max}=2$.
 Similarly the strength of density dependence is calculated at sites 1 and 2 as $\beta(1)=\beta(1+\tilde{rv}_1)$ and $\beta(2)=\beta(1+\tilde{rv}_2)$ where $\tilde{rv}_{1,2}$ are independently drawn from $\rm{Normal}(0,\alpha)$ and $\beta=0.001$.
-Steady states for populations with increasingly asymmetric values ($\alpha\rightarrow 0.1$) for vital rates $r_{\rm max}$ and $\beta$ are shown in warmer tones.
+Steady-states for populations with increasingly asymmetric values ($\alpha\rightarrow 0.1$) for vital rates $r_{\rm max}$ and $\beta$ are shown in warmer tones.
 } \label{fig:symmetry}

### 2017-09-23 model/KevinExtinction.jl

@@ -400,7 +400,7 @@ lines($mvec,$(ma_rth2[:,2]),col=palsub[2],cex=0.5,lwd=2)
 lines($mvec,$(ma_rth2[:,3]),col=palsub[3],cex=0.5,lwd=2)
 text(par('usr')[1]-0.09,max($(ma_rth2)),'(a)', xpd=TRUE)
 arrows($(mvec[indmax(ma_rth2[1:100,1])]),10,$(mvec[indmax(ma_rth2[1:100,1])]),12,length=0.05,angle=40,lwd=3)
-text($(mvec[indmax(ma_rth2[1:100,1])]),9,'PFB')
+text($(mvec[indmax(ma_rth2[1:100,1])]),9,'DCB')
 
 
 plot($(ma1_m2_ddm[:,1]),$(ma_rth2_ddm[:,1]),col=palsub[1],log='y',cex=0.5,pch=16,xlab='m*',ylab='Recovery time',ylim=c(min($rt_ddm),max($(ma_rth2_ddm))),las=1)
@@ -419,7 +419,7 @@ text(-0.072,max($(ma_rth2_ddm)),'(b)', xpd=TRUE)
 types = c('subordinate extinct','dominant extinct','near-collapse')
 legend(x=0.18,y=130,legend=types,col=palsub,pch=22,xpd=TRUE,pt.bg=palsub,cex=1, bty="n") #,title=expression(paste(Delta,theta))
 arrows($(ma1_m2_ddm[indmax(ma_rth2_ddm[1:100,3]),3]),10,$(ma1_m2_ddm[indmax(ma_rth2_ddm[1:100,3]),3]),12,length=0.05,angle=40,lwd=3)
-text($(ma1_m2_ddm[indmax(ma_rth2_ddm[1:100,3]),3]),9,'PFB')
+text($(ma1_m2_ddm[indmax(ma_rth2_ddm[1:100,3]),3]),9,'DCB')
 
 
 dev.off()

### 2017-09-23 model/KevinSimulations_mtheta2.jl

@@ -477,7 +477,7 @@ dev.off()
 namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/Manuscript/FinalDraft_rev/fig_mtheta_rt.pdf");
 R"""
 library(RColorBrewer)
-library(shape)
+#library(shape)
 pal = brewer.pal(9,'Greys')
 pdf($namespace,height=4,width=5)
 palsub = pal[c(4,6,8)];
@@ -488,7 +488,7 @@ lines($mvec2,$(ma_rth[:,3]),col=palsub[3],cex=0.5,lwd=2)
 arrows(0.122,2800,0.122,1800,length=0.05,angle=40,lwd=3)
 types = c('subordinate extinct','dominant extinct','near-collapse')
 legend(x=0.14,y=8000,legend=types,col=palsub,pch=22,xpd=TRUE,pt.bg=palsub,cex=0.9, bty="n") #,title=expression(paste(Delta,theta))
-text(0.122,4000,'PFB')
+text(0.122,4000,'DCB')
 text(0.034,90,'*',cex=2)
 dev.off()
 """

### 2017-09-23 model/mvsddm2.jl

@@ -244,7 +244,7 @@ pdf($namespace,height=8,width=5)
 par(mfrow=c(2,1),mai = c(0.8, 0.8, 0.1, 0.1))
 plot($(ma_mvecsmooth),$(ma_medpe5smooth),col=pal[2],type='l',ylim=c(1,max($([ma_medpe3[mlist] ma_medpe5[mlist] ma_medpe8[mlist]]))),xlab='m, m*',ylab='PE',lwd=3,las=1)
 arrows(0.053,2.48,0.053,2.35,length=0.05,angle=40,lwd=3)
-text(0.053,2.55,'PFB')
+text(0.053,2.55,'DCB')
 
 points($(ma_medindm5max[mddmlist]),$(ma_medpe5_ddm[mddmlist]),col=pal[2],pch=16,cex=0.8)
 points($(ma_medindm5min[mddmlist]),$(ma_medpe5_ddm[mddmlist]),col=pal[2],pch=16,cex=0.8)
@@ -267,7 +267,7 @@ points($(ma_medindm8max[mddmlist]),$(ma_medpe8_ddm[mddmlist]),col=pal[3],pch=16,
 points($(ma_medindm8min[mddmlist]),$(ma_medpe8_ddm[mddmlist]),col=pal[3],pch=16,cex=0.8)
 l = length($(ma_medindm5max[mddmlist]))
 for (i in 1:l) {
-  segments($(ma_medindm3max[mddmlist])[i],$(ma_medpe3_ddm[mddmlist])[i],$(ma_medindm3min[mddmlist])[i],$(ma_medpe3_ddm[mddmlist])[i],col=paste(pal[2],'60',sep=''))
+  segments($(ma_medindm3max[mddmlist])[i],$(ma_medpe3_ddm[mddmlist])[i],$(ma_medindm3min[mddmlist])[i],$(ma_medpe3_ddm[mddmlist])[i],col=paste(pal[1],'60',sep=''))
   segments($(ma_medindm5max[mddmlist])[i],$(ma_medpe5_ddm[mddmlist])[i],$(ma_medindm5min[mddmlist])[i],$(ma_medpe5_ddm[mddmlist])[i],col=paste(pal[2],'60',sep=''))
   segments($(ma_medindm8max[mddmlist])[i],$(ma_medpe8_ddm[mddmlist])[i],$(ma_medindm8min[mddmlist])[i],$(ma_medpe8_ddm[mddmlist])[i],col=paste(pal[3],'60',sep=''))
 }

### 2017-11-15 Manuscript/FinalDraft_rev/msv6.tex
fdraft 3

--- 2017-09-19 Manuscript/FinalDraft_rev/msv6.tex
@@
 
-\usepackage[comma,sort&compress]{natbib}
+% \usepackage[comma,sort&compress]{natbib}
 % \bibpunct{[}{]}{,}{a}{}{;}
@@
 Although the dispersal of individuals from donor to recipient populations provides connections within the metapopulation, promoting demographic and evolutionary rescue, it may also introduce maladapted individuals into habitats host to different environmental conditions, potentially lowering the fitness of the recipient population.
-Here we explore a minimal model of the eco-evolutionary dynamics between two populations connected by dispersal, where the productivity of each is defined by a trait complex that is subject to local selection.
-Although general in nature, our model is inspired by salmon metapopulations, where dispersal between populations is defined in terms of the \emph{straying rate}, which has been shown to be density dependent, and recently proposed to be shaped by social interactions consistent with collective movement.
-%possibly emerging as a consequence of collective migration.
+Thus, dispersal plays a dual role in both promoting and inhibiting local adaptation.
+Here we explore a model of the eco-evolutionary dynamics between two populations connected by dispersal, where the productivity of each is defined by a trait complex that is subject to local selection.
+Although general in nature, our model is inspired by salmon metapopulations, where dispersal between populations is defined in terms of the straying rate, which has been shown to be density-dependent, and recently proposed to be shaped by social interactions consistent with collective movement.
 The results of our model reveal that increased straying between evolving populations leads to alternative stable states, which has large and nonlinear effects on two measures of metapopulation robustness: the portfolio effect and the time to recovery following an induced disturbance.
 We show that intermediate levels of straying result in large gains in robustness, and that increased habitat heterogeneity promotes robustness when straying rates are low, and erodes robustness when straying rates are high.
-Finally, we show that density-dependent straying promotes robustness, particularly when the aggregate biomass is low and straying is correspondingly high, and this has important ramifications for the conservation of salmon metapopulations facing both natural and anthropogenic disturbances.
+Finally, we show that density-dependent straying promotes robustness, particularly when the aggregate biomass is low and straying is correspondingly high, which has important ramifications for the conservation of salmon metapopulations facing both natural and anthropogenic disturbances.
 \end{abstract}
@@
 \vspace{2mm}
-\noindent {\bf Media Summary} The rate at which individuals move within a metapopulation has large impacts on extinction risk.
-Among migratory species such as salmon, the rate of dispersal, or straying, may be density dependent due to collective decision-making.
+\noindent {\bf Media Summary} 
+Many migratory species, such as salmon, are remarkable in finding their way home.
+This homing has allowed fine-scale adaptations to the environments in which they evolve.
+But some individuals do not find their way home and instead stray to other locations, especially when there are fewer individuals to help with collective decision-making.
+With an eco-evolutionary model, we discovered that an intermediate and density-dependent straying rate allows linked populations to be robust to disturbance but maintain local adaptations.\\
+% The rate at which individuals move within a metapopulation has large impacts on extinction risk.
+Among migratory species such as salmon, the rate of dispersal, also referred to as straying, may be density-dependent due to collective decision-making.
 We explore an eco-evolutionary model of two populations connected by dispersal, where the productivity of each is defined by a trait complex subject to local selection.
-We show that intermediate straying rates result in large gains in robustness, and that density-dependent straying promotes robustness.
+We show that intermediate straying rates result in faster recovery times and more stable population dynamics, and that density-dependent straying tends to promote this robustness.
 These results have important ramifications for the conservation of salmon metapopulations facing both natural and anthropogenic disturbances.\\
@@
 % Here we ask the overarching question: how does collective behavior mediated dispersal and gene flow interact to influence the robustness of locally adapted populations?
-To address this question we construct a minimal eco-evolutionary model of two populations occupying different sites that are linked by straying individuals, each with an associated trait distribution subject to natural selection determined by local conditions.
-Specifically we compare (a) different rates of straying and (b) the influence of collective movement, across (c) increasing environmental heterogeneity, by assessing two measures of metapopulation robustness: the portfolio effect and the time required for a population(s) to recover following an induced disturbance.
+To address this question we constructed a eco-evolutionary model of two populations occupying different sites that are linked by straying individuals, each with an associated trait distribution subject to natural selection determined by local conditions.
+Specifically we compared (a) different rates of straying and (b) the influence of collective movement, across (c) increasing environmental heterogeneity, by assessing two measures of metapopulation robustness: the portfolio effect and the time required for a population(s) to recover following an induced disturbance.
 % This model enables us to explore the multiple and potentially opposing pathways by which straying influences metapopulation robustness such as the potentially detrimental erosion of local adaptation vs.
@@
   &R_i[\mu_i(t)] = \\ \nonumber
-  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i(t)-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i(t),\mu_i,\sigma^2) {\rm d}x_i(t) +\tilde{P}_i\\ \nonumber
+  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i,\mu_i(t),\sigma^2) {\rm d}x_i +\tilde{P}_i\\ \nonumber
   &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-\mu_i(t))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\} +\tilde{P}_i,
@@
 % \noindent{\bf (b) Recruitment over a selective landscape}
-\noindent Because individuals from the local population are mixed with individuals from the remote population via straying and subsequent reproduction, the resulting trait distribution is a mixed normal with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and straying individuals, $1-w_i$, where
+Because individuals from the local population are mixed with individuals from the remote population via straying and subsequent reproduction, the resulting trait distribution is a complex mixture of trait distributions (see Supplemental Materials: Appendix I for the derivation, Fig.
+\ref{fig:PDF}).
+We make two simplifying assumptions.
+First, we approximate the distribution resulting from the mix of remote and local individuals following reproduction as a Gaussian where $X_i=x_i$ with probability $g(x_i)$.
+The expectation of the actual trait distribution as well as the Gaussian approximation are the same, such that ${\rm E}\{g(x_i)\} = w_i\mu_i + (1-w_i)\mu_j$, with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and straying individuals, $1-w_i$, where
 \begin{equation}
@@
 \end{equation}
-We make two simplifying assumptions.
-First, we assume that the distribution resulting from the mix of remote and local individuals, following reproduction, is also normal with a mean value equal to that of the mean for the mixed-normal distribution.
-Thus, we assume that strays can successfully reproduce and introduce their genotypes into the recipient population, which is supported by observations in wild populations \citep{Jasper:2013cc}.
-Second, we assume that changes in trait variance through time are minimal, such that $\sigma^2$ is assumed to be constant, which is a common simplification in eco-evolutionary models of population dynamics \citep{Lande:1976ga,Schreiber:2011wx,Gilbert:2014ee,Gibert:2015kc}.
+We show that the Gaussian approximation is not only exact with respect to the mean, but also is a good approximation of variance in most relevant cases (see Appendix I in the Supplemental Materials, Fig.
+\ref{fig:PDF}).
+Thus, strays can successfully reproduce and introduce their genotypes into the recipient population, which is supported by observations in wild populations \citep{Jasper:2013cc}.
+Secondly, we assumed that changes in trait variance through time are minimal, such that $\sigma^2$ is constant over time, which is a common simplification in eco-evolutionary models of population dynamics \citep{Lande:1976ga,Schreiber:2011wx,Gilbert:2014ee,Gibert:2015kc}.
 
 % An increasing flow of incoming strays is generally expected to pull the mean trait value of the local population away from its optimum over time, which will decrease its rate of recruitment.
-Following Lande \citep{Lande:1976ga}, the mean trait value thus changes through time according to the difference equation
+Following Lande \citep{Lande:1976ga}, and given our assumption of trait distribution normality, the mean trait value thus changes through time according to the difference equation
 
@@
 \noindent {\bf (b) Density-dependent straying}
-%Density dependent m
-\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is mounting evidence that at least in some species (including salmon) the straying rate is density-dependent, a signature of collective navigation \citep{Berdahl:2016dx,Bett:2017ha}.
+%Density-dependent m
+\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is mounting evidence that at least in some species (including salmon) the straying rate is density-dependent, a signature of collective navigation \citep{Berdahl:2014bl,Berdahl:2017uu}.
 Specifically, the rate at which individuals stray has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tend to decrease the rate at which individuals err, reducing the overall proportion of a population that strays.
-According to Berdahl et al. \citep{Berdahl:2014bl,Berdahl:2016dx}, given the probability that an individual strays is $m_0$, the proportion of the local population $N_i(t)$ that strays is
+According to Berdahl et al. \citep{Berdahl:2016dx}, given the probability that an individual strays is $m_0$, the proportion of the local population $N_i(t)$ that strays is
 
@@
 \noindent {\bf (d) Measuring metapopulation robustness}
-\noindent We evaluated metapopulation robustness by measuring the average-CV portfolio effect (PE) \citep{Anderson:2014cx,Schindler:2015gf} as well as the time required for the system to return to a steady state following an induced disturbance to one or both of the populations \citep{Ovaskainen:2002il}.
+\noindent We evaluated metapopulation robustness by measuring the average-CV portfolio effect (PE) \citep{Anderson:2014cx,Schindler:2015gf} as well as the recovery time, which is the time required for the system to return to a steady-state following an induced disturbance to one or both of the populations \citep{Ovaskainen:2002il}.
+Throughout, we refer to an increase in portfolio effects and/or reduction in recovery time as promoting metapopulation robustness.
+
 The average-CV portfolio effect is, as the name implies, the average CV across each population $N_i$ divided by the CV of the aggregate $N_T=\sum_i N_i$ \citep{Anderson:2013gb}, such that
@@
 
-\noindent where in this case the number of populations is limited to $X=2$ and the expectations $\rm E(\cdot)$ and variances $\rm VAR(\cdot)$ are evaluated at the steady state.
-The steady state condition is denoted by `$*$'.
+\noindent where in this case the number of populations is limited to $X=2$ and the expectations $\rm E(\cdot)$ and variances $\rm VAR(\cdot)$ are evaluated at the steady-state.
+The steady-state condition is denoted by `$*$'.
 As the CV of $N_T^*$ decreases relative to that of the constituent populations, $\langle{\rm PE}\rangle > 1$, and the metapopulation is presumed to be more stable because the aggregate has functioned to dampen population-level variance.
@@
 %Although there are more robust ways to measure the portfolio effect (in particular for comparing the PE across species with different mean-variance relationships; REF: Anderson), its only utlity here is to provide a simple statistical measure of metapopulation persistence.
-A more direct way to measure system robustness is to measure the time that the system (measured as the aggregate steady state biomass $N_T^*$) takes to return to a steady state following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
-Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation \citep{GuckHolmes}, because we aimed to 1) assess the effects of a large perturbation, and 2) estimate the time required for all transient effects to decay (including dampened oscillations), we used a simulation-based numerical procedure.
+A more direct way to measure system robustness is to measure the time that the system (measured as the aggregate steady-state biomass $N_T^*$) takes to return to a steady-state following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
+Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation \citep{GuckHolmes}, because we aimed to 
+1) assess the effects of a large perturbation far from the steady-state, and 
+2) estimate the time required for all transient effects to decay following this perturbation (including dampened oscillations), we used a simulation-based numerical procedure.
 Recovery time was calculated by initiating a disturbance at $t=t_d$, and monitoring $N_T(t_d+t)$ as $t\rightarrow T$, where $T$ is large. 
@@
 
-Numerically estimating the time that it takes for a perturbed system to recover also permits a more detailed perspective of metapopulation fragility.
+Numerically estimating the time that it takes for a perturbed system to recover also permits a more detailed perspective of metapopulation robustness.
 For example, if populations settle to alternative stable states (one at low- and one at high-density), comparing recovery times after a disturbance applied to individual populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery.
@@
 We measured recovery time following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} are equivalent if populations have the same density); (\emph{iii}) near-collapse of both populations where just 1.0\% of each survives.
-Throughout, we will refer to an increase in the portfolio effects and/or reduction in recovery times as promoting metapopulation robustness.
 \\
@@
 \noindent{\bf (a) Nonlinear effects of straying on metapopulation robustness} \\
-\noindent Regardless of density dependence, straying lowers steady state densities for both populations by (\emph{i}) the donor population losing locally-adapted individuals to the recipient population and (\emph{ii}) the introduction of maladapted individuals to the recipient population from the donor population (Fig.
+\noindent Regardless of density dependence, straying lowers steady-state densities for both populations by (\emph{i}) the donor population losing locally-adapted individuals to the recipient population and (\emph{ii}) the introduction of maladapted individuals to the recipient population from the donor population (Fig.
 \ref{fig:traj}).
 This prediction is in accordance with observations from natural populations \citep{Bett:2017ha}. %, regardless of trait heritability and variance or habitat heterogeneity
-The decline in steady state densities is not gradual: as straying increases, the system crosses a discrete cusp bifurcation (DCB) \citep{AleksandrovichKuznetsov:1995p2580} whereby the single steady state for the metapopulation bifurcates into two basins of attraction: one at high biomass, and one at low biomass density (figure \ref{fig:traj}a, \ref{fig:PE}a).
+The decline in steady-state densities is not gradual: as straying increases, the system crosses a discrete cusp bifurcation (DCB) \citep{AleksandrovichKuznetsov:1995p2580} whereby the single steady-state for the metapopulation bifurcates into two basins of attraction: one at high biomass, and one at low biomass density (figure \ref{fig:traj}a, \ref{fig:PE}a).
 Mean trait values for both populations bifurcate similarly (figure \ref{fig:traj}b). 
 In discrete systems, the cusp bifurcation is defined by two fold bifurcations intersecting at a cusp point \citep{AleksandrovichKuznetsov:1995p2580}, and is observed when the real part of the dominant eigenvalue of the Jacobian matrix crosses the unit circle at +1 (figure \ref{fig:eigs}).
-Visually, the dynamics are similar to those observed at a pitchfork bifurcation in continuous systems, where a single steady state gives rise to two steady states separated by an unstable fixed point.
+Visually, the dynamics are similar to those observed at a pitchfork bifurcation in continuous systems, where a single steady-state gives rise to two alternative steady-states.
 
 Above the threshold straying rate defined by the DCB, there are two alternative eco-evolutionary states: the \emph{dominant state} population will have a higher density and greater degree of local adaptation (smaller trait offset from the local optimum), while the \emph{subordinate state} population will have lower density with maladapted trait values (larger trait offset from the local optimum).
-Hysteresis is observed to occur at this transition, such that the single steady state regime cannot easily be recovered by reducing straying after the system attains alternative steady states (figure \ref{fig:hysteresis}).
+Hysteresis is observed to occur at this transition, such that the single steady-state regime cannot easily be recovered by reducing straying after the system attains alternative steady-states (figure \ref{fig:hysteresis}).
 These dynamics are also observed in the Ronce and Kirkpatrick model, where populations are described as transitioning between symmetric to asymmetric states \citep{Ronce:2001dp}.
@@
 
-Trait heritability $h^2$ has a large effect on the degree to which straying affects both the aggregate population steady state density ($N^*_T$; figure \ref{fig:PE}a) as well as the difference between steady state densities (the distance between alternative stable states: $\Delta N^*=|N^*_1-N^*_2|$; figure \ref{fig:PE}b).
+Trait heritability $h^2$ has a large effect on the degree to which straying affects both the aggregate population steady-state density ($N^*_T$; figure \ref{fig:PE}a) as well as the difference between steady-state densities (the distance between alternative stable states: $\Delta N^*=|N^*_1-N^*_2|$; figure \ref{fig:PE}b).
 Greater trait heritability results in a steeper decline in $N_T^*$ with increasing straying rates $m$, but leads to only moderate changes to $\Delta N^*$.
@@
 \ref{fig:PE}a-c) occurs at lower values of the straying rate $m$ with lower heritability (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
-Although trait heritability among salmonids is variable, most life history traits have an $h^2 <0.5$ \citep{Carlson:2008hl}, and we largely focus our efforts on that range.
+Although trait heritability among salmonids is variable, most life-history traits have an $h^2 <0.5$ \citep{Carlson:2008hl}, and we largely focus our efforts on that range.
 
@@
 % Return time as a measure of system persistence and relation to PE
-As the portfolio effect is highly sensitive to the rate of straying between populations, so is the time required for the system to recover to a steady state following a large disturbance.
+As the portfolio effect is highly sensitive to the rate of straying between populations, so is the time required for the system to recover to a steady-state following a large disturbance.
 In general, we find that the average-CV portfolio effect is negatively correlated with recovery time (figure \ref{fig:PE}d), indicating that, for our system, both measures are valuable indicators of metapopulation robustness.
@@
 Straying had non-linear impacts on the recovery time of populations. 
-When the dominant state (well adapted and high density) is wiped out, high rates of straying allow it to recover quickly (figure \ref{fig:relax}a) because the surviving population has a mean trait value skewed towards the optimum of the recovering population (figure \ref{fig:relaxtraj_hdlh}).
+When the dominant state population (well adapted and high density) goes extinct, high rates of straying allow recolonization of the extirpated habitat and quick recovery (figure \ref{fig:relax}a) because the surviving population has a mean trait value skewed towards the optimum of the affected habitat (figure \ref{fig:relaxtraj_hdlh}).
 Yet, as straying decreases, recovery time for the disturbed dominant state population increases, in part because there is time for the trait distribution to move back towards the trait optimum of the subordinate state population.
 In contrast, when the subordinate state population (maladapted and low density) is wiped out, recovery rates are fastest at low to intermediate levels of straying.
-Because the mean trait values of both populations are skewed towards those of the dominant population, when the subordinate population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into the recovering population extends the length of time required for it to return to its steady state (figure \ref{fig:relaxtraj_ldlh}).
+Because the mean trait values of both populations are skewed towards those of the dominant population, when the subordinate population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into the recovering population extends the length of time required for it to return to its steady-state (figure \ref{fig:relaxtraj_ldlh}).
 When both populations are both dramatically reduced, recovery time is generally fastest at lower levels of straying due to reduced mixing of maladaptive phenotypes.
@@
 %Lowering Phenotypic diversity with increasing straying rate
-Increased rates of straying lowers phenotypic diversity ($\Delta \mu^* = |\mu_i^*-\mu_j^*|$, evaluated at the steady state) because both local and remote populations are increasingly homogenized.
-The loss of phenotypic diversity with increased straying is greater if trait heritability is low because traits take longer to go back to their local optima than they do when heritability is high. 
+Increased rates of straying lowers phenotypic diversity ($\Delta \mu^* = |\mu_i^*-\mu_j^*|$, evaluated at the steady-state) because both local and remote populations are increasingly homogenized.
+The loss of phenotypic diversity is greater with increased straying when trait heritability is low because traits take longer to go back to their local optima than they do when heritability is high. 
 Hence straying counters the effect of diversifying local adaptation. 
@@
 
-\noindent{\bf (b) The effects of collective migration and density dependent straying} \\
+\noindent{\bf (b) The effects of collective navigation and density-dependent straying} \\
 %Portfolio effects
@@
 \ref{eq:ddm}).
-%Because our analyses concern steady state conditions, the inclusion of density-dependent straying does not have a large impact on the qualitative nature of our findings (figure \ref{fig:ddm}).
-Density dependence alters the straying rate at steady state population densities because $0 < m^* < m_0$, and this serves to rescale both the strength of the PE as well as the recovery time, but does not change the qualitative nature of our findings.
-In the alternative stable state regime, because each population exists at different steady state densities, there are likewise two alternative straying rates $(m_i^*,m_j^*)$: the higher straying rate is associated with the low-density population, and the lower straying rate is associated with the high-density population.
+%Because our analyses concern steady-state conditions, the inclusion of density-dependent straying does not have a large impact on the qualitative nature of our findings (figure \ref{fig:ddm}).
+Density dependence alters the straying rate at steady-state population densities because $0 < m^* < m_0$, and this serves to rescale both the strength of the PE as well as the recovery time, but does not change the qualitative nature of our findings.
+In the alternative stable state regime, because each population exists at different steady-state densities, there are likewise two alternative straying rates $(m_i^*,m_j^*)$: the higher straying rate is associated with the low-density population, and the lower straying rate is associated with the high-density population.
 We assessed metapopulation robustness across a range of $(m_i^*,m_j^*)$ values by varying the probability that an individual strays $m_0$, which is positively and linearly related to $(m_i^*,m_j^*)$.
@@
 % As such, the difference in $m^*$ between subordinate and dominant populations appears to matter less than the lower straying rate generated by the dominant population.
-As $m^*$ increases, we observe an increase in the PE than for systems with constant $m$.
+As $m^*$ increases, we observe an increase in the PE relative to systems with constant $m$.
 
@@
 %whereas in the alternative stable state regime (higher $m^*$), the recovery times for different disturbance types were very similar to systems with a constant $m$ (figure \ref{fig:relax}b; note difference in x-axis scales).
-Trait heritability had a large effect on recovery times, with near-collapse requiring a more protracted time to recover in the alternative steady state regime (figure \ref{fig:relax_highh}).
+Trait heritability had a large effect on recovery times, with near-collapse requiring a more protracted time to recover in the alternative steady-state regime (figure \ref{fig:relax_highh}).
 % As trait heritability increased, the metapopulation always recovered more quickly if the subordinate population was lost (figure \ref{fig:relax_highh}).
@@
 % This higher PE at increased $m^*$ mirrors recovery times that are also lower than those observed for systems with constant $m$ (figure \ref{fig:relax}).
-Together, analysis of both PE and recovery time suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our minimal model, it does appear to promote robustness, particularly in the case of near-collapse of both populations when straying is high.
+Together, analysis of both PE and recovery time suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our model, it does appear to promote robustness, particularly in the case of near-collapse of both populations when straying is high.
 \\
@@
 \noindent{\bf (c) The role of habitat heterogeneity and changing selective landscapes}\\ 
-\noindent With the onset of straying, we find that increasingly divergent trait optima generally lower $N_T^*$ and exaggerate $\Delta N^*$, and this is particularly pronounced for density dependent straying (figure \ref{fig:thetadiffN}).
+\noindent With the onset of straying, we find that increasingly divergent trait optima generally lower $N_T^*$ and exaggerate $\Delta N^*$, and this is particularly pronounced for density-dependent straying (figure \ref{fig:thetadiffN}).
 %, such that the biomass distribution becomes increasingly uneven.
@@
 %The rate of straying that gives rise to a large PE at the cusp bifurcation is conditioned on the difference in optimal trait values between the local and remote populations, which we call $\Delta\theta = \sqrt{(\theta_i - \theta_j)^2}$.
-As habitat heterogeneity increases, alternative steady states appear at lower straying rates -- with the crossing of the cusp bifurcation, accompanied by a peak in the PE -- whereas the magnitude of increase in the PE also increases (figure \ref{fig:thetaPE}b), lowering recovery time (figure \ref{fig:relaxtheta}).
+As habitat heterogeneity increases, alternative steady-states appear at lower straying rates -- with the crossing of the cusp bifurcation, accompanied by a peak in the PE -- whereas the magnitude of increase in the PE also increases (figure \ref{fig:thetaPE}b), lowering recovery time (figure \ref{fig:relaxtheta}).
 For increased rates of straying, greater habitat heterogeneity erodes the PE (figure \ref{fig:thetaPE}b) and increases the recovery times (figure \ref{fig:relaxtheta}).
@@
 1) the average-CV portfolio effect \citep{Anderson:2013gb,Anonymous:2015gf}, a statistical metric commonly used to assess the buffering capacity of metapopulations, and
-2) the recovery time, defined here as the time required for the aggregate metapopulation biomass $N_T^*$ to return to its steady state following an induced disturbance, which is mechanistically linked to persistence \citep{Ovaskainen:2002il}.
-In our minimal eco-evolutionary model of dispersal and natural selection between two populations, we show that these statistical and mechanistic descriptors of metapopulation dynamics and robustness are tightly coupled (figure \ref{fig:PE}d), which is not uncommon for diverse metrics of stability \citep{Donohue:2013iu}.
+2) the recovery time, defined here as the time required for the aggregate metapopulation biomass $N_T^*$ to return to its steady-state following an induced disturbance, which is mechanistically linked to persistence \citep{Ovaskainen:2002il}.
+In our eco-evolutionary model of dispersal and natural selection between two populations, we show that these statistical and mechanistic descriptors of metapopulation dynamics and robustness are tightly coupled (figure \ref{fig:PE}d), which is not uncommon for diverse metrics of stability \citep{Donohue:2013iu}.
 Taken as a whole, our results point to an important role of density-dependent straying in the colonization and recovery dynamics within metapopulations, while also underscoring the risk of straying by individuals with maladaptive traits to reduce the productivity of locally adapted stock complexes.
@@
 % %The bifurcation and critical slowing down
-% [This paragraph is interesting to me but could be cut] Although the portfolio effect is negatively correlated with recovery time, there is one exception: the point of transition from a single steady state to alternative stable states, which occurs at the pitchfork bifurcation (REF).
+% [This paragraph is interesting to me but could be cut] Although the portfolio effect is negatively correlated with recovery time, there is one exception: the point of transition from a single steady-state to alternative stable states, which occurs at the pitchfork bifurcation (REF).
 % At this point, the portfolio effect is maximized, but so is the recovery time (figures \ref{fig:PE}, \ref{fig:relax}).
@@
 Although the emergence of alternative stable states via a cusp bifurcation is characteristic of symmetrical systems, we find that increasing the asymmetry in the vital rates of populations across sites does not significantly alter the presence or position of the DCB (figure \ref{fig:symmetry}).
-That these patterns arise in alternative formulations and are insensitive to parameter asymmetry suggests that the emergence of dominant and subordinate states directly from eco-evolutionary dynamics may be a generally relevant phenomenon in ecological systems.
-%, though this requires additional investigation and scrutiny.
+That these patterns arise in alternative formulations and are relatively insensitive to parameter asymmetry suggests that the dynamical features observed have potentially widespread ramifications for the evolutionary dynamics of spatially connected populations in general.
+% suggests that the emergence of dominant and subordinate states directly from eco-evolutionary dynamics may be a generally relevant phenomenon in ecological systems.
+% Importantly, we show that the resulting dynamics are qualitatively similar, suggesting that the dynamical features present in both of these approaches have potentially widespread ramifications for the evolutionary dynamics of spatially connected populations, regardless of the specific formulation implemented.
+<%, though this requires additional investigation and scrutiny.
 
@@
 Although high levels of dispersal in our system generally supports this finding, the interplay between dispersal and PE is more subtle when selection for local adaptations is considered.
-Low to intermediate levels of density-dependent straying result in an elevated PE, increasing the buffering capacity of the metapopulation.
-Although PE is measured at the steady state, low to intermediate rates of straying also appear to have a beneficial effect on transient dynamics.
+Low to intermediate levels of straying result in an elevated PE, increasing the buffering capacity of the metapopulation.
+
+Although PE is measured at the steady-state, low to intermediate rates of straying also appear to have a beneficial effect on transient dynamics.
 When there is just enough straying to cause alternative stable states, the time to recovery following an induced disturbance declines, though -- as with the PE -- it then grows if straying becomes large (figure \ref{fig:relax}a).
-
-% %Density dependent straying
+In the alternative steady-state regime, a lower rate of straying inhibits admixture of maladapted individuals.
+Following a large disturbance, such as the near-collapse of both dominant and subordinate populations, this limited mixing increases the growth rates of both populations, permitting faster recovery times.
+If the rate of straying becomes too high, an influx of maladapted individuals into both populations inhibits local growth rates, and recovery slows.
+
+% %Density-dependent straying
 % This themed issue is largely concerned with the notion that straying in biological systems where movement is a function of collective navigation may be density-dependent.
@@
 We highlight three important findings that contribute to our understanding of collective movement suggesting that density-dependent straying may play an important role in the persistence of metapopulations over evolutionary time.
-First, the inclusion of density-dependent straying does not qualitatively alter either (\emph{i}) steady state or (\emph{ii}) transient dynamics of our minimal eco-evolutionary model, but effectively rescales measures of robustness to the lower straying rates that emerge as a consequence of the coupled dynamics.
+First, the inclusion of density-dependent straying does not qualitatively alter either (\emph{i}) steady-state or (\emph{ii}) transient dynamics of our eco-evolutionary model, but effectively rescales measures of robustness to the lower straying rates that emerge as a consequence of the coupled dynamics.
 Second, compared to systems with constant dispersal, density-dependent straying appears to increase the portfolio effect across a range of straying rates (figure \ref{fig:thetaPE}a). 
@@
 %mtheta
-In other cases, habitats that are closer in space can be assumed to have greater similarity in environmental conditions than those that are geographically distant, and phenotypes of more proximately located populations should be more similar \citep{Westley:2012ui}.
+In other cases, habitats that are closer in space can be assumed to have greater similarity in environmental conditions than those that are geographically distant, and phenotypes of more proximately located populations should be more similar \citep{Reisenbichler:1988ex,Fraser:2011co,Westley:2012ui}.
 It is thus reasonable to expect a larger number of straying individuals between sites that are geographically proximate and indeed evidence corroborates this prediction \citep{Candy:2000hu,JPE:JPE1383}.
@@
 %there is enough isolation to allow the surviving subordinate population to adapt towards its own optimum.
-Morever, there is an inversion in the alternative stable states following the disturbance, resulting in a state shift in dominance.
-This may have particular conservation implications when aiding dispersal following disturbances or reconnections of lost habitats \citep{Anderson:2013bf,Pess:2014isa}.
+Such a dynamic is accompanied by an inversion in the alternative stable states following the disturbance, resulting in a state shift in dominance.
+Thus, management activities that alter dispersal rates by outplanting individuals or reconnecting disconnected habitats could have unintended eco-evolutionary consequences  \citep{Anderson:2013bf,Pess:2014isa}.
+% This may have particular conservation implications when aiding dispersal following disturbances or reconnections of lost habitats \citep{Anderson:2013bf,Pess:2014isa}.
 % Moreover, local temperature regimes are known to play a central role in dictating local adaptation of salmon populations.
@@
 % Population declines following these disturbances would also magnify the rate of straying, and our results suggest that this response in the straying rate would quicken the recovery of the small populations that survive.
-% Indeed, we found that density-dependent migration appears to increase the robustness of metapopulations faced with large-scale disturbance, particularly in regimes where populations have lower steady states and correspondingly higher rates of straying.
+% Indeed, we found that density-dependent migration appears to increase the robustness of metapopulations faced with large-scale disturbance, particularly in regimes where populations have lower steady-states and correspondingly higher rates of straying.
 % As population numbers increase and stray rates decrease, the metapopulation could achieve higher production due to fine-scale local adaptation unhindered by high rates of maladaptive strays.
 
-The portfolio effect and the time to recovery following a disturbance are independent and correlated measures of metapopulation robustness that take into account both steady state and transient dynamics.
+The portfolio effect and the time to recovery following a disturbance are independent and correlated measures of metapopulation robustness that take into account both steady-state and transient dynamics.
 We show that these measures of robustness are strongly influenced by the rate at which individuals from donor populations stray into habitats occupied by recipient populations. 
 Importantly, density-dependent straying, which may occur when individuals collectively navigate, can both increase the portfolio effect and lower the time to recovery following a disturbance, which is anticipated to promote persistence.
+Therefore, preserving the biological processes that facilitate this collective behavior may be an important conservation target in its own right, echoing the sentiments of  Hardesty-Moore et al.
+\citep{HardestyMoore:wg}.
 We suggest that understanding the spatial complexity of metapopulations dispersing across heterogeneous environments, in tandem with the mosaic of selective forces acting on those environments, may be key to uncovering those factors that promote persistence in the wild.
@@
 % \bibliography{aa_kevin}
-% 
 \begin{thebibliography}{10}
@@
 \newblock \emph{Movement Ecology} \textbf{4}, 1--12
+
+\bibitem{HardestyMoore:wg}
+Hardesty-Moore M, \emph{et~al.}, 2017 {Migration in the Anthropocene}.
+\newblock \emph{Philos. T. Roy. Soc. B} \textbf{This volume}
 
@@
 
+\bibitem{Berdahl:2017uu}
+Berdahl A, 2017 {Berdahl et al. contribution}.
+\newblock \emph{Philos. T. Roy. Soc. B} \textbf{This volume}
+
 \bibitem{Moore:2010gs}
@@
 
+\bibitem{Reisenbichler:1988ex}
+Reisenbichler RR, 1988 {Relation between Distance Transferred from Natal Stream
+  and Recovery Rate for Hatchery Coho Salmon}.
+\newblock \emph{N. Am. J. Fish. Manage.} \textbf{8}, 172--174
+
+\bibitem{Fraser:2011co}
+Fraser DJ, Weir LK, Bernatchez L, Hansen MM, Taylor EB, 2011 {Extent and scale
+  of local adaptation in salmonid fishes: review and meta-analysis.}
+\newblock \emph{Heredity} \textbf{106}, 404--420
+
 \bibitem{Westley:2012ui}
@@
 
-% \begin{table}[!t]
+\begin{table}[!t]
 \begin{center}
-\begin{tabular}{ l|l }
+\begin{tabular}{ l l l }
 \hline
-Parameter & Definition \\
+Parameter & Definition & Value \\
 \hline
-$N_i(t)$, $N_T(t)$ & Individual, aggregate population over time\\
-$x_i$ & Trait value for an individual in population $i$\\
-$\mu_i(t)$ & Mean of $x$ for population $i$ over time\\
-$\sigma^2$ & Genetic variance of trait $x$\\
-$m$, $m(t)$ & Constant, density dependent straying rate\\
-$m_0$ & Straying rate of an individual\\
-$R_i[\mu(t)]$ & Recruitment rate of population $i$\\
-$r_{\rm max}$ & Maximum recruitment rate\\
-${\rm e}^{-Z}$ & Survival rate\\
-$\beta$ & Strength of density dependence\\
-$\theta_i$ & Optimal trait value for habitat $i$\\
-$\Delta\theta$ & Habitat heterogeneity\\
-$\tau$ & Strength of selection\\
-$h^2$ & Heritability\\
-$C$ & Half saturation constant for $m(t)$\\
-${\rm PE}$ & Portfolio Effect\\
+$N_i(t)$, $N_T(t)$ & Individual, aggregate population over time & {\rm var.}\\
+$x_i$ & Trait value for an individual in population $i$ & {\rm var.}\\
+$\mu_i(t)$ & Mean of $x$ for population $i$ over time & {\rm var.}\\
+$\sigma^2$ & Genetic variance of trait $x$ & 1.0\\
+$m$, $m(t)$ & Constant, density-dependent straying rate & {\rm var.}\\
+$m_0$ & Straying rate of an individual & {\rm var.}\\
+$R_i[\mu(t)]$ & Recruitment rate of population $i$ & {\rm var.}\\
+$r_{\rm max}$ & Maximum recruitment rate & 2.0\\
+${\rm e}^{-Z}$ & Survival rate & 0.6\\
+$\beta$ & Strength of density dependence & $10^{-3}$\\
+$\theta_i$ & Optimal trait value for habitat $i$ & 5.0\\
+$\Delta\theta$ & Habitat heterogeneity & {\rm var.}\\
+$\tau$ & Strength of selection & 1.0\\
+$h^2$ & Heritability & {\rm var.}\\
+$C$ & Half saturation constant for $m(t)$ &  $10^3$\\
+${\rm PE}$ & Portfolio Effect & {\rm var.}\\
+$T$ & Terminal simulation time & $10^5$\\
 \hline
@@
 \end{center}
-% \end{table}
-
-\clearpage
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.4\textwidth]{fig_traj.pdf}
-\caption{
-(a) The steady state densities of $N_1$ and $N_2$ as a function of a constant stray rate $m$.
-Which population attains the low- or high-density state is random due to small applied fluctuations in the initial conditions.
-(b) The steady state trait values measured as $\theta_i - x_i$, as a function of a constant stray rate $m$. 
-DCB marks the discrete cusp bifurcation.
-Unless otherwise indicated, the default parameter values used are: $r_{\rm max}=2$; $Z=0.5$; $\beta=0.001$; $\theta_1=5$; $\Delta\theta=5$; $\tau=1$; $\sigma=1$; $T=1\times10^5$.
-} \label{fig:traj}
-\end{figure}
+\caption{Table of parameters, definitions, and assigned values (var. = variable).}
+\end{table}
+
 % tmax=10000;
@@
 
+\clearpage
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.4\textwidth]{fig_traj.pdf}
+\caption{
+(a) The steady-state densities of $N_1$ and $N_2$ vs. the stray rate $m$. Which population attains the low- or high-density state is random due to small applied fluctuations in the initial conditions.
+(b) The steady-state trait values measured as the offset from the local optimum $\theta_i - mu_i$, vs. the stray rate $m$. 
+DCB marks the discrete cusp bifurcation.
+% Unless otherwise indicated, the default parameter values used are: $r_{\rm max}=2$; $Z=0.5$; $\beta=0.001$; $\theta_1=5$; $\Delta\theta=5$; $\tau=1$; $\sigma=1$; $T=1\times10^5$.
+} \label{fig:traj}
+\end{figure}
+
 \begin{figure*}
@@
 (c) the portfolio effect PE as a function of heritability $h^2$ and a constant stray rate $m$. Light colors = high values.
-The black line shows the cusp bifurcation separating a single steady state (left) from alternative stable states (right).
+The black line shows the cusp bifurcation separating a single steady-state (left) from alternative stable states (right).
 (d) The relationship between the time to recovery following a disturbance and the portfolio effect. Portfolio effects greater than unity corresponds to less synchronization.
@@
 \caption{
-Recovery time of $N_T^*$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.2$.
+Recovery time of $N_T^*$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady-state $m^*$) with trait heritability $h^2=0.2$.
 If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
@@
 \caption{
-Phenotypic diversity ($\Delta \mu^*$) evaluated at the steady state as a function of straying rate $m$ and trait heritability $h^2$.
+Phenotypic diversity ($\Delta \mu^*$) evaluated at the steady-state as a function of straying rate $m$ and trait heritability $h^2$.
 The jump occurs as the system crosses the cusp bifurcation; lower phenoytic diversity emerges with higher straying rates and in the alternative stable state regime.
@@
 \centering
-\includegraphics[width=0.4\textwidth]{fig_thetaPEmvm.pdf}
+\includegraphics[width=0.4\textwidth]{fig_thetaPEmvm.png}
 \caption{
@@
   \caption{
-  Distance dependent recovery times for three disturbance types. When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases.
-  The `$*$' marks the value of $(m,\Delta\theta)$ below which there is a switch in subordinate/dominant states following extinction of the dominant population.
+  Recovery times under three disturbance types for systems where higher habitat heterogeneity $\Delta \theta$ corresponds to lower straying $m$ (see figure \ref{fig:mthetarelation}), and vice versa. 
+  % When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases.
+  The `$*$' marks the value of $m$ below which (low straying between highly heterogeneous environments) there is an inversion between subordinate/dominant states following extinction of the dominant population.
   DCB marks the discrete cusp bifurcation.
@@
 (b) The same as (a) but including recovery times when straying is density-dependent evaluated at $m^*$, shown by linked point pairs.
-Recovery times for systems with density-dependent straying are longer when straying is low and shorter when straying is high, mirroring the change in portfolio effects with respect to density dependent straying shown in figure \ref{fig:thetaPE}.
+Recovery times for systems with density-dependent straying are longer when straying is low and shorter when straying is high, mirroring the change in portfolio effects with respect to density-dependent straying shown in figure \ref{fig:thetaPE}.
 } \label{fig:relaxtheta}

### 2017-11-15 manuscript/FinalDraft3/msv6_arxiv.tex
fdraft 3

--- 2017-11-15 Manuscript/FinalDraft_rev/msv6.tex
@@
-\RequirePackage{lineno}
-\documentclass{revtex4}
+% \RequirePackage{lineno}
+\documentclass[twocolumn]{revtex4}
 %\usepackage[pdftex]{graphicx}
@@
 
-\linespread{1.5}
+\linespread{1}
 
@@
 % \pagewiselinenumbers
-\linenumbers
+% \linenumbers
 % \setlength\linenumbersep{3pt}
@@
 
-\centerline{Salmon metapopulations, Straying, Dispersal, Eco-evolutionary dynamics, Alternative stable states}
+% \centerline{Salmon metapopulations, Straying, Dispersal, Eco-evolutionary dynamics, Alternative stable states}
 \vspace{2mm}
@@
 % Thus, the dual nature of straying as both promoter of connections among metapopulation demes and potential eroder of locally adapted gene complexes highlights the interplay between ecological dynamics of connected populations and the evolutionary dynamics of mixed trait distributions that respond to heterogeneous local conditions.
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.45\textwidth]{fig_traj.pdf}
+\caption{
+(a) The steady-state densities of $N_1$ and $N_2$ vs. the stray rate $m$. Which population attains the low- or high-density state is random due to small applied fluctuations in the initial conditions.
+(b) The steady-state trait values measured as the offset from the local optimum $\theta_i - mu_i$, vs. the stray rate $m$. 
+DCB marks the discrete cusp bifurcation.
+Unless otherwise indicated, the default parameter values used are: $r_{\rm max}=2$; $Z=0.5$; $\beta=0.001$; $\theta_1=5$; $\Delta\theta=5$; $\tau=1$; $\sigma=1$; $T=1\times10^5$.
+} \label{fig:traj}
+\end{figure}
 
@@
 
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=1\textwidth]{fig_MDPE_hm.pdf}
+\caption{
+(a) Total means $N_t^*$, 
+(b) difference in means $\Delta N^*$, and 
+(c) the portfolio effect PE as a function of heritability $h^2$ and a constant stray rate $m$. Light colors = high values.
+The black line shows the cusp bifurcation separating a single steady-state (left) from alternative stable states (right).
+(d) The relationship between the time to recovery following a disturbance and the portfolio effect. Portfolio effects greater than unity corresponds to less synchronization.
+} \label{fig:PE}
+\end{figure*}
+
 \section{Model Description \& Analysis}
@@
 % \noindent{\bf (b) Recruitment over a selective landscape}
-Because individuals from the local population are mixed with individuals from the remote population via straying and subsequent reproduction, the resulting trait distribution is a complex mixture of trait distributions (see Supplemental Materials: Appendix I for the derivation, Fig.
-\ref{fig:PDF}).
-We make two simplifying assumptions.
-First, we approximate the distribution resulting from the mix of remote and local individuals following reproduction as a Gaussian where $X_i=x_i$ with probability $g(x_i)$.
-The expectation of the actual trait distribution as well as the Gaussian approximation are the same, such that ${\rm E}\{g(x_i)\} = w_i\mu_i + (1-w_i)\mu_j$, with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and straying individuals, $1-w_i$, where
+Because individuals from the local population are mixed with individuals from the remote population via straying and subsequent reproduction, the resulting trait distribution is a mixed normal with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and straying individuals, $1-w_i$, where
 \begin{equation}
@@
 \end{equation}
-We show that the Gaussian approximation is not only exact with respect to the mean, but also is a good approximation of variance in most relevant cases (see Appendix I in the Supplemental Materials, Fig.
-\ref{fig:PDF}).
+We made two simplifying assumptions.
+First, we assumed that the distribution resulting from the mix of remote and local individuals, following reproduction, is also normal with a mean value equal to that of the mean for the mixed-normal distribution.
 Thus, strays can successfully reproduce and introduce their genotypes into the recipient population, which is supported by observations in wild populations \citep{Jasper:2013cc}.
-Secondly, we assumed that changes in trait variance through time are minimal, such that $\sigma^2$ is constant over time, which is a common simplification in eco-evolutionary models of population dynamics \citep{Lande:1976ga,Schreiber:2011wx,Gilbert:2014ee,Gibert:2015kc}.
+Second, we assumed that changes in trait variance through time are minimal, such that $\sigma^2$ is constant over time, which is a common simplification in eco-evolutionary models of population dynamics \citep{Lande:1976ga,Schreiber:2011wx,Gilbert:2014ee,Gibert:2015kc}.
 
 % An increasing flow of incoming strays is generally expected to pull the mean trait value of the local population away from its optimum over time, which will decrease its rate of recruitment.
-Following Lande \citep{Lande:1976ga}, and given our assumption of trait distribution normality, the mean trait value thus changes through time according to the difference equation
+Following Lande \citep{Lande:1976ga}, the mean trait value thus changes through time according to the difference equation
 
@@
 
+\begin{table}
+\begin{center}
+  \caption{Parameters and definitions}
+\begin{tabular}{ l l }
+\hline
+Parameter & Definition \\
+\hline
+$N_i(t)$, $N_T(t)$ & Individual, aggregate population over time\\
+$x_i$ & Trait value for an individual in population $i$\\
+$\mu_i(t)$ & Mean of $x$ for population $i$ over time\\
+$\sigma^2$ & Genetic variance of trait $x$\\
+$m$, $m(t)$ & Constant, density-dependent straying rate\\
+$m_0$ & Straying rate of an individual\\
+$R_i[\mu(t)]$ & Recruitment rate of population $i$\\
+$r_{\rm max}$ & Maximum recruitment rate\\
+${\rm e}^{-Z}$ & Survival rate\\
+$\beta$ & Strength of density dependence\\
+$\theta_i$ & Optimal trait value for habitat $i$\\
+$\Delta\theta$ & Habitat heterogeneity\\
+$\tau$ & Strength of selection\\
+$h^2$ & Heritability\\
+$C$ & Half saturation constant for $m(t)$\\
+${\rm PE}$ & Portfolio Effect\\
+\hline
+\end{tabular}
+\end{center}
+\end{table}
+
 \noindent {\bf (b) Density-dependent straying}
@@
 For realistic population densities, $m(t) < m_0$.\\
+
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.9\textwidth]{fig_relax_lowh.pdf}
+\caption{
+Recovery time of $N_T^*$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady-state $m^*$) with trait heritability $h^2=0.2$.
+If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
+DCB marks the discrete cusp bifurcation.
+} \label{fig:relax}
+\end{figure*}
 
@@
 
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.5\textwidth]{fig_traitdiff.pdf}
+\caption{
+Phenotypic diversity ($\Delta \mu^*$) evaluated at the steady-state as a function of straying rate $m$ and trait heritability $h^2$.
+The jump occurs as the system crosses the cusp bifurcation; lower phenoytic diversity emerges with higher straying rates and in the alternative stable state regime.
+} \label{fig:traitdiff}
+\end{figure}
+
 \section{Results}
@@
 Whether a specific population goes to one state or the other in our model is random, due to a small amount of introduced variance in the initial conditions.
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.4\textwidth]{fig_thetaPEmvm.pdf}
+\caption{
+(a) Median portfolio effect as a function of a constant stray rate $m$ (solid line) and density-dependent stray rate (point pairs) given heritability is $h^2 < 0.5$ and $\Delta\theta=5$.
+Point pairs connected by a horizontal line represent the PE as a function of density-dependent straying rates, evaluated for both low- and high-density populations at equilibrium.
+The lower straying rate of a pair is for the larger population; the higher straying rate is for the smaller population.
+(b) Median portfolio effects for habitats with increasing heterogeneity as measured by the difference in regional trait optima $\Delta \theta$ for both constant and density-dependent stray rates as shown in (a).
+Portfolio effects greater than unity corresponds to less synchronization.
+DCB marks the discrete cusp bifurcation.
+} \label{fig:thetaPE}
+\end{figure}
 
@@
 %(for the discussion: an example of this situation may be the salmon populations during the last glacial maximum, where any mixing would be from geographically distant populations)
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+  \centering
+  \includegraphics[width=0.5\textwidth]{fig_mtheta_rt.pdf}
+  \caption{
+  Recovery times under three disturbance types for systems where higher habitat heterogeneity $\Delta \theta$ corresponds to lower straying $m$ (see figure \ref{fig:mthetarelation}), and vice versa. 
+  % When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases.
+  The `$*$' marks the value of $m$ below which (low straying between highly heterogeneous environments) there is an inversion between subordinate/dominant states following extinction of the dominant population.
+  DCB marks the discrete cusp bifurcation.
+  } \label{fig:mtheta}
+\end{figure}
 
@@
 
-\begin{table}[!t]
-\begin{center}
-\begin{tabular}{ l l l }
-\hline
-Parameter & Definition & Value \\
-\hline
-$N_i(t)$, $N_T(t)$ & Individual, aggregate population over time & {\rm var.}\\
-$x_i$ & Trait value for an individual in population $i$ & {\rm var.}\\
-$\mu_i(t)$ & Mean of $x$ for population $i$ over time & {\rm var.}\\
-$\sigma^2$ & Genetic variance of trait $x$ & 1.0\\
-$m$, $m(t)$ & Constant, density-dependent straying rate & {\rm var.}\\
-$m_0$ & Straying rate of an individual & {\rm var.}\\
-$R_i[\mu(t)]$ & Recruitment rate of population $i$ & {\rm var.}\\
-$r_{\rm max}$ & Maximum recruitment rate & 2.0\\
-${\rm e}^{-Z}$ & Survival rate & 0.6\\
-$\beta$ & Strength of density dependence & $10^{-3}$\\
-$\theta_i$ & Optimal trait value for habitat $i$ & 5.0\\
-$\Delta\theta$ & Habitat heterogeneity & {\rm var.}\\
-$\tau$ & Strength of selection & 1.0\\
-$h^2$ & Heritability & {\rm var.}\\
-$C$ & Half saturation constant for $m(t)$ &  $10^3$\\
-${\rm PE}$ & Portfolio Effect & {\rm var.}\\
-$T$ & Terminal simulation time & $10^5$\\
-\hline
-\end{tabular}
-\end{center}
-\caption{Table of parameters, definitions, and assigned values (var. = variable).}
-\end{table}
-
-% tmax=10000;
-% z=0.5;
-% rmax=2.0;
-% beta=0.001;
-% theta1=5.0;
-% thetadiff=5.0;
-% tau=1.0;
-% h=0.49;
-% sigmaE=0;
-% sigmaG=1.0;
-% m=0.35;
-% perror=0.01;
-
-\clearpage
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.4\textwidth]{fig_traj.pdf}
-\caption{
-(a) The steady-state densities of $N_1$ and $N_2$ vs. the stray rate $m$. Which population attains the low- or high-density state is random due to small applied fluctuations in the initial conditions.
-(b) The steady-state trait values measured as the offset from the local optimum $\theta_i - mu_i$, vs. the stray rate $m$. 
-DCB marks the discrete cusp bifurcation.
-% Unless otherwise indicated, the default parameter values used are: $r_{\rm max}=2$; $Z=0.5$; $\beta=0.001$; $\theta_1=5$; $\Delta\theta=5$; $\tau=1$; $\sigma=1$; $T=1\times10^5$.
-} \label{fig:traj}
-\end{figure}
-
-\begin{figure*}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=1\textwidth]{fig_MDPE_hm.pdf}
-\caption{
-(a) Total means $N_t^*$, 
-(b) difference in means $\Delta N^*$, and 
-(c) the portfolio effect PE as a function of heritability $h^2$ and a constant stray rate $m$. Light colors = high values.
-The black line shows the cusp bifurcation separating a single steady-state (left) from alternative stable states (right).
-(d) The relationship between the time to recovery following a disturbance and the portfolio effect. Portfolio effects greater than unity corresponds to less synchronization.
-} \label{fig:PE}
-\end{figure*}
-
-\begin{figure*}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.9\textwidth]{fig_relax_lowh.pdf}
-\caption{
-Recovery time of $N_T^*$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady-state $m^*$) with trait heritability $h^2=0.2$.
-If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
-DCB marks the discrete cusp bifurcation.
-} \label{fig:relax}
-\end{figure*}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.5\textwidth]{fig_traitdiff.pdf}
-\caption{
-Phenotypic diversity ($\Delta \mu^*$) evaluated at the steady-state as a function of straying rate $m$ and trait heritability $h^2$.
-The jump occurs as the system crosses the cusp bifurcation; lower phenoytic diversity emerges with higher straying rates and in the alternative stable state regime.
-} \label{fig:traitdiff}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.4\textwidth]{fig_thetaPEmvm.png}
-\caption{
-(a) Median portfolio effect as a function of a constant stray rate $m$ (solid line) and density-dependent stray rate (point pairs) given heritability is $h^2 < 0.5$ and $\Delta\theta=5$.
-Point pairs connected by a horizontal line represent the PE as a function of density-dependent straying rates, evaluated for both low- and high-density populations at equilibrium.
-The lower straying rate of a pair is for the larger population; the higher straying rate is for the smaller population.
-(b) Median portfolio effects for habitats with increasing heterogeneity as measured by the difference in regional trait optima $\Delta \theta$ for both constant and density-dependent stray rates as shown in (a).
-Portfolio effects greater than unity corresponds to less synchronization.
-DCB marks the discrete cusp bifurcation.
-} \label{fig:thetaPE}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-  \centering
-  \includegraphics[width=0.5\textwidth]{fig_mtheta_rt.pdf}
-  \caption{
-  Recovery times under three disturbance types for systems where higher habitat heterogeneity $\Delta \theta$ corresponds to lower straying $m$ (see figure \ref{fig:mthetarelation}), and vice versa. 
-  % When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases.
-  The `$*$' marks the value of $m$ below which (low straying between highly heterogeneous environments) there is an inversion between subordinate/dominant states following extinction of the dominant population.
-  DCB marks the discrete cusp bifurcation.
-  } \label{fig:mtheta}
-\end{figure}
+% \clearpage
 
@@
 
+\clearpage
+
 \beginsupplement
 
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-% \includegraphics[width=0.35\textwidth]{fig_mthetarelation.pdf}
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.35\textwidth]{fig_mthetarelation.pdf}
 \caption{
@@
 } \label{fig:mthetarelation}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-% \includegraphics[width=0.35\textwidth]{fig_recovery.pdf}
+\end{figure*}
+
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.35\textwidth]{fig_recovery.pdf}
 \caption{
@@
 } \label{fig:recovery}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-% \includegraphics[width=0.35\textwidth]{fig_eigs.pdf}
+\end{figure*}
+
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.35\textwidth]{fig_eigs.pdf}
 \caption{
@@
 } \label{fig:eigs}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-% \includegraphics[width=0.35\textwidth]{fig_hysteresis.pdf}
+\end{figure*}
+
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.35\textwidth]{fig_hysteresis.pdf}
 \caption{
@@
 } \label{fig:hysteresis}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-% \includegraphics[width=0.35\textwidth]{fig_relax_large.pdf}
+\end{figure*}
+
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.35\textwidth]{fig_relax_large.pdf}
 \caption{
@@
 } \label{fig:relaxtraj_hdlh}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-% \includegraphics[width=0.35\textwidth]{fig_relax_small.pdf}
+\end{figure*}
+
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.35\textwidth]{fig_relax_small.pdf}
 \caption{
@@
 } \label{fig:relaxtraj_ldlh}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-% \includegraphics[width=0.9\textwidth]{fig_relax_highh.pdf}
+\end{figure*}
+
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.9\textwidth]{fig_relax_highh.pdf}
 \caption{
@@
 } \label{fig:relax_highh}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-% \includegraphics[width=0.4\textwidth]{fig_thetadiffN.pdf}
+\end{figure*}
+
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.4\textwidth]{fig_thetadiffN.pdf}
 \caption{
@@
 Solid lines are for constant $m$; dashed lines are for density-dependent $m$.} \label{fig:thetadiffN}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-% \includegraphics[width=0.8\textwidth]{fig_relaxtheta2.pdf}
+\end{figure*}
+
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.8\textwidth]{fig_relaxtheta2.pdf}
 \caption{
@@
 } \label{fig:relaxtheta}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-  \centering
-  % \includegraphics[width=0.35\textwidth]{fig_MDPE_hm_mtheta_rt.pdf}
+\end{figure*}
+
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+  \centering
+  \includegraphics[width=0.35\textwidth]{fig_MDPE_hm_mtheta_rt.pdf}
   \caption{
@@
   } \label{fig:mthetaPE}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-% \includegraphics[width=0.35\textwidth]{fig_relax_inertia.pdf}
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.35\textwidth]{fig_relax_inertia.pdf}
 \caption{
@@
 } \label{fig:inertia}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-  \centering
-  % \includegraphics[width=0.8\textwidth]{fig_relax_mtheta.pdf}
+\end{figure*}
+
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+  \centering
+  \includegraphics[width=0.8\textwidth]{fig_relax_mtheta.pdf}
   \caption{
@@
   } \label{fig:mthetamvm}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-% \includegraphics[width=0.35\textwidth]{fig_asymdensity.pdf}
+\end{figure*}
+
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.35\textwidth]{fig_asymdensity.pdf}
 \caption{
@@
 } \label{fig:symmetry}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-% \includegraphics[width=0.35\textwidth]{fig_relax_both_lowh.pdf}
+\end{figure*}
+
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.35\textwidth]{fig_relax_both_lowh.pdf}
 \caption{
@@
 } \label{fig:relaxtraj_bothlh}
-\end{figure}
+\end{figure*}
 

### 2017-11-15 manuscript/FinalDraft3/msv7.tex
fdraft 3

--- 2017-11-15 Manuscript/FinalDraft_rev/msv6.tex
@@
 We first assumed that a proportion ${\rm e}^{-Z}$ of both populations survive such that the surviving aggregated population, composed of both local individuals (at site $i$) and incoming strays (from site $j$), is $\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z}$.
-Because local individuals will recruit differently than incoming strays, the recruitment of the aggregate must incorporate two recruitment functions, given by $\left(R_i[\mu_i(t)] (1-m)N_i(t) + R_i[\mu_j(t)] m N_j(t)\right)$.
+Because local individuals will recruit differently than incoming strays, the recruitment of the aggregate must incorporate two recruitment functions, given by $R_i[w_i\mu_i(t) + (1-w_i)\mu_j(t)]\left((1-m)N_i + mN_j\right)$ (Eq.
+\ref{eq:R}).
 This mix of individuals is subject to identical compensatory effects, which is determined by the parameter $\beta$.
@@
   &\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z} \\ \nonumber
-  &+ \left(R_i[\mu_i(t)] (1-m)N_i(t) + R_i[\mu_j(t)] m N_j(t)\right) \\ \nonumber
+  &+ R_i[w_i\mu_i(t) + (1-w_i)\mu_j(t)]\left((1-m)N_i + mN_j\right) \\ \nonumber
   &\times {\rm e}^{-\beta ((1-m)N_i(t) + m N_j(t))},
@@
 
-\noindent where the equation for $N_j$ mirrors that for $N_i$.
+\noindent where
+
+\begin{equation}
+w_i=\frac{(1-m)N_i(t)}{(1-m) N_i(t) + m N_j(t)},
+\end{equation} 
+and the equation for $N_j$ mirrors that for $N_i$.
 
@@
 \begin{align}
-  &R_i[\mu_i(t)] = \\ \nonumber
-  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i,\mu_i(t),\sigma^2) {\rm d}x_i +\tilde{P}_i\\ \nonumber
-  &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-\mu_i(t))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\} +\tilde{P}_i,
-  \label{eq:R}
+  &R_i[w_i\mu_i(t) + (1-w_i)\mu_j(t)] = \\ \nonumber
+  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i,w_i\mu_i(t) + (1-w_i)\mu_j(t),\sigma^2) {\rm d}x_i +\tilde{P}_i\\ \nonumber
+  &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-(w_i\mu_i(t) + (1-w_i)\mu_j(t)))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\} +\tilde{P}_i,
 \end{align}
@@
 First, we approximate the distribution resulting from the mix of remote and local individuals following reproduction as a Gaussian where $X_i=x_i$ with probability $g(x_i)$.
-The expectation of the actual trait distribution as well as the Gaussian approximation are the same, such that ${\rm E}\{g(x_i)\} = w_i\mu_i + (1-w_i)\mu_j$, with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and straying individuals, $1-w_i$, where
-\begin{equation}
-w_i=\frac{(1-m)N_i(t)}{(1-m) N_i(t) + m N_j(t)}.
-\end{equation}
+The expectation of the actual trait distribution as well as the Gaussian approximation are the same, such that ${\rm E}\{g(x_i)\} = w_i\mu_i + (1-w_i)\mu_j$, with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and straying individuals, $1-w_i$.
 We show that the Gaussian approximation is not only exact with respect to the mean, but also is a good approximation of variance in most relevant cases (see Appendix I in the Supplemental Materials, Fig.
@@
   \mu_i(t+1) &= w_i\mu_i(t) + (1-w_i)\mu_j(t) \\ \nonumber
-  &+ h^2\sigma^2\frac{\partial}{\partial \mu_i}\ln\left(w_i R_i[\mu_i(t)] + (1-w_i)R_i[\mu_j(t)]  \right),
+  &+ h^2\sigma^2\frac{\partial}{\partial \mu_i}\ln\left(R_i[w_i \mu_i(t)+ (1-w_i)\mu_j(t)] \right),
 \end{align}

### 2017-11-15 manuscript/FinalDraft3/supp.tex
fdraft 3

--- 2017-09-23 Manuscript/FinalDraft_rev/supp.tex
@@
 \beginsupplement
+
+\section*{I. Gaussian approximation for the mixed trait distribution}
+We assume a Gaussian distribution to approximate the more complex trait distribution that results when two populations with two distinct trait distributions are linked by gene flow and reproduction.
+Here and throughout, we refer to the resident population as $i$ and the straying population as $j$, though both populations stray simultaneously.
+The Gaussian approximation is assumed to have a mean that is equal to the mean of the mixed trait distribution $(1-m)\mu_i + m\mu_j$ with a constant variance $\sigma^2$, equal to the original trait variance prior to mixing.
+We employ this approximation because a mixed trait distribution that remixes and recombines (through reproduction) at each time-step becomes complicated very quickly, as we shall demonstrate.
+\emph{Importantly, because the evolutionary framework tracks only the mean and not the variability through time, this approximation is exact with respect to the first moment.}
+Accordingly, the only attribute that is approximate is the assumption of normality.
+Below we will show that a constant variance of $\sigma^2$ for the mixed trait distribution is a good approximation in most relevant cases, and becomes increasingly accurate when 1) generations become less overlapping (i.e.
+the case of semelparity), and 2) with density-dependent straying ratios.
+
+{\bf The trait distribution at $\bm{t=0}$.} At time $t=0$, it is assumed that individuals from two separate populations have not yet mixed. 
+Thus, the mean trait values for both populations are equal to the optimal trait values that accord to their respective habitats $\mu_i(t=0) = \theta_i$, and that their trait values are distributed as $X_i=x_i \sim g(x_i,\mu_i,\sigma) = \mathcal{N}(\mu_i,\sigma)$.
+Here and henceforth we use upper-case to denote random variables, and lower-case to denote specific values of random variables.
+
+{\bf The trait distribution at $\bm{t=1}$.} At time $t=1$, dispersal of individuals from site $j$ to site $i$, and vice versa, has begun.
+The proportion of individuals that disperse to and from each population is given by the stray ratio $m$.
+The calculation of the mixed distribution that results from dispersal and subsequent recruitment incorporates the following processes:
+\emph{1)} dispersal of strays into the local population from the remote population, and vice versa;
+\emph{2)} recruitment of juveniles that are produced from this mixed distribution (under the assumption of random mating); 
+\emph{3)} mortality of individuals in the adult class.
+These events occur in the order in which they are listed.
+
+The mixed trait distribution for site $i$ immediately following dispersal is simply a mixed Gaussian with weights determined by the proportion of the population that are residents $i$ and the proportion that are strays $j$, such that
+\begin{equation}
+  p(x_{\rm disperse}) = w_i g(x_i,\mu_i,\sigma) + (1-w_i)g(x_j,\mu_j,\sigma),
+\end{equation}
+where the proportion of the population $i$ composed of \emph{resident} individuals is
+\begin{equation}
+  w_i = \frac{(1-m)N_i}{(1-m)N_i + mN_j},
+\end{equation}
+and the proportion of population $i$ composed of strays is $(1-w_i)$.
+Once individuals in both habitats mix, they mate without preference.
+Recruits are assumed to have trait values intermediate to their parents such that $x_{\rm r} = (x_{1} + x_{2})/2$ where trait values $X_1=x_{1}\sim p(x_{\rm disperse})$ and $X_2=x_{2}\sim p(x_{\rm disperse})$.
+We thus assume that parental traits are drawn independently from the mixed distribution $p(x_{\rm disperse})$, conditional on the requirement that $x_2 = 2x_r - x_1$, which we get when rearranging the equation for $x_{\rm r}$.
+Incorporating this condition and that traits are independently drawn from the mix of residents and strays yields the probability density for the trait value of recruits,
+\begin{equation}
+  q(x_{\rm r}) = \int_{-\infty}^\infty 2p(x_1)p(2x_{\rm r}-x_1){\rm d}x_1,
+\end{equation}
+where the factor of 2 comes from...
+The mean of the recruit trait distribution is calculated to be
+\begin{equation}
+{\rm E}\{q(x_{\rm r})\}=(1-m)\mu_i + m\mu_j,
+\end{equation} 
+and the variance is
+\begin{equation}
+  {\rm Var}\{q(x_{\rm r})\} = \frac{m^2 N_j^2 \left(4 \mu_j^2+\sigma ^2\right)+(m-1)^2 N_i^2 \left(4 \mu_i^2+\sigma ^2\right)-(m-1) m N_i N_j \left(\mu_i^2+6 \mu_i \mu_j+\mu_j^2+2 \sigma ^2\right)}{2 (-m N_i+m N_j+N_i)^2}.
+\end{equation}
+
+The relative contribution of trait values from resident parents, stray parents, and recruits depends on the numbers of each within the mixed population.
+The number if residents is $(1-m)N_i$, and the number of strays is $mN_j$.
+However, because adult mortality occurs after reproduction, we must adjust these numbers to account for only those who survive, which is given by ${\rm e}^{-Z}$.
+As $Z\rightarrow \infty$, the population becomes semelparous, as the adult class gives way to new recruits at each timestep. 
+Accordingly, the number of adult residents and strays is given by $(1-m)N_i{\rm e}^{-Z}$ and  $mN_i{\rm e}^{-Z}$, respectively.
+In contrast, the number of recruits is given by the combined contribution of resident and straying individuals to recruit production as $N_r = ((1-m)N_1 R[\mu_i] + mN_2 R[\mu_j]){\rm e}^{(-\beta((1-m)N_i + mN_j))}$ where $R$ is the recruitment rate, which differs for resident vs.
+straying individuals (see manuscript, Eq 2).
+
+\begin{figure}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=0.7\textwidth]{fig_PDF.pdf}
+\caption{
+Clockwise from the upper-left.
+The calculated probability distribution function (PDF) for individuals in population $i$ after immigration from population $j$, recruitment, and mortality (blue) vs.
+the Gaussian approximation used for the dynamical system (green), with
+a) low stray ratio, large straying population, overlapping generations: $m=0.1$, $Z=0.5$;
+b) high stray ratio, overlapping generations: $m=0.5$, $Z=0.5$;
+c) high stray ratio, non-overlapping generations: $m=0.5$, $Z=1000$;
+d) density-dependent stray ratio, non-overlapping generations: $m=m_0(1-N_i/(C+N_i))$, $Z=1000$, where $m_0=0.2$ and $C=1000$.
+The mean and variability of both distributions is shown by a point (mean) and line ($\sigma$) at the base of each distribution.
+} \label{fig:PDF}
+\end{figure}
+
+The proportion of individuals from the resident, straying, and recruitment classes following mortality at site $i$ are then $c_i$, $c_j$, and $c_r$ respectively, where
+\begin{align}
+c_i &= \frac{(1-m)N_i{\rm e}^{-Z}}{(1-m)N_i{\rm e}^{-Z} + m N_j{\rm e}^{-Z} + N_r} \\ \nonumber
+c_j &= \frac{m N_j{\rm e}^{-Z}}{(1-m)N_i{\rm e}^{-Z} + m N_j{\rm e}^{-Z} + N_r} \\ \nonumber
+c_r &= \frac{N_r}{(1-m)N_i{\rm e}^{-Z} + m N_j{\rm e}^{-Z} + N_r}.
+\end{align}
+where $c_i + c_j + c_r= 1$.
+The trait value distribution for time $t=1$ is then calculated as a mix of trait distributions from 1) surviving residents, 2) surviving strays, and 3) their offspring, such that
+\begin{equation}
+  g_{{\rm next}}(x_i) = c_i g(x_i) + c_j g(x_j) + c_r q(x_r).
+\end{equation}
+The expectation of this distribution is
+\begin{equation}
+{\rm E}\{g_{\rm next}(x_i)\}=(1-m)\mu_i + m\mu_j,
+\end{equation} and the variance is
+% variance a that has a closed form solution, but that is not easily readable.
+\begin{align}
+  {\rm Var}\{g_{{\rm next}}(x_i)\}=&\frac{(\mu_i-\mu_j)^2 (2 (m-1)^2 N_i^2 c_j^2-(m-1) N_i c_j (2 (m-1) N_i+m N_j (1-4 c_i))}{2 (-m N_i+m N_j+N_i)^2} \\ \nonumber
+  &+ \frac{m N_j (c_i-1) (-m N_i+2 m N_j c_i+N_i))-\sigma ^2 (c_i+c_j+1) (-m N_i+m N_j+N_i)^2}{2 (-m N_i+m N_j+N_i)^2}.
+\end{align}
+
+Importantly, ${\rm E}\{g_{{\rm next}}(x_i)\}$ is exactly equal to the trait mean for the original mix, as well as the trait mean for recruits.
+Thus, a Gaussian approximation distributed as $\mathcal{N}(w_i\mu_i + (1-w_i)\mu_j,\sigma)$ will not alter the dynamics under the assumption of constant variance in Eq 4.
+Moreover, the mixed distribution described above becomes more complex at the next timestep, as two multi-modal trait distributions then mix and recombine through recruitment.
+Although the complexity of the true distribution is increased and becomes quickly intractable, one expects the actual form to eventually approximate a Gaussian based on the central limit theorem.
+One could make the argument that a Gaussian approximation does not represent the multi-modality in trait values inherent to a mixed distribution, and thus underestimates variance.
+Although the Gaussian approximation exactly captures the expectation, we next explore to what extent the approximation does and does not capture the variability expressed in $g(x_{\rm next})$.
+
+%Discussion of factors that increase or decrease accuracy of the approximation
+The Gaussian approximation $\mathcal{N}(w_i\mu_i + (1-w_i)\mu_j,\sigma)$ will generally underestimate the true variance of the mix, but not by much.
+If the straying ratio is low, (ca. 0.0 to 0.2), the Gaussian approximation tends to be very similar to the actual distribution (Fig. \ref{fig:PDF}a).
+As the straying ratio increases, so does the bimodality of the actual distribution, and this magnifies our underestimate of variance (Fig. \ref{fig:PDF}b).
+The interesting dynamics observed in our model generally occur between $m=0.05$ and $m=0.2$, where estimates of variance are similar.
+Decreasing the overlap between generations also leads to greater similarity in the actual distribution and the Gaussian approximation, and the potential bias introduced by bimodality will be less for cases of non-overlapping generations, or semelparity (Fig.
+\ref{fig:PDF}c).
+Most importantly, density-dependent straying increases the accuracy of our Gaussian approximation by suppressing the straying ratio (Fig. \ref{fig:PDF}d).
+
+Because the Gaussian approximation is exact to first order, and generally provides a good approximation of realized variance, we suggest that the use of Lande's formulation for the change in mean trait values through time under the assumption of constant variance is appropriate (see Eq.
+4 in main text).
+Because our estimate of variance is generally an underestimate, the rate of recruitment will be a slight overestimate (Eq.
+2, main text), and the timescale of trait evolution will generally be slower (Eq.
+4, main text).
+Changes in the rate of recruitment (Fig.
+\ref{fig:symmetry}) and heritability do not have an impact on our qualitative findings, and this suggests any bias that our approximation introduces by underestimating true variance is minimal.
+
+\section*{II. Additional Figures}
+
+% \begin{figure}
+%   \captionsetup{justification=raggedright,
+% singlelinecheck=false
+% }
+% \centering
+% \includegraphics[width=0.7\textwidth]{fig_contSD.pdf}
+% \caption{
+% Magnitude of difference in the standard deviation for the calculated PDF for individuals in population $i$ after 1) immigration from population $j$, 2) recruitment, and 3) adult mortality vs.
+the Gaussian approximation where $\sigma=1$.
+% Values $<1$ reflect underestimates of variability; values $>1$ reflect overestimates.
+% Top row is for constant straying ratio, and the bottom row is for density-dependent straying ratio under different treatments of habitat heterogeneity (left column $\Delta \theta=5$; right column $(\Delta \theta=8$).
+% } \label{fig:contourSD}
+% \end{figure}
+%
 

### 2017-11-15 model/KevinExtinction.jl

@@ -459,3 +459,177 @@ for (i in 1:length($mvec)) {
      
 dev.off()
 """
+
+
+
+
+
+#############
+# Calculate return time as a function of 'm' and 'Z'
+#############
+
+
+
+@everywhere using Distributions
+@everywhere using RCall
+@everywhere using JLD
+@everywhere using HDF5
+
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve_ddm.jl")
+
+
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolveExtinct.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolveExtinct_ddm.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/timeSS.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/movingaverage.jl")
+
+
+#Analysis over m
+tmax=10000;
+
+mvec = collect(0.0001:0.01:0.5);
+pvec = ["small","large","both"];
+h = 0.2;
+zvec = logspace(-1,log10(5),50);
+reps = 50;
+
+# z=0.5;
+rmax=2.0;
+beta=0.001;
+theta1=5.0;
+thetadiff=5;
+tau=1.0;
+C=1000;
+sigmaE=0;
+sigmaG=1;
+perror=0.01;
+refuge=0.01;
+
+burnin = 0.8;
+
+# n1v = SharedArray{Float64}(reps,length(mvec),length(pvec),tmax);
+# n2v = SharedArray{Float64}(reps,length(mvec),length(pvec),tmax);
+# x1v = SharedArray{Float64}(reps,length(mvec),length(pvec),tmax);
+# x2v = SharedArray{Float64}(reps,length(mvec),length(pvec),tmax);
+
+t_ext = Int64(round(tmax/2));
+
+
+rt = SharedArray{Float64}(reps,length(mvec),length(pvec),length(zvec));
+rt_ddm = SharedArray{Float64}(reps,length(mvec),length(pvec),length(zvec));
[…]

### 2017-11-15 model/KevinFinalRun.jl

@@ -1,0 +1,555 @@
+
+
+
+#Figure 1
+
+#Analysis over m
+tmax=10000;
+mvec = collect(0.0:0.001:0.5);
+n1ts = zeros(Float64,length(mvec),tmax);
+n2ts = zeros(Float64,length(mvec),tmax);
+n1mean=zeros(Float64,length(mvec));
+n2mean=zeros(Float64,length(mvec));
+n1sd=zeros(Float64,length(mvec));
+n2sd=zeros(Float64,length(mvec));
+aggmean=zeros(Float64,length(mvec));
+aggsd=zeros(Float64,length(mvec));
+x1mean=zeros(Float64,length(mvec));
+x2mean=zeros(Float64,length(mvec));
+pe=zeros(Float64,length(mvec));
+eigs = Array(Array{Complex{Float64}},length(mvec));
+maxeigs = Array{Float64}(length(mvec));
+maximeigs = Array{Float64}(length(mvec));
+
+z=0.5;
+rmax=2.0;
+beta=0.001;
+theta1=5.0;
+thetadiff=3;
+tau=1.0;
+h=0.2;
+sigmaE=0;
+sigmaG=1;
+perror=0.01;
+
+burnin=0.80
+@time for i=1:length(mvec)
+  m=mvec[i];
+  
+  n1, n2, x1, x2, w1, w2 = 
+  KevinEvolve(
+    tmax,
+    z,
+    rmax,
+    beta,
+    theta1,
+    thetadiff,
+    tau,
+    h,
+    sigmaE,
+    sigmaG,
+    m,
+    perror
+    );
+
+  n1trim = n1[Int64(floor(tmax*burnin)):tmax];
+  n2trim = n2[Int64(floor(tmax*burnin)):tmax];
+  x1trim = x1[Int64(floor(tmax*burnin)):tmax];
+  x2trim = x2[Int64(floor(tmax*burnin)):tmax]
+  
+  n1ts[i,:] = n1;
+  n2ts[i,:] = n2;
+  
+  n1mean[i] = mean(n1trim);
+  n2mean[i] = mean(n2trim);
+  n1sd[i] = std(n1trim);
+  n2sd[i] = std(n2trim);
+  
+  aggmean[i] = mean(n1trim+n2trim);
+  aggsd[i] = std(n1trim+n2trim);
+
+  x1mean[i] = theta1-mean(x1trim);
+  x2mean[i] = (theta1+thetadiff)-mean(x2trim);
+  
+  # #Calculate the Jacobian
[…]

### 2017-11-15 model/model2/Figure3_S8.jl

@@ -1,0 +1,226 @@
+#Figure 3
+#Figure S8
+using Distributions, RCall, JLD, HDF5
+
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve_ddm.jl")
+
+
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolveExtinct.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolveExtinct_ddm.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/timeSS.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/movingaverage.jl")
+
+
+#Analysis over m
+tmax=10000;
+
+mvec = collect(0.0001:0.001:0.5);
+pvec = ["small","large","both"];
+hvec = [0.2, 0.8]
+reps = 100;
+
+z=0.5;
+rmax=2.0;
+beta=0.001;
+theta1=5.0;
+thetadiff=5;
+tau=1.0;
+C=1000;
+sigmaE=0;
+sigmaG=1;
+perror=0.01;
+refuge=0.01;
+
+burnin = 0.8;
+
+# n1v = SharedArray{Float64}(reps,length(mvec),length(pvec),tmax);
+# n2v = SharedArray{Float64}(reps,length(mvec),length(pvec),tmax);
+# x1v = SharedArray{Float64}(reps,length(mvec),length(pvec),tmax);
+# x2v = SharedArray{Float64}(reps,length(mvec),length(pvec),tmax);
+
+t_ext = Int64(round(tmax/2));
+
+
+rt = SharedArray{Float64}(reps,length(mvec),length(pvec),length(hvec));
+rt_ddm = SharedArray{Float64}(reps,length(mvec),length(pvec),length(hvec));
+m1mean = SharedArray{Float64}(reps,length(mvec),length(pvec),length(hvec));
[…]

### 2017-11-15 model/model2/Figure5.jl

@@ -1,0 +1,654 @@
+#Figure 5
+
+
+using Distributions
+using RCall
+using HDF5
+using JLD
+
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/movingaverage.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve_ddm.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/qualsfunc.jl")
+
+
+
+# Figure 5
+
+#Analysis over m & theta divergence
+mvec=collect(0.0:0.001:0.45);
+hvec = collect(0.0:0.01:1.0);
+
+
+n1mean=SharedArray(Float64,length(hvec),length(mvec));
+n2mean=SharedArray(Float64,length(hvec),length(mvec));
+x1mean=SharedArray(Float64,length(hvec),length(mvec));
+x2mean=SharedArray(Float64,length(hvec),length(mvec));
+pe=SharedArray(Float64,length(hvec),length(mvec));
+
+tmax=10000;
+z=0.5;
+rmax=2.0;
+beta=0.001;
+theta1=5.0;
+thetadiff=5.0;
+tau=1.0;
+sigmaE=0;
+sigmaG=1;
+
+perror=0.01;
+
+@sync @parallel for k=1:length(hvec)
+    
+    h = hvec[k];
+    
+    for j=1:length(mvec)
+      m = mvec[j];
+      
+      n1, n2, x1, x2, w1, w2 = 
+      KevinEvolve(
+        tmax,
+        z,
+        rmax,
+        beta,
+        theta1,
+        thetadiff,
+        tau,
+        h,
+        sigmaE,
+        sigmaG,
+        m,
+        perror
+        );
+      burnin=0.8
+      n1trim = n1[Int64(floor(tmax*burnin)):tmax];
+      n2trim = n2[Int64(floor(tmax*burnin)):tmax];
+      x1trim = x1[Int64(floor(tmax*burnin)):tmax];
[…]

### 2017-11-15 model/model2/FigureS2_S4_S5.jl

@@ -1,0 +1,275 @@
+#Figure S2
+#Figure S4
+
+using RCall
+
+
+thetascale = 2;
+m=collect(0:0.01:0.5);
+thetadiff=zeros(length(m));
+for i = 1:length(m)
+    thetadiff[i] = (1-2*m[i])/(thetascale*m[i])
+end
+
+namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/FinalDraft_rev/fig_mthetarelation.pdf");
+R"""
+pdf($namespace,height=5,width=6)
+plot($thetadiff,$m,type='l',ylab='Straying rate m',xlab=expression(paste('Habitat heterogeneity ',Delta,theta)),ylim=c(0,0.5))
+dev.off()
+"""
+
+using Distributions
+using RCall
+
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve.jl")
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinJacobian.jl")
+
+
+
+#Analysis over m
+tmax=10000;
+mvec = collect(0.0:0.0001:1);
+n1ts = zeros(Float64,length(mvec),tmax);
+n2ts = zeros(Float64,length(mvec),tmax);
+n1mean=zeros(Float64,length(mvec));
+n2mean=zeros(Float64,length(mvec));
+n1sd=zeros(Float64,length(mvec));
+n2sd=zeros(Float64,length(mvec));
+aggmean=zeros(Float64,length(mvec));
+aggsd=zeros(Float64,length(mvec));
+x1mean=zeros(Float64,length(mvec));
+x2mean=zeros(Float64,length(mvec));
+pe=zeros(Float64,length(mvec));
+eigs = Array(Array{Complex{Float64}},length(mvec));
+maxeigs = Array{Float64}(length(mvec));
+mineigs = Array{Float64}(length(mvec));
+maximeigs = Array{Float64}(length(mvec));
+minimeigs = Array{Float64}(length(mvec));
+
+z=0.5;
+rmax=2.0;
+beta=0.001;
+theta1=5.0;
+thetadiff=4;
+tau=1.0;
+h=0.5;
+sigmaE=0;
+sigmaG=1;
+perror=0.01;
+
+burnin=0.80
[…]

### 2017-11-15 model/model2/FigureS3_S6_S7_S12_S15.jl

@@ -1,0 +1,157 @@
+#Figure S12
+
+#Note Figures S6 and S7 have to be recreated and renamed!
+
+
+using Distributions
+using RCall
+
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve_mtheta.jl")
+
+
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve.jl")
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve_ddm.jl")
+
+
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolveExtinct.jl")
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/timeSS.jl")
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/movingaverage.jl")
+
+
+
+tmax=10000;
+z=0.5;
+rmax=2.0;
+beta=0.001;
+theta1=5.0;
+thetascale=2.0;
+tau=1.0;
+h=0.2;
+sigmaE=0;
+sigmaG=1;
+perror=0.01;
+refuge=0.01;
+m=0.02;
+
+thetadiff = (1-2*m)/(thetascale*m);
+
+extpop = "large";
+t_ext = Int64(round(tmax/2));
+
+n1, n2, x1, x2, w1, w2 = 
+KevinEvolveExtinct(tmax, 
+z, 
+rmax,
+beta,
+theta1,
+thetadiff,
+tau,
+h,
+sigmaE,
+sigmaG,
+m,
+perror,
+extpop,
+t_ext,
+refuge
+);
+
+t_ss, relaxtime = timeSS(n1,n2,t_ext);
+
+
+R"""
+library(RColorBrewer)
+pal = brewer.pal(3,'Set1')
+par(mfrow=c(2,1),mai = c(0.8, 0.8, 0.4, 0.1))
+plot($n1,type='l',col=pal[1],xlim=c($t_ext,$t_ss+50),ylim=c(0,max($n1,$n2)),ylab='Population density',xlab='Time',main=paste(c('RT=',$relaxtime)))
+lines($n2,col=pal[2])
+lines(c($t_ss,$t_ss),c(-10,10000))
[…]

### 2017-11-15 model/model2/FigureS5_S9_S10_S14.jl

@@ -1,0 +1,501 @@
+using RCall
+using Distributions
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolveSS.jl")
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinJacobian.jl")
+
+
+#Analysis over m
+tmax=10000;
+mvec1 = collect(0.0:0.0001:0.15);
+mvec = [mvec1 ; reverse(mvec1)];
+n1ts = zeros(Float64,length(mvec),tmax);
+n2ts = zeros(Float64,length(mvec),tmax);
+n1mean=zeros(Float64,length(mvec));
+n2mean=zeros(Float64,length(mvec));
+n1sd=zeros(Float64,length(mvec));
+n2sd=zeros(Float64,length(mvec));
+aggmean=zeros(Float64,length(mvec));
+aggsd=zeros(Float64,length(mvec));
+x1mean=zeros(Float64,length(mvec));
+x2mean=zeros(Float64,length(mvec));
+pe=zeros(Float64,length(mvec));
+eigs = Array(Array{Complex{Float64}},length(mvec));
+maxeigs = Array{Float64}(length(mvec));
+maximeigs = Array{Float64}(length(mvec));
+mineigs = Array{Float64}(length(mvec));
+minimeigs = Array{Float64}(length(mvec));
+
+z=0.5;
+rmax=2.0;
+beta=0.001;
+theta1=5.0;
+thetadiff=6;
+tau=1;
+h=0.7;
+sigmaE=0;
+sigmaG=1;
+perror=0.01;
+
+burnin=0.80
+@time for i=1:length(mvec)
+  
+  m=mvec[i];
+  
+  if i == 1
+      n0 = [2,2];
+    #   x0 = [theta1 + rand(Normal(0,0.0001)),(theta1 + thetadiff)  + rand(Normal(0,0.0001))];
+    x0 = [theta1 ,(theta1 + thetadiff)];
+  else
+      n0 = [n1mean[i-1],n2mean[i-1]];
+      x0 = [x1mean[i-1],x2mean[i-1]];
+  end
+  
+  n1, n2, x1, x2, w1, w2 = 
+  KevinEvolveSS(
+    n0,
+    x0,
+    tmax,
+    z,
+    rmax,
+    beta,
+    theta1,
+    thetadiff,
+    tau,
+    h,
+    sigmaE,
+    sigmaG,
+    m,
+    perror
+    );
+
[…]

### 2017-11-15 model/src/KevinEvolve.jl

@@ -38,27 +38,41 @@ function KevinEvolve(
   
   for t=1:tmax-1
     
+    w1[t] = (n1[t]-m*n1[t])/(n1[t]-m*n1[t]+m*n2[t]);
+      
+    w2[t] = (n2[t]-m*n2[t])/(n2[t]-m*n2[t]+m*n1[t]);
+    
+    # n1[t+1] = ((1-m)*n1[t] + m*n2[t])*exp(-z) +
+    # ((rfunc(x1[t],theta1) + rand(pdist))*(n1[t]-m*n1[t]) +
+    # (rfunc(x2[t],theta1) + rand(pdist))*m*n2[t]) * 
+    # exp(-beta*((1-m)*n1[t]+m*n2[t]));
+    # 
+    # n2[t+1] = ((1-m)*n2[t] + m*n1[t])*exp(-z) +
+    # ((rfunc(x2[t],(theta1 + thetadiff)) + rand(pdist))*(n2[t]-m*n2[t]) +
+    # (rfunc(x1[t],(theta1 + thetadiff)) + rand(pdist))*m*n1[t]) * 
+    # exp(-beta*((1-m)*n2[t]+m*n1[t]));
+    
     n1[t+1] = ((1-m)*n1[t] + m*n2[t])*exp(-z) +
-    ((rfunc(x1[t],theta1) + rand(pdist))*(n1[t]-m*n1[t]) +
-    (rfunc(x2[t],theta1) + rand(pdist))*m*n2[t]) * 
+    ((rfunc(w1[t]*x1[t] + (1-w1[t])*x2[t],theta1) + rand(pdist))*((1-m)*n1[t]+m*n2[t])) * 
     exp(-beta*((1-m)*n1[t]+m*n2[t]));
     
     n2[t+1] = ((1-m)*n2[t] + m*n1[t])*exp(-z) +
-    ((rfunc(x2[t],(theta1 + thetadiff)) + rand(pdist))*(n2[t]-m*n2[t]) +
-    (rfunc(x1[t],(theta1 + thetadiff)) + rand(pdist))*m*n1[t]) * 
+    ((rfunc(w2[t]*x2[t] + (1-w2[t])*x1[t],(theta1 + thetadiff)) + rand(pdist))*((1-m)*n2[t]+m*n1[t])) * 
[…]

### 2017-11-15 model/src/gausapprox.jl

@@ -1,0 +1,139 @@
+using Distributions, RCall
+
+
+#simulated mixture
+m=0.1;
+
+tau = 1;
+theta1 = 0;
+thetadiff = 5;
+theta2 = theta1+thetadiff;
+rmax = 2;
+tau = 1;
+sigmaE=0;
+sigmaG=1;
+beta = 0.001;
+Z=0.5;
+
+N1 = 10;
+N2 = 10;
+
+tmax = 100;
+
+
+rfunc = function(mu,theta)
+  r = rmax*tau*(exp(-((theta-mu)^2)/(2*((sigmaE^2 + sigmaG^2)+tau^2))))/(sqrt(tau^2+(sigmaE^2 + sigmaG^2)))
+end
+
+
+mu1 = theta1;
+mu2 = mu1+thetadiff;
+
+
+
+d1 = Normal(mu1,1);
+d2 = Normal(mu2,1);
+t1 = rand(d1,N1);
+t2 = rand(d2,N2);
+
+Pop1 = Array{Int64}(tmax);
+Pop2 = Array{Int64}(tmax);
+tnext1 = Array{Array{Float64}}(tmax);
+tnext2 = Array{Array{Float64}}(tmax);
+a_mean1 = Array{Float64}(tmax);
+a_mean2 = Array{Float64}(tmax);
+for t = 1:tmax
+    #the number of strays leaving site 2 to site 1
+    N1s = Int64(floor(N2*m));
+    #the number of strays leaving site 1 to site 2
+    N2s = Int64(floor(N1*m));
+    
+    #Proportion of the site that is local
+    w1 = ((1-m)*N1)/((1-m)*N1 + m*N2);
+    w2 = ((1-m)*N2)/((1-m)*N2 + m*N1);
+    
+    #strays leave the population
+    N1 = N1 - N2s;
+    N2 = N2 - N1s;
+    
+    t1_orig = copy(t1);
+    t2_orig = copy(t2);
+    
+    #mixed trait distribution after dispersal
+    t1strayloc = sort(unique(rand(collect(1:N1),N2s)));
+    t2strayloc = sort(unique(rand(collect(1:N2),N1s)));
+    
+    t1_minus_strays = deleteat!(t1,t1strayloc);
+    t2_minus_strays = deleteat!(t2,t2strayloc);
+    
+    td1 = [t1_minus_strays ; t2_orig[t2strayloc]];
+    
[…]

### 2017-11-17 manuscript/FinalDraft3/eq_4.tex
figures revision 3

\documentclass[a4paper]{article}

%% Language and font encodings
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}

%% Sets page size and margins
\usepackage[a4paper,top=3cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

%% Useful packages
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}

\title{Infamous Equation 4}
\author{Jean P. Gibert}

\begin{document}
\maketitle

\section{Phenotypic evolution in two sites}
In this paper, we aim at understanding how straying may affect phenotypic evolution, and how in turn this evolutionary change may lead to changes in ecological dynamics as they unfold. To do so, we build upon ideas from classic papers (Lande 1976, most notably). In what follows, we show how the application of Lande's equation for a situation like the one we are keeping track of leads to overly-complicated equations of trait evolution that make our model difficult to track, and our understanding of the processes in play, difficult to achieve.

\subsection{Lande's classic work}
To keep track of the evolution of a phenotypic trait $z$ with probability density function $p$ and mean $\overline{z}$ and variance $\sigma^{2}$, we need a few basic ingredients. First, we define the mean fitness,$\overline{W}$ of the population of individuals bearing the trait $z$ as:

\begin{equation}\label{1}
\overline{W}=\int_{} p(z)W(z)\,dz.
\end{equation}
We also define the mean phenotype in the population as

\begin{equation}\label{2}
\overline{z}=\int_{} zp(z)\,dz,
\end{equation}
and the mean phenotype after selection (but before reproduction), $\overline{z}_{w}$ (Falconer 1960), as:

\begin{equation}\label{3}
\overline{z}_{w}=\frac{1}{\overline{W}}\int_{} zp(z)W(z)\,dz.
\end{equation}
Then, we can assess how the change in the phenotype $z$ occurs over time following
\begin{equation}\label{4}
\Delta z=z(t+1)-z(t),
\end{equation}
which can be rewritten as (Falconer 1960, Lande 1976):
\begin{equation}\label{5}
\Delta z=h^{2}(\overline{z}_{w}-z(t)),
\end{equation}
where $h^{2}$ is the heritability of the trait, and the difference $\overline{z}_{w}-z(t)$ is the classical "selection differential". In other words, \ref{5} is none other that the Breeder's equation.

To understand how $z$ changes over time, we first need to understand how the mean fitness of the population changes with a change in the mean trait value or $\frac{\delta \overline{W}}{\delta \overline{z}}$ (i.e. the shape of the "adaptive landscape"). Using \ref{1} we have:

\begin{equation}\label{6}
\frac{\delta \overline{W}}{\delta \overline{z}}=\frac{\delta}{\delta \overline{z}}\int_{} p(z)W(z)\,dz.
\end{equation}
Using Leibniz rule we can pass the derivative under the integral sign, 
\begin{equation}\label{7}
\frac{\delta \overline{W}}{\delta \overline{z}}=\int_{} \frac{\delta p(z)}{\delta \overline{z}}W(z)\,dz,
\end{equation}
and assuming $p$ is Gaussian (i.e. of the form $\frac{1}{\sqrt{2 \pi \sigma^{2}}} e^{-\frac{1}{2}\frac{(z-\overline{z})^{2}}{\sigma^{2}}} $), we calculate the derivative of $p$ with respect to $\overline{z}$, which leads to:

\begin{equation}\label{8}
\frac{\delta \overline{W}}{\delta \overline{z}}=\int_{} \Big[\frac{z-\overline{z}}{\sigma^{2}}\Big] p(z)W(z)\,dz.
\end{equation}
We can now expand in \ref{8} to get,
\begin{equation}\label{9}
\frac{\delta \overline{W}}{\delta \overline{z}}=
\frac{1}{\sigma^{2}}\int_{}zp(z)W(z)\,dz-\frac{\overline{z}}{\sigma^{2}}\int_{}p(z)W(z)\,dz,
\end{equation}
which making use of equations \ref{1} and \ref{3}, becomes $\frac{\delta \overline{W}}{\delta \overline{z}}=
\frac{\overline{W}}{\sigma^{2}}\overline{z}_{w}-\frac{\overline{z}}{\sigma^{2}}\overline{W}$, which can be factored as $\frac{\delta \overline{W}}{\delta \overline{z}}=
\frac{\overline{W}}{\sigma^{2}}(\overline{z}_{w}-\overline{z})$. Using \ref{5}, we can rearrange this expression to obtain an equation that related the change in the average phenotype from one time step to the next to the amount of heritable variation, $\sigma^{2}h^{2}$, and the adaptive landscape $\frac{\delta \overline{W}}{\delta \overline{z}}$:
\begin{equation}\label{10}
\Delta \overline{z}=\frac{\sigma^{2}h^{2}}{\overline{W}}\frac{\delta \overline{W}}{\delta \overline{z}}=\sigma^{2}h^{2}\frac{\delta \ln\overline{W}}{\delta \overline{z}}.
\end{equation}
Equation \ref{10} (Lande 1976's equation 7), is a staple of evolutionary biology. It is also possible to use \ref{10} and \ref{4} to write a recurrence relationship that allows to simulating evolutionary change over time in a trait $z$ as:
\begin{equation}\label{11}
\overline{z}(t+1)=\overline{z}(t)+\sigma^{2}h^{2}\frac{\delta \ln\overline{W}}{\delta \overline{z}}.
\end{equation}

\subsection{Using Lande's approach for two populations}
In what follows we show how Lande's approach can be used in the case this paper focuses on (two populations with one trait but two different means in each site) to calculate the magnitude of the phenotypic change analytically, why it is not necessarily useful to use the obtained expression in a simulation context, and what we can do to circumvent that issue.

In our case, we have two populations that exchange migrants, with a normally distributed trait $z$ that controls recruitment (as explained in the main text), with each population having a different mean, $\mu_{1}$ and $\mu_{2}$, but same variance $\sigma^{2}$. For simplicity, we will focus on population 1, but all the results we present in what follows are analogous in population 2 as well.

When individuals from population 2 stray into population 1, and individuals of population 1 stray into population 2, the phenotypic distribution of trait $z$ in population 1 is a mixture distribution, $p(z)$, of the form:

\begin{equation}\label{12}
p(z)=\omega p(z,\mu_{1})+(1-\omega)p(z,\mu_{2}).
\end{equation}
Such distribution has a mean $\overline{z}=\omega \mu_{1}+(1-\omega)\mu_{2}$. As before, we define the mean fitness as in equation \ref{1}, only that now $p$ is a mixture distribution. Let us use Lande's approach to obtain an expression for the change in the mean phenotype of the mixture distribution from one time step to the next. We thus ask, what is the change in $\overline{W}$ with a change in $\overline{z}$? We can use Lande's approach all the way to equation \ref{7}, then replace $p(z)$ by the mixture (equation \ref{12}):

\begin{equation}\label{13}
\frac{\delta \overline{W}}{\delta \overline{z}}=\int_{} \frac{\delta (\omega p(z,\mu_{1})+(1-\omega)p(z,\mu_{2}))}{\delta \overline{z}}W(z)\,dz.
\end{equation}
To take the derivative in equation \ref{13}, we need a change of variables, which is provided by the fact that $\overline{z}=\omega \mu_{1}+(1-\omega)\mu_{2}$:
\begin{equation} \label{14}
	\begin{cases} 
		\mu_{1}=\frac{\overline{z}-(1-\omega)\mu_{2}}{\omega}\\
		\mu_{2}=\frac{\overline{z}-\omega\mu_{1}}{1-\omega}\\
	\end{cases}.
\end{equation}
We can thus replace equation \ref{14} into \ref{13},
\begin{equation}\label{15}
\frac{\delta \overline{W}}{\delta \overline{z}}=\int_{} \frac{\delta (\omega p(z,\frac{\overline{z}-(1-\omega)\mu_{2}}{\omega})+(1-\omega)p(z,\frac{\overline{z}-\omega\mu_{1}}{1-\omega}))}{\delta \overline{z}}W(z)\,dz,
\end{equation} 
which, after expanding and taking the derivatives with respect to $\overline{z}$ becomes:

\begin{multline}\label{16}
\frac{\delta \overline{W}}{\delta \overline{z}}=\int_{} \omega\Bigg[\frac{z-\frac{\overline{z}-(1-\omega)\mu_{2}}{\omega}}{\omega \sigma^{2}}\Bigg] p(z,\frac{\overline{z}-(1-\omega)\mu_{2}}{\omega})W(z)\,dz + \\ + \int_{} (1-\omega)\Bigg[\frac{z-\frac{\overline{z}-\omega\mu_{1}}{1-\omega}}{(1-\omega) \sigma^{2}}\Bigg] p(z,\frac{\overline{z}-\omega\mu_{1}}{1-\omega})W(z)\,dz.
\end{multline}
To simplify notation, at this point we can come back to the original variables $\mu_{1}$ and $\mu_{2}$. We then expand and collect terms to get:
\begin{multline}\label{17}
\frac{\delta \overline{W}}{\delta \overline{z}}=\frac{1}{\sigma^{2}}\Bigg[\int_{}zp(z,\mu_{1})W(z)\,dz + \mu_{1}\int_{}p(z,\mu_{1})W(z)\,dz + \\ + \int_{}zp(z,\mu_{2})W(z)\,dz + \mu_{2}\int_{}p(z,\mu_{2})W(z)\,dz\Bigg].
\end{multline}
Using equations \ref{1} and \ref{3}, we can define the following quantities,

\begin{equation}\label{18}
\overline{W}_{1}=\int_{} p(z,\mu_{1})W(z)\,dz 
\end{equation}
\begin{equation}\label{19}
\overline{W}_{2}=\int_{} p(z,\mu_{2})W(z)\,dz  
\end{equation}
\begin{equation}\label{20}
\overline{z}_{w,1}=\frac{1}{\overline{W}}\int_{}zp(z,\mu_{1})W(z)\,dz
\end{equation}
\begin{equation}\label{21}
\overline{z}_{w,2}=\frac{1}{\overline{W}}\int_{}zp(z,\mu_{2})W(z)\,dz, 
\end{equation}
then replace them into equation \ref{17}:
\begin{equation}\label{22}
\begin{aligned}
\frac{\delta \overline{W}}{\delta \overline{z}} &=\frac{1}{\sigma^{2}}\Big[\overline{z}_{w,1} \overline{W}_{1} + \mu_{1}\overline{W}_{1} + \overline{z}_{w,2} \overline{W}_{2} + \mu_{2}\overline{W}_{2}\Big]  \\ 
&= \frac{1}{\sigma^{2}}\Big[ \overline{W}_{1}(\overline{z}_{w,1}-\mu_{1}) + \overline{W}_{2}(\overline{z}_{w,2}-\mu_{2})\Big].
\end{aligned}
\end{equation}
Using equation \ref{5} and rearranging terms we get:
\begin{equation}\label{23}
\overline{W}_{1}\Delta \mu_{1} + \overline{W}_{2}\Delta\mu_{2} = \sigma^{2}h^{2}\frac{\delta \overline{W}}{\delta \overline{z}} ,
\end{equation}
with $\overline{W}=\omega\overline{W}_{1}+(1-\omega)\overline{W}_{2}$. Equation \ref{23} relates how changes in either $\mu_{1}$ or $\mu_{2}$ lead to changes in fitness, and viceversa. However, it doesn't tell us anything as to how straying may lead to evolutionary change, and how that may feedback onto ecological dynamics. Because of this, we will once again use the change of variables in \ref{14}, to rewrite \ref{23} as:
\begin{equation}\label{24}
\overline{W}_{1}\Big(\frac{\Delta\overline{z}-(1-\omega)\Delta\mu_{2}}{\omega} \Big)+\overline{W}_{2}\Big(\frac{\Delta\overline{z}-\Delta\mu_{1}}{1-\omega} \Big)= \sigma^{2}h^{2}\frac{\delta \overline{W}}{\delta \overline{z}}, 
\end{equation}
which, can be simplified and rearranged to obtain an expression for the change in the mean phenotype of the mixture distribution,
\begin{equation}\label{25}
\Delta\overline{z}=\sigma^{2}h^{2}\frac{\delta \overline{W}}{\delta \overline{z}}\Big[\frac{(1-\omega)}{\omega}\overline{W}_{1}\Delta\mu_{2}+\frac{\omega}{1-\omega}\overline{W}_{2}\Delta\mu_{1} \Big]\frac{\omega}{\overline{W}_{1}+\overline{W}_{2}}.
\end{equation}
By multiplying \ref{25} by $\frac{\overline{W}}{\overline{W}}$, we can further rewrite the equation as:
\begin{equation}\label{26}
\Delta\overline{z}=\sigma^{2}h^{2}\frac{\delta \ln \overline{W}}{\delta \overline{z}}\Big[\frac{(1-\omega)}{\omega}\overline{W}_{1}\Delta\mu_{2}+\frac{\omega}{1-\omega}\overline{W}_{2}\Delta\mu_{1} \Big]\frac{\omega \overline{W}}{\overline{W}_{1}+\overline{W}_{2}},
\end{equation}
which has a similar form to Lande's equation (\ref{10}), but shows the explicit dependence of the change in the mean of the mix on the proportion of straying individuals. A similar expression can be derived for the other site.

While some understanding can be gained using equation \ref{26}, and we now present that in our main text, the expression does not easily allow us to simulate the change in $\overline{z}$ over time because the change in $\mu_{1}$ and $\mu_{2}$ is difficult to track in closed form. Moreover, this equation only holds for the first time step, but as new migrant arrive and the trait distributions get updated by the influx of individuals from the other site, the equation does not hold anymore.

This leaves us between a rock and a hard place: we can either track this change exactly for one generation, or we can do so through an approximation. Given that our paper aims to gain understanding about the feedback between evolutionary and ecological processes in salmon populations with spatial structure, we decided to simplify our problem by assuming that, as new migrants arrive, the mixture distribution can be locally approximated by a gaussian distribution whose mean equals the mean of the mix. Below, we show why this approximation, albeit coarse at first view, actually does a good job at describing the phenotypic distributions found in our model system under some conditions. Also, because of this assumption, we can use Lande's equation (here equation \ref{10}), without violating the original assumptions that led to its current form.

% For simplicity, let us assume the existence of a population of organisms with a normally distributed trait $z$, with mean $\mu$ and variance $\sigma^{2}$. Further assuming that the trait $z$ controls recruitment, $R$, and is under stabilizing selection, we show in the main text that recruitment is a function of the trait mean, $\mu$, of the form:

% \begin{equation}\label{recruitment_func}
% 	R(\mu)=\frac{r_{max}\tau}{\sqrt{\sigma^{2}+\tau^{2}}} e^{-\frac{(\theta-\mu)^{2}}{2(\sigma^{2}+\tau^{2})}}    .
% \end{equation}

% Under these assumptions, the evolution of $\mu$ over time is controlled by the following recurrence relationship:

% \begin{equation}\label{recurrenceONE}
% \mu_{t+1}=\mu_{t}+S,
% \end{equation}
% where $S$ has been shown by Lande (1976) to be equal to,

% \begin{equation}\label{selectionLande}
% \sigma^{2}h^{2}\frac{\partial \ln(W)}{\partial \mu_{t}} 
% \end{equation}
% with $\sigma^{2}h^{2}$ being the total heritable variation of trait $z$, and $W$ being the average fitness of the organisms. In this paper, we define our recruitment function as our measure of fitness, which allows us to write \ref{recurrenceONE} as:

% \begin{equation}\label{recurrenceTWO}
% \mu_{t+1}=\mu_{t}+\sigma^{2}h^{2}\frac{\partial \ln(R(\mu_{t}))}{\partial \mu_{t}} .
% \end{equation}

% Let us now assume the existence of two populations, as is done in the main text, where both population share the trait $z$, but with different means $\mu_1$ and $\mu_2$ (variance is kept the same for both populations). While we do not contend that equation \ref{recurrenceTWO} holds in a situation where $z$ is not normally distributed –it simply does not– we notice our simplifying assumption, once one population receives migrants from the second population (quoting from the main text): "we assume that the distribution resulting from the mix of remote and local individuals, following reproduction, is also normal with a mean value being that of the mixed-normal". Under these conditions, recruitment in the focal population (population 1) depends on: 1) recruitment from the local individuals, which now make up a proportion $\omega$ of the population, and 2) recruitment from the individuals that came from the other populations, which now make up a proportion $1-\omega$ of the focal population. In other words, recruitment in the focal population is now $\omega R(\mu_{1})+(1-\omega) R(\mu_{1})$. Because we assume that immediately after reproduction the ensuing recruits have a trait distribution that is normally distributed, that allows us to simply replace $\omega R(\mu_{1})+(1-\omega) R(\mu_{1})$ as our fitness in \ref{recurrenceTWO}, to obtain equation XXX
%  of the main text.

\end{document}

### 2017-11-17 manuscript/FinalDraft3/supp.tex
figures revision 3

@@
 %\title{Community assembly and dynamics by the deconstruction of species interactions}
-\title{Supplementary Figures: Eco-evolutionary dynamics and collective dispersal: implications for salmon metapopulation robustness}
+\title{Supplementary Materials: Eco-evolutionary dynamics and collective dispersal: implications for salmon metapopulation robustness}
 \author{
@@
 
-\section*{I. Gaussian approximation for the mixed trait distribution}
+We aim to understand how straying between populations may affect phenotypic evolution, and how in turn this evolutionary change may lead to changes in ecological dynamics as they unfold. 
+To do so, we build upon ideas from classic papers (Lande 1976, most notably). 
+In what follows, we show 
+1) how the application of Lande's equation for the case of two populations linked by dispersal leads to complex equations of trait evolution that are not easily modeled,
+2) that this complex trait distribution can be approximated as a single Gaussian distribution, and
+3) under what conditions this approximation is appropriate.
+
+\section*{I. Phenotypic evolution from one to two sites}
+
+\textbf{Trait evolution in one site} To keep track of the evolution of a phenotypic trait $x$ with probability density function $p$ and mean $\overline{x}$ and variance $\sigma^{2}$, we need a few basic ingredients.
+First, we define the mean fitness,$\overline{W}$ of the population of individuals bearing the trait $x$ as:
+
+\begin{equation}\label{1}
+\overline{W}=\int_{} p(x)W(x)\,dx.
+\end{equation}
+We also define the mean phenotype in the population as
+
+\begin{equation}\label{2}
+\overline{x}=\int_{} xp(x)\,dx,
+\end{equation}
+and the mean phenotype after selection (but before reproduction), $\overline{x}_{w}$ \cite{Falconer:1967tg}, as:
+
+\begin{equation}\label{3}
+\overline{x}_{w}=\frac{1}{\overline{W}}\int_{} xp(x)W(x)\,dx.
+\end{equation}
+Then, we can assess how the change in the phenotype $x$ occurs over time following
+\begin{equation}\label{4}
+\Delta x=x(t+1)-x(t),
+\end{equation}
+which can be rewritten as:
+\begin{equation}\label{5}
+\Delta x=h^{2}(\overline{x}_{w}-x(t)),
+\end{equation}
+where $h^{2}$ is the heritability of the trait, and the difference $\overline{x}_{w}-x(t)$ is the classical "selection differential"  \cite{Falconer:1967tg,Lande:1976ga}.
+In other words, \ref{5} is none other that the Breeder's equation.
+
+To understand how $x$ changes over time, we first need to understand how the mean fitness of the population changes with a change in the mean trait value or $\frac{\partial \overline{W}}{\partial \overline{x}}$ (i.e.
+the shape of the "adaptive landscape").
+Using \ref{1} we have:
+
+\begin{equation}\label{6}
+\frac{\partial \overline{W}}{\partial \overline{x}}=\frac{\partial}{\partial \overline{x}}\int_{} p(x)W(x)\,dx.
+\end{equation}
+Using Leibniz rule we can pass the derivative under the integral sign, 
+\begin{equation}\label{7}
+\frac{\partial \overline{W}}{\partial \overline{x}}=\int_{} \frac{\partial p(x)}{\partial \overline{x}}W(x)\,dx,
+\end{equation}
+and assuming $p$ is Gaussian (i.e.
+of the form $\frac{1}{\sqrt{2 \pi \sigma^{2}}} e^{-\frac{1}{2}\frac{(x-\overline{x})^{2}}{\sigma^{2}}} $), we calculate the derivative of $p$ with respect to $\overline{x}$, which leads to:
+
+\begin{equation}\label{8}
+\frac{\partial \overline{W}}{\partial \overline{x}}=\int_{} \Big[\frac{x-\overline{x}}{\sigma^{2}}\Big] p(x)W(x)\,dx.
+\end{equation}
+We can now expand in \ref{8} to get,
+\begin{equation}\label{9}
+\frac{\partial \overline{W}}{\partial \overline{x}}=
+\frac{1}{\sigma^{2}}\int_{}xp(x)W(x)\,dx-\frac{\overline{x}}{\sigma^{2}}\int_{}p(x)W(x)\,dx,
+\end{equation}
+which making use of equations \ref{1} and \ref{3}, becomes $\frac{\partial \overline{W}}{\partial \overline{x}}=
+\frac{\overline{W}}{\sigma^{2}}\overline{x}_{w}-\frac{\overline{x}}{\sigma^{2}}\overline{W}$, which can be factored as $\frac{\partial \overline{W}}{\partial \overline{x}}=
+\frac{\overline{W}}{\sigma^{2}}(\overline{x}_{w}-\overline{x})$.
+Using \ref{5}, we can rearrange this expression to obtain an equation that related the change in the average phenotype from one time step to the next to the amount of heritable variation, $\sigma^{2}h^{2}$, and the adaptive landscape $\frac{\partial \overline{W}}{\partial \overline{x}}$:
+\begin{equation}\label{10}
+\Delta \overline{x}=\frac{\sigma^{2}h^{2}}{\overline{W}}\frac{\partial \overline{W}}{\partial \overline{x}}=\sigma^{2}h^{2}\frac{\partial \ln\overline{W}}{\partial \overline{x}}.
+\end{equation}
+Equation \ref{10} (Lande 1976's equation 7), is a staple of evolutionary biology.
+It is also possible to use \ref{10} and \ref{4} to write a recurrence relationship that allows to simulating evolutionary change over time in a trait $x$ as:
+\begin{equation}\label{11}
+\overline{x}(t+1)=\overline{x}(t)+\sigma^{2}h^{2}\frac{\partial \ln\overline{W}}{\partial \overline{x}}.
+\end{equation}
+
+\textbf{Expanding the single site approach to two sites linked by dispersal} In what follows we show how Lande's approach can be used to calculate the magnitude of the phenotypic change between two populations with one trait with two different means in each site.
+We will then 1) show why it is not necessarily useful to use the obtained expression in a simulation context, and 2) introduce an approximation that retains the features important for exploring how trait and population dynamics impact each other in the case of dispersing populations.
+In Section II, we will explore how well the approximation reproduces the underlying trait dynamics.
+Specifically, we will discuss in what cases it is exact and for which cases the approximation becomes problematic, such that it should be avoided.
+
+We explore the case of two populations that exchange migrants, with a normally distributed trait $x$ that controls recruitment (as explained in the main text), with each population having a different mean, $\mu_{i}$ and $\mu_{j}$, and same variance $\sigma^{2}$.
+For simplicity, we will focus on a local population $i$ relative to the connected population $j$.
+
+When individuals from population $j$ stray into population $i$, and individuals of population $i$ stray into population $j$, the phenotypic distribution of trait $x$ in population $i$ is a mixture distribution, $p(x)$, of the form:
+
+\begin{equation}\label{12}
+p(x)=\omega_i g(x,\mu_{i})+(1-\omega_i)g(x,\mu_{j}).
+\end{equation}
+where $g(\cdot)$ is the Gaussian probability density function, and $\omega_i$ is the proportion of the mixed population that is composed of individuals local to site $i$ relative to those that have dispersed into site $i$ from site $j$.
+This distribution has a mean $\overline{x}=\omega_i \mu_{i}+(1-\omega_i)\mu_{j}$.
+As before, we define the mean fitness as in equation \ref{1}, only that now $p$ is a mixture distribution.
+Let us use Lande's approach to obtain an expression for the change in the mean phenotype of the mixture distribution from one time step to the next.
+We thus ask, what is the change in $\overline{W}$ with a change in $\overline{x}$?
+We can use Lande's approach \cite{Lande:1976ga} to derive equation \ref{7}, replacing $p(x)$ by the mixture (equation \ref{12}) to obtain
+
+\begin{equation}\label{13}
+\frac{\partial \overline{W}}{\partial \overline{x}}=\int_{} \frac{\partial (\omega_i g(x,\mu_{i})+(1-\omega_i)g(x,\mu_{j}))}{\partial \overline{x}}W(x)\,dx.
+\end{equation}
+To obtain the derivative of equation \ref{13}, we introduce a change in variables, which is provided by the fact that $\overline{x}=\omega_i \mu_{i}+(1-\omega_i)\mu_{j}$:
+\begin{equation} \label{14}
+	\begin{cases} 
+		\mu_{i}=\frac{\overline{x}-(1-\omega_i)\mu_{j}}{\omega_i}\\
+		\mu_{j}=\frac{\overline{x}-\omega_i\mu_{i}}{1-\omega_i}\\
+	\end{cases}.
+\end{equation}
+We can thus replace equation \ref{14} into \ref{13},
+\begin{equation}\label{15}
+\frac{\partial \overline{W}}{\partial \overline{x}}=\int_{} \frac{\partial (\omega_i g(x,\frac{\overline{x}-(1-\omega_i)\mu_{j}}{\omega_i})+(1-\omega_i)g(x,\frac{\overline{x}-\omega_i\mu_{i}}{1-\omega_i}))}{\partial \overline{x}}W(x)\,dx,
+\end{equation} 
+which, after expanding and taking the derivatives with respect to $\overline{x}$ becomes:
+
+\begin{align}\label{16}
+\frac{\partial \overline{W}}{\partial \overline{x}}=&\int_{} \omega_i\Bigg[\frac{x-\frac{\overline{x}-(1-\omega_i)\mu_{j}}{\omega_i}}{\omega_i \sigma^{2}}\Bigg] g(x,\frac{\overline{x}-(1-\omega_i)\mu_{j}}{\omega_i})W(x)\,dx \\ \nonumber
+&+ \int_{} (1-\omega_i)\Bigg[\frac{x-\frac{\overline{x}-\omega_i\mu_{i}}{1-\omega_i}}{(1-\omega_i) \sigma^{2}}\Bigg] g(x,\frac{\overline{x}-\omega_i\mu_{i}}{1-\omega_i})W(x)\,dx.
+\end{align}
+To simplify notation, at this point we can come back to the original variables $\mu_{i}$ and $\mu_{j}$. We then expand and collect terms to get:
+\begin{align}\label{17}
+\frac{\partial \overline{W}}{\partial \overline{x}}&=\frac{1}{\sigma^{2}}\Bigg[\int_{}xg(x,\mu_{i})W(x)\,dx - \mu_{i}\int_{}g(x,\mu_{i})W(x)\,dx \\ \nonumber 
+&+ \int_{}xg(x,\mu_{j})W(x)\,dx - \mu_{j}\int_{}g(x,\mu_{j})W(x)\,dx\Bigg].
+\end{align}
+Using equations \ref{1} and \ref{3}, we can define the following quantities,
+
+\begin{equation}\label{18}
+\overline{W}_{i}=\int_{} g(x,\mu_{i})W(x)\,dx 
+\end{equation}
+\begin{equation}\label{19}
+\overline{W}_{j}=\int_{} g(x,\mu_{j})W(x)\,dx  
+\end{equation}
+\begin{equation}\label{20}
+\overline{x}_{w,i}=\frac{1}{\overline{W}}\int_{}xg(x,\mu_{i})W(x)\,dx
+\end{equation}
+\begin{equation}\label{21}
+\overline{x}_{w,j}=\frac{1}{\overline{W}}\int_{}xg(x,\mu_{j})W(x)\,dx, 
+\end{equation}
+then replace them into equation \ref{17}:
+\begin{equation}\label{22}
+\begin{aligned}
+\frac{\partial \overline{W}}{\partial \overline{x}} &=\frac{1}{\sigma^{2}}\Big[\overline{x}_{w,i} \overline{W}_{i} - \mu_{i}\overline{W}_{i} + \overline{x}_{w,j} \overline{W}_{j} - \mu_{j}\overline{W}_{j}\Big]  \\
+&= \frac{1}{\sigma^{2}}\Big[ \overline{W}_{i}(\overline{x}_{w,i}-\mu_{i}) + \overline{W}_{j}(\overline{x}_{w,j}-\mu_{j})\Big].
+\end{aligned}
+\end{equation}
+Using equation \ref{5} and rearranging terms we get:
+\begin{equation}\label{23}
+\overline{W}_{i}\Delta \mu_{i} + \overline{W}_{j}\Delta\mu_{j} = \sigma^{2}h^{2}\frac{\partial \overline{W}}{\partial \overline{x}} ,
+\end{equation}
+with $\overline{W}=\omega_i\overline{W}_{i}+(1-\omega_i)\overline{W}_{j}$.
+Equation \ref{23} relates how changes in either $\mu_{i}$ or $\mu_{j}$ lead to changes in fitness, and vice-versa.
+However, it doesn't tell us anything as to how straying may lead to evolutionary change, or how that may feedback onto ecological dynamics.
+Because of this, we will once again use the change of variables in \ref{14}, to rewrite \ref{23} as:
+\begin{equation}\label{24}
+\overline{W}_{i}\Big(\frac{\Delta\overline{x}-(1-\omega_i)\Delta\mu_{j}}{\omega_i} \Big)+\overline{W}_{j}\Big(\frac{\Delta\overline{x}-\omega_i\Delta\mu_{i}}{1-\omega_i} \Big)= \sigma^{2}h^{2}\frac{\partial \overline{W}}{\partial \overline{x}},
+\end{equation}
+which can be simplified and rearranged to obtain an expression for the change in the mean phenotype of the mixture distribution,
+\begin{equation}\label{25}
+\Delta\overline{x}=\Big[ \sigma^{2}h^{2}\frac{\partial \overline{W}}{\partial \overline{x}} + \Big(\frac{(1-\omega_i)}{\omega_i}\overline{W}_{i}\Delta\mu_{j}+\frac{\omega_i}{1-\omega_i}\overline{W}_{j}\Delta\mu_{i} \Big) \Big] \frac{\omega_i(1-\omega_i)}{(1-\omega_i)\overline{W}_{i}+\omega_i\overline{W}_{j}}.
+\end{equation}
+By multiplying \ref{25} by $\frac{\overline{W}}{\overline{W}}$, we can further rewrite the equation as:
+\begin{equation}\label{26}
+\Delta\overline{x}=\Big[ \sigma^{2}h^{2}\frac{\partial \ln\overline{W}}{\partial \overline{x}} + \frac{1}{\overline{W}}\Big(\frac{(1-\omega_i)}{\omega_i}\overline{W}_{i}\Delta\mu_{j}+\frac{\omega_i}{1-\omega_i}\overline{W}_{j}\Delta\mu_{i} \Big) \Big] \frac{\omega_i(1-\omega_i)\overline{W}}{(1-\omega_i)\overline{W}_{i}+\omega_i\overline{W}_{j}},
+\end{equation}
+which has a similar form to Lande's equation (\ref{10}), but shows the explicit dependence of the change in the mean of the mix on the proportion of straying individuals.
+A symmetrical expression can be derived for the other site.
+
+While some understanding can be gained using equation \ref{26}, the expression does not easily allow us to simulate the change in $\overline{x}$ over time because the change in $\mu_{i}$ and $\mu_{j}$ is difficult to track in closed form.
+Moreover, this equation only holds for the first time step, and as new dispersing individuals arrive and the trait distribution is again updated by the influx of individuals from the other site, the mix becomes a mix of mixes, such that the analytical expressions presented above no longer hold.
+
+This presents a tradeoff: we can either track changes in phenotype exactly for a single generation, or we can track changes in phenotype across multiple generations using an approximation.
+Given that our central goal is to gain an understanding of the feedback between evolutionary and ecological processes in salmon populations with dispersal in explicit space, we simplify the complexities created by phenotypic evolution by assuming that, as new migrants arrive, the mixture distribution can be locally approximated by a Gaussian distribution with a mean equal to that of the mix.
+Below, we show why this approximation is appropriate for describing the phenotypic distributions under most conditions relevant to our questions.
+Moreover, based on this assumption, we are able to use Lande's equation (here equation \ref{10}), without violating the original assumptions that led to its current form.
+
+\section*{II. Gaussian approximation for the mixed trait distribution}
 We assume a Gaussian distribution to approximate the more complex trait distribution that results when two populations with two distinct trait distributions are linked by gene flow and reproduction.
-Here and throughout, we refer to the resident population as $i$ and the straying population as $j$, though both populations stray simultaneously.
-The Gaussian approximation is assumed to have a mean that is equal to the mean of the mixed trait distribution $(1-m)\mu_i + m\mu_j$ with a constant variance $\sigma^2$, equal to the original trait variance prior to mixing.
-We employ this approximation because a mixed trait distribution that remixes and recombines (through reproduction) at each time-step becomes complicated very quickly, as we shall demonstrate.
+Here and throughout, we refer to the resident population as $i$ and the connected population as $j$, though both populations stray simultaneously.
+The Gaussian approximation is assumed to have a mean that is equal to the mean of the mixed trait distribution $\omega_i\mu_i + (1-\omega_i)\mu_j$ with a constant variance $\sigma^2$, equal to the original trait variance prior to mixing.
+We employ this approximation because a mixed trait distribution that remixes and recombines (through reproduction) at each time-step becomes complicated very quickly, as we shall demonstrate from a mechanistic perspective.
 \emph{Importantly, because the evolutionary framework tracks only the mean and not the variability through time, this approximation is exact with respect to the first moment.}
@@
 \begin{equation}
-  p(x_{\rm disperse}) = w_i g(x_i,\mu_i,\sigma) + (1-w_i)g(x_j,\mu_j,\sigma),
+  p(x_{\rm disperse}) = \omega_i g(x_i,\mu_i,\sigma) + (1-\omega_i)g(x_j,\mu_j,\sigma),
 \end{equation}
@@
 \begin{equation}
-  w_i = \frac{(1-m)N_i}{(1-m)N_i + mN_j},
+  \omega_i = \frac{(1-m)N_i}{(1-m)N_i + mN_j},
 \end{equation}
-and the proportion of population $i$ composed of strays is $(1-w_i)$.
+and the proportion of population $i$ composed of strays is $(1-\omega_i)$.
 Once individuals in both habitats mix, they mate without preference.
@@
 \begin{equation}
-  q(x_{\rm r}) = \int_{-\infty}^\infty 2p(x_1)p(2x_{\rm r}-x_1){\rm d}x_1,
+  q(x_{\rm r}) = \int_{-\infty}^\infty 2p(x_1)p(2x_{\rm r}-x_1){\rm d}x_1.
 \end{equation}
-where the factor of 2 comes from...
 The mean of the recruit trait distribution is calculated to be
 \begin{equation}
-{\rm E}\{q(x_{\rm r})\}=(1-m)\mu_i + m\mu_j,
+{\rm E}\{q(x_{\rm r})\}=\omega_i\mu_i + (1-\omega_i)\mu_j,
 \end{equation} 
@@
 The relative contribution of trait values from resident parents, stray parents, and recruits depends on the numbers of each within the mixed population.
-The number if residents is $(1-m)N_i$, and the number of strays is $mN_j$.
+The number of residents is $(1-m)N_i$, and the number of strays is $mN_j$.
 However, because adult mortality occurs after reproduction, we must adjust these numbers to account for only those who survive, which is given by ${\rm e}^{-Z}$.
@@
 Accordingly, the number of adult residents and strays is given by $(1-m)N_i{\rm e}^{-Z}$ and  $mN_i{\rm e}^{-Z}$, respectively.
-In contrast, the number of recruits is given by the combined contribution of resident and straying individuals to recruit production as $N_r = ((1-m)N_1 R[\mu_i] + mN_2 R[\mu_j]){\rm e}^{(-\beta((1-m)N_i + mN_j))}$ where $R$ is the recruitment rate, which differs for resident vs.
-straying individuals (see manuscript, Eq 2).
+In contrast, the number of recruits is given by the combined contribution of resident and straying individuals to recruit production as 
+\begin{equation}
+N_r = \int_{-\infty}^{\infty}r_{\rm max}{\rm e}^{\frac{(x_{\rm r}-\theta_i)^2}{2\tau^2}}q(x_{\rm r})dx_{\rm r}((1-m)N_i + mN_j){\rm e}^{(-\beta((1-m)N_i + mN_j))},
+\end{equation}
+where $\theta_i$ is the optimal trait value in site $i$, $r_{\rm max}$ is the maximum recruitment rate, and $\tau$ is the strength of selection.
 
@@
 a) low stray ratio, large straying population, overlapping generations: $m=0.1$, $Z=0.5$;
-b) high stray ratio, overlapping generations: $m=0.5$, $Z=0.5$;
-c) high stray ratio, non-overlapping generations: $m=0.5$, $Z=1000$;
+b) maximal stray ratio, overlapping generations: $m=0.5$, $Z=0.5$;
+c) maximal stray ratio, non-overlapping generations: $m=0.5$, $Z=1000$;
 d) density-dependent stray ratio, non-overlapping generations: $m=m_0(1-N_i/(C+N_i))$, $Z=1000$, where $m_0=0.2$ and $C=1000$.
@@
 \begin{equation}
-{\rm E}\{g_{\rm next}(x_i)\}=(1-m)\mu_i + m\mu_j,
-\end{equation} and the variance is
+{\rm E}\{g_{\rm next}(x_i)\}=\omega_i\mu_i + (1-\omega_i)\mu_j.
+\end{equation}
+
 % variance a that has a closed form solution, but that is not easily readable.
-\begin{align}
-  {\rm Var}\{g_{{\rm next}}(x_i)\}=&\frac{(\mu_i-\mu_j)^2 (2 (m-1)^2 N_i^2 c_j^2-(m-1) N_i c_j (2 (m-1) N_i+m N_j (1-4 c_i))}{2 (-m N_i+m N_j+N_i)^2} \\ \nonumber
-  &+ \frac{m N_j (c_i-1) (-m N_i+2 m N_j c_i+N_i))-\sigma ^2 (c_i+c_j+1) (-m N_i+m N_j+N_i)^2}{2 (-m N_i+m N_j+N_i)^2}.
-\end{align}
+% \begin{align}
+%   {\rm Var}\{g_{{\rm next}}(x_i)\}=&\frac{(\mu_i-\mu_j)^2 (2 (m-1)^2 N_i^2 c_j^2-(m-1) N_i c_j (2 (m-1) N_i+m N_j (1-4 c_i))}{2 (-m N_i+m N_j+N_i)^2} \\ \nonumber
+%   &+ \frac{m N_j (c_i-1) (-m N_i+2 m N_j c_i+N_i))-\sigma ^2 (c_i+c_j+1) (-m N_i+m N_j+N_i)^2}{2 (-m N_i+m N_j+N_i)^2}.
+% \end{align}
 
 Importantly, ${\rm E}\{g_{{\rm next}}(x_i)\}$ is exactly equal to the trait mean for the original mix, as well as the trait mean for recruits.
-Thus, a Gaussian approximation distributed as $\mathcal{N}(w_i\mu_i + (1-w_i)\mu_j,\sigma)$ will not alter the dynamics under the assumption of constant variance in Eq 4.
+Thus, a Gaussian approximation distributed as $\mathcal{N}(\omega_i\mu_i + (1-\omega_i)\mu_j,\sigma)$ will not alter the dynamics under the assumption of constant variance in Eq 4.
 Moreover, the mixed distribution described above becomes more complex at the next timestep, as two multi-modal trait distributions then mix and recombine through recruitment.
-Although the complexity of the true distribution is increased and becomes quickly intractable, one expects the actual form to eventually approximate a Gaussian based on the central limit theorem.
 One could make the argument that a Gaussian approximation does not represent the multi-modality in trait values inherent to a mixed distribution, and thus underestimates variance.
@@
 %Discussion of factors that increase or decrease accuracy of the approximation
-The Gaussian approximation $\mathcal{N}(w_i\mu_i + (1-w_i)\mu_j,\sigma)$ will generally underestimate the true variance of the mix, but not by much.
-If the straying ratio is low, (ca. 0.0 to 0.2), the Gaussian approximation tends to be very similar to the actual distribution (Fig. \ref{fig:PDF}a).
+The Gaussian approximation $\mathcal{N}(\omega_i\mu_i + (1-\omega_i)\mu_j,\sigma)$ will generally underestimate the true variance of the mix, but not by much.
+If the straying ratio is low, (ca. 0.0 to 0.2), the Gaussian approximation is very similar to the actual distribution (Fig. \ref{fig:PDF}a).
 As the straying ratio increases, so does the bimodality of the actual distribution, and this magnifies our underestimate of variance (Fig. \ref{fig:PDF}b).
-The interesting dynamics observed in our model generally occur between $m=0.05$ and $m=0.2$, where estimates of variance are similar.
+The relevant dynamics observed in our model generally occur between $m=0.05$ and $m=0.2$, where estimates of variance are similar.
 Decreasing the overlap between generations also leads to greater similarity in the actual distribution and the Gaussian approximation, and the potential bias introduced by bimodality will be less for cases of non-overlapping generations, or semelparity (Fig.
 \ref{fig:PDF}c).
-Most importantly, density-dependent straying increases the accuracy of our Gaussian approximation by suppressing the straying ratio (Fig. \ref{fig:PDF}d).
+Most importantly, density-dependent straying increases the accuracy of our Gaussian approximation by actively suppressing the straying ratio (Fig. \ref{fig:PDF}d).
 
-Because the Gaussian approximation is exact to first order, and generally provides a good approximation of realized variance, we suggest that the use of Lande's formulation for the change in mean trait values through time under the assumption of constant variance is appropriate (see Eq.
-4 in main text).
-Because our estimate of variance is generally an underestimate, the rate of recruitment will be a slight overestimate (Eq.
-2, main text), and the timescale of trait evolution will generally be slower (Eq.
-4, main text).
+Because the Gaussian approximation is exact to first order, and generally provides a good approximation of realized variance, we suggest that the use of Lande's formulation for the change in mean trait values through time under the assumption of constant variance is appropriate.
+Because our estimate of variance is generally an underestimate, the rate of recruitment will be a slight overestimate, and the timescale of trait evolution will generally be slower.
 Changes in the rate of recruitment (Fig.
 \ref{fig:symmetry}) and heritability do not have an impact on our qualitative findings, and this suggests any bias that our approximation introduces by underestimating true variance is minimal.
+
+Under these assumptions, the governing equations for the dynamic system are then written as
+
+\begin{align}
+  &N_i(t+1) = \\ \nonumber
+  &\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z} \\ \nonumber
+  &+ R_i[\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\theta_i]\left((1-m)N_i + mN_j\right) \\ \nonumber
+  &\times {\rm e}^{-\beta ((1-m)N_i(t) + m N_j(t))},
+  \label{eq:N}
+\end{align}
+where the trait distribution for recruits is also approximated by the Gaussian 
+${\rm pr}(x_i) \approx \mathcal{N}(\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\sigma^2)$, where the mean is the analytical expectation.
+This gives the rate of recruitment as 
+\begin{align}
+  &R_i[\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\theta_i] = \\ \nonumber
+  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i,\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\sigma^2) {\rm d}x_i +\tilde{P}_i\\ \nonumber
+  &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-(\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t)))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\} +\tilde{P}_i.
+\end{align}
+The mean of the trait distribution then changes as
+\begin{align}
+  \label{eq:mu}
+  \mu_i(t+1) &= \omega_i\mu_i(t) + (1-\omega_i)\mu_j(t) + h^2\sigma^2\frac{\partial}{\partial \mu^\prime}\ln\left(R_i[\mu^\prime,\theta_i] \right), \\ \nonumber
+  &= \omega_i\mu_i(t) + (1-\omega_i)\mu_j(t) + h^2\sigma^2\left(\frac{\theta_i - \omega_i\mu_i - (1-\omega_i)\mu_j}{\sigma^2+\tau^2} \right),
+\end{align}
+given $\mu^\prime = \omega_i \mu_i(t)+ (1-\omega_i)\mu_j(t)$.
+Under the condition that $\sigma=1$ and $\tau = 1$ as is set in the manuscript, this simplifies to
+\begin{equation}
+  \mu_i(t+1) = \omega_i\mu_i(t) + (1-\omega_i)\mu_j(t) + h^2\left(\theta_i - \omega_i\mu_i - (1-\omega_i)\mu_j\right).
+\end{equation}
+
+\bibliography{aa_kevin}
+
+\clearpage
 
@@
 
-\begin{figure}
+\begin{figure}[h]
   \captionsetup{justification=raggedright,

### 2017-11-17 model/model2/Figure1.jl

@@ -1,8 +1,13 @@
 #Figure 1
+using Distributions
+using RCall
+
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve.jl")
+include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinJacobian.jl")
 
 #Analysis over m
 tmax=10000;
-mvec = collect(0.0:0.001:0.5);
+mvec = collect(0.0:0.001:0.3);
 n1ts = zeros(Float64,length(mvec),tmax);
 n2ts = zeros(Float64,length(mvec),tmax);
 n1mean=zeros(Float64,length(mvec));
@@ -22,7 +27,7 @@ z=0.5;
 rmax=2.0;
 beta=0.001;
 theta1=5.0;
-thetadiff=3;
+thetadiff=3.3;
 tau=1.0;
 h=0.2;
 sigmaE=0;
@@ -86,22 +91,22 @@ burnin=0.80
   
 end
 #Steady state plot
-namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/FinalDraft_rev/fig_traj.pdf");
+namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/FinalDraft3/fig_traj.pdf");
 R"""
 library(RColorBrewer)
 cols = brewer.pal(4,'Set1')
-#pdf($namespace,height=8,width=5)
+pdf($namespace,height=8,width=5)
 par(mfrow=c(2,1),mai = c(0.8, 0.8, 0.1, 0.1))
-plot($mvec,$n1mean,pch='.',col=cols[4],xlab="Straying rate m",ylab="Steady state biomass",cex=0.5,ylim=c(0,max($n1mean)),las=1)
+plot($mvec,$n1mean,pch='.',col=cols[4],xlab="Straying rate m",ylab="Steady state biomass",cex=0.5,ylim=c(0,max($n1mean)+200),las=1)
 points($mvec,$n2mean,pch='.',col=cols[4],cex=0.5)
-arrows($(mvec[indmax(pe)]),1200,$(mvec[indmax(pe)]),1100,length=0.05,angle=40,lwd=3)
-text($(mvec[indmax(pe)]),1275,'DCB')
[…]

### 2017-11-17 model/model2/Figure3_S8.jl

@@ -1,6 +1,6 @@
 #Figure 3
 #Figure S8
-using Distributions, RCall, JLD, HDF5
+@everywhere using Distributions, RCall, JLD, HDF5
 
 @everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve.jl")
 @everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve_ddm.jl")
@@ -15,7 +15,7 @@ using Distributions, RCall, JLD, HDF5
 #Analysis over m
 tmax=10000;
 
-mvec = collect(0.0001:0.001:0.5);
+mvec = collect(0.0001:0.001:0.3);
 pvec = ["small","large","both"];
 hvec = [0.2, 0.8]
 reps = 100;
@@ -24,7 +24,7 @@ z=0.5;
 rmax=2.0;
 beta=0.001;
 theta1=5.0;
-thetadiff=5;
+thetadiff=4;
 tau=1.0;
 C=1000;
 sigmaE=0;
@@ -125,9 +125,9 @@ m2mean = SharedArray{Float64}(reps,length(mvec),length(pvec),length(hvec));
   end
 end
 
-save(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data2/data_relax.jld"),"rt",rt,"rt_ddm",rt_ddm,"m1mean",m1mean,"m2mean",m2mean);
+save(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data3/data_relax.jld"),"rt",rt,"rt_ddm",rt_ddm,"m1mean",m1mean,"m2mean",m2mean);
 
-d = load(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data2/data_relax.jld"));
+d = load(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data3/data_relax.jld"));
 #This loads the dictionary
 rt = d["rt"];
 rt_ddm = d["rt_ddm"];
@@ -137,9 +137,6 @@ m2mean = d["m2mean"];
 
 ma_rth2 = mean([rt[i,:,:,1] for i=1:reps]);
 ma_rth8 = mean([rt[i,:,:,2] for i=1:reps]);
-
-
-
 
[…]

### 2017-11-17 model/model2/Figure5.jl

@@ -1,12 +1,9 @@
 #Figure 5
 
 
-using Distributions
-using RCall
-using HDF5
-using JLD
+@everywhere using Distributions, RCall, JLD, HDF5
 
-include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/movingaverage.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/movingaverage.jl")
 @everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve.jl")
 @everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve_ddm.jl")
 @everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/qualsfunc.jl")
@@ -16,7 +13,7 @@ include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/movingaverage.
 # Figure 5
 
 #Analysis over m & theta divergence
-mvec=collect(0.0:0.001:0.45);
+mvec=collect(0.0:0.001:0.3);
 hvec = collect(0.0:0.01:1.0);
 
 
@@ -75,11 +72,11 @@ perror=0.01;
       (1/(std(n1trim+n2trim)/mean(n1trim+n2trim)))
     end
 end
-save(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data2/data_sig_h_m.jld"),"n1mean",n1mean,"n2mean",n2mean,"x1mean",x1mean,"x2mean",x2mean,"pe",pe);
+save(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data3/data_sig_h_m.jld"),"n1mean",n1mean,"n2mean",n2mean,"x1mean",x1mean,"x2mean",x2mean,"pe",pe);
 
 
 #Analysis over m & theta divergence
-mvec=collect(0.0:0.001:0.45);
+mvec=collect(0.0:0.001:0.3);
 hvec = collect(0.0:0.01:1.0);
 
 
@@ -138,12 +135,12 @@ perror=0.01;
     (1/(std(n1trim+n2trim)/mean(n1trim+n2trim)))
[…]

### 2017-11-17 model/model2/Figure6_S11_S13.jl

@@ -3,7 +3,7 @@
 #Figure S13
 
 
-using Distributions, RCall, HDF5, JLD
+@everywhere using Distributions, RCall, HDF5, JLD
 
 @everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve_mtheta.jl")
 
@@ -30,6 +30,7 @@ rt = SharedArray{Float64}(reps,length(mvec2),length(pvec));
 rt_ddm = SharedArray{Float64}(reps,length(mvec2),length(pvec));
 m1mean = SharedArray{Float64}(reps,length(mvec2),length(pvec));
 m2mean = SharedArray{Float64}(reps,length(mvec2),length(pvec));
+pe = SharedArray{Float64}(reps,length(mvec2),length(pvec));
 
 
 z=0.5;
@@ -50,7 +51,7 @@ t_ext = Int64(round(tmax/2));
 @sync @parallel for r=1:reps
   for j=1:length(pvec)
     extpop = pvec[j]
-    for i=1:length(mvec)
+    for i=1:length(mvec2)
       
       m = mvec2[i];
       a0 = 1-m;
@@ -79,6 +80,14 @@ t_ext = Int64(round(tmax/2));
       
       rt[r,i,j] = relaxtime;
       
+      burnin=0.80
+      n1trim = n1[Int64(floor(tmax*burnin)):tmax];
+      n2trim = n2[Int64(floor(tmax*burnin)):tmax];
+      x1trim = x1[Int64(floor(tmax*burnin)):tmax];
+      x2trim = x2[Int64(floor(tmax*burnin)):tmax]
+      
+      pe[r,i,j] = mean([(std(n1trim)/mean(n1trim)),(std(n2trim)/mean(n2trim))])*
+      (1/(std(n1trim+n2trim)/mean(n1trim+n2trim)))
       
       n1_ddm, n2_ddm, x1_ddm, x2_ddm, w1_ddm, w2_ddm, m1_ddm, m2_ddm = 
       KevinEvolveExtinct_ddm(
@@ -115,11 +124,11 @@ t_ext = Int64(round(tmax/2));
   end
 end
 
[…]

### 2017-11-17 model/model2/FigureS2_S4_S5.jl

@@ -5,16 +5,16 @@ using RCall
 
 
 thetascale = 2;
-m=collect(0:0.01:0.5);
+m=collect(0:0.01:0.3);
 thetadiff=zeros(length(m));
 for i = 1:length(m)
     thetadiff[i] = (1-2*m[i])/(thetascale*m[i])
 end
 
-namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/FinalDraft_rev/fig_mthetarelation.pdf");
+namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/FinalDraft3/fig_mthetarelation.pdf");
 R"""
 pdf($namespace,height=5,width=6)
-plot($thetadiff,$m,type='l',ylab='Straying rate m',xlab=expression(paste('Habitat heterogeneity ',Delta,theta)),ylim=c(0,0.5))
+plot($thetadiff,$m,type='l',ylab='Straying rate m',xlab=expression(paste('Habitat heterogeneity ',Delta,theta)),ylim=c(0,0.3),xlim=c(0,10))
 dev.off()
 """
 
@@ -149,15 +149,14 @@ dev.off()
 """
 
 
-using RCall
-using Distributions
-include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolveSS.jl")
-include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinJacobian.jl")
+@everywhere using RCall, Distributions
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolveSS.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinJacobian.jl")
 
 
 #Analysis over m
 tmax=10000;
-mvec1 = collect(0.0:0.0001:0.15);
+mvec1 = collect(0.0:0.0001:0.2);
 mvec = [mvec1 ; reverse(mvec1)];
 n1ts = zeros(Float64,length(mvec),tmax);
 n2ts = zeros(Float64,length(mvec),tmax);
@@ -170,19 +169,19 @@ aggsd=zeros(Float64,length(mvec));
 x1mean=zeros(Float64,length(mvec));
 x2mean=zeros(Float64,length(mvec));
[…]

### 2017-11-17 model/model2/FigureS5_S9_S10_S14.jl

@@ -142,9 +142,9 @@ dev.off()
 
 
 #Analysis over m & theta divergence
-mvec = collect(0.0001:0.001:0.5);
-thetadiffvec = collect(0.0:0.1:10.0);
-hvec = collect(0.1:0.1:0.9);
+mvec = collect(0.0001:0.001:0.3);
+thetadiffvec = collect(0.0:0.1:6.0);
+hvec = collect(0.1:0.1:0.8);
 reps = 50;
 
 thetadiffarray=SharedArray(Float64,reps,length(thetadiffvec),length(hvec),length(mvec));
@@ -284,9 +284,9 @@ t_ext = Int64(round(tmax/2));
   end
 end
 
-save(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data2/data_habitathetero_ext.jld"),"n1mean",n1mean,"n2mean",n2mean,"x1mean",x1mean,"x2mean",x2mean,"pe",pe,"rt",rt,"n1mean_ddm",n1mean_ddm,"n2mean_ddm",n2mean_ddm,"x1mean_ddm",x1mean_ddm,"x2mean_ddm",x2mean_ddm,"rt_ddm",rt_ddm,"m1mean",m1mean,"m2mean",m2mean,"pe_ddm",pe_ddm,"thetadiffarray",thetadiffarray);
+save(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data3/data_habitathetero_ext.jld"),"n1mean",n1mean,"n2mean",n2mean,"x1mean",x1mean,"x2mean",x2mean,"pe",pe,"rt",rt,"n1mean_ddm",n1mean_ddm,"n2mean_ddm",n2mean_ddm,"x1mean_ddm",x1mean_ddm,"x2mean_ddm",x2mean_ddm,"rt_ddm",rt_ddm,"m1mean",m1mean,"m2mean",m2mean,"pe_ddm",pe_ddm,"thetadiffarray",thetadiffarray);
 
-d = load(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data2/data_habitathetero_ext.jld"));
+d = load(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data3/data_habitathetero_ext.jld"));
 #This loads the dictionary
 n1mean = d["n1mean"];
 n2mean = d["n2mean"];
@@ -322,7 +322,7 @@ for i=1:length(hvec)
   end
 end
 
[…]

### 2017-11-17 model/src/KevinEvolve.jl

@@ -70,9 +70,9 @@ function KevinEvolve(
     # ((((sigmaE^2 + sigmaG^2) + tau^2)^(3/2))*(((exp(-(((theta1 + thetadiff)-x1[t])^2)/(2*((sigmaE^2 + sigmaG^2)+tau^2)))*rmax*(1-w2[t])*tau)/
     # (sqrt((sigmaE^2 + sigmaG^2)+tau^2))) +  ((exp(-(((theta1 + thetadiff)-x2[t])^2)/(2*((sigmaE^2 + sigmaG^2)+tau^2)))*rmax*w2[t]*tau)/(sqrt((sigmaE^2 + sigmaG^2)+tau^2)))));
     
-    meanfit1[t] = (w1[t]*(theta1 - w1[t]*x1[t] + (w1[t]-1)*x2[t]))/((sigmaE^2 + sigmaG^2) + tau^2);
+    meanfit1[t] = (theta1 - w1[t]*x1[t] - (1 - w1[t])*x2[t])/((sigmaE^2 + sigmaG^2) + tau^2);
     
-    meanfit2[t] = (w2[t]*((theta1 + thetadiff) - w2[t]*x2[t] + (w2[t]-1)*x1[t]))/((sigmaE^2 + sigmaG^2) + tau^2);
+    meanfit2[t] = ((theta1 + thetadiff) - w2[t]*x2[t] - (1 - w2[t])*x1[t])/((sigmaE^2 + sigmaG^2) + tau^2);
     
     x1[t+1] = w1[t]*x1[t] + (1-w1[t])*x2[t] + (h)*(( sigmaG^2))*meanfit1[t];
     

### 2017-11-17 model/src/KevinEvolveExtinct_ddm.jl

@@ -76,27 +76,69 @@ function KevinEvolveExtinct_ddm(
     m1[t] = 1 - (a0 + (1-a0)*(n1[t]/(C+n1[t])));
     m2[t] = 1 - (a0 + (1-a0)*(n2[t]/(C+n2[t])));
     
+    # n1[t+1] = ((1-m1[t])*n1[t] + m2[t]*n2[t])*exp(-z) +
+    # ((rfunc(x1[t],theta1) + rand(pdist))*(n1[t]-m1[t]*n1[t]) +
+    # (rfunc(x2[t],theta1) + rand(pdist))*m2[t]*n2[t]) * 
+    # exp(-beta*((1-m1[t])*n1[t]+m2[t]*n2[t]));
+    # 
+    # n2[t+1] = ((1-m2[t])*n2[t] + m1[t]*n1[t])*exp(-z) +
+    # ((rfunc(x2[t],(theta1 + thetadiff)) + rand(pdist))*(n2[t]-m2[t]*n2[t]) +
+    # (rfunc(x1[t],(theta1 + thetadiff)) + rand(pdist))*m1[t]*n1[t]) * 
+    # exp(-beta*((1-m2[t])*n2[t]+m1[t]*n1[t]));
+    # 
+    # w1[t] = (n1[t]-m1[t]*n1[t])/(n1[t]-m1[t]*n1[t]+m2[t]*n2[t]);
+    # 
+    # w2[t] = (n2[t]-m2[t]*n2[t])/(n2[t]-m2[t]*n2[t]+m1[t]*n1[t]);
+    # 
+    # meanfit1[t] = (rmax*w1[t]*(theta1-x1[t])*tau*exp(-((theta1-x1[t])^2)/(2*((sigmaE^2 + sigmaG^2)+tau^2)))) / 
+    # ((((sigmaE^2 + sigmaG^2) + tau^2)^(3/2))*(((exp(-((theta1-x2[t])^2)/(2*((sigmaE^2 + sigmaG^2)+tau^2)))*rmax*(1-w1[t])*tau)/
+    # (sqrt((sigmaE^2 + sigmaG^2)+tau^2))) +  ((exp(-((theta1-x1[t])^2)/(2*((sigmaE^2 + sigmaG^2)+tau^2)))*rmax*w1[t]*tau)/(sqrt((sigmaE^2 + sigmaG^2)+tau^2)))));
+    # 
[…]

### 2017-11-17 model/src/KevinEvolveSS.jl

@@ -42,31 +42,47 @@ function KevinEvolveSS(
   
   for t=1:tmax-1
     
-    n1[t+1] = ((1-m)*n1[t] + m*n2[t])*exp(-z) +
-    ((rfunc(x1[t],theta1) + rand(pdist))*(n1[t]-m*n1[t]) +
-    (rfunc(x2[t],theta1) + rand(pdist))*m*n2[t]) * 
-    exp(-beta*((1-m)*n1[t]+m*n2[t]));
-    
-    n2[t+1] = ((1-m)*n2[t] + m*n1[t])*exp(-z) +
-    ((rfunc(x2[t],(theta1 + thetadiff)) + rand(pdist))*(n2[t]-m*n2[t]) +
-    (rfunc(x1[t],(theta1 + thetadiff)) + rand(pdist))*m*n1[t]) * 
-    exp(-beta*((1-m)*n2[t]+m*n1[t]));
-    
-    w1[t] = (n1[t]-m*n1[t])/(n1[t]-m*n1[t]+m*n2[t]);
-    
-    w2[t] = (n2[t]-m*n2[t])/(n2[t]-m*n2[t]+m*n1[t]);
-    
-    meanfit1[t] = (rmax*w1[t]*(theta1-x1[t])*tau*exp(-((theta1-x1[t])^2)/(2*((sigmaE^2 + sigmaG^2)+tau^2)))) / 
-    ((((sigmaE^2 + sigmaG^2) + tau^2)^(3/2))*(((exp(-((theta1-x2[t])^2)/(2*((sigmaE^2 + sigmaG^2)+tau^2)))*rmax*(1-w1[t])*tau)/
-    (sqrt((sigmaE^2 + sigmaG^2)+tau^2))) +  ((exp(-((theta1-x1[t])^2)/(2*((sigmaE^2 + sigmaG^2)+tau^2)))*rmax*w1[t]*tau)/(sqrt((sigmaE^2 + sigmaG^2)+tau^2)))));
-    
-    meanfit2[t] = (rmax*w2[t]*((theta1 + thetadiff)-x2[t])*tau*exp(-(((theta1 + thetadiff)-x2[t])^2)/(2*((sigmaE^2 + sigmaG^2)+tau^2)))) / 
[…]

### 2017-11-17 model/src/KevinEvolve_ddm.jl

@@ -46,31 +46,74 @@ function KevinEvolve_ddm(
     m1[t] = 1 - (a0 + (1-a0)*(n1[t]/(C+n1[t])));
     m2[t] = 1 - (a0 + (1-a0)*(n2[t]/(C+n2[t])));
     
+    # n1[t+1] = ((1-m1[t])*n1[t] + m2[t]*n2[t])*exp(-z) +
+    # ((rfunc(x1[t],theta1) + rand(pdist))*(n1[t]-m1[t]*n1[t]) +
+    # (rfunc(x2[t],theta1) + rand(pdist))*m2[t]*n2[t]) * 
+    # exp(-beta*((1-m1[t])*n1[t]+m2[t]*n2[t]));
+    # 
+    # n2[t+1] = ((1-m2[t])*n2[t] + m1[t]*n1[t])*exp(-z) +
+    # ((rfunc(x2[t],(theta1 + thetadiff)) + rand(pdist))*(n2[t]-m2[t]*n2[t]) +
+    # (rfunc(x1[t],(theta1 + thetadiff)) + rand(pdist))*m1[t]*n1[t]) * 
+    # exp(-beta*((1-m2[t])*n2[t]+m1[t]*n1[t]));
+    # 
+    # w1[t] = (n1[t]-m1[t]*n1[t])/(n1[t]-m1[t]*n1[t]+m2[t]*n2[t]);
+    # 
+    # w2[t] = (n2[t]-m2[t]*n2[t])/(n2[t]-m2[t]*n2[t]+m1[t]*n1[t]);
+    # 
+    # meanfit1[t] = (rmax*w1[t]*(theta1-x1[t])*tau*exp(-((theta1-x1[t])^2)/(2*((sigmaE^2 + sigmaG^2)+tau^2)))) / 
+    # ((((sigmaE^2 + sigmaG^2) + tau^2)^(3/2))*(((exp(-((theta1-x2[t])^2)/(2*((sigmaE^2 + sigmaG^2)+tau^2)))*rmax*(1-w1[t])*tau)/
+    # (sqrt((sigmaE^2 + sigmaG^2)+tau^2))) +  ((exp(-((theta1-x1[t])^2)/(2*((sigmaE^2 + sigmaG^2)+tau^2)))*rmax*w1[t]*tau)/(sqrt((sigmaE^2 + sigmaG^2)+tau^2)))));
+    # 
[…]

### 2017-11-17 model/src/KevinJacobian.jl

@@ -1,3 +1,5 @@
+# THIS NEEDS TO BE UPDATED! 11/16/2017
+
 function KevinJacobian(
   N1star,
   N2star,

### 2017-11-29 manuscript/FinalDraft3/msv7.tex
removed overlapping generations dynamic

@@
 We first assumed that a proportion ${\rm e}^{-Z}$ of both populations survive such that the surviving aggregated population, composed of both local individuals (at site $i$) and incoming strays (from site $j$), is $\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z}$.
-Because local individuals will recruit differently than incoming strays, the recruitment of the aggregate must incorporate two recruitment functions, given by $R_i[w_i\mu_i(t) + (1-w_i)\mu_j(t)]\left((1-m)N_i + mN_j\right)$ (Eq.
+Because local individuals will recruit differently than incoming strays, the recruitment of the aggregate must incorporate two recruitment functions, given by $R_i[\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t)]\left((1-m)N_i + mN_j\right)$ (Eq.
 \ref{eq:R}).
@@
 \begin{align}
-  &N_i(t+1) = \\ \nonumber
-  &\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z} \\ \nonumber
-  &+ R_i[w_i\mu_i(t) + (1-w_i)\mu_j(t)]\left((1-m)N_i + mN_j\right) \\ \nonumber
-  &\times {\rm e}^{-\beta ((1-m)N_i(t) + m N_j(t))},
+  &N_i(t+1) = R_i[\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\theta_i] \\ \nonumber
+  &\times \left((1-m)N_i + mN_j\right){\rm e}^{-\beta ((1-m)N_i(t) + m N_j(t))},
   \label{eq:N}
@@
 
-\noindent where
+\noindent where it is assumed that $Z>0$ and that
 
 \begin{equation}
-w_i=\frac{(1-m)N_i(t)}{(1-m) N_i(t) + m N_j(t)},
+\omega_i=\frac{(1-m)N_i(t)}{(1-m) N_i(t) + m N_j(t)},
 \end{equation} 
@@
 \begin{align}
-  &R_i[w_i\mu_i(t) + (1-w_i)\mu_j(t)] = \\ \nonumber
-  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i,w_i\mu_i(t) + (1-w_i)\mu_j(t),\sigma^2) {\rm d}x_i +\tilde{P}_i\\ \nonumber
-  &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-(w_i\mu_i(t) + (1-w_i)\mu_j(t)))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\} +\tilde{P}_i,
+  &R_i[\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\theta_i] = \\ \nonumber
+  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i,\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\sigma^2) {\rm d}x_i +\tilde{P}_i\\ \nonumber
+  &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-(\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t)))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\} +\tilde{P}_i.
 \end{align}
@@
 First, we approximate the distribution resulting from the mix of remote and local individuals following reproduction as a Gaussian where $X_i=x_i$ with probability $g(x_i)$.
-The expectation of the actual trait distribution as well as the Gaussian approximation are the same, such that ${\rm E}\{g(x_i)\} = w_i\mu_i + (1-w_i)\mu_j$, with weights corresponding to the proportion of the mixed population that are local individuals, $w_i$, and straying individuals, $1-w_i$.
+The expectation of the actual trait distribution as well as the Gaussian approximation are the same, such that ${\rm E}\{g(x_i)\} = \omega_i\mu_i + (1-\omega_i)\mu_j$, with weights corresponding to the proportion of the mixed population that are local individuals, $\omega_i$, and straying individuals, $1-\omega_i$.
 We show that the Gaussian approximation is not only exact with respect to the mean, but also is a good approximation of variance in most relevant cases (see Appendix I in the Supplemental Materials, Fig.
@@
   \label{eq:mu}
-  \mu_i(t+1) &= w_i\mu_i(t) + (1-w_i)\mu_j(t) \\ \nonumber
-  &+ h^2\sigma^2\frac{\partial}{\partial \mu_i}\ln\left(R_i[w_i \mu_i(t)+ (1-w_i)\mu_j(t)] \right),
+  \mu_i(t+1) &= \omega_i\mu_i(t) + (1-\omega_i)\mu_j(t) + h^2\sigma^2\frac{\partial}{\partial \mu^\prime}\ln\left(R_i[\mu^\prime,\theta_i] \right), \\ \nonumber
+  &= \omega_i\mu_i(t) + (1-\omega_i)\mu_j(t) + h^2\sigma^2\left(\frac{\theta_i - \omega_i\mu_i - (1-\omega_i)\mu_j}{\sigma^2+\tau^2} \right),
 \end{align}
-
-\noindent where the first two components determine the mixed normal average of the aggregated local and remote populations.
-The partial derivative in Eq.
-\ref{eq:mu} determines how the mean trait changes through time due to natural selection \citep{Lande:1976ga}, which is proportional to the change in mean fitness with respect to $\mu_i$.
-We note that the derivative is dependent on both $\mu_i$ and $\mu_j$ due to the influence of the logarithm on the sum.
-%Comparison to the Ronce and Kirkpatrick paper
-This model formulation has parallels to that proposed by Ronce and Kirkpatrick \citep{Ronce:2001dp}, where habitat specialization evolves between two populations as a function of dispersal, yet differs in that we treat trait evolution mechanistically at some cost to analytical tractability.
+given $\mu^\prime = \omega_i \mu_i(t)+ (1-\omega_i)\mu_j(t)$.
+This model formulation parallels that proposed by Ronce and Kirkpatrick \citep{Ronce:2001dp}, where habitat specialization evolves between two populations as a function of dispersal.
 \\
@@
 
-% \bibliography{aa_kevin}
-\begin{thebibliography}{10}
-\expandafter\ifx\csname urlstyle\endcsname\relax
-  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
-  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
-  \urlstyle{rm}\Url}\fi
-
-\bibitem{Brown:1977gk}
-Brown JH, Kodric-Brown A, 1977 {Turnover rates in insular biogeography: effect
-  of immigration on extinction}.
-\newblock \emph{Ecology} \textbf{58}, 445--449
-
-\bibitem{Earn:2000fm}
-Earn DJD, Levin SA, Rohani P, 2000 {Coherence and conservation}.
-\newblock \emph{Science} \textbf{290}, 1360--1364
-
-\bibitem{Anonymous:2015gf}
-Schindler DE, Armstrong JB, Reed TE, 2015 {The portfolio concept in ecology and
-  evolution}.
-\newblock \emph{Front. Ecol. Environ.} \textbf{13}, 257--263
-
-\bibitem{Abbott:2017hl}
-Abbott RE, Doak DF, Peterson ML, 2017 {Portfolio effects, climate change, and
-  the persistence of small populations: analyses on the rare plant Saussurea
-  weberi.}
-\newblock \emph{Ecology} \textbf{98}, 1071--1081
-
-\bibitem{Thibaut:2012km}
-Thibaut LM, Connolly SR, 2013 {Understanding diversity-stability relationships:
-  towards a unified model of portfolio effects.}
-\newblock \emph{Ecol. Lett.} \textbf{16}, 140--150
-
-\bibitem{MilnerGulland:2011vm}
-Milner-Gulland EJ, Fryxell JM, Sinclair ARE, 2011 \emph{{Animal Migration}}.
-\newblock Oxford: Oxford University Press
-
-\bibitem{Bell:2011ki}
-Bell G, Gonzalez A, 2011 {Adaptation and evolutionary rescue in metapopulations
-  experiencing environmental deterioration}.
-\newblock \emph{Science} \textbf{332}, 1327--1330
-
-\bibitem{Carlson:2014is}
-Carlson SM, Cunningham CJ, Westley PAH, 2014 {Evolutionary rescue in a changing
-  world}.
-\newblock \emph{Trends Ecol. Evol.} \textbf{29}, 521--530
-
-\bibitem{Muhlfeld:2014hs}
-Muhlfeld CC, Kovach RP, Jones LA, Al-Chokhachy R, Boyer MC, Leary RF, Lowe WH,
-  Luikart G, Allendorf FW, 2014 {Invasive hybridization in a threatened species
-  is accelerated by climate change}.
-\newblock \emph{Nature Climate Change} \textbf{4}, 620--624
-
-\bibitem{Lowe:2015ft}
-Lowe WH, Muhlfeld CC, Allendorf FW, 2015 {Spatial sorting promotes the spread
-  of maladaptive hybridization}.
-\newblock \emph{Trends Ecol. Evol.} \textbf{30}, 456--462
-
-\bibitem{H:2013fs}
-Westley PAH, Quinn TP, 2013 {Rates of straying by hatchery-produced Pacific
-  salmon (\emph{Oncorhynchus} spp.) and steelhead (\emph{Oncorhynchus mykiss})
-  differ among species, life history types, and populations}.
-\newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{70}, 735--746
-
-\bibitem{Keefer:2014gg}
-Keefer ML, Caudill CC, 2014 {Homing and straying by anadromous salmonids: a
-  review of mechanisms and rates}.
-\newblock \emph{Rev Fish Biol Fisher} \textbf{24}, 333--368
-
-\bibitem{Bett:2017ha}
-Bett NN, Hinch SG, Burnett NJ, Donaldson MR, Naman SM, 2017 {Causes and
-  consequences of straying into small populations of Pacific salmon}.
-\newblock \emph{Fisheries} \textbf{42}, 220--230
-
-\bibitem{Berdahl:2015kv}
-Berdahl A, Torney CJ, Schertzer E, Levin SA, 2015 {On the evolutionary
-  interplay between dispersal and local adaptation in heterogeneous
-  environments}.
-\newblock \emph{Evolution} \textbf{69}, 1390--1405
-
-\bibitem{Berdahl:2016dx}
-Berdahl A, 2016 {Collective behavior as a driver of critical transitions in
-  migratory populations}.
-\newblock \emph{Movement Ecology} \textbf{4}, 1--12
-
-\bibitem{HardestyMoore:wg}
-Hardesty-Moore M, \emph{et~al.}, 2017 {Migration in the Anthropocene}.
-\newblock \emph{Philos. T. Roy. Soc. B} \textbf{This volume}
-
-\bibitem{Simons:2004jo}
-Simons AM, 2004 {Many wrongs: The advantage of group navigation}.
-\newblock \emph{Trends Ecol. Evol.} \textbf{19}, 453--455
-
-\bibitem{Brenner:2012gl}
-Brenner RE, Moffitt SD, Grant WS, 2012 {Straying of hatchery salmon in Prince
-  William Sound, Alaska}.
-\newblock \emph{Environ Biol Fish} \textbf{94}, 179--195
-
-\bibitem{Johnson:2012fe}
-Johnson RC, Weber PK, Wikert JD, Workman ML, MacFarlane RB, Grove MJ, Schmitt
-  AK, 2012 {Managed metapopulations: do salmon hatchery
-  {\textquoteleft}sources{\textquoteright} lead to in-river
-  {\textquoteleft}sinks{\textquoteright} in conservation?}
-\newblock \emph{PLoS ONE} \textbf{7}, e28880--11
-
-\bibitem{Fullerton:2011ii}
-Fullerton AH, Lindley ST, Pess GR, Feist BE, Steel EA, Mcelhany P, 2011 {Human
-  influence on the spatial structure of threatened Pacific salmon
-  metapopulations}.
-\newblock \emph{Conserv Biol} \textbf{25}, 932--944
-
-\bibitem{Quinn:2011tf}
-Quinn TP, 2011 \emph{{The Behavior and Ecology of Pacific Salmon and Trout}}.
-\newblock Vancouver: UBC Press
-
-\bibitem{Jonsson:2011kg}
-Jonsson B, Jonsson N, 2011 \emph{{Ecology of Atlantic Salmon and Brown Trout}}.
-\newblock Dordrecht: Springer Netherlands
-
-\bibitem{Quinn:1993ge}
-Quinn TP, 1993 {A review of homing and straying of wild and hatchery-produced
-  salmon}.
-\newblock \emph{Fish Res} \textbf{18}, 29--44
-
-\bibitem{Hendry:2004wf}
-Hendry AP, Bohlin T, Jonsson B, Berg OK, 2004 {The evolution of philopatry and
-  dispersal: homing versus straying in salmonids}.
-\newblock In AP~Hendry, SC~Stearns, eds., \emph{Evolution Illuminated}. Oxford:
-  Oxford University Press on Demand
-
-\bibitem{Schtickzelle:2007wb}
-Schtickzelle N, Quinn TP, 2007 {A metapopulation perspective for salmon and
-  other anadromous fish}.
-\newblock \emph{Fish Fish.} \textbf{8}, 297--314
-
-\bibitem{Anderson:2014cx}
-Anderson SC, Moore JW, McClure MM, Dulvy NK, Cooper AB, 2015 {Portfolio
-  conservation of metapopulations under climate change}.
-\newblock \emph{Ecol. Appl.} \textbf{25}, 559--572
-
-\bibitem{Keefer:2008bs}
-Keefer ML, Caudill CC, Peery CA, Lee SR, 2008 {Transporting juvenile salmonids
-  around dams impairs adult migration}.
-\newblock \emph{Ecol. Appl.} \textbf{18}, 1888--1900
-
-\bibitem{Westley:2015to}
-Westley PAH, Dittman AH, Ward EJ, Quinn TP, 2015 {Signals of climate,
-  conspecific density, and watershed features in patterns of homing and
-  dispersal by Pacific salmon.}
-\newblock \emph{Ecology} \textbf{96}, 2823--2833
-
-\bibitem{Bond:2016dz}
-Bond MH, Westley PAH, Dittman AH, Holecek D, Marsh T, Quinn TP, 2016 {Combined
-  effects of barge transportation, river environment, and rearing location on
-  straying and migration of adult snake river fall-run Chinook salmon}.
-\newblock \emph{T Am Fish Soc} \textbf{146}, 60--73
-
-\bibitem{Peterson:2014gy}
-Peterson DA, Hilborn R, Hauser L, 2014 {Local adaptation limits lifetime
-  reproductive success of dispersers in a wild salmon metapopulation}.
-\newblock \emph{Nat Commun} \textbf{5}, 3696
-
-\bibitem{Berdahl:2017uu}
-Berdahl A, 2017 {Berdahl et al. contribution}.
-\newblock \emph{Philos. T. Roy. Soc. B} \textbf{This volume}
-
-\bibitem{Moore:2010gs}
-Moore JW, McClure M, Rogers LA, Schindler DE, 2010 {Synchronization and
-  portfolio performance of threatened salmon}.
-\newblock \emph{Conserv. Lett.} \textbf{3}, 340--348
-
-\bibitem{Carlson:2011ce}
-Carlson SM, Satterthwaite WH, Fleming IA, 2011 {Weakened portfolio effect in a
-  collapsed salmon population complex}.
-\newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{68}, 1579--1589
-
-\bibitem{Braun:2016ib}
-Braun DC, Moore JW, Candy J, Bailey RE, 2016 {Population diversity in salmon:
-  linkages among response, genetic and life history diversity}.
-\newblock \emph{Ecography} \textbf{39}, 317--328
-
-\bibitem{simpson1953major}
-Simpson GG, 1953 \emph{{The Major Features of Evolution}}.
-\newblock New York: Simon and Schuster
-
-\bibitem{Lande:1976ga}
-Lande R, 1976 {Natural selection and random genetic drift in phenotypic
-  evolution}.
-\newblock \emph{Evolution} \textbf{30}, 314--334
-
-\bibitem{Shelton:2011eq}
-Shelton AO, Mangel M, 2011 {Fluctuations of fish populations and the magnifying
-  effects of fishing.}
-\newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{108}, 7075--7080
-
-\bibitem{Jasper:2013cc}
-Jasper JR, Habicht C, Moffitt S, Brenner R, Marsh J, Lewis B, Fox EC, Grauvogel
-  Z, Olive SDR, Grant WS, 2013 {Source-sink estimates of genetic introgression
-  show influence of hatchery strays on wild chum salmon populations in Prince
-  William Sound, Alaska}.
-\newblock \emph{PLoS ONE} \textbf{8}, e81916
-
-\bibitem{Schreiber:2011wx}
-Schreiber SJ, B{\"u}rger R, Bolnick DI, 2011 {The community effects of
-  phenotypic and genetic variation within a predator population.}
-\newblock \emph{Ecology} \textbf{92}, 1582--1593
-
-\bibitem{Gilbert:2014ee}
-Gibert JP, Brassil CE, 2014 {Individual phenotypic variation reduces
-  interaction strengths in a consumer{\textendash}resource system}.
-\newblock \emph{Ecol Evol} \textbf{4}, 3703--3713
-
-\bibitem{Gibert:2015kc}
-Gibert JP, DeLong JP, 2015 {Individual variation decreases interference
-  competition but increases species persistence}.
-\newblock \emph{Adv Ecol Res} \textbf{52}, 45--64
-
-\bibitem{Ronce:2001dp}
-Ronce O, Kirkpatrick M, 2001 {When sources become sinks: migrational meltdown
-  in heterogeneous habitats}.
-\newblock \emph{Evolution} \textbf{55}, 1520
-
-\bibitem{Berdahl:2014bl}
-Berdahl A, Westley PAH, Levin SA, Couzin ID, Quinn TP, 2014 {A collective
-  navigation hypothesis for homeward migration in anadromous salmonids}.
-\newblock \emph{Fish Fish.} \textbf{17}, 525--542
-
-\bibitem{Candy:2000hu}
-Candy JR, Beacham TD, 2000 {Patterns of homing and straying in southern British
-  Columbia coded-wire tagged Chinook salmon (\emph{Oncorhynchus tshawytscha})
-  populations}.
-\newblock \emph{Fish Res} \textbf{47}, 41--56
-
-\bibitem{JPE:JPE1383}
-Schick RS, Lindley ST, 2007 {Directed connectivity among fish populations in a
-  riverine network}.
-\newblock \emph{J. Appl. Ecol.} \textbf{44}, 1116--1126
-
-\bibitem{Schindler:2015gf}
-Schindler DE, Armstrong JB, Reed TE, 2015 {The portfolio concept in ecology and
-  evolution}.
-\newblock \emph{Front. Ecol. Environ.} \textbf{13}, 257--263
-
-\bibitem{Ovaskainen:2002il}
-Ovaskainen O, Hanski I, 2002 {Transient dynamics in metapopulation response to
-  perturbation}.
-\newblock \emph{Theor Popul Biol} \textbf{61}, 285--295
-
-\bibitem{Anderson:2013gb}
-Anderson SC, Cooper AB, Dulvy NK, 2013 {Ecological prophets: quantifying
-  metapopulation portfolio effects}.
-\newblock \emph{Methods Ecol Evol} \textbf{4}, 971--981
-
-\bibitem{Loreau:2008ju}
-Loreau M, de~Mazancourt C, 2008 {Species synchrony and its drivers: neutral and
-  nonneutral community dynamics in fluctuating environments}.
-\newblock \emph{Am. Nat.} \textbf{172}, E48--E66
-
-\bibitem{Yeakel:2013vz}
-Yeakel JD, Moore JW, Guimar{\~a}es~Jr PR, de~Aguiar MAM, 2014 {Synchronisation
-  and stability in river metapopulation networks}.
-\newblock \emph{Ecol. Lett.} \textbf{17}, 273--283
-
-\bibitem{GuckHolmes}
-Guckenheimer J, Holmes P, 1983 \emph{{Nonlinear Oscillations, Dynamical
-  Systems, and Bifurcations of Vector Fields}}.
-\newblock New York: Springer
-
-\bibitem{AleksandrovichKuznetsov:1995p2580}
-Kuznetsov Y, 1998 \emph{{Elements of Applied Bifurcation Theory}}.
-\newblock New York: Springer
-
-\bibitem{Carlson:2008hl}
-Carlson SM, Seamons TR, 2008 {A review of quantitative genetic components of
-  fitness in salmonids: implications for adaptation to future change}.
-\newblock \emph{Evol Appl} \textbf{1}, 222--238
-
-\bibitem{Scheffer:2009gg}
-Scheffer M, Bascompte J, Brock WA, Brovkin V, Carpenter SR, Dakos V, Held H,
-  van Nes EH, Rietkerk M, Sugihara G, 2009 {Early-warning signals for critical
-  transitions}.
-\newblock \emph{Nature} \textbf{461}, 53--59
-
-\bibitem{Lade:2012eu}
-Lade SJ, Gross T, 2012 {Early warning signals for critical transitions: A
-  generalized modeling approach}.
-\newblock \emph{PLoS Comp. Biol.} \textbf{8}, e1002360
-
-\bibitem{Anonymous:2013br}
-Boettiger C, Ross N, Hastings A, 2013 {Early warning signals: the charted and
-  uncharted territories}.
-\newblock \emph{Theor. Ecol.} \textbf{6}, 255--264
-
-\bibitem{Dakos:2014br}
-Dakos V, Bascompte J, 2014 {Critical slowing down as early warning for the
-  onset of collapse in mutualistic communities}.
-\newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{111}, 17546--17551
-
-\bibitem{Krkosek:2014ch}
-Krko{\v s}ek M, Drake JM, 2014 {On signals of phase transitions in salmon
-  population dynamics}.
-\newblock \emph{Proc. Roy. Soc. B} \textbf{281}, 20133221
-
-\bibitem{Kuehn:2010p2591}
-Kuehn C, 2011 {A mathematical framework for critical transitions: bifurcations,
-  fast-slow systems and stochastic dynamics}.
-\newblock \emph{Physica D} \textbf{240}, 1020--1035
-
-\bibitem{Donohue:2013iu}
-Donohue I, Petchey OL, Montoya JM, Jackson AL, McNally L, Viana M, Healy K,
-  Lurgi M, O'Connor NE, Emmerson MC, 2013 {On the dimensionality of ecological
-  stability.}
-\newblock \emph{Ecol. Lett.} \textbf{16}, 421--429
-
-\bibitem{Secor:2009ena}
-Secor DH, Kerr LA, Cadrin SX, 2009 {Connectivity effects on productivity,
-  stability, and persistence in a herring metapopulation model}.
-\newblock \emph{ICES J Mar Sci} \textbf{66}, 1726--1732
-
-\bibitem{Christie:2012bj}
-Christie MR, Marine ML, French RA, Blouin MS, 2012 {Genetic adaptation to
-  captivity can occur in a single generation.}
-\newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{109}, 238--242
-
-\bibitem{Chilcote:2003bb}
-Chilcote MW, 2003 {Relationship between natural productivity and the frequency
-  of wild fish in mixed spawning populations of wild and hatchery steelhead
-  (\emph{Oncorhynchus mykiss})}.
-\newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{60}, 1057--1067
-
-\bibitem{Araki:2007cm}
-Araki H, Cooper B, Blouin MS, 2007 {Genetic effects of captive breeding cause a
-  rapid, cumulative fitness decline in the wild}.
-\newblock \emph{Science} \textbf{318}, 100--103
-
-\bibitem{Reisenbichler:1988ex}
-Reisenbichler RR, 1988 {Relation between Distance Transferred from Natal Stream
-  and Recovery Rate for Hatchery Coho Salmon}.
-\newblock \emph{N. Am. J. Fish. Manage.} \textbf{8}, 172--174
-
-\bibitem{Fraser:2011co}
-Fraser DJ, Weir LK, Bernatchez L, Hansen MM, Taylor EB, 2011 {Extent and scale
-  of local adaptation in salmonid fishes: review and meta-analysis.}
-\newblock \emph{Heredity} \textbf{106}, 404--420
-
-\bibitem{Westley:2012ui}
-Westley PAH, { Corinne M Conway}, Fleming IA, 2012 {Phenotypic divergence of
-  exotic fish populations is shaped by spatial proximity and habitat
-  differences across an invaded landscape}.
-\newblock \emph{Evol. Ecol. Res.} \textbf{14}, 147--167
-
-\bibitem{Anderson:2013bf}
-Anderson JH, Faulds PL, Atlas WI, Quinn TP, 2013 {Reproductive success of
-  captively bred and naturally spawned Chinook salmon colonizing newly
-  accessible habitat}.
-\newblock \emph{Evol Appl} \textbf{6}, 165--179
-
-\bibitem{Pess:2014isa}
-Pess GR, Quinn TP, Gephard SR, Saunders R, 2014 {Re-colonization of Atlantic
-  and Pacific rivers by anadromous fishes: linkages between life history and
-  the benefits of barrier removal}.
-\newblock \emph{Rev Fish Biol Fisher} \textbf{24}, 881--900
-
-\bibitem{Ford:2002ip}
-Ford MJ, 2002 {Selection in captivity during supportive breeding may reduce
-  fitness in the wild}.
-\newblock \emph{Conserv Biol} \textbf{16}, 815--825
-
-\bibitem{Liermann:2017gj}
-Liermann M, PESS G, McHenry M, McMillan J, Elofson M, Bennett T, Moses R, 2017
-  {Relocation and recolonization of Coho salmon \emph{Oncorhynchus kisutch} in
-  two tributaries to the elwha river: Implications for management and
-  monitoring}.
-\newblock \emph{T Am Fish Soc} \textbf{17}, 1--39
-
-\end{thebibliography}
+\bibliography{aa_kevin}
+% \begin{thebibliography}{10}
+% \expandafter\ifx\csname urlstyle\endcsname\relax
+%   \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
+%   \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
+%   \urlstyle{rm}\Url}\fi
+% 
+% \bibitem{Brown:1977gk}
+% Brown JH, Kodric-Brown A, 1977 {Turnover rates in insular biogeography: effect
+%   of immigration on extinction}.
+% \newblock \emph{Ecology} \textbf{58}, 445--449
+% 
+% \bibitem{Earn:2000fm}
+% Earn DJD, Levin SA, Rohani P, 2000 {Coherence and conservation}.
+% \newblock \emph{Science} \textbf{290}, 1360--1364
+% 
+% \bibitem{Anonymous:2015gf}
+% Schindler DE, Armstrong JB, Reed TE, 2015 {The portfolio concept in ecology and
+%   evolution}.
+% \newblock \emph{Front. Ecol. Environ.} \textbf{13}, 257--263
+% 
+% \bibitem{Abbott:2017hl}
+% Abbott RE, Doak DF, Peterson ML, 2017 {Portfolio effects, climate change, and
+%   the persistence of small populations: analyses on the rare plant Saussurea
+%   weberi.}
+% \newblock \emph{Ecology} \textbf{98}, 1071--1081
+% 
+% \bibitem{Thibaut:2012km}
+% Thibaut LM, Connolly SR, 2013 {Understanding diversity-stability relationships:
+%   towards a unified model of portfolio effects.}
+% \newblock \emph{Ecol. Lett.} \textbf{16}, 140--150
+% 
+% \bibitem{MilnerGulland:2011vm}
+% Milner-Gulland EJ, Fryxell JM, Sinclair ARE, 2011 \emph{{Animal Migration}}.
+% \newblock Oxford: Oxford University Press
+% 
+% \bibitem{Bell:2011ki}
+% Bell G, Gonzalez A, 2011 {Adaptation and evolutionary rescue in metapopulations
+%   experiencing environmental deterioration}.
+% \newblock \emph{Science} \textbf{332}, 1327--1330
+% 
+% \bibitem{Carlson:2014is}
+% Carlson SM, Cunningham CJ, Westley PAH, 2014 {Evolutionary rescue in a changing
+%   world}.
+% \newblock \emph{Trends Ecol. Evol.} \textbf{29}, 521--530
+% 
+% \bibitem{Muhlfeld:2014hs}
+% Muhlfeld CC, Kovach RP, Jones LA, Al-Chokhachy R, Boyer MC, Leary RF, Lowe WH,
+%   Luikart G, Allendorf FW, 2014 {Invasive hybridization in a threatened species
+%   is accelerated by climate change}.
+% \newblock \emph{Nature Climate Change} \textbf{4}, 620--624
+% 
+% \bibitem{Lowe:2015ft}
+% Lowe WH, Muhlfeld CC, Allendorf FW, 2015 {Spatial sorting promotes the spread
+%   of maladaptive hybridization}.
+% \newblock \emph{Trends Ecol. Evol.} \textbf{30}, 456--462
+% 
+% \bibitem{H:2013fs}
+% Westley PAH, Quinn TP, 2013 {Rates of straying by hatchery-produced Pacific
+%   salmon (\emph{Oncorhynchus} spp.) and steelhead (\emph{Oncorhynchus mykiss})
+%   differ among species, life history types, and populations}.
+% \newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{70}, 735--746
+% 
+% \bibitem{Keefer:2014gg}
+% Keefer ML, Caudill CC, 2014 {Homing and straying by anadromous salmonids: a
+%   review of mechanisms and rates}.
+% \newblock \emph{Rev Fish Biol Fisher} \textbf{24}, 333--368
+% 
+% \bibitem{Bett:2017ha}
+% Bett NN, Hinch SG, Burnett NJ, Donaldson MR, Naman SM, 2017 {Causes and
+%   consequences of straying into small populations of Pacific salmon}.
+% \newblock \emph{Fisheries} \textbf{42}, 220--230
+% 
+% \bibitem{Berdahl:2015kv}
+% Berdahl A, Torney CJ, Schertzer E, Levin SA, 2015 {On the evolutionary
+%   interplay between dispersal and local adaptation in heterogeneous
+%   environments}.
+% \newblock \emph{Evolution} \textbf{69}, 1390--1405
+% 
+% \bibitem{Berdahl:2016dx}
+% Berdahl A, 2016 {Collective behavior as a driver of critical transitions in
+%   migratory populations}.
+% \newblock \emph{Movement Ecology} \textbf{4}, 1--12
+% 
+% \bibitem{HardestyMoore:wg}
+% Hardesty-Moore M, \emph{et~al.}, 2017 {Migration in the Anthropocene}.
+% \newblock \emph{Philos. T. Roy. Soc. B} \textbf{This volume}
+% 
+% \bibitem{Simons:2004jo}
+% Simons AM, 2004 {Many wrongs: The advantage of group navigation}.
+% \newblock \emph{Trends Ecol. Evol.} \textbf{19}, 453--455
+% 
+% \bibitem{Brenner:2012gl}
+% Brenner RE, Moffitt SD, Grant WS, 2012 {Straying of hatchery salmon in Prince
+%   William Sound, Alaska}.
+% \newblock \emph{Environ Biol Fish} \textbf{94}, 179--195
+% 
+% \bibitem{Johnson:2012fe}
+% Johnson RC, Weber PK, Wikert JD, Workman ML, MacFarlane RB, Grove MJ, Schmitt
+%   AK, 2012 {Managed metapopulations: do salmon hatchery
+%   {\textquoteleft}sources{\textquoteright} lead to in-river
+%   {\textquoteleft}sinks{\textquoteright} in conservation?}
+% \newblock \emph{PLoS ONE} \textbf{7}, e28880--11
+% 
+% \bibitem{Fullerton:2011ii}
+% Fullerton AH, Lindley ST, Pess GR, Feist BE, Steel EA, Mcelhany P, 2011 {Human
+%   influence on the spatial structure of threatened Pacific salmon
+%   metapopulations}.
+% \newblock \emph{Conserv Biol} \textbf{25}, 932--944
+% 
+% \bibitem{Quinn:2011tf}
+% Quinn TP, 2011 \emph{{The Behavior and Ecology of Pacific Salmon and Trout}}.
+% \newblock Vancouver: UBC Press
+% 
+% \bibitem{Jonsson:2011kg}
+% Jonsson B, Jonsson N, 2011 \emph{{Ecology of Atlantic Salmon and Brown Trout}}.
+% \newblock Dordrecht: Springer Netherlands
+% 
+% \bibitem{Quinn:1993ge}
+% Quinn TP, 1993 {A review of homing and straying of wild and hatchery-produced
+%   salmon}.
+% \newblock \emph{Fish Res} \textbf{18}, 29--44
+% 
+% \bibitem{Hendry:2004wf}
+% Hendry AP, Bohlin T, Jonsson B, Berg OK, 2004 {The evolution of philopatry and
+%   dispersal: homing versus straying in salmonids}.
+% \newblock In AP~Hendry, SC~Stearns, eds., \emph{Evolution Illuminated}. Oxford:
+%   Oxford University Press on Demand
+% 
+% \bibitem{Schtickzelle:2007wb}
+% Schtickzelle N, Quinn TP, 2007 {A metapopulation perspective for salmon and
+%   other anadromous fish}.
+% \newblock \emph{Fish Fish.} \textbf{8}, 297--314
+% 
+% \bibitem{Anderson:2014cx}
+% Anderson SC, Moore JW, McClure MM, Dulvy NK, Cooper AB, 2015 {Portfolio
+%   conservation of metapopulations under climate change}.
+% \newblock \emph{Ecol. Appl.} \textbf{25}, 559--572
+% 
+% \bibitem{Keefer:2008bs}
+% Keefer ML, Caudill CC, Peery CA, Lee SR, 2008 {Transporting juvenile salmonids
+%   around dams impairs adult migration}.
+% \newblock \emph{Ecol. Appl.} \textbf{18}, 1888--1900
+% 
+% \bibitem{Westley:2015to}
+% Westley PAH, Dittman AH, Ward EJ, Quinn TP, 2015 {Signals of climate,
+%   conspecific density, and watershed features in patterns of homing and
+%   dispersal by Pacific salmon.}
+% \newblock \emph{Ecology} \textbf{96}, 2823--2833
+% 
+% \bibitem{Bond:2016dz}
+% Bond MH, Westley PAH, Dittman AH, Holecek D, Marsh T, Quinn TP, 2016 {Combined
+%   effects of barge transportation, river environment, and rearing location on
+%   straying and migration of adult snake river fall-run Chinook salmon}.
+% \newblock \emph{T Am Fish Soc} \textbf{146}, 60--73
+% 
+% \bibitem{Peterson:2014gy}
+% Peterson DA, Hilborn R, Hauser L, 2014 {Local adaptation limits lifetime
+%   reproductive success of dispersers in a wild salmon metapopulation}.
+% \newblock \emph{Nat Commun} \textbf{5}, 3696
+% 
+% \bibitem{Berdahl:2017uu}
+% Berdahl A, 2017 {Berdahl et al. contribution}.
+% \newblock \emph{Philos. T. Roy. Soc. B} \textbf{This volume}
+% 
+% \bibitem{Moore:2010gs}
+% Moore JW, McClure M, Rogers LA, Schindler DE, 2010 {Synchronization and
+%   portfolio performance of threatened salmon}.
+% \newblock \emph{Conserv. Lett.} \textbf{3}, 340--348
+% 
+% \bibitem{Carlson:2011ce}
+% Carlson SM, Satterthwaite WH, Fleming IA, 2011 {Weakened portfolio effect in a
+%   collapsed salmon population complex}.
+% \newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{68}, 1579--1589
+% 
+% \bibitem{Braun:2016ib}
+% Braun DC, Moore JW, Candy J, Bailey RE, 2016 {Population diversity in salmon:
+%   linkages among response, genetic and life history diversity}.
+% \newblock \emph{Ecography} \textbf{39}, 317--328
+% 
+% \bibitem{simpson1953major}
+% Simpson GG, 1953 \emph{{The Major Features of Evolution}}.
+% \newblock New York: Simon and Schuster
+% 
+% \bibitem{Lande:1976ga}
+% Lande R, 1976 {Natural selection and random genetic drift in phenotypic
+%   evolution}.
+% \newblock \emph{Evolution} \textbf{30}, 314--334
+% 
+% \bibitem{Shelton:2011eq}
+% Shelton AO, Mangel M, 2011 {Fluctuations of fish populations and the magnifying
+%   effects of fishing.}
+% \newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{108}, 7075--7080
+% 
+% \bibitem{Jasper:2013cc}
+% Jasper JR, Habicht C, Moffitt S, Brenner R, Marsh J, Lewis B, Fox EC, Grauvogel
+%   Z, Olive SDR, Grant WS, 2013 {Source-sink estimates of genetic introgression
+%   show influence of hatchery strays on wild chum salmon populations in Prince
+%   William Sound, Alaska}.
+% \newblock \emph{PLoS ONE} \textbf{8}, e81916
+% 
+% \bibitem{Schreiber:2011wx}
+% Schreiber SJ, B{\"u}rger R, Bolnick DI, 2011 {The community effects of
+%   phenotypic and genetic variation within a predator population.}
+% \newblock \emph{Ecology} \textbf{92}, 1582--1593
+% 
+% \bibitem{Gilbert:2014ee}
+% Gibert JP, Brassil CE, 2014 {Individual phenotypic variation reduces
+%   interaction strengths in a consumer{\textendash}resource system}.
+% \newblock \emph{Ecol Evol} \textbf{4}, 3703--3713
+% 
+% \bibitem{Gibert:2015kc}
+% Gibert JP, DeLong JP, 2015 {Individual variation decreases interference
+%   competition but increases species persistence}.
+% \newblock \emph{Adv Ecol Res} \textbf{52}, 45--64
+% 
+% \bibitem{Ronce:2001dp}
+% Ronce O, Kirkpatrick M, 2001 {When sources become sinks: migrational meltdown
+%   in heterogeneous habitats}.
+% \newblock \emph{Evolution} \textbf{55}, 1520
+% 
+% \bibitem{Berdahl:2014bl}
+% Berdahl A, Westley PAH, Levin SA, Couzin ID, Quinn TP, 2014 {A collective
+%   navigation hypothesis for homeward migration in anadromous salmonids}.
+% \newblock \emph{Fish Fish.} \textbf{17}, 525--542
+% 
+% \bibitem{Candy:2000hu}
+% Candy JR, Beacham TD, 2000 {Patterns of homing and straying in southern British
+%   Columbia coded-wire tagged Chinook salmon (\emph{Oncorhynchus tshawytscha})
+%   populations}.
+% \newblock \emph{Fish Res} \textbf{47}, 41--56
+% 
+% \bibitem{JPE:JPE1383}
+% Schick RS, Lindley ST, 2007 {Directed connectivity among fish populations in a
+%   riverine network}.
+% \newblock \emph{J. Appl. Ecol.} \textbf{44}, 1116--1126
+% 
+% \bibitem{Schindler:2015gf}
+% Schindler DE, Armstrong JB, Reed TE, 2015 {The portfolio concept in ecology and
+%   evolution}.
+% \newblock \emph{Front. Ecol. Environ.} \textbf{13}, 257--263
+% 
+% \bibitem{Ovaskainen:2002il}
+% Ovaskainen O, Hanski I, 2002 {Transient dynamics in metapopulation response to
+%   perturbation}.
+% \newblock \emph{Theor Popul Biol} \textbf{61}, 285--295
+% 
+% \bibitem{Anderson:2013gb}
+% Anderson SC, Cooper AB, Dulvy NK, 2013 {Ecological prophets: quantifying
+%   metapopulation portfolio effects}.
+% \newblock \emph{Methods Ecol Evol} \textbf{4}, 971--981
+% 
+% \bibitem{Loreau:2008ju}
+% Loreau M, de~Mazancourt C, 2008 {Species synchrony and its drivers: neutral and
+%   nonneutral community dynamics in fluctuating environments}.
+% \newblock \emph{Am. Nat.} \textbf{172}, E48--E66
+% 
+% \bibitem{Yeakel:2013vz}
+% Yeakel JD, Moore JW, Guimar{\~a}es~Jr PR, de~Aguiar MAM, 2014 {Synchronisation
+%   and stability in river metapopulation networks}.
+% \newblock \emph{Ecol. Lett.} \textbf{17}, 273--283
+% 
+% \bibitem{GuckHolmes}
+% Guckenheimer J, Holmes P, 1983 \emph{{Nonlinear Oscillations, Dynamical
+%   Systems, and Bifurcations of Vector Fields}}.
+% \newblock New York: Springer
+% 
+% \bibitem{AleksandrovichKuznetsov:1995p2580}
+% Kuznetsov Y, 1998 \emph{{Elements of Applied Bifurcation Theory}}.
+% \newblock New York: Springer
+% 
+% \bibitem{Carlson:2008hl}
+% Carlson SM, Seamons TR, 2008 {A review of quantitative genetic components of
+%   fitness in salmonids: implications for adaptation to future change}.
+% \newblock \emph{Evol Appl} \textbf{1}, 222--238
+% 
+% \bibitem{Scheffer:2009gg}
+% Scheffer M, Bascompte J, Brock WA, Brovkin V, Carpenter SR, Dakos V, Held H,
+%   van Nes EH, Rietkerk M, Sugihara G, 2009 {Early-warning signals for critical
+%   transitions}.
+% \newblock \emph{Nature} \textbf{461}, 53--59
+% 
+% \bibitem{Lade:2012eu}
+% Lade SJ, Gross T, 2012 {Early warning signals for critical transitions: A
+%   generalized modeling approach}.
+% \newblock \emph{PLoS Comp. Biol.} \textbf{8}, e1002360
+% 
+% \bibitem{Anonymous:2013br}
+% Boettiger C, Ross N, Hastings A, 2013 {Early warning signals: the charted and
+%   uncharted territories}.
+% \newblock \emph{Theor. Ecol.} \textbf{6}, 255--264
+% 
+% \bibitem{Dakos:2014br}
+% Dakos V, Bascompte J, 2014 {Critical slowing down as early warning for the
+%   onset of collapse in mutualistic communities}.
+% \newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{111}, 17546--17551
+% 
+% \bibitem{Krkosek:2014ch}
+% Krko{\v s}ek M, Drake JM, 2014 {On signals of phase transitions in salmon
+%   population dynamics}.
+% \newblock \emph{Proc. Roy. Soc. B} \textbf{281}, 20133221
+% 
+% \bibitem{Kuehn:2010p2591}
+% Kuehn C, 2011 {A mathematical framework for critical transitions: bifurcations,
+%   fast-slow systems and stochastic dynamics}.
+% \newblock \emph{Physica D} \textbf{240}, 1020--1035
+% 
+% \bibitem{Donohue:2013iu}
+% Donohue I, Petchey OL, Montoya JM, Jackson AL, McNally L, Viana M, Healy K,
+%   Lurgi M, O'Connor NE, Emmerson MC, 2013 {On the dimensionality of ecological
+%   stability.}
+% \newblock \emph{Ecol. Lett.} \textbf{16}, 421--429
+% 
+% \bibitem{Secor:2009ena}
+% Secor DH, Kerr LA, Cadrin SX, 2009 {Connectivity effects on productivity,
+%   stability, and persistence in a herring metapopulation model}.
+% \newblock \emph{ICES J Mar Sci} \textbf{66}, 1726--1732
+% 
+% \bibitem{Christie:2012bj}
+% Christie MR, Marine ML, French RA, Blouin MS, 2012 {Genetic adaptation to
+%   captivity can occur in a single generation.}
+% \newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{109}, 238--242
+% 
+% \bibitem{Chilcote:2003bb}
+% Chilcote MW, 2003 {Relationship between natural productivity and the frequency
+%   of wild fish in mixed spawning populations of wild and hatchery steelhead
+%   (\emph{Oncorhynchus mykiss})}.
+% \newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{60}, 1057--1067
+% 
+% \bibitem{Araki:2007cm}
+% Araki H, Cooper B, Blouin MS, 2007 {Genetic effects of captive breeding cause a
+%   rapid, cumulative fitness decline in the wild}.
+% \newblock \emph{Science} \textbf{318}, 100--103
+% 
+% \bibitem{Reisenbichler:1988ex}
+% Reisenbichler RR, 1988 {Relation between Distance Transferred from Natal Stream
+%   and Recovery Rate for Hatchery Coho Salmon}.
+% \newblock \emph{N. Am. J. Fish. Manage.} \textbf{8}, 172--174
+% 
+% \bibitem{Fraser:2011co}
+% Fraser DJ, Weir LK, Bernatchez L, Hansen MM, Taylor EB, 2011 {Extent and scale
+%   of local adaptation in salmonid fishes: review and meta-analysis.}
+% \newblock \emph{Heredity} \textbf{106}, 404--420
+% 
+% \bibitem{Westley:2012ui}
+% Westley PAH, { Corinne M Conway}, Fleming IA, 2012 {Phenotypic divergence of
+%   exotic fish populations is shaped by spatial proximity and habitat
+%   differences across an invaded landscape}.
+% \newblock \emph{Evol. Ecol. Res.} \textbf{14}, 147--167
+% 
+% \bibitem{Anderson:2013bf}
+% Anderson JH, Faulds PL, Atlas WI, Quinn TP, 2013 {Reproductive success of
+%   captively bred and naturally spawned Chinook salmon colonizing newly
+%   accessible habitat}.
+% \newblock \emph{Evol Appl} \textbf{6}, 165--179
+% 
+% \bibitem{Pess:2014isa}
+% Pess GR, Quinn TP, Gephard SR, Saunders R, 2014 {Re-colonization of Atlantic
+%   and Pacific rivers by anadromous fishes: linkages between life history and
+%   the benefits of barrier removal}.
+% \newblock \emph{Rev Fish Biol Fisher} \textbf{24}, 881--900
+% 
+% \bibitem{Ford:2002ip}
+% Ford MJ, 2002 {Selection in captivity during supportive breeding may reduce
+%   fitness in the wild}.
+% \newblock \emph{Conserv Biol} \textbf{16}, 815--825
+% 
+% \bibitem{Liermann:2017gj}
+% Liermann M, PESS G, McHenry M, McMillan J, Elofson M, Bennett T, Moses R, 2017
+%   {Relocation and recolonization of Coho salmon \emph{Oncorhynchus kisutch} in
+%   two tributaries to the elwha river: Implications for management and
+%   monitoring}.
+% \newblock \emph{T Am Fish Soc} \textbf{17}, 1--39
+% 
+% \end{thebibliography}
+%
 

### 2017-11-29 model/KevinExtinction.jl

@@ -459,177 +459,3 @@ for (i in 1:length($mvec)) {
      
 dev.off()
 """
-
-
-
-
-
-#############
-# Calculate return time as a function of 'm' and 'Z'
-#############
-
-
-
-@everywhere using Distributions
-@everywhere using RCall
-@everywhere using JLD
-@everywhere using HDF5
-
-@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve.jl")
-@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve_ddm.jl")
-
-
-@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolveExtinct.jl")
-@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolveExtinct_ddm.jl")
-@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/timeSS.jl")
-@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/movingaverage.jl")
-
-
-#Analysis over m
-tmax=10000;
-
-mvec = collect(0.0001:0.01:0.5);
-pvec = ["small","large","both"];
-h = 0.2;
-zvec = logspace(-1,log10(5),50);
-reps = 50;
-
-# z=0.5;
-rmax=2.0;
-beta=0.001;
-theta1=5.0;
-thetadiff=5;
-tau=1.0;
-C=1000;
-sigmaE=0;
-sigmaG=1;
-perror=0.01;
-refuge=0.01;
-
-burnin = 0.8;
-
-# n1v = SharedArray{Float64}(reps,length(mvec),length(pvec),tmax);
-# n2v = SharedArray{Float64}(reps,length(mvec),length(pvec),tmax);
-# x1v = SharedArray{Float64}(reps,length(mvec),length(pvec),tmax);
-# x2v = SharedArray{Float64}(reps,length(mvec),length(pvec),tmax);
-
-t_ext = Int64(round(tmax/2));
-
-
-rt = SharedArray{Float64}(reps,length(mvec),length(pvec),length(zvec));
-rt_ddm = SharedArray{Float64}(reps,length(mvec),length(pvec),length(zvec));
[…]

### 2017-11-29 model/model2/Figure2_4.jl

@@ -33,11 +33,11 @@ peE=SharedArray(Float64,length(hvec),length(mvec));
 rtE=SharedArray(Float64,length(hvec),length(mvec));
 
 tmax=10000;
-z=0.5;
+z=2;
 rmax=2.0;
 beta=0.001;
 theta1=5.0;
-thetadiff=4.0;
+thetadiff=2.4;
 tau=1.0;
 sigmaE=0;
 sigmaG=1;
@@ -151,16 +151,16 @@ library(RColorBrewer)
 pal = rev(brewer.pal(9,"Blues"))
 pdf($namespace,height=3,width=10)
 par(mfrow=c(1,4))
-image(x=$mvec,y=$hvec,z=t($(n1mean[:,:]))+t($(n2mean[:,:])),zlim=c(0,3000),col=pal,xlab='m',ylab=expression(paste(h^2)),las=1)
+image(x=$mvec,y=$hvec,z=t($(n1mean[:,:]))+t($(n2mean[:,:])),zlim=c(0,1000),col=pal,xlab='m',ylab=expression(paste(h^2)),las=1)
 #points($(bifvalue),type='l',cex=1)
 text(par('usr')[1]-0.12,1.05,'(a)', xpd=TRUE)
-image(x=$mvec,y=$hvec,z=sqrt((t($(n1mean[:,:]))-t($(n2mean[:,:])))^2),zlim=c(0,1500),col=pal,xlab='m',ylab=expression(paste(h^2)),las=1)
+image(x=$mvec,y=$hvec,z=sqrt((t($(n1mean[:,:]))-t($(n2mean[:,:])))^2),zlim=c(0,500),col=pal,xlab='m',ylab=expression(paste(h^2)),las=1)
 #points($(bifvalue),type='l',cex=1)
 text(par('usr')[1]-0.12,1.05,'(b)', xpd=TRUE)
 image(x=$mvec,y=$hvec,z=t($(pe[:,:])),zlim=c(1,2),col=pal,xlab='m',ylab=expression(paste(h^2)),las=1)
 #points($(bifvalue),type='l',cex=1)
 text(par('usr')[1]-0.12,1.05,'(c)', xpd=TRUE)
-plot($(rt_ext[1:50,:]),$(pe_ext[1:50,:]),log='xy',ylim=c(1,3),xlim=c(18,150),pch='.',xlab='Recovery time',ylab='PE',las=1)
[…]

### 2017-11-29 model/model2/Figure5.jl

@@ -1,5 +1,8 @@
 #Figure 5
 
+#3 = Low habitat heterogeneity
+#5 = Moderate habitat heterogeneity
+#8 = High habitat heterogeneity
 
 @everywhere using Distributions, RCall, JLD, HDF5
 
@@ -14,7 +17,7 @@
 
 #Analysis over m & theta divergence
 mvec=collect(0.0:0.001:0.3);
-hvec = collect(0.0:0.01:1.0);
+hvec = collect(0.0:0.01:0.5);
 
 
 n1mean=SharedArray(Float64,length(hvec),length(mvec));
@@ -28,7 +31,7 @@ z=0.5;
 rmax=2.0;
 beta=0.001;
 theta1=5.0;
-thetadiff=5.0;
+thetadiff=2.5;
 tau=1.0;
 sigmaE=0;
 sigmaG=1;
@@ -77,7 +80,7 @@ save(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data3/data_sig
 
 #Analysis over m & theta divergence
 mvec=collect(0.0:0.001:0.3);
-hvec = collect(0.0:0.01:1.0);
+hvec = collect(0.0:0.01:0.5);
 
 
 n1mean=SharedArray(Float64,length(hvec),length(mvec));
@@ -91,7 +94,7 @@ z=0.5;
 rmax=2.0;
 beta=0.001;
 theta1=5.0;
-thetadiff=3.0;
+thetadiff=2.0;
 tau=1.0;
 sigmaE=0;
 sigmaG=1;
@@ -141,7 +144,7 @@ save(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data3/data_sig
 
 #Analysis over m & theta divergence
 mvec=collect(0.0:0.001:0.3);
-hvec = collect(0.0:0.01:1.0);
+hvec = collect(0.0:0.01:0.5);
 
 
 n1mean=SharedArray(Float64,length(hvec),length(mvec));
@@ -155,7 +158,7 @@ z=0.5;
 rmax=2.0;
 beta=0.001;
 theta1=5.0;
-thetadiff=8.0;
+thetadiff=3.0;
 tau=1.0;
 sigmaE=0;
 sigmaG=1;
@@ -203,7 +206,7 @@ save(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data3/data_sig
 
[…]

### 2017-11-29 model/model2/Figure_overlappingpops.jl

@@ -1,0 +1,190 @@
+
+
+
+#############
+# Calculate return time as a function of 'm' and 'Z'
+#############
+
+
+
+@everywhere using Distributions
+@everywhere using RCall
+@everywhere using JLD
+@everywhere using HDF5
+
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve_ddm.jl")
+
+
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolveExtinct.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolveExtinct_ddm.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/timeSS.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/movingaverage.jl")
+
+
+#Analysis over m
+tmax=10000;
+
+mvec = collect(0.0001:0.001:0.3);
+pvec = ["both"];
+h = 0.2;
+zvec = logspace(-1,log10(5),200);
+reps = 50;
+
+# z=0.5;
+rmax=2.0;
+beta=0.001;
+theta1=5.0;
+thetadiff=5;
+tau=1.0;
+C=1000;
+sigmaE=0;
+sigmaG=1;
+perror=0.01;
+refuge=0.01;
+
+burnin = 0.8;
+
+# n1v = SharedArray{Float64}(reps,length(mvec),length(pvec),tmax);
+# n2v = SharedArray{Float64}(reps,length(mvec),length(pvec),tmax);
+# x1v = SharedArray{Float64}(reps,length(mvec),length(pvec),tmax);
+# x2v = SharedArray{Float64}(reps,length(mvec),length(pvec),tmax);
+
+t_ext = Int64(round(tmax/2));
+
+
+rt = SharedArray{Float64}(reps,length(mvec),length(pvec),length(zvec));
+rt_ddm = SharedArray{Float64}(reps,length(mvec),length(pvec),length(zvec));
+m1mean = SharedArray{Float64}(reps,length(mvec),length(pvec),length(zvec));
[…]

### 2017-11-29 model/src/KevinEvolve.jl

@@ -52,12 +52,18 @@ function KevinEvolve(
     # (rfunc(x1[t],(theta1 + thetadiff)) + rand(pdist))*m*n1[t]) * 
     # exp(-beta*((1-m)*n2[t]+m*n1[t]));
     
-    n1[t+1] = ((1-m)*n1[t] + m*n2[t])*exp(-z) +
-    ((rfunc(w1[t]*x1[t] + (1-w1[t])*x2[t],theta1) + rand(pdist))*((1-m)*n1[t]+m*n2[t])) * 
+    # n1[t+1] = ((1-m)*n1[t] + m*n2[t])*exp(-z) +
+    # ((rfunc(w1[t]*x1[t] + (1-w1[t])*x2[t],theta1) + rand(pdist))*((1-m)*n1[t]+m*n2[t])) * 
+    # exp(-beta*((1-m)*n1[t]+m*n2[t]));
+    # 
+    # n2[t+1] = ((1-m)*n2[t] + m*n1[t])*exp(-z) +
+    # ((rfunc(w2[t]*x2[t] + (1-w2[t])*x1[t],(theta1 + thetadiff)) + rand(pdist))*((1-m)*n2[t]+m*n1[t])) * 
+    # exp(-beta*((1-m)*n2[t]+m*n1[t]));
+    
+    n1[t+1] = ((rfunc(w1[t]*x1[t] + (1-w1[t])*x2[t],theta1) + rand(pdist))*((1-m)*n1[t]+m*n2[t])) * 
     exp(-beta*((1-m)*n1[t]+m*n2[t]));
     
-    n2[t+1] = ((1-m)*n2[t] + m*n1[t])*exp(-z) +
-    ((rfunc(w2[t]*x2[t] + (1-w2[t])*x1[t],(theta1 + thetadiff)) + rand(pdist))*((1-m)*n2[t]+m*n1[t])) * 
+    n2[t+1] = ((rfunc(w2[t]*x2[t] + (1-w2[t])*x1[t],(theta1 + thetadiff)) + rand(pdist))*((1-m)*n2[t]+m*n1[t])) * 
     exp(-beta*((1-m)*n2[t]+m*n1[t]));
     
 

### 2017-11-29 model/src/KevinEvolveExtinct.jl

@@ -85,12 +85,18 @@ function KevinEvolveExtinct(
     # (rfunc(x1[t],(theta1 + thetadiff)) + rand(pdist))*m*n1[t]) * 
     # exp(-beta*((1-m)*n2[t]+m*n1[t]));
     
-    n1[t+1] = ((1-m)*n1[t] + m*n2[t])*exp(-z) +
-    ((rfunc(w1[t]*x1[t] + (1-w1[t])*x2[t],theta1) + rand(pdist))*((1-m)*n1[t]+m*n2[t])) * 
+    # n1[t+1] = ((1-m)*n1[t] + m*n2[t])*exp(-z) +
+    # ((rfunc(w1[t]*x1[t] + (1-w1[t])*x2[t],theta1) + rand(pdist))*((1-m)*n1[t]+m*n2[t])) * 
+    # exp(-beta*((1-m)*n1[t]+m*n2[t]));
+    # 
+    # n2[t+1] = ((1-m)*n2[t] + m*n1[t])*exp(-z) +
+    # ((rfunc(w2[t]*x2[t] + (1-w2[t])*x1[t],(theta1 + thetadiff)) + rand(pdist))*((1-m)*n2[t]+m*n1[t])) * 
+    # exp(-beta*((1-m)*n2[t]+m*n1[t]));
+    
+    n1[t+1] = ((rfunc(w1[t]*x1[t] + (1-w1[t])*x2[t],theta1) + rand(pdist))*((1-m)*n1[t]+m*n2[t])) * 
     exp(-beta*((1-m)*n1[t]+m*n2[t]));
     
-    n2[t+1] = ((1-m)*n2[t] + m*n1[t])*exp(-z) +
-    ((rfunc(w2[t]*x2[t] + (1-w2[t])*x1[t],(theta1 + thetadiff)) + rand(pdist))*((1-m)*n2[t]+m*n1[t])) * 
+    n2[t+1] = ((rfunc(w2[t]*x2[t] + (1-w2[t])*x1[t],(theta1 + thetadiff)) + rand(pdist))*((1-m)*n2[t]+m*n1[t])) * 
     exp(-beta*((1-m)*n2[t]+m*n1[t]));
     
 

### 2017-11-29 model/src/KevinEvolveExtinct_ddm.jl

@@ -118,12 +118,18 @@ function KevinEvolveExtinct_ddm(
     # (rfunc(x1[t],(theta1 + thetadiff)) + rand(pdist))*m*n1[t]) * 
     # exp(-beta*((1-m)*n2[t]+m*n1[t]));
     
-    n1[t+1] = ((1-m1[t])*n1[t] + m2[t]*n2[t])*exp(-z) +
-    ((rfunc(w1[t]*x1[t] + (1-w1[t])*x2[t],theta1) + rand(pdist))*((1-m1[t])*n1[t]+m2[t]*n2[t])) * 
+    # n1[t+1] = ((1-m1[t])*n1[t] + m2[t]*n2[t])*exp(-z) +
+    # ((rfunc(w1[t]*x1[t] + (1-w1[t])*x2[t],theta1) + rand(pdist))*((1-m1[t])*n1[t]+m2[t]*n2[t])) * 
+    # exp(-beta*((1-m1[t])*n1[t]+m2[t]*n2[t]));
+    # 
+    # n2[t+1] = ((1-m2[t])*n2[t] + m1[t]*n1[t])*exp(-z) +
+    # ((rfunc(w2[t]*x2[t] + (1-w2[t])*x1[t],(theta1 + thetadiff)) + rand(pdist))*((1-m2[t])*n2[t]+m1[t]*n1[t])) * 
+    # exp(-beta*((1-m2[t])*n2[t]+m1[t]*n1[t]));
+    
+    n1[t+1] = ((rfunc(w1[t]*x1[t] + (1-w1[t])*x2[t],theta1) + rand(pdist))*((1-m1[t])*n1[t]+m2[t]*n2[t])) * 
     exp(-beta*((1-m1[t])*n1[t]+m2[t]*n2[t]));
     
-    n2[t+1] = ((1-m2[t])*n2[t] + m1[t]*n1[t])*exp(-z) +
-    ((rfunc(w2[t]*x2[t] + (1-w2[t])*x1[t],(theta1 + thetadiff)) + rand(pdist))*((1-m2[t])*n2[t]+m1[t]*n1[t])) * 
+    n2[t+1] = ((rfunc(w2[t]*x2[t] + (1-w2[t])*x1[t],(theta1 + thetadiff)) + rand(pdist))*((1-m2[t])*n2[t]+m1[t]*n1[t])) * 
     exp(-beta*((1-m2[t])*n2[t]+m1[t]*n1[t]));
     
 

### 2017-11-29 model/src/KevinEvolve_asym.jl

@@ -17,8 +17,10 @@ function KevinEvolve_asym(
   rv1 = 0.0;
   rv2 = 0.0;
   if asym > 0.0
-      rv1 = abs(rand(Normal(0,asym)));
-      rv2 = -abs(rand(Normal(0,asym)));
+      # rv1 = abs(rand(Normal(0,asym)));
+      # rv2 = -abs(rand(Normal(0,asym)));
+      rv1 = asym;
+      rv2 = -asym;
   end
   rmax1 = rmax*(1+rv1);
   rmax2 = rmax1*(1+rv2);
@@ -68,12 +70,18 @@ function KevinEvolve_asym(
       # (rfunc(x1[t],(theta1 + thetadiff)) + rand(pdist))*m*n1[t]) * 
       # exp(-beta*((1-m)*n2[t]+m*n1[t]));
       
-      n1[t+1] = ((1-m)*n1[t] + m*n2[t])*exp(-z) +
-      ((rfunc(w1[t]*x1[t] + (1-w1[t])*x2[t],theta1,rmax1) + rand(pdist))*((1-m)*n1[t]+m*n2[t])) * 
+      # n1[t+1] = ((1-m)*n1[t] + m*n2[t])*exp(-z) +
+      # ((rfunc(w1[t]*x1[t] + (1-w1[t])*x2[t],theta1,rmax1) + rand(pdist))*((1-m)*n1[t]+m*n2[t])) * 
+      # exp(-beta1*((1-m)*n1[t]+m*n2[t]));
+      # 
+      # n2[t+1] = ((1-m)*n2[t] + m*n1[t])*exp(-z) +
+      # ((rfunc(w2[t]*x2[t] + (1-w2[t])*x1[t],(theta1 + thetadiff),rmax2) + rand(pdist))*((1-m)*n2[t]+m*n1[t])) * 
+      # exp(-beta2*((1-m)*n2[t]+m*n1[t]));
+      
+      n1[t+1] = ((rfunc(w1[t]*x1[t] + (1-w1[t])*x2[t],theta1,rmax1) + rand(pdist))*((1-m)*n1[t]+m*n2[t])) * 
       exp(-beta1*((1-m)*n1[t]+m*n2[t]));
       
-      n2[t+1] = ((1-m)*n2[t] + m*n1[t])*exp(-z) +
[…]

### 2017-11-29 model/src/KevinEvolve_ddm.jl

@@ -88,12 +88,18 @@ function KevinEvolve_ddm(
     # (rfunc(x1[t],(theta1 + thetadiff)) + rand(pdist))*m*n1[t]) * 
     # exp(-beta*((1-m)*n2[t]+m*n1[t]));
     
-    n1[t+1] = ((1-m1[t])*n1[t] + m2[t]*n2[t])*exp(-z) +
-    ((rfunc(w1[t]*x1[t] + (1-w1[t])*x2[t],theta1) + rand(pdist))*((1-m1[t])*n1[t]+m2[t]*n2[t])) * 
+    # n1[t+1] = ((1-m1[t])*n1[t] + m2[t]*n2[t])*exp(-z) +
+    # ((rfunc(w1[t]*x1[t] + (1-w1[t])*x2[t],theta1) + rand(pdist))*((1-m1[t])*n1[t]+m2[t]*n2[t])) * 
+    # exp(-beta*((1-m1[t])*n1[t]+m2[t]*n2[t]));
+    # 
+    # n2[t+1] = ((1-m2[t])*n2[t] + m1[t]*n1[t])*exp(-z) +
+    # ((rfunc(w2[t]*x2[t] + (1-w2[t])*x1[t],(theta1 + thetadiff)) + rand(pdist))*((1-m2[t])*n2[t]+m1[t]*n1[t])) * 
+    # exp(-beta*((1-m2[t])*n2[t]+m1[t]*n1[t]));
+    
+    n1[t+1] = ((rfunc(w1[t]*x1[t] + (1-w1[t])*x2[t],theta1) + rand(pdist))*((1-m1[t])*n1[t]+m2[t]*n2[t])) * 
     exp(-beta*((1-m1[t])*n1[t]+m2[t]*n2[t]));
     
-    n2[t+1] = ((1-m2[t])*n2[t] + m1[t]*n1[t])*exp(-z) +
-    ((rfunc(w2[t]*x2[t] + (1-w2[t])*x1[t],(theta1 + thetadiff)) + rand(pdist))*((1-m2[t])*n2[t]+m1[t]*n1[t])) * 
+    n2[t+1] = ((rfunc(w2[t]*x2[t] + (1-w2[t])*x1[t],(theta1 + thetadiff)) + rand(pdist))*((1-m2[t])*n2[t]+m1[t]*n1[t])) * 
     exp(-beta*((1-m2[t])*n2[t]+m1[t]*n1[t]));
     
 

### 2017-11-29 model/src/KevinEvolve_mtheta_ddm.jl

@@ -91,14 +91,19 @@ function KevinEvolve_mtheta_ddm(
       # (rfunc(x1[t],(theta1 + thetadiff)) + rand(pdist))*m*n1[t]) * 
       # exp(-beta*((1-m)*n2[t]+m*n1[t]));
       
-      n1[t+1] = ((1-m1[t])*n1[t] + m2[t]*n2[t])*exp(-z) +
-      ((rfunc(w1[t]*x1[t] + (1-w1[t])*x2[t],theta1) + rand(pdist))*((1-m1[t])*n1[t]+m2[t]*n2[t])) * 
+      # n1[t+1] = ((1-m1[t])*n1[t] + m2[t]*n2[t])*exp(-z) +
+      # ((rfunc(w1[t]*x1[t] + (1-w1[t])*x2[t],theta1) + rand(pdist))*((1-m1[t])*n1[t]+m2[t]*n2[t])) * 
+      # exp(-beta*((1-m1[t])*n1[t]+m2[t]*n2[t]));
+      # 
+      # n2[t+1] = ((1-m2[t])*n2[t] + m1[t]*n1[t])*exp(-z) +
+      # ((rfunc(w2[t]*x2[t] + (1-w2[t])*x1[t],(theta1 + thetadiff)) + rand(pdist))*((1-m2[t])*n2[t]+m1[t]*n1[t])) * 
+      # exp(-beta*((1-m2[t])*n2[t]+m1[t]*n1[t]));
+      
+      n1[t+1] = ((rfunc(w1[t]*x1[t] + (1-w1[t])*x2[t],theta1) + rand(pdist))*((1-m1[t])*n1[t]+m2[t]*n2[t])) * 
       exp(-beta*((1-m1[t])*n1[t]+m2[t]*n2[t]));
       
-      n2[t+1] = ((1-m2[t])*n2[t] + m1[t]*n1[t])*exp(-z) +
-      ((rfunc(w2[t]*x2[t] + (1-w2[t])*x1[t],(theta1 + thetadiff)) + rand(pdist))*((1-m2[t])*n2[t]+m1[t]*n1[t])) * 
+      n2[t+1] = ((rfunc(w2[t]*x2[t] + (1-w2[t])*x1[t],(theta1 + thetadiff)) + rand(pdist))*((1-m2[t])*n2[t]+m1[t]*n1[t])) * 
       exp(-beta*((1-m2[t])*n2[t]+m1[t]*n1[t]));
-      
[…]

### 2017-12-19 Manuscript/FinalDraft3/msv8.tex
final draft submission to PhilTrans

--- 2017-11-29 manuscript/FinalDraft3/msv7.tex
@@
 \author{
-Justin D. Yeakel${}^{1,2,*}$, Jean P. Gibert${}^{1}$, Peter A. H. Westley${}^{3}$, \& Jonathan W. Moore${}^{4}$ \\
+Justin D. Yeakel${}^{1,2,*}$, Jean P. Gibert${}^{1}$, Thilo Gross${}^3$, Peter A. H. Westley${}^{4}$, \& Jonathan W. Moore${}^{5}$ \\
 ${}^1$School of Natural Sciences, University of California, Merced, Merced CA, USA \\
 ${}^2$The Santa Fe Institute, Santa Fe NM, USA \\
-${}^3$College of Fisheries and Ocean Sciences, University of Alaska, Fairbanks, Fairbanks AK, USA \\
-${}^4$Earth${}_2$Oceans Research Group, Simon Fraser University, Burnaby BC, Canada \\
+${}^3$University of Bristol, Bristol, UK\\
+${}^4$College of Fisheries and Ocean Sciences, University of Alaska, Fairbanks, Fairbanks AK, USA \\
+${}^5$Earth${}_2$Oceans Research Group, Simon Fraser University, Burnaby BC, Canada \\
 ${}^*$To whom correspondence should be addressed: jdyeakel@gmail.com
@@
 \noindent{\bf (a) Metapopulation framework}\\
-\noindent We considered two populations $N_1$ and $N_2$ that inhabit two distinct habitats, each with trait values $x_1$ and $x_2$ determining recruitment rates.
-We assumed that there is an optimum trait value $\theta_1$ and $\theta_2$ associated with each habitat, where recruitment is maximized if the trait value of the local population equals the optimum, such that $x = \theta$.
-Moreover, we assumed that $x_{1,2}$ are normally distributed with means $\mu_1$ and $\mu_2$ and have the same standard deviation $\sigma$.
-As such, the recruitment rate $R_1[\mu_1(t),\theta_1]$ for both populations is determined by the mean trait value of the local population relative to optimal value at that site.
-Trait means for each population are subject to selection, the strength of which is proportional to the difference between the trait mean and the local trait optimum at a given point in time \citep{simpson1953major,Lande:1976ga}.
+\noindent We follow the basic framework provided by Ronce and Kirkpatrick \cite{Ronce:2001dp}, where dispersal connects two populations $N_i$ and $N_j$ that inhabit two distinct habitats, each with trait values $x_i$ and $x_j$ determining phenotype.
+In our version of the model, there is an optimum trait value $\theta_i$ and $\theta_j$ associated with each habitat, where recruitment is maximized if the trait value of the local population equals the optimum, such that $x = \theta$.
+Moreover, we assumed that $x_{i,j}$ are normally distributed with means $\mu_i$ and $\mu_j$ and have the same standard deviation $\sigma$.
+As such, the recruitment rate $R_i[\mu_i(t),\theta_i]$ for both populations is determined by the mean trait value of the local population relative to optimal value at that site.
+Trait means for each population are subject to selection, the strength of which is proportional to the difference between the trait mean and the local trait optimum at a given point in time \citep{simpson1953major,Lande:1976ga,Ronce:2001dp}.
+% This framework mirrors the Ronce and Kirkpatrick model, except that selection is assumed to act on the rate of recruitment, rather than mortality.
 
@@
 Again, straying can be thought of as both from the donor or recipient population perspective.
-The two populations occur in spatially separate sites that are close enough that a proportion of the population $m$ can stray into the other site, and where mortality occurs before individuals return to reproduce.
-If there is no straying between these populations, then the mean trait evolves towards the optimal value for each site $\mu_1 \rightarrow \theta_1$, and the recruitment rate for each population is maximized.
+The two populations occur in spatially separate sites that are close enough that a proportion of the population $m$ strays into the other site.
+If there is no straying between these populations, then the mean trait evolves towards the optimal value for each site $\mu_i \rightarrow \theta_i$, and the recruitment rate for each population is maximized.
 If there is straying between populations at rate $m$, then the traits in each respective location will be pulled away from the optimum, and recruitment rates will be lowered.
-As $m \rightarrow 0.5$, the populations are perfectly mixed, acting as a single population.
-
-We used the discrete Ricker framework described by Shelton and Mangel \citep{Shelton:2011eq} as the basis for our two-site metapopulation model, with the added effect of the local population $N_i$ mixing with a set proportion $m$ of a remote population $N_j$ that is straying into it.
+As $m \rightarrow 0.5$, the populations are perfectly mixed, effectively acting as a single panmictic population.
+
+We used the discrete Ricker framework described by Shelton and Mangel \citep{Shelton:2011eq} as the basis for our two-site metapopulation model, with the added effect of the local population $N_i$ mixing with a proportion $m$ of individuals from a remote population $N_j$, and where there is no demographic overlap between generations.
 In this sense, both populations serve simultaneously as donor and recipient populations.
-We first assumed that a proportion ${\rm e}^{-Z}$ of both populations survive such that the surviving aggregated population, composed of both local individuals (at site $i$) and incoming strays (from site $j$), is $\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z}$.
-Because local individuals will recruit differently than incoming strays, the recruitment of the aggregate must incorporate two recruitment functions, given by $R_i[\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t)]\left((1-m)N_i + mN_j\right)$ (Eq.
+Because total recruitment will be determined by both locals (with a mean trait value closer to the site optimum) and strays (with mean trait values farther from the local optimum), the recruitment of the aggregate is given by $R_i[\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\theta_i]$ (Eq.
 \ref{eq:R}).
 This mix of individuals is subject to identical compensatory effects, which is determined by the parameter $\beta$.
-Taken together, the difference equation that determines changes in population size is
+Taken together, the difference equation that determines changes in population size from time $t$ to $t+1$ is
 
@@
 
-\noindent where it is assumed that $Z>0$ and that
+\noindent where
 
@@
 
-The recruitment of local individuals $(1-m)N_i(t)$ as a function of their mean trait value at time $t$ and the local trait optimum, is
+The recruitment of the population as a function of the mean trait value at time $t$ and the local trait optimum is
 
@@
 We make two simplifying assumptions.
-First, we approximate the distribution resulting from the mix of remote and local individuals following reproduction as a Gaussian where $X_i=x_i$ with probability $g(x_i)$.
+First, we approximate the distribution resulting from the mix of remote and local individuals prior to reproduction as a Gaussian, where $X_i=x_i$ with probability $g(x_i)$.
 The expectation of the actual trait distribution as well as the Gaussian approximation are the same, such that ${\rm E}\{g(x_i)\} = \omega_i\mu_i + (1-\omega_i)\mu_j$, with weights corresponding to the proportion of the mixed population that are local individuals, $\omega_i$, and straying individuals, $1-\omega_i$.
-We show that the Gaussian approximation is not only exact with respect to the mean, but also is a good approximation of variance in most relevant cases (see Appendix I in the Supplemental Materials, Fig.
-\ref{fig:PDF}).
 Thus, strays can successfully reproduce and introduce their genotypes into the recipient population, which is supported by observations in wild populations \citep{Jasper:2013cc}.
-Secondly, we assumed that changes in trait variance through time are minimal, such that $\sigma^2$ is constant over time, which is a common simplification in eco-evolutionary models of population dynamics \citep{Lande:1976ga,Schreiber:2011wx,Gilbert:2014ee,Gibert:2015kc}.
+Second, we assumed that changes in trait variance through time are minimal, such that $\sigma^2$ is constant over time, which is a common simplification in eco-evolutionary models of population dynamics \citep{Lande:1976ga,Ronce:2001dp,Schreiber:2011wx,Gilbert:2014ee,Gibert:2015kc}.
+These simplifications are the same as those introduced by Ronce and Kirkpatrick, and were shown to have negligible impacts on dynamics \cite{Ronce:2001dp}.
 
@@
 given $\mu^\prime = \omega_i \mu_i(t)+ (1-\omega_i)\mu_j(t)$.
-This model formulation parallels that proposed by Ronce and Kirkpatrick \citep{Ronce:2001dp}, where habitat specialization evolves between two populations as a function of dispersal.
+% This model formulation parallels that proposed by Ronce and Kirkpatrick \citep{Ronce:2001dp}, where habitat specialization evolves between two populations as a function of dispersal.
 \\
@@
 %Density-dependent m
-\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is mounting evidence that at least in some species (including salmon) the straying rate is density-dependent, a signature of collective navigation \citep{Berdahl:2014bl,Berdahl:2017uu}.
+\noindent We have so far assumed that the proportion of strays leaving and entering a population is constant, however there is mounting evidence that at least in some species (including salmon) straying is density-dependent, a signature of collective navigation \citep{Berdahl:2014bl,Berdahl:2017uu}.
 Specifically, the rate at which individuals stray has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tend to decrease the rate at which individuals err, reducing the overall proportion of a population that strays.
@@
 
-\noindent where $C$ is a half-saturation constant and is set to $C=1000$ throughout.
+\noindent where $C$ is a half-saturation constant, and determines to what extent collective behavior, as a function of group size, diminishes straying.
+If $C$ is small, relatively small groups of organisms `correct' for even high individual error rates, suppressing straying between sites.
+In the context of our model, small values of $C$ indicate that the effects of collective behavior on modifying straying are strong.
+As $C \rightarrow \infty$, groups of any size have no impact on straying, such that $m(t) \rightarrow m_0$, and the model reverts to that of constant straying.
+When $C \ll \infty$ and the population density is high, $m(t) \rightarrow 0$, whereas if the population is small, individuals operate without regard to the collective, such that $m(t) \rightarrow m_0$, such that in all cases, $m(t) < m_0$.
+% For realistic population densities, $m(t) < m_0$.\\
 % We note that at the limit $C\rightarrow \infty$, the density-dependent straying rate becomes constant such that $m(t) \rightarrow m_0$, and this corresponds to the original formulation where $m=m_0$.
-When the population density is very high, $m(t) \rightarrow 0$, and when the population is small, individuals operate without regard to collective behavior, such that $m(t) \rightarrow m_0$.
-For realistic population densities, $m(t) < m_0$.\\
-
-\noindent {\bf (c) Habitat heterogeneity}
-\noindent Increasing differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater regional differences in the conditions that favor alternative trait complexes, which can be interpreted as increased habitat heterogeneity.
-If both populations are isolated, natural selection will direct the mean trait values of both populations towards their respective optima, such that $\mu_i(t) \rightarrow \theta_i$ as $t\rightarrow\infty$.
-%difference in optimal trait values between the local and remote populations, which we call . 
-%Linking m/m0 and thetadiff
-Habitat heterogeneity and the rate of straying are treated both independently, and as parameters that covary.
-In the latter instance, we examined a case where it is assumed that increased habitat heterogeneity correlates with lower straying rates, and vice versa (illustrated in figure \ref{fig:mthetarelation}).
-Two scenarios may lead to this correlation: 
-(\emph{i}) sites may be distributed over greater spatial distances, where habitat differences are assumed to be exaggerated and the likelihood of straying over greater distances is lower \citep{Candy:2000hu,JPE:JPE1383};
-(\emph{ii}) individuals may have behaviors promoting dispersal between habitats with structural or physiognomic similarities \citep{Peterson:2014gy}.
-In this case, the rate of straying would be greater between habitats with smaller differences in trait optima (lower $\Delta\theta$) and lesser between habitats with greater differences in trait optima (higher $\Delta\theta$).
-\\
-% Where indicated, we account for correlation between the straying rate and the difference between optimal trait values by assuming that $m$ (if the stray rate is constant) or $m_0$ (if the stray rate is density-dependent) is a function of the difference between optimal trait values between sites $\Delta\theta$.
-% The straying rate is maximized at $m_{\rm max} = 0.5$, assuming both sites are equally attractive to the respective populations.
-% Thus, we can integrate these two variables by setting $m,m_0 = (1/m_{\rm max} + \epsilon \Delta\theta)^{-1}$, where $\epsilon$ sets the sensitivity of a declining $m$ to increasing distance (greater values of $\Delta\theta$), which we set to $\epsilon=2$.\\
-
-\noindent {\bf (d) Measuring metapopulation robustness}
-\noindent We evaluated metapopulation robustness by measuring the average-CV portfolio effect (PE) \citep{Anderson:2014cx,Schindler:2015gf} as well as the recovery time, which is the time required for the system to return to a steady-state following an induced disturbance to one or both of the populations \citep{Ovaskainen:2002il}.
+
+% \noindent {\bf (c) Habitat heterogeneity}
+% \noindent Increasing differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater regional differences in the conditions that favor alternative trait complexes, which can be interpreted as increased habitat heterogeneity.
+% If both populations are isolated, natural selection will direct the mean trait values of both populations towards their respective optima, such that $\mu_i(t) \rightarrow \theta_i$ as $t\rightarrow\infty$.
+% Habitat heterogeneity and the rate of straying are treated both independently, and as parameters that covary.
+% In the latter instance, we examined a case where it is assumed that increased habitat heterogeneity correlates with lower straying rates, and vice versa (illustrated in figure \ref{fig:mthetarelation}).
+% Two scenarios may lead to this correlation: 
+% (\emph{i}) sites may be distributed over greater spatial distances, where habitat differences are assumed to be exaggerated and the likelihood of straying over greater distances is lower \citep{Candy:2000hu,JPE:JPE1383};
+% (\emph{ii}) individuals may have behaviors promoting dispersal between habitats with structural or physiognomic similarities \citep{Peterson:2014gy}.
+% In this case, the rate of straying would be greater between habitats with smaller differences in trait optima (lower $\Delta\theta$) and lesser between habitats with greater differences in trait optima (higher $\Delta\theta$).
+% \\
+
+\noindent {\bf (c) Measuring metapopulation robustness}
+\noindent We evaluated metapopulation robustness by measuring the average-CV portfolio effect (PE) \citep{Anderson:2014cx,Schindler:2015gf} as well as the recovery time, which is the time required for the system to return to a steady-state following an induced disturbance to one or both populations \citep{Ovaskainen:2002il}.
 Throughout, we refer to an increase in portfolio effects and/or reduction in recovery time as promoting metapopulation robustness.
@@
 Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation \citep{GuckHolmes}, because we aimed to 
-1) assess the effects of a large perturbation far from the steady-state, and 
+1) assess the effects of a large perturbation far from the stable state, and 
 2) estimate the time required for all transient effects to decay following this perturbation (including dampened oscillations), we used a simulation-based numerical procedure.
 Recovery time was calculated by initiating a disturbance at $t=t_d$, and monitoring $N_T(t_d+t)$ as $t\rightarrow T$, where $T$ is large. 
-The aggregate was deemed recovered at $t_r$, such that recovery time was calculated as $t_r-t_d$, and recovery at $t=t_r$ was measured as the initial $t$ where $N_T(t) < {\rm SD}\left( N_T^* \right)$ for $t\in(t_r,T)$, where $\rm{SD}(\cdot)$ is standard deviation (illustrated in figure \ref{fig:recovery}).
-% The system was considered `recovered' at time $t=t_r$ where $N_T(t_d+t_r) < \pm {\rm SD}[N_T(t=T)]$ for $t = (t_r,T)$, assuming SD is standard deviation and $T$ is large.
-
-Numerically estimating the time that it takes for a perturbed system to recover also permits a more detailed perspective of metapopulation robustness.
-For example, if populations settle to alternative stable states (one at low- and one at high-density), comparing recovery times after a disturbance applied to individual populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery.
-% (alternative stable states in our model requiring one population to be high-density and one low-density)
+The aggregate was deemed recovered at $t_r$, such that recovery time was calculated as $t_r-t_d$, and recovery at $t=t_r$ was determined by the initial $t$ where $N_T(t) < N_T^*\pm{\rm SD}\left( N_T^* \right)$ for $t\in(t_r,T)$, where $\rm{SD}(\cdot)$ is standard deviation (illustrated in figure \ref{fig:recovery}).
+
+Numerically estimating the time that it takes for a perturbed system to recover also permits a more nuanced perspective of metapopulation robustness.
+For example, if populations settle to alternative stable states, comparing recovery times after a disturbance applied to individual populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery.
 We measured recovery time following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} are equivalent if populations have the same density); (\emph{iii}) near-collapse of both populations where just 1.0\% of each survives.
@@
 \noindent{\bf (a) Nonlinear effects of straying on metapopulation robustness} \\
-\noindent Regardless of density dependence, straying lowers steady-state densities for both populations by (\emph{i}) the donor population losing locally-adapted individuals to the recipient population and (\emph{ii}) the introduction of maladapted individuals to the recipient population from the donor population (Fig.
+\noindent Regardless of density dependence, straying lowers steady state densities for both populations by (\emph{i}) the donor population losing locally-adapted individuals to the recipient population and (\emph{ii}) the introduction of maladapted individuals to the recipient population from the donor population (Fig.
 \ref{fig:traj}).
 This prediction is in accordance with observations from natural populations \citep{Bett:2017ha}. %, regardless of trait heritability and variance or habitat heterogeneity
-The decline in steady-state densities is not gradual: as straying increases, the system crosses a discrete cusp bifurcation (DCB) \citep{AleksandrovichKuznetsov:1995p2580} whereby the single steady-state for the metapopulation bifurcates into two basins of attraction: one at high biomass, and one at low biomass density (figure \ref{fig:traj}a, \ref{fig:PE}a).
-Mean trait values for both populations bifurcate similarly (figure \ref{fig:traj}b). 
-In discrete systems, the cusp bifurcation is defined by two fold bifurcations intersecting at a cusp point \citep{AleksandrovichKuznetsov:1995p2580}, and is observed when the real part of the dominant eigenvalue of the Jacobian matrix crosses the unit circle at +1 (figure \ref{fig:eigs}).
-Visually, the dynamics are similar to those observed at a pitchfork bifurcation in continuous systems, where a single steady-state gives rise to two alternative steady-states.
-
-Above the threshold straying rate defined by the DCB, there are two alternative eco-evolutionary states: the \emph{dominant state} population will have a higher density and greater degree of local adaptation (smaller trait offset from the local optimum), while the \emph{subordinate state} population will have lower density with maladapted trait values (larger trait offset from the local optimum).
-Hysteresis is observed to occur at this transition, such that the single steady-state regime cannot easily be recovered by reducing straying after the system attains alternative steady-states (figure \ref{fig:hysteresis}).
-These dynamics are also observed in the Ronce and Kirkpatrick model, where populations are described as transitioning between symmetric to asymmetric states \citep{Ronce:2001dp}.
-Whether a specific population goes to one state or the other in our model is random, due to a small amount of introduced variance in the initial conditions.
+The decline in steady-state densities is not gradual: as straying increases, the system crosses a pitchfork bifurcation (PB) \citep{AleksandrovichKuznetsov:1995p2580} whereby the single steady-state for the metapopulation bifurcates into two basins of attraction: one at high biomass, and one at low biomass density (figure \ref{fig:traj}a, \ref{fig:PE}a).
+Accordingly, if straying is high enough, the populations assume asymmetric densities, referred to by Ronce and Kirkpatrick as \emph{migrational meltdown} \citep{Ronce:2001dp}.
+Mean trait values for both populations bifurcate similarly (figure \ref{fig:traj}b), where the trait optimum for the dominant population skews both dominant and subordinate trait means. 
+If straying is increased enough, the alternative fixed points revert to a single fixed point, and the populations assume symmetric, low biomass densities.
+Notably, if straying is subsequently reduced \emph{after} it has increased (such that hysteresis effects are observed), the system passes the PB without change to the dynamics until it crosses a second bifurcation (a fold bifurcation; FB) at lower values of $m$ (figure \ref{fig:traj}).
+Thus, the eco-evolutionary model is characterized by two separate dynamic regimes that can lead to alternative stable states:
+regime I, where two (asymmetric) alternative stable states (dominant and subordinate) are separated by an unstable steady state, and
+regime II, where the default state is a symmetric stable state at intermediate biomass densities, but that harbours a high- and low-density basin of attraction.
+% which can be encountered if straying is first increased past the PB and then decreased (i.e. hysteresis), or if a larger disturbance is applied to one of the populations.
+% 
+% two (asymmetric) alternative stable states (a dominant state population at high biomass density, and a subordinate state population at low biomass density) separated by an intermediate-density basin of attraction.
+% In the latter case, dominant and subordinate densities are only encountered 
+% 
+% that can be encountered as a function of the straying ratio, which are likewise observed in the Ronce and Kirkpatrick model:
+% 1) a symmetric high biomass steady state (low values of $m$),
+% 2) two (asymmetric) alternative stable states (a dominant state population at high biomass density, and a subordinate state population at low biomass density) separated by a stable steady state (low/intermediate values of $m$),
+% 3) two (asymmetric) alternative stable states (dominant and subordinate) separated by an unstable steady state (intermediate values of $m$),
+% 4) a symmetric low biomass steady state (high values of $m$).
+% % As we shall see, all four regimes have important ramifications for system robustness.
 
 Trait heritability $h^2$ has a large effect on the degree to which straying affects both the aggregate population steady-state density ($N^*_T$; figure \ref{fig:PE}a) as well as the difference between steady-state densities (the distance between alternative stable states: $\Delta N^*=|N^*_1-N^*_2|$; figure \ref{fig:PE}b).
-Greater trait heritability results in a steeper decline in $N_T^*$ with increasing straying rates $m$, but leads to only moderate changes to $\Delta N^*$.
-Lower trait heritability has little effect on the total biomass density but contrastingly large effects on $\Delta N^*$.
-The cusp bifurcation (the black line in Figs.
-\ref{fig:PE}a-c) occurs at lower values of the straying rate $m$ with lower heritability (Fig \ref{fig:PE}a,b), indicating that weaker coupling between ecological and evolutionary dynamics in addition to higher rates of straying promotes the appearance of alternative stable states.
-Although trait heritability among salmonids is variable, most life-history traits have an $h^2 <0.5$ \citep{Carlson:2008hl}, and we largely focus our efforts on that range.
+Increased trait heritability is observed to prevent the development of the alternative stable state regime, thereby maintaining greater densities in $N_T^*$, and a relatively uniform PE across increased straying ratios.
+Lower trait heritability ($h^2<0.8$) introduces alternative stable state regimes, and the pitchfork bifurcation that marks the onset of this regime occurs for lower straying rates for traits with lower $h^2$ (Fig \ref{fig:PE}a,b).
+This indicates that weaker coupling between ecological and evolutionary dynamics in addition to increased straying promotes the appearance of alternative stable states.
+Although trait heritability among salmonids is variable, most life-history traits have an $h^2 < 0.5$ \citep{Carlson:2008hl}, and for all additional analyses we have set $h^2=0.2$.
 
 %Critical slowing down
-As the cusp bifurcation is approached with increasing $m$, the portfolio effect increases sharply due to an amplification in variance within both donor and recipient populations.
-This amplification in variance is the product of a dynamical process known as \emph{critical slowing down} that can occur near bifurcations \citep{Scheffer:2009gg}, a phenomenon that some have suggested may serve as an early warning indicator for approaching phase transitions \citep{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br,Krkosek:2014ch}.
-For larger values of $m$ (to the right of the cusp bifurcation in Fig \ref{fig:PE}a-c), where alternative stable states occur, the portfolio effect declines steadily as the CV of $N_T^*$ increases.
-The decline over $m$ is more gradual if trait heritability is low, and steeper if trait heritability is high (figure \ref{fig:PE}c).
+As the alternative stable state regime is approached with increasing $m$, the portfolio effect increases sharply due to an amplification in variance within both populations.
+This amplification in variance is the product of \emph{critical slowing down}, which occurs near some bifurcations \citep{Scheffer:2009gg} and has been suggested to serve as an early warning indicator for approaching phase transitions \citep{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br,Krkosek:2014ch}.
+For larger values of $m$ (to the right of the PB in Fig \ref{fig:PE}a-c), where alternative stable states occur, the portfolio effect declines steadily as the CV of $N_T^*$ decreases.
+The decline over $m$ is more gradual if trait heritability is low, and steeper if trait heritability is high, suggesting that the benefits provided by a higher PE are greater for traits that have a low $h^2$ and when the straying ratio is near the pitchfork bifurcation in regime I.
 
@@
 % Return time as a measure of system persistence and relation to PE
-As the portfolio effect is highly sensitive to the rate of straying between populations, so is the time required for the system to recover to a steady-state following a large disturbance.
-In general, we find that the average-CV portfolio effect is negatively correlated with recovery time (figure \ref{fig:PE}d), indicating that, for our system, both measures are valuable indicators of metapopulation robustness.
-Because we can assess the time to recovery in response to the various disturbance types described above, this allows us to gain an in-depth perspective into the fragility of the metapopulation as a function of straying rate.
-
-Straying had non-linear impacts on the recovery time of populations. 
-When the dominant state population (well adapted and high density) goes extinct, high rates of straying allow recolonization of the extirpated habitat and quick recovery (figure \ref{fig:relax}a) because the surviving population has a mean trait value skewed towards the optimum of the affected habitat (figure \ref{fig:relaxtraj_hdlh}).
-Yet, as straying decreases, recovery time for the disturbed dominant state population increases, in part because there is time for the trait distribution to move back towards the trait optimum of the subordinate state population.
-In contrast, when the subordinate state population (maladapted and low density) is wiped out, recovery rates are fastest at low to intermediate levels of straying.
-Because the mean trait values of both populations are skewed towards those of the dominant population, when the subordinate population collapses under high rates of straying, selection against the flood of maladapted individuals that stray into the recovering population extends the length of time required for it to return to its steady-state (figure \ref{fig:relaxtraj_ldlh}).
-When both populations are both dramatically reduced, recovery time is generally fastest at lower levels of straying due to reduced mixing of maladaptive phenotypes.
-Near the onset of the cusp bifurcation, recovery time increases explosively, however this is -- as the name implies -- characteristic of \emph{slow} dynamics occurring near critical transitions \citep{Scheffer:2009gg,Kuehn:2010p2591}.
-
-%Lowering Phenotypic diversity with increasing straying rate
-Increased rates of straying lowers phenotypic diversity ($\Delta \mu^* = |\mu_i^*-\mu_j^*|$, evaluated at the steady-state) because both local and remote populations are increasingly homogenized.
-The loss of phenotypic diversity is greater with increased straying when trait heritability is low because traits take longer to go back to their local optima than they do when heritability is high. 
-Hence straying counters the effect of diversifying local adaptation. 
-% The loss of phenotypic diversity with increased straying is greater if trait heritability is low because the selective forces acting against trait means far from the local optima are lessened.
-% with decreased $h^2$.
-Less intuitively, we observe a discrete jump towards low phenotypic diversity as the cusp bifurcation is crossed (figure \ref{fig:traitdiff}).
-Although the development of alternative stable states elevates the portfolio effect due to the variance-dampening effects of the aggregate, entering this dynamic regime also results in a substantial decline in phenotypic diversity, which may have less predictable adverse effects on the population.\\
+As the portfolio effect is highly sensitive to straying between populations, so is the time required for the system to recover to a steady state following a large disturbance.
+We find that the average-CV portfolio effect is negatively correlated with recovery time (figure \ref{fig:PE}d), indicating that both measures are valuable indicators of metapopulation robustness, and where recovery time is measured for \emph{i}) near-collapse of both populations, \emph{ii}) extinction of the dominant population, and \emph{iii}) extinction of the subordinate population.
+%NOTE RECOVERY TIME DESCRIPTION (SHORTER!)
+% Accordingly, we find that if straying is high enough such that there are alternative stable states, recovery time is generally suppressed.
+% Accordingly, changes in PE as a function of straying are mirrored by changes in the time required for recovery following near-collapse of both populations or the complete extinction of either.
+We find that if straying is low, recovery following near-collapse requires the most time (figure \ref{fig:relax}a).
+However, as straying increases and the system enters the alternative stable state regime, the extinction of the dominant population has a recovery time on-par with recovery following near-collapse, due to greater straying from a population hosting a mean trait value farther from the optimum of the recovering population.
+% Although near-collapse of both populations requires the longest time to recovery if straying is very low or very high, extinction of the dominant population requires the longest time to recovery if straying is just large enough to produce alternative stable states (figure \ref{fig:relax}a).
+If straying continues to increase, the time required to recover from any of the disturbance scenarios grows due to the increasingly negative effects of demographic mixing on the rate of recruitment.
+Near the onset of the pitchfork bifurcation, recovery time increases explosively, however this is -- as the name implies -- characteristic of slow dynamics occurring near critical transitions \citep{Scheffer:2009gg,Kuehn:2010p2591}.
+Given the small parameter range where this effect occurs, we do not consider slow dynamics very close to the critical transition to be ecologically significant.
 
@@
 %Portfolio effects
-If we assume that the rate of straying is density-dependent, the probability that an individual strays $m_0$ determines the rate of straying within the population, such that $m(t)$ becomes lower as $N(t)$ increases, likely due to the effects of collective decision-making \citep{Berdahl:2016dx} (Eq.
-\ref{eq:ddm}).
+If we assume that the rate of straying is density-dependent, the probability that an individual errs $m_0$ in part determines the magnitude of straying within the population (Eq.
+\ref{eq:ddm}), such that $m(t)$ becomes lower as $N(t)$ increases, which we assume here to be the consequence of collective decision-making behaviors \citep{Berdahl:2016dx}.
+% Organisms that employ collection decision-making to aid in navigation have reduced straying rates.
+The parameter $C$ in Eq.
+\ref{eq:ddm} determines to what extent straying is reduced as group size increases, where low values correspond to the effects of strong collective behavior, and high values correspond to the effects of weak collective behavior \cite{Berdahl:2016dx}.
+To enable comparisons between models with constant and density-dependent straying, we examine constant $m$ with respect to both the individual straying probability $m_0$ as well as the value of straying observed at the steady state $m^*$.
+In the alternative stable state regime, the dispersing populations are characterized by alternative $m^*$ values, corresponding to dominant and subordinate densities.
+% , such that small increases in group size have a large effect on straying, and where $C \rightarrow \infty$ corresponds to constant straying independent of group size .
 %Because our analyses concern steady-state conditions, the inclusion of density-dependent straying does not have a large impact on the qualitative nature of our findings (figure \ref{fig:ddm}).
-Density dependence alters the straying rate at steady-state population densities because $0 < m^* < m_0$, and this serves to rescale both the strength of the PE as well as the recovery time, but does not change the qualitative nature of our findings.
-In the alternative stable state regime, because each population exists at different steady-state densities, there are likewise two alternative straying rates $(m_i^*,m_j^*)$: the higher straying rate is associated with the low-density population, and the lower straying rate is associated with the high-density population.
-We assessed metapopulation robustness across a range of $(m_i^*,m_j^*)$ values by varying the probability that an individual strays $m_0$, which is positively and linearly related to $(m_i^*,m_j^*)$.
-We find that the portfolio effects generated in systems with density-dependent straying are qualitatively similar to systems with constant straying, however there are some important quantitative differences.
-First, the PE associated with the high-density (low $m^*$) population is the same as that for a system with a constant $m$ (figure \ref{fig:thetaPE}a).
-% As such, the difference in $m^*$ between subordinate and dominant populations appears to matter less than the lower straying rate generated by the dominant population.
-As $m^*$ increases, we observe an increase in the PE relative to systems with constant $m$.
-
-%Recovery times
-Density-dependent straying alters these recovery times (figure \ref{fig:relax}b; note difference in x-axis scale). 
-First, in comparison with constant stray rates, density-dependent straying lowered recovery times at elevated straying rates for near-collapse of both populations and extirpation of the subordinate population.
-%Recovery times
-%The recovery times for systems with constant and density-dependent straying both show similar trends for a given $m$ and $m^*$ (figure \ref{fig:relax}c,d), however there are some distinct quantitative differences.
-In contrast, at low straying rates, near-collapse of both populations required more time to recover.
-%whereas in the alternative stable state regime (higher $m^*$), the recovery times for different disturbance types were very similar to systems with a constant $m$ (figure \ref{fig:relax}b; note difference in x-axis scales).
-Trait heritability had a large effect on recovery times, with near-collapse requiring a more protracted time to recover in the alternative steady-state regime (figure \ref{fig:relax_highh}).
-% As trait heritability increased, the metapopulation always recovered more quickly if the subordinate population was lost (figure \ref{fig:relax_highh}).
-% There was greater similarity in the recovery times for constant and density-dependent $m$ systems when trait heritability was high (figure \ref{fig:relax_highh}).
-In general, the lower recovery time for systems with increased $m^*$ mirrors an elevated PE with higher density-dependent straying rates (figure \ref{fig:thetaPE}a).
-% This higher PE at increased $m^*$ mirrors recovery times that are also lower than those observed for systems with constant $m$ (figure \ref{fig:relax}).
-Together, analysis of both PE and recovery time suggests that although density-dependent straying does not appear to change the `dynamic landscape' in our model, it does appear to promote robustness, particularly in the case of near-collapse of both populations when straying is high.
-\\
+% Density dependence alters the straying ratio at steady-state population densities because $0 < m^* < m_0$, and this serves to rescale both the strength of the PE as well as the recovery time, but does not change the qualitative nature of our findings.
+
+When collective behavior is very strong, such that $C$ is very low, small increases in population density begit large reductions in straying.
+These reductions can be large enough such that the system avoids the alternative stable state regime altogether (figure \ref{fig:cb}, left inset).
+Conversely, when collective behavior is very weak, such that $C$ is very high, there is effectively no reduction in straying with increased group size, and the dynamics conform to those of the constant straying model (figure \ref{fig:cb}, right inset).
+However, when collective behavior is of intermediate strength ($10^{1.5} \lesssim C \lesssim 10^3$), the dynamics are altered in two important ways.
+First, in the alternative stable state regime, the low density subordinate population has correspondingly higher $m^*$, whereas the high density dominant population has correspondingly lower $m^*$ (figure \ref{fig:cb}, center inset).
+Second, the alternative stable state regime results in a $\Delta N^*$ that is reduced when individual error rates are low, and magnified when individual error rates are high (figure \ref{fig:cb}, main).
+In other words, when collective behavior is of intermediate strength, a higher individual rate of error exaggerates the differences between the steady state densities, effectively pushing the subordinate population closer to extinction.
+
+In both the constant- and dynamic straying model, there are two regimes that give rise to alternative stable states: one that is produced by a pitchfork bifurcation (regime I), and one that is produced by a fold bifurcation (regime II; figure \ref{fig:bifurcationsddm}).
+When the strength of collective behavior is intermediate, the cause of the exaggerated difference in steady state densities at high individual straying rates is due to an increase in the dominance of regime I over a greater range of $m_0$ (figure \ref{fig:bifurcationsddm}).
+If the effects of collective behavior are very strong, regime I vanishes, and the populations generally assume symmetric densities.
+However, while regime I vanishes, the alternative stable state regime created by the fold bifurcation expands (regime II; figure \ref{fig:bifurcationsddm}).
+In regime II there are two possible steady state configurations: 
+1) the steady states of both populations can either have symmetric densities, or 
+2) one or both populations can revert to asymmetric dominant/subordinate states as in regime I.
+In this case, populations will attain asymmetric densites if $m_0$ is first increased and subsequently decreased (introducing a hysteresis effect), or if a sufficiently large perturbation directs one of the populations towards the dominant or subordinate basins of attraction.
+
+%PE, Recovery times (near-collapse)
+Dynamic straying via collective behavior directly alters the dynamic regimes of the model, and this has an impact on measures of robustness such as the portfolio effect and recovery time.
+% When collective behavior is weak (large $C$), the alternative stable state regime occurs for lower values of $m_0$.
+% As collective behavior is strengthened (smaller $C$), Regime I first expands for larger values of $m_0$ and collapses for $C \lesssim 80$  (black region in figure \ref{fig:bifurcationsddm}).
+When the effects of collective behavior are weak (large $C$) the portfolio effects and recovery times conform to those examined in the case of constant straying.
+As the effects of collective behavior are increased to intermediate values (intermediate $C$), the expansion of regime I has contrasting effects on metapopulation robustness as measured by the portfolio effect and time to recovery following near-collapse of both populations (figure \ref{fig:pert}):
+\emph{i}) if individual straying is low, regime I is avoided;
+\emph{ii}) if individual straying is intermediate, the system enters regime I, resulting in an increase in robustness as observed by an elevated portfolio effect and suppressed recovery time;
+\emph{iii}) if individual straying is high, the time to recovery becomes extreme, such that robustness is eroded.
+When the effects of collective behavior are unrealistically strong (low $C$), the portfolio effect remains relatively constant and the recovery time increases with higher values of $m_0$, tracking a gradual decline in steady state densities.
+% , inhibiting the strength of natural selection.
+
+%Fold bifurcation effect (single extinctions)
+As is described above, the alternative stable state regime II created by the fold bifurcation expands with lower $C$ (figure \ref{fig:bifurcationsddm}).
+Here, the lone extinction of either population will push the system into a different configuration.
+Directly after extinction and as the site is being recolonized by the undisturbed population, instead of returning to its previously higher (symmetric) density that it shared with the undisturbed population, it becomes trapped in the basin of attraction defined by the subordinate state that is generally not observed barring the occurrence of hysteresis.
+The extinction of either population in regime II therefore has a shorter time to recovery (figure \ref{fig:pertlh}), but this is in part driven by the newly acquired subordinate steady state density.
+% , which requires less time to colonize because it is closer to extinction.
+Importantly, this low-density basins of attraction, which is particularly prevelant across $m_0$ when the effects of collective behavior are strong, would not be easy to anticipate or detect prior to a large disturbance.
+% Accordingly, if individual straying is low and the effects of collective behavior are strong, there exists low density basins of attraction that can trap disturbed populations.
 
 \noindent{\bf (c) The role of habitat heterogeneity and changing selective landscapes}\\
-\noindent With the onset of straying, we find that increasingly divergent trait optima generally lower $N_T^*$ and exaggerate $\Delta N^*$, and this is particularly pronounced for density-dependent straying (figure \ref{fig:thetadiffN}).
-%, such that the biomass distribution becomes increasingly uneven.
-%I've checked this - it's a theshold type (logistic) response Supp fig?
-The impact of habitat heterogeneity on the portfolio effect and recovery time is more complex, serving to emphasize the nonlinear relationship between rates of straying and metapopulation robustness.
-%regardless of heritability.
-%The rate of straying that gives rise to a large PE at the cusp bifurcation is conditioned on the difference in optimal trait values between the local and remote populations, which we call $\Delta\theta = \sqrt{(\theta_i - \theta_j)^2}$.
-As habitat heterogeneity increases, alternative steady-states appear at lower straying rates -- with the crossing of the cusp bifurcation, accompanied by a peak in the PE -- whereas the magnitude of increase in the PE also increases (figure \ref{fig:thetaPE}b), lowering recovery time (figure \ref{fig:relaxtheta}).
-For increased rates of straying, greater habitat heterogeneity erodes the PE (figure \ref{fig:thetaPE}b) and increases the recovery times (figure \ref{fig:relaxtheta}).
-In tandem, these results suggest that habitat heterogeneity promotes robustness when straying rates are low, and erodes robustness when straying rates are high.
-%This means that alternative stable states can emerge with relatively little dispersal between sites (figure \ref{fig:thetaPE}), however the increased PE (and shorter recovery times).
-
-Until now, we have treated the rate of straying and habitat heterogeneity as independent parameters, however they may also be assumed to covary.
-For instance, if sites are separated by greater distance, they may be assumed to have increased habitat heterogeneity as well as lower rates of straying.
-Alternatively, individuals may be genetically predisposed to stray into sites that are more similar, such that greater between-site heterogeneity will correspond to lower straying rates.
-% We next investigate the dynamics across different types of sites that host 
-% Sites that are farther apart are likely to be more heterogeneous, while simultaneously facilitating lower rates of straying between populations.
-% If we assume that environmental heterogeneity increases with distance between sites, or that individuals may have behaviors that minimize their chances of straying into very different habitats, the rate of straying will decline with habitat heterogeneity.
-% We can implement this dependence by allowing $\Delta\theta$ to decrease as $m$ increases, such that there are low rates of straying between dissimilar populations and high rates of straying between similar populations.
-We implemented this inverse relationship by setting $m = 0.5(1 + \Delta\theta)^{-1}$ where maximum straying is assumed to occur at $m=0.5$ (perfect mixing; figure \ref{fig:mthetarelation}).
-This assumes that $m$ is greater for lower $\Delta\theta$, such that there are low rates of straying between dissimilar (distant) sites and high straying rates between similar (close) sites.
-Under these conditions we find that alternative stable states appear for very low rates of straying (figure \ref{fig:mthetaPE}). % $0 < m \leq 0.43$.
-As the straying rate increases and $\Delta\theta$ decreases, a single stable state emerges as the cusp bifurcation is crossed, which is opposite the pattern observed when straying is independent of habitat heterogeneity.
-
-There are two notable dynamics that emerge following extinction of the dominant population at low rates of straying between dissimilar (high $\Delta\theta$) sites (figure \ref{fig:mtheta}).
-(\emph{i}) Above a threshold $m$ value, the dominant population recovers quickly enough that the evolving subordinate phenotype is overwhelmed by incoming strays, and it shifts back to its pre-disturbance subordinate state;
-(\emph{ii}) below a threshold $m$ value, there is an \emph{inversion} between subordinate and dominant states: because there is enough time and isolation for the subordinate trait mean to shift towards its local optimum, and away from that of the recovering dominant population, the dominant population becomes subordinate, and the subordinate population becomes dominant (figure \ref{fig:inertia}).
-This threshold value of $m$, below which the inversion dynamic behavior occurs, is marked by the asterisk in figure \ref{fig:mtheta}, and holds for both constant and density-dependent straying (figure \ref{fig:mthetamvm}).
-
-% At low rates of straying, there is little mixing between dissimilar populations, and this leads to the formation of alternative stable states.
-% Here we find that if the subordinate population is wiped out, the time to recovery is minimized, whereas if the dominant population is wiped out, the time to recovery is maximized; near-collapse of both results in intermediate recovery times.
-% When straying rates are low and the difference in trait optima are correspondingly high, we might assume that extinction of the dominant population would result in relatively faster recovery.
-% This would be a logical assumption because the mean trait value of the subordinate population would be skewed towards that of the dominant (larger) population, such that recruitment at the disturbed site should be high and the population should recovery more quickly.
-% Our results show, however, that \emph{this assumption is wrong:} although the mean trait value of the subordinate population \emph{is} skewed towards the optimum of the dominant population, when the latter is wiped out and the rate of straying is low, there is enough time and isolation for the subordinate trait mean to shift towards its own optimum, and away from that of the recovering dominant population (figure \ref{fig:inertia}).
-% This \emph{selective inertia} occurs until the dominant population grows large enough that the evolving subordinate phenotype is overwhelmed by incoming strays, shifting it back to its pre-disturbance (subordinate) state.
-% This effect can lead to much longer recovery times for the dominant population.
-% Importantly, if the rate of straying is below a threshold value, it can also result in a state-switch, wherein increased isolation permits the subordinate population to \emph{escape} the selective pull of the dominant population, leading to a switching in which population is in the dominant/subordinate state.
-
-% That alternative stable states appear and that PE becomes severely depressed for very low rates of straying is surprising: this means that even a small amount of mixing of populations from distant or dissimilar populations can qualitatively alter the dynamics of the metapopulation, regardless of trait heritability.
-%(for the discussion: an example of this situation may be the salmon populations during the last glacial maximum, where any mixing would be from geographically distant populations)
+As habitat heterogeneity ($\Delta\theta$) increases, even small amounts of straying can lead to the appearance of alternative stable states, and this is in direct accord with the findings of Ronce and Kirkpatrick \citep{Ronce:2001dp}.
+However, if straying is density dependent, the strength of collective behavior  has a large influence on the occurrence of alternative stable states defined by both regimes I and II.
+When habitat heterogeneity is low and the effects of collective behavior are weak (large $C$), regime I occurs for small-intermediate $m_0$, and regime II does not play a role (figure \ref{fig:hystheta}a).
+The absence of regime II implies that there are no hidden steady state configurations that might trap a disturbed population.
+As the strength of collective behavior increases, regime II appears at a cusp and becomes prevalent with increased $m_0$.
+
+When habitat heterogeneity is high, the proportion of $(m_0,~C)$-space that is dominated by alternative stable state regimes expands (figure \ref{fig:hystheta}b,c).
+Only when habitat heterogeneity is intermediate does regime II play a significant role as straying is relatively constant (figure \ref{fig:hystheta}b).
+For high $\Delta\theta$, there is no longer any sanctum from regimes I or II except for very high individual straying.
+In this case, regime I dominates when the strength of collective behavior is weak.
+When the strength of collective behavior is strong, regime II is dominant when individual straying is low, and regime I is dominant when individual straying is high.
+
+Until now, we have treated straying and habitat heterogeneity as independent parameters, however they may also be assumed to covary.
+For instance, if sites are separated by greater distance, they may be assumed to have increased habitat heterogeneity as well as less straying.
+Alternatively, individuals may be genetically predisposed to stray into sites that are more similar, such that higher straying can be assumed to occur between sites that are more homogeneous in aspect.
+We implemented this inverse relationship by setting $m_0 = 1/(2+\epsilon\Delta\theta)$ where $\epsilon$ controls the degree to which an increase in $\Delta\theta$ lowers $m_0$ (figure \ref{fig:mtheta}, inset).
+Accordingly, that $m_0$ is greater for lower $\Delta\theta$, such that there is lower straying between dissimilar sites and higher straying between similar sites.
+Under these conditions we find that regime I appears for very low $m_0$, and regime II appears for higher $m_0$ (figure \ref{fig:mtheta}). % $0 < m \leq 0.43$.
+As the straying rate increases and $\Delta\theta$ decreases, a single (symmetric) steady state emerges as the fold bifurcation is crossed, which is opposite the pattern observed when straying is independent of habitat heterogeneity.
+
+% There are two notable dynamics that emerge following extinction of the dominant population at low rates of straying between dissimilar (high $\Delta\theta$) sites (figure \ref{fig:mtheta}).
+% (\emph{i}) Above a threshold $m$ value, the dominant population recovers quickly enough that the evolving subordinate phenotype is overwhelmed by incoming strays, and it shifts back to its pre-disturbance subordinate state;
+% (\emph{ii}) below a threshold $m$ value, there is an \emph{inversion} between subordinate and dominant states: because there is enough time and isolation for the subordinate trait mean to shift towards its local optimum, and away from that of the recovering dominant population, the dominant population becomes subordinate, and the subordinate population becomes dominant (figure \ref{fig:inertia}).
+% This threshold value of $m$, below which the inversion dynamic behavior occurs, is marked by the asterisk in figure \ref{fig:mtheta}, and holds for both constant and density-dependent straying (figure \ref{fig:mthetamvm}).
 
@@
 
-% 
-% %The bifurcation and critical slowing down
-% [This paragraph is interesting to me but could be cut] Although the portfolio effect is negatively correlated with recovery time, there is one exception: the point of transition from a single steady-state to alternative stable states, which occurs at the pitchfork bifurcation (REF).
-% At this point, the portfolio effect is maximized, but so is the recovery time (figures \ref{fig:PE}, \ref{fig:relax}).
-% This suggests that while the sharply increasing variance of the individual populations is dampened by aggregation, a large disturbance will have a much greater adverse effect due to exponentially longer recovery times.
-% Whether the change in variance, also known as critical slowing down, as the alternative stable state regime is approached could be used as an early warning signal of an oncoming phase transition is a hotly debated topic (REFS).
-% The detectability of such changes in dynamical behavior among salmonids appears to be idiosyncratic across species, though the difficulty in measuring critical slowing down may - ironically - be masked by large portfolio effects \citep{Krkosek:2014ch}.
-%
-
-%alternative stable states & critical slowing
-% A salient finding from our results was that straying can lead to the emergence of alternative stable states, pushing one of the populations to high density (the \emph{dominant state}), and one to low density (the \emph{subordinate state}).
-% This asymmetry in population densities also leads to an asymmetry in the mean trait values, both of which are skewed towards the local optimum of the dominant population. 
-% The formation of alternative stable states in our system is a relatively common example of spatial pattern formation. 
-% Pattern formation can occur as a consequence of myriad ecological processes, including habitat selection, aggregation, local environmental conditions, and/or interspecific interactions such as competition or predation, and is inexorably linked to scale \citep{Levin:1992p1472}.
-% Pattern formation is also known to occur with the onset of spatially or diffusion-induced instabilities, or Turing instabilities \citep{Meinhardt:2012hw}. 
-% Here, we describe an additional mechanism by which pattern formation in local densities can occur, when the onset of collective movement-mediated maladaptation leads to local differences in reproductive rates resulting in alternative stable states across sites.
-% While metapopulation approaches have widely recognized the importance of source and sink populations \citep{Hanski:1999vu} as well as the occurrence of alternative states \citep{Boughton:1999fa}, here we show that the coupled forces of ecological and evolutionary processes can result in similar populations diverging into alternative states where one population is large and well-adapted, while the other is smaller and maladapted.
-% The appearance of these alternative states is largely dependent on the dispersal, or rate of straying, between sites.
-
-A salient finding from our results is that straying can lead to the emergence of alternative stable states, pushing one of the populations to high density (the \emph{dominant state}), and one to low density (the \emph{subordinate state}).
-This pattern has been observed in other models of eco-evolutionary dynamics with explicit space \citep{Ronce:2001dp}, suggesting that it may be a general feature of spatially-linked populations that evolve toward local optima while being hindered by dispersal.
-An important aspect of our framework is that there are similar forces that dictate interactions within and between sites, and this naturally results in a symmetry that could be perceived as limiting the relevance of our findings for natural (and inherently less symmetric) systems.
-Although the emergence of alternative stable states via a cusp bifurcation is characteristic of symmetrical systems, we find that increasing the asymmetry in the vital rates of populations across sites does not significantly alter the presence or position of the DCB (figure \ref{fig:symmetry}).
-That these patterns arise in alternative formulations and are relatively insensitive to parameter asymmetry suggests that the dynamical features observed have potentially widespread ramifications for the evolutionary dynamics of spatially connected populations in general.
-% suggests that the emergence of dominant and subordinate states directly from eco-evolutionary dynamics may be a generally relevant phenomenon in ecological systems.
-% Importantly, we show that the resulting dynamics are qualitatively similar, suggesting that the dynamical features present in both of these approaches have potentially widespread ramifications for the evolutionary dynamics of spatially connected populations, regardless of the specific formulation implemented.
-<%, though this requires additional investigation and scrutiny.
-
-% 
-% To what extent the existence of alternative stable states promotes or diminishes the system as a whole is a matter of perspective, though with clear conservation implications.
-% At face value, the subordinate population is pushed closer to extinction (figure \ref{fig:traj}), suggesting that the system may be more fragile.
-% On the other hand, the presence of \emph{just enough} straying to cause formation of alternative stable states both increases the portfolio effect (figure \ref{fig:PE}c) and reduces the rate of return -- particularly if straying is density-dependent (figures \ref{thetaPE}a, \ref{fig:relax}b).
-% This asymmetry in population densities also leads an asymmetry in the mean trait values, skewed towards the local optimum of the dominant population, one, and this has important -- and often non-intuitive -- ramifications for metapopulation robustness.
-%, as measured by both the portfolio effect as well as recovery time.
-
-An intermediate straying rate increases metapopulation robustness. 
-Results from our model reveal that the presence of just enough straying to cause formation of alternative stable states increases the portfolio effect (figure \ref{fig:PE}a). 
-We note that we do not consider the extremely high PE at the DCB, matched by an extremely long recovery time, to be an indicator of robustness, as the dynamics exactly at or very close to a bifurcation are unlikely to be realized in nature.
-Previous theoretical work has shown that increased connectivity may erode portfolio effects in herring metapopulations, where straying is also thought to be density-dependent \citep{Secor:2009ena}.
-Although high levels of dispersal in our system generally supports this finding, the interplay between dispersal and PE is more subtle when selection for local adaptations is considered.
-Low to intermediate levels of straying result in an elevated PE, increasing the buffering capacity of the metapopulation.
-
-Although PE is measured at the steady-state, low to intermediate rates of straying also appear to have a beneficial effect on transient dynamics.
-When there is just enough straying to cause alternative stable states, the time to recovery following an induced disturbance declines, though -- as with the PE -- it then grows if straying becomes large (figure \ref{fig:relax}a).
-In the alternative steady-state regime, a lower rate of straying inhibits admixture of maladapted individuals.
+A central dynamic of the model is that straying can lead to the emergence of asymmetric alternative stable states, or \emph{migrational meltdown} \citep{Ronce:2001dp}, pushing one of the populations to a dominant, high density, state, and one to a subordinate, low density, state.
+In regime I (figures \ref{fig:traj}a, \ref{bifurcations}), there exists only one configuration for alternative stable states: dominant and subordinate.
+By contrast, in regime II, there exist two configurations: a symmetric intermediate density, or the asymmetric dominant/subordinate state.
+These general dynamic features are exactly those observed for the eco-evolutionary model investigated by Ronce and Kirkpatrick \citep{Ronce:2001dp}, where dispersal is assumed to be symmetric between sites and constant over time.
+% , suggesting that it may be a general feature of spatially-linked populations that evolve toward local optima while being hindered by dispersal.
+An important aspect of our eco-evolutionary model is that there are similar forces that dictate interactions within and between sites, and this naturally results in a symmetry that could limit the relevance of our findings for natural (and inherently less symmetric) systems.
+Although the emergence of alternative stable states via the combined effects of the fold and pitchfork bifurcations are characteristic of symmetrical systems, we find that increasing the asymmetry in the vital rates of populations between sites does not significantly alter the presence or position of alternative stable states (figure \ref{fig:symmetry}).
+
+% Under the assumption of constant straying, we find that an intermediate degree of straying  increases metapopulation robustness. 
+% The presence of just enough straying to cause formation of alternative stable states increases the portfolio effect (figure \ref{fig:PE}a). 
+% % We note that we do not consider the extremely high PE at the bifurcation, matched by an extremely long recovery time, to be an indicator of robustness, as the dynamics exactly at or very close to a bifurcation are unlikely to be realized in nature.
+% Previous theoretical work has shown that increased connectivity may erode portfolio effects in herring metapopulations, however straying in these populations is thought to be density-dependent \citep{Secor:2009ena}.
+% Although high levels of dispersal in our system generally supports this finding, the interplay between dispersal and PE is more subtle when selection for local adaptations is considered.
+% Low to intermediate levels of straying result in an elevated PE, increasing the buffering capacity of the metapopulation.
+
+Under the assumption of constant straying, we find that an intermediate degree of straying increases metapopulation robustness, whereas low to intermediate rates of straying appear to have a beneficial effect on transient dynamics, which is measured by determining the time to recovery following an induced disturbance.
+When there is just enough straying to cause the system to enter regime I, the portfolio effect at first increases and the recovery time at first declines: signs of increased metapopulation robustness.
+In the alternative stable-state regime, a lower rate of straying inhibits admixture of maladapted individuals.
 Following a large disturbance, such as the near-collapse of both dominant and subordinate populations, this limited mixing increases the growth rates of both populations, permitting faster recovery times.
-If the rate of straying becomes too high, an influx of maladapted individuals into both populations inhibits local growth rates, and recovery slows.
-
-% %Density-dependent straying
-% This themed issue is largely concerned with the notion that straying in biological systems where movement is a function of collective navigation may be density-dependent.
-% Berdahl et al.
-(REF) provided a mechanistic hypothesis for density-dependent straying where the proportion of the population that strays is less than the probability that a single individual takes the wrong turn.
-%(as illustrated by \emph{Kevin}; figure \ref{fig:xkcd}) takes the wrong turn.
-% Neither the transient nor asymptotic dynamics that we describe here qualitatively differ as a function of whether the rate of straying is constant or density-dependent, however we observe important quantitative differences that suggest density-dependent straying may play an important role in the persistence of metapopulations over evolutionary time.
+As straying becomes very large large, robustness declines, as observed by a decrease in PE (figure \ref{fig:PE}c) and increase in recovery time (figure \ref{fig:relax}a).
+In this case, increased straying promotes the influx of maladapted individuals into both populations, inhibiting local growth rates, lengthening recovery.
+
+%%%%%%%%%%%%%%%%%%%%%%%
+% THIS IS THE SECTION TO EXPAND
+%%%%%%%%%%%%%%%%%%%%%%%
 
 This themed issue formalizes the role of collective movement in the ecology of natural systems and illuminates a signature of collective navigation in animal populations on the move.
+When straying is dynamic, straying at the steady state is determined by a) the probability that individuals err $m_0$, and b) the influence that group size has on straying at the level of the population, which we assume here is the product of collective decision-making, determined by the parameter $C$ (small values of $C$ mean that the influence of collective behavior on reducing straying is large; Eq.
+\ref{eq:ddm}).
 We highlight three important findings that contribute to our understanding of collective movement suggesting that density-dependent straying may play an important role in the persistence of metapopulations over evolutionary time.
-First, the inclusion of density-dependent straying does not qualitatively alter either (\emph{i}) steady-state or (\emph{ii}) transient dynamics of our eco-evolutionary model, but effectively rescales measures of robustness to the lower straying rates that emerge as a consequence of the coupled dynamics.
-Second, compared to systems with constant dispersal, density-dependent straying appears to increase the portfolio effect across a range of straying rates (figure \ref{fig:thetaPE}a). 
-% and that at low to intermediate levels of straying, PE is high.
-Third, density-dependent straying reduces the time to recovery following disturbance, and this is particularly true in the case of near-collapse of the metapopulation (figure \ref{fig:relax}b).
-In the case of near-collapse, although both populations inherit low population densities, the mean trait values of both are skewed towards the optimum of the dominant population.
-Due to density-dependent straying, low population densities lead to greater dispersal, and while this increased connectivity primarily facilities the growth of the dominant population (because the trait means are closer to the dominant optimum), because the dominant population contains the bulk of the aggregate biomass, the overall recovery time is lessened significantly (figures \ref{fig:relax}b, \ref{fig:relaxtraj_bothlh}).
-% Third, in habitats with greater heterogeneity between sites (measured by an increase in $\Delta\theta$), density-dependent straying diminished robustness at low rates of straying (marginally lower portfolio effects and substantially longer return times; figures \ref{fig:thetaPE}b, \ref{fig:relaxtheta}a,b).
-% Conversely, for increased straying rates, greater and exaggerates the positive effects when $m^*$ is high (higher portfolio effects, and shorter return times).
-
-% %Comparing recovery times
-% Previous theoretical work has shown that the largest perturbations lead to the longest recovery times \citep{Ovaskainen:2002il}.
-% In terms of relative biomass lost, the largest perturbation that we investigate is the near-collapse of both populations, where only 1\% of the pre-perturbation densities survive, however whether this disturbance results in the longest recovery time depends largely on the rate of straying and trait heritability.
-% For example, when straying and trait heritability are low (figure \ref{fig:relax}) the extinction of the larger population maximizes return times, whereas if straying is high, near-collapse maximizes return time.
-% The latter is always true in the alternative stable state regime if trait heritability is high (figure \ref{fig:relax_highh}).
+
+First, if the effects of collective behavior are very strong, the alternative stable state regime I can be avoided altogether.
+This means that - despite potentially high individual error rates - group formation minimizes straying, and the pitchfork bifurcation can be avoided altogether.
+However, this occurs when groups of $\leq 10^2$ individuals significantly minimizes straying, which may be unrealistic.
+As the heterogeneity between habitats increases, avoidance of the pitchfork bifurcation becomes impossible, and alternative stable state regime I becomes unavoidable.
+Second, if collective behavior is of intermediate strength, the differences between the steady state densities are exaggerated, effectively pushing the subordinate population closer to extinction (figure \ref{fig:cb}).
+Moreover, although robustness generally declines with increased individual straying (quantified by a decrease in PE and an increase in recovery time), it remains elevated iff collective behavior is of intermediate strength (figure \ref{pert}), but this effect is eroded as $m_0 \rightarrow 0.5$\\.
+
+%implications of single extinctions
+Third, we find that the alternative stable state regime II expands across a larger range of $m_0$ as the effects of collective behavior grow.
+In regime II, alternative stable states are generally avoided, however there exist hidden low-density basins of attraction, which may trap a population in the wake of a large disturbance.
+We observe this by instituting the extinction of a single population.
+If the system is outside of regime II, the population will rebound to its previous state as it is colonized by the surviving population.
+If the system is within regime II, the recovering population will become trapped in the subordinate basin of attraction, and will not recover its previous density.
+So although strong collective behavior can provide benefits to a metapopulation by either
+1) avoiding the alternative steady state regime I, or
+2) increasing robustness for higher individual straying rates,
+there is an unavoidable cost, which lies in the expanded parameter space dominated by these hidden low-density basins of attraction.
+
+[Beta] The dynamics of our eco-evolutionary model with either constant or density dependent straying can be divided into: a high density symmetric state, the alternative stable state regimes I and II, and a low density symmetric state.
+It is tempting to speculate that high density symmetric states are always more preferable, that alternative stable states are always less preferable, and that the low density symmetric state is to be avoided.
+Although the latter is undoubtedly true from a conservation perspective, it is less clear to what extent high density symmetric states are `better' or `worse' than the alternative stable states found in regimes I or II.
+For example, although high densities buffer populations against large disturbances, symmetric densities are by definition synchronized, which eliminates the potential for ecological rescue.
+By contrast, although regime I subjects one population to a subordinate state, and regime II harbours low-density basins of attraction, these dynamic features decrease synchronization and increase measures of robustness such as PE and recovery time.
+What can be said is that alternative stable state regimes increase the potential complexity of the system, which may introduce considerable challenges from a management perspective.
+Of particular concern for management is the expansion of regime II for organisms with non-neglible individual straying and that navigate by collective decision-making, due to the likely undetectable basins of attraction that lurk at low population densities.
 
@@
 Our surrogate measure for habitat heterogeneity is the difference in trait optima between sites $\Delta\theta$.
-In general, our findings indicate that increased habitat heterogeneity promotes robustness (higher PE, shorter time to recovery) when straying rates are low, but may erode robustness when straying rates are high (figure \ref{fig:thetaPE}b, solid lines).
+% In general, our findings indicate that increased habitat heterogeneity promotes robustness (higher PE, shorter time to recovery) when straying rates are low, but may erode robustness when straying rates are high (figure \ref{fig:thetaPE}b, solid lines).
+We show that as habitat heterogeneity increases, the occurrence of alternative steady states associated with regime I becomes unavoidable, particularly for $0.1 \leq m_0 \leq 0.4$, and regime II becomes minimized (figure \ref{fig:hystheta}).
 This may be particularly consequential for populations that are spatially adjacent but separated by sharp environmental boundaries, such that trait optima are divergent yet dispersal is relatively high.
@@
 Alternatively, salmon that cue to specific environmental conditions may be more likely to stray into sites that are structurally and physiognamically more similar \citep{Peterson:2014gy}.
-These considerations justify imposing a direct relationship between the rate of straying $m$ and habitat heterogeneity: as site dissimilarity increases, so too should the straying rate decrease (figure \ref{fig:mthetarelation}).
+These considerations justify imposing a negative correlation between habitat heterogeneity and individual straying $m_0$: as site heterogeneity increases, so too should straying decrease (figure \ref{fig:mtheta}).
 % As site dissimilarity increases, so to should the optima in trait values for the resident populations.
-When habitat heterogeneity and the rate of straying are linked in this way, we show that very small amounts of either constant or density-dependent straying result in long recovery times for the dominant population because there is time for selection to push the subordinate trait mean away from the optimum of the dominant population (figure \ref{fig:mtheta}).
+When habitat heterogeneity and straying are linked in this way, we show that very small amounts of straying give rise to regime I, and that regime II occurs for higher values of $m_0$ (figure \ref{fig:mtheta}).
+This pattern is opposite that observed for scenarios were habitat heterogeneity and straying are assumed to be independent, and suggests that increases in straying that are associated with growing similarities between habitats can push a metapopulation into a regime where hidden low-density basins of attraction exist.
+Thus, management activities that alter dispersal rates by outplanting individuals or reconnecting disconnected habitats could have unintended eco-evolutionary consequences \citep{Anderson:2013bf,Pess:2014isa}.
+
+% in long recovery times for the dominant population because there is time for selection to push the subordinate trait mean away from the optimum of the dominant population (figure \ref{fig:mtheta}).
 %there is enough isolation to allow the surviving subordinate population to adapt towards its own optimum.
-Such a dynamic is accompanied by an inversion in the alternative stable states following the disturbance, resulting in a state shift in dominance.
-Thus, management activities that alter dispersal rates by outplanting individuals or reconnecting disconnected habitats could have unintended eco-evolutionary consequences  \citep{Anderson:2013bf,Pess:2014isa}.
-% This may have particular conservation implications when aiding dispersal following disturbances or reconnections of lost habitats \citep{Anderson:2013bf,Pess:2014isa}.
-% Moreover, local temperature regimes are known to play a central role in dictating local adaptation of salmon populations.
-% If temperature is the primary determinant of $\Delta\theta$ in terms of our model, this would suggest that populations spanning North-South gradients are more predisposed to distance dependent dynamics.
-% In contrast, the relationship between the rate of straying and habitat heterogeneity may be less clear-cut for populations spanning an East-West gradient, where distant habitats may equally fitting for straying populations.
-
-% Our model is a simple framework for an initial consideration of the eco-evolutionary dynamics of metapopulation and accordingly made assumptions that are important to clarify. 
-% First, our model did not consider the role of straying and dispersal in introducing new genetic material that can combat genetic challenges of small populations such as inbreeding. 
-% Second, our model did not incorporate the potential genetic underpinnings of response diversity and asynchrony. 
-% We anticipate that including a link between genetic divergence and population stochasticity would exacerbate the rate at which higher straying would synchronize populations. 
-% Third, our model assumed that strays were incorporated into the trait distribution of the recipient populations—in natural systems it is possible that strays may die prior to reproduction when the two habitats are highly divergent.
-% Fourth, natural systems are quite temporally? variable, which could push systems out of steady-state dynamics, which is what our model focused on. 
-% Fifth, we focused on an extremely simple system of two linked populations.
-While more linked populations would introduce additional complexity, we predict that our qualitative results would apply to more complicated systems.
-% It is clear that understanding the eco-evolutionary dynamics of metapopulations is a rich area for further research.
-
+% Such a dynamic is accompanied by an inversion in the alternative stable states following the disturbance, resulting in a state shift in dominance.
+% 
 Although our study was inspired by salmon metapopulations, the results have general implications for the conservation and management of other migratory metapopulations as well. 
-Because changes in straying rates can have large and nonlinear impacts on robustness, human activities that alter straying rates could have unintended consequences. 
+Because changes in straying can have large and nonlinear impacts on robustness, human activities that alter straying could have unintended consequences. 
 For example, salmon produced by hatcheries often stray into proximate wild populations \citep{Brenner:2012gl}, and these recipient populations can have lower fitness due in part to the introduction of maladapted genes \citep{Ford:2002ip}.
-We show that there is an intermediate straying rate where disturbed populations that are recovering by the introduction of maladapted strays recover fastest: if the straying rate is too low or too high, recovery times increase (figure \ref{fig:relax}).
+We show that an intermediate individual straying rate can result in faster recovery times following a large disturbance, but that this is largely determined by the strength of collective behavior.
 This finding suggests that salmon stocking efforts that aim to lower recovery times following dam removal could actually prolong recovery if the rate at which individuals are introduced and the suitability of those fish in that habitat (i.e.
@@
 
-The portfolio effect and the time to recovery following a disturbance are independent and correlated measures of metapopulation robustness that take into account both steady-state and transient dynamics.
-We show that these measures of robustness are strongly influenced by the rate at which individuals from donor populations stray into habitats occupied by recipient populations. 
-Importantly, density-dependent straying, which may occur when individuals collectively navigate, can both increase the portfolio effect and lower the time to recovery following a disturbance, which is anticipated to promote persistence.
-Therefore, preserving the biological processes that facilitate this collective behavior may be an important conservation target in its own right, echoing the sentiments of  Hardesty-Moore et al.
+Density dependent straying, whether it is caused by collective decision-making or otherwise, has a large influence on the dynamics of populations in the presence of local adaptation.
+The rate at which individual err, and the influence of group size on straying at the population level, are two important components of dynamic straying \cite{Berdahl:2016dx}.
+We show that changes in these characteristics can alter the occurrence and positioning of two different alternative stable state regimes, one of which may harbor hidden low-density basins of attractions that can effectively trap populations after a large disturbance.
+Generally, increasing the strength of collective behavior mitigates the potentially negative impacts of what Ronce and Kirkpatrick describe as \emph{migrational meltdown} \cite{Ronce:2001dp}.
+Preserving the biological processes that facilitate this collective behavior may be an important conservation target in its own right, echoing the sentiments of  Hardesty-Moore et al.
 \citep{HardestyMoore:wg}.
@@
 \\ \\
-
-% 
-% 
-% %Concluding
-% The importance of selection and its influence on the dynamics of populations is increasingly recognized as vital for assessing species' fragility.
-% Of particular interest is how selection might influence spatially coupled metapopulations distributed across environmental gradients.
-% We have shown that two measures of metapopulation robustness -- the portfolio effect and the time to recovery following a disturbance -- are strongly influenced by the rate at which individuals from a donor population stray into habitats occupied by a recipient population.
-% Importantly, density-dependent straying, which may occur when individuals collectively navigate, can both increase the portfolio and lower the time to recovery following a disturbance, which is anticipated to promote persistence.
-% We suggest that understanding the spatial complexity of metapopulations dispersing across heterogeneous environments, combined with the mosaic of selective forces acting on those environments, may be key to discovering those factors that promote persistence.
+% The portfolio effect and the time to recovery following a disturbance are independent and correlated measures of metapopulation robustness that take into account both steady-state and transient dynamics.
+% We show that these measures of robustness are strongly influenced by the rate at which individuals from donor populations stray into habitats occupied by recipient populations. 
+% Importantly, density-dependent straying, which may occur when individuals collectively navigate, can both increase the portfolio effect and lower the time to recovery following a disturbance, which is anticipated to promote persistence.
+% Therefore, preserving the biological processes that facilitate this collective behavior may be an important conservation target in its own right, echoing the sentiments of  Hardesty-Moore et al.
+\citep{HardestyMoore:wg}.
+% We suggest that understanding the spatial complexity of metapopulations dispersing across heterogeneous environments, in tandem with the mosaic of selective forces acting on those environments, may be key to uncovering those factors that promote persistence in the wild.
 
@@
 \bibliography{aa_kevin}
+
 % \begin{thebibliography}{10}
@@
 \centering
-\includegraphics[width=0.9\textwidth]{fig_relax_lowh.pdf}
+\includegraphics[width=1\textwidth]{fig_meandiff.pdf}
 \caption{
-Recovery time of $N_T^*$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady-state $m^*$) with trait heritability $h^2=0.2$.
-If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
-DCB marks the discrete cusp bifurcation.
-} \label{fig:relax}
+Inset: steady states for the constant straying model (purple) and density-dependent straying model (green) for different strengths of collective behavior.
+Top row shows steady state densities as a function of individual straying $m_0$; bottom row shows steady state densities as a function of steady state straying $m(t)^*$
+Main: Difference in steady states averaged across low straying values ($0 < m,m_0 < 0.25$; blue) and high straying ($0.25 < m,m_0 < 0.5$; red).
+} \label{fig:cb}
 \end{figure*}
 
-\begin{figure}
+\begin{figure*}
   \captionsetup{justification=raggedright,
@@
 \centering
-\includegraphics[width=0.5\textwidth]{fig_traitdiff.pdf}
+\includegraphics[width=1\textwidth]{fig_hysteresis_ddm.pdf}
 \caption{
-Phenotypic diversity ($\Delta \mu^*$) evaluated at the steady-state as a function of straying rate $m$ and trait heritability $h^2$.
-The jump occurs as the system crosses the cusp bifurcation; lower phenoytic diversity emerges with higher straying rates and in the alternative stable state regime.
-} \label{fig:traitdiff}
-\end{figure}
-
-\begin{figure}
+Regime I (black); Regime II (gray). Leftmost white is symmetric high density stable state regime; white space to left is symmetric low density stable state regime. 
+} \label{fig:bifurcationsddm}
+\end{figure*}
+
+\begin{figure*}
   \captionsetup{justification=raggedright,
@@
 \centering
-\includegraphics[width=0.4\textwidth]{fig_thetaPEmvm.png}
+\includegraphics[width=1\textwidth]{fig_rtpe_ddm.pdf}
 \caption{
-(a) Median portfolio effect as a function of a constant stray rate $m$ (solid line) and density-dependent stray rate (point pairs) given heritability is $h^2 < 0.5$ and $\Delta\theta=5$.
-Point pairs connected by a horizontal line represent the PE as a function of density-dependent straying rates, evaluated for both low- and high-density populations at equilibrium.
-The lower straying rate of a pair is for the larger population; the higher straying rate is for the smaller population.
-(b) Median portfolio effects for habitats with increasing heterogeneity as measured by the difference in regional trait optima $\Delta \theta$ for both constant and density-dependent stray rates as shown in (a).
-Portfolio effects greater than unity corresponds to less synchronization.
-DCB marks the discrete cusp bifurcation.
-} \label{fig:thetaPE}
-\end{figure}
-
-\begin{figure}
+a) Portfolio effect
+b) Recovery time following near-collapse of both populations.
+} \label{fig:pert}
+\end{figure*}
+
+\begin{figure*}
   \captionsetup{justification=raggedright,
@@
 \centering
-  \includegraphics[width=0.5\textwidth]{fig_mtheta_rt.pdf}
+\includegraphics[width=1\textwidth]{fig_hysteresis_ddm_theta.pdf}
 \caption{
-  Recovery times under three disturbance types for systems where higher habitat heterogeneity $\Delta \theta$ corresponds to lower straying $m$ (see figure \ref{fig:mthetarelation}), and vice versa. 
-  % When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases.
-  The `$*$' marks the value of $m$ below which (low straying between highly heterogeneous environments) there is an inversion between subordinate/dominant states following extinction of the dominant population.
-  DCB marks the discrete cusp bifurcation.
+Each subfig: Regime I (black); Regime II (gray).
+Leftmost white is symmetric high density stable state regime; white space to left is symmetric low density stable state regime.
+The cutoff white area on rightmost fig is an artifact.
+} \label{fig:hystheta}
+\end{figure*}
+
+\begin{figure*}
+  \captionsetup{justification=raggedright,
+singlelinecheck=false
+}
+\centering
+\includegraphics[width=1\textwidth]{fig_hysteresis_ddm_mtheta.pdf}
+\caption{
+Inset: relationship imposed between $m_0$ and $\Delta\theta$.
+Regime I (black); Regime II (gray). Rightmost white is symmetric (low-density) stable state.
 } \label{fig:mtheta}
-\end{figure}
+\end{figure*}
+
+% 
+% \begin{figure}
+%   \captionsetup{justification=raggedright,
+% singlelinecheck=false
+% }
+% \centering
+% \includegraphics[width=0.4\textwidth]{fig_thetaPEmvm.png}
+% \caption{
+% (a) Median portfolio effect as a function of a constant stray rate $m$ (solid line) and density-dependent stray rate (point pairs) given heritability is $h^2 < 0.5$ and $\Delta\theta=5$.
+% Point pairs connected by a horizontal line represent the PE as a function of density-dependent straying rates, evaluated for both low- and high-density populations at equilibrium.
+The lower straying rate of a pair is for the larger population; the higher straying rate is for the smaller population.
+% (b) Median portfolio effects for habitats with increasing heterogeneity as measured by the difference in regional trait optima $\Delta \theta$ for both constant and density-dependent stray rates as shown in (a).
+% Portfolio effects greater than unity corresponds to less synchronization.
+% DCB marks the discrete cusp bifurcation.
+% } \label{fig:thetaPE}
+% \end{figure}
+% 
+% 
+% \begin{figure}
+%   \captionsetup{justification=raggedright,
+% singlelinecheck=false
+% }
+%   \centering
+%   \includegraphics[width=0.5\textwidth]{fig_mtheta_rt.pdf}
+%   \caption{
+%   Recovery times under three disturbance types for systems where higher habitat heterogeneity $\Delta \theta$ corresponds to lower straying $m$ (see figure \ref{fig:mthetarelation}), and vice versa.
+%   % When straying is distance dependent, $m$ increases as $\Delta\theta$ decreases.
+%   The `$*$' marks the value of $m$ below which (low straying between highly heterogeneous environments) there is an inversion between subordinate/dominant states following extinction of the dominant population.
+%   DCB marks the discrete cusp bifurcation.
+%   } \label{fig:mtheta}
+% \end{figure}
+% 
+%
 
@@
 \beginsupplement
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-% \includegraphics[width=0.35\textwidth]{fig_mthetarelation.pdf}
-\caption{
-In some cases, habitat heterogeneity may be assumed to determine the rate of straying, if for example:
-1) sites are distributed over greater spatial distances, where habitat differences are assumed to be greater between more distant sites, or 2) individuals have behaviors promoting dispersal between habitats that are more similar.
-To examine such cases, we use the relationship $m= 0.5(1 + \Delta\theta)^{-1}$ where maximum straying is assumed to occur at $m=0.5$ (perfect mixing).
-} \label{fig:mthetarelation}
-\end{figure}
 
@@
 
+% 
+% \begin{figure}
+%   \captionsetup{justification=raggedright,
+% singlelinecheck=false
+% }
+% \centering
+% \includegraphics[width=0.35\textwidth]{fig_hysteresis2Dhys.pdf}
+% \caption{
+% White (low $m$)
+% Gray
+% Black
+% White (high $m$)
+% The white/black region is separated by a sub-critical pitchfork bifurcation ($\rm PB_*$); the gray/black region is separated by a super-critical pitchfork bifurcation ($\rm PB^*$); the white/gray region is separated by a fold bifurcation (FB).
+% } \label{fig:bifurcations}
+% \end{figure}
+
 \begin{figure}
@@
 
-\begin{figure}
+\begin{figure*}
   \captionsetup{justification=raggedright,
@@
 \centering
-% \includegraphics[width=0.35\textwidth]{fig_hysteresis.pdf}
+% \includegraphics[width=0.9\textwidth]{fig_relax_lowh.pdf}
 \caption{
-Increasing the straying rate results in the transition from a single steady-state for both populations to a dominant and subordinate states.
-If the straying rate is subsequently lowered, the single steady-state is not easily obtained, which is the hallmark of hysteresis.
-} \label{fig:hysteresis}
-\end{figure}
-
-\begin{figure}
+Recovery time of $N_T^*$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady-state $m^*$) with trait heritability $h^2=0.2$.
+If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
+DCB marks the discrete cusp bifurcation.
+} \label{fig:relax}
+\end{figure*}
+
+\begin{figure*}
   \captionsetup{justification=raggedright,
@@
 \centering
-% \includegraphics[width=0.35\textwidth]{fig_relax_large.pdf}
+% \includegraphics[width=0.9\textwidth]{fig_rtpe_ddmsl.pdf}
 \caption{
-Extinction of high-density population with a high straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
-Black line marks the calculated point of recovery post-perturbation.
-Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
-} \label{fig:relaxtraj_hdlh}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-% \includegraphics[width=0.35\textwidth]{fig_relax_small.pdf}
-\caption{
-Extinction of low-density population with a high constant straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
-Black line marks the calculated point of recovery post-perturbation.
-Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
-} \label{fig:relaxtraj_ldlh}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-% \includegraphics[width=0.9\textwidth]{fig_relax_highh.pdf}
-\caption{
-Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady-state $m^*$) with trait heritability $h^2=0.8$.
-If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
-} \label{fig:relax_highh}
-\end{figure}
-
-\clearpage
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-% \includegraphics[width=0.4\textwidth]{fig_thetadiffN.pdf}
-\caption{
-Median difference in population densities taken over the straying rate as a function of habitat heterogeneity $\Delta\theta$.
-Solid lines are for constant $m$; dashed lines are for density-dependent $m$.} \label{fig:thetadiffN}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-% \includegraphics[width=0.8\textwidth]{fig_relaxtheta2.pdf}
-\caption{
-(a) Recovery time after near collapse of both populations as a function of straying rate $m$ and habitat heterogeneity $\Delta\theta$.
-(b) The same as (a) but including recovery times when straying is density-dependent evaluated at $m^*$, shown by linked point pairs.
-Recovery times for systems with density-dependent straying are longer when straying is low and shorter when straying is high, mirroring the change in portfolio effects with respect to density-dependent straying shown in figure \ref{fig:thetaPE}.
-} \label{fig:relaxtheta}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-  \centering
-  % \includegraphics[width=0.35\textwidth]{fig_MDPE_hm_mtheta_rt.pdf}
-  \caption{
-  Portfolio effects as a function of straying rate $m$ and trait heritability $h^2$ when the rate of straying is $m = 0.5(1 + \Delta\theta)^{-1}$.
-Alternative steady-states emerge for low values of $m$ (left of the cusp bifurcation, denoted by the black line), whereas a single steady-state exists for high $m$.
-  } \label{fig:mthetaPE}
-\end{figure}
-
-% \clearpage
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-% \includegraphics[width=0.35\textwidth]{fig_relax_inertia.pdf}
-\caption{
-Population inversion where increased differences in trait optima between sites $\Delta\theta$ corresponds to lower rates of straying $m$.
-At low rates of straying $m=0.02$ ($\Delta\theta=24$), extinction of the dominant population leads to slower-than-expected recovery times because the subordinate population is isolated enough to evolve towards its own trait optimum.
-%until growth of the dominant population overwhelms this local selection.
-In this case, $m$ is less than $m=0.034$ (denoted by the asterisk in figure \ref{fig:mtheta}), such that isolation allows the subordinate population to `run away' from the influence of the dominant population.
-This leads to a switch in subordinate/dominant states for the two populations.
-If $m$ is low but greater than $0.034$, isolation permits the subordinate population to `run away' from the influence of the dominant population, until it is overwhelmed by the recovering dominant population, and reverts back to its previous trait mean prior to disturbance.
-} \label{fig:inertia}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-  \centering
-  % \includegraphics[width=0.8\textwidth]{fig_relax_mtheta.pdf}
-  \caption{
-  Recovery times for three disturbance types when the straying rate covaries with habitat heterogeneity as $m = 0.5(1 + \Delta\theta)^{-1}$ for constant (a) and density-dependent (b) straying rates.
-  The cusp bifurcation is not as clear in (b) because $\Delta\theta$ is a function of the individual straying rate $m_0$, whereas the x-axis in (b) is the straying rate at the steady-state $m^*$.
-  Despite this difference, the general behavior shown in (a) are the same in (b).
-  } \label{fig:mthetamvm}
-\end{figure}
+PE and recovery time after the sole extinction of the subordinate and dominate populations.
+} \label{fig:pertlh}
+\end{figure*}
 
@@
 
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-% \includegraphics[width=0.35\textwidth]{fig_relax_both_lowh.pdf}
-\caption{
-Near collapse of both populations with a low straying rate $m=0.1$ and low trait heritability $h^2=0.2$ (see figure \ref{fig:relax}a).
-Black line marks the calculated point of recovery post-perturbation.
-Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
-} \label{fig:relaxtraj_bothlh}
-\end{figure}
-
 \end{document}

### 2017-12-19 Manuscript/FinalDraft3/msv9.tex
final draft submission to PhilTrans

\RequirePackage{lineno}
\documentclass{revtex4}
%\usepackage[pdftex]{graphicx}

\usepackage{amsmath,amsfonts,amssymb}
\usepackage[english]{babel}
\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{color}
\usepackage{float}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{mathtools}
\usepackage{stmaryrd}
\usepackage{anyfontsize}

\usepackage[font={small}]{caption}
\usepackage{subcaption}
\captionsetup{compatibility=false}
% \usepackage{xr}
% \externaldocument[extfig-]{supp}
%\usepackage{epstopdf}
%\usepackage{array}
%\usepackage{tabularx}
%\usepackage{multirow}
\usepackage{color}
%\usepackage{multibox}
%\usepackage{rotating}
% \usepackage{lineno}

%\usepackage[left]{lineno}
%\usepackage[comma,sort&compress]{natbib}
%\usepackage{authblk}
%\usepackage{multicol}

% \usepackage{longtable}

\linespread{1.5}

% \usepackage[comma,sort&compress]{natbib}
% \bibpunct{[}{]}{,}{a}{}{;}

\renewcommand{\thesection}{\arabic{section}}

\bibliographystyle{prsb}

% \usepackage{bibunits}

\newcommand{\beginsupplement}{%
        \clearpage
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }

% \pagewiselinenumbers
\linenumbers
% \setlength\linenumbersep{3pt}

\begin{document}

% \Blindtext
%\title{Simple rules yield complex communities: deconstructed species interactions and the assembly of communities}
%\title{Community assembly and dynamics by the deconstruction of species interactions}
\title{Eco-evolutionary dynamics, density dependent dispersal, and collective behavior: implications for salmon metapopulation robustness}
\author{
Justin D. Yeakel${}^{1,2,*}$, Jean P. Gibert${}^{1}$, Thilo Gross${}^3$, Peter A. H. Westley${}^{4}$, \& Jonathan W. Moore${}^{5}$ \\
${}^1$School of Natural Sciences, University of California, Merced, Merced CA, USA \\
${}^2$The Santa Fe Institute, Santa Fe NM, USA \\
${}^3$University of Bristol, Bristol, UK\\
${}^4$College of Fisheries and Ocean Sciences, University of Alaska, Fairbanks, Fairbanks AK, USA \\
${}^5$Earth${}_2$Oceans Research Group, Simon Fraser University, Burnaby BC, Canada \\
${}^*$To whom correspondence should be addressed: jdyeakel@gmail.com
}

\begin{abstract} %250 words (now 255)
The spatial dispersal of individuals plays an important role in the dynamics of populations, and is central to metapopulation theory. Although the dispersal of individuals from donor to recipient populations provides connections within the metapopulation, promoting demographic and evolutionary rescue, it may also introduce maladapted individuals into habitats where they are ill-suited, potentially lowering the fitness of recipient populations through introgression of heritable traits. To explore this dual nature of dispersal, we modify a well-established eco-evolutionary model of two locally adapted populations and their associated mean trait values, to examine recruiting salmon populations that are connected by density-dependent dispersal, which we assume emerges from collective navigation. When the strength of collective behaviour is weak such that straying is effectively constant, we show that a low level of straying is associated with the highest gains in metapopulation robustness and that high straying serves to erode robustness. Moreover we find that as the strength of collective behaviour increases, metapopulation robustness is enhanced, but this relationship depends on the rate at which individuals stray. We find that strong collective behaviour increases the presence of hidden low-density basins of attraction, which may serve to trap disturbed populations, and that this is exacerbated by increased habitat heterogeneity. Taken as a whole, our findings suggest that density dependent straying may be a currently unappreciated adaptation in salmon that have evolved in dynamic landscapes and which has important ramifications for the conservation of salmon metapopulations facing both natural and anthropogenic contemporary disturbances.
\end{abstract}

% Although general in nature, our model is inspired by salmon metapopulations, where dispersal between populations is defined in terms of straying, which is known to be density-dependent, and recently proposed to be shaped by social interactions consistent with collective behaviour.
% Thus, dispersal may play contrasting roles as promoter or inhibitor of metapopulation robustness.
% When the rate at which individuals stray is low metapopulation robustness is increased, however it is substantially eroded as individual straying increases.
% Critically, results indicate that density-dependent straying promotes robustness, particularly when population biomass is low and straying is correspondingly high, but also promotes local adaptation and reproductive isolation as populations recover and straying correspondingly declines.

\maketitle

\centerline{Salmon metapopulations, Straying, Dispersal, Eco-evolutionary dynamics, Alternative stable states}
\vspace{2mm}
\noindent {\bf Media Summary} 
Many migratory species, such as salmon, are remarkable in finding their way home. This homing has allowed fine-scale adaptations to the environments in which they evolve. But some individuals do not find their way home and instead stray to other locations, especially when there are fewer individuals to help with collective decision-making. With an eco-evolutionary model, we discovered that low density-dependent straying when the strength of collective behaviour is intermediate  allows linked populations to both be robust to disturbance and maintain local adaptations to their respective habitats.\\

\section{Introduction}

Intraspecific diversity can increase the resilience and stability of species or metapopulations. 
This diversity-stability linkage occurs when there are asynchronous population dynamics within the metapopulation. 
This asynchrony will increase the potential for demographic rescue \citep{Brown:1977gk,Earn:2000fm} and also decrease the variability of processes that integrate across the metapopulation \citep{Anonymous:2015gf}. 
For example, different responses to climate variability within populations of a rare plant reduced fluctuations in abundance \citep{Abbott:2017hl}. 
This statistical buffer has traditionally been quantified as the Portfolio Effect (PE), which is the ratio of the population CV to the CV of the aggregated metapopulation \citep{Thibaut:2012km}. 
Strengthened portfolio effects are expected to increase the robustness of metapopulations to external disturbances, and by extension promote persistence \citep{Thibaut:2012km}.
In contrast, homogenization of populations leading to greater synchronization and weakened PE may be a harbinger of metapopulation collapse and extinction \citep{Carlson:2011ce}.

% Although dispersal and the consequent gene flow can influence the evolutionary dynamics of metapopulations, the interplay between selection, population dynamics, and persistence is not well understood.
Movement of individuals among local populations (i.e. dispersal) can have a large influence on metapopulation persistence \citep{MilnerGulland:2011vm}. 
Dispersal facilitates evolutionary rescue, whereby immigration of individuals with heritable adapative traits can rescue small populations from local extinction in the context of maladaptive environmental change \citep{Bell:2011ki,Carlson:2014is}.
Dispersal also enables demographic rescue, when depressed or extirpated populations are recolonized by immigrants from the rest of the metapopulation.
On the other hand, high rates of dispersal may synchronize populations and increase the risk of extinction of the entire metapopulation \citep{Earn:2000fm,Satterthwaite:2015ge}. 
% Dispersal will also influence the evolutionary dynamics of the metapopulation.
% though how the interplay between selection and population dynamics might impact persistence is less well understood. 
Lastly, dispersal may also introduce maladapted individuals into habitats that are host to different environmental conditions, possibly lowering the mean fitness of the recipient population \citep{Ronce:2001dp,Muhlfeld:2014hs}. 
More broadly, dispersal can provide a mechanism by which phenotypes are sorted in space rather than time and facilitates the spread of potentially maladaptive genes \citep{Lowe:2015ft}.
Dispersal in this case may lead to genetic homogenization that erodes the asynchrony underpinning portfolio effects and metapopulation persistence. 
% Thus, straying can influence the resilience and robustness of metapopulations through both ecological and evolutionary processes.
% Thus, the dual nature of straying as both promoter of connections among metapopulation demes and potential eroder of locally adapted gene complexes highlights the interplay between ecological dynamics of connected populations and the evolutionary dynamics of mixed trait distributions that respond to heterogeneous local conditions.

%Movement through space (link back to straying)
% Migratory populations that return to a breeding ground or natal stream to reproduce are linked to each other by some proportion of the population that permanently disperse, or stray into the `wrong' site. %; we might say that there is at least one \emph{Kevin} in every school or flock (figure \ref{fig:xkcd}).
There is growing appreciation that a combination of abiotic, biotic, and anthropogenic factors can control the rate of dispersal among populations \citep{Travis:2013en,H:2013fs,Keefer:2014gg,Bett:2017ha}.
Migratory populations that return to breeding sites for reproduction can be linked to each other by some proportion of the population that permanently disperses into the `wrong' site. 
Recently, the role of social interactions and collective navigation has been hypothesized  as a mechanism shaping the success of philopatric migrations \citep{Berdahl:2015kv,Berdahl:2016dx,HardestyMoore:wg,Berdahl:2017uu}.
The rate at which individuals disperse may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \citep{Simons:2004jo,Berdahl:2015kv,Berdahl:2016dx,Berdahl:2017uu}.
Thus, dispersal rates can be higher at lower population abundances \citep{Berdahl:2014bl}, which may profoundly influence the eco-evolutionary dynamics of metapopulations. 
% The potential influence of collective dispersal on the dynamics of individual populations and the metapopulation as a whole is a topic of considerable interest that has tangible conservation implications \citep{Brenner:2012gl,Johnson:2012fe,Fullerton:2011ii}.

%Eco-evolutionary dynamics in space
%% From geographic mosaic of coevolution to...
% That evolutionary forces of selection and gene flow play out heterogeneously across geographic mosaics is now a foundational concept in ecology and evolutionary biology \citep{Nuismer:1999ko,Thompson:2015hq,Nuismer:2000bb,Thompson:2005wf}.
% These mosaics are in part driven by environmental differences between habitats that alter the selective forces acting on different phenotypes \citep{Endler:1986tz}, and a principle underlying assumption is that there is gene flow such that individuals from different habitats mix over space \citep{Gomulkiewicz:2000bz,Thompson:2002hr,Nuismer:2003eh,Nuismer:2006be,Forde:2008kc,Guimaraes:2011hu,Gibert:2013hp}.
% Although the evolutionary outcomes of these spatial processes have been explored in depth (REFS), it is less well understood how selective mosaics and their consequent evolutionary forces impact population dynamics as they unpitchfork \citep{Hendry:2016un}.

%Motivation: straying
The eco-evolutionary impacts of dispersal likely have implications for conservation and management in key taxa such as in migratory salmon \citep{Brenner:2012gl,Johnson:2012fe,Fullerton:2011ii}.
While anadromous salmonid fishes (genera \emph{Oncorhynchus} and \emph{Salmo}) are renown for returning to their natal spawning habitats with high accuracy and precision after years at sea \citep{Quinn:2011tf,Jonsson:2011kg,Keefer:2014gg}, there are generally some individuals that disperse (termed `straying' and used synonymously with dispersal hereafter) to non-natal sites to spawn \citep{Quinn:1993ge,Hendry:2004wf}.
Straying is fundamental to the evolutionary legacy of populations as it has provided a mechanism for the colonization of new or connected habitats following glacial retreat or large scale geomorphic landscape change \citep{Hendry:2004wf}.
Salmon may operate as metapopulations, where populations are genetically distinct but linked by some level of straying \citep{Schtickzelle:2007wb,Anderson:2014cx}.
Although extensive work has been done to document the extent of straying from donor populations into recipient populations \citep{Keefer:2014gg,Bett:2017ha}, only recently have the abiotic, biotic, and anthropogenic influences of straying behaviours been investigated systemically \citep{Keefer:2008bs,Westley:2015to,Bond:2016dz}.
Straying among salmon may be influenced by environmental factors such as water temperature, human activities such as hatchery practices, and population density as predicted by the collective navigation hypothesis \citep{Peterson:2014gy,Berdahl:2017uu}.
Straying can introduce new maladaptive genotypes into the recipient population, while the ensuing genetic homogenization could synchronize population dynamics and erode portfolio effects \citep{Moore:2010gs,Carlson:2011ce,Braun:2016ib}.
Given that locally distinct populations are often linked by straying, there is an opportunity and need to understand the fundamental and applied consequences of straying on metapopulation persistence, conservation, and management.

%Here we
Here we seek to explore how collective density-dependent straying influences the stability and robustness of metapopulations through ecological and evolutionary processes.
To address this question we build upon an established eco-evolutionary model of two populations occupying different sites that are linked by straying individuals, each with an associated trait distribution subject to natural selection determined by local conditions \citep{Ronce:2001dp}.
Specifically we compared (a) density-independent (constant) straying with (b) density-dependent straying as a function of the rate at which individuals stray and the strength of collective behaviour across (c) increasing environmental heterogeneity, by assessing two measures of metapopulation robustness: the portfolio effect and the time required for a population(s) to recover following an induced disturbance. 
This model enables us to explore the tradeoff between the potentially detrimental erosion of local adaptation vs. the positive effects of demographic and evolutionary rescue, both of which are facilitated by straying and the effects of collective navigation.

\section{Model Description \& Analysis}

\noindent{\bf (a) Metapopulation framework}\\
\noindent We follow the basic framework described by Ronce and Kirkpatrick \citep{Ronce:2001dp}, where dispersal connects two populations $N_i$ and $N_j$ that inhabit two distinct habitats, each with trait values $x_i$ and $x_j$.
In our version of the model, $N_i$ and $N_j$ are locally adapted to site-specific conditions such that there is an optimum trait value $\theta_i$ and $\theta_j$ associated with each habitat, where recruitment is maximized if the trait value of the local population equals the optimum, such that $x = \theta$.
Moreover, we assumed that $x_{i,j}$ are normally distributed with means $\mu_{i,j}$ and have the same standard deviation $\sigma$.
As such, the recruitment rate $R_i(\mu_i(t),\theta_i)$ for both populations is determined by the mean trait value of the local population relative to the optimal value at that site.
Mean trait values for both populations are dynamic variables and change over time in response to differences in recruitment as individuals mix between sites.
Trait means for each population are thus subject to selection, the strength of which is proportional to the difference between the trait mean and the local trait optimum at a given point in time \citep{simpson1953major,Lande:1976ga,Ronce:2001dp}.
This is broadly consistent with empirical patterns observed in pacific salmon dynamics \citep{Reisenbichler:1988ex}.
The two populations occur in spatially separate sites that are close enough that a proportion of the population $m$ strays into the wrong site.
If there is no straying between these populations, then the mean trait evolves towards the optimal value for each site $\mu_i \rightarrow \theta_i$, and the recruitment rate for each population is maximized.
If there is straying between populations, then the trait means in each respective location will be pulled away from their optima, and recruitment rates will decline.
As $m \rightarrow 0.5$, the populations are perfectly mixed.

We used the discrete Ricker framework described by Shelton and Mangel \citep{Shelton:2011eq} as the basis for our two-site metapopulation model, with the added effect that the size of the local population $N_i$ is altered by mixing $mN_j$ individuals from the remote population.
Moreover, we assume that there is no demographic overlap between generations, consistent with the life history of many populations of pink salmon (\emph{O. gorbuscha}) who all mature at two years of age and die after one reproductive season..
In this sense, both populations serve simultaneously as donor and recipient populations.
Because total recruitment will be determined by both locals (with a mean trait value closer to the site optimum) and strays (with mean trait values farther from the local optimum), the recruitment of the aggregate is determined by the mean of the trait mix $R_i(\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\theta_i)$, where
\begin{equation}
\omega_i=\frac{(1-m)N_i(t)}{(1-m) N_i(t) + m N_j(t)}.
\end{equation}
This mix of individuals is subject to identical compensatory effects, which is determined by the parameter $\beta$.
Taken together, the difference equation that determines changes in population size from time $t$ to $t+1$ is

\begin{align}
  \label{eq:N}
  &N_i(t+1) = R_i(\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\theta_i) \\ \nonumber
  &\cdot \left((1-m)N_i + mN_j\right){\rm e}^{-\beta ((1-m)N_i(t) + m N_j(t))},
\end{align}

\noindent where the recruitment of the population as a function of the mean trait value at time $t$ and the local trait optimum is

\begin{align}
  \label{eq:R}
  &R_i[\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\theta_i] = \\ \nonumber
  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i,\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\sigma^2) {\rm d}x_i +\tilde{P}_i\\ \nonumber
  &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-(\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t)))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\} +\tilde{P}_i.
\end{align}

\noindent The mismatch between the mean of the local trait mix $\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t)$ and the local optimum $\theta_i$ scales the recruitment rate for the population, and $\tilde{P}_i\sim {\rm Normal}(0,0.01)$ introduces a small amount of demographic variability.
The parameter $\tau$ is the strength of selection, and controls the sensitivity of recruitment to changes in the mean trait value away from the optimum.
Because straying individuals are emigrating from a population with a mean trait value farther from the local optimum, their rate of recruitment is diminished.
Recent studies of wild sockeye salmon have indeed found that straying individuals have lower life-time fitness than individuals that do not stray, although it is unknown at what life-stage this selection occurs \citep{Peterson:2014gy}.
% The compensatory effects are then determined by the exponential, following the Ricker stock-recruitment relationship.

% \noindent{\bf (b) Recruitment over a selective landscape}
Because individuals from the local population are mixed with individuals from the remote population via straying and subsequent reproduction, the resulting trait distribution is a complex mixture of trait distributions.
We make two simplifying assumptions.
First, we approximate the distribution resulting from the mix of remote and local individuals prior to reproduction as a Gaussian distribution, where $X_i=x_i$ with probability $g(x_i)$.
The expectation of the actual trait distribution as well as the Gaussian approximation are the same, such that ${\rm E}\{g(x_i)\} = \omega_i\mu_i + (1-\omega_i)\mu_j$, with weights corresponding to the proportion of the mixed population that are local individuals, $\omega_i$, and straying individuals, $1-\omega_i$.
Thus, strays can successfully reproduce and introduce their genotypes into the recipient population, which is supported by observations in wild populations \citep{Jasper:2013cc}.
Second, we assumed that changes in trait variance through time are minimal, such that $\sigma^2$ is constant over time, which is a common simplification in eco-evolutionary models of population dynamics \citep{Lande:1976ga,Ronce:2001dp,Schreiber:2011wx,Gilbert:2014ee,Gibert:2015kc}.
These simplifications are the same as those introduced by Ronce and Kirkpatrick, and were shown to have negligible impacts on dynamics \citep{Ronce:2001dp}.

% An increasing flow of incoming strays is generally expected to pull the mean trait value of the local population away from its optimum over time, which will decrease its rate of recruitment.
Following Lande \citep{Lande:1976ga}, and given our assumption of trait distribution normality, the mean trait value thus changes through time according to the difference equation

\begin{align}
  \label{eq:mu}
  \mu_i(t+1) &= \omega_i\mu_i(t) + (1-\omega_i)\mu_j(t) + h^2\sigma^2\frac{\partial}{\partial \mu^\prime}\ln\left(R_i[\mu^\prime,\theta_i] \right), \\ \nonumber
  &= \omega_i\mu_i(t) + (1-\omega_i)\mu_j(t) + h^2\sigma^2\left(\frac{\theta_i - \omega_i\mu_i - (1-\omega_i)\mu_j}{\sigma^2+\tau^2} \right),
\end{align}
with $\mu^\prime = \omega_i \mu_i(t)+ (1-\omega_i)\mu_j(t)$.
Although trait heritability among salmonids is variable, most life-history traits have an $h^2 < 0.5$ \citep{Carlson:2008hl}, and for all additional analyses we have conservatively set $h^2=0.2$.\\

\noindent {\bf (b) Density-dependent straying}
%Density-dependent m
\noindent There is mounting evidence that straying is density-dependent, perhaps reflecting a signature of collective navigation \citep{Berdahl:2014bl,Berdahl:2017uu}.
Specifically, straying has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tend to decrease the rate at which individuals err, reducing the overall proportion of a population that strays.
According to Berdahl et al. \citep{Berdahl:2016dx}, given the probability that an individual strays is $m_0$, the proportion of the local population $N_i(t)$ that strays is

\begin{equation}
  m(t) = m_0\left(1- \frac{N_i(t)}{C+N_i(t)}\right),
  \label{eq:ddm}
\end{equation}

\noindent where $C$ is a half-saturation constant, and determines to what extent collective behaviour, as a function of group size, diminishes straying.
If $C$ is small, relatively small groups of organisms `correct' for even high individual error rates, suppressing straying between sites.
In the context of our model, small values of $C$ indicate that the effects of collective behaviour on modifying straying are strong.
As $C \rightarrow \infty$, groups of any size have no impact on straying, such that $m(t) \rightarrow m_0$, and the model reverts to that of constant straying.
When $C \ll \infty$ and the population density is high, $m(t) \rightarrow 0$, whereas if the population is small, individuals operate without regard to the collective, such that $m(t) \rightarrow m_0$, such that in all cases, $m(t) < m_0$.
% For realistic population densities, $m(t) < m_0$.\\
% We note that at the limit $C\rightarrow \infty$, the density-dependent straying rate becomes constant such that $m(t) \rightarrow m_0$, and this corresponds to the original formulation where $m=m_0$.

% \noindent {\bf (c) Habitat heterogeneity}
% \noindent Increasing differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater regional differences in the conditions that favor alternative trait complexes, which can be interpreted as increased habitat heterogeneity.
% If both populations are isolated, natural selection will direct the mean trait values of both populations towards their respective optima, such that $\mu_i(t) \rightarrow \theta_i$ as $t\rightarrow\infty$.
% Habitat heterogeneity and the rate of straying are treated both independently, and as parameters that covary.
% In the latter instance, we examined a case where it is assumed that increased habitat heterogeneity correlates with lower straying rates, and vice versa (illustrated in figure \ref{fig:mthetarelation}).
% Two scenarios may lead to this correlation: 
% (\emph{i}) sites may be distributed over greater spatial distances, where habitat differences are assumed to be exaggerated and the likelihood of straying over greater distances is lower \citep{Candy:2000hu,JPE:JPE1383};
% (\emph{ii}) individuals may have behaviours promoting dispersal between habitats with structural or physiognomic similarities \citep{Peterson:2014gy}.
% In this case, the rate of straying would be greater between habitats with smaller differences in trait optima (lower $\Delta\theta$) and lesser between habitats with greater differences in trait optima (higher $\Delta\theta$).
% \\

\noindent {\bf (c) Measuring metapopulation robustness}
\noindent We evaluated metapopulation robustness by measuring the average-CV portfolio effect (PE) \citep{Anderson:2014cx,Schindler:2015gf} as well as the recovery time, which is the time required for the system to return to a steady-state following an induced disturbance to one or both populations \citep{Ovaskainen:2002il}.
Throughout, we refer to an increase in portfolio effects and/or reduction in recovery time as promoting metapopulation robustness.

The average-CV portfolio effect is, as the name implies, the average CV across each population $N_i$ divided by the CV of the aggregate $N_T=\sum_i N_i$ \citep{Anderson:2013gb}, such that

\begin{equation}
\langle{\rm PE}\rangle =\frac{1}{X}\sum_{i=1}^{X} \frac{\sqrt{{\rm VAR}(N_i^*)}}{{\rm E}(N_i^*)}\cdot \frac{{\rm E}(N_T^*)}{\sqrt{{\rm VAR}(N_T^*)}},
\label{eq:pe}
\end{equation}

\noindent where in this case the number of populations is limited to $X=2$ and the expectations $\rm E(\cdot)$ and variances $\rm VAR(\cdot)$ are evaluated at the steady-state.
The steady-state condition is denoted by `$*$'.
As the CV of $N_T^*$ decreases relative to that of the constituent populations, $\langle{\rm PE}\rangle > 1$, and the metapopulation is presumed to be more stable because the aggregate has functioned to dampen population-level variance.
Moreover, portfolio effects greater than unity correspond to less synchronization  \citep{Loreau:2008ju,Anderson:2014cx,Yeakel:2013vz} and thus a greater potential for demographic rescue among populations, buffering the system as a whole against extinction.

%Although there are more robust ways to measure the portfolio effect (in particular for comparing the PE across species with different mean-variance relationships; REF: Anderson), its only utlity here is to provide a simple statistical measure of metapopulation persistence.
A more direct way to measure system robustness is to measure the time required for the system (measured as the aggregate steady-state biomass $N_T^*$) to recover following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation \citep{GuckHolmes}, because we aimed to 
1) assess the effects of a large perturbation far from the stable state, and 
2) estimate the time required for all transient effects to decay following this perturbation (including dampened oscillations), we used a simulation-based numerical procedure.
Recovery time was calculated by initiating a disturbance at $t=t_d$, and monitoring $N_T(t_d+t)$ as $t\rightarrow T$, where $T$ is large. 
The aggregate was deemed recovered at $t_r$, such that recovery time was calculated as $t_r-t_d$, and recovery at $t=t_r$ was determined by the initial $t$ where $N_T(t) < N_T^*\pm{\rm SD}\left( N_T^* \right)$ for $t\in(t_r,T)$, where $\rm{SD}(\cdot)$ is standard deviation (illustrated in figure \ref{fig:recovery}).

Numerically estimating the time that it takes for a perturbed system to recover also permits a more nuanced perspective of metapopulation robustness.
For example, if populations settle to alternative stable states, comparing recovery times after a disturbance applied to individual populations allows for an assessment of which component of the metapopulation has a longer-lasting influence on the system's recovery.
We measured recovery time following three types of induced disturbance: (\emph{i}) extinction of the low-density population; (\emph{ii}) extinction of the high-density population (scenarios \emph{i} and \emph{ii} are equivalent if populations have the same density); (\emph{iii}) near-collapse of both populations where just 1.0\% of each survives.
\\

\section{Results}

%Alternative stable states
% \noindent Density-independent straying lowers stable state densities for both populations by (\emph{i}) the numerical loss of locally-adapted individuals from donor to recipient populations and (\emph{ii}) the introduction of maladapted individuals and their associated traits to the recipient population from the donor population (figure \ref{fig:traj}).
% The decline in steady-state densities is abrupt: as straying increases, the system moves from a single stable state where the populations have the same (symmetric) densities, to a region with alternative stable states.
% Two types of alternative stable state regimes occur in this system, and have important effects on robustness.
% The first that is encountered at lower straying $m$, separated by a fold bifurcation \citep{AleksandrovichKuznetsov:1995p2580}, has an intermediate stable state, as well as a dominant and subordinate stable state (regime II in figure \ref{fig:traj}).
% Because the intermediate state is stable, the dominant and subordinate densities are not encountered unless the system is pushed into basins of attraction that define these states (dashed lines in figure \ref{fig:traj}).
% The second type of alternative stable state regime (regime I) is separated from regime II by a pitchfork bifurcation.
% Here the only stable points are the dominant (high) and subordinate (low) stable state densities (regime I in figure \ref{fig:traj}).
% Importantly, we find that increasing the asymmetry in the vital rates of populations between sites does not significantly alter the presence or position of these different regimes (figure \ref{fig:symmetry}).\\

\noindent At low values of density-independent straying the system approaches a fixed point at which both populations persist at equal population size. 
As we increase the straying rate a fold bifurcation occurs in which another fixed point is created (figure \ref{fig:traj}, inset). 
In this alternative state the population sizes are asymmetric. 
The population with the larger population size, i.e. the dominant population, is well-adapted and reproduces quickly. 
A small fraction of this population that strays to the subordinate site constitutes a considerable inflow, such that the population in the subordinate site is not as well adapted, which in turn reduces reproduction and stabilizes the asymmetry.

In a regime found at low straying rates (regime I) both the symmetric and the asymmetric state are stable fixed points of the system (figure \ref{fig:traj}, dashed lines). 
Which of these fixed points is approached depends on the initial conditions. 
The dividing line that separates initial conditions that approach the symmetric solution and initial states that approach the asymmetric solution is a separatrix that runs through an unstable fixed point (figure \ref{fig:traj}, inset) that was also created in the fold bifurcation. 
The set of all initial states from which the system approaches the symmetric state is said to be that state's basin of attraction. 
As we increase the straying rate this basin of attraction shrinks and and eventually vanishes. At this point the unstable asymmetric state collides with the stable symmetric state. 
A bifurcation occurs in which the the unstable asymmetric state vanishes and the symmetric state is destabilized. 
This is a subcritical pitchfork bifurcation that only occurs in systems with symmetries. 
After this bifurcation the stable asymmetric state is the only attractor.

We find a wide regime (regime II) where the system will always approach an asymmetric state where one population is suppressed. 
However, if the straying is increased further the imbalance in population sizes becomes harder to maintain. 
As the straying rate is increased we reach a critical point where the stable asymmetric fixed point becomes symmetric and collides with the unstable symmetric fixed point. 
The system undergoes a bifurcation that is the supercritical form of the subcritical bifurcation encountered earlier. 
In this bifurcation the stable asymmetric fixed point vanishes while the unstable symmetric state is stabilized. 
After this bifurcation the symmetric state is the only attractor in the system.
Importantly, we find that increasing the asymmetry in the vital rates of populations between sites does not significantly alter the presence or position of these different regimes (figure \ref{fig:symmetry}).

%ROBUSTNESS
\noindent{\bf (a) Nonlinear effects of straying on metapopulation robustness} \\
Straying has a large effect on metapopulation robustness, measured by the portfolio effect and the time to recovery following the three types of induced disturbance: near-collapse of both populations, the extinction of the dominant population, and the extinction of the subordinate population (figure \ref{fig:mpert}).
Importantly, the presence of alternative stable state regimes I and II both have a direct impact on robustness as a function of straying $m$.
We observe that as straying increases, the portfolio effect increases sharply as regime I or II is entered, and then declines gradually (figure \ref{fig:mpert}a).
Thus, low levels of straying (2-10\% of the population) are associated with the strongest portfolio effects.
% This non-linear relationship between straying and portfolio effects is linked to the alternative stable state regimes I and II (figure \ref{fig:traj}).

Different types of disturbance lead to different relationships between straying and portfolio effects.
When either population suffers extinction, the PE is shown to increase with lower straying; in the case of near-collapse of both populations, PE increases when straying is higher (figure \ref{fig:mpert}a).
This difference is due to the hidden basin of attraction at low population densities that only plays a role when a disturbance impacts a single population.
In other words, disturbance to a single population can push that population into a low-density alternative state, which in turn contributes to higher PE.
The increase in PE for the synchronous near-collapse scenario occurs at higher values of $m$ when the system enters regime II, where there exists only an asymmetric dominant (high-density) and subordinate (low-density) state.
The PE spikes again when straying is very high and the system leaves regime II, entering a symmetric low-density state.

Similar patterns are observed with respect to the recovery time as straying is increased (figure \ref{fig:mpert}b).
For lower $m$, recovery of single extinctions is impacted by the appearance of low-density basins of attraction in regime I, whereas recovery following near-collapse is not.
For intermediate values of $m$ (regime II), the time to recovery is only diminished when the subordinate population becomes extinct, whereas the time to recovery following near-collapse and the extinction of the dominant population are similar and grow until Regime II is exited at high $m$.
In the case where the subordinate population is extirpated, the most rapid recovery rate occurs when straying is low ($m = 0.08$). 
In contrast, when the dominant population goes extinct, the most rapid recovery is associated with minimal straying. 
It should be noted that when there is no straying ($m = 0$), recovery from extinction is infinite and these values are not shown. 
The relationship between straying and recovery time in the case of synchronous population disturbance is generally positive.

Collectively, these patterns in recovery time and PE are influenced by the different alternative stable states regimes. 
As the alternative stable state regime is approached with increasing $m$, both measures of robustness increase sharply due to an amplification in variance within both populations.
This amplification in variance is the product of \emph{critical slowing down}, which occurs near some bifurcations \citep{Scheffer:2009gg} and has been suggested to serve as an early warning indicator for approaching phase transitions \citep{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br,Krkosek:2014ch}.
At this point, PE peaks along with recovery time, suggesting the former is not a good indicator of robustness very close to the bifurcation.
Because these large increases in robustness pertain to a very small range of $m$, we do not consider them to be biologically relevant, and are primarily useful in this context for observing transitions between dynamic regimes.
In general, high PE corresponds with shorter recovery times, and low PE corresponds with longer recovery times (figure \ref{fig:pevsrt}).
Together, these results suggest that under the assumption of constant (and symmetric) dispersal, robustness depends strongly on the magnitude of straying as well as the type of disturbance experienced by the metapopulation.
% Although the emergence of alternative stable states presents their own risks to the persistence of populations -- including the appearance of potentially hidden low-density basins of attraction -- low-intermediate $m$ also serves to increase robustness by dampening the variance of the aggregate metapopulation (increasing PE), though the effect on recovery time is more idiosyncratic.
We next examine how density-dependent straying challenges these expectations.\\

\noindent{\bf (b) The effects of collective navigation and density-dependent straying} \\
%Portfolio effects
If we assume that straying is density-dependent, the probability that an individual errs $m_0$ in part determines the magnitude of straying within the population (Eq. \ref{eq:ddm}), such that $m(t)$ becomes lower as $N(t)$ increases, which we assume here to be the consequence of collective decision-making behaviours \citep{Berdahl:2016dx}.
% Organisms that employ collection decision-making to aid in navigation have reduced straying rates.
The parameter $C$ in Eq. \ref{eq:ddm} determines to what extent straying is reduced as group size increases, where low values correspond to the effects of strong collective behaviour, and high values correspond to the effects of weak collective behaviour, such that straying is effectively constant \citep{Berdahl:2016dx}.
To enable comparisons between models with constant and density-dependent straying, we compare constant $m$ to both the individual straying probability $m_0$ as well as the value of straying observed at the stable state $m^*$.
In the alternative stable state regimes, the dispersing populations are characterized by alternative $m^*$ values if the populations have asymmetric densities, corresponding to dominant and subordinate states.
% , such that small increases in group size have a large effect on straying, and where $C \rightarrow \infty$ corresponds to constant straying independent of group size .
%Because our analyses concern steady-state conditions, the inclusion of density-dependent straying does not have a large impact on the qualitative nature of our findings (figure \ref{fig:ddm}).
% Density dependence alters the straying ratio at steady-state population densities because $0 < m^* < m_0$, and this serves to rescale both the strength of the PE as well as the recovery time, but does not change the qualitative nature of our findings.

When collective behaviour is very strong, such that $C$ is very low, small increases in population density beget large reductions in straying.
These reductions can be large enough that the system avoids the alternative stable state regime altogether (figure \ref{fig:cb}, left inset).
Conversely, when collective behaviour is very weak, such that $C$ is very high, there is effectively no reduction in straying with increased group size, and the dynamics are those expected if straying were constant (figure \ref{fig:cb}, right inset).
However, when collective behaviour is of intermediate strength ($10^{2} \lesssim C \lesssim 10^3$), the dynamics are altered in two important ways.
First, in the alternative stable state regime, the low density subordinate population has correspondingly higher $m^*$, whereas the high density dominant population has correspondingly lower $m^*$ (figure \ref{fig:cb}, center inset).
Second, the alternative stable state regime results in a $\Delta N^*$ that is reduced when individual straying is low, and magnified when individual straying is high (figure \ref{fig:cb}, main).
In other words, when collective behaviour is of intermediate strength -- the more realistic range for species that navigate via collective decision-making -- increased individual straying exaggerates the differences between the stable state densities, effectively pushing the subordinate population closer to extinction.

%PE, Recovery times (near-collapse)
Density-dependent straying directly alters the dynamic regimes of the model, and this has a large effect on metapopulation robustness.
When the effects of collective behaviour are weak (high $C$) the portfolio effects and recovery times conform to those examined in the case of constant straying (figures \ref{fig:mpert}).
When the effects of collective behaviour are very strong (low $C$), we observe that recovery times are shorter in the case of near-collapse (figure \ref{fig:pert}b).
Recovery times also tend to be shorter when a single population goes extinct except for at very low $m_0$, in which case the time to recovery is much longer (figure \ref{fig:pert}c,d).
As before, when the strength of collective behaviour is intermediate ($10^{2} \lesssim C \lesssim 10^3$), the relationships are more complex.
With intermediate $C$ there is a general increase in robustness if straying is low-intermediate (higher portfolio effect, shorter recovery time), followed by an erosion in robustness as straying becomes high.
In this parameter space, collective migration can mean that the low-density population will stray more, losing well-adapted local individuals, while still receiving some maladapted strays from the larger population, thereby increasing the likelihood of stochastic extinction.

Observed changes in metapopulation robustness are due to changes in alternative stable state regimes I and II as the strength of collective behaviour increases (lower $C$, figure \ref{fig:bifurcationsddm}a). 
When collective behaviour is weak (large $C$), alternative stable states tend to occur at low-intermediate values of individual straying $m_0$.
As collective behaviour is strengthened (smaller $C$), Regime II is avoided at lower valus of $m_0$ and expands for higher values of $m_0$.
When the effects of collective behavior are very strong, Regime II then collapses (black region in figure \ref{fig:bifurcationsddm}a) and gives way to regime I, which plays a larger role over a larger range of $m_0$ when $C$ is low (gray region in figure \ref{fig:bifurcationsddm}a).
Importantly, when the strength of collective behaviour is intermediate, both regimes I and II are relevant at low-intermediate $m_0$.
% In this case, the combined effect of regimes I and II leads to the observed increase in robustness at low-intermediate $m_0$ and subsequent decline with high $m_0$.

% the lone extinction of either population will push the system into a different configuration.
% Directly after extinction and as the site is being recolonized by the undisturbed population, instead of returning to its previously higher (symmetric) density that it shared with the undisturbed population, it becomes trapped in the basin of attraction defined by the subordinate state that is generally not observed barring the occurrence of hysteresis.
% The extinction of either population in regime II therefore has a shorter time to recovery (figure \ref{fig:pertlh}), but this is in part driven by the newly acquired subordinate stable state density.
% Importantly, this low-density basins of attraction, which is particularly prevelant across $m_0$ when the effects of collective behaviour are strong, would not be easy to anticipate or detect prior to a large disturbance.

% 
% In both the constant- and dynamic straying model, there are two regimes that give rise to alternative stable states: one that is produced by a pitchfork bifurcation (regime I), and one that is produced by a fold bifurcation (regime II; figure \ref{fig:bifurcationsddm}).
% When the strength of collective behaviour is intermediate, the cause of the exaggerated difference in stable state densities at high individual straying rates is due to an increase in the dominance of regime I over a greater range of $m_0$ (figure \ref{fig:bifurcationsddm}).
% If the effects of collective behaviour are very strong, regime I vanishes, and the populations generally assume symmetric densities.
% However, while regime I vanishes, the alternative stable state regime created by the fold bifurcation expands (regime II; figure \ref{fig:bifurcationsddm}).
% In regime II there are two possible stable state configurations: 
% 1) the stable states of both populations can either have symmetric densities, or 
% 2) one or both populations can revert to asymmetric dominant/subordinate states as in regime I.
% In this case, populations will attain asymmetric densites if $m_0$ is first increased and subsequently decreased (introducing a hysteresis effect), or if a sufficiently large perturbation directs one of the populations towards the dominant or subordinate basins of attraction.
%

\noindent{\bf (c) The role of habitat heterogeneity and changing selective landscapes}\\
As habitat heterogeneity ($\Delta\theta$) increases, even small amounts of straying can lead to the appearance of alternative stable states, and this is in direct accord with the findings of Ronce and Kirkpatrick \citep{Ronce:2001dp}.
However, if straying is density-dependent, the strength of collective behaviour has a large influence on the occurrence of both alternative stable state regimes I and II.
When heterogeneity is low and the effects of collective behaviour are weak such that straying is constant (high $C$), regime II occurs for small-intermediate $m_0$, and regime I does not play a role (figure \ref{fig:bifurcationsddm}b), mirroring the findings of Ronce and Kirkpatrick \citep{Ronce:2001dp}.
The absence of regime I implies that there are no hidden stable state configurations that might trap a disturbed population in a low-density state.
As the strength of collective behaviour increases, regime I appears at a cusp and becomes increasingly dominant with greater individual straying.
For sites distributed across more heterogeneous habitats, the alternative stable state regimes I and II expand (figure \ref{fig:bifurcationsddm}a,c).
% Only when habitat heterogeneity is intermediate does regime II play a significant role as straying is relatively constant (figure \ref{fig:hystheta}b).
% For high $\Delta\theta$, there is no longer any sanctum from regimes I or II except for very high individual straying.
Regime II dominates all but very high individual straying when the effects of collective behaviour are weak (high $C$), and very low individual straying when the effects of collective behaviour are strong (low $C$).
Moreover, in highly heterogeneous habitats, if the effects of collective behaviour are strong and straying is low (low $m_0$ and $C$ values), regime I, which harbors low-density basins of attraction, cannot be avoided.

Until now, we have treated straying and habitat heterogeneity as independent parameters, however they could covary.
For instance, if sites are separated by greater distance, they may be assumed to have increased habitat heterogeneity as well as less straying.
Alternatively, individuals may be genetically predisposed to stray into sites that are more similar \citep{Peterson:2014gy,Lin:2008kl}, such that higher straying can be assumed to occur between sites that are more homogeneous in aspect.
We implemented this inverse relationship by setting $m_0 = 1/(2+\epsilon\Delta\theta)$ where $\epsilon$ controls the degree to which an increase in $\Delta\theta$ lowers $m_0$ (figure \ref{fig:mtheta}, inset).
Accordingly, $m_0$ is greater for lower $\Delta\theta$, such that there is less straying between dissimilar sites and more straying between similar sites.
Under these conditions we find that regime II appears for very low $m_0$, and regime I appears for higher $m_0$ (figure \ref{fig:mtheta}), which is opposite the case where $m_0$ and $\Delta\theta$ are independent. % $0 < m \leq 0.43$.
As the straying rate increases and $\Delta\theta$ decreases, a single (symmetric) stable state emerges as the fold bifurcation is crossed, which is opposite the pattern observed when straying is independent of habitat heterogeneity.

% not unlike the allee effects observed in Berdahl et al.

\section{Discussion}
%1) Introductory paragraph with take home message
%2) Paragraph on the importance of straying on robustness
%3) Paragraph on the importance of collective straying rates on robustness
%4) Paragraph on habitat heterogeneity and extent of local adaptation on robustness
%5) Paragraph on the emergence of alternative eco-evolutionary states (NO)
%6) Paragraph on implications for salmon management and conservation
%7) Paragraph for beyond salmon

%1) Introductory paragraph with take home message
In this paper we show that density-dependent straying between populations consistent with collective navigation, coupled with localized selection against immigrant phenotypes has large, nonlinear impacts on metapopulation robustness.
Building upon the dynamical framework introduced by Ronce and Kirkpatrick \citep{Ronce:2001dp}, we assess robustness by measuring 
1) the average-CV portfolio effect \citep{Anderson:2013gb,Anonymous:2015gf}, a statistical metric commonly used to assess the buffering capacity of metapopulations, and
2) the recovery time, defined here as the time required for the aggregate metapopulation biomass $N_T$ to return to its steady-state following an induced disturbance, which is mechanistically linked to persistence \citep{Ovaskainen:2002il}.
These statistical and mechanistic descriptors of metapopulation dynamics and robustness are tightly coupled (figure \ref{fig:pevsrt}), which is not uncommon for diverse metrics of stability \citep{Donohue:2013iu}.
% We discovered that a low level of straying is associated with the highest metapopulation robustness for most types of disturbance.
We introduce density-dependent straying by assuming that larger group sizes lower population-level straying from the baseline probability than an individual errs $m_0$, with the strength of this effect determined by $C$ in equation \ref{eq:ddm} (lower values of $C$ indicate that the effects of collective behaviour are strong).
%Main takehome
Generally, we find that when the effects of collective behaviour are strong, metapopulation robustness is enhanced.
However, empirical observations of natural populations suggest that the effects of collective behaviour are intermediate (e.g. $10^{2} \lesssim C \lesssim 10^3$) \citep{Berdahl:2014bl,Berdahl:2016dx}.
% , and that the rate at which individuals err is such that $m_0 \ll 0.5$.
In this case, we find that the robustness of the metapopulation is increased only if the probability that individuals stray is low, and is substantially eroded if the probability that individuals err is high.

% We then expand this approach by introducing density-dependent straying, which is assumed to occur as a function of collective decision-making, and examine how these dynamics are effected by both the rate at which individuals stray $m_0$ as well as the effects of group size on straying $C$.

% We have shown that density-dependent straying between populations consistent with collective navigation, coupled with localized selection against donor phenotypes, has a large and nonlinear impact on dynamic properties that affect metapopulation robustness.
% We measured robustness as:
% 1) the average-CV portfolio effect \citep{Anderson:2013gb,Anonymous:2015gf}, a statistical metric commonly used to assess the buffering capacity of metapopulations, and
% 2) the recovery time, defined here as the time required for the aggregate metapopulation biomass $N_T^*$ to return to its steady-state following an induced disturbance, which is mechanistically linked to persistence \citep{Ovaskainen:2002il}.
% In our eco-evolutionary model of dispersal and natural selection between two populations, we show that these statistical and mechanistic descriptors of metapopulation dynamics and robustness are tightly coupled (figure \ref{fig:PE}d), which is not uncommon for diverse metrics of stability \citep{Donohue:2013iu}.
% Taken as a whole, our results point to an important role of density-dependent straying in the colonization and recovery dynamics within metapopulations, while also underscoring the risk of straying by individuals with maladaptive traits to reduce the productivity of locally adapted stock complexes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% POINT BY POINT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%2) Paragraph on the importance of straying on robustness
Metapopulation robustness was found to depend strongly on the magnitude of straying between sites.
We generally found that metapopulation robustness was highest (as indicated by higher PE and lower recovery times) when straying was at a low-intermediate level. 
% Exceedingly low levels of straying did not permit the metapopulation to buffer itself against variability (figure \ref{fig:pert}a), whereas excess straying led to genetic homogenization and maladaptation, and consequently lower abundances, which conspired to lengthen recovery time (figure \ref{fig:pert}b). 
% Thus, we speculate that low-intermediate levels of straying generally enables populations to solve the ecological and evolutionary challenges posed by disturbance and local adaptation in meta-populations.
A central dynamic of the model is that straying can lead to the emergence of asymmetric alternative stable states, or \emph{migrational meltdown} \citep{Ronce:2001dp}, pushing one of the populations to a dominant, well-adapted, high density, state, and one to a subordinate, maladapted, low density, state.
Although there are subtle differences in our model and the general framework presented in Ref. \citep{Ronce:2001dp}, the general dynamic features are the same if we assume that dispersal is symmetric between sites and density-independent (which occurs when $C\rightarrow\infty$).
The dynamic regimes that emerge from the eco-evolutionary model -- in particular the occurrence of alternative stable state regimes I and II (figure \ref{fig:traj}) -- have large effects on both the portfolio effect and the recovery time following an induced disturbance (figure \ref{fig:pert}).
In general, we find that intermediate straying increases the PE and lowers the time to recovery, particularly in the case of the extinction of the subordinate (low-density) population.
In this case, elevated PE occurs when the system enters either regime I or II (depending on the initial conditions), where one population assumes a subordinate low-density state. 
Given that the time to recovery following near-collapse of both populations increases with straying (figure \ref{fig:pert}b), it would suggest that all but the lowest values of density-independent straying erode robustness, regardless of the increase in PE observed at more intermediate values.

% As straying continues to increase, robustness declines.

% Under the assumption of constant straying, we find that an intermediate degree of straying increases metapopulation robustness, whereas low to intermediate rates of straying appear to have a beneficial effect on transient dynamics, which is measured by determining the time to recovery following an induced disturbance.
% When there is just enough straying to cause the system to enter regime I, the portfolio effect at first increases and the recovery time at first declines: signs of increased metapopulation robustness.
% In the alternative stable-state regime, a lower rate of straying inhibits admixture of maladapted individuals.
% Following a large disturbance, such as the near-collapse of both dominant and subordinate populations, this limited mixing increases the growth rates of both populations, permitting faster recovery times. 
% As straying becomes very large large, robustness declines, as observed by a decrease in PE (figure \ref{fig:PE}c) and increase in recovery time (figure \ref{fig:relax}a).
% In this case, increased straying promotes the influx of maladapted individuals into both populations, inhibiting local growth rates, lengthening recovery.

% In regime I (figures \ref{fig:traj}a, \ref{bifurcations}), there exists only one configuration for alternative stable states: dominant and subordinate.
% By contrast, in regime II, there exist two configurations: a symmetric intermediate density, or the asymmetric dominant/subordinate state.
% These general dynamic features are exactly those observed for the eco-evolutionary model investigated by Ronce and Kirkpatrick \citep{Ronce:2001dp}, where dispersal is assumed to be symmetric between sites and constant over time.
% An important aspect of our eco-evolutionary model is that there are similar forces that dictate interactions within and between sites, and this naturally results in a symmetry that could limit the relevance of our findings for natural (and inherently less symmetric) systems.

%3) Paragraph on the importance of collective straying rates on robustness
This themed issue formalizes the role of collective movement in the ecology of natural systems and illuminates a signature of collective navigation in animal populations on the move.
Here we explore the implications of this collective navigation on metapopulations.
% When straying is dynamic, straying at the stable state is determined by a) the probability that individuals err $m_0$, and b) the influence that group size has on straying at the level of the population $C$ (small values of $C$ mean that the influence of collective behaviour on reducing straying is large; Eq. \ref{eq:ddm}).
We highlight three important findings that contribute to our understanding of collective movement suggesting that density-dependent straying may play an important role in the persistence of metapopulations over evolutionary time.
% We will first examine the potentially unrealistic case of very strong collective behaviour, and then discuss the more complicated, but more realistic case of intermediate collective behaviour.

First, if the effects of collective behaviour are very strong (low $C$), metapopulation robustness is increased, due primarily to the avoidance of alternative stable state regime II (figures \ref{fig:pert},\ref{fig:bifurcationsddm}a).
This means that - despite potentially high individual error rates - group formation minimizes straying. 
This occurs when groups of $\leq 10^2$ individuals significantly minimizes straying, which is likely unrealistic.
Moreover, when the effects of collective behaviour are strong, regime II gives way to the dominance of regime I, which harbors low-density basins of attraction (dashed lines in figure \ref{fig:traj}) that can effectively trap disturbed populations in a subordinate stable state, not unlike the allee effects observed in the collective migration model explored by Berdahl et al. \citep{Berdahl:2016dx}.
% The existence of these hidden basins of attraction could present large challenges to the management of populations in these regimes.

Our second important finding reveals that when the effects of collective behaviour are intermediate, metapopulation robustness is impacted in three ways, depending on the magnitude of individual straying.
% Empirical studies suggest that collective behaviour occurs when group sizes are around 100-1000 (REFSXX).
Here, the system is generally in alternative stable state regime I or II except for perhaps unrealistically low levels of individual straying (figure \ref{fig:bifurcationsddm}a).
If individual straying is high ($m_0>0.25$),
1) there is a magnified difference between the numerical densities of the subordinate and dominant populations, effectively pushing the subordinate population to lower steady state densities (figure \ref{fig:cb}),
2) the portfolio effect is low, such that the coefficient of variance for the aggregate metapopulation biomass is on-par with the coefficient of variance of its constituent populations, and
3) more time is generally required for the population(s) to recover following an induced disturbance, and this is particularly true for the recovery of the system following near-collapse of both populations (figure \ref{fig:pert}b).
Together, this suggests that when the effects of collective behaviour are intermediate, and straying is high, there is an overall reduction in metapopulation robustness, thereby reducing persistence.

% The decline in metapopulation robustness only holds if individual straying is very high.
Empirical observations of straying support low-intermediate levels of individual error rates in most species \citep{H:2013fs,Keefer:2014gg}.
If $m_0$ is low ($m_0<0.25$),
1) the differences between the numerical densities of the subordinate and dominant populations are less exaggerated, and the range of $m_0$ that avoids the alternative stable state regime II is larger (figures \ref{fig:cb}, \ref{fig:bifurcationsddm}a),
2) the portfolio effect is exaggerated, meaning that the variance of the combined metapopulation is dampening variance observed in the constituent populations (figure \ref{fig:pert}a), and
3) the time required for the population(s) to recover following an induced disturbance is lower (figure \ref{fig:pert}b).
Interestingly, the largest portfolio effects are observed when straying is just large enough to enter regime II, where one population assumes a subordinate state.
This does not appear, in fact, to be a robust condition because the system relies to a large extent on the dominant population as the source, whereas the subordinate population assumes the role of a sink.
However, recovery time was measured with respect to the aggregate biomass of the metapopulation ($N_T = N_i + N_j$), and despite the source-sink dynamics that emerge in regime II, the aggregate biomass of the system recovers more quickly in this region following the near-collapse of both populations (figure \ref{fig:mpert}b).
If the perspective is shifted to the robustness of the aggregate biomass of the metapopulation, the existence of asymmetric dominant/subordinate alternative stable states could in one sense be considered robust, despite the fact that one population is always at greater risk of stochastic extinction.

% 
% if collective behaviour is of intermediate strength, the differences between the stable state densities are exaggerated, effectively pushing the subordinate population closer to extinction (figure \ref{fig:cb}).
% Moreover, although robustness generally declines with increased individual straying (quantified by a decrease in PE and an increase in recovery time), it remains elevated iff collective behaviour is of intermediate strength (figure \ref{pert}), but this effect is eroded as $m_0 \rightarrow 0.5$.
% 
% %implications of single extinctions
% Third, we find that the alternative stable state regime II expands across a larger range of $m_0$ as the effects of collective behaviour grow.
% In regime II, alternative stable states are generally avoided, however there exist hidden low-density basins of attraction, which may trap a population in the wake of a large disturbance.
% We observe this by instituting the extinction of a single population.
% If the system is outside of regime II, the population will rebound to its previous state as it is colonized by the surviving population.
% If the system is within regime II, the recovering population will become trapped in the subordinate basin of attraction, and will not recover its previous density.
% So although strong collective behaviour can provide benefits to a metapopulation by either
% 1) avoiding the alternative stable state regime I, or
% 2) increasing robustness for higher individual straying rates,
% there is an unavoidable cost, which lies in the expanded parameter space dominated by these hidden low-density basins of attraction.

%4) Paragraph on habitat heterogeneity and extent of local adaptation on robustness
Third, we find that greater habitat heterogeneity increases the role of alternative stable state regimes, particularly when the effects of collective behaviour are strong (high $\Delta\theta$, low $C$; figure \ref{fig:bifurcationsddm}c), and this increases the potential complexity of metapopulation dynamics.
Salmon are distributed and stray across a diverse range of habitats, and the rates of straying between geographically diverse sites can be plastic and idiosyncratic \citep{Westley:2015to}.
Our surrogate measure for habitat heterogeneity is the difference in trait optima between sites $\Delta\theta$.
We show that as habitat heterogeneity increases, the occurrence of alternative stable states associated with regime II becomes unavoidable, particularly for $0.1 \leq m_0 \leq 0.4$, and regime I is minimized (figure \ref{fig:bifurcationsddm}).
This may be particularly consequential for populations that are spatially adjacent but separated by sharp environmental boundaries, such that trait optima are divergent yet dispersal is relatively high.
Such a scenario plays out repeatedly in the context of wild and hatchery-produced salmon. 
Although wild and hatchery populations may occur close on the landscape, and indeed often are sympatric within the same river network, the selective environments to which they are locally adapted differ dramatically \citep{Christie:2012bj}. 
Straying of domesticated hatchery-produced fish from release sites and spawning in the wild drastically reduces the productivity of wild populations through competition and outbreeding depression \citep{Chilcote:2003bb,Araki:2007cm}.

%mtheta
In other cases, habitats that are closer in space can be assumed to have greater similarity in environmental conditions than those that are geographically distant, and phenotypes of more proximately located populations should be more similar \citep{Reisenbichler:1988ex,Fraser:2011co,Westley:2012ui}.
It is thus reasonable to expect a larger number of straying individuals between sites that are geographically proximate and indeed evidence corroborates this prediction \citep{Candy:2000hu,JPE:JPE1383}.
Alternatively, salmon that cue to specific environmental conditions may be more likely to stray into sites that are structurally and physiognamically more similar \citep{Peterson:2014gy}.
These considerations justify imposing a negative correlation between habitat heterogeneity and individual straying $m_0$: as site heterogeneity increases, so too should straying decrease (figure \ref{fig:mtheta}, inset).
When habitat heterogeneity and straying are linked in this way, we show that very small amounts of straying gives rise to regime II, and that regime I occurs for higher values of $m_0$ (figure \ref{fig:mtheta}).
This pattern is opposite that observed for scenarios where habitat heterogeneity and straying are assumed to be independent, and suggests that increases in straying that are associated with growing similarities between habitats can push a metapopulation into a regime where hidden low-density basins of attraction exist.
Thus, management activities that alter dispersal rates by outplanting individuals or reconnecting disconnected habitats could have unintended eco-evolutionary consequences \citep{Anderson:2013bf,Pess:2014isa}, introducing particular management challenges.

%%%%%%%%%%%%%%%%%%%%%
% SUMMING UP
%%%%%%%%%%%%%%%%%%%%%
% [this part still needs work - something about greater complexity with low-intermediate straying, intermediate strength of collective behaviour and high habitat heterogeneity]

A general message from our theoretical framework is that the emergence of alternative eco-evolutionary states depends jointly on the strength of collective behaviour and individual straying, and that this has large implications for metapopulation robustness.
Although robustness is in many cases aided by increasing the strength of collective behaviour, the greater role of both alternative stable state regimes I and II portends additional complexity in eco-evolutionary dynamics, and this could serve to hinder effective management.
Moreover, this increased complexity at empirically observed levels of straying \citep{H:2013fs} and at realistic (intermediate) ranges of the strength of collective behaviour, is only magnified with increasing habitat heterogeneity and when heterogeneity itself is linked to individual straying.
An additional component that we have not explored here, but that may be particularly relevant to consider, are the effects of including additional sites, as well as the patterns of dispersal that connect these sites.
The structure of dispersal has been shown to have a large influence on population dynamics \citep{Heino:1997vf,Carrara:2014cy,Yeakel:2013vz,Gilarranz:2017fs}, and to what extent density-dependent straying influences the eco-evolutionary dynamics of populations in large spatially-structure networks is of considerable interest.
We are hopeful that these predictions may inspire future theoretical and empirical studies that aim to expand upon the relationships that we have explored.

While our theoretical work is a vast simplification of natural systems, our results have important potential implications for understanding the eco-evolutionary dynamics of spatially structured populations such as in salmon. 
A particularly salient finding was that density-dependent straying may serve to promote or inhibit population robustness, depending on the strength of the collective behaviour, and the underlying magnitude of straying.
When the strength of collective behaviour is at plausible intermediate levels, straying increased population robustness through enhanced portfolio effects and reduced time to recovery following disturbance. 
Salmon have evolved within the context of dynamic geomorphic landscapes where habitat quantity and quality shifts as a mosaic through time \citep{Stanford:2017bu}. 
Our results provide evidence supporting the hypothesis described in Berdahl et al. \citep{Berdahl:2014bl} that collective behaviour may underpin rapid habitat colonization following natural disturbance such as volcanic eruptions \citep{Leider:1989gx}, reconnected habitats following restoration \citep{Pess:2012by,Pess:2014isa}, or in the context of glacial retreat and climate warming \citep{Milner:2008gb}. 
Moreover, our results are consistent with the role of collective behaviour in facilitating reproductive isolation and local adaptation to site-specific selection in populations that recover following disturbance to the extent that straying decreases as population sizes increase. 
Additionally, collective behaviour may be beneficial in facilitating navigation though increasingly modified and fragmented habitats \citep{Berdahl:2017uu}. 
On the other hand, collective behaviour coupled with high straying may push populations to extirpation. 
Thus, collective behaviour could provide both resilience to salmon metapopulations but also vulnerabilities.

Our study broadly indicates that management activities that alter patterns of straying could have profound implications for metapopulation robustness and adaptive potential. 
High rates of straying are predicted to decrease metapopulation robustness, and there are a series of common practices in salmon management that may be elevating straying rates \citep{Ford:2002ip,Brenner:2012gl}.
For example, transporting young salmon downstream via barge to avoid infrastructure (e.g., dams) or dangerous habitats may disrupt the processes involved with critical periods of imprinting prior to or during downstream migration by sea-going individuals and increase straying by adults later in life. 
Our results support the conservation concern that large-scale releases of salmon produced in hatcheries that stray could have profound impacts on the robustness of salmon metapopulations through the erosion of portfolio effects. 
Moreover, hatchery environments are associated with marked changes in fish social behaviour that may increase collective dynamics of migrating groups \citep{Ruzzante:1994bu}, consistent with the findings of Jonsson and Jonsson \citep{Jonsson:2011kg} who report stronger associations between straying and abundance in escaped aquaculture produced Atlantic salmon than their wild counterparts. 
Thus, management activities that have the unintended consequence of altering straying may compromise recovery efforts.

%Paragraph for beyond salmon
% Although our study was inspired by salmon metapopulations, the results have general implications for the conservation and management of other migratory metapopulations as well. Human activities that influence dispersal among populations will influence the eco-evolutionary dynamics of the metapopulation and thus its robustness. For example, human infrastructure such as roads can act as barriers to dispersal. 
% Alternatively, human activities may increase dispersal. 
% Furthermore, if dispersal is density-dependent (REFS), then human activities such as harvest that reduce abundances could dramatically increase straying rates. Harvest may be pushing metapopulations towards higher straying rates that swamp local adaptations and decrease population productivities. 
% Thus, there is building appreciation (REFS) that dispersal is a key parameter in the conservation and management of metapopulations.

% Although our study was inspired by salmon metapopulations, the results have general implications for the conservation and management of other migratory metapopulations as well. 
% Because changes in straying can have large and nonlinear impacts on robustness, human activities that alter straying could have unintended consequences. 
% For example, salmon produced by hatcheries often stray into proximate wild populations \citep{Brenner:2012gl}, and these recipient populations can have lower fitness due in part to the introduction of maladapted genes \citep{Ford:2002ip}. 
% We show that an intermediate individual straying rate can result in faster recovery times following a large disturbance, but that this is largely determined by the strength of collective behaviour.
% This finding suggests that salmon stocking efforts that aim to lower recovery times following dam removal could actually prolong recovery if the rate at which individuals are introduced and the suitability of those fish in that habitat (i.e. their measure of pre-adaptation) is not taken into account.
% Ongoing examinations of experimental restocking in the recently re-opened Elwha River (Washington State) will provide empirical insight into the potential short- and long-term consequences of facilitated recovery \citep{Liermann:2017gj}.

%Concluding
Beyond salmon, density-dependent dispersal, whether it is caused by collective decision-making or other factors, has a large influence on the dynamics of populations in the presence of local adaptation.
The rate at which individual err, and the influence of group size on navigation at the population level, are two important components of dynamic dispersal \citep{Berdahl:2016dx}.
We show that changes in these characteristics can alter the occurrence and positioning of two different alternative stable state regimes, one of which may harbor hidden low-density basins of attractions that can effectively trap populations after a large disturbance.
Generally, increasing the strength of collective behaviour mitigates the potentially negative impacts of so-called \emph{migrational meltdown} \citep{Ronce:2001dp}.
Thus, preserving the biological processes that facilitate collective behaviour of migratory species may be an important conservation target in its own right, echoing the sentiments of Hardesty-Moore et al. \citep{HardestyMoore:wg}. 
We suggest that increased understanding the proximate and ultimate factors governing dispersal among local populations within metapopulations, across heterogeneous environments, in tandem with the mosaic of selective forces acting on those environments, may be key to promoting persistence in the wild.
\\ \\
% The portfolio effect and the time to recovery following a disturbance are independent and correlated measures of metapopulation robustness that take into account both steady-state and transient dynamics.
% We show that these measures of robustness are strongly influenced by the rate at which individuals from donor populations stray into habitats occupied by recipient populations. 
% Importantly, density-dependent straying, which may occur when individuals collectively navigate, can both increase the portfolio effect and lower the time to recovery following a disturbance, which is anticipated to promote persistence. 
% Therefore, preserving the biological processes that facilitate this collective behaviour may be an important conservation target in its own right, echoing the sentiments of  Hardesty-Moore et al. \citep{HardestyMoore:wg}. 
% We suggest that understanding the spatial complexity of metapopulations dispersing across heterogeneous environments, in tandem with the mosaic of selective forces acting on those environments, may be key to uncovering those factors that promote persistence in the wild.

\noindent {\bf Competing interests:} The authors declare no competing interests
\\
\noindent {\bf Author contributions:} JDY and JWM conceived of the initial project design. JDY and JPG designed the modeling framework and conducted the analyses. JDY, JPG, PAHW, and JWM interpreted the results, and drafted and wrote the manuscript.
\\
\noindent {\bf Data Accessibility:} Code is made available at https://github.com/jdyeakel/SalmonStrays
\\
\noindent {\bf Acknowledgements:} We thank Sean Anderson for helpful discussions and comments on the manuscript. We also thank the guest editors Andrew Berdahl, Dora Biro, and Colin Torney, for inviting us to contribute to this themed issue, and two anonymous reviewers for their insightful comments and critiques. J.D.Y. was supported by startup funds at the University of California, Merced and an Omidyar Postdoctoral Fellowship at the Santa Fe Institute. J.P.G. was supported by a James S. McDonnell Foundation Postdoctoral Fellowship in Complex Systems at the University of California, Merced. T.G. was supported by the EPSRC (EP/N034384/1). P.A.H.W. was supported by the UA Foundation at the University of Alaska Fairbanks. J.W.M. was supported by the Liber Ero Research Chair in Coastal Science and Management at Simon Fraser University.

% \bibliography{aa_kevin}

\begin{thebibliography}{10}
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
  \urlstyle{rm}\Url}\fi

\bibitem{Brown:1977gk}
Brown JH, Kodric-Brown A, 1977 {Turnover rates in insular biogeography: effect
  of immigration on extinction}.
\newblock \emph{Ecology} \textbf{58}, 445--449

\bibitem{Earn:2000fm}
Earn DJD, Levin SA, Rohani P, 2000 {Coherence and conservation}.
\newblock \emph{Science} \textbf{290}, 1360--1364

\bibitem{Anonymous:2015gf}
Schindler DE, Armstrong JB, Reed TE, 2015 {The portfolio concept in ecology and
  evolution}.
\newblock \emph{Front. Ecol. Environ.} \textbf{13}, 257--263

\bibitem{Abbott:2017hl}
Abbott RE, Doak DF, Peterson ML, 2017 {Portfolio effects, climate change, and
  the persistence of small populations: analyses on the rare plant Saussurea
  weberi.}
\newblock \emph{Ecology} \textbf{98}, 1071--1081

\bibitem{Thibaut:2012km}
Thibaut LM, Connolly SR, 2013 {Understanding diversity-stability relationships:
  towards a unified model of portfolio effects.}
\newblock \emph{Ecol. Lett.} \textbf{16}, 140--150

\bibitem{Carlson:2011ce}
Carlson SM, Satterthwaite WH, 2011 {Weakened portfolio effect in a collapsed
  salmon population complex}.
\newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{68}, 1579--1589

\bibitem{MilnerGulland:2011vm}
Milner-Gulland EJ, Fryxell JM, Sinclair ARE, 2011 \emph{{Animal Migration}}.
\newblock Oxford: Oxford University Press

\bibitem{Bell:2011ki}
Bell G, Gonzalez A, 2011 {Adaptation and evolutionary rescue in metapopulations
  experiencing environmental deterioration}.
\newblock \emph{Science} \textbf{332}, 1327--1330

\bibitem{Carlson:2014is}
Carlson SM, Cunningham CJ, Westley PAH, 2014 {Evolutionary rescue in a changing
  world}.
\newblock \emph{Trends Ecol. Evol.} \textbf{29}, 521--530

\bibitem{Satterthwaite:2015ge}
Satterthwaite WH, Carlson SM, Bradbury I, 2015 {Weakening portfolio effect
  strength in a hatchery-supplemented Chinook salmon population complex}.
\newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{72}, 1860--1875

\bibitem{Ronce:2001dp}
Ronce O, Kirkpatrick M, 2001 {When sources become sinks: migrational meltdown
  in heterogeneous habitats}.
\newblock \emph{Evolution} \textbf{55}, 1520

\bibitem{Muhlfeld:2014hs}
Muhlfeld CC, Kovach RP, Jones LA, Al-Chokhachy R, Boyer MC, Leary RF, Lowe WH,
  Luikart G, Allendorf FW, 2014 {Invasive hybridization in a threatened species
  is accelerated by climate change}.
\newblock \emph{Nature Climate Change} \textbf{4}, 620--624

\bibitem{Lowe:2015ft}
Lowe WH, Muhlfeld CC, Allendorf FW, 2015 {Spatial sorting promotes the spread
  of maladaptive hybridization}.
\newblock \emph{Trends Ecol. Evol.} \textbf{30}, 456--462

\bibitem{Travis:2013en}
Travis JMJ, \emph{et~al.}, 2013 {Dispersal and species{\textquoteright}
  responses to climate change}.
\newblock \emph{Oikos} \textbf{122}, 1532--1540

\bibitem{H:2013fs}
Westley PAH, Quinn TP, Dittman AH, 2013 {Rates of straying by hatchery-produced
  Pacific salmon (\emph{Oncorhynchus} spp.) and steelhead (\emph{Oncorhynchus
  mykiss}) differ among species, life history types, and populations}.
\newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{70}, 735--746

\bibitem{Keefer:2014gg}
Keefer ML, Caudill CC, 2014 {Homing and straying by anadromous salmonids: a
  review of mechanisms and rates}.
\newblock \emph{Rev Fish Biol Fisher} \textbf{24}, 333--368

\bibitem{Bett:2017ha}
Bett NN, Hinch SG, Burnett NJ, Donaldson MR, Naman SM, 2017 {Causes and
  consequences of straying into small populations of Pacific salmon}.
\newblock \emph{Fisheries} \textbf{42}, 220--230

\bibitem{Berdahl:2015kv}
Berdahl A, Torney CJ, Schertzer E, Levin SA, 2015 {On the evolutionary
  interplay between dispersal and local adaptation in heterogeneous
  environments}.
\newblock \emph{Evolution} \textbf{69}, 1390--1405

\bibitem{Berdahl:2016dx}
Berdahl A, 2016 {Collective behavior as a driver of critical transitions in
  migratory populations}.
\newblock \emph{Movement Ecology} \textbf{4}, 1--12

\bibitem{HardestyMoore:wg}
Hardesty-Moore M, \emph{et~al.}, 2017 {Migration in the Anthropocene}.
\newblock \emph{Philos. T. Roy. Soc. B} \textbf{This volume}

\bibitem{Berdahl:2017uu}
Berdahl A, Kao A, Flack J, Westley PAH, Codling E, Couzin ID, Dell AI, Biro D,
  2017 {Collective animal navigation and migratory culture: from theoretical
  models to empirical evidence.}
\newblock \emph{Philos. T. Roy. Soc. B} \textbf{This volume}

\bibitem{Simons:2004jo}
Simons AM, 2004 {Many wrongs: The advantage of group navigation}.
\newblock \emph{Trends Ecol. Evol.} \textbf{19}, 453--455

\bibitem{Berdahl:2014bl}
Berdahl A, Westley PAH, Levin SA, Couzin ID, Quinn TP, 2014 {A collective
  navigation hypothesis for homeward migration in anadromous salmonids}.
\newblock \emph{Fish Fish.} \textbf{17}, 525--542

\bibitem{Brenner:2012gl}
Brenner RE, Moffitt SD, Grant WS, 2012 {Straying of hatchery salmon in Prince
  William Sound, Alaska}.
\newblock \emph{Environ Biol Fish} \textbf{94}, 179--195

\bibitem{Johnson:2012fe}
Johnson RC, Weber PK, Wikert JD, Workman ML, MacFarlane RB, Grove MJ, Schmitt
  AK, 2012 {Managed metapopulations: do salmon hatchery
  {\textquoteleft}sources{\textquoteright} lead to in-river
  {\textquoteleft}sinks{\textquoteright} in conservation?}
\newblock \emph{PLoS ONE} \textbf{7}, e28880--11

\bibitem{Fullerton:2011ii}
Fullerton AH, Lindley ST, Pess GR, Feist BE, Steel EA, Mcelhany P, 2011 {Human
  influence on the spatial structure of threatened Pacific salmon
  metapopulations}.
\newblock \emph{Conserv Biol} \textbf{25}, 932--944

\bibitem{Quinn:2011tf}
Quinn TP, 2011 \emph{{The Behavior and Ecology of Pacific Salmon and Trout}}.
\newblock Vancouver: UBC Press

\bibitem{Jonsson:2011kg}
Jonsson B, Jonsson N, 2011 \emph{{Ecology of Atlantic Salmon and Brown Trout}}.
\newblock Dordrecht: Springer Netherlands

\bibitem{Quinn:1993ge}
Quinn TP, 1993 {A review of homing and straying of wild and hatchery-produced
  salmon}.
\newblock \emph{Fish Res} \textbf{18}, 29--44

\bibitem{Hendry:2004wf}
Hendry AP, Bohlin T, Jonsson B, Berg OK, 2004 {The evolution of philopatry and
  dispersal: homing versus straying in salmonids}.
\newblock In AP~Hendry, SC~Stearns, eds., \emph{Evolution Illuminated}. Oxford:
  Oxford University Press on Demand

\bibitem{Schtickzelle:2007wb}
Schtickzelle N, Quinn TP, 2007 {A metapopulation perspective for salmon and
  other anadromous fish}.
\newblock \emph{Fish Fish.} \textbf{8}, 297--314

\bibitem{Anderson:2014cx}
Anderson SC, Moore JW, McClure MM, Dulvy NK, Cooper AB, 2015 {Portfolio
  conservation of metapopulations under climate change}.
\newblock \emph{Ecol. Appl.} \textbf{25}, 559--572

\bibitem{Keefer:2008bs}
Keefer ML, Caudill CC, Peery CA, Lee SR, 2008 {Transporting juvenile salmonids
  around dams impairs adult migration}.
\newblock \emph{Ecol. Appl.} \textbf{18}, 1888--1900

\bibitem{Westley:2015to}
Westley PAH, Dittman AH, Ward EJ, Quinn TP, 2015 {Signals of climate,
  conspecific density, and watershed features in patterns of homing and
  dispersal by Pacific salmon.}
\newblock \emph{Ecology} \textbf{96}, 2823--2833

\bibitem{Bond:2016dz}
Bond MH, Westley PAH, Dittman AH, Holecek D, Marsh T, Quinn TP, 2016 {Combined
  effects of barge transportation, river environment, and rearing location on
  straying and migration of adult snake river fall-run Chinook salmon}.
\newblock \emph{T Am Fish Soc} \textbf{146}, 60--73

\bibitem{Peterson:2014gy}
Peterson DA, Hilborn R, Hauser L, 2014 {Local adaptation limits lifetime
  reproductive success of dispersers in a wild salmon metapopulation}.
\newblock \emph{Nat Commun} \textbf{5}, 3696

\bibitem{Moore:2010gs}
Moore JW, McClure M, Rogers LA, Schindler DE, 2010 {Synchronization and
  portfolio performance of threatened salmon}.
\newblock \emph{Conserv. Lett.} \textbf{3}, 340--348

\bibitem{Braun:2016ib}
Braun DC, Moore JW, Candy J, Bailey RE, 2016 {Population diversity in salmon:
  linkages among response, genetic and life history diversity}.
\newblock \emph{Ecography} \textbf{39}, 317--328

\bibitem{simpson1953major}
Simpson GG, 1953 \emph{{The Major Features of Evolution}}.
\newblock New York: Simon and Schuster

\bibitem{Lande:1976ga}
Lande R, 1976 {Natural selection and random genetic drift in phenotypic
  evolution}.
\newblock \emph{Evolution} \textbf{30}, 314--334

\bibitem{Reisenbichler:1988ex}
Reisenbichler RR, 1988 {Relation between Distance Transferred from Natal Stream
  and Recovery Rate for Hatchery Coho Salmon}.
\newblock \emph{N. Am. J. Fish. Manage.} \textbf{8}, 172--174

\bibitem{Shelton:2011eq}
Shelton AO, Mangel M, 2011 {Fluctuations of fish populations and the magnifying
  effects of fishing.}
\newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{108}, 7075--7080

\bibitem{Jasper:2013cc}
Jasper JR, Habicht C, Moffitt S, Brenner R, Marsh J, Lewis B, Fox EC, Grauvogel
  Z, Olive SDR, Grant WS, 2013 {Source-sink estimates of genetic introgression
  show influence of hatchery strays on wild chum salmon populations in Prince
  William Sound, Alaska}.
\newblock \emph{PLoS ONE} \textbf{8}, e81916

\bibitem{Schreiber:2011wx}
Schreiber SJ, B{\"u}rger R, Bolnick DI, 2011 {The community effects of
  phenotypic and genetic variation within a predator population.}
\newblock \emph{Ecology} \textbf{92}, 1582--1593

\bibitem{Gilbert:2014ee}
Gibert JP, Brassil CE, 2014 {Individual phenotypic variation reduces
  interaction strengths in a consumer{\textendash}resource system}.
\newblock \emph{Ecol Evol} \textbf{4}, 3703--3713

\bibitem{Gibert:2015kc}
Gibert JP, DeLong JP, 2015 {Individual variation decreases interference
  competition but increases species persistence}.
\newblock \emph{Adv Ecol Res} \textbf{52}, 45--64

\bibitem{Carlson:2008hl}
Carlson SM, Seamons TR, 2008 {A review of quantitative genetic components of
  fitness in salmonids: implications for adaptation to future change}.
\newblock \emph{Evol Appl} \textbf{1}, 222--238

\bibitem{Schindler:2015gf}
Schindler DE, Armstrong JB, Reed TE, 2015 {The portfolio concept in ecology and
  evolution}.
\newblock \emph{Front. Ecol. Environ.} \textbf{13}, 257--263

\bibitem{Ovaskainen:2002il}
Ovaskainen O, Hanski I, 2002 {Transient dynamics in metapopulation response to
  perturbation}.
\newblock \emph{Theor Popul Biol} \textbf{61}, 285--295

\bibitem{Anderson:2013gb}
Anderson SC, Cooper AB, Dulvy NK, 2013 {Ecological prophets: quantifying
  metapopulation portfolio effects}.
\newblock \emph{Methods Ecol Evol} \textbf{4}, 971--981

\bibitem{Loreau:2008ju}
Loreau M, de~Mazancourt C, 2008 {Species synchrony and its drivers: neutral and
  nonneutral community dynamics in fluctuating environments}.
\newblock \emph{Am. Nat.} \textbf{172}, E48--E66

\bibitem{Yeakel:2013vz}
Yeakel JD, Moore JW, Guimar{\~a}es~Jr PR, de~Aguiar MAM, 2014 {Synchronisation
  and stability in river metapopulation networks}.
\newblock \emph{Ecol. Lett.} \textbf{17}, 273--283

\bibitem{GuckHolmes}
Guckenheimer J, Holmes P, 1983 \emph{{Nonlinear Oscillations, Dynamical
  Systems, and Bifurcations of Vector Fields}}.
\newblock New York: Springer

\bibitem{Scheffer:2009gg}
Scheffer M, Bascompte J, Brock WA, Brovkin V, Carpenter SR, Dakos V, Held H,
  van Nes EH, Rietkerk M, Sugihara G, 2009 {Early-warning signals for critical
  transitions}.
\newblock \emph{Nature} \textbf{461}, 53--59

\bibitem{Lade:2012eu}
Lade SJ, Gross T, 2012 {Early warning signals for critical transitions: A
  generalized modeling approach}.
\newblock \emph{PLoS Comp. Biol.} \textbf{8}, e1002360

\bibitem{Anonymous:2013br}
Boettiger C, Ross N, Hastings A, 2013 {Early warning signals: the charted and
  uncharted territories}.
\newblock \emph{Theor. Ecol.} \textbf{6}, 255--264

\bibitem{Dakos:2014br}
Dakos V, Bascompte J, 2014 {Critical slowing down as early warning for the
  onset of collapse in mutualistic communities}.
\newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{111}, 17546--17551

\bibitem{Krkosek:2014ch}
Krko{\v s}ek M, Drake JM, 2014 {On signals of phase transitions in salmon
  population dynamics}.
\newblock \emph{Proc. Roy. Soc. B} \textbf{281}, 20133221

\bibitem{Lin:2008kl}
Lin J, Quinn TP, Hilborn R, Hauser L, 2008 {Fine-scale differentiation between
  sockeye salmon ecotypes and the effect of phenotype on straying}.
\newblock \emph{Heredity} \textbf{101}, 341--350

\bibitem{Donohue:2013iu}
Donohue I, Petchey OL, Montoya JM, Jackson AL, McNally L, Viana M, Healy K,
  Lurgi M, O'Connor NE, Emmerson MC, 2013 {On the dimensionality of ecological
  stability.}
\newblock \emph{Ecol. Lett.} \textbf{16}, 421--429

\bibitem{Christie:2012bj}
Christie MR, Marine ML, French RA, Blouin MS, 2012 {Genetic adaptation to
  captivity can occur in a single generation.}
\newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{109}, 238--242

\bibitem{Chilcote:2003bb}
Chilcote MW, 2003 {Relationship between natural productivity and the frequency
  of wild fish in mixed spawning populations of wild and hatchery steelhead
  (\emph{Oncorhynchus mykiss})}.
\newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{60}, 1057--1067

\bibitem{Araki:2007cm}
Araki H, Cooper B, Blouin MS, 2007 {Genetic effects of captive breeding cause a
  rapid, cumulative fitness decline in the wild}.
\newblock \emph{Science} \textbf{318}, 100--103

\bibitem{Fraser:2011co}
Fraser DJ, Weir LK, Bernatchez L, Hansen MM, Taylor EB, 2011 {Extent and scale
  of local adaptation in salmonid fishes: review and meta-analysis.}
\newblock \emph{Heredity} \textbf{106}, 404--420

\bibitem{Westley:2012ui}
Westley PAH, { Corinne M Conway}, Fleming IA, 2012 {Phenotypic divergence of
  exotic fish populations is shaped by spatial proximity and habitat
  differences across an invaded landscape}.
\newblock \emph{Evol. Ecol. Res.} \textbf{14}, 147--167

\bibitem{Candy:2000hu}
Candy JR, Beacham TD, 2000 {Patterns of homing and straying in southern British
  Columbia coded-wire tagged Chinook salmon (\emph{Oncorhynchus tshawytscha})
  populations}.
\newblock \emph{Fish Res} \textbf{47}, 41--56

\bibitem{JPE:JPE1383}
Schick RS, Lindley ST, 2007 {Directed connectivity among fish populations in a
  riverine network}.
\newblock \emph{J. Appl. Ecol.} \textbf{44}, 1116--1126

\bibitem{Anderson:2013bf}
Anderson JH, Faulds PL, Atlas WI, Quinn TP, 2013 {Reproductive success of
  captively bred and naturally spawned Chinook salmon colonizing newly
  accessible habitat}.
\newblock \emph{Evol Appl} \textbf{6}, 165--179

\bibitem{Pess:2014isa}
Pess GR, Quinn TP, Gephard SR, Saunders R, 2014 {Re-colonization of Atlantic
  and Pacific rivers by anadromous fishes: linkages between life history and
  the benefits of barrier removal}.
\newblock \emph{Rev Fish Biol Fisher} \textbf{24}, 881--900

\bibitem{Heino:1997vf}
Heino M, Kaitala V, Ranta E, Lindstrom J, 1997 {Synchronous dynamics and rates
  of extinction in spatially structured populations}.
\newblock \emph{Proc. Roy. Soc. B} \textbf{264}, 481--486

\bibitem{Carrara:2014cy}
Carrara F, Rinaldo A, Giometto A, Altermatt F, 2014 {Complex Interaction of
  Dendritic Connectivity and Hierarchical Patch Size on Biodiversity in
  River-Like Landscapes}.
\newblock \emph{Am. Nat.} \textbf{183}, 13--25

\bibitem{Gilarranz:2017fs}
Gilarranz LJ, Rayfield B, Li{\~n}{\'a}n-Cembrano G, Bascompte J, Gonzalez A,
  2017 {Effects of network modularity on the spread of perturbation impact in
  experimental metapopulations}.
\newblock \emph{Science} \textbf{357}, 199--201

\bibitem{Stanford:2017bu}
Stanford JA, Lorang MS, Hauer FR, 2017 {The shifting habitat mosaic of river
  ecosystems}.
\newblock \emph{Verh. Internat. Verein. Limnol.} \textbf{29}, 123--136

\bibitem{Leider:1989gx}
Leider SA, 1989 {Increased straying by adult steelhead trout, \emph{Salmo
  gairdneri}, following the 1980 eruption of Mount St. Helens}.
\newblock \emph{Environ Biol Fish} \textbf{24}, 219--229

\bibitem{Pess:2012by}
Pess GR, Hilborn R, Kloehn K, Quinn TP, 2012 {The influence of population
  dynamics and environmental conditions on pink salmon (\emph{Oncorhynchus
  gorbuscha}) recolonization after barrier removal in the Fraser River, British
  Columbia, Canada}.
\newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{69}, 970--982

\bibitem{Milner:2008gb}
Milner AM, Robertson AL, Monaghan KA, Veal AJ, Flory EA, 2008 {Colonization and
  development of an Alaskan stream community over 28 years}.
\newblock \emph{Front. Ecol. Environ.} \textbf{6}, 413--419

\bibitem{Ford:2002ip}
Ford MJ, 2002 {Selection in captivity during supportive breeding may reduce
  fitness in the wild}.
\newblock \emph{Conserv Biol} \textbf{16}, 815--825

\bibitem{Ruzzante:1994bu}
Ruzzante DE, 1994 {Domestication effects on aggressive and schooling behavior
  in fish}.
\newblock \emph{Aquaculture} \textbf{120}, 1--24

\end{thebibliography}

\clearpage

\begin{table}[!t]
\begin{center}
\begin{tabular}{ l l l }
\hline
Parameter & Definition & Value/Range \\
\hline
$N_i(t)$, $N_T(t)$ & Individual, aggregate population over time & dyn.\\
$x_i$ & Trait value for an individual in population $i$ & dyn.\\
$\mu_i(t)$ & Mean of $x$ for population $i$ over time & dyn.\\
$m$; $m(t)$ & Constant, density-dependent straying & $(0,0.5)$; dyn.\\
$m_0$ & Individual straying probability & $(0,0.5)$\\
$C$ & Strength of collective behavior (low=strong) & $(10^1,10^5)$\\
$r_{\rm max}$ & Maximum recruitment rate & 2.0 \\
$\beta$ & Strength of density dependence & $10^{-3}$\\
$\theta_i$ & Optimal trait value for habitat $i$ & 5.0\\
$\sigma^2$ & Genetic variance of trait $x$ & 1.0\\
$\tau$ & Strength of selection & 1.0\\
$h^2$ & Heritability & 0.2\\
$\Delta\theta$ & Habitat heterogeneity & 2.0\\
$\epsilon$ & Sensitivity of $m_0$ to changes in $\Delta\theta$ & 20.0\\
${\rm PE}$ & Portfolio Effect & $\geq1$\\
$T$ & Terminal simulation time & $10^5$\\
\hline
\end{tabular}
\end{center}
\caption{Table of parameters, definitions, and assigned values or ranges.}
\end{table}

% tmax=10000;
% z=0.5;
% rmax=2.0;
% beta=0.001;
% theta1=5.0;
% thetadiff=5.0;
% tau=1.0;
% h=0.49;
% sigmaE=0;
% sigmaG=1.0;
% m=0.35;
% perror=0.01;

\clearpage

\begin{figure}
  \captionsetup{justification=raggedright,
singlelinecheck=false
}
\centering
\includegraphics[width=0.75\textwidth]{fig_traj.pdf}
\caption{
The steady-state densities of $N_i$ and $N_j$ vs. straying $m$ for the constant straying model. 
Alternative stable states exist for regimes I and II, labeled RI and RII, respectively.
Two configurations exist in regime I: a symmetric, intermediate state (solid line), and an asymmetric dominant/subordinate state (red and blue dashed lines, respectively).
One configuration exists in regime II: an asymmetric dominant/subordinate state (red and blue lines, respectively).
Inset: An illustration of the alternative stable states (solid lines) in regime I (light gray area) and II (dark gray area).
The dividing line that separates initial conditions that approach the symmetric (sym.) and asymmetric (asym.) solution is a separatrix (dashed line) that runs through an unstable fixed point.
} \label{fig:traj}
\end{figure}

\begin{figure*}
  \captionsetup{justification=raggedright,
singlelinecheck=false
}
\centering
\includegraphics[width=1\textwidth]{fig_mpert.pdf}
\caption{
Measures of metapopulation robustness for the constant straying model as a function of straying $m$.
Alternative stable state regimes I and II corresponding to those in figure \ref{fig:traj} are labeled RI and RII, respectively.
(a) Portfolio effect as a function of $m$.
(b) Recovery time as a function of $m$.
Measures of metapopulation robustness are shown with respect to different induced disturbances: the near-collapse of both populations (black), and the lone extinction of either the dominant (dark gray) or subordinate (light gray) population.
Portfolio effects are different for the near-collapse and single extinction scenarios due to one population becoming trapped in a low-density basin of attraction in the alternative stable state regime I.
} \label{fig:mpert}
\end{figure*}

\begin{figure*}
  \captionsetup{justification=raggedright,
singlelinecheck=false
}
\centering
\includegraphics[width=0.75\textwidth]{fig_meandiff.pdf}
\caption{
Comparison of steady state population densities for the constant straying model and density-dependent straying model.
Inset: stable state densities for the constant straying model (purple) and density-dependent straying model (green) for different strengths of collective behaviour.
Low $C$ corresponds to strong effects of collective behaviour.
The top row shows stable state densities as a function of individual straying $m_0$; the bottom row shows stable state densities as a function of straying at the stable state $m(t)^*$
Main: The absolute difference in stable state densities averaged across intervals of low straying ($0 < m,m_0 < 0.25$; blue) and high straying ($0.25 < m,m_0 < 0.5$; red).
} \label{fig:cb}
\end{figure*}

\begin{figure*}
  \captionsetup{justification=raggedright,
singlelinecheck=false
}
\centering
\includegraphics[width=0.75\textwidth]{fig_rtpe_ddm.pdf}
\caption{
Measures of metapopulation robustness for the density-dependent straying model as a function of individual straying $m_0$ and the strength of collective behaviour $C$ (note the $\rm log_10$ scale, including 
a) the Portfolio effect,
b) the time to recovery following near-collapse of both populations,
c) the time to recovery following the extinction of the subordinate population, and
d) the time to recovery following the extinction of the dominant population.
} \label{fig:pert}
\end{figure*}

\begin{figure*}
  \captionsetup{justification=raggedright,
singlelinecheck=false
}
\centering
\includegraphics[width=0.75\textwidth]{fig_hysteresis_ddm.pdf}
\caption{
Alternative stable state regimes I (gray) and II (black) as a function of individual straying $m_0$ and the strength of collective behaviour $C$ (note the $\rm log_10$ scale).
Regime I signifies parameter space where there is either 1) an intermediate-density, symmetric, steady state, or 2) an asymmetric dominant/subordinate density.
Regime II signifies parameter space where there is an asymmetric dominant/subordinate steady state density.
The white space to the left (lower values of $m_0$) signifies high-density, symmetric, stable states, and the white space to the right (higher values of $m_0$) signifies low-density, symmetric, steady states.
Relationships are shown for (a) intermediate habitat heterogeneity ($\Delta\theta$), as well as (b) low habitat heterogeneity and (c) high heterogeneity.
} \label{fig:bifurcationsddm}
\end{figure*}

\begin{figure*}
  \captionsetup{justification=raggedright,
singlelinecheck=false
}
\centering
\includegraphics[width=0.5\textwidth]{fig_hysteresis_ddm_mtheta.pdf}
\caption{
Alternative stable state regimes I (gray) and II (black) as a function of individual straying $m_0$ and the strength of collective behaviour $C$ (note the $\rm log_10$ scale), for the case where individual straying increases with lower habitat heterogeneity (inset).
Regime I signifies parameter space where there is either 1) an intermediate-density, symmetric, steady state, or 2) an asymmetric dominant/subordinate density.
Regime II signifies parameter space where there is an asymmetric dominant/subordinate steady state density.
} \label{fig:mtheta}
\end{figure*}

%%%%%%%%%%%%%
% SUPPLEMENTARY FIGURES
%%%%%%%%%%%%%

\beginsupplement

\begin{figure}
  \captionsetup{justification=raggedright,
singlelinecheck=false
}
\centering
% \includegraphics[width=0.35\textwidth]{fig_recovery.pdf}
\caption{
Example of the numerical procedure used to estimate recovery time. After a disturbance is introduced, the recovery time is calculated by measuring the point in time where $N_T$ (in black), which is the aggregate of both populations (blue, red), settles to within one standard deviation of the new equilibrium $N_T^*$. 
} \label{fig:recovery}
\end{figure}

\begin{figure}
  \captionsetup{justification=raggedright,
singlelinecheck=false
}
\centering
% \includegraphics[width=0.35\textwidth]{fig_density.png}
\caption{
Steady-state densities of both populations as a function of $m$: one in a dominant state and one in a subordinate state.
Steady-states for populations with symmetrical values ($\alpha=0$) in the vital rates $r_{\rm max}$ and $\beta$ are shown with cool tones.
As the asymmetry among populations between sites increases ($\alpha>0$), their vital rates diverge, such that the maximal growth at sites 1 and 2 is now $r_{\rm max,1}=(1 + \alpha)r_{\rm max,2}$ and $\beta_1=(1+\alpha)\beta_2$ where $\alpha$ is increased from $0$ to $0.1$, thereby increasing the asymmetry in vital rates.
Steady-states for populations with increasingly asymmetric values ($\alpha\rightarrow 0.1$) for vital rates $r_{\rm max}$ and $\beta$ are shown in warmer tones.
The lower-density values that appear between $m=0.06$ and $m=0.1$ represent populations that have been trapped in the low-density basins of attraction associated with regime I.
Increasing the asymmetry in vital rates does not impact the qualitative nature of the dynamics.
} \label{fig:symmetry}
\end{figure}

\begin{figure}
  \captionsetup{justification=raggedright,
singlelinecheck=false
}
\centering
% \includegraphics[width=0.35\textwidth]{fig_pevsrt.pdf}
\caption{
Comparison of portfolio effects vs. recovery time following the near-collapse of both populations.
} \label{fig:pevsrt}
\end{figure}

\end{document}

### 2017-12-19 Manuscript/FinalDraft3/supp2.tex
final draft submission to PhilTrans

--- 2017-11-17 manuscript/FinalDraft3/supp.tex
@@
 \usepackage{xr}
-\externaldocument[extfig-]{msv6}
+\externaldocument[extfig-]{msv8}
 %\usepackage{epstopdf}
@@
 given $\mu^\prime = \omega_i \mu_i(t)+ (1-\omega_i)\mu_j(t)$.
-Under the condition that $\sigma=1$ and $\tau = 1$ as is set in the manuscript, this simplifies to
-\begin{equation}
-  \mu_i(t+1) = \omega_i\mu_i(t) + (1-\omega_i)\mu_j(t) + h^2\left(\theta_i - \omega_i\mu_i - (1-\omega_i)\mu_j\right).
-\end{equation}
 
@@
 \section*{II. Additional Figures}
-
-% \begin{figure}
-%   \captionsetup{justification=raggedright,
-% singlelinecheck=false
-% }
-% \centering
-% \includegraphics[width=0.7\textwidth]{fig_contSD.pdf}
-% \caption{
-% Magnitude of difference in the standard deviation for the calculated PDF for individuals in population $i$ after 1) immigration from population $j$, 2) recruitment, and 3) adult mortality vs.
-the Gaussian approximation where $\sigma=1$.
-% Values $<1$ reflect underestimates of variability; values $>1$ reflect overestimates.
-% Top row is for constant straying ratio, and the bottom row is for density-dependent straying ratio under different treatments of habitat heterogeneity (left column $\Delta \theta=5$; right column $(\Delta \theta=8$).
-% } \label{fig:contourSD}
-% \end{figure}
-%
-
-\begin{figure}[h]
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.35\textwidth]{fig_mthetarelation.pdf}
-\caption{
-In some cases, habitat heterogeneity may be assumed to determine the rate of straying, if for example:
-1) sites are distributed over greater spatial distances, where habitat differences are assumed to be greater between more distant sites, or 2) individuals have behaviors promoting dispersal between habitats that are more similar.
-To examine such cases, we use the relationship $m= 0.5(1 + \Delta\theta)^{-1}$ where maximum straying is assumed to occur at $m=0.5$ (perfect mixing).
-} \label{fig:mthetarelation}
-\end{figure}
 
@@
 \centering
-\includegraphics[width=0.35\textwidth]{fig_hysteresis.pdf}
+\includegraphics[width=0.9\textwidth]{fig_relax_lowh.pdf}
 \caption{
-Increasing the straying rate results in the transition from a single steady-state for both populations to a dominant and subordinate states.
-If the straying rate is subsequently lowered, the single steady-state is not easily obtained, which is the hallmark of hysteresis.
-} \label{fig:hysteresis}
-\end{figure}
+Recovery time of $N_T^*$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady-state $m^*$) with trait heritability $h^2=0.2$.
+If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
+DCB marks the discrete cusp bifurcation.
+} \label{fig:relax}
+\end{figure*}
 
-\begin{figure}
+\begin{figure*}
   \captionsetup{justification=raggedright,
@@
 \centering
-\includegraphics[width=0.35\textwidth]{fig_relax_large.pdf}
+\includegraphics[width=0.9\textwidth]{fig_rtpe_ddmsl.pdf}
 \caption{
-Extinction of high-density population with a high straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{extfig-fig:relax}a).
-Black line marks the calculated point of recovery post-perturbation.
-Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
-} \label{fig:relaxtraj_hdlh}
-\end{figure}
+} \label{fig:pertlh}
+\end{figure*}
 
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.35\textwidth]{fig_relax_small.pdf}
-\caption{
-Extinction of low-density population with a high constant straying rate $m=0.4$ and low trait heritability $h^2=0.2$ (see figure \ref{extfig-fig:relax}a).
-Black line marks the calculated point of recovery post-perturbation.
-Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
-} \label{fig:relaxtraj_ldlh}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.9\textwidth]{fig_relax_highh.pdf}
-\caption{
-Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady-state $m^*$) with trait heritability $h^2=0.8$.
-If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
-} \label{fig:relax_highh}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.4\textwidth]{fig_thetadiffN.pdf}
-\caption{
-Median difference in population densities taken over the straying rate as a function of habitat heterogeneity $\Delta\theta$.
-Solid lines are for constant $m$; dashed lines are for density-dependent $m$.} \label{fig:thetadiffN}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.8\textwidth]{fig_relaxtheta2.pdf}
-\caption{
-(a) Recovery time after near collapse of both populations as a function of straying rate $m$ and habitat heterogeneity $\Delta\theta$.
-(b) The same as (a) but including recovery times when straying is density-dependent evaluated at $m^*$, shown by linked point pairs.
-Recovery times for systems with density-dependent straying are longer when straying is low and shorter when straying is high, mirroring the change in portfolio effects with respect to density-dependent straying shown in figure \ref{extfig-fig:thetaPE}.
-} \label{fig:relaxtheta}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-  \centering
-  \includegraphics[width=0.35\textwidth]{fig_MDPE_hm_mtheta_rt.pdf}
-  \caption{
-  Portfolio effects as a function of straying rate $m$ and trait heritability $h^2$ when the rate of straying is $m = 0.5(1 + \Delta\theta)^{-1}$.
-Alternative steady-states emerge for low values of $m$ (left of the cusp bifurcation, denoted by the black line), whereas a single steady-state exists for high $m$.
-  } \label{fig:mthetaPE}
-\end{figure}
-
-% \clearpage
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.35\textwidth]{fig_relax_inertia.pdf}
-\caption{
-Population inversion where increased differences in trait optima between sites $\Delta\theta$ corresponds to lower rates of straying $m$.
-At low rates of straying $m=0.02$ ($\Delta\theta=24$), extinction of the dominant population leads to slower-than-expected recovery times because the subordinate population is isolated enough to evolve towards its own trait optimum.
-%until growth of the dominant population overwhelms this local selection.
-In this case, $m$ is less than $m=0.034$ (denoted by the asterisk in figure \ref{extfig-fig:mtheta}), such that isolation allows the subordinate population to `run away' from the influence of the dominant population.
-This leads to a switch in subordinate/dominant states for the two populations.
-If $m$ is low but greater than $0.034$, isolation permits the subordinate population to `run away' from the influence of the dominant population, until it is overwhelmed by the recovering dominant population, and reverts back to its previous trait mean prior to disturbance.
-} \label{fig:inertia}
-\end{figure}
-
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-  \centering
-  \includegraphics[width=0.8\textwidth]{fig_relax_mtheta.pdf}
-  \caption{
-  Recovery times for three disturbance types when the straying rate covaries with habitat heterogeneity as $m = 0.5(1 + \Delta\theta)^{-1}$ for constant (a) and density-dependent (b) straying rates.
-  The cusp bifurcation is not as clear in (b) because $\Delta\theta$ is a function of the individual straying rate $m_0$, whereas the x-axis in (b) is the straying rate at the steady-state $m^*$.
-  Despite this difference, the general behavior shown in (a) are the same in (b).
-  } \label{fig:mthetamvm}
-\end{figure}
+% 
+% \begin{figure}
+%   \captionsetup{justification=raggedright,
+% singlelinecheck=false
+% }
+% \centering
+% \includegraphics[width=0.35\textwidth]{fig_hysteresis.pdf}
+% \caption{
+% Increasing the straying rate results in the transition from a single steady-state for both populations to a dominant and subordinate states.
+If the straying rate is subsequently lowered, the single steady-state is not easily obtained, which is the hallmark of hysteresis.
+% } \label{fig:hysteresis}
+% \end{figure}
+% 
+%
 
@@
 
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.35\textwidth]{fig_relax_both_lowh.pdf}
-\caption{
-Near collapse of both populations with a low straying rate $m=0.1$ and low trait heritability $h^2=0.2$ (see figure \ref{extfig-fig:relax}a).
-Black line marks the calculated point of recovery post-perturbation.
-Trait optima are $\theta_1 = 10$ (blue population trajectory) and $\theta_2 = 5$ (red population).
-} \label{fig:relaxtraj_bothlh}
-\end{figure}
-
 \end{document}

### 2017-12-19 Manuscript/FinalDraft3/supp3.tex
final draft submission to PhilTrans

--- 2017-12-19 Manuscript/FinalDraft3/supp2.tex
@@
-\RequirePackage{lineno}
 \documentclass{revtex4}
-%\usepackage[pdftex]{graphicx}
 
-\usepackage{amsmath,amsfonts,amssymb}
+%% Language and font encodings
 \usepackage[english]{babel}
-\usepackage[latin1]{inputenc}
+\usepackage[utf8x]{inputenc}
 \usepackage[T1]{fontenc}
-\usepackage{color}
-\usepackage{float}
-\usepackage{verbatim}
+
+%% Sets page size and margins
+\usepackage[a4paper,top=3cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}
+
+%% Useful packages
+\usepackage{amsmath}
+\usepackage{graphicx}
+\usepackage[colorinlistoftodos]{todonotes}
+\usepackage[colorlinks=true, allcolors=blue]{hyperref}
+
 \usepackage{graphicx}
@@
 \captionsetup{compatibility=false}
-\usepackage{xr}
-\externaldocument[extfig-]{msv8}
-%\usepackage{epstopdf}
-%\usepackage{array}
-%\usepackage{tabularx}
-%\usepackage{multirow}
-\usepackage{color}
-%\usepackage{multibox}
-%\usepackage{rotating}
-% \usepackage{lineno}
-
-%\usepackage[left]{lineno}
-%\usepackage[comma,sort&compress]{natbib}
-%\usepackage{authblk}
-%\usepackage{multicol}
-
-\linespread{1.5}
-
-\usepackage{natbib}
-\bibpunct{(}{)}{,}{a}{}{;}
-
-% \bibliographystyle{prsb}
-
-% \usepackage{bibunits}
-
-\newcommand{\beginsupplement}{%
-        \clearpage
-        \setcounter{table}{0}
-        \renewcommand{\thetable}{S\arabic{table}}%
-        \setcounter{figure}{0}
-        \renewcommand{\thefigure}{S\arabic{figure}}%
-     }
-
-% \pagewiselinenumbers
-\linenumbers
-% \setlength\linenumbersep{3pt}
 
 \begin{document}
-
-% \Blindtext
-%\title{Simple rules yield complex communities: deconstructed species interactions and the assembly of communities}
-%\title{Community assembly and dynamics by the deconstruction of species interactions}
-\title{Supplementary Materials: Eco-evolutionary dynamics and collective dispersal: implications for salmon metapopulation robustness}
-\author{
-Justin D. Yeakel${}^{1,2,*}$, Jean P. Gibert${}^{1}$, Peter A. H. Westley${}^{3}$, \& Jonathan W. Moore${}^{4}$ \\
-${}^1$School of Natural Sciences, University of California, Merced, Merced CA, USA \\
-${}^2$The Santa Fe Institute, Santa Fe NM, USA \\
-${}^3$College of Fisheries and Ocean Sciences, University of Alaska, Fairbanks, Fairbanks AK, USA \\
-${}^4$Earth${}_2$Oceans Research Group, Simon Fraser University, Vancouver BC, Canada \\
-${}^*$To whom correspondence should be addressed: jdyeakel@gmail.com
-}
-
+\title{Supplementary Materials: Eco-evolutionary dynamics, density-dependent dispersal, and collective behavior: implications for salmon metapopulation robustness}
+% \author{
+% Justin D. Yeakel${}^{1,2,*}$, Jean P. Gibert${}^{1}$, Peter A. H. Westley${}^{3}$, \& Jonathan W. Moore${}^{4}$ \\
+% ${}^1$School of Natural Sciences, University of California, Merced, Merced CA, USA \\
+% ${}^2$The Santa Fe Institute, Santa Fe NM, USA \\
+% ${}^3$College of Fisheries and Ocean Sciences, University of Alaska, Fairbanks, Fairbanks AK, USA \\
+% ${}^4$Earth${}_2$Oceans Research Group, Simon Fraser University, Vancouver BC, Canada \\
+% ${}^*$To whom correspondence should be addressed: jdyeakel@gmail.com
+% }
 \maketitle
 
-\beginsupplement
-
 We aim to understand how straying between populations may affect phenotypic evolution, and how in turn this evolutionary change may lead to changes in ecological dynamics as they unfold. 
-To do so, we build upon ideas from classic papers (Lande 1976, most notably). 
-In what follows, we show 
-1) how the application of Lande's equation for the case of two populations linked by dispersal leads to complex equations of trait evolution that are not easily modeled,
-2) that this complex trait distribution can be approximated as a single Gaussian distribution, and
-3) under what conditions this approximation is appropriate.
+To do so, we build upon Lande's classic formulation of phenotypic evolution \citep{Lande:1976ga}. In what follows, we  
+1) re-derive Lande's phenotypic evolution equation for one population to introduce the reader to the main concepts and assumptions of the framework used, and,
+2) explain how we used this framework for the case of two populations connected by dispersal.
 
@@
 \end{equation}
-Then, we can assess how the change in the phenotype $x$ occurs over time following
+Then, we can assess how the change in the phenotype $x$ occurs over time using
 \begin{equation}\label{4}
@@
 which making use of equations \ref{1} and \ref{3}, becomes $\frac{\partial \overline{W}}{\partial \overline{x}}=
-\frac{\overline{W}}{\sigma^{2}}\overline{x}_{w}-\frac{\overline{x}}{\sigma^{2}}\overline{W}$, which can be factored as $\frac{\partial \overline{W}}{\partial \overline{x}}=
+\frac{\overline{W}}{\sigma^{2}}\overline{x}_{w}-\frac{\overline{x}}{\sigma^{2}}\overline{W}$, and can be factored as $\frac{\partial \overline{W}}{\partial \overline{x}}=
 \frac{\overline{W}}{\sigma^{2}}(\overline{x}_{w}-\overline{x})$.
-Using \ref{5}, we can rearrange this expression to obtain an equation that related the change in the average phenotype from one time step to the next to the amount of heritable variation, $\sigma^{2}h^{2}$, and the adaptive landscape $\frac{\partial \overline{W}}{\partial \overline{x}}$:
+Using \ref{5}, we can rearrange this expression to obtain an equation that relates the change in the average phenotype from one time step to the next to the amount of heritable variation, $\sigma^{2}h^{2}$, and the adaptive landscape $\frac{\partial \overline{W}}{\partial \overline{x}}$:
 \begin{equation}\label{10}
@@
 \end{equation}
+Equation \ref{11} thus assumes the existence of one population with a normally distributed trait in which selection occurs before reproduction, the probability of surviving to reproduce depends on the value of the trait different individuals have, and that the trait is normally distributed after reproduction.
+\\
 
@@
 We will then 1) show why it is not necessarily useful to use the obtained expression in a simulation context, and 2) introduce an approximation that retains the features important for exploring how trait and population dynamics impact each other in the case of dispersing populations.
-In Section II, we will explore how well the approximation reproduces the underlying trait dynamics.
-Specifically, we will discuss in what cases it is exact and for which cases the approximation becomes problematic, such that it should be avoided.
 
@@
 Moreover, this equation only holds for the first time step, and as new dispersing individuals arrive and the trait distribution is again updated by the influx of individuals from the other site, the mix becomes a mix of mixes, such that the analytical expressions presented above no longer hold.
+\\
 
-This presents a tradeoff: we can either track changes in phenotype exactly for a single generation, or we can track changes in phenotype across multiple generations using an approximation.
+\textbf{The Gaussian approximation used in the main text}
+The above issue presents us with a tradeoff: we can either track changes in phenotype exactly for a single generation, or we can track changes in phenotype across multiple generations using an approximation.
 Given that our central goal is to gain an understanding of the feedback between evolutionary and ecological processes in salmon populations with dispersal in explicit space, we simplify the complexities created by phenotypic evolution by assuming that, as new migrants arrive, the mixture distribution can be locally approximated by a Gaussian distribution with a mean equal to that of the mix.
-Below, we show why this approximation is appropriate for describing the phenotypic distributions under most conditions relevant to our questions.
-Moreover, based on this assumption, we are able to use Lande's equation (here equation \ref{10}), without violating the original assumptions that led to its current form.
+Based on this assumption, we can directly use Lande's equation (here equation \ref{10}), without violating the original assumptions that led to its current form.
 
-\section*{II. Gaussian approximation for the mixed trait distribution}
-We assume a Gaussian distribution to approximate the more complex trait distribution that results when two populations with two distinct trait distributions are linked by gene flow and reproduction.
-Here and throughout, we refer to the resident population as $i$ and the connected population as $j$, though both populations stray simultaneously.
-The Gaussian approximation is assumed to have a mean that is equal to the mean of the mixed trait distribution $\omega_i\mu_i + (1-\omega_i)\mu_j$ with a constant variance $\sigma^2$, equal to the original trait variance prior to mixing.
-We employ this approximation because a mixed trait distribution that remixes and recombines (through reproduction) at each time-step becomes complicated very quickly, as we shall demonstrate from a mechanistic perspective.
-\emph{Importantly, because the evolutionary framework tracks only the mean and not the variability through time, this approximation is exact with respect to the first moment.}
-Accordingly, the only attribute that is approximate is the assumption of normality.
-Below we will show that a constant variance of $\sigma^2$ for the mixed trait distribution is a good approximation in most relevant cases, and becomes increasingly accurate when 1) generations become less overlapping (i.e.
-the case of semelparity), and 2) with density-dependent straying ratios.
+The model in the main text thus assumes the existence of a normally distributed trait in each population with means $\mu_{i}$ and $\mu_{j}$.
+Then straying occurs, and the the populations now have a mixture distribution as in \ref{12}.
+To keep track of the mean of that mix, we assume the mixture distribution to be Gaussian with mean $\overline{x}=\omega_i \mu_{i}+(1-\omega_i)\mu_{j}$, which is true whenever straying is low enough, or the two means are not too different from one another.
+Then selection and reproduction occur at the same time leading to a new generation with a normally distributed trait distribution whose.
 
-{\bf The trait distribution at $\bm{t=0}$.} At time $t=0$, it is assumed that individuals from two separate populations have not yet mixed. 
-Thus, the mean trait values for both populations are equal to the optimal trait values that accord to their respective habitats $\mu_i(t=0) = \theta_i$, and that their trait values are distributed as $X_i=x_i \sim g(x_i,\mu_i,\sigma) = \mathcal{N}(\mu_i,\sigma)$.
-Here and henceforth we use upper-case to denote random variables, and lower-case to denote specific values of random variables.
+In a previous paper by Ronce and Kirkpatrick \citep{Ronce:2001dp}, it was shown that assumptions similar to the ones we make do not qualitatively alter the ecological or evolutionary dynamics observed.
 
-{\bf The trait distribution at $\bm{t=1}$.} At time $t=1$, dispersal of individuals from site $j$ to site $i$, and vice versa, has begun.
-The proportion of individuals that disperse to and from each population is given by the stray ratio $m$.
-The calculation of the mixed distribution that results from dispersal and subsequent recruitment incorporates the following processes:
-\emph{1)} dispersal of strays into the local population from the remote population, and vice versa;
-\emph{2)} recruitment of juveniles that are produced from this mixed distribution (under the assumption of random mating); 
-\emph{3)} mortality of individuals in the adult class.
-These events occur in the order in which they are listed.
+% \section*{II. Additional Figures}
+\bibliographystyle{prsb}
+% \bibliography{aa_kevin}
 
-The mixed trait distribution for site $i$ immediately following dispersal is simply a mixed Gaussian with weights determined by the proportion of the population that are residents $i$ and the proportion that are strays $j$, such that
-\begin{equation}
-  p(x_{\rm disperse}) = \omega_i g(x_i,\mu_i,\sigma) + (1-\omega_i)g(x_j,\mu_j,\sigma),
-\end{equation}
-where the proportion of the population $i$ composed of \emph{resident} individuals is
-\begin{equation}
-  \omega_i = \frac{(1-m)N_i}{(1-m)N_i + mN_j},
-\end{equation}
-and the proportion of population $i$ composed of strays is $(1-\omega_i)$.
-Once individuals in both habitats mix, they mate without preference.
-Recruits are assumed to have trait values intermediate to their parents such that $x_{\rm r} = (x_{1} + x_{2})/2$ where trait values $X_1=x_{1}\sim p(x_{\rm disperse})$ and $X_2=x_{2}\sim p(x_{\rm disperse})$.
-We thus assume that parental traits are drawn independently from the mixed distribution $p(x_{\rm disperse})$, conditional on the requirement that $x_2 = 2x_r - x_1$, which we get when rearranging the equation for $x_{\rm r}$.
-Incorporating this condition and that traits are independently drawn from the mix of residents and strays yields the probability density for the trait value of recruits,
-\begin{equation}
-  q(x_{\rm r}) = \int_{-\infty}^\infty 2p(x_1)p(2x_{\rm r}-x_1){\rm d}x_1.
-\end{equation}
-The mean of the recruit trait distribution is calculated to be
-\begin{equation}
-{\rm E}\{q(x_{\rm r})\}=\omega_i\mu_i + (1-\omega_i)\mu_j,
-\end{equation} 
-and the variance is
-\begin{equation}
-  {\rm Var}\{q(x_{\rm r})\} = \frac{m^2 N_j^2 \left(4 \mu_j^2+\sigma ^2\right)+(m-1)^2 N_i^2 \left(4 \mu_i^2+\sigma ^2\right)-(m-1) m N_i N_j \left(\mu_i^2+6 \mu_i \mu_j+\mu_j^2+2 \sigma ^2\right)}{2 (-m N_i+m N_j+N_i)^2}.
-\end{equation}
+\begin{thebibliography}{1}
+\expandafter\ifx\csname urlstyle\endcsname\relax
+  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
+  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
+  \urlstyle{rm}\Url}\fi
 
-The relative contribution of trait values from resident parents, stray parents, and recruits depends on the numbers of each within the mixed population.
-The number of residents is $(1-m)N_i$, and the number of strays is $mN_j$.
-However, because adult mortality occurs after reproduction, we must adjust these numbers to account for only those who survive, which is given by ${\rm e}^{-Z}$.
-As $Z\rightarrow \infty$, the population becomes semelparous, as the adult class gives way to new recruits at each timestep. 
-Accordingly, the number of adult residents and strays is given by $(1-m)N_i{\rm e}^{-Z}$ and  $mN_i{\rm e}^{-Z}$, respectively.
-In contrast, the number of recruits is given by the combined contribution of resident and straying individuals to recruit production as 
-\begin{equation}
-N_r = \int_{-\infty}^{\infty}r_{\rm max}{\rm e}^{\frac{(x_{\rm r}-\theta_i)^2}{2\tau^2}}q(x_{\rm r})dx_{\rm r}((1-m)N_i + mN_j){\rm e}^{(-\beta((1-m)N_i + mN_j))},
-\end{equation}
-where $\theta_i$ is the optimal trait value in site $i$, $r_{\rm max}$ is the maximum recruitment rate, and $\tau$ is the strength of selection.
+\bibitem{Falconer:1967tg}
+Falconer DS, 1967 \emph{{Introduction to Quantitative Genetics}}.
+\newblock Edinburgh: Oliver and Boyd LTD
 
-\begin{figure}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.7\textwidth]{fig_PDF.pdf}
-\caption{
-Clockwise from the upper-left.
-The calculated probability distribution function (PDF) for individuals in population $i$ after immigration from population $j$, recruitment, and mortality (blue) vs.
-the Gaussian approximation used for the dynamical system (green), with
-a) low stray ratio, large straying population, overlapping generations: $m=0.1$, $Z=0.5$;
-b) maximal stray ratio, overlapping generations: $m=0.5$, $Z=0.5$;
-c) maximal stray ratio, non-overlapping generations: $m=0.5$, $Z=1000$;
-d) density-dependent stray ratio, non-overlapping generations: $m=m_0(1-N_i/(C+N_i))$, $Z=1000$, where $m_0=0.2$ and $C=1000$.
-The mean and variability of both distributions is shown by a point (mean) and line ($\sigma$) at the base of each distribution.
-} \label{fig:PDF}
-\end{figure}
+\bibitem{Lande:1976ga}
+Lande R, 1976 {Natural selection and random genetic drift in phenotypic
+  evolution}.
+\newblock \emph{Evolution} \textbf{30}, 314--334
 
-The proportion of individuals from the resident, straying, and recruitment classes following mortality at site $i$ are then $c_i$, $c_j$, and $c_r$ respectively, where
-\begin{align}
-c_i &= \frac{(1-m)N_i{\rm e}^{-Z}}{(1-m)N_i{\rm e}^{-Z} + m N_j{\rm e}^{-Z} + N_r} \\ \nonumber
-c_j &= \frac{m N_j{\rm e}^{-Z}}{(1-m)N_i{\rm e}^{-Z} + m N_j{\rm e}^{-Z} + N_r} \\ \nonumber
-c_r &= \frac{N_r}{(1-m)N_i{\rm e}^{-Z} + m N_j{\rm e}^{-Z} + N_r}.
-\end{align}
-where $c_i + c_j + c_r= 1$.
-The trait value distribution for time $t=1$ is then calculated as a mix of trait distributions from 1) surviving residents, 2) surviving strays, and 3) their offspring, such that
-\begin{equation}
-  g_{{\rm next}}(x_i) = c_i g(x_i) + c_j g(x_j) + c_r q(x_r).
-\end{equation}
-The expectation of this distribution is
-\begin{equation}
-{\rm E}\{g_{\rm next}(x_i)\}=\omega_i\mu_i + (1-\omega_i)\mu_j.
-\end{equation}
+\bibitem{Ronce:2001dp}
+Ronce O, Kirkpatrick M, 2001 {When sources become sinks: migrational meltdown
+  in heterogeneous habitats}.
+\newblock \emph{Evolution} \textbf{55}, 1520
 
-% variance a that has a closed form solution, but that is not easily readable.
-% \begin{align}
-%   {\rm Var}\{g_{{\rm next}}(x_i)\}=&\frac{(\mu_i-\mu_j)^2 (2 (m-1)^2 N_i^2 c_j^2-(m-1) N_i c_j (2 (m-1) N_i+m N_j (1-4 c_i))}{2 (-m N_i+m N_j+N_i)^2} \\ \nonumber
-%   &+ \frac{m N_j (c_i-1) (-m N_i+2 m N_j c_i+N_i))-\sigma ^2 (c_i+c_j+1) (-m N_i+m N_j+N_i)^2}{2 (-m N_i+m N_j+N_i)^2}.
-% \end{align}
-
-Importantly, ${\rm E}\{g_{{\rm next}}(x_i)\}$ is exactly equal to the trait mean for the original mix, as well as the trait mean for recruits.
-Thus, a Gaussian approximation distributed as $\mathcal{N}(\omega_i\mu_i + (1-\omega_i)\mu_j,\sigma)$ will not alter the dynamics under the assumption of constant variance in Eq 4.
-Moreover, the mixed distribution described above becomes more complex at the next timestep, as two multi-modal trait distributions then mix and recombine through recruitment.
-One could make the argument that a Gaussian approximation does not represent the multi-modality in trait values inherent to a mixed distribution, and thus underestimates variance.
-Although the Gaussian approximation exactly captures the expectation, we next explore to what extent the approximation does and does not capture the variability expressed in $g(x_{\rm next})$.
-
-%Discussion of factors that increase or decrease accuracy of the approximation
-The Gaussian approximation $\mathcal{N}(\omega_i\mu_i + (1-\omega_i)\mu_j,\sigma)$ will generally underestimate the true variance of the mix, but not by much.
-If the straying ratio is low, (ca. 0.0 to 0.2), the Gaussian approximation is very similar to the actual distribution (Fig. \ref{fig:PDF}a).
-As the straying ratio increases, so does the bimodality of the actual distribution, and this magnifies our underestimate of variance (Fig. \ref{fig:PDF}b).
-The relevant dynamics observed in our model generally occur between $m=0.05$ and $m=0.2$, where estimates of variance are similar.
-Decreasing the overlap between generations also leads to greater similarity in the actual distribution and the Gaussian approximation, and the potential bias introduced by bimodality will be less for cases of non-overlapping generations, or semelparity (Fig.
-\ref{fig:PDF}c).
-Most importantly, density-dependent straying increases the accuracy of our Gaussian approximation by actively suppressing the straying ratio (Fig. \ref{fig:PDF}d).
-
-Because the Gaussian approximation is exact to first order, and generally provides a good approximation of realized variance, we suggest that the use of Lande's formulation for the change in mean trait values through time under the assumption of constant variance is appropriate.
-Because our estimate of variance is generally an underestimate, the rate of recruitment will be a slight overestimate, and the timescale of trait evolution will generally be slower.
-Changes in the rate of recruitment (Fig.
-\ref{fig:symmetry}) and heritability do not have an impact on our qualitative findings, and this suggests any bias that our approximation introduces by underestimating true variance is minimal.
-
-Under these assumptions, the governing equations for the dynamic system are then written as
-
-\begin{align}
-  &N_i(t+1) = \\ \nonumber
-  &\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z} \\ \nonumber
-  &+ R_i[\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\theta_i]\left((1-m)N_i + mN_j\right) \\ \nonumber
-  &\times {\rm e}^{-\beta ((1-m)N_i(t) + m N_j(t))},
-  \label{eq:N}
-\end{align}
-where the trait distribution for recruits is also approximated by the Gaussian 
-${\rm pr}(x_i) \approx \mathcal{N}(\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\sigma^2)$, where the mean is the analytical expectation.
-This gives the rate of recruitment as 
-\begin{align}
-  &R_i[\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\theta_i] = \\ \nonumber
-  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i,\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\sigma^2) {\rm d}x_i +\tilde{P}_i\\ \nonumber
-  &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-(\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t)))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\} +\tilde{P}_i.
-\end{align}
-The mean of the trait distribution then changes as
-\begin{align}
-  \label{eq:mu}
-  \mu_i(t+1) &= \omega_i\mu_i(t) + (1-\omega_i)\mu_j(t) + h^2\sigma^2\frac{\partial}{\partial \mu^\prime}\ln\left(R_i[\mu^\prime,\theta_i] \right), \\ \nonumber
-  &= \omega_i\mu_i(t) + (1-\omega_i)\mu_j(t) + h^2\sigma^2\left(\frac{\theta_i - \omega_i\mu_i - (1-\omega_i)\mu_j}{\sigma^2+\tau^2} \right),
-\end{align}
-given $\mu^\prime = \omega_i \mu_i(t)+ (1-\omega_i)\mu_j(t)$.
-
-\bibliography{aa_kevin}
+\end{thebibliography}
 
 \clearpage
-
-\section*{II. Additional Figures}
 
@@
 Example of the numerical procedure used to estimate recovery time.
-After a disturbance is introduced, the recovery time is calculated by measuring the point in time where $N_T$ (in black), which is the aggregate of both populations (blue, red) settles to within one standard deviation of the new equilibrium $N_T^*$.
+After a disturbance is introduced, the recovery time is calculated by measuring the point in time where $N_T$ (in black), which is the aggregate of both populations (blue, red), settles to within one standard deviation of the new equilibrium $N_T^*$.
 } \label{fig:recovery}
 \end{figure}
-
-% 
-% \begin{figure}
-%   \captionsetup{justification=raggedright,
-% singlelinecheck=false
-% }
-% \centering
-% \includegraphics[width=0.35\textwidth]{fig_hysteresis2Dhys.pdf}
-% \caption{
-% White (low $m$)
-% Gray
-% Black
-% White (high $m$)
-% The white/black region is separated by a sub-critical pitchfork bifurcation ($\rm PB_*$); the gray/black region is separated by a super-critical pitchfork bifurcation ($\rm PB^*$); the white/gray region is separated by a fold bifurcation (FB).
-% } \label{fig:bifurcations}
-% \end{figure}
 
@@
 \centering
-\includegraphics[width=0.35\textwidth]{fig_eigs.pdf}
+\includegraphics[width=0.35\textwidth]{fig_density.png}
 \caption{
-The real parts of the four eigenvalues for the Jacobian matrix of the 4-dimensional system.
-The cusp bifurcation occurs when the dominant eigenvalue crosses the unit circle at $+1$. 
-} \label{fig:eigs}
+Steady-state densities of both populations as a function of $m$: one in a dominant state and one in a subordinate state.
+Steady-states for populations with symmetrical values ($\alpha=0$) in the vital rates $r_{\rm max}$ and $\beta$ are shown with cool tones.
+As the asymmetry among populations between sites increases ($\alpha>0$), their vital rates diverge, such that the maximal growth at sites 1 and 2 is now $r_{\rm max,1}=(1 + \alpha)r_{\rm max,2}$ and $\beta_1=(1+\alpha)\beta_2$ where $\alpha$ is increased from $0$ to $0.1$, thereby increasing the asymmetry in vital rates.
+Steady-states for populations with increasingly asymmetric values ($\alpha\rightarrow 0.1$) for vital rates $r_{\rm max}$ and $\beta$ are shown in warmer tones.
+The lower-density values that appear between $m=0.06$ and $m=0.1$ represent populations that have been trapped in the low-density basins of attraction associated with regime I.
+Increasing the asymmetry in vital rates does not impact the qualitative nature of the dynamics.
+} \label{fig:symmetry}
 \end{figure}
-
-\begin{figure*}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.9\textwidth]{fig_relax_lowh.pdf}
-\caption{
-Recovery time of $N_T^*$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady-state $m^*$) with trait heritability $h^2=0.2$.
-If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.
-DCB marks the discrete cusp bifurcation.
-} \label{fig:relax}
-\end{figure*}
-
-\begin{figure*}
-  \captionsetup{justification=raggedright,
-singlelinecheck=false
-}
-\centering
-\includegraphics[width=0.9\textwidth]{fig_rtpe_ddmsl.pdf}
-\caption{
-} \label{fig:pertlh}
-\end{figure*}
-
-% 
-% \begin{figure}
-%   \captionsetup{justification=raggedright,
-% singlelinecheck=false
-% }
-% \centering
-% \includegraphics[width=0.35\textwidth]{fig_hysteresis.pdf}
-% \caption{
-% Increasing the straying rate results in the transition from a single steady-state for both populations to a dominant and subordinate states.
-If the straying rate is subsequently lowered, the single steady-state is not easily obtained, which is the hallmark of hysteresis.
-% } \label{fig:hysteresis}
-% \end{figure}
-% 
-%
 
@@
 \centering
-\includegraphics[width=0.35\textwidth]{fig_asymdensity.pdf}
+\includegraphics[width=0.35\textwidth]{fig_pevsrt.pdf}
 \caption{
-Steady-state densities of both populations as a function of $m$, where a cusp bifurcation indicates the emergence of alternative steady-states: one in a dominant state and one in a subordinate state.
-Steady-states for populations with symmetrical values ($\alpha=0$) in the vital rates $r_{\rm max}$ and $\beta$ are shown with cool tones.
-As the asymmetry among populations between sites increases ($\alpha>0$), their vital rates diverge, such that the maximal growth at sites 1 and 2 is now $r_{\rm max}(1)=r_{\rm max}(1+\tilde{rv}_1)$ and $r_{\rm max}(2)=r_{\rm max}(1+\tilde{rv}_2)$ where $rv_{1,2}$ are independently drawn from $\rm{Normal}(0,\alpha)$ and $r_{\rm max}=2$.
-Similarly the strength of density dependence is calculated at sites 1 and 2 as $\beta(1)=\beta(1+\tilde{rv}_1)$ and $\beta(2)=\beta(1+\tilde{rv}_2)$ where $\tilde{rv}_{1,2}$ are independently drawn from $\rm{Normal}(0,\alpha)$ and $\beta=0.001$.
-Steady-states for populations with increasingly asymmetric values ($\alpha\rightarrow 0.1$) for vital rates $r_{\rm max}$ and $\beta$ are shown in warmer tones.
-} \label{fig:symmetry}
+Comparison of portfolio effects vs. recovery time following the near-collapse of both populations.
+} \label{fig:pevsrt}
 \end{figure}

### 2017-12-19 model/model2/Figure1.jl

@@ -38,7 +38,7 @@ burnin=0.80
 @time for i=1:length(mvec)
   m=mvec[i];
   
-  n1, n2, x1, x2, w1, w2 = 
+  n1_pre, n2_pre, x1_pre, x2_pre, w1, w2 = 
   KevinEvolve(
     tmax,
     z,
@@ -53,11 +53,25 @@ burnin=0.80
     m,
     perror
     );
-
+    
+    # if n1_pre[tmax-1] < n2_pre[tmax-1]
+        n1 = copy(n2_pre);
+        n2 = copy(n1_pre);
+        x1 = copy(x2_pre);
+        x2 = copy(x1_pre);
+    # else
+    #     n1 = copy(n1_pre);
+    #     n2 = copy(n2_pre);
+    #     x1 = copy(x1_pre);
+    #     x2 = copy(x2_pre);
+    # end
+    
   n1trim = n1[Int64(floor(tmax*burnin)):tmax];
   n2trim = n2[Int64(floor(tmax*burnin)):tmax];
   x1trim = x1[Int64(floor(tmax*burnin)):tmax];
-  x2trim = x2[Int64(floor(tmax*burnin)):tmax]
+  x2trim = x2[Int64(floor(tmax*burnin)):tmax];
+  
+  
   
   n1ts[i,:] = n1;
   n2ts[i,:] = n2;
@@ -92,7 +106,7 @@ burnin=0.80
 end
 diffn = n1mean .- n2mean;
 #Steady state plot
-namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/FinalDraft3/fig_traj.pdf");
+namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/Manuscript/FinalDraft3/fig_traj.pdf");
 R"""
 library(RColorBrewer)
 cols = brewer.pal(4,'Set1')
@@ -100,9 +114,14 @@ pdf($namespace,height=8,width=5)
 par(mfrow=c(2,1),mai = c(0.8, 0.8, 0.1, 0.1))
 plot($mvec,$n1mean,pch='.',col=cols[4],xlab="Straying rate m",ylab="Steady state biomass",cex=0.5,ylim=c(0,max($n1mean)+200),las=1)
 points($mvec,$n2mean,pch='.',col=cols[4],cex=0.5)
[…]

### 2017-12-19 model/model2/Figure2_4.jl

@@ -11,26 +11,26 @@
 @everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/timeSS.jl")
 @everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/qualsfunc.jl")
 @everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/bifdet.jl")
-
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/movingaverage.jl")
 
 
 #Analysis over m & theta divergence
-mvec=collect(0.0:0.001:0.3);
-hvec = collect(0.0:0.005:1.0);
+mvec=collect(0.0:0.0005:0.3);
+hvec = collect(0.0:0.01:0.5);
 
 
-n1mean=SharedArray(Float64,length(hvec),length(mvec));
-n2mean=SharedArray(Float64,length(hvec),length(mvec));
-x1mean=SharedArray(Float64,length(hvec),length(mvec));
-x2mean=SharedArray(Float64,length(hvec),length(mvec));
-pe=SharedArray(Float64,length(hvec),length(mvec));
+n1mean=SharedArray{Float64}(length(hvec),length(mvec));
+n2mean=SharedArray{Float64}(length(hvec),length(mvec));
+x1mean=SharedArray{Float64}(length(hvec),length(mvec));
+x2mean=SharedArray{Float64}(length(hvec),length(mvec));
+pe=SharedArray{Float64}(length(hvec),length(mvec));
 
-n1meanE=SharedArray(Float64,length(hvec),length(mvec));
-n2meanE=SharedArray(Float64,length(hvec),length(mvec));
-x1meanE=SharedArray(Float64,length(hvec),length(mvec));
-x2meanE=SharedArray(Float64,length(hvec),length(mvec));
-peE=SharedArray(Float64,length(hvec),length(mvec));
-rtE=SharedArray(Float64,length(hvec),length(mvec));
+n1meanE=SharedArray{Float64}(length(hvec),length(mvec));
+n2meanE=SharedArray{Float64}(length(hvec),length(mvec));
[…]

### 2017-12-19 model/model2/Figure3_S8.jl

@@ -15,16 +15,16 @@
 #Analysis over m
 tmax=10000;
 
-mvec = collect(0.0001:0.001:0.3);
+mvec = collect(0.0001:0.001:0.5);
 pvec = ["small","large","both"];
-hvec = [0.2, 0.8]
+hvec = 0.2;
 reps = 100;
 
 z=0.5;
 rmax=2.0;
 beta=0.001;
 theta1=5.0;
-thetadiff=2.4;
+thetadiff=2.0;
 tau=1.0;
 C=300;
 sigmaE=0;
@@ -43,9 +43,12 @@ t_ext = Int64(round(tmax/2));
 
 
 rt = SharedArray{Float64}(reps,length(mvec),length(pvec),length(hvec));
+pe = SharedArray{Float64}(reps,length(mvec),length(pvec),length(hvec));
+
 rt_ddm = SharedArray{Float64}(reps,length(mvec),length(pvec),length(hvec));
 m1mean = SharedArray{Float64}(reps,length(mvec),length(pvec),length(hvec));
 m2mean = SharedArray{Float64}(reps,length(mvec),length(pvec),length(hvec));
+pe = SharedArray{Float64}(reps,length(mvec),length(pvec),length(hvec));
 
 @sync @parallel for r=1:reps
   
@@ -58,7 +61,7 @@ m2mean = SharedArray{Float64}(reps,length(mvec),length(pvec),length(hvec));
       
       extpop = pvec[j];
       
-      for k=1:2
+      for k=1:1
         
         h = hvec[k];
         
@@ -82,42 +85,50 @@ m2mean = SharedArray{Float64}(reps,length(mvec),length(pvec),length(hvec));
         
         t_ss, relaxtime = timeSS(n1,n2,t_ext);
         
+        n1trim = n1[Int64(floor(tmax*burnin)):tmax-1];
+        n2trim = n2[Int64(floor(tmax*burnin)):tmax-1];
+        
         rt[r,i,j,k] = relaxtime;
         
-
-        #Do the same thing for density-dependent m
+        pe[r,i,j,k] = mean([(std(n1trim)/mean(n1trim)),(std(n2trim)/mean(n2trim))])*
+        (1/(std(n1trim+n2trim)/mean(n1trim+n2trim)))
[…]

### 2017-12-19 model/model2/FigureS2_S4_S5.jl

@@ -4,11 +4,11 @@
 using RCall
 
 
-thetascale = 20;
-m=collect(0:0.01:0.5);
-thetadiff=zeros(length(m));
-for i = 1:length(m)
-    thetadiff[i] = (1-2*m[i])/(thetascale*m[i])
+thetascale=20;
+thetadiff=collect(0:0.01:3.0);
+m = zeros(length(thetadiff));
+for i=1:length(thetadiff)
+    m[i] = 1/(2+thetascale*thetadiff[i])
 end
 
 namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/Manuscript/FinalDraft3/fig_mthetarelation.pdf");
@@ -17,6 +17,8 @@ pdf($namespace,height=5,width=6)
 plot($thetadiff,$m,type='l',ylab='Straying rate m',xlab=expression(paste('Habitat heterogeneity ',Delta,theta)),ylim=c(0,0.5),xlim=c(0,3))
 dev.off()
 """
+
+
 
 using Distributions
 using RCall

### 2017-12-19 model/model2/Figure_4regimes_ddm.jl

@@ -1,0 +1,1866 @@
+using Distributions, RCall
+
+
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve_ddm.jl")
+@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinJacobian.jl")
+
+#Analysis over m
+tmax=10000;
+mvec = collect(0.0:0.001:0.5);
+n1ts = zeros(Float64,length(mvec),tmax);
+n2ts = zeros(Float64,length(mvec),tmax);
+n1mean=zeros(Float64,length(mvec));
+n2mean=zeros(Float64,length(mvec));
+n1sd=zeros(Float64,length(mvec));
+n2sd=zeros(Float64,length(mvec));
+aggmean=zeros(Float64,length(mvec));
+aggsd=zeros(Float64,length(mvec));
+x1mean=zeros(Float64,length(mvec));
+x2mean=zeros(Float64,length(mvec));
+pe=zeros(Float64,length(mvec));
+
+n1ts_ddm = zeros(Float64,length(mvec),tmax);
+n2ts_ddm = zeros(Float64,length(mvec),tmax);
+n1mean_ddm=zeros(Float64,length(mvec));
+n2mean_ddm=zeros(Float64,length(mvec));
+n1sd_ddm=zeros(Float64,length(mvec));
+n2sd_ddm=zeros(Float64,length(mvec));
+aggmean_ddm=zeros(Float64,length(mvec));
+aggsd_ddm=zeros(Float64,length(mvec));
+x1mean_ddm=zeros(Float64,length(mvec));
+x2mean_ddm=zeros(Float64,length(mvec));
+m1mean_ddm=zeros(Float64,length(mvec));
+m2mean_ddm=zeros(Float64,length(mvec));
+pe_ddm=zeros(Float64,length(mvec));
+
+z=2;
+rmax=2.0;
+beta=0.001;
+theta1=5.0;
+thetadiff=2.0;
+tau=1.0;
+h=0.2;
+sigmaE=0;
+sigmaG=1;
[…]

### 2017-12-19 model/model2/scenarios.jl

@@ -1,0 +1,60 @@
+
+tmax=10000;
+z=0.5;
+rmax=2.0;
+beta=0.001;
+theta1=5.0;
+thetadiff=2.0;
+tau=1.0;
+h=0.2;
+sigmaE=0;
+sigmaG=1;
+perror=0.01;
+refuge=0.01;
+m=0.15;
+a0=1-m;
+C=10^4;
+
+extpop = "large";
+t_ext = Int64(round(tmax/2));
+
+n1, n2, x1, x2, w1, w2 = 
+KevinEvolveExtinct_ddm( 
+tmax, 
+z, 
+rmax,
+beta,
+theta1,
+thetadiff,
+tau,
+h,
+a0,
+C,
+sigmaE,
+sigmaG,
+perror,
+extpop,
+t_ext,
+refuge
+);
+
+t_ss, relaxtime = timeSS(n1,n2,t_ext);
+
+
+R"""
+library(RColorBrewer)
+pal = brewer.pal(3,'Set1')
+#pdf($namespace,height=5,width=6)
+par(mfrow=c(2,1),mai = c(0.8, 0.8, 0.4, 0.1))
+plot($n1,type='l',col=pal[1],xlim=c($t_ext-50,$t_ss+50),ylim=c(0,max($n1+$n2)),ylab='Population density',xlab='Time')
+lines($n2,col=pal[2])
+lines($n1+$n2,col='black')
+lines(c($t_ss,$t_ss),c(-10,10000))
+points($t_ss,$(n1[t_ss])+$(n2[t_ss]),pch=16)
+text(x=$t_ss+18,y=1,paste(c('Recovery time=',$relaxtime),collapse=''))
+plot($x1,type='l',col=pal[1],ylim=c(5,10),xlim=c($t_ext-50,$t_ss+50),ylab='Trait means',xlab='Time')
+lines($x2,col=pal[2])
+lines(c($t_ss,$t_ss),c(0,10000))
+#dev.off()
+"""
+

### 2017-12-19 model/src/KevinEvolveSS_ddm.jl

@@ -1,0 +1,110 @@
+function KevinEvolveSS_ddm(
+  n0,
+  x0,
+  tmax, 
+  z, 
+  rmax,
+  beta,
+  theta1,
+  thetadiff,
+  tau,
+  h,
+  a0,
+  C,
+  sigmaE,
+  sigmaG,
+  perror
+  )
+  
+  #Initialize variables
+  n1 = zeros(Float64,tmax); n1[1]=n0[1];
+  n2 = zeros(Float64,tmax); n2[1]=n0[2];
+  
+  m1 = zeros(Float64,tmax-1);
+  m2 = zeros(Float64,tmax-1);
+  
+  # x1 = zeros(Float64,tmax); x1[1]=theta1 + rand(Normal(0,0.0001));
+  # x2 = zeros(Float64,tmax); x2[1]=(theta1 + thetadiff)  + rand(Normal(0,0.0001));
+  x1 = zeros(Float64,tmax); x1[1] = x0[1];
+  x2 = zeros(Float64,tmax); x2[1] = x0[1];
+  
+  w1 = zeros(Float64,tmax-1);
+  w2 = zeros(Float64,tmax-1);
+  meanfit1 = zeros(Float64,tmax-1);
+  meanfit2 = zeros(Float64,tmax-1);
+  
+  #Process Error Distribution
+  if perror == 0
+    pdist = [0,0];
+  else
+    pdist = Normal(0,perror);
+  end
+  
+  rfunc = function(mu,theta)
+    r = rmax*tau*(exp(-((theta-mu)^2)/(2*((sigmaE^2 + sigmaG^2)+tau^2))))/(sqrt(tau^2+(sigmaE^2 + sigmaG^2)))
+  end
+  
+  for t=1:tmax-1
+      
+      m1[t] = 1 - (a0 + (1-a0)*(n1[t]/(C+n1[t])));
+      m2[t] = 1 - (a0 + (1-a0)*(n2[t]/(C+n2[t])));
+      
+      w1[t] = (n1[t]-m1[t]*n1[t])/(n1[t]-m1[t]*n1[t]+m2[t]*n2[t]);
+        
+      w2[t] = (n2[t]-m2[t]*n2[t])/(n2[t]-m2[t]*n2[t]+m1[t]*n1[t]);
+      
[…]

### 2018-01-12 manuscript/oldversions/Description.tex
penultimate

\documentclass[twocolumn,preprintnumbers,amsmath,amssymb,superscriptaddress]{revtex4}
%\usepackage[pdftex]{graphicx}

\usepackage{amsmath,amsfonts,amssymb}
\usepackage[english]{babel}
\usepackage[latin1]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{color}
\usepackage{float}
\usepackage{verbatim}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{mathtools}
\usepackage{stmaryrd}
\usepackage{anyfontsize}

%\usepackage{epstopdf}
%\usepackage{array}
%\usepackage{tabularx}
%\usepackage{multirow}
\usepackage{color}
%\usepackage{multibox}
%\usepackage{rotating}
%\usepackage{lineno}
%\usepackage[left]{lineno}
%\usepackage[comma,sort&compress]{natbib}
%\usepackage{authblk}
%\usepackage{multicol}

%\bibliographystyle{ieeetr}

%\linenumbers
%\setlength\linenumbersep{3pt}

\begin{document}

%\title{Simple rules yield complex communities: deconstructed species interactions and the assembly of communities}
%\title{Community assembly and dynamics by the deconstruction of species interactions}
\title{Salmon straying description}
%\author{Justin D. Yeakel${}^{1,2,*}$, Christopher P. Kempes${}^{2}$, \& Sidney Redner${}^{2,3}$ \\ \\
%${}^1$School of Natural Science, University of California Merced, Merced, CA \\
%${}^2$The Santa Fe Institute, Santa Fe, NM \\
%${}^3$Department of Physics, Boston University, Boston MA \\
%${}^*$To whom correspondence should be addressed: jdyeakel@gmail.com
%}

\maketitle

\section*{Model Description}

\noindent {\bf Goal:} incorporate the effects of straying into an eco-evolutionary model of population dynamics.

\begin{itemize}
\item Consider a 2-site system, where each site has a different optimum trait value $(\theta_1, \theta_2)$, where if the population has a trait value equal to the optimum $x_1=\theta_1$, the growth rate is maximized.
\item When strays come into the population, they bring trait values equal to those of their native habitat (likely farther from the local optimum), and this will tend to push the local population away from the optimum if the rate of straying is too high
\item The traits evolve through time in response to changes in fitness of the populations
\item Though straying will generally be bad for a local population, they also serve to rescue the population from local extinction. We aim to explore this tradeoff
\end{itemize}

\noindent A more rigorous explanation:

Let's consider 2 populations that change over time according to the following population dynamics:

\begin{eqnarray}
  \dot{N_1} &= \bar{a_1}(\bar{x_1})R(N_1,N_2)-\mu N_1 \\
  \dot{N_2} &= \bar{a_2}(\bar{x_2})R(N_1,N_2)-\mu N_2
\end{eqnarray}

\noindent where $\bar{a_i}$ is the mean growth rate of population $i$, which depends on the mean trait value of that population $\bar{x_i}$, $R(N_1,N_2)$ is the recruitment function, which depends on the density of the local population $N_1$ as well as those that stray in from population $N_2$, and $\mu$ is the mortality rate... let's say it is the same for both, and that most mortality is happening out in the ocean (where the populations are mixed).

The growth rate is maximized if the trait value is equal to the optimum trait value $\theta_i$, and falls off in a Gaussian manner as you move away from this optimum, so

\begin{equation}
  a_i(x_i) = \alpha_i\exp\left(\frac{(x_i-\theta_i)^2}{2\tau_i^2}\right)
\end{equation}

\noindent where $\alpha_i$ is the maximal growth rate, and $\tau_i$ determines how steeply the growth rate falls as distance to the optimum increases.
If we assume that the trait $x_i$ is normally distributed with a weighted mean composed of both resident $i$ and straying $j$ individuals where $\eta$ is the proportion of the mixed population that are strays, then $\bar{x_i} = (1-\eta)x_i + \eta x_j$.
We then assume that the trait distribution has some variance $\sigma^2$ and that the pdf is given $p(x,\bar{x_i})$, such that the mean growth rate is $\bar{a_i}(\bar{x_i}) = \int_{-\infty}^\infty a_i(x)p(x,\bar{x_i}){\rm d}x$, which is written

\begin{equation}
  \bar{a_i}(\bar{x_i}) = \frac{\alpha_i \tau_i}{\sqrt{\sigma^2+\tau^2}}\exp\left(-\frac{\bar{x_i}-\theta_i)^2}{2(\sigma^2+\tau_i^2)}\right)
\end{equation}

We have 2 differential equations describing the two populations through time, and now we need to show how the trait values change through time.
This depends on how we define the relative fitness of the populations, which in this case is pretty straightforward.
We define fitness as the per capita population change

\begin{equation}
  W_i = a_i(x_i)\frac{R(N_i,N_j)}{N_i}-\mu,
\end{equation}

\noindent where $R(N_i,N_j)$ is some unspecified recruitment function (say Shephard), but one that acts on the total reproducing population, adjusting for incoming and outgoing strays $N_i + \eta(N_j - N_i)$.
The mean fitness is then $\bar{W_i}=\int_{-\infty}^\infty W(a_i,N_i,N_j)p(x_i,\bar{x_i}){\rm d}x$ which is $\bar{W_i}=\bar{a_i}(\bar{x_i})\frac{R(N_i,N_j)}{N_i}-\mu$.

Finally, from Lande (1976), we get the differential equation describing how the mean trait value changes over time, where

\begin{equation}
  \dot{\bar{x_i}} = \sigma^2_{\rm G} \frac{d\bar{W_i}}{d\bar{x_i}}
\end{equation}

\noindent where $\sigma^2_{\rm G}$ is the component of the trait variation that is genetic.
So the full system for the 2-site system would be

\begin{align}
  \dot{N_1} &= \bar{a_1}(\bar{x_1})R(N_1,N_2)-\mu N_1 \\
  \dot{N_2} &= \bar{a_2}(\bar{x_2})R(N_1,N_2)-\mu N_2 \\
  \dot{\bar{x_1}} &= \sigma^2_{\rm G} \frac{d\bar{W_1}}{d\bar{x_1}} \\
  \dot{\bar{x_2}} &= \sigma^2_{\rm G} \frac{d\bar{W_2}}{d\bar{x_2}}
\end{align}

SO: the populations change over time, and the rate at which they change depends on the mismatch between the optimum trait value of the site and the mean trait value of the population.
The fitness of that population determines the rate of change for the trait value itself, which is a mix of local and straying trait values... as the rate of straying increases, the more difficulty the system should have to zero in towards the optimum trait value.

A few things I should mention: This works because we assume the variance of the trait values is static.
Of course, as we mix trait values from local and straying populations, we do change the variance a little.
We are ignoring this.
I'd argue that we can ignore this as long as the proportion of strays entering the local population is relatively small.
Changes in variance can be addressed, but the problem becomes so complicated that it would be better to resort to simulation.
On that line, it might be good to write a simulation version of this model so that we could examine whether changes in trait variance over time are really that important... in my opinion, this is a second or third order problem that we could explore later.
There is a lot to explore if you guys accept that small amounts of straying shouldn't change the variance too much.

I'm currently displaying the proportion of the population that are strays as a single parameter $\eta = cN_j/(N_i+cN_j)$ where $c$ is the proportion of the remote population that is straying... i.e. if $c$ is small, so is $\eta$ as long as $N_1$ and $N_2$ are similar in magnitude.
I think that it would be really fun to explore the case where $c$ is density dependent on $N_j$, as you guys were describing in our last meeting.
If it was a relatively simple function of $N_j$, that would be pretty easy to incorporate.

Okay - that's all for now! I hope that this is coherent - I'm on beer 3.

\end{document}

### 2018-01-12 manuscript/oldversions/msv3.tex
penultimate

--- 2017-06-01 manuscript/msv3.tex
@@
 
-\bibliographystyle{ieeetr}
+\usepackage{natbib}
+\bibpunct{(}{)}{;}{a}{}{;}
+
+\bibliographystyle{prsb}
 
@@
 
+\begin{abstract} %250 words
+The spatial dispersal of individuals is known to play an important role in the dynamics of populations, and is central to metapopulation theory.
+At the same time, local adaptation to different environmental conditions creates a geographic mosaic of evolutionary forces, where the combined drivers of selection and gene flow interact.
+Although the dispersal of individuals from donor to recipient populations provides connections within the metapopulation, potentially promoting demographic and evolutionary rescue, it may also introduce maladapted individuals into habitats that are host to different environmental conditions, potentially lowering the mean fitness of the recipient population.
+Here we explore a minimal model of the eco-evolutionary dynamics between two populations connected by dispersal, where the productivity of each is defined by a trait complex that is subject to local selection.
+Although general in nature, our model is inspired by salmon metapopulations, where dispersal between populations is defined in terms of the \emph{straying rate}, which has been shown to be density dependent, possibly emerging as a consequence of collective migration.
+The results of our model reveal that increased straying between evolving populations leads to alternative stable states, which has large and nonlinear effects on two measures of metapopulation robustness: the portfolio effect and the time to recovery following an induced disturbance.
+We show that intermediate levels of straying result in large gains in robustness, and that increased differences between sites (increased habitat heterogeneity) promotes robustness when straying rates are low, and erodes robustness when straying rates are high.
+Finally, we show that density-dependent straying promotes robustness, particularly when the aggregate biomass is low and straying is correspondingly high, and this has important ramifications for the conservation of salmon metapopulations facing both natural and anthropogenic disturbances.
+\end{abstract} %250 words
+
 \maketitle
 
-\section*{Abstract}
-{\small
-We show that taking straying into account leads to alternative stable states in population densities and trait values, which has consequences for maladaptation, intraspecific trait variability, and long-term robustness of salmon metapopulations.
-}
-\section{Introduction}
-
 Intraspecific diversity can increase the resilience and stability of species or metapopulations. 
-This diversity-stability linkage can occur when there are asynchronous population dynamics, where the changes in population vary across a metapopulation. 
+This diversity-stability linkage can occur when there are asynchronous population dynamics, where the changes in population size varies across the metapopulation. 
 This asynchrony will increase the potential for demographic rescue \cite{Brown:1977gk,Earn:2000fm} and also decrease the variability of processes that integrate across the metapopulation \cite{Anonymous:2015gf}.
@@
 
-It is clear that dispersal (i.e., movement with consequences for gene flow) can influence the evolutionary dynamics of the metapopulation, the interplay between selection, population dynamics, and persistence is not well understood.
+% Although dispersal and the consequent gene flow can influence the evolutionary dynamics of metapopulations, the interplay between selection, population dynamics, and persistence is not well understood.
 Movement of individuals among local populations (i.e. dispersal) can have a large influence on metapopulation persistence \cite{MilnerGulland:2011vm}. 
@@
 
-We use the discrete Ricker population dynamic framework described by Shelton and Mangel \cite{Shelton:2011eq} as the basis for our two-site model, with the added effect of the local population $N_i$ mixing with a set proportion $m$ of a remote population $N_j$ that is straying into it.
-In this sense, both populations serve as donor and recipient populations.
+We use discrete Ricker framework described by Shelton and Mangel \cite{Shelton:2011eq} as the basis for our two-site metapopulation model, with the added effect of the local population $N_i$ mixing with a set proportion $m$ of a remote population $N_j$ that is straying into it.
+In this sense, both populations serve simultaneously as donor and recipient populations.
 We first assume that the proportion ${\rm e}^{-Z}$ of both populations survive such that the surviving aggregated population, composed of both local individuals (at site $i$) and incoming strays (from site $j$), is $\left((1-m)N_i(t) + m N_j(t) \right){\rm e}^{-Z}$.
@@
 Because straying individuals are emigrating from a population with a mean trait value farther from the local optimum, their rate of recruitment is diminished.
-Recent studies of wild sockeye have indeed found that straying individuals have lower life-time fitness than individuals that do not stray, although it unknown at what life-stage this selection occurs \cite{Peterson:2014gy}.
+Recent studies of wild sockeye salmon have indeed found that straying individuals have lower life-time fitness than individuals that do not stray, although it unknown at what life-stage this selection occurs \cite{Peterson:2014gy}.
 % The compensatory effects are then determined by the exponential, following the Ricker stock-recruitment relationship.
@@
 Thus, we assume that strays can successfully reproduce and introduce their genotypes into the recipient population, which is supported by observations in wild populations \cite{Jasper:2013cc}.
-Second, we assume that changes in trait variance through time are minimal, such that $\sigma^2$ is assumed to be constant, and this is a common simplification in eco-evolutionary models of population dynamics \cite{Lande:1976ga,Schreiber:2011wx,Gilbert:2014ee,Gibert:2015kc}.
+Second, we assume that changes in trait variance through time are minimal, such that $\sigma^2$ is assumed to be constant, which is a common simplification in eco-evolutionary models of population dynamics \cite{Lande:1976ga,Schreiber:2011wx,Gilbert:2014ee,Gibert:2015kc}.
 
-An increasing flow of incoming strays is generally expected to pull the mean trait value of the local population away from its optimum over time, which will decrease its rate of recruitment.
+% An increasing flow of incoming strays is generally expected to pull the mean trait value of the local population away from its optimum over time, which will decrease its rate of recruitment.
 Following Lande \cite{Lande:1976ga}, the mean trait value thus changes through time according to the difference equation
@@
 
-\noindent where the first two factors determine the mixed normal average of the aggregated local and remote populations.
+\noindent where the first two components determine the mixed normal average of the aggregated local and remote populations.
 The partial derivative in the Eq.
@@
 \noindent {\bf (c) Habitat heterogeneity}
+\noindent Increasing differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater regional differences in the conditions that favor alternative trait complexes, which we interpret here as increased habitat heterogeneity.
+If both populations are isolated, natural selection will direct the mean trait values of both populations towards their respective optima, such that $\mu_i(t) \rightarrow \theta_i$ as $t\rightarrow\infty$.
+%difference in optimal trait values between the local and remote populations, which we call . 
+%Linking m/m0 and thetadiff
+Although we largely treat habitat heterogeneity and the rate of straying as independent parameters, we evaluate a case where we assume thgat increased habitat heterogeneity correlates with lower straying rates, and vice versa.
+Two scenarios may lead to this correlation: 
+(\emph{i}) sites may be distributed over greater spatial distances, where habitat dissimilarity is assumed to be larger between more distant sites, whereas the likelihood of straying over greater distances would likewise be lower \cite{Candy:2000hu,JPE:JPE1383};
+(\emph{ii}) individuals may have behaviors promoting dispersal between habitats with structural or physiognomic similarities \cite{Peterson:2014gy}.
 
-%Linking m/m0 and thetadiff
-The straying rate is largely influenced by the spatial distance between the donor and recipient population.
-The greater the distance between two populations, the lower the expected rate of straying \cite{H:2013fs,Westley:2015to}.
-We account for this interdependence in our model by assuming that $m$ (if the stray rate is constant) or $m_0$ (if the stray rate is density-dependent) is a function of the difference between optimal trait values between sites $\theta_i-\theta_j$, which can be assumed to be large if the remote site $j$ is a great distance away from the local site $i$.
-If sites $i$ and $j$ are very close, the stray rate is maximized at $m_{\rm max} = 0.5$, assuming both sites are equally attractive to the respective populations.
-Thus, we can integrate these two variables by setting $m,m_0 = (1/m_{\rm max} + \epsilon (\theta_i-\theta_j))^{-1}$, where $\epsilon$ sets the sensitivity of a declining $m$ to increasing distance (greater values of $\theta_i-\theta_j$).\\
+Where indicated, we account for correlation between the straying rate and the differnce between optimal trait values by assuming that $m$ (if the stray rate is constant) or $m_0$ (if the stray rate is density-dependent) is a function of the difference between optimal trait values between sites $\Delta\theta$.
+The straying rate is maximized at $m_{\rm max} = 0.5$, assuming both sites are equally attractive to the respective populations.
+Thus, we can integrate these two variables by setting $m,m_0 = (1/m_{\rm max} + \epsilon \Delta\theta)^{-1}$, where $\epsilon$ sets the sensitivity of a declining $m$ to increasing distance (greater values of $\Delta\theta$), which we set to $\epsilon=2$.\\
 
@@
 %Although there are more robust ways to measure the portfolio effect (in particular for comparing the PE across species with different mean-variance relationships; REF: Anderson), its only utlity here is to provide a simple statistical measure of metapopulation persistence.
-A more direct way to measure system robustness is to measure the time that it takes the system (measured as the aggregate biomass $N_T$) to recover its steady state abundance following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
+A more direct way to measure system robustness is to measure the time that the system (measured as the aggregate biomass $N_T$) takes to recover its steady state abundance following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
 Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation \cite{GuckHolmes}, because we aimed to 1) assess the effects of a large perturbation, and 2) estimate the time required for all transient effects to decay (including dampened oscillations), we used a simulation-based numerical procedure.
@@
 %Critical slowing down
-As the fold bifurcation is approached with increasing $m$, the portfolio effect increases sharply due to an amplification in variance within both donor and recipient populations ${\rm VAR}(N_{i,j})$.
-This variance increase is the product of a dynamical process known as \emph{critical slowing down} that occurs near fold bifurcations \cite{Scheffer:2009gg}, a phenomenon that some have suggested may serve as an early warning indicator for approaching phase transitions \cite{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br}.
+As the fold bifurcation is approached with increasing $m$, the portfolio effect increases sharply due to an amplification in variance within both donor and recipient populations.
+This amplification in variance is the product of a dynamical process known as \emph{critical slowing down} that occurs near fold bifurcations \cite{Scheffer:2009gg}, a phenomenon that some have suggested may serve as an early warning indicator for approaching phase transitions \cite{Scheffer:2009gg,Lade:2012eu,Anonymous:2013br,Dakos:2014br,Krkosek:2014ch}.
 For larger values of $m$ (to the right of the fold bifurcation in Fig \ref{fig:PE}a-c), where alternative stable states occur, the portfolio effect declines steadily as the CV of $N_T$ increases.
@@
 \noindent{\bf (c) The role of habitat heterogeneity and changing selective landscapes}\\ 
-\noindent Increasing differences in optimal trait values between sites ($\Delta\theta = \left|\theta_i - \theta_j\right|$) corresponds to greater regional differences in the conditions that favor alternative trait complexes, which we interpret here as greater habitat heterogeneity (REFS).
-If both populations are isolated, natural selection will direct the mean trait values of both populations towards their respective optima, such that $\mu_i(t) \rightarrow \theta_i$ as $t\rightarrow\infty$.
-%difference in optimal trait values between the local and remote populations, which we call .
-With the onset of straying, we find that increasingly divergent trait optima generally lower $N_T$ and exaggerate $\Delta N$ (figure \ref{fig:thetadiffN}), such that the biomass distribution becomes increasingly uneven.
+\noindent With the onset of straying, we find that increasingly divergent trait optima generally lower $N_T$ and exaggerate $\Delta N$ (figure \ref{fig:thetadiffN}), such that the biomass distribution becomes increasingly uneven.
 %I've checked this - it's a theshold type (logistic) response Supp fig?
@@
 (\emph{i}) above a threshold $m$ value, the dominant population recovers quickly enough that the evolving subordinate phenotype is overwhelmed by incoming strays, and it shifts back to its pre-disturbance (subordinate) state;
-(\emph{ii}) below a threshold $m$ value, there is enough time and isolation for the subordinate trait mean to shift towards its local optimum, and away from that of the recovering dominant population, such that the dominant population becomes subordinate, and the subordinate population becomes dominant (figure \ref{fig:inertia}).
-This threshold value of $m$, below which this behavior occurs, is marked by the asterisk in figure \ref{fig:mtheta}, and holds for both constant and density-dependent straying (figure \ref{fig:mthetamvm}).
+(\emph{ii}) below a threshold $m$ value, there is an \emph{inversion} between subordinate and dominant states: because there is enough time and isolation for the subordinate trait mean to shift towards its local optimum, and away from that of the recovering dominant population, the dominant population becomes subordinate, and the subordinate population becomes dominant (figure \ref{fig:inertia}).
+This threshold value of $m$, below which the inversion dynamic behavior occurs, is marked by the asterisk in figure \ref{fig:mtheta}, and holds for both constant and density-dependent straying (figure \ref{fig:mthetamvm}).
 
@@
 
-We have shown that density-dependent straying between populations consistent with collective navigation, coupled with localized selection against donor phenotypes, has a large and nonlinear impact on dynamic properties directly impacting the robustness of metapopulations against external disturbances.
-Specifically, we observed robustness in terms of: 
+We have shown that density-dependent straying between populations consistent with collective navigation, coupled with localized selection against donor phenotypes, has a large and nonlinear impact on dynamic properties that affect metapopulation robustness.
+We measured robustness as:
 1) the average-CV portfolio effect \cite{Anderson:2013gb,Anonymous:2015gf}, a statistical metric commonly used to assess the buffering capacity of metapopulations, and
 2) the recovery time, defined here as the time required for the aggregate metapopulation biomass $N_T$ to return to its steady state following an induced disturbance, which is mechanistically linked to persistence \cite{Ovaskainen:2002il}.
-In our minimal eco-evolutionary model of dispersal and natural selection between two populations, we show that these statistical and mechanistic descriptors of metapopulation dynamics and robustness are tightly coupled (figure \ref{fig:PE}d), which is not uncommon among different metrics of stability \cite{Donohue:2013iu}.
-Taken as a whole, we interpret our results to suggest an important role of density-dependent straying in the colonization and recovery dynamics for populations in shifting river mosaics, while also underscoring the risk of straying by individuals with maladaptive traits to reduce the productivity of locally adapted stock complexes.
+In our minimal eco-evolutionary model of dispersal and natural selection between two populations, we show that these statistical and mechanistic descriptors of metapopulation dynamics and robustness are tightly coupled (figure \ref{fig:PE}d), which is not uncommon for diverse metrics of stability \cite{Donohue:2013iu}.
+Taken as a whole, our results point to an important role of density-dependent straying in the colonization and recovery dynamics within metapopulations, while also underscoring the risk of straying by individuals with maladaptive traits to reduce the productivity of locally adapted stock complexes.
 
@@
 %alternative stable states & critical slowing
-A salient finding from our results was that straying can push metapopulations into alternative stable states, pushing one of the populations to a high density population (the \emph{dominant state}), and one to a low density population (the \emph{subordinate population}).
+A salient finding from our results was that straying can lead to the emergence of alternative stable states, pushing one of the populations to high density (the \emph{dominant state}), and one to low density (the \emph{subordinate population}).
 This asymmetry in population densities also leads to an asymmetry in the mean trait values, both of which are skewed towards the local optimum of the dominant population. 
-The formation of alternative stable states in our system is a relatively common example of spatial pattern formation, which only occurs above a threshold straying rate, the value of which depends on trait heritability and the difference in trait optima between sites.
+The formation of alternative stable states in our system is a relatively common example of spatial pattern formation.
+%, only occurring above a threshold straying rate that largely depends on trait heritability and the difference in trait optima between sites.
 Pattern formation can occur as a consequence of myriad ecological processes, including habitat selection, aggregation, local environmental conditions, and/or interspecific interactions such as competition or predation, and is inexorably linked to scale \cite{Levin:1992p1472}.
-Additionally, pattern formation is also well known to occur with the onset of spatially or diffusion-induced instabilities, or Turing instabilities \cite{Meinhardt:2012hw}. 
-Here, we describe an additional mechanism through which pattern formation in local densities can occur, when the onset of collective movement-mediated maladaptation leads to local differences in reproductive rates resulting in alternative stable states across sites.
+Pattern formation is also known to occur with the onset of spatially or diffusion-induced instabilities, or Turing instabilities \cite{Meinhardt:2012hw}. 
+Here, we describe an additional mechanism by which pattern formation in local densities can occur, when the onset of collective movement-mediated maladaptation leads to local differences in reproductive rates resulting in alternative stable states across sites.
 %This mechanism of pattern formation through the simultaneous interplay between ecological and evolutionary dynamics may open up new areas of research.
-While metapopulation approaches have widely recognized the importance of source and sink populations \cite{Hanski:1999vu} and have observed formation of alternative states \cite{Boughton:1999fa}, here we show that even similar populations can evolve into divergent states where one population is large and the other is smaller and maladapted.
+While metapopulation approaches have widely recognized the importance of source and sink populations \cite{Hanski:1999vu} as well as the occurrence of alternative states \cite{Boughton:1999fa}, here we show that the coupled forces of ecological and evolutionary processes can result in similar populations diverging into alternative states where one population is large and well-adapted, while the other is smaller and maladapted.
+The appearance of these alternative states is largely dependent on the dispersal, or rate of straying, between sites.
 
@@
 Previous theoretical work has shown that increased connectivity may erode portfolio effects in herring metapopulations, where straying is also thought to be density-dependent \cite{Secor:2009ena}.
-Although high levels of dispersal in our system shows a similar trend, the interplay between dispersal and PE is more subtle when selection for local adaptations is considered, such that low to intermediate levels of density-dependent straying result in an elevated PE, increasing the buffering capacity of the metapopulation.
+Although high levels of dispersal in our system generally supports this finding, the interplay between dispersal and PE is more subtle when selection for local adaptations is considered, such that low to intermediate levels of density-dependent straying result in an elevated PE, thus increasing the buffering capacity of the metapopulation.
 Although PE is measured at the steady state, low to intermediate rates of straying also appear to have a beneficial effect on transient dynamics.
-When there is just enough straying to cause alternative stable states, recovery times following induced disturbances decline, though -- as with the PE -- the time to recovery then grows if straying becomes too large (figure \ref{fig:relax}a).
+When there is just enough straying to cause alternative stable states, the time to recovery following an induced disturbance declines, though -- as with the PE -- recovery time then grows if straying becomes very large (figure \ref{fig:relax}a).
 
@@
 We observed three salient results that contribute to our understanding of collective movement that combine to suggest that density-dependent straying may play an important role in the persistence of metapopulations over evolutionary time.
-First, the inclusion of density-dependent straying does not qualitatively alter either (\emph{i}) steady state or (\emph{ii}) transient dynamics of our minimal eco-evolutionary model, but effectively rescales measures of robustness to the straying rates the emerge from the coupled dynamics.
+First, the inclusion of density-dependent straying does not qualitatively alter either (\emph{i}) steady state or (\emph{ii}) transient dynamics of our minimal eco-evolutionary model, but effectively rescales measures of robustness to the lower straying rates that emerge as a consequence of the coupled dynamics.
 Second, compared to systems with constant dispersal, density-dependent straying appears to increase the portfolio effect across a range of straying rates (figure \ref{fig:thetaPE}a). 
@@
 In the case of near-collapse, although both populations inherit low population densities, the mean trait values of both are skewed towards the optimum of the dominant population.
-Due to density-dependent straying, low population densities evince greater dispersal, and while this increased connectivity primarily facilities the growth of the dominant population (because the trait means are closer to the dominant optimum), because the dominant population is the bulk of the aggregate biomass, the overall recovery time is significantly lessened (figures \ref{fig:relax}b, \ref{fig:relaxtraj_bothlh}).
+Due to density-dependent straying, low population densities evince greater dispersal, and while this increased connectivity primarily facilities the growth of the dominant population (because the trait means are closer to the dominant optimum), because the dominant population contains the bulk of the aggregate biomass, the overall recovery time is lessened significantly (figures \ref{fig:relax}b, \ref{fig:relaxtraj_bothlh}).
 % Third, in habitats with greater heterogeneity between sites (measured by an increase in $\Delta\theta$), density-dependent straying diminished robustness at low rates of straying (marginally lower portfolio effects and substantially longer return times; figures \ref{fig:thetaPE}b, \ref{fig:relaxtheta}a,b).
@@
 This may be particularly consequential for populations that are spatially adjacent but separated by sharp environmental boundaries, such that trait optima are divergent yet dispersal is relatively high.
-Such a scenario plays out repeatedly in the context of adjacent wild and hatchery-produce salmon.
+Such a scenario plays out repeatedly in the context of adjacent wild and hatchery-produced salmon. 
 Although wild and hatchery populations may occur close on the landscape, and indeed often are sympatric within the same river network, the selective environments to which they are locally adapted differ dramatically \cite{Christie:2012bj}.
@@
 All else being equal, habitats that are closer in space generally have greater similarity in environmental conditions than habitats that are geographically distant, and correspondingly phenotypes of more proximately located populations should be more similar to those that are distant \cite{Westley:2012ui}.
-Correspondingly, it is reasonable to expect a larger number of straying individuals between sites that are geographically proximate and indeed evidence corroborates this prediction \cite{Candy:2000hu,JPE:JPE1383}.
+It is thus reasonable to expect a larger number of straying individuals between sites that are geographically proximate and indeed evidence corroborates this prediction \cite{Candy:2000hu,JPE:JPE1383}.
 Alternatively, salmon that cue to specific environmental conditions may be more likely to stray into sites that have structurally and physiognamically similar habitats \cite{Peterson:2014gy} even if other potential sites are closer.
 These considerations justify imposing a direct relationship between the rate of straying $m$ and habitat heterogeneity: as site dissimilarity increases, so too should the straying rate decrease.
-As site dissimilarity increases, so to should the optima in trait values for the resident populations.
+% As site dissimilarity increases, so to should the optima in trait values for the resident populations.
+When habitat heterogeneity and the rate of straying are linked in this way, we show that very small amounts of either constant or density-dependent straying result in long recovery times for the dominant population because there is time for selection to push the subordinate trait mean away from the optimum of the dominant population (figure \ref{fig:mtheta}).
+%there is enough isolation to allow the surviving subordinate population to adapt towards its own optimum.
+If the straying rate is very low, there is an inversion in the alternative stable states following the disturbance, resulting in a state shift in dominance.
+This may have particular conservation implications for considerations of aiding dispersal following disturbances or reconnections of lost habitats \cite{Anderson:2013bf,Pess:2014isa}.
+% Moreover, local temperature regimes are known to play a central role in dictating local adaptation of salmon populations.
+% If temperature is the primary determinant of $\Delta\theta$ in terms of our model, this would suggest that populations spanning North-South gradients are more predisposed to distance dependent dynamics.
+% In contrast, the relationship between the rate of straying and habitat heterogeneity may be less clear-cut for populations spanning an East-West gradient, where distant habitats may equally fitting for straying populations.
 
-When habitat heterogeneity and the rate of straying are linked, we show that small amounts of either constant or density-dependent straying result in longer recovery times for the dominant population because there is enough isolation to allow the surviving subordinate population to slowly adapt towards its own optimum and therefore strays from that population will be maladaptive in the dominant site.
-As selection pushes the subordinate population away from the dominant trait optimum, the recovery of the extinct population is slowed (figure \ref{fig:mtheta}).
-This may have particular conservation implications for considerations of aiding dispersal following disturbances or reconnections of lost habitats \cite{Anderson:2013bf,Pess:2014isa}.
-Moreover, local temperature regimes are known to play a central role in dictating local adaptation of salmon populations (REFS).
-If temperature is the primary determinant of $\Delta\theta$ in terms of our model, this would suggest that populations spanning North-South gradients are more predisposed to distance dependent dynamics.
-In contrast, the relationship between the rate of straying and habitat heterogeneity may be less clear-cut for populations spanning an East-West gradient, where distant habitats may equally fitting for straying populations.
+% Our model is a simple framework for an initial consideration of the eco-evolutionary dynamics of metapopulation and accordingly made assumptions that are important to clarify. 
+% First, our model did not consider the role of straying and dispersal in introducing new genetic material that can combat genetic challenges of small populations such as inbreeding. 
+% Second, our model did not incorporate the potential genetic underpinnings of response diversity and asynchrony. 
+% We anticipate that including a link between genetic divergence and population stochasticity would exacerbate the rate at which higher straying would synchronize populations. 
+% Third, our model assumed that strays were incorporated into the trait distribution of the recipient populations—in natural systems it is possible that strays may die prior to reproduction when the two habitats are highly divergent.
+% Fourth, natural systems are quite temporally? variable, which could push systems out of steady-state dynamics, which is what our model focused on. 
+% Fifth, we focused on an extremely simple system of two linked populations.
+While more linked populations would introduce additional complexity, we predict that our qualitative results would apply to more complicated systems.
+% It is clear that understanding the eco-evolutionary dynamics of metapopulations is a rich area for further research.
 
-%I'm here
+Although our study was inspired by salmon metapopulations, the results have general implications for the conservation and management of other migratory metapopulations as well. 
+Because changes in straying rates can have large and nonlinear impacts on robustness, human activities that alter straying rates could have unintended consequences. 
+For example, salmon produced by hatcheries often stray into proximate wild populations \cite{Brenner:2012gl}, and these recipient populations can have lower fitness due in part to the introduction of maladapted genes \cite{Ford:2002ip}.
+We show that there is an intermediate straying rate where disturbed populations that are recovering by the introduction of maladapted strays recover fastest: if the straying rate is too low or too high, recovery times increase (figure \ref{fig:relax}).
+This finding suggests that salmon stocking efforts that aim to lower recovery times following dam removal could actually prolong recovery if the rate at which individuals are introduced is not taken into account.
+Ongoing examinations of experimental restocking in the recently re-opened Elwha River (Washington State) will provide empirical insight into the potential short- and long-term consequences of facilitated recovery \cite{Liermann:2017gj}.
 
-%Concluding
-The importance of selection and its influence on the dynamics of populations is increasingly recognized as vital for assessing species' fragility.
-Of particular interest is how selection might influence spatially coupled metapopulations distributed across environmental gradients.
-We have shown that two measures of metapopulation robustness -- the portfolio effect and the time to recovery following a disturbance -- are strongly influenced by the rate at which individuals from a donor population stray into habitats occupied by a recipient population.
-Importantly, density-dependent straying, which may occur when individuals collectively navigate, can both increase the portfolio and lower the time to recovery following a disturbance, which is anticipated to promote persistence.
-We suggest that understanding the spatial complexity of metapopulations dispersing across heterogeneous environments, combined with the mosaic of selective forces acting on those environments, may be key to discovering those factors that promote persistence.
+% More generally, the current attributes and resilience of salmon metapopulations are likely a function of their history of evolving in dynamic river systems that periodically experienced massive disturbances such as landslides, glaciers, and floods over different climatic regimes (Waples et al.
+2009; Waples et al.
+2008).
+% Population declines following these disturbances would also magnify the rate of straying, and our results suggest that this response in the straying rate would quicken the recovery of the small populations that survive.
+% Indeed, we found that density-dependent migration appears to increase the robustness of metapopulations faced with large-scale disturbance, particularly in regimes where populations have lower steady states and correspondingly higher rates of straying.
+% As population numbers increase and stray rates decrease, the metapopulation could achieve higher production due to fine-scale local adaptation unhindered by high rates of maladaptive strays.
+
+The portfolio effect and the time to recovery following a disturbance are independent and correlated measures of metapopulation robustness that take into account both steady state and transient dynamics.
+We show that these measures of robustness are strongly influenced by the rate at which individuals from donor populations stray into habitats occupied by recipient populations. 
+Importantly, density-dependent straying, which may occur when individuals collectively navigate, can both increase the portfolio effect and lower the time to recovery following a disturbance, which is anticipated to promote persistence.
+We suggest that understanding the spatial complexity of metapopulations dispersing across heterogeneous environments, in tandem with the mosaic of selective forces acting on those environments, may be key to uncovering those factors that promote persistence in the wild.
+
+% 
+% 
+% %Concluding
+% The importance of selection and its influence on the dynamics of populations is increasingly recognized as vital for assessing species' fragility.
+% Of particular interest is how selection might influence spatially coupled metapopulations distributed across environmental gradients.
+% We have shown that two measures of metapopulation robustness -- the portfolio effect and the time to recovery following a disturbance -- are strongly influenced by the rate at which individuals from a donor population stray into habitats occupied by a recipient population.
+% Importantly, density-dependent straying, which may occur when individuals collectively navigate, can both increase the portfolio and lower the time to recovery following a disturbance, which is anticipated to promote persistence.
+% We suggest that understanding the spatial complexity of metapopulations dispersing across heterogeneous environments, combined with the mosaic of selective forces acting on those environments, may be key to discovering those factors that promote persistence.
 

### 2018-01-12 manuscript/oldversions/msv4.tex
penultimate

--- 2017-06-03 manuscript/msv4.tex
@@
 The results of our model reveal that increased straying between evolving populations leads to alternative stable states, which has large and nonlinear effects on two measures of metapopulation robustness: the portfolio effect and the time to recovery following an induced disturbance.
-We show that intermediate levels of straying result in large gains in robustness, and that increased differences between sites (increased habitat heterogeneity) promotes robustness when straying rates are low, and erodes robustness when straying rates are high.
+We show that intermediate levels of straying result in large gains in robustness, and that increased habitat heterogeneity promotes robustness when straying rates are low, and erodes robustness when straying rates are high.
 Finally, we show that density-dependent straying promotes robustness, particularly when the aggregate biomass is low and straying is correspondingly high, and this has important ramifications for the conservation of salmon metapopulations facing both natural and anthropogenic disturbances.
-\end{abstract} %250 words
+\end{abstract}
 
 \maketitle
+
+\centerline{Salmon metapopulations, Straying, Dispersal, Eco-evolutionary dynamics, Alternative stable states}
 
@@
 \noindent{\bf (c) The role of habitat heterogeneity and changing selective landscapes}\\ 
-\noindent With the onset of straying, we find that increasingly divergent trait optima generally lower $N_T$ and exaggerate $\Delta N$ (figure \ref{extfig-fig:thetadiffN}), such that the biomass distribution becomes increasingly uneven.
+\noindent With the onset of straying, we find that increasingly divergent trait optima generally lower $N_T^*$ and exaggerate $\Delta N^*$ (figure \ref{extfig-fig:thetadiffN}), such that the biomass distribution becomes increasingly uneven.
 %I've checked this - it's a theshold type (logistic) response Supp fig?
@@
 
-\bibliography{aa_kevin}
+\noindent {\bf Competing interests:} The authors declare no competing interests
+\\
+\noindent {\bf Author contributions:} JDY and JWM conceived of the initial project design.
+JDY and JPG designed models and conducted analyses.
+JDY, JPG, PAHW, and JWM interpreted the results, and drafted and wrote the manuscript.
+\\
+\noindent {\bf Data Accessibility:} Code is made available at https://github.com/jdyeakel/SalmonStrays
+\\
+\noindent {\bf Acknowledgements:} We thank Sean Anderson for helpful discussions and comments on the manuscript.
+J.D.Y.
+was supported by startup funds at the University of California, Merced, and an Omidyar Postdoctoral Fellowship at the Santa Fe Institute.
+J.P.G.
+was supported by a James S.
+McDonnell Foundation Postdoctoral Fellowship at the University of California, Merced.
+P.A.H.W.
+was supported by...
+J.W.M.
+was supported by Simon Fraser University, the Liber Ero Foundation and the Natural Sciences and Engineering Research Council of Canada.
+
+% \bibliography{aa_kevin}
+
+\begin{thebibliography}{10}
+\expandafter\ifx\csname urlstyle\endcsname\relax
+  \providecommand{\doi}[1]{doi:\discretionary{}{}{}#1}\else
+  \providecommand{\doi}{doi:\discretionary{}{}{}\begingroup
+  \urlstyle{rm}\Url}\fi
+
+\bibitem{Brown:1977gk}
+Brown JH, Kodric-Brown A, 1977 {Turnover rates in insular biogeography: effect
+  of immigration on extinction}.
+\newblock \emph{Ecology} \textbf{58}, 445--449
+
+\bibitem{Earn:2000fm}
+Earn DJD, Levin SA, Rohani P, 2000 {Coherence and conservation}.
+\newblock \emph{Science} \textbf{290}, 1360--1364
+
+\bibitem{Anonymous:2015gf}
+Schindler DE, Armstrong JB, Reed TE, 2015 {The portfolio concept in ecology and
+  evolution}.
+\newblock \emph{Front. Ecol. Environ.} \textbf{13}, 257--263
+
+\bibitem{Abbott:2017hl}
+Abbott RE, Doak DF, Peterson ML, 2017 {Portfolio effects, climate change, and
+  the persistence of small populations: analyses on the rare plant Saussurea
+  weberi.}
+\newblock \emph{Ecology} \textbf{98}, 1071--1081
+
+\bibitem{Thibaut:2012km}
+Thibaut LM, Connolly SR, 2013 {Understanding diversity-stability relationships:
+  towards a unified model of portfolio effects.}
+\newblock \emph{Ecol. Lett.} \textbf{16}, 140--150
+
+\bibitem{MilnerGulland:2011vm}
+Milner-Gulland EJ, Fryxell JM, Sinclair ARE, 2011 \emph{{Animal Migration}}.
+\newblock Oxford: Oxford University Press
+
+\bibitem{Bell:2011ki}
+Bell G, Gonzalez A, 2011 {Adaptation and evolutionary rescue in metapopulations
+  experiencing environmental deterioration}.
+\newblock \emph{Science} \textbf{332}, 1327--1330
+
+\bibitem{Carlson:2014is}
+Carlson SM, Cunningham CJ, Westley PAH, 2014 {Evolutionary rescue in a changing
+  world}.
+\newblock \emph{Trends Ecol. Evol.} \textbf{29}, 521--530
+
+\bibitem{Muhlfeld:2014hs}
+Muhlfeld CC, Kovach RP, Jones LA, Al-Chokhachy R, Boyer MC, Leary RF, Lowe WH,
+  Luikart G, Allendorf FW, 2014 {Invasive hybridization in a threatened species
+  is accelerated by climate change}.
+\newblock \emph{Nature Climate Change} \textbf{4}, 620--624
+
+\bibitem{Lowe:2015ft}
+Lowe WH, Muhlfeld CC, Allendorf FW, 2015 {Spatial sorting promotes the spread
+  of maladaptive hybridization}.
+\newblock \emph{Trends Ecol. Evol.} \textbf{30}, 456--462
+
+\bibitem{H:2013fs}
+Westley PAH, Quinn TP, 2013 {Rates of straying by hatchery-produced Pacific
+  salmon (\emph{Oncorhynchus} spp.) and steelhead (\emph{Oncorhynchus mykiss})
+  differ among species, life history types, and populations}.
+\newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{70}, 735--746
+
+\bibitem{Keefer:2014gg}
+Keefer ML, Caudill CC, 2014 {Homing and straying by anadromous salmonids: a
+  review of mechanisms and rates}.
+\newblock \emph{Rev Fish Biol Fisher} \textbf{24}, 333--368
+
+\bibitem{Bett:2017ha}
+Bett NN, Hinch SG, Burnett NJ, Donaldson MR, Naman SM, 2017 {Causes and
+  consequences of straying into small populations of Pacific salmon}.
+\newblock \emph{Fisheries} \textbf{42}, 220--230
+
+\bibitem{Berdahl:2015kv}
+Berdahl A, Torney CJ, Schertzer E, Levin SA, 2015 {On the evolutionary
+  interplay between dispersal and local adaptation in heterogeneous
+  environments}.
+\newblock \emph{Evolution} \textbf{69}, 1390--1405
+
+\bibitem{Berdahl:2016dx}
+Berdahl A, 2016 {Collective behavior as a driver of critical transitions in
+  migratory populations}.
+\newblock \emph{Movement Ecology} \textbf{4}, 1--12
+
+\bibitem{Simons:2004jo}
+Simons AM, 2004 {Many wrongs: The advantage of group navigation}.
+\newblock \emph{Trends Ecol. Evol.} \textbf{19}, 453--455
+
+\bibitem{Brenner:2012gl}
+Brenner RE, Moffitt SD, Grant WS, 2012 {Straying of hatchery salmon in Prince
+  William Sound, Alaska}.
+\newblock \emph{Environ Biol Fish} \textbf{94}, 179--195
+
+\bibitem{Johnson:2012fe}
+Johnson RC, Weber PK, Wikert JD, Workman ML, MacFarlane RB, Grove MJ, Schmitt
+  AK, 2012 {Managed metapopulations: do salmon hatchery
+  {\textquoteleft}sources{\textquoteright} lead to in-river
+  {\textquoteleft}sinks{\textquoteright} in conservation?}
+\newblock \emph{PLoS ONE} \textbf{7}, e28880--11
+
+\bibitem{Fullerton:2011ii}
+Fullerton AH, Lindley ST, Pess GR, Feist BE, Steel EA, Mcelhany P, 2011 {Human
+  influence on the spatial structure of threatened Pacific salmon
+  metapopulations}.
+\newblock \emph{Conserv Biol} \textbf{25}, 932--944
+
+\bibitem{Quinn:2011tf}
+Quinn TP, 2011 \emph{{The Behavior and Ecology of Pacific Salmon and Trout}}.
+\newblock Vancouver: UBC Press
+
+\bibitem{Jonsson:2011kg}
+Jonsson B, Jonsson N, 2011 \emph{{Ecology of Atlantic Salmon and Brown Trout}}.
+\newblock Dordrecht: Springer Netherlands
+
+\bibitem{Quinn:1993ge}
+Quinn TP, 1993 {A review of homing and straying of wild and hatchery-produced
+  salmon}.
+\newblock \emph{Fish Res} \textbf{18}, 29--44
+
+\bibitem{Hendry:2004wf}
+Hendry AP, Bohlin T, Jonsson B, Berg OK, 2004 {The evolution of philopatry and
+  dispersal: homing versus straying in salmonids}.
+\newblock In AP~Hendry, SC~Stearns, eds., \emph{Evolution Illuminated}. Oxford:
+  Oxford University Press on Demand
+
+\bibitem{Schtickzelle:2007wb}
+Schtickzelle N, Quinn TP, 2007 {A metapopulation perspective for salmon and
+  other anadromous fish}.
+\newblock \emph{Fish Fish.} \textbf{8}, 297--314
+
+\bibitem{Anderson:2014cx}
+Anderson SC, Moore JW, McClure MM, Dulvy NK, Cooper AB, 2015 {Portfolio
+  conservation of metapopulations under climate change}.
+\newblock \emph{Ecol. Appl.} \textbf{25}, 559--572
+
+\bibitem{Keefer:2008bs}
+Keefer ML, Caudill CC, Peery CA, Lee SR, 2008 {Transporting juvenile salmonids
+  around dams impairs adult migration}.
+\newblock \emph{Ecol. Appl.} \textbf{18}, 1888--1900
+
+\bibitem{Westley:2015to}
+Westley PAH, Dittman AH, Ward EJ, Quinn TP, 2015 {Signals of climate,
+  conspecific density, and watershed features in patterns of homing and
+  dispersal by Pacific salmon.}
+\newblock \emph{Ecology} \textbf{96}, 2823--2833
+
+\bibitem{Bond:2016dz}
+Bond MH, Westley PAH, Dittman AH, Holecek D, Marsh T, Quinn TP, 2016 {Combined
+  effects of barge transportation, river environment, and rearing location on
+  straying and migration of adult snake river fall-run Chinook salmon}.
+\newblock \emph{T Am Fish Soc} \textbf{146}, 60--73
+
+\bibitem{Peterson:2014gy}
+Peterson DA, Hilborn R, Hauser L, 2014 {Local adaptation limits lifetime
+  reproductive success of dispersers in a wild salmon metapopulation}.
+\newblock \emph{Nat Commun} \textbf{5}, 3696
+
+\bibitem{Moore:2010gs}
+Moore JW, McClure M, Rogers LA, Schindler DE, 2010 {Synchronization and
+  portfolio performance of threatened salmon}.
+\newblock \emph{Conserv. Lett.} \textbf{3}, 340--348
+
+\bibitem{Carlson:2011ce}
+Carlson SM, Satterthwaite WH, Fleming IA, 2011 {Weakened portfolio effect in a
+  collapsed salmon population complex}.
+\newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{68}, 1579--1589
+
+\bibitem{Braun:2016ib}
+Braun DC, Moore JW, Candy J, Bailey RE, 2016 {Population diversity in salmon:
+  linkages among response, genetic and life history diversity}.
+\newblock \emph{Ecography} \textbf{39}, 317--328
+
+\bibitem{simpson1953major}
+Simpson GG, 1953 \emph{{The Major Features of Evolution}}.
+\newblock New York: Simon and Schuster
+
+\bibitem{Lande:1976ga}
+Lande R, 1976 {Natural selection and random genetic drift in phenotypic
+  evolution}.
+\newblock \emph{Evolution} \textbf{30}, 314--334
+
+\bibitem{Shelton:2011eq}
+Shelton AO, Mangel M, 2011 {Fluctuations of fish populations and the magnifying
+  effects of fishing.}
+\newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{108}, 7075--7080
+
+\bibitem{Jasper:2013cc}
+Jasper JR, Habicht C, Moffitt S, Brenner R, Marsh J, Lewis B, Fox EC, Grauvogel
+  Z, Olive SDR, Grant WS, 2013 {Source-sink estimates of genetic introgression
+  show influence of hatchery strays on wild chum salmon populations in Prince
+  William Sound, Alaska}.
+\newblock \emph{PLoS ONE} \textbf{8}, e81916
+
+\bibitem{Schreiber:2011wx}
+Schreiber SJ, B{\"u}rger R, Bolnick DI, 2011 {The community effects of
+  phenotypic and genetic variation within a predator population.}
+\newblock \emph{Ecology} \textbf{92}, 1582--1593
+
+\bibitem{Gilbert:2014ee}
+Gibert JP, Brassil CE, 2014 {Individual phenotypic variation reduces
+  interaction strengths in a consumer{\textendash}resource system}.
+\newblock \emph{Ecol Evol} \textbf{4}, 3703--3713
+
+\bibitem{Gibert:2015kc}
+Gibert JP, DeLong JP, 2015 {Individual variation decreases interference
+  competition but increases species persistence}.
+\newblock \emph{Adv Ecol Res} \textbf{52}, 45--64
+
+\bibitem{Berdahl:2014bl}
+Berdahl A, Westley PAH, Levin SA, Couzin ID, Quinn TP, 2014 {A collective
+  navigation hypothesis for homeward migration in anadromous salmonids}.
+\newblock \emph{Fish Fish.} \textbf{17}, 525--542
+
+\bibitem{Candy:2000hu}
+Candy JR, Beacham TD, 2000 {Patterns of homing and straying in southern British
+  Columbia coded-wire tagged Chinook salmon (\emph{Oncorhynchus tshawytscha})
+  populations}.
+\newblock \emph{Fish Res} \textbf{47}, 41--56
+
+\bibitem{JPE:JPE1383}
+Schick RS, Lindley ST, 2007 {Directed connectivity among fish populations in a
+  riverine network}.
+\newblock \emph{J. Appl. Ecol.} \textbf{44}, 1116--1126
+
+\bibitem{Schindler:2015gf}
+Schindler DE, Armstrong JB, Reed TE, 2015 {The portfolio concept in ecology and
+  evolution}.
+\newblock \emph{Front. Ecol. Environ.} \textbf{13}, 257--263
+
+\bibitem{Ovaskainen:2002il}
+Ovaskainen O, Hanski I, 2002 {Transient dynamics in metapopulation response to
+  perturbation}.
+\newblock \emph{Theor Popul Biol} \textbf{61}, 285--295
+
+\bibitem{Anderson:2013gb}
+Anderson SC, Cooper AB, Dulvy NK, 2013 {Ecological prophets: quantifying
+  metapopulation portfolio effects}.
+\newblock \emph{Methods Ecol Evol} \textbf{4}, 971--981
+
+\bibitem{Loreau:2008ju}
+Loreau M, de~Mazancourt C, 2008 {Species synchrony and its drivers: neutral and
+  nonneutral community dynamics in fluctuating environments}.
+\newblock \emph{Am. Nat.} \textbf{172}, E48--E66
+
+\bibitem{Yeakel:2013vz}
+Yeakel JD, Moore JW, Guimar{\~a}es~Jr PR, de~Aguiar MAM, 2014 {Synchronisation
+  and stability in river metapopulation networks}.
+\newblock \emph{Ecol. Lett.} \textbf{17}, 273--283
+
+\bibitem{GuckHolmes}
+Guckenheimer J, Holmes P, 1983 \emph{{Nonlinear Oscillations, Dynamical
+  Systems, and Bifurcations of Vector Fields}}.
+\newblock New York: Springer
+
+\bibitem{Carlson:2008hl}
+Carlson SM, Seamons TR, 2008 {A review of quantitative genetic components of
+  fitness in salmonids: implications for adaptation to future change}.
+\newblock \emph{Evol Appl} \textbf{1}, 222--238
+
+\bibitem{Scheffer:2009gg}
+Scheffer M, Bascompte J, Brock WA, Brovkin V, Carpenter SR, Dakos V, Held H,
+  van Nes EH, Rietkerk M, Sugihara G, 2009 {Early-warning signals for critical
+  transitions}.
+\newblock \emph{Nature} \textbf{461}, 53--59
+
+\bibitem{Lade:2012eu}
+Lade SJ, Gross T, 2012 {Early warning signals for critical transitions: A
+  generalized modeling approach}.
+\newblock \emph{PLoS Comp. Biol.} \textbf{8}, e1002360
+
+\bibitem{Anonymous:2013br}
+Boettiger C, Ross N, Hastings A, 2013 {Early warning signals: the charted and
+  uncharted territories}.
+\newblock \emph{Theor. Ecol.} \textbf{6}, 255--264
+
+\bibitem{Dakos:2014br}
+Dakos V, Bascompte J, 2014 {Critical slowing down as early warning for the
+  onset of collapse in mutualistic communities}.
+\newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{111}, 17546--17551
+
+\bibitem{Krkosek:2014ch}
+Krko{\v s}ek M, Drake JM, 2014 {On signals of phase transitions in salmon
+  population dynamics}.
+\newblock \emph{Proc. Roy. Soc. B} \textbf{281}, 20133221
+
+\bibitem{Kuehn:2010p2591}
+Kuehn C, 2011 {A mathematical framework for critical transitions: bifurcations,
+  fast-slow systems and stochastic dynamics}.
+\newblock \emph{Physica D} \textbf{240}, 1020--1035
+
+\bibitem{Donohue:2013iu}
+Donohue I, Petchey OL, Montoya JM, Jackson AL, McNally L, Viana M, Healy K,
+  Lurgi M, O'Connor NE, Emmerson MC, 2013 {On the dimensionality of ecological
+  stability.}
+\newblock \emph{Ecol. Lett.} \textbf{16}, 421--429
+
+\bibitem{Levin:1992p1472}
+Levin SA, 1992 {The problem of pattern and scale in ecology}.
+\newblock \emph{Ecology} \textbf{73}, 1943--1967
+
+\bibitem{Meinhardt:2012hw}
+Meinhardt H, 2012 {Turing's theory of morphogenesis of 1952 and the subsequent
+  discovery of the crucial role of local self-enhancement and long-range
+  inhibition}.
+\newblock \emph{Interface Focus} \textbf{2}, 407--416
+
+\bibitem{Hanski:1999vu}
+Hanski I, 1999 \emph{{Metapopulation Ecology}}.
+\newblock Oxford: Oxford University Press
+
+\bibitem{Boughton:1999fa}
+Boughton DA, 1999 {Empirical evidence for complex source{\textendash}sink
+  dynamics with alternative states in a butterfly metapopulation}.
+\newblock \emph{Ecology} \textbf{80}, 2727--2739
+
+\bibitem{Secor:2009ena}
+Secor DH, Kerr LA, Cadrin SX, 2009 {Connectivity effects on productivity,
+  stability, and persistence in a herring metapopulation model}.
+\newblock \emph{ICES J Mar Sci} \textbf{66}, 1726--1732
+
+\bibitem{Christie:2012bj}
+Christie MR, Marine ML, French RA, Blouin MS, 2012 {Genetic adaptation to
+  captivity can occur in a single generation.}
+\newblock \emph{Proc. Natl. Acad. Sci. USA} \textbf{109}, 238--242
+
+\bibitem{Chilcote:2003bb}
+Chilcote MW, 2003 {Relationship between natural productivity and the frequency
+  of wild fish in mixed spawning populations of wild and hatchery steelhead
+  (\emph{Oncorhynchus mykiss})}.
+\newblock \emph{Can. J. Fish. Aquat. Sci.} \textbf{60}, 1057--1067
+
+\bibitem{Araki:2007cm}
+Araki H, Cooper B, Blouin MS, 2007 {Genetic effects of captive breeding cause a
+  rapid, cumulative fitness decline in the wild}.
+\newblock \emph{Science} \textbf{318}, 100--103
+
+\bibitem{Westley:2012ui}
+Westley PAH, { Corinne M Conway}, Fleming IA, 2012 {Phenotypic divergence of
+  exotic fish populations is shaped by spatial proximity and habitat
+  differences across an invaded landscape}.
+\newblock \emph{Evol. Ecol. Res.} \textbf{14}, 147--167
+
+\bibitem{Anderson:2013bf}
+Anderson JH, Faulds PL, Atlas WI, Quinn TP, 2013 {Reproductive success of
+  captively bred and naturally spawned Chinook salmon colonizing newly
+  accessible habitat}.
+\newblock \emph{Evol Appl} \textbf{6}, 165--179
+
+\bibitem{Pess:2014isa}
+Pess GR, Quinn TP, Gephard SR, Saunders R, 2014 {Re-colonization of Atlantic
+  and Pacific rivers by anadromous fishes: linkages between life history and
+  the benefits of barrier removal}.
+\newblock \emph{Rev Fish Biol Fisher} \textbf{24}, 881--900
+
+\bibitem{Ford:2002ip}
+Ford MJ, 2002 {Selection in captivity during supportive breeding may reduce
+  fitness in the wild}.
+\newblock \emph{Conserv Biol} \textbf{16}, 815--825
+
+\bibitem{Liermann:2017gj}
+Liermann M, PESS G, McHenry M, McMillan J, Elofson M, Bennett T, Moses R, 2017
+  {Relocation and recolonization of Coho salmon \emph{Oncorhynchus kisutch} in
+  two tributaries to the elwha river: Implications for management and
+  monitoring}.
+\newblock \emph{T Am Fish Soc} \textbf{17}, 1--39
+
+\end{thebibliography}
+
+\clearpage
 
@@
 \caption{
-(a) Total means $N_t$, 
-(b) difference in means $\Delta N$, and 
+(a) Total means $N_t^*$, 
+(b) difference in means $\Delta N^*$, and 
 (c) the portfolio effect PE as a function of heritability $h^2$ and a constant stray rate $m$. Light colors = high values.
@@
 \caption{
-Recovery time of $N_T$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.2$.
+Recovery time of $N_T^*$ following the extinction of either the low-density (light gray) or high-density (gray) population, or the near-collapse of both (dark gray) assuming (a) constant straying rates $m$ and (b) density-dependent straying rates (evaluated at the steady state $m^*$) with trait heritability $h^2=0.2$.
 If $m$ is density-dependent, in the alternative stable state regime there are two straying rates observed: one each for the low- and high-density populations, respectively, which are linked by a horizontal line.

### 2018-01-17 Manuscript/FinalDraft3/msv10.tex
found error in figure 3 - fixed

--- 2017-12-19 Manuscript/FinalDraft3/msv9.tex
@@
 For example, different responses to climate variability within populations of a rare plant reduced fluctuations in abundance \citep{Abbott:2017hl}. 
-This statistical buffer has traditionally been quantified as the Portfolio Effect (PE), which is the ratio of the population CV to the CV of the aggregated metapopulation \citep{Thibaut:2012km}. 
+This statistical buffer has traditionally been quantified as the Portfolio Effect (PE), which is the ratio of the population coefficient of variation (CV) to the CV of the aggregated metapopulation \citep{Thibaut:2012km}.
 Strengthened portfolio effects are expected to increase the robustness of metapopulations to external disturbances, and by extension promote persistence \citep{Thibaut:2012km}.
@@
 Migratory populations that return to breeding sites for reproduction can be linked to each other by some proportion of the population that permanently disperses into the `wrong' site. 
-Recently, the role of social interactions and collective navigation has been hypothesized  as a mechanism shaping the success of philopatric migrations \citep{Berdahl:2015kv,Berdahl:2016dx,HardestyMoore:wg,Berdahl:2017uu}.
-The rate at which individuals disperse may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \citep{Simons:2004jo,Berdahl:2015kv,Berdahl:2016dx,Berdahl:2017uu}.
+Recently, the role of social interactions and collective navigation has been hypothesized  as a mechanism shaping the success of philopatric migrations \citep{Berdahl:2016dx,HardestyMoore:wg,Berdahl:2017uu}.
+The rate at which individuals disperse may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \citep{Simons:2004jo,Berdahl:2016dx,Berdahl:2017uu}.
 Thus, dispersal rates can be higher at lower population abundances \citep{Berdahl:2014bl}, which may profoundly influence the eco-evolutionary dynamics of metapopulations. 
@@
 This model enables us to explore the tradeoff between the potentially detrimental erosion of local adaptation vs.
-the positive effects of demographic and evolutionary rescue, both of which are facilitated by straying and the effects of collective navigation.
+the positive effects of demographic rescue, both of which are facilitated by straying and the effects of collective navigation.
 
@@
 Moreover, we assume that there is no demographic overlap between generations, consistent with the life history of many populations of pink salmon (\emph{O.
-gorbuscha}) who all mature at two years of age and die after one reproductive season..
-In this sense, both populations serve simultaneously as donor and recipient populations.
+gorbuscha}) who all mature at two years of age and die after one reproductive season.
+% In this sense, both populations serve simultaneously as donor and recipient populations.
 Because total recruitment will be determined by both locals (with a mean trait value closer to the site optimum) and strays (with mean trait values farther from the local optimum), the recruitment of the aggregate is determined by the mean of the trait mix $R_i(\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\theta_i)$, where
@@
   &R_i[\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\theta_i] = \\ \nonumber
-  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i-\theta_i)^2}{2\tau^2}\right\} {\rm pr}(x_i,\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\sigma^2) {\rm d}x_i +\tilde{P}_i\\ \nonumber
+  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i-\theta_i)^2}{2\tau^2}\right\} {\rm p}(x_i,\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\sigma^2) {\rm d}x_i +\tilde{P}_i\\ \nonumber
   &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-(\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t)))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\} +\tilde{P}_i.
@@
 First, we approximate the distribution resulting from the mix of remote and local individuals prior to reproduction as a Gaussian distribution, where $X_i=x_i$ with probability $g(x_i)$.
-The expectation of the actual trait distribution as well as the Gaussian approximation are the same, such that ${\rm E}\{g(x_i)\} = \omega_i\mu_i + (1-\omega_i)\mu_j$, with weights corresponding to the proportion of the mixed population that are local individuals, $\omega_i$, and straying individuals, $1-\omega_i$.
+The expectation of the actual trait distribution as well as the Gaussian approximation are the same, such that ${\rm E}\{X_i\} = \omega_i\mu_i + (1-\omega_i)\mu_j$, with weights corresponding to the proportion of the mixed population that are local individuals, $\omega_i$, and straying individuals, $1-\omega_i$.
 Thus, strays can successfully reproduce and introduce their genotypes into the recipient population, which is supported by observations in wild populations \citep{Jasper:2013cc}.
@@
 with $\mu^\prime = \omega_i \mu_i(t)+ (1-\omega_i)\mu_j(t)$.
-Although trait heritability among salmonids is variable, most life-history traits have an $h^2 < 0.5$ \citep{Carlson:2008hl}, and for all additional analyses we have conservatively set $h^2=0.2$.\\
+Although trait heritability among salmonids is variable, most life-history traits have an $h^2 < 0.5$ \citep{Carlson:2008hl}, and for all additional analyses we have conservatively set $h^2=0.2$.
+Together, equations \ref{eq:N} and \ref{eq:mu} for two linked habitats $i$ and $j$ define the 4-dimensional system of difference equations that describe the eco-evolutionary dynamics of the metapopulation.
+\\
 
@@
 Specifically, straying has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tend to decrease the rate at which individuals err, reducing the overall proportion of a population that strays.
-According to Berdahl et al. \citep{Berdahl:2016dx}, given the probability that an individual strays is $m_0$, the proportion of the local population $N_i(t)$ that strays is
+Following Berdahl et al. \citep{Berdahl:2016dx}, given the probability that an individual strays is $m_0$, the proportion of the local population $N_i(t)$ that strays is
 
@@
 
-\noindent where $C$ is a half-saturation constant, and determines to what extent collective behaviour, as a function of group size, diminishes straying.
+\noindent where $C$ is a half-saturation constant, determining to what extent collective behaviour, as a function of group size, diminishes straying.
 If $C$ is small, relatively small groups of organisms `correct' for even high individual error rates, suppressing straying between sites.
 In the context of our model, small values of $C$ indicate that the effects of collective behaviour on modifying straying are strong.
-As $C \rightarrow \infty$, groups of any size have no impact on straying, such that $m(t) \rightarrow m_0$, and the model reverts to that of constant straying.
-When $C \ll \infty$ and the population density is high, $m(t) \rightarrow 0$, whereas if the population is small, individuals operate without regard to the collective, such that $m(t) \rightarrow m_0$, such that in all cases, $m(t) < m_0$.
+% As $C \rightarrow \infty$, groups of any size have no impact on straying, such that $m(t) \rightarrow m_0$, and the model reverts to that of constant straying.
+% When $C \ll \infty$ and the population density is high, $m(t) \rightarrow 0$, whereas if the population is small, individuals operate without regard to the collective, such that $m(t) \rightarrow m_0$, such that in all cases, $m(t) < m_0$.
+Henceforth, we refer to $C$ as determining the strength of collective behavior: as $C\rightarrow\infty$, the effect of collective behavior becomes weaker, such that the size of the population has no impact on straying, and $m(t)\rightarrow m_0$.
+In contrast, if $C \ll \infty$, the effect of collective behavior is strong, and smaller populations reduce $m(t)$.
+Although the strength of collective behavior depends both on $C$ as well as the probability that an individual strays $m_0$, for a specific value of $m_0$, a lower $C$ reduces straying for smaller population sizes, increasing the strength of collective behavior.
+Accordingly, although the strength of collective behavior is not exclusively determined by $C$, we believe that it is an effective moniker in this case, and use it for this purpose while acknowledging that it is not a perfect descriptor.
 % For realistic population densities, $m(t) < m_0$.\\
@@
 Recovery time was calculated by initiating a disturbance at $t=t_d$, and monitoring $N_T(t_d+t)$ as $t\rightarrow T$, where $T$ is large. 
-The aggregate was deemed recovered at $t_r$, such that recovery time was calculated as $t_r-t_d$, and recovery at $t=t_r$ was determined by the initial $t$ where $N_T(t) < N_T^*\pm{\rm SD}\left( N_T^* \right)$ for $t\in(t_r,T)$, where $\rm{SD}(\cdot)$ is standard deviation (illustrated in figure \ref{fig:recovery}).
+The aggregate was deemed recovered at $t_r$, such that recovery time was calculated as $t_r-t_d$, and recovery at $t=t_r$ was determined by the initial $t$ where $N_T(t) < {\rm E}(N_T^*)\pm{\rm SD}\left( N_T^* \right)$ for $t\in(t_r,T)$, where $\rm{SD}(\cdot)$ is standard deviation (illustrated in figure \ref{fig:recovery}).
+If the system recovers to a different basin of attraction after the perturbation is applied, recovery time is calculated with respect to the newly acquired steady state.
 
@@
 In contrast, when the dominant population goes extinct, the most rapid recovery is associated with minimal straying. 
-It should be noted that when there is no straying ($m = 0$), recovery from extinction is infinite and these values are not shown. 
+It should be noted that when there is no straying ($m = 0$), recovery time is infinite and these values are not shown. 
 The relationship between straying and recovery time in the case of synchronous population disturbance is generally positive.
@@
 
-Observed changes in metapopulation robustness are due to changes in alternative stable state regimes I and II as the strength of collective behaviour increases (lower $C$, figure \ref{fig:bifurcationsddm}a).
+Sharp changes in metapopulation robustness are due to changes in alternative stable state regimes I and II as the strength of collective behaviour increases (lower $C$, figure \ref{fig:bifurcationsddm}a).
 When collective behaviour is weak (large $C$), alternative stable states tend to occur at low-intermediate values of individual straying $m_0$.
@@
 When heterogeneity is low and the effects of collective behaviour are weak such that straying is constant (high $C$), regime II occurs for small-intermediate $m_0$, and regime I does not play a role (figure \ref{fig:bifurcationsddm}b), mirroring the findings of Ronce and Kirkpatrick \citep{Ronce:2001dp}.
-The absence of regime I implies that there are no hidden stable state configurations that might trap a disturbed population in a low-density state.
+The absence of regime I implies that there are no hidden stable state configurations that might trap a disturbed population in an asymmetric low-density state.
 As the strength of collective behaviour increases, regime I appears at a cusp and becomes increasingly dominant with greater individual straying.
@@
 This occurs when groups of $\leq 10^2$ individuals significantly minimizes straying, which is likely unrealistic.
-Moreover, when the effects of collective behaviour are strong, regime II gives way to the dominance of regime I, which harbors low-density basins of attraction (dashed lines in figure \ref{fig:traj}) that can effectively trap disturbed populations in a subordinate stable state, not unlike the allee effects observed in the collective migration model explored by Berdahl et al.
+Moreover, when the effects of collective behaviour are strong, regime II gives way to the dominance of regime I, which harbors low-density basins of attraction (dashed lines in figure \ref{fig:traj}) that can effectively trap disturbed populations in a subordinate stable state, not unlike the Allee effects observed in the collective migration model explored by Berdahl et al.
 \citep{Berdahl:2016dx}.
@@
 If $m_0$ is low ($m_0<0.25$),
-1) the differences between the numerical densities of the subordinate and dominant populations are less exaggerated, and the range of $m_0$ that avoids the alternative stable state regime II is larger (figures \ref{fig:cb}, \ref{fig:bifurcationsddm}a),
-2) the portfolio effect is exaggerated, meaning that the variance of the combined metapopulation is dampening variance observed in the constituent populations (figure \ref{fig:pert}a), and
+1) alternative stable state regime II tends to be avoided for a larger range of $m_0$ (figures \ref{fig:cb}, \ref{fig:bifurcationsddm}a),
+2) the portfolio effect is exaggerated, meaning that the variance of the combined metapopulation has dampened variance relative to its constituent populations (figure \ref{fig:pert}a), and
 3) the time required for the population(s) to recover following an induced disturbance is lower (figure \ref{fig:pert}b).
-Interestingly, the largest portfolio effects are observed when straying is just large enough to enter regime II, where one population assumes a subordinate state.
+Interestingly, the largest portfolio effects are observed when straying is just large enough to enter regime II, where one population assumes a subordinate state, and the differences between the subordinate and dominant population densities is largest.
 This does not appear, in fact, to be a robust condition because the system relies to a large extent on the dominant population as the source, whereas the subordinate population assumes the role of a sink.
 However, recovery time was measured with respect to the aggregate biomass of the metapopulation ($N_T = N_i + N_j$), and despite the source-sink dynamics that emerge in regime II, the aggregate biomass of the system recovers more quickly in this region following the near-collapse of both populations (figure \ref{fig:mpert}b).
-If the perspective is shifted to the robustness of the aggregate biomass of the metapopulation, the existence of asymmetric dominant/subordinate alternative stable states could in one sense be considered robust, despite the fact that one population is always at greater risk of stochastic extinction.
+From this perspective, the existence of asymmetric dominant/subordinate alternative stable states could be considered to be more robust with respect to the recovery time of the total biomass, or less robust because one population is always at greater risk of stochastic extinction.
 
@@
 
-\bibitem{Berdahl:2015kv}
-Berdahl A, Torney CJ, Schertzer E, Levin SA, 2015 {On the evolutionary
-  interplay between dispersal and local adaptation in heterogeneous
-  environments}.
-\newblock \emph{Evolution} \textbf{69}, 1390--1405
+% \bibitem{Berdahl:2015kv}
+% Berdahl A, Torney CJ, Schertzer E, Levin SA, 2015 {On the evolutionary
+%   interplay between dispersal and local adaptation in heterogeneous
+%   environments}.
+% \newblock \emph{Evolution} \textbf{69}, 1390--1405
 
@@
 Measures of metapopulation robustness are shown with respect to different induced disturbances: the near-collapse of both populations (black), and the lone extinction of either the dominant (dark gray) or subordinate (light gray) population.
-Portfolio effects are different for the near-collapse and single extinction scenarios due to one population becoming trapped in a low-density basin of attraction in the alternative stable state regime I.
+Portfolio effects are different for the near-collapse and single extinction scenarios due to different CVs for the populations and aggregate in alternative basins of attraction that exist in regimes I and II.
 } \label{fig:mpert}
@@
 \caption{
-Measures of metapopulation robustness for the density-dependent straying model as a function of individual straying $m_0$ and the strength of collective behaviour $C$ (note the $\rm log_10$ scale, including
+Measures of metapopulation robustness for the density-dependent straying model as a function of individual straying $m_0$ and the strength of collective behaviour $C$ (note the $\rm log_{10}$ scale, including
 a) the Portfolio effect,
@@
 \caption{
-Alternative stable state regimes I (gray) and II (black) as a function of individual straying $m_0$ and the strength of collective behaviour $C$ (note the $\rm log_10$ scale).
+Alternative stable state regimes I (gray) and II (black) as a function of individual straying $m_0$ and the strength of collective behaviour $C$ (note the $\rm log_{10}$ scale).
 Regime I signifies parameter space where there is either 1) an intermediate-density, symmetric, steady state, or 2) an asymmetric dominant/subordinate density.
@@
 \caption{
-Alternative stable state regimes I (gray) and II (black) as a function of individual straying $m_0$ and the strength of collective behaviour $C$ (note the $\rm log_10$ scale), for the case where individual straying increases with lower habitat heterogeneity (inset).
+Alternative stable state regimes I (gray) and II (black) as a function of individual straying $m_0$ and the strength of collective behaviour $C$ (note the $\rm log_{10}$ scale), for the case where individual straying increases with lower habitat heterogeneity (inset).
 Regime I signifies parameter space where there is either 1) an intermediate-density, symmetric, steady state, or 2) an asymmetric dominant/subordinate density.

### 2018-01-17 model/model2/FigureS5_S9_S10_S14.jl

@@ -114,11 +114,11 @@ library(RColorBrewer)
 cols = brewer.pal(3,'Set1')
 pdf($namespace,height=5,width=6)
 plot($(mvec[1:midpoint]),$(n1mean[1:midpoint]),col=cols[1],xlab="Straying (m)",ylab="Steady state biomass",xlim=c(0,0.5),ylim=c(1,400),type='l')
-lines($(mvec[1:midpoint]),$(n2mean[1:midpoint]),col=cols[2])
-lines($(mvec[midpoint+1:length(mvec)]),$(n1mean[midpoint+1:length(n1mean)]),col=cols[1],lty=2)
-lines($(mvec[midpoint+1:length(mvec)]),$(n2mean[midpoint+1:length(n2mean)]),col=cols[2],lty=2)
-text(0.04,380,'RII', xpd=TRUE)
-text(0.18,380,'RI', xpd=TRUE)
+points($(mvec[1:midpoint]),$(n2mean[1:midpoint]),col=cols[2])
+points($(mvec[midpoint+1:length(mvec)]),$(n1mean[midpoint+1:length(n1mean)]),col=cols[1],lty=2)
+points($(mvec[midpoint+1:length(mvec)]),$(n2mean[midpoint+1:length(n2mean)]),col=cols[2],lty=2)
+text(0.04,380,'RI', xpd=TRUE)
+text(0.18,380,'RII', xpd=TRUE)
 types = c('Increasing m','Decreasing m')
 #legend(x=0.1,y=500,legend=types,col='black',lty=c(1,2),xpd=TRUE,cex=0.9, bty="n")
 dev.off()

### 2018-01-19 Manuscript/FinalDraft3/msv10.tex
R4 edits

@@
 For example, different responses to climate variability within populations of a rare plant reduced fluctuations in abundance \citep{Abbott:2017hl}. 
-This statistical buffer has traditionally been quantified as the Portfolio Effect (PE), which is the ratio of the population coefficient of variation (CV) to the CV of the aggregated metapopulation \citep{Thibaut:2012km}.
+This statistical buffer has traditionally been quantified as the Portfolio Effect (PE), which is the ratio of the population's coefficient of variation (CV) to the CV of the aggregated metapopulation \citep{Thibaut:2012km}.
 Strengthened portfolio effects are expected to increase the robustness of metapopulations to external disturbances, and by extension promote persistence \citep{Thibaut:2012km}.
@@
 
-%Alternative stable states
-% \noindent Density-independent straying lowers stable state densities for both populations by (\emph{i}) the numerical loss of locally-adapted individuals from donor to recipient populations and (\emph{ii}) the introduction of maladapted individuals and their associated traits to the recipient population from the donor population (figure \ref{fig:traj}).
-% The decline in steady-state densities is abrupt: as straying increases, the system moves from a single stable state where the populations have the same (symmetric) densities, to a region with alternative stable states.
-% Two types of alternative stable state regimes occur in this system, and have important effects on robustness.
-% The first that is encountered at lower straying $m$, separated by a fold bifurcation \citep{AleksandrovichKuznetsov:1995p2580}, has an intermediate stable state, as well as a dominant and subordinate stable state (regime II in figure \ref{fig:traj}).
-% Because the intermediate state is stable, the dominant and subordinate densities are not encountered unless the system is pushed into basins of attraction that define these states (dashed lines in figure \ref{fig:traj}).
-% The second type of alternative stable state regime (regime I) is separated from regime II by a pitchfork bifurcation.
-% Here the only stable points are the dominant (high) and subordinate (low) stable state densities (regime I in figure \ref{fig:traj}).
-% Importantly, we find that increasing the asymmetry in the vital rates of populations between sites does not significantly alter the presence or position of these different regimes (figure \ref{fig:symmetry}).\\
-
-\noindent At low values of density-independent straying the system approaches a fixed point at which both populations persist at equal population size. 
-As we increase the straying rate a fold bifurcation occurs in which another fixed point is created (figure \ref{fig:traj}, inset). 
-In this alternative state the population sizes are asymmetric. 
-The population with the larger population size, i.e. the dominant population, is well-adapted and reproduces quickly. 
-A small fraction of this population that strays to the subordinate site constitutes a considerable inflow, such that the population in the subordinate site is not as well adapted, which in turn reduces reproduction and stabilizes the asymmetry.
-
-In a regime found at low straying rates (regime I) both the symmetric and the asymmetric state are stable fixed points of the system (figure \ref{fig:traj}, dashed lines). 
+% 
+% \noindent At low values of density-independent straying the system approaches a fixed point at which both populations persist at equal population size. 
+% As we increase the straying rate a fold bifurcation occurs in which another fixed point is created (figure \ref{fig:traj}, inset). 
+% In this alternative state the population sizes are asymmetric. 
+% The population with the larger population size, i.e. the dominant population, is well-adapted and reproduces quickly. 
+% A small fraction of this population that strays to the subordinate site constitutes a considerable inflow, such that the population in the subordinate site is not as well adapted, which in turn reduces reproduction and stabilizes the asymmetry.
+% 
+% In a regime found at low straying rates (regime I) both the symmetric and the asymmetric state are stable fixed points of the system (figure \ref{fig:traj}, dashed lines). 
+% Which of these fixed points is approached depends on the initial conditions. 
+% The dividing line that separates initial conditions that approach the symmetric solution and initial states that approach the asymmetric solution is a separatrix that runs through an unstable fixed point (figure \ref{fig:traj}, inset) that was also created in the fold bifurcation.
+% The set of all initial states from which the system approaches the symmetric state is said to be that state's basin of attraction. 
+% As we increase the straying rate this basin of attraction shrinks and and eventually vanishes. At this point the unstable asymmetric state collides with the stable symmetric state. 
+% A bifurcation occurs in which the the unstable asymmetric state vanishes and the symmetric state is destabilized. 
+% This is a subcritical pitchfork bifurcation that only occurs in systems with symmetries. 
+% After this bifurcation the stable asymmetric state is the only attractor. 
+% 
+% We find a wide regime (regime II) where the system will always approach an asymmetric state where one population is suppressed. 
+% However, if the straying is increased further the imbalance in population sizes becomes harder to maintain. 
+% As the straying rate is increased we reach a critical point where the stable asymmetric fixed point becomes symmetric and collides with the unstable symmetric fixed point. 
+% The system undergoes a bifurcation that is the supercritical form of the subcritical bifurcation encountered earlier. 
+% In this bifurcation the stable asymmetric fixed point vanishes while the unstable symmetric state is stabilized. 
+% After this bifurcation the symmetric state is the only attractor in the system.
+% Importantly, we find that increasing the asymmetry in the vital rates of populations between sites does not significantly alter the presence or position of these different regimes (figure \ref{fig:symmetry}).
+
+\noindent At low values of density-independent straying the system approaches a fixed point at which both populations persist at equal population size, but as we increase straying, other fixed points are created in which the population sizes are asymmetric (figure \ref{fig:traj}, inset).
+The system's underlying symmetry implies that for every asymmetric fixed point there must be another `mirror-image' fixed point in which we find the same population sizes, but where the identities of the populations are reversed.
+Asymmetric fixed points appear in bifurcations as a critical value of the straying parameter is crossed. 
+As the noise in the system is negligible for the purposes of the bifurcation diagram we can use concepts of deterministic bifurcation theory. 
+Based on the Jacobian eigenvalues we conjecture that these bifurcations are fold bifurcations. 
+In a generic fold bifurcation of maps, two new fixed points are created, one of which is unstable, while the other is stable. 
+In the present case, two of these bifurcations occur at the same time, one which creates fixed points where the first population is dominant in the stable fixed point, whereas the second bifurcation creates the mirror-image fixed points where the second population is dominant in the stable fixed point.
+
+In the asymmetric states the dominant population is well-adapted and reproduces quickly. 
+The (small) fraction of this population that strays to the subordinate site constitutes a considerable inflow, such that the population in the subordinate site is not as well adapted, which in turn reduces reproduction and stabilizes the asymmetry.
+In a regime found at low straying rates (regime I) both the symmetric and the asymmetric states are stable fixed points of the system (figure \ref{fig:traj}, dashed lines). 
 Which of these fixed points is approached depends on the initial conditions.
-The dividing line that separates initial conditions that approach the symmetric solution and initial states that approach the asymmetric solution is a separatrix that runs through an unstable fixed point (figure \ref{fig:traj}, inset) that was also created in the fold bifurcation.
-The set of all initial states from which the system approaches the symmetric state is said to be that state's basin of attraction. 
-As we increase the straying rate this basin of attraction shrinks and and eventually vanishes. At this point the unstable asymmetric state collides with the stable symmetric state. 
-A bifurcation occurs in which the the unstable asymmetric state vanishes and the symmetric state is destabilized. 
-This is a subcritical pitchfork bifurcation that only occurs in systems with symmetries. 
-After this bifurcation the stable asymmetric state is the only attractor.
-
+
+As we increase straying, the asymmetric states eventually collide with the stable symmetric state. 
+A subcritical pitchfork bifurcation occurs in which the unstable asymmetric states vanishes and the symmetric state is destabilized. 
+After this bifurcation the stable asymmetric states are the only attractors. 
 We find a wide regime (regime II) where the system will always approach an asymmetric state where one population is suppressed. 
 However, if the straying is increased further the imbalance in population sizes becomes harder to maintain. 
-As the straying rate is increased we reach a critical point where the stable asymmetric fixed point becomes symmetric and collides with the unstable symmetric fixed point. 
-The system undergoes a bifurcation that is the supercritical form of the subcritical bifurcation encountered earlier. 
-In this bifurcation the stable asymmetric fixed point vanishes while the unstable symmetric state is stabilized. 
-After this bifurcation the symmetric state is the only attractor in the system.
-Importantly, we find that increasing the asymmetry in the vital rates of populations between sites does not significantly alter the presence or position of these different regimes (figure \ref{fig:symmetry}).
+Eventually we reach a critical point where the stable asymmetric fixed points become symmetric and collide with the unstable symmetric fixed point. 
+The system undergoes a supercritical pitchfork bifurcation, in which the stable asymmetric fixed points vanish while the unstable symmetric fixed point is stabilized. 
+After this bifurcation the symmetric fixed point is the only attractor in the system. 
+Importantly, we find that increasing the asymmetry in the vital rates of populations between sites does not significantly alter the presence or position of these different regimes (figure \ref{symmetry}).
 
@@
 
-Sharp changes in metapopulation robustness are due to changes in alternative stable state regimes I and II as the strength of collective behaviour increases (lower $C$, figure \ref{fig:bifurcationsddm}a).
+Sharp changes in metapopulation robustness are due to changes in alternative stable state regimes I and II as the strength of collective behaviour increases (lower $C$, figure \ref{fig:bifurcationsddm}b).
 When collective behaviour is weak (large $C$), alternative stable states tend to occur at low-intermediate values of individual straying $m_0$.
 As collective behaviour is strengthened (smaller $C$), Regime II is avoided at lower valus of $m_0$ and expands for higher values of $m_0$.
-When the effects of collective behavior are very strong, Regime II then collapses (black region in figure \ref{fig:bifurcationsddm}a) and gives way to regime I, which plays a larger role over a larger range of $m_0$ when $C$ is low (gray region in figure \ref{fig:bifurcationsddm}a).
+When the effects of collective behavior are very strong, Regime II then collapses (black region in figure \ref{fig:bifurcationsddm}b) and gives way to regime I, which plays a larger role over a larger range of $m_0$ when $C$ is low (gray region in figure \ref{fig:bifurcationsddm}b).
 Importantly, when the strength of collective behaviour is intermediate, both regimes I and II are relevant at low-intermediate $m_0$.
@@
 However, if straying is density-dependent, the strength of collective behaviour has a large influence on the occurrence of both alternative stable state regimes I and II.
-When heterogeneity is low and the effects of collective behaviour are weak such that straying is constant (high $C$), regime II occurs for small-intermediate $m_0$, and regime I does not play a role (figure \ref{fig:bifurcationsddm}b), mirroring the findings of Ronce and Kirkpatrick \citep{Ronce:2001dp}.
+When heterogeneity is low and the effects of collective behaviour are weak such that straying is constant (high $C$), regime II occurs for small-intermediate $m_0$, and regime I does not play a role (figure \ref{fig:bifurcationsddm}a), mirroring the findings of Ronce and Kirkpatrick \citep{Ronce:2001dp}.
 The absence of regime I implies that there are no hidden stable state configurations that might trap a disturbed population in an asymmetric low-density state.
 As the strength of collective behaviour increases, regime I appears at a cusp and becomes increasingly dominant with greater individual straying.
-For sites distributed across more heterogeneous habitats, the alternative stable state regimes I and II expand (figure \ref{fig:bifurcationsddm}a,c).
+For sites distributed across more heterogeneous habitats, the alternative stable state regimes I and II expand (figure \ref{fig:bifurcationsddm}b,c).
 % Only when habitat heterogeneity is intermediate does regime II play a significant role as straying is relatively constant (figure \ref{fig:hystheta}b).
@@
 
-First, if the effects of collective behaviour are very strong (low $C$), metapopulation robustness is increased, due primarily to the avoidance of alternative stable state regime II (figures \ref{fig:pert},\ref{fig:bifurcationsddm}a).
+First, if the effects of collective behaviour are very strong (low $C$), metapopulation robustness is increased, due primarily to the avoidance of alternative stable state regime II (figures \ref{fig:pert},\ref{fig:bifurcationsddm}b).
 This means that - despite potentially high individual error rates - group formation minimizes straying. 
@@
 % Empirical studies suggest that collective behaviour occurs when group sizes are around 100-1000 (REFSXX).
-Here, the system is generally in alternative stable state regime I or II except for perhaps unrealistically low levels of individual straying (figure \ref{fig:bifurcationsddm}a).
+Here, the system is generally in alternative stable state regime I or II except for perhaps unrealistically low levels of individual straying (figure \ref{fig:bifurcationsddm}b).
 If individual straying is high ($m_0>0.25$),
@@
 If $m_0$ is low ($m_0<0.25$),
-1) alternative stable state regime II tends to be avoided for a larger range of $m_0$ (figures \ref{fig:cb}, \ref{fig:bifurcationsddm}a),
+1) alternative stable state regime II tends to be avoided for a larger range of $m_0$ (figures \ref{fig:cb}, \ref{fig:bifurcationsddm}b),
 2) the portfolio effect is exaggerated, meaning that the variance of the combined metapopulation has dampened variance relative to its constituent populations (figure \ref{fig:pert}a), and
@@
 \centering
-\includegraphics[width=0.75\textwidth]{fig_traj.pdf}
+\includegraphics[width=0.75\textwidth]{fig_traj2.pdf}
 \caption{
-The steady-state densities of $N_i$ and $N_j$ vs. straying $m$ for the constant straying model. 
+% The steady-state densities of $N_i$ and $N_j$ vs. straying $m$ for the constant straying model. 
+% Alternative stable states exist for regimes I and II, labeled RI and RII, respectively.
+% Two configurations exist in regime I: a symmetric, intermediate state (solid line), and an asymmetric dominant/subordinate state (red and blue dashed lines, respectively).
+% One configuration exists in regime II: an asymmetric dominant/subordinate state (red and blue lines, respectively).
+% Inset: An illustration of the alternative stable states (solid lines) in regime I (light gray area) and II (dark gray area).
+% The dividing line that separates initial conditions that approach the symmetric (sym.) and asymmetric (asym.) solution is a separatrix (dashed line) that runs through an unstable fixed point.
+The steady-state densities of $N_i$ and $N_j$ vs.~straying $m$ for the constant straying model. 
 Alternative stable states exist for regimes I and II, labeled RI and RII, respectively. 
-Two configurations exist in regime I: a symmetric, intermediate state (solid line), and an asymmetric dominant/subordinate state (red and blue dashed lines, respectively).
-One configuration exists in regime II: an asymmetric dominant/subordinate state (red and blue lines, respectively).
-Inset: An illustration of the alternative stable states (solid lines) in regime I (light gray area) and II (dark gray area).
-The dividing line that separates initial conditions that approach the symmetric (sym.) and asymmetric (asym.) solution is a separatrix (dashed line) that runs through an unstable fixed point.
+In regime I the system can approach qualitatively different states: a symmetric, intermediate state (purple points), and asymmetric dominant/subordinate states (red and blue points, respectively). 
+In regime II only one type of attractor exists: an asymmetric dominant/subordinate state (red and blue points, respectively), and its mirror image where identities of dominant and subordinate are exchanged.
+Inset: A qualitative sketch of the bifurcation diagram, showing the stable (solid lines) and unstable (dashed line) fixed points in regime I (light gray area) and II (dark gray area).
+The symmetric condition (sym.) is the horizontal line at the base of the inset, whereas the asymmetric condition (asym.) is represented by the curved line.
 } \label{fig:traj}
@@
 \centering
-\includegraphics[width=0.75\textwidth]{fig_meandiff.pdf}
+\includegraphics[width=0.75\textwidth]{fig_meandiff2.pdf}
 \caption{
@@
 Low $C$ corresponds to strong effects of collective behaviour.
-The top row shows stable state densities as a function of individual straying $m_0$; the bottom row shows stable state densities as a function of straying at the stable state $m(t)^*$
+The top row shows stable state densities as a function of individual straying $m_0$; the bottom row shows stable state densities as a function of straying at the stable state $m(t)^*$.
+Vertical green lines link paired subordinate and dominant population densities.
 Main: The absolute difference in stable state densities averaged across intervals of low straying ($0 < m,m_0 < 0.25$; blue) and high straying ($0.25 < m,m_0 < 0.5$; red).

### 2018-01-20 Manuscript/FinalDraft3/msv10.tex
asym updates

@@
 In a generic fold bifurcation of maps, two new fixed points are created, one of which is unstable, while the other is stable. 
-In the present case, two of these bifurcations occur at the same time, one which creates fixed points where the first population is dominant in the stable fixed point, whereas the second bifurcation creates the mirror-image fixed points where the second population is dominant in the stable fixed point.
+In the present case, two of these bifurcations occur at the same time, one which creates fixed points where the first population is dominant in the stable fixed point, whereas the second bifurcation creates the mirror-image fixed points where the second population is dominant.
 
@@
 The (small) fraction of this population that strays to the subordinate site constitutes a considerable inflow, such that the population in the subordinate site is not as well adapted, which in turn reduces reproduction and stabilizes the asymmetry.
-In a regime found at low straying rates (regime I) both the symmetric and the asymmetric states are stable fixed points of the system (figure \ref{fig:traj}, dashed lines). 
+In a regime found at low straying rates (regime I) both the symmetric and the asymmetric states are stable fixed points of the system (figure \ref{fig:traj}, purple points). 
 Which of these fixed points is approached depends on the initial conditions.
@@
 After this bifurcation the symmetric fixed point is the only attractor in the system. 
-Importantly, we find that increasing the asymmetry in the vital rates of populations between sites does not significantly alter the presence or position of these different regimes (figure \ref{symmetry}).
+Importantly, we find that increasing the asymmetry in the vital rates of populations between sites does not significantly alter the presence or position of these different regimes (figure \ref{fig:symmetry}).\\
 
@@
 \centering
-\includegraphics[width=0.75\textwidth]{fig_rtpe_ddm.pdf}
+\includegraphics[width=1\textwidth]{fig_rtpe_ddm.pdf}
 \caption{
@@
 \centering
-\includegraphics[width=0.75\textwidth]{fig_hysteresis_ddm.pdf}
+\includegraphics[width=1\textwidth]{fig_hysteresis_ddm.pdf}
 \caption{

### 2018-01-20 model/KevinSimulations.jl

@@ -12,11 +12,11 @@ beta=0.001;
 theta1=5.0;
 thetadiff=5.0;
 tau=1.0;
-h=0.49;
+h=1.0;
 sigmaE=0;
 sigmaG=1.0;
-m=0.35;
-perror=0.01;
+m=0.49;
+perror=0.05;
 n1, n2, x1, x2, w1, w2 = 
 KevinEvolve(
   tmax,
@@ -83,12 +83,12 @@ beta=0.001;
 theta1=5.0;
 thetadiff=5;
 tau=1.0;
-h=0.8;
+h=0.2;
 sigmaE=0;
-sigmaG=1;
-perror=0.01;
+sigmaG=0.2;
+perror=0.05;
 
-burnin=0.80
+burnin=0.99
 @time for i=1:length(mvec)
   m=mvec[i];
   
@@ -137,10 +137,7 @@ burnin=0.80
   # maxeigs[i] = maximum(re);
   # maximeigs[i] = maximum(im);
   
-  # pe[i] = (mean([std(n1trim),std(n2trim)])/mean([mean(n1trim),mean(n2trim)])) *
-  # (1/(std(n1trim+n2trim)/mean(n1trim+n2trim)))
-  
-  pe[i] = mean([(std(n1trim)/mean(n1trim)),(std(n2trim)/mean(n2trim))])*
+  pe[i] = (mean([std(n1trim),std(n2trim)])/mean([mean(n1trim),mean(n2trim)])) *
   (1/(std(n1trim+n2trim)/mean(n1trim+n2trim)))
   
 end
@@ -154,7 +151,7 @@ points($mvec,$n2mean,pch='.',col=cols[2],cex=0.5)
 
 
 #Steady state plot
-namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/figs/fig_SSm.pdf");
+namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/fig_SSm.pdf");
 R"""
 library(RColorBrewer)
 cols = brewer.pal(3,'Set1')
@@ -164,7 +161,7 @@ points($mvec,$n2mean,pch='.',col=cols[2],cex=0.5)
 dev.off()
 """
 #Trait offset plot
-namespace = string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/manuscript/figs/fig_Traitm.pdf");
[…]

### 2018-01-20 model/model2/Figure_4regimes_ddm.jl

@@ -5,6 +5,10 @@ using Distributions, RCall
 @everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolve_ddm.jl")
 @everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinJacobian.jl")
 
+@everywhere include("$(homedir())/src/KevinEvolve.jl")
+@everywhere include("$(homedir())/src/KevinEvolve_ddm.jl")
+@everywhere include("$(homedir())/src/KevinJacobian.jl")
+
 #Analysis over m
 tmax=10000;
 mvec = collect(0.0:0.001:0.5);
@@ -783,15 +787,15 @@ pal = rev(brewer.pal(9,"Blues"))
 pal2 = rev(brewer.pal(11,"Spectral"))
 pdf($namespace,height=8,width=10)
 par(mfrow=c(2,2),mai = c(0.8, 0.8, 0.3, 0.9))
-image.plot(x=$mvec,y=$cexpvec,z=$mpe_ddm,zlim=c(min($(mpe_ddm[!isnan(mpe_ddm)])),2),col=pal,xlab=expression(paste(m[0])),ylab=expression(paste(log[10],' C')))
+image.plot(x=$mvec,y=$cexpvec,z=$mpe_ddm,zlim=c(min($(mpe_ddm[!isnan(mpe_ddm)])),2),col=pal,xlab=expression(paste('Individual straying ',m[0])),ylab=expression(paste(log[10],' C')))
 text(x=0.25,y=5.2,'Portfolio effect',xpd=T)
-image.plot(x=$mvec,y=$cexpvec,z=log($mrt_ddm,10),zlim=c(min(log($mrt_ddm,10)),2.2),col=pal2,xlab=expression(paste(m[0])),ylab=expression(paste(log[10],' C')))
+image.plot(x=$mvec,y=$cexpvec,z=log($mrt_ddm,10),zlim=c(min(log($mrt_ddm,10)),2.2),col=pal2,xlab=expression(paste('Individual straying ',m[0])),ylab=expression(paste(log[10],' C')))
 text(x=0.43,y=4.8,'Near-collapse',xpd=T,col='black')
 text(x=0.25,y=5.2,expression(paste(log[10],'Recovery time')),xpd=T)
[…]

### 2018-01-21 Manuscript/FinalDraft3/msv10.tex
labels to figure 4

@@
 ${}^2$The Santa Fe Institute, Santa Fe NM, USA \\
-${}^3$University of Bristol, Bristol, UK\\
+${}^3$Department of Engineering Mathematics, University of Bristol, Bristol, UK\\
 ${}^4$College of Fisheries and Ocean Sciences, University of Alaska, Fairbanks, Fairbanks AK, USA \\
@@
 Intraspecific diversity can increase the resilience and stability of species or metapopulations. 
-This diversity-stability linkage occurs when there are asynchronous population dynamics within the metapopulation. 
-This asynchrony will increase the potential for demographic rescue \citep{Brown:1977gk,Earn:2000fm} and also decrease the variability of processes that integrate across the metapopulation \citep{Anonymous:2015gf}.
+This diversity-stability linkage can arise when there are asynchronous population dynamics within the metapopulation. 
+Such asynchrony will increase the potential for demographic rescue \citep{Brown:1977gk,Earn:2000fm} and also decrease the variability of processes that integrate across the metapopulation \citep{Anonymous:2015gf}.
 For example, different responses to climate variability within populations of a rare plant reduced fluctuations in abundance \citep{Abbott:2017hl}. 
@@
   &R_i[\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\theta_i] = \\ \nonumber
-  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i-\theta_i)^2}{2\tau^2}\right\} {\rm p}(x_i,\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\sigma^2) {\rm d}x_i +\tilde{P}_i\\ \nonumber
-  &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-(\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t)))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\} +\tilde{P}_i.
+  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i-\theta_i)^2}{2\tau^2}\right\} g(x_i,\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\sigma^2) {\rm d}x_i +\tilde{P}_i\\ \nonumber
+  &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-(\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t)))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\} +\tilde{P}_i,
 \end{align}
 
-\noindent The mismatch between the mean of the local trait mix $\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t)$ and the local optimum $\theta_i$ scales the recruitment rate for the population, and $\tilde{P}_i\sim {\rm Normal}(0,0.01)$ introduces a small amount of demographic variability.
+\noindent where $g(x_i)$ is the Gaussian probability density function for the trait $x_i$.
+The mismatch between the mean of the local trait mix $\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t)$ and the local optimum $\theta_i$ scales the recruitment rate for the population, and $\tilde{P}_i\sim {\rm Normal}(0,0.01)$ introduces a small amount of demographic variability.
 The parameter $\tau$ is the strength of selection, and controls the sensitivity of recruitment to changes in the mean trait value away from the optimum.

### 2018-01-21 Manuscript/FinalDraft3/supp3.tex
labels to figure 4

@@
 
-We aim to understand how straying between populations may affect phenotypic evolution, and how in turn this evolutionary change may lead to changes in ecological dynamics as they unfold. 
+We aim to understand how straying between populations may affect phenotypic evolution, and how this evolutionary change may in turn lead to changes in ecological dynamics as they unfold. 
 To do so, we build upon Lande's classic formulation of phenotypic evolution \citep{Lande:1976ga}. In what follows, we  
@@
 
-The model in the main text thus assumes the existence of a normally distributed trait in each population with means $\mu_{i}$ and $\mu_{j}$.
-Then straying occurs, and the the populations now have a mixture distribution as in \ref{12}.
-To keep track of the mean of that mix, we assume the mixture distribution to be Gaussian with mean $\overline{x}=\omega_i \mu_{i}+(1-\omega_i)\mu_{j}$, which is true whenever straying is low enough, or the two means are not too different from one another.
-Then selection and reproduction occur at the same time leading to a new generation with a normally distributed trait distribution whose.
-
-In a previous paper by Ronce and Kirkpatrick \citep{Ronce:2001dp}, it was shown that assumptions similar to the ones we make do not qualitatively alter the ecological or evolutionary dynamics observed.
+The model used in the main text thus assumes the existence of a normally distributed trait in each population with means $\mu_{i}$ and $\mu_{j}$. 
+After straying occurs, the populations now have a mixture distribution as in \ref{12}. 
+To keep track of the mean of that mix, we approximate the mixture distribution as a Gaussian with mean $\overline{x}=\omega_i \mu_{i}+(1-\omega_i)\mu_{j}$. 
+Selection and reproduction thus occur at the same time leading to a new generation with a normally distributed trait distribution.
+In a previous paper by Ronce and Kirkpatrick \citep{Ronce:2001dp}, it was shown that assumptions similar to those made here do not qualitatively alter the observed ecological or evolutionary dynamics.
 
@@
 \centering
-\includegraphics[width=0.35\textwidth]{fig_density.png}
+\includegraphics[width=0.35\textwidth]{fig_density2.pdf}
 \caption{

### 2018-01-21 model/model2/Figure_4regimes_ddm.jl

@@ -789,15 +789,19 @@ pdf($namespace,height=8,width=10)
 par(mfrow=c(2,2),mai = c(0.8, 0.8, 0.3, 0.9))
 image.plot(x=$mvec,y=$cexpvec,z=$mpe_ddm,zlim=c(min($(mpe_ddm[!isnan(mpe_ddm)])),2),col=pal,xlab=expression(paste('Individual straying ',m[0])),ylab=expression(paste(log[10],' C')))
 text(x=0.25,y=5.2,'Portfolio effect',xpd=T)
+text(par('usr')[1]-0.088,max($(cexpvec))+0.2,'(a)', xpd=TRUE,cex=1.2)
 image.plot(x=$mvec,y=$cexpvec,z=log($mrt_ddm,10),zlim=c(min(log($mrt_ddm,10)),2.2),col=pal2,xlab=expression(paste('Individual straying ',m[0])),ylab=expression(paste(log[10],' C')))
 text(x=0.43,y=4.8,'Near-collapse',xpd=T,col='black')
-text(x=0.25,y=5.2,expression(paste(log[10],'Recovery time')),xpd=T)
+text(x=0.25,y=5.2,expression(paste(log[10],' Recovery time')),xpd=T)
+text(par('usr')[1]-0.088,max($(cexpvec))+0.2,'(b)', xpd=TRUE,cex=1.2)
 image.plot(x=$mvec,y=$cexpvec,z=log($mrt_ddm_s,10),zlim=c(min(log($mrt_ddm_s,10)),2.2),col=pal2,xlab=expression(paste('Individual straying ',m[0])),ylab=expression(paste(log[10],' C')))
-text(x=0.25,y=5.2,expression(paste(log[10],'Recovery time')),xpd=T)
+text(x=0.25,y=5.2,expression(paste(log[10],' Recovery time')),xpd=T)
 text(x=0.4,y=4.8,'Subordinate extinct',xpd=T,col='black')
+text(par('usr')[1]-0.088,max($(cexpvec))+0.2,'(c)', xpd=TRUE,cex=1.2)
 image.plot(x=$mvec,y=$cexpvec,z=log($mrt_ddm_l,10),zlim=c(min(log($mrt_ddm_l,10)),2.2),col=pal2,xlab=expression(paste('Individual straying ',m[0])),ylab=expression(paste(log[10],' C')))
[…]

### 2018-01-23 Manuscript/FinalDraft3/msv10.tex
coauthor edits

@@
 %\title{Community assembly and dynamics by the deconstruction of species interactions}
-\title{Eco-evolutionary dynamics, density dependent dispersal, and collective behavior: implications for salmon metapopulation robustness}
+\title{Eco-evolutionary dynamics, density-dependent dispersal, and collective behaviour: implications for salmon metapopulation robustness}
 \author{
@@
 The spatial dispersal of individuals plays an important role in the dynamics of populations, and is central to metapopulation theory.
-Although the dispersal of individuals from donor to recipient populations provides connections within the metapopulation, promoting demographic and evolutionary rescue, it may also introduce maladapted individuals into habitats where they are ill-suited, potentially lowering the fitness of recipient populations through introgression of heritable traits.
-To explore this dual nature of dispersal, we modify a well-established eco-evolutionary model of two locally adapted populations and their associated mean trait values, to examine recruiting salmon populations that are connected by density-dependent dispersal, which we assume emerges from collective navigation.
+Dispersal provides connections within metapopulations, promoting demographic and evolutionary rescue, but may also introduce maladapted individuals, potentially lowering the fitness of recipient populations through introgression of heritable traits.
+To explore this dual nature of dispersal, we modify a well-established eco-evolutionary model of two locally adapted populations and their associated mean trait values, to examine recruiting salmon populations that are connected by density-dependent dispersal, consistent with collective migratory behaviour that promotes navigation.
 When the strength of collective behaviour is weak such that straying is effectively constant, we show that a low level of straying is associated with the highest gains in metapopulation robustness and that high straying serves to erode robustness.
 Moreover we find that as the strength of collective behaviour increases, metapopulation robustness is enhanced, but this relationship depends on the rate at which individuals stray.
-We find that strong collective behaviour increases the presence of hidden low-density basins of attraction, which may serve to trap disturbed populations, and that this is exacerbated by increased habitat heterogeneity.
-Taken as a whole, our findings suggest that density dependent straying may be a currently unappreciated adaptation in salmon that have evolved in dynamic landscapes and which has important ramifications for the conservation of salmon metapopulations facing both natural and anthropogenic contemporary disturbances.
+Specifically, strong collective behaviour increases the presence of hidden low-density basins of attraction, which may serve to trap disturbed populations, and this is exacerbated by increased habitat heterogeneity.
+Taken as a whole, our findings suggest that density-dependent straying and collective migratory behaviour may help enable metapopulations such as in salmon thrive in dynamic landscapes.
+Given the pervasive eco-evolutionary impacts of dispersal on metapopulations, these findings have important ramifications for the conservation of salmon metapopulations facing both natural and anthropogenic contemporary disturbances.
 \end{abstract}
@@
 
-Intraspecific diversity can increase the resilience and stability of species or metapopulations. 
+Intraspecific diversity can increase the resilience and stability of species or metapopulations \citep{Gibert:2016cu}. 
 This diversity-stability linkage can arise when there are asynchronous population dynamics within the metapopulation. 
@@
 This statistical buffer has traditionally been quantified as the Portfolio Effect (PE), which is the ratio of the population's coefficient of variation (CV) to the CV of the aggregated metapopulation \citep{Thibaut:2012km}.
-Strengthened portfolio effects are expected to increase the robustness of metapopulations to external disturbances, and by extension promote persistence \citep{Thibaut:2012km}.
+Larger portfolio effects are expected to increase the robustness of metapopulations to external disturbances, and by extension promote persistence \citep{Thibaut:2012km}.
 In contrast, homogenization of populations leading to greater synchronization and weakened PE may be a harbinger of metapopulation collapse and extinction \citep{Carlson:2011ce}.
@@
 % Although dispersal and the consequent gene flow can influence the evolutionary dynamics of metapopulations, the interplay between selection, population dynamics, and persistence is not well understood.
-Movement of individuals among local populations (i.e. dispersal) can have a large influence on metapopulation persistence \citep{MilnerGulland:2011vm}. 
-Dispersal facilitates evolutionary rescue, whereby immigration of individuals with heritable adapative traits can rescue small populations from local extinction in the context of maladaptive environmental change \citep{Bell:2011ki,Carlson:2014is}.
+Permanent movement of individuals among local populations (i.e. dispersal) can have a large influence on metapopulation persistence \citep{MilnerGulland:2011vm}. 
+Dispersal facilitates evolutionary rescue, whereby immigration of individuals with heritable adaptive traits can rescue small populations from local extinction in the context of maladaptive environmental change \citep{Bell:2011ki,Carlson:2014is}.
 Dispersal also enables demographic rescue, when depressed or extirpated populations are recolonized by immigrants from the rest of the metapopulation.
-On the other hand, high rates of dispersal may synchronize populations and increase the risk of extinction of the entire metapopulation \citep{Earn:2000fm,Satterthwaite:2015ge}. 
+On the other hand, high rates of dispersal may synchronize the dynamics of populations and subsequently increase the risk of extinction of the entire metapopulation \citep{Earn:2000fm,Satterthwaite:2015ge}.
 % Dispersal will also influence the evolutionary dynamics of the metapopulation.
 % though how the interplay between selection and population dynamics might impact persistence is less well understood. 
-Lastly, dispersal may also introduce maladapted individuals into habitats that are host to different environmental conditions, possibly lowering the mean fitness of the recipient population \citep{Ronce:2001dp,Muhlfeld:2014hs}.
+Additionally, dispersal may also introduce maladapted individuals into habitats that are host to different environmental conditions, possibly lowering the mean fitness of the recipient population \citep{Ronce:2001dp,Muhlfeld:2014hs}.
 More broadly, dispersal can provide a mechanism by which phenotypes are sorted in space rather than time and facilitates the spread of potentially maladaptive genes \citep{Lowe:2015ft}.
@@
 There is growing appreciation that a combination of abiotic, biotic, and anthropogenic factors can control the rate of dispersal among populations \citep{Travis:2013en,H:2013fs,Keefer:2014gg,Bett:2017ha}.
-Migratory populations that return to breeding sites for reproduction can be linked to each other by some proportion of the population that permanently disperses into the `wrong' site. 
-Recently, the role of social interactions and collective navigation has been hypothesized  as a mechanism shaping the success of philopatric migrations \citep{Berdahl:2016dx,HardestyMoore:wg,Berdahl:2017uu}.
-The rate at which individuals disperse may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \citep{Simons:2004jo,Berdahl:2016dx,Berdahl:2017uu}.
-Thus, dispersal rates can be higher at lower population abundances \citep{Berdahl:2014bl}, which may profoundly influence the eco-evolutionary dynamics of metapopulations. 
+Migratory populations that return to breeding sites for reproduction can be linked to each other by some proportion of the population that disperses into the `wrong' site. 
+Recently, the role of social interactions to lead to collective navigation has been hypothesized  as a mechanism shaping the success of philopatric migrations \citep{Berdahl:2016dx,HardestyMoore:wg,Berdahl:2017uu}.
+The collective navigation hypothesis posits that rate at which individuals disperse may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \citep{Simons:2004jo,Berdahl:2016dx,Berdahl:2017uu}.
+Thus, dispersal rates can be higher at lower population abundances \citep{Berdahl:2014bl}, which can in turn profoundly influence the eco-evolutionary dynamics of metapopulations. 
 % The potential influence of collective dispersal on the dynamics of individual populations and the metapopulation as a whole is a topic of considerable interest that has tangible conservation implications \citep{Brenner:2012gl,Johnson:2012fe,Fullerton:2011ii}.
@@
 The eco-evolutionary impacts of dispersal likely have implications for conservation and management in key taxa such as in migratory salmon \citep{Brenner:2012gl,Johnson:2012fe,Fullerton:2011ii}.
-While anadromous salmonid fishes (genera \emph{Oncorhynchus} and \emph{Salmo}) are renown for returning to their natal spawning habitats with high accuracy and precision after years at sea \citep{Quinn:2011tf,Jonsson:2011kg,Keefer:2014gg}, there are generally some individuals that disperse (termed `straying' and used synonymously with dispersal hereafter) to non-natal sites to spawn \citep{Quinn:1993ge,Hendry:2004wf}.
-Straying is fundamental to the evolutionary legacy of populations as it has provided a mechanism for the colonization of new or connected habitats following glacial retreat or large scale geomorphic landscape change \citep{Hendry:2004wf}.
-Salmon may operate as metapopulations, where populations are genetically distinct but linked by some level of straying \citep{Schtickzelle:2007wb,Anderson:2014cx}.
+While anadromous salmonid fishes (genera \emph{Oncorhynchus} and \emph{Salmo}) are renown for returning to their natal spawning habitats with high accuracy and precision after years at sea \citep{Quinn:2011tf,Jonsson:2011kg,Keefer:2014gg}, while some individuals disperse (termed `straying' and used synonymously with dispersal hereafter) to non-natal sites to spawn \citep{Quinn:1993ge,Hendry:2004wf}.
+Straying provides a mechanism for the colonization of new or connected habitats following glacial retreat or large-scale geomorphic landscape change \citep{Hendry:2004wf}.
+Salmon appear to operate as metapopulations, where populations are in part reproductively isolated in discrete habitat patches, but linked by some level of straying \citep{Schtickzelle:2007wb,Anderson:2014cx}.
 Although extensive work has been done to document the extent of straying from donor populations into recipient populations \citep{Keefer:2014gg,Bett:2017ha}, only recently have the abiotic, biotic, and anthropogenic influences of straying behaviours been investigated systemically \citep{Keefer:2008bs,Westley:2015to,Bond:2016dz}.
@@
 To address this question we build upon an established eco-evolutionary model of two populations occupying different sites that are linked by straying individuals, each with an associated trait distribution subject to natural selection determined by local conditions \citep{Ronce:2001dp}.
-Specifically we compared (a) density-independent (constant) straying with (b) density-dependent straying as a function of the rate at which individuals stray and the strength of collective behaviour across (c) increasing environmental heterogeneity, by assessing two measures of metapopulation robustness: the portfolio effect and the time required for a population(s) to recover following an induced disturbance.
+Specifically we compared (a) density-independent (constant) straying with (b) density-dependent straying as a function of the rate at which individuals stray and the strength of collective behaviour across (c) increasing environmental heterogeneity, by assessing two measures of metapopulation robustness: the portfolio effect and the time required for recovery following an induced disturbance.
 This model enables us to explore the tradeoff between the potentially detrimental erosion of local adaptation vs.
-the positive effects of demographic rescue, both of which are facilitated by straying and the effects of collective navigation.
+the positive effects of demographic rescue, both of which are facilitated by straying and potentially moderated by the effects of collective navigation.
 
@@
 \noindent{\bf (a) Metapopulation framework}\\
-\noindent We follow the basic framework described by Ronce and Kirkpatrick \citep{Ronce:2001dp}, where dispersal connects two populations $N_i$ and $N_j$ that inhabit two distinct habitats, each with trait values $x_i$ and $x_j$.
-In our version of the model, $N_i$ and $N_j$ are locally adapted to site-specific conditions such that there is an optimum trait value $\theta_i$ and $\theta_j$ associated with each habitat, where recruitment is maximized if the trait value of the local population equals the optimum, such that $x = \theta$.
+\noindent We follow the basic framework described by Ronce and Kirkpatrick \citep{Ronce:2001dp}, where dispersal connects two populations $i$ and $j$ that inhabit two distinct habitats, each with densities $N_i$ and $N_j$ and trait values $x_i$ and $x_j$.
+In our version of the model, $i$ and $j$ are locally adapted to site-specific conditions such that there is an optimum trait value $\theta_i$ and $\theta_j$ associated with each habitat, where recruitment is maximized if the trait value of the local population equals the optimum ($x = \theta$).
 Moreover, we assumed that $x_{i,j}$ are normally distributed with means $\mu_{i,j}$ and have the same standard deviation $\sigma$.
@@
 The mismatch between the mean of the local trait mix $\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t)$ and the local optimum $\theta_i$ scales the recruitment rate for the population, and $\tilde{P}_i\sim {\rm Normal}(0,0.01)$ introduces a small amount of demographic variability.
-The parameter $\tau$ is the strength of selection, and controls the sensitivity of recruitment to changes in the mean trait value away from the optimum.
+The parameter $\tau$ is the strength of selection and controls the sensitivity of recruitment to changes in the mean trait value away from the optimum.
 Because straying individuals are emigrating from a population with a mean trait value farther from the local optimum, their rate of recruitment is diminished.
-Recent studies of wild sockeye salmon have indeed found that straying individuals have lower life-time fitness than individuals that do not stray, although it is unknown at what life-stage this selection occurs \citep{Peterson:2014gy}.
+Recent studies of wild sockeye salmon have indeed found that straying individuals have lower lifetime fitness than individuals that do not stray, although it is unknown at what life stage this selection occurs \citep{Peterson:2014gy}.
 % The compensatory effects are then determined by the exponential, following the Ricker stock-recruitment relationship.
@@
 Although trait heritability among salmonids is variable, most life-history traits have an $h^2 < 0.5$ \citep{Carlson:2008hl}, and for all additional analyses we have conservatively set $h^2=0.2$.
-Together, equations \ref{eq:N} and \ref{eq:mu} for two linked habitats $i$ and $j$ define the 4-dimensional system of difference equations that describe the eco-evolutionary dynamics of the metapopulation.
+Together, equations \ref{eq:N} and \ref{eq:mu} for two linked populations $i$ and $j$ define the 4-dimensional system of difference equations that describe the eco-evolutionary dynamics of the metapopulation.
 \\
@@
 %Density-dependent m
-\noindent There is mounting evidence that straying is density-dependent, perhaps reflecting a signature of collective navigation \citep{Berdahl:2014bl,Berdahl:2017uu}.
+\noindent There is mounting evidence that straying is density-dependent, consistent with predictions of the collective navigation hypothesis \citep{Berdahl:2014bl,Berdahl:2017uu}.
 Specifically, straying has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tend to decrease the rate at which individuals err, reducing the overall proportion of a population that strays.
@@
 If $C$ is small, relatively small groups of organisms `correct' for even high individual error rates, suppressing straying between sites.
-In the context of our model, small values of $C$ indicate that the effects of collective behaviour on modifying straying are strong.
+In the context of our model, small values of $C$ indicate that the effects of collective behaviour on modifying straying -- thus leading to collective navigation -- are strong.
 % As $C \rightarrow \infty$, groups of any size have no impact on straying, such that $m(t) \rightarrow m_0$, and the model reverts to that of constant straying.
 % When $C \ll \infty$ and the population density is high, $m(t) \rightarrow 0$, whereas if the population is small, individuals operate without regard to the collective, such that $m(t) \rightarrow m_0$, such that in all cases, $m(t) < m_0$.
-Henceforth, we refer to $C$ as determining the strength of collective behavior: as $C\rightarrow\infty$, the effect of collective behavior becomes weaker, such that the size of the population has no impact on straying, and $m(t)\rightarrow m_0$.
-In contrast, if $C \ll \infty$, the effect of collective behavior is strong, and smaller populations reduce $m(t)$.
-Although the strength of collective behavior depends both on $C$ as well as the probability that an individual strays $m_0$, for a specific value of $m_0$, a lower $C$ reduces straying for smaller population sizes, increasing the strength of collective behavior.
-Accordingly, although the strength of collective behavior is not exclusively determined by $C$, we believe that it is an effective moniker in this case, and use it for this purpose while acknowledging that it is not a perfect descriptor.
+Henceforth, we refer to $C$ as determining the strength of collective behaviour: as $C\rightarrow\infty$, the effect of collective behaviour becomes weaker, such that the size of the population has no impact on straying, and $m(t)\rightarrow m_0$.
+In contrast, if $C \ll \infty$, the effect of collective behaviour is strong, and smaller populations reduce $m(t)$.
+Although the strength of collective behaviour depends both on $C$ as well as the probability that an individual strays $m_0$, for a specific value of $m_0$, a lower $C$ reduces straying for smaller population sizes, increasing the strength of collective behaviour.
+We acknowledge that the strength of collective behaviour is not exclusively determined by $C$, but suggest that it is an effective proxy for the role of collective navigation in this model.
 % For realistic population densities, $m(t) < m_0$.\\
@@
 \noindent {\bf (c) Measuring metapopulation robustness}
-\noindent We evaluated metapopulation robustness by measuring the average-CV portfolio effect (PE) \citep{Anderson:2014cx,Schindler:2015gf} as well as the recovery time, which is the time required for the system to return to a steady-state following an induced disturbance to one or both populations \citep{Ovaskainen:2002il}.
+\noindent We evaluated two complementary measures of metapopulation robustness by quantifying \emph{i}) the average-CV portfolio effect (PE) \citep{Anderson:2014cx,Schindler:2015gf} and \emph{ii}) the recovery time, which is the time required for the system to return to a steady-state following an induced disturbance to one or both populations \citep{Ovaskainen:2002il}.
 Throughout, we refer to an increase in portfolio effects and/or reduction in recovery time as promoting metapopulation robustness.
@@
 A more direct way to measure system robustness is to measure the time required for the system (measured as the aggregate steady-state biomass $N_T^*$) to recover following an induced disturbance: systems that recover quickly (shorter recovery times) are more robust than those that recover more slowly (longer recovery times).
-Although there is a direct eigenvalue relationship between the rate of return following a small pulse perturbation \citep{GuckHolmes}, because we aimed to 
+Although there is a direct relationship between the rate of return following a small pulse perturbation and the magnitude of the leading eigenvalue of the Jacobian matrix \citep{GuckHolmes}, because we aimed to
 1) assess the effects of a large perturbation far from the stable state, and 
@@
 
-% 
-% \noindent At low values of density-independent straying the system approaches a fixed point at which both populations persist at equal population size. 
-% As we increase the straying rate a fold bifurcation occurs in which another fixed point is created (figure \ref{fig:traj}, inset). 
-% In this alternative state the population sizes are asymmetric. 
-% The population with the larger population size, i.e. the dominant population, is well-adapted and reproduces quickly. 
-% A small fraction of this population that strays to the subordinate site constitutes a considerable inflow, such that the population in the subordinate site is not as well adapted, which in turn reduces reproduction and stabilizes the asymmetry.
-% 
-% In a regime found at low straying rates (regime I) both the symmetric and the asymmetric state are stable fixed points of the system (figure \ref{fig:traj}, dashed lines). 
-% Which of these fixed points is approached depends on the initial conditions. 
-% The dividing line that separates initial conditions that approach the symmetric solution and initial states that approach the asymmetric solution is a separatrix that runs through an unstable fixed point (figure \ref{fig:traj}, inset) that was also created in the fold bifurcation.
-% The set of all initial states from which the system approaches the symmetric state is said to be that state's basin of attraction. 
-% As we increase the straying rate this basin of attraction shrinks and and eventually vanishes. At this point the unstable asymmetric state collides with the stable symmetric state. 
-% A bifurcation occurs in which the the unstable asymmetric state vanishes and the symmetric state is destabilized. 
-% This is a subcritical pitchfork bifurcation that only occurs in systems with symmetries. 
-% After this bifurcation the stable asymmetric state is the only attractor. 
-% 
-% We find a wide regime (regime II) where the system will always approach an asymmetric state where one population is suppressed. 
-% However, if the straying is increased further the imbalance in population sizes becomes harder to maintain. 
-% As the straying rate is increased we reach a critical point where the stable asymmetric fixed point becomes symmetric and collides with the unstable symmetric fixed point. 
-% The system undergoes a bifurcation that is the supercritical form of the subcritical bifurcation encountered earlier. 
-% In this bifurcation the stable asymmetric fixed point vanishes while the unstable symmetric state is stabilized. 
-% After this bifurcation the symmetric state is the only attractor in the system.
-% Importantly, we find that increasing the asymmetry in the vital rates of populations between sites does not significantly alter the presence or position of these different regimes (figure \ref{fig:symmetry}).
-
-\noindent At low values of density-independent straying the system approaches a fixed point at which both populations persist at equal population size, but as we increase straying, other fixed points are created in which the population sizes are asymmetric (figure \ref{fig:traj}, inset).
+\noindent At low values of density-dependent straying or collective navigation, the system approaches a fixed point at which both populations persist at equal population size, but as we increase straying, other fixed points are created in which the population sizes are asymmetric (figure \ref{fig:traj}, inset).
 The system's underlying symmetry implies that for every asymmetric fixed point there must be another `mirror-image' fixed point in which we find the same population sizes, but where the identities of the populations are reversed.
@@
 
-In the asymmetric states the dominant population is well-adapted and reproduces quickly. 
-The (small) fraction of this population that strays to the subordinate site constitutes a considerable inflow, such that the population in the subordinate site is not as well adapted, which in turn reduces reproduction and stabilizes the asymmetry.
+In the asymmetric states the dominant population is well-adapted and has a high rate of recruitment. 
+The (small) fraction of this population that strays to the subordinate site constitutes a considerable inflow of individuals, such that the population in the subordinate site is not as well adapted.
+This in turn reduces reproduction and stabilizes the asymmetry. 
 In a regime found at low straying rates (regime I) both the symmetric and the asymmetric states are stable fixed points of the system (figure \ref{fig:traj}, purple points). 
@@
 %Portfolio effects
-If we assume that straying is density-dependent, the probability that an individual errs $m_0$ in part determines the magnitude of straying within the population (Eq.
+% If we assume that straying is density-dependent, the probability that an individual errs $m_0$ in part determines the magnitude of straying within the population (Eq.
 \ref{eq:ddm}), such that $m(t)$ becomes lower as $N(t)$ increases, which we assume here to be the consequence of collective decision-making behaviours \citep{Berdahl:2016dx}.
-% Organisms that employ collection decision-making to aid in navigation have reduced straying rates.
-The parameter $C$ in Eq.
+% The parameter $C$ in Eq.
 \ref{eq:ddm} determines to what extent straying is reduced as group size increases, where low values correspond to the effects of strong collective behaviour, and high values correspond to the effects of weak collective behaviour, such that straying is effectively constant \citep{Berdahl:2016dx}.
-To enable comparisons between models with constant and density-dependent straying, we compare constant $m$ to both the individual straying probability $m_0$ as well as the value of straying observed at the stable state $m^*$.
-In the alternative stable state regimes, the dispersing populations are characterized by alternative $m^*$ values if the populations have asymmetric densities, corresponding to dominant and subordinate states.
-% , such that small increases in group size have a large effect on straying, and where $C \rightarrow \infty$ corresponds to constant straying independent of group size .
-%Because our analyses concern steady-state conditions, the inclusion of density-dependent straying does not have a large impact on the qualitative nature of our findings (figure \ref{fig:ddm}).
-% Density dependence alters the straying ratio at steady-state population densities because $0 < m^* < m_0$, and this serves to rescale both the strength of the PE as well as the recovery time, but does not change the qualitative nature of our findings.
-
-When collective behaviour is very strong, such that $C$ is very low, small increases in population density beget large reductions in straying.
+% To enable comparisons between models with constant and density-dependent straying, we compare constant $m$ to both the individual straying probability $m_0$ as well as the value of straying observed at the stable state $m^*$.
+% In the alternative stable state regimes, the dispersing populations are characterized by alternative $m^*$ values if the populations have asymmetric densities, corresponding to dominant and subordinate states.
+When collective behaviour is very strong, (small values of $C$), small increases in population density beget large reductions in straying.
 These reductions can be large enough that the system avoids the alternative stable state regime altogether (figure \ref{fig:cb}, left inset).
@@
 With intermediate $C$ there is a general increase in robustness if straying is low-intermediate (higher portfolio effect, shorter recovery time), followed by an erosion in robustness as straying becomes high.
-In this parameter space, collective migration can mean that the low-density population will stray more, losing well-adapted local individuals, while still receiving some maladapted strays from the larger population, thereby increasing the likelihood of stochastic extinction.
+In this parameter space, collective navigation results in the low-density population straying more, losing well-adapted local individuals, while still receiving some maladapted strays from the larger population, thereby increasing the likelihood of stochastic extinction.
 
@@
 As collective behaviour is strengthened (smaller $C$), Regime II is avoided at lower valus of $m_0$ and expands for higher values of $m_0$.
-When the effects of collective behavior are very strong, Regime II then collapses (black region in figure \ref{fig:bifurcationsddm}b) and gives way to regime I, which plays a larger role over a larger range of $m_0$ when $C$ is low (gray region in figure \ref{fig:bifurcationsddm}b).
+When the effects of collective behaviour are very strong, Regime II then collapses (black region in figure \ref{fig:bifurcationsddm}b) and gives way to regime I, which plays a larger role over a larger range of $m_0$ when $C$ is low (gray region in figure \ref{fig:bifurcationsddm}b).
 Importantly, when the strength of collective behaviour is intermediate, both regimes I and II are relevant at low-intermediate $m_0$.
@@
 \noindent{\bf (c) The role of habitat heterogeneity and changing selective landscapes}\\
-As habitat heterogeneity ($\Delta\theta$) increases, even small amounts of straying can lead to the appearance of alternative stable states, and this is in direct accord with the findings of Ronce and Kirkpatrick \citep{Ronce:2001dp}.
+As habitat heterogeneity ($\Delta\theta$) increases, even small amounts of straying can lead to the appearance of alternative stable states.
+%, and this is in direct accord with the findings of Ronce and Kirkpatrick \citep{Ronce:2001dp}.
 However, if straying is density-dependent, the strength of collective behaviour has a large influence on the occurrence of both alternative stable state regimes I and II.
-When heterogeneity is low and the effects of collective behaviour are weak such that straying is constant (high $C$), regime II occurs for small-intermediate $m_0$, and regime I does not play a role (figure \ref{fig:bifurcationsddm}a), mirroring the findings of Ronce and Kirkpatrick \citep{Ronce:2001dp}.
+When heterogeneity is low and the effects of collective behaviour are weak such that straying is constant (high $C$), regime II occurs for small-intermediate $m_0$, and regime I does not play a role (figure \ref{fig:bifurcationsddm}a).%, mirroring the findings of Ronce and Kirkpatrick \citep{Ronce:2001dp}.
 The absence of regime I implies that there are no hidden stable state configurations that might trap a disturbed population in an asymmetric low-density state.
@@
 %Main takehome
-Generally, we find that when the effects of collective behaviour are strong, metapopulation robustness is enhanced.
+Generally, we find that when the effects of collective behaviour are strong and collective navigation is likely, metapopulation robustness is enhanced.
 However, empirical observations of natural populations suggest that the effects of collective behaviour are intermediate (e.g. $10^{2} \lesssim C \lesssim 10^3$) \citep{Berdahl:2014bl,Berdahl:2016dx}.
@@
 A central dynamic of the model is that straying can lead to the emergence of asymmetric alternative stable states, or \emph{migrational meltdown} \citep{Ronce:2001dp}, pushing one of the populations to a dominant, well-adapted, high density, state, and one to a subordinate, maladapted, low density, state.
-Although there are subtle differences in our model and the general framework presented in Ref.
-\citep{Ronce:2001dp}, the general dynamic features are the same if we assume that dispersal is symmetric between sites and density-independent (which occurs when $C\rightarrow\infty$).
+Although there are subtle differences in our model and the general framework presented by Ronce and Kirkpatrick \citep{Ronce:2001dp}, the general dynamic features are the same if we assume that dispersal is symmetric between sites and density-independent (which occurs when $C\rightarrow\infty$).
 The dynamic regimes that emerge from the eco-evolutionary model -- in particular the occurrence of alternative stable state regimes I and II (figure \ref{fig:traj}) -- have large effects on both the portfolio effect and the recovery time following an induced disturbance (figure \ref{fig:pert}).
@@
 This means that - despite potentially high individual error rates - group formation minimizes straying. 
-This occurs when groups of $\leq 10^2$ individuals significantly minimizes straying, which is likely unrealistic.
+This occurs when groups of $\leq 100$ individuals significantly minimizes straying, which is likely unrealistic.
 Moreover, when the effects of collective behaviour are strong, regime II gives way to the dominance of regime I, which harbors low-density basins of attraction (dashed lines in figure \ref{fig:traj}) that can effectively trap disturbed populations in a subordinate stable state, not unlike the Allee effects observed in the collective migration model explored by Berdahl et al.
@@
 This may be particularly consequential for populations that are spatially adjacent but separated by sharp environmental boundaries, such that trait optima are divergent yet dispersal is relatively high.
-Such a scenario plays out repeatedly in the context of wild and hatchery-produced salmon. 
+Such a scenario plays out repeatedly in the context of interactions between wild and hatchery-produced salmon. 
 Although wild and hatchery populations may occur close on the landscape, and indeed often are sympatric within the same river network, the selective environments to which they are locally adapted differ dramatically \citep{Christie:2012bj}.
-Straying of domesticated hatchery-produced fish from release sites and spawning in the wild drastically reduces the productivity of wild populations through competition and outbreeding depression \citep{Chilcote:2003bb,Araki:2007cm}.
+Straying of domesticated hatchery-produced fish from release sites and spawning in the wild reduces the productivity of wild populations through competition and outbreeding depression \citep{Chilcote:2003bb,Araki:2007cm}.
 
@@
 This pattern is opposite that observed for scenarios where habitat heterogeneity and straying are assumed to be independent, and suggests that increases in straying that are associated with growing similarities between habitats can push a metapopulation into a regime where hidden low-density basins of attraction exist.
-Thus, management activities that alter dispersal rates by outplanting individuals or reconnecting disconnected habitats could have unintended eco-evolutionary consequences \citep{Anderson:2013bf,Pess:2014isa}, introducing particular management challenges.
+Thus, management activities that alter dispersal rates by outplanting individuals or reconnecting disconnected habitats could have complex eco-evolutionary consequences \citep{Anderson:2013bf,Pess:2014isa}, and compromise management or conservation objectives.
 
@@
 
-A general message from our theoretical framework is that the emergence of alternative eco-evolutionary states depends jointly on the strength of collective behaviour and individual straying, and that this has large implications for metapopulation robustness.
+A general message from our theoretical framework is that the emergence of alternative eco-evolutionary states depends jointly on the strength of collective behaviour and level of individual straying, and that this has large implications for metapopulation robustness.
 Although robustness is in many cases aided by increasing the strength of collective behaviour, the greater role of both alternative stable state regimes I and II portends additional complexity in eco-evolutionary dynamics, and this could serve to hinder effective management.
 Moreover, this increased complexity at empirically observed levels of straying \citep{H:2013fs} and at realistic (intermediate) ranges of the strength of collective behaviour, is only magnified with increasing habitat heterogeneity and when heterogeneity itself is linked to individual straying.
-An additional component that we have not explored here, but that may be particularly relevant to consider, are the effects of including additional sites, as well as the patterns of dispersal that connect these sites.
+An additional component that we have not explored here, but that may be particularly relevant to consider, are the effects of including additional sites within the metapopulation network, as well as the patterns of dispersal that connect these sites.
 The structure of dispersal has been shown to have a large influence on population dynamics \citep{Heino:1997vf,Carrara:2014cy,Yeakel:2013vz,Gilarranz:2017fs}, and to what extent density-dependent straying influences the eco-evolutionary dynamics of populations in large spatially-structure networks is of considerable interest.
-We are hopeful that these predictions may inspire future theoretical and empirical studies that aim to expand upon the relationships that we have explored.
-
-While our theoretical work is a vast simplification of natural systems, our results have important potential implications for understanding the eco-evolutionary dynamics of spatially structured populations such as in salmon.
-A particularly salient finding was that density-dependent straying may serve to promote or inhibit population robustness, depending on the strength of the collective behaviour, and the underlying magnitude of straying.
-When the strength of collective behaviour is at plausible intermediate levels, straying increased population robustness through enhanced portfolio effects and reduced time to recovery following disturbance.
+We are hopeful that these predictions will inspire future theoretical and empirical studies that aim to expand upon the relationships that we have explored.
+
+% While our theoretical work is a vast simplification of natural systems, our results have important potential implications for understanding the eco-evolutionary dynamics of spatially structured populations such as in salmon.
+A particularly salient finding of our work was that density-dependent straying may serve to promote or inhibit population robustness, depending on the strength of the collective behaviour, and the underlying magnitude of straying.
+When the strength of collective behaviour is at plausible intermediate levels, a moderate level of straying increased population robustness by reducing the time to recovery following metapopulation collapse (figure \ref{fig:pert}b).
 Salmon have evolved within the context of dynamic geomorphic landscapes where habitat quantity and quality shifts as a mosaic through time \citep{Stanford:2017bu}. 
 Our results provide evidence supporting the hypothesis described in Berdahl et al.
-\citep{Berdahl:2014bl} that collective behaviour may underpin rapid habitat colonization following natural disturbance such as volcanic eruptions \citep{Leider:1989gx}, reconnected habitats following restoration \citep{Pess:2012by,Pess:2014isa}, or in the context of glacial retreat and climate warming \citep{Milner:2008gb}.
+\citep{Berdahl:2014bl} that collective behaviour may support rapid habitat colonization following natural disturbance such as volcanic eruptions \citep{Leider:1989gx}, reconnected habitats following restoration \citep{Pess:2012by,Pess:2014isa}, or in the context of glacial retreat and climate warming \citep{Milner:2008gb}.
 Moreover, our results are consistent with the role of collective behaviour in facilitating reproductive isolation and local adaptation to site-specific selection in populations that recover following disturbance to the extent that straying decreases as population sizes increase.
@@
 High rates of straying are predicted to decrease metapopulation robustness, and there are a series of common practices in salmon management that may be elevating straying rates \citep{Ford:2002ip,Brenner:2012gl}.
-For example, transporting young salmon downstream via barge to avoid infrastructure (e.g., dams) or dangerous habitats may disrupt the processes involved with critical periods of imprinting prior to or during downstream migration by sea-going individuals and increase straying by adults later in life.
-Our results support the conservation concern that large-scale releases of salmon produced in hatcheries that stray could have profound impacts on the robustness of salmon metapopulations through the erosion of portfolio effects.
+For example, transporting young salmon downstream to increase survival during outmigration may disrupt the processes involved with critical periods of imprinting prior to or during downstream migration by sea-going individuals and increase straying by adults later in life.
+Our results support the conservation concern that large-scale releases of salmon produced in hatcheries that stray could decrease robustness of salmon metapopulations through the erosion of portfolio effects and increase in recovery times.
 Moreover, hatchery environments are associated with marked changes in fish social behaviour that may increase collective dynamics of migrating groups \citep{Ruzzante:1994bu}, consistent with the findings of Jonsson and Jonsson \citep{Jonsson:2011kg} who report stronger associations between straying and abundance in escaped aquaculture produced Atlantic salmon than their wild counterparts.
@@
 \citep{HardestyMoore:wg}.
-We suggest that increased understanding the proximate and ultimate factors governing dispersal among local populations within metapopulations, across heterogeneous environments, in tandem with the mosaic of selective forces acting on those environments, may be key to promoting persistence in the wild.
+We suggest that increased understanding of the proximate and ultimate factors governing dispersal among local populations within metapopulations, across heterogeneous environments, in tandem with the mosaic of selective forces acting on those environments, may be key to promoting persistence in the wild \cite{Westley:uh}.
 \\ \\
@@
 P.A.H.W.
-was supported by the UA Foundation at the University of Alaska Fairbanks.
+was supported by the UA Foundation at the University of Alaska Fairbanks and the Gordon and Betty Moore Foundation.
 J.W.M.
@@
 
+\bibitem{Gibert:2016cu}
+Gibert JP, 2016 {The effect of phenotypic variation on metapopulation
+  persistence}.
+\newblock \emph{Population Ecology} \textbf{58}, 345--355
+
 \bibitem{Brown:1977gk}
@@
 
-% \bibitem{Berdahl:2015kv}
-% Berdahl A, Torney CJ, Schertzer E, Levin SA, 2015 {On the evolutionary
-%   interplay between dispersal and local adaptation in heterogeneous
-%   environments}.
-% \newblock \emph{Evolution} \textbf{69}, 1390--1405
-
 \bibitem{Berdahl:2016dx}
@@
 Berdahl A, Kao A, Flack J, Westley PAH, Codling E, Couzin ID, Dell AI, Biro D,
-  2017 {Collective animal navigation and migratory culture: from theoretical
+  2018 {Collective animal navigation and migratory culture: from theoretical
   models to empirical evidence.}
@@
 \newblock \emph{Aquaculture} \textbf{120}, 1--24
+
+\bibitem{Westley:uh}
+Westley PAH, Berdahl A, Torney CJ, Biro D, 2018 {Emerging study of collective
+  movement ecology}.
+\newblock \emph{Philos. T. Roy. Soc. B} \textbf{This volume}
 
@@
 $m_0$ & Individual straying probability & $(0,0.5)$\\
-$C$ & Strength of collective behavior (low=strong) & $(10^1,10^5)$\\
+$C$ & Strength of collective behaviour (low=strong) & $(10^1,10^5)$\\
 $r_{\rm max}$ & Maximum recruitment rate & 2.0 \\

### 2018-01-26 Manuscript/FinalDraft3/msv10.tex
final revision

@@
 Larger portfolio effects are expected to increase the robustness of metapopulations to external disturbances, and by extension promote persistence \citep{Thibaut:2012km}.
-In contrast, homogenization of populations leading to greater synchronization and weakened PE may be a harbinger of metapopulation collapse and extinction \citep{Carlson:2011ce}.
+In contrast, homogenization of populations leading to greater synchronisation and weakened PE may be a harbinger of metapopulation collapse and extinction \citep{Carlson:2011ce}.
 
@@
 Dispersal also enables demographic rescue, when depressed or extirpated populations are recolonized by immigrants from the rest of the metapopulation.
-On the other hand, high rates of dispersal may synchronize the dynamics of populations and subsequently increase the risk of extinction of the entire metapopulation \citep{Earn:2000fm,Satterthwaite:2015ge}.
+On the other hand, high rates of dispersal may synchronise the dynamics of populations and subsequently increase the risk of extinction of the entire metapopulation \citep{Earn:2000fm,Satterthwaite:2015ge}.
 % Dispersal will also influence the evolutionary dynamics of the metapopulation.
 % though how the interplay between selection and population dynamics might impact persistence is less well understood. 
-Additionally, dispersal may also introduce maladapted individuals into habitats that are host to different environmental conditions, possibly lowering the mean fitness of the recipient population \citep{Ronce:2001dp,Muhlfeld:2014hs}.
+Dispersal may also introduce maladapted individuals into habitats that are host to different environmental conditions, possibly lowering the mean fitness of the recipient population \citep{Ronce:2001dp,Muhlfeld:2014hs}.
 More broadly, dispersal can provide a mechanism by which phenotypes are sorted in space rather than time and facilitates the spread of potentially maladaptive genes \citep{Lowe:2015ft}.
@@
 Recently, the role of social interactions to lead to collective navigation has been hypothesized  as a mechanism shaping the success of philopatric migrations \citep{Berdahl:2016dx,HardestyMoore:wg,Berdahl:2017uu}.
-The collective navigation hypothesis posits that rate at which individuals disperse may be linked to errors made at an individual-level that are themselves diminished by migrating in groups and pooling individual choices \citep{Simons:2004jo,Berdahl:2016dx,Berdahl:2017uu}.
+The collective navigation hypothesis posits that the rate at which individuals disperse may be linked to individual-level error, which is diminished by migrating in groups and pooling individual choices \citep{Simons:2004jo,Berdahl:2016dx,Berdahl:2017uu}.
 Thus, dispersal rates can be higher at lower population abundances \citep{Berdahl:2014bl}, which can in turn profoundly influence the eco-evolutionary dynamics of metapopulations. 
@@
 The eco-evolutionary impacts of dispersal likely have implications for conservation and management in key taxa such as in migratory salmon \citep{Brenner:2012gl,Johnson:2012fe,Fullerton:2011ii}.
-While anadromous salmonid fishes (genera \emph{Oncorhynchus} and \emph{Salmo}) are renown for returning to their natal spawning habitats with high accuracy and precision after years at sea \citep{Quinn:2011tf,Jonsson:2011kg,Keefer:2014gg}, while some individuals disperse (termed `straying' and used synonymously with dispersal hereafter) to non-natal sites to spawn \citep{Quinn:1993ge,Hendry:2004wf}.
+While anadromous salmonid fishes (genera \emph{Oncorhynchus} and \emph{Salmo}) are renown for returning to their natal spawning habitats with high accuracy and precision after years at sea \citep{Quinn:2011tf,Jonsson:2011kg,Keefer:2014gg}, some individuals disperse (termed `straying' and used synonymously with dispersal hereafter) to non-natal sites to spawn \citep{Quinn:1993ge,Hendry:2004wf}.
 Straying provides a mechanism for the colonization of new or connected habitats following glacial retreat or large-scale geomorphic landscape change \citep{Hendry:2004wf}.
@@
 Straying among salmon may be influenced by environmental factors such as water temperature, human activities such as hatchery practices, and population density as predicted by the collective navigation hypothesis \citep{Peterson:2014gy,Berdahl:2017uu}.
-Straying can introduce new maladaptive genotypes into the recipient population, while the ensuing genetic homogenization could synchronize population dynamics and erode portfolio effects \citep{Moore:2010gs,Carlson:2011ce,Braun:2016ib}.
+Straying can introduce new maladaptive genotypes into the recipient population, while the ensuing genetic homogenization could synchronise population dynamics and erode portfolio effects \citep{Moore:2010gs,Carlson:2011ce,Braun:2016ib}.
 Given that locally distinct populations are often linked by straying, there is an opportunity and need to understand the fundamental and applied consequences of straying on metapopulation persistence, conservation, and management.
@@
 \noindent{\bf (a) Metapopulation framework}\\
-\noindent We follow the basic framework described by Ronce and Kirkpatrick \citep{Ronce:2001dp}, where dispersal connects two populations $i$ and $j$ that inhabit two distinct habitats, each with densities $N_i$ and $N_j$ and trait values $x_i$ and $x_j$.
+\noindent We follow the basic framework described by Ronce and Kirkpatrick \citep{Ronce:2001dp}, where dispersal connects two populations $i$ and $j$ that inhabit two distinct habitats, each with abundances $N_i$ and $N_j$ and trait values $x_i$ and $x_j$.
 In our version of the model, $i$ and $j$ are locally adapted to site-specific conditions such that there is an optimum trait value $\theta_i$ and $\theta_j$ associated with each habitat, where recruitment is maximized if the trait value of the local population equals the optimum ($x = \theta$).
 Moreover, we assumed that $x_{i,j}$ are normally distributed with means $\mu_{i,j}$ and have the same standard deviation $\sigma$.
-As such, the recruitment rate $R_i(\mu_i(t),\theta_i)$ for both populations is determined by the mean trait value of the local population relative to the optimal value at that site.
+As such, the recruitment rate $R(\mu(t),\theta)$ for both populations is determined by the mean trait value of the local population relative to the optimal value at that site.
 Mean trait values for both populations are dynamic variables and change over time in response to differences in recruitment as individuals mix between sites.
 Trait means for each population are thus subject to selection, the strength of which is proportional to the difference between the trait mean and the local trait optimum at a given point in time \citep{simpson1953major,Lande:1976ga,Ronce:2001dp}.
-This is broadly consistent with empirical patterns observed in pacific salmon dynamics \citep{Reisenbichler:1988ex}.
+This is broadly consistent with empirical patterns observed in Pacific salmon dynamics \citep{Reisenbichler:1988ex}.
 The two populations occur in spatially separate sites that are close enough that a proportion of the population $m$ strays into the wrong site.
@@
 % In this sense, both populations serve simultaneously as donor and recipient populations.
-Because total recruitment will be determined by both locals (with a mean trait value closer to the site optimum) and strays (with mean trait values farther from the local optimum), the recruitment of the aggregate is determined by the mean of the trait mix $R_i(\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\theta_i)$, where
+Because total recruitment will be determined by both locals (with a mean trait value closer to the site optimum) and strays (with mean trait values farther from the local optimum), the recruitment of the aggregate for population $i$ is determined by the mean of the trait mix $R_i(\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\theta_i)$, where
 \begin{equation}
@@
 \noindent where $g(x_i)$ is the Gaussian probability density function for the trait $x_i$.
-The mismatch between the mean of the local trait mix $\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t)$ and the local optimum $\theta_i$ scales the recruitment rate for the population, and $\tilde{P}_i\sim {\rm Normal}(0,0.01)$ introduces a small amount of demographic variability.
+The mismatch between the mean of the local trait mix $\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t)$ and the local optimum $\theta_i$ scales the recruitment rate for the population, and $\tilde{P}_i\sim {\rm Normal}(0,0.01)$ introduces a small amount of demographic stochasticity.
 The parameter $\tau$ is the strength of selection and controls the sensitivity of recruitment to changes in the mean trait value away from the optimum.
@@
 Specifically, straying has been linked directly to a collective decision-making phenomenon, where greater numbers of individuals tend to decrease the rate at which individuals err, reducing the overall proportion of a population that strays.
-Following Berdahl et al. \citep{Berdahl:2016dx}, given the probability that an individual strays is $m_0$, the proportion of the local population $N_i(t)$ that strays is
+Following Berdahl et al. \citep{Berdahl:2016dx}, given the probability that an individual strays is $m_0$, the proportion of the local population $i$ that strays is
 
@@
 \noindent where $C$ is a half-saturation constant, determining to what extent collective behaviour, as a function of group size, diminishes straying.
-If $C$ is small, relatively small groups of organisms `correct' for even high individual error rates, suppressing straying between sites.
-In the context of our model, small values of $C$ indicate that the effects of collective behaviour on modifying straying -- thus leading to collective navigation -- are strong.
+For a given $m_0$, if $C$ is small, relatively smaller groups of organisms `correct' for higher individual error rates, suppressing straying between sites.
+Small values of $C$ indicate that the effects of collective behaviour on modifying straying -- thus leading to collective navigation -- are strong.
 % As $C \rightarrow \infty$, groups of any size have no impact on straying, such that $m(t) \rightarrow m_0$, and the model reverts to that of constant straying.
@@
 Henceforth, we refer to $C$ as determining the strength of collective behaviour: as $C\rightarrow\infty$, the effect of collective behaviour becomes weaker, such that the size of the population has no impact on straying, and $m(t)\rightarrow m_0$.
-In contrast, if $C \ll \infty$, the effect of collective behaviour is strong, and smaller populations reduce $m(t)$.
-Although the strength of collective behaviour depends both on $C$ as well as the probability that an individual strays $m_0$, for a specific value of $m_0$, a lower $C$ reduces straying for smaller population sizes, increasing the strength of collective behaviour.
-We acknowledge that the strength of collective behaviour is not exclusively determined by $C$, but suggest that it is an effective proxy for the role of collective navigation in this model.
+% In contrast, if $C \ll \infty$, the effect of collective behaviour is strong, and smaller populations reduce $m(t)$.
+Thus, although the strength of collective behaviour depends both on $C$ as well as $m_0$, for a given $m_0$, $C$ is an effective proxy for the strength of collective behaviour.
+% a lower $C$ reduces straying for smaller population sizes, increasing the strength of collective behaviour.
+% We acknowledge that the strength of collective behaviour is not exclusively determined by $C$, but suggest that it
 % For realistic population densities, $m(t) < m_0$.\\
@@
 
-The average-CV portfolio effect is, as the name implies, the average CV across each population $N_i$ divided by the CV of the aggregate $N_T=\sum_i N_i$ \citep{Anderson:2013gb}, such that
+The average-CV portfolio effect is, as the name implies, the average CV of the population biomass $N_i$ divided by the CV of the aggregate biomass $N_T=\sum_i N_i$ \citep{Anderson:2013gb}, such that
 
@@
 
-\noindent where in this case the number of populations is limited to $X=2$ and the expectations $\rm E(\cdot)$ and variances $\rm VAR(\cdot)$ are evaluated at the steady-state.
-The steady-state condition is denoted by `$*$'.
+\noindent where in this case the number of populations is limited to $X=2$ and the expectations $\rm E(\cdot)$ and variances $\rm VAR(\cdot)$ are evaluated at the steady-state, denoted by `$*$'.
 As the CV of $N_T^*$ decreases relative to that of the constituent populations, $\langle{\rm PE}\rangle > 1$, and the metapopulation is presumed to be more stable because the aggregate has functioned to dampen population-level variance.
-Moreover, portfolio effects greater than unity correspond to less synchronization  \citep{Loreau:2008ju,Anderson:2014cx,Yeakel:2013vz} and thus a greater potential for demographic rescue among populations, buffering the system as a whole against extinction.
+Moreover, portfolio effects greater than unity correspond to less synchronisation  \citep{Loreau:2008ju,Anderson:2014cx,Yeakel:2013vz} and thus a greater potential for demographic rescue among populations, buffering the system as a whole against extinction.
 
@@
 Although there is a direct relationship between the rate of return following a small pulse perturbation and the magnitude of the leading eigenvalue of the Jacobian matrix \citep{GuckHolmes}, because we aimed to
-1) assess the effects of a large perturbation far from the stable state, and 
-2) estimate the time required for all transient effects to decay following this perturbation (including dampened oscillations), we used a simulation-based numerical procedure.
+1) assess the effects of a large perturbation far from the steady state, and 
+2) estimate the time required for all transient effects to decay following this perturbation, we used a simulation-based numerical procedure.
 Recovery time was calculated by initiating a disturbance at $t=t_d$, and monitoring $N_T(t_d+t)$ as $t\rightarrow T$, where $T$ is large. 
@@
 
-\noindent At low values of density-dependent straying or collective navigation, the system approaches a fixed point at which both populations persist at equal population size, but as we increase straying, other fixed points are created in which the population sizes are asymmetric (figure \ref{fig:traj}, inset).
+\noindent At low values of density-independent straying the system approaches a fixed point at which both populations persist at equal population size, but as we increase straying, other fixed points are created in which the population sizes are asymmetric (figure \ref{fig:traj}, inset).
 The system's underlying symmetry implies that for every asymmetric fixed point there must be another `mirror-image' fixed point in which we find the same population sizes, but where the identities of the populations are reversed.
@@
 Based on the Jacobian eigenvalues we conjecture that these bifurcations are fold bifurcations. 
-In a generic fold bifurcation of maps, two new fixed points are created, one of which is unstable, while the other is stable. 
+In a generic fold bifurcation of maps, two new fixed points are created, one of which is unstable, while the other is stable \citep{GuckHolmes}. 
 In the present case, two of these bifurcations occur at the same time, one which creates fixed points where the first population is dominant in the stable fixed point, whereas the second bifurcation creates the mirror-image fixed points where the second population is dominant.
@@
 This in turn reduces reproduction and stabilizes the asymmetry. 
-In a regime found at low straying rates (regime I) both the symmetric and the asymmetric states are stable fixed points of the system (figure \ref{fig:traj}, purple points). 
+In a regime found where straying is low (regime I) both the symmetric and the asymmetric states are stable fixed points of the system (figure \ref{fig:traj}). 
 Which of these fixed points is approached depends on the initial conditions.
@@
 As we increase straying, the asymmetric states eventually collide with the stable symmetric state. 
-A subcritical pitchfork bifurcation occurs in which the unstable asymmetric states vanishes and the symmetric state is destabilized. 
+A subcritical pitchfork bifurcation occurs in which the unstable asymmetric states vanish and the symmetric state is destabilized. 
 After this bifurcation the stable asymmetric states are the only attractors. 
 We find a wide regime (regime II) where the system will always approach an asymmetric state where one population is suppressed. 
-However, if the straying is increased further the imbalance in population sizes becomes harder to maintain. 
+However, if straying is increased further the imbalance in population sizes becomes harder to maintain. 
 Eventually we reach a critical point where the stable asymmetric fixed points become symmetric and collide with the unstable symmetric fixed point. 
@@
 Similar patterns are observed with respect to the recovery time as straying is increased (figure \ref{fig:mpert}b).
-For lower $m$, recovery of single extinctions is impacted by the appearance of low-density basins of attraction in regime I, whereas recovery following near-collapse is not.
+For lower $m$, recovery following individual extinctions is impacted by the appearance of low-density basins of attraction in regime I, whereas recovery following near-collapse is not.
 For intermediate values of $m$ (regime II), the time to recovery is only diminished when the subordinate population becomes extinct, whereas the time to recovery following near-collapse and the extinction of the dominant population are similar and grow until Regime II is exited at high $m$.
-In the case where the subordinate population is extirpated, the most rapid recovery rate occurs when straying is low ($m = 0.08$). 
+In the case where the subordinate population is extirpated, the most rapid recovery occurs when straying is low ($m = 0.08$). 
 In contrast, when the dominant population goes extinct, the most rapid recovery is associated with minimal straying. 
 It should be noted that when there is no straying ($m = 0$), recovery time is infinite and these values are not shown. 
-The relationship between straying and recovery time in the case of synchronous population disturbance is generally positive.
+Increased straying generally leads to longer recovery times when both populations suffer near-collapse.
+% The relationship between straying and recovery time in the case of synchronous population disturbance is generally positive.
 
@@
 First, in the alternative stable state regime, the low density subordinate population has correspondingly higher $m^*$, whereas the high density dominant population has correspondingly lower $m^*$ (figure \ref{fig:cb}, center inset).
-Second, the alternative stable state regime results in a $\Delta N^*$ that is reduced when individual straying is low, and magnified when individual straying is high (figure \ref{fig:cb}, main).
-In other words, when collective behaviour is of intermediate strength -- the more realistic range for species that navigate via collective decision-making -- increased individual straying exaggerates the differences between the stable state densities, effectively pushing the subordinate population closer to extinction.
+Second, the alternative stable state regime results in a $\Delta N^*$ that is reduced or negligible when individual straying is low, and magnified when individual straying is high (figure \ref{fig:cb}, main).
+In other words, when collective behaviour is of intermediate strength -- the more realistic range for species that navigate via collective decision-making -- increased individual straying exaggerates the differences between the steady state densities, effectively pushing the subordinate population closer to extinction.
 
@@
 With intermediate $C$ there is a general increase in robustness if straying is low-intermediate (higher portfolio effect, shorter recovery time), followed by an erosion in robustness as straying becomes high.
-In this parameter space, collective navigation results in the low-density population straying more, losing well-adapted local individuals, while still receiving some maladapted strays from the larger population, thereby increasing the likelihood of stochastic extinction.
-
-Sharp changes in metapopulation robustness are due to changes in alternative stable state regimes I and II as the strength of collective behaviour increases (lower $C$, figure \ref{fig:bifurcationsddm}b).
+In this parameter space, collective navigation results in the low-density population that is straying more, losing well-adapted local individuals, while still receiving some maladapted strays from the larger population, thereby increasing the likelihood of stochastic extinction.
+
+Sharp changes in metapopulation robustness are due to changes in alternative stable state regimes I and II as the strength of collective behaviour increases (lower $C$, figure \ref{fig:bifurcationsddm}).
 When collective behaviour is weak (large $C$), alternative stable states tend to occur at low-intermediate values of individual straying $m_0$.
-As collective behaviour is strengthened (smaller $C$), Regime II is avoided at lower valus of $m_0$ and expands for higher values of $m_0$.
-When the effects of collective behaviour are very strong, Regime II then collapses (black region in figure \ref{fig:bifurcationsddm}b) and gives way to regime I, which plays a larger role over a larger range of $m_0$ when $C$ is low (gray region in figure \ref{fig:bifurcationsddm}b).
-Importantly, when the strength of collective behaviour is intermediate, both regimes I and II are relevant at low-intermediate $m_0$.
+As collective behaviour is strengthened (smaller $C$), Regime II is avoided at lower values of $m_0$ and expands at higher values of $m_0$.
+When the effects of collective behaviour are very strong, Regime II collapses (black region in figure \ref{fig:bifurcationsddm}b) and gives way to regime I, which plays a larger role over a larger range of $m_0$ when $C$ is low (gray region in figure \ref{fig:bifurcationsddm}b).
+Importantly, when the strength of collective behaviour is intermediate, both regimes I and II are relevant at low-intermediate $m_0$.\\
 % In this case, the combined effect of regimes I and II leads to the observed increase in robustness at low-intermediate $m_0$ and subsequent decline with high $m_0$.
@@
 However, if straying is density-dependent, the strength of collective behaviour has a large influence on the occurrence of both alternative stable state regimes I and II.
-When heterogeneity is low and the effects of collective behaviour are weak such that straying is constant (high $C$), regime II occurs for small-intermediate $m_0$, and regime I does not play a role (figure \ref{fig:bifurcationsddm}a).%, mirroring the findings of Ronce and Kirkpatrick \citep{Ronce:2001dp}.
-The absence of regime I implies that there are no hidden stable state configurations that might trap a disturbed population in an asymmetric low-density state.
+When heterogeneity is low and the effects of collective behaviour are weak such that straying is constant (high $C$), regime II occurs for small-intermediate $m_0$, and regime I does not play a role (figure \ref{fig:bifurcationsddm}a).
+%, mirroring the findings of Ronce and Kirkpatrick \citep{Ronce:2001dp}.
+The absence of regime I implies that there are no hidden steady state configurations that might trap a disturbed population in an asymmetric low-density state.
 As the strength of collective behaviour increases, regime I appears at a cusp and becomes increasingly dominant with greater individual straying.
@@
 % For high $\Delta\theta$, there is no longer any sanctum from regimes I or II except for very high individual straying.
-Regime II dominates all but very high individual straying when the effects of collective behaviour are weak (high $C$), and very low individual straying when the effects of collective behaviour are strong (low $C$).
-Moreover, in highly heterogeneous habitats, if the effects of collective behaviour are strong and straying is low (low $m_0$ and $C$ values), regime I, which harbors low-density basins of attraction, cannot be avoided.
+Regime II dominates at all but very high individual straying when the effects of collective behaviour are weak (high $C$), and very low individual straying when the effects of collective behaviour are strong (low $C$).
+Moreover, in highly heterogeneous habitats, if the effects of collective behaviour are strong and straying is low (low $m_0$ and low $C$), regime I, which harbors low-density basins of attraction, cannot be avoided.
 
@@
 We implemented this inverse relationship by setting $m_0 = 1/(2+\epsilon\Delta\theta)$ where $\epsilon$ controls the degree to which an increase in $\Delta\theta$ lowers $m_0$ (figure \ref{fig:mtheta}, inset).
-Accordingly, $m_0$ is greater for lower $\Delta\theta$, such that there is less straying between dissimilar sites and more straying between similar sites.
+Accordingly, $m_0$ is increased for lower $\Delta\theta$ and decreased for higher $\Delta\theta$ , such that there is less straying between dissimilar sites and more straying between similar sites.
 Under these conditions we find that regime II appears for very low $m_0$, and regime I appears for higher $m_0$ (figure \ref{fig:mtheta}), which is opposite the case where $m_0$ and $\Delta\theta$ are independent.
 % $0 < m \leq 0.43$.
-As the straying rate increases and $\Delta\theta$ decreases, a single (symmetric) stable state emerges as the fold bifurcation is crossed, which is opposite the pattern observed when straying is independent of habitat heterogeneity.
+In this case, as straying increases and $\Delta\theta$ decreases, a single (symmetric) steady state emerges as the fold bifurcation is crossed.
+%, which is opposite the pattern observed when straying is independent of habitat heterogeneity.
 
@@
 %Main takehome
-Generally, we find that when the effects of collective behaviour are strong and collective navigation is likely, metapopulation robustness is enhanced.
+Generally, we find that when the effects of collective behaviour are strong such that collective navigation occurs, metapopulation robustness is enhanced.
 However, empirical observations of natural populations suggest that the effects of collective behaviour are intermediate (e.g. $10^{2} \lesssim C \lesssim 10^3$) \citep{Berdahl:2014bl,Berdahl:2016dx}.
@@
 A central dynamic of the model is that straying can lead to the emergence of asymmetric alternative stable states, or \emph{migrational meltdown} \citep{Ronce:2001dp}, pushing one of the populations to a dominant, well-adapted, high density, state, and one to a subordinate, maladapted, low density, state.
-Although there are subtle differences in our model and the general framework presented by Ronce and Kirkpatrick \citep{Ronce:2001dp}, the general dynamic features are the same if we assume that dispersal is symmetric between sites and density-independent (which occurs when $C\rightarrow\infty$).
+Although there are subtle differences in our model and the framework presented by Ronce and Kirkpatrick \citep{Ronce:2001dp}, the general dynamic features are the same if we assume that dispersal is symmetric between sites and density-independent (which occurs when $C\rightarrow\infty$).
 The dynamic regimes that emerge from the eco-evolutionary model -- in particular the occurrence of alternative stable state regimes I and II (figure \ref{fig:traj}) -- have large effects on both the portfolio effect and the recovery time following an induced disturbance (figure \ref{fig:pert}).
@@
 This means that - despite potentially high individual error rates - group formation minimizes straying. 
-This occurs when groups of $\leq 100$ individuals significantly minimizes straying, which is likely unrealistic.
-Moreover, when the effects of collective behaviour are strong, regime II gives way to the dominance of regime I, which harbors low-density basins of attraction (dashed lines in figure \ref{fig:traj}) that can effectively trap disturbed populations in a subordinate stable state, not unlike the Allee effects observed in the collective migration model explored by Berdahl et al.
+This occurs when groups of $\leq 100$ individuals significantly minimize straying, which is likely unrealistic.
+Moreover, when the effects of collective behaviour are strong, regime II gives way to the dominance of regime I, which harbors low-density basins of attraction (figure \ref{fig:traj}).
+The presence of low-density basins of attraction can effectively trap disturbed populations in a subordinate steady state, not unlike the Allee effects observed in the collective migration model explored by Berdahl et al.
 \citep{Berdahl:2016dx}.
@@
 1) there is a magnified difference between the numerical densities of the subordinate and dominant populations, effectively pushing the subordinate population to lower steady state densities (figure \ref{fig:cb}),
-2) the portfolio effect is low, such that the coefficient of variance for the aggregate metapopulation biomass is on-par with the coefficient of variance of its constituent populations, and
+2) the portfolio effect is low, such that the coefficient of variation for the aggregate metapopulation biomass is on-par with the coefficient of variation for its constituent populations, and
 3) more time is generally required for the population(s) to recover following an induced disturbance, and this is particularly true for the recovery of the system following near-collapse of both populations (figure \ref{fig:pert}b).
@@
 This does not appear, in fact, to be a robust condition because the system relies to a large extent on the dominant population as the source, whereas the subordinate population assumes the role of a sink.
-However, recovery time was measured with respect to the aggregate biomass of the metapopulation ($N_T = N_i + N_j$), and despite the source-sink dynamics that emerge in regime II, the aggregate biomass of the system recovers more quickly in this region following the near-collapse of both populations (figure \ref{fig:mpert}b).
+However, recovery time was measured with respect to the aggregate biomass of the metapopulation ($N_T = N_i + N_j$), and despite the source-sink dynamics that emerge in regime II, the aggregate biomass of the system recovers more quickly in this region following the near-collapse of both populations (figure \ref{fig:pert}b).
 From this perspective, the existence of asymmetric dominant/subordinate alternative stable states could be considered to be more robust with respect to the recovery time of the total biomass, or less robust because one population is always at greater risk of stochastic extinction.
@@
 Our surrogate measure for habitat heterogeneity is the difference in trait optima between sites $\Delta\theta$.
-We show that as habitat heterogeneity increases, the occurrence of alternative stable states associated with regime II becomes unavoidable, particularly for $0.1 \leq m_0 \leq 0.4$, and regime I is minimized (figure \ref{fig:bifurcationsddm}).
+We show that as habitat heterogeneity increases, the occurrence of alternative stable states associated with regime II becomes unavoidable, particularly for $0.1 \leq m_0 \leq 0.4$, and regime I is minimized.
 This may be particularly consequential for populations that are spatially adjacent but separated by sharp environmental boundaries, such that trait optima are divergent yet dispersal is relatively high.
@@
 Alternatively, salmon that cue to specific environmental conditions may be more likely to stray into sites that are structurally and physiognamically more similar \citep{Peterson:2014gy}.
-These considerations justify imposing a negative correlation between habitat heterogeneity and individual straying $m_0$: as site heterogeneity increases, so too should straying decrease (figure \ref{fig:mtheta}, inset).
-When habitat heterogeneity and straying are linked in this way, we show that very small amounts of straying gives rise to regime II, and that regime I occurs for higher values of $m_0$ (figure \ref{fig:mtheta}).
+These considerations justify imposing a negative correlation between habitat heterogeneity and individual straying: as site heterogeneity increases, so too should individual straying decrease (figure \ref{fig:mtheta}, inset).
+When habitat heterogeneity and individual straying are linked in this way, we show that very small amounts of individual straying gives rise to regime II, and that regime I occurs for higher values of $m_0$ (figure \ref{fig:mtheta}).
 This pattern is opposite that observed for scenarios where habitat heterogeneity and straying are assumed to be independent, and suggests that increases in straying that are associated with growing similarities between habitats can push a metapopulation into a regime where hidden low-density basins of attraction exist.
@@
 Although robustness is in many cases aided by increasing the strength of collective behaviour, the greater role of both alternative stable state regimes I and II portends additional complexity in eco-evolutionary dynamics, and this could serve to hinder effective management.
-Moreover, this increased complexity at empirically observed levels of straying \citep{H:2013fs} and at realistic (intermediate) ranges of the strength of collective behaviour, is only magnified with increasing habitat heterogeneity and when heterogeneity itself is linked to individual straying.
+Moreover, this increased complexity at empirically observed levels of straying \citep{H:2013fs} and at realistic (intermediate) ranges for the strength of collective behaviour, is only magnified with increasing habitat heterogeneity and when heterogeneity itself is linked to individual straying.
 An additional component that we have not explored here, but that may be particularly relevant to consider, are the effects of including additional sites within the metapopulation network, as well as the patterns of dispersal that connect these sites.
@@
 % While our theoretical work is a vast simplification of natural systems, our results have important potential implications for understanding the eco-evolutionary dynamics of spatially structured populations such as in salmon.
-A particularly salient finding of our work was that density-dependent straying may serve to promote or inhibit population robustness, depending on the strength of the collective behaviour, and the underlying magnitude of straying.
-When the strength of collective behaviour is at plausible intermediate levels, a moderate level of straying increased population robustness by reducing the time to recovery following metapopulation collapse (figure \ref{fig:pert}b).
+A particularly salient finding of our work was that density-dependent straying may serve to promote or inhibit population robustness, depending on the strength of the collective behaviour and the underlying magnitude of straying.
+% A particularly salient finding of our work was that an intermediate strength of collective behaviour and moderate levels of straying increased population robustness by reducing the time to recovery following metapopulation collapse (figure \ref{fig:pert}b).
 Salmon have evolved within the context of dynamic geomorphic landscapes where habitat quantity and quality shifts as a mosaic through time \citep{Stanford:2017bu}. 
@@
 The rate at which individual err, and the influence of group size on navigation at the population level, are two important components of dynamic dispersal \citep{Berdahl:2016dx}.
-We show that changes in these characteristics can alter the occurrence and positioning of two different alternative stable state regimes, one of which may harbor hidden low-density basins of attractions that can effectively trap populations after a large disturbance.
+We show that changes in these characteristics can alter the occurrence and positioning of two different alternative stable state regimes, one of which may harbor hidden low-density basins of attractions that can effectively trap populations after large disturbances.
 Generally, increasing the strength of collective behaviour mitigates the potentially negative impacts of so-called \emph{migrational meltdown} \citep{Ronce:2001dp}.
@@
 \citep{HardestyMoore:wg}.
-We suggest that increased understanding of the proximate and ultimate factors governing dispersal among local populations within metapopulations, across heterogeneous environments, in tandem with the mosaic of selective forces acting on those environments, may be key to promoting persistence in the wild \cite{Westley:uh}.
+We suggest that an increased understanding of the proximate and ultimate factors governing dispersal among local populations within metapopulations, across heterogeneous environments, in tandem with the mosaic of selective forces acting on those environments, may be key to promoting persistence in the wild \cite{Westley:uh}.
 \\ \\
@@
 \noindent {\bf Author contributions:} JDY and JWM conceived of the initial project design.
-JDY and JPG designed the modeling framework and conducted the analyses.
-JDY, JPG, PAHW, and JWM interpreted the results, and drafted and wrote the manuscript.
+JDY, JPG, and TG designed the modeling framework and conducted the analyses.
+All authors contributed to interpreting the results and writing the manuscript.
 \\
@@
 \bibitem{Moore:2010gs}
-Moore JW, McClure M, Rogers LA, Schindler DE, 2010 {Synchronization and
+Moore JW, McClure M, Rogers LA, Schindler DE, 2010 {synchronisation and
   portfolio performance of threatened salmon}.
@@
 Comparison of steady state population densities for the constant straying model and density-dependent straying model.
-Inset: stable state densities for the constant straying model (purple) and density-dependent straying model (green) for different strengths of collective behaviour.
+Inset: steady state densities for the constant straying model (purple) and density-dependent straying model (green) for different strengths of collective behaviour.
 Low $C$ corresponds to strong effects of collective behaviour.
-The top row shows stable state densities as a function of individual straying $m_0$; the bottom row shows stable state densities as a function of straying at the stable state $m(t)^*$.
+The top row shows steady state densities as a function of individual straying $m_0$; the bottom row shows steady state densities as a function of straying at the steady state $m(t)^*$.
 Vertical green lines link paired subordinate and dominant population densities.
-Main: The absolute difference in stable state densities averaged across intervals of low straying ($0 < m,m_0 < 0.25$; blue) and high straying ($0.25 < m,m_0 < 0.5$; red).
+Main: The absolute difference in steady state densities averaged across intervals of low straying ($0 < m,m_0 < 0.25$; blue) and high straying ($0.25 < m,m_0 < 0.5$; red).
+Horizontal dashed lines correspond to the mean absolute differences in steady state densities for low (blue) and high (red) density-independent straying.
+As $C\rightarrow\infty$, mean absolute differences in steady state densities become equivalent.
 } \label{fig:cb}
@@
 Regime II signifies parameter space where there is an asymmetric dominant/subordinate steady state density.
-The white space to the left (lower values of $m_0$) signifies high-density, symmetric, stable states, and the white space to the right (higher values of $m_0$) signifies low-density, symmetric, steady states.
-Relationships are shown for (a) intermediate habitat heterogeneity ($\Delta\theta$), as well as (b) low habitat heterogeneity and (c) high heterogeneity.
+The white space to the left (lower values of $m_0$) signifies high-density, symmetric, steady states, and the white space to the right (higher values of $m_0$) signifies low-density, symmetric, steady states.
+Relationships are shown for (a) low habitat heterogeneity ($\Delta\theta$), (b) intermediate habitat heterogeneity and (c) high heterogeneity.
+The horizontal cutoff of Region I at low values of $C$ in (a) is due to numerical limitations. 
 } \label{fig:bifurcationsddm}
@@
 Example of the numerical procedure used to estimate recovery time.
-After a disturbance is introduced, the recovery time is calculated by measuring the point in time where $N_T$ (in black), which is the aggregate of both populations (blue, red), settles to within one standard deviation of the new equilibrium $N_T^*$.
+After a disturbance is introduced, the recovery time is calculated by measuring the point in time where $N_T$ (black), which is the aggregate of both populations (blue, red), settles to within one standard deviation of the new steady state expectation ${\rm E}(N_T^*)$.
 } \label{fig:recovery}
@@
 \caption{
-Steady-state densities of both populations as a function of $m$: one in a dominant state and one in a subordinate state.
-Steady-states for populations with symmetrical values ($\alpha=0$) in the vital rates $r_{\rm max}$ and $\beta$ are shown with cool tones.
-As the asymmetry among populations between sites increases ($\alpha>0$), their vital rates diverge, such that the maximal growth at sites 1 and 2 is now $r_{\rm max,1}=(1 + \alpha)r_{\rm max,2}$ and $\beta_1=(1+\alpha)\beta_2$ where $\alpha$ is increased from $0$ to $0.1$, thereby increasing the asymmetry in vital rates.
-Steady-states for populations with increasingly asymmetric values ($\alpha\rightarrow 0.1$) for vital rates $r_{\rm max}$ and $\beta$ are shown in warmer tones.
-The lower-density values that appear between $m=0.06$ and $m=0.1$ represent populations that have been trapped in the low-density basins of attraction associated with regime I.
-Increasing the asymmetry in vital rates does not impact the qualitative nature of the dynamics.
+The effects of asymmetrical vital rates on the formation of dominant/subordinate states as a function of straying $m$. 
+(a) Steady state densities for populations with symmetrical values ($\alpha=0$) in vital rates $r_{\rm max}$ and $\beta$ are shown with cool (blue) tones.
+As the asymmetry among populations between sites increases ($\alpha>0$), vital rates diverge, such that the maximal growth at sites 1 and 2 is now $r_{\rm max,1}=(1 + \alpha)r_{\rm max,2}$ and $\beta_1=(1+\alpha)\beta_2$ where $\alpha$ is increased from $0$ to $0.1$.
+Steady-states for populations with increasingly asymmetric values are shown in warmer (red) tones.
+(b) Steady states for populations with higher asymmetric values ($0.1 \leq \alpha \leq 0.4$) for vital rates $r_{\rm max}$ and $\beta$. Note the change in scale compared to panel (a). 
+If asymmetry in vital rates is very high, the pitchfork bifurcation occurs at values of $m<0$.
+We observe that increasing the asymmetry in vital rates -- even to very high levels -- does not impact the qualitative nature of the dynamics.
 } \label{fig:symmetry}
@@
 \caption{
-Comparison of portfolio effects vs. recovery time following the near-collapse of both populations.
-} \label{fig:pevsrt}
+Comparison of portfolio effects vs. recovery time following the near-collapse of both populations where straying is assumed to be density-independent.
+Parameter values used are those reported in table 1, with values of $ 0 < h^2 \leq 0.5$ and for $0 < m \leq 0.5$.} \label{fig:pevsrt}
 \end{figure}

### 2018-01-26 Manuscript/FinalDraft3/supp3.tex
final revision

@@
 \begin{document}
-\title{Supplementary Materials: Eco-evolutionary dynamics, density-dependent dispersal, and collective behavior: implications for salmon metapopulation robustness}
+\title{Supplementary Materials: Eco-evolutionary dynamics, density-dependent dispersal, and collective behaviour: implications for salmon metapopulation robustness}
 % \author{
@@
 \centering
-\includegraphics[width=0.35\textwidth]{fig_density2.pdf}
+\includegraphics[width=0.8\textwidth]{fig_density.pdf}
 \caption{
-Steady-state densities of both populations as a function of $m$: one in a dominant state and one in a subordinate state.
-Steady-states for populations with symmetrical values ($\alpha=0$) in the vital rates $r_{\rm max}$ and $\beta$ are shown with cool tones.
-As the asymmetry among populations between sites increases ($\alpha>0$), their vital rates diverge, such that the maximal growth at sites 1 and 2 is now $r_{\rm max,1}=(1 + \alpha)r_{\rm max,2}$ and $\beta_1=(1+\alpha)\beta_2$ where $\alpha$ is increased from $0$ to $0.1$, thereby increasing the asymmetry in vital rates.
-Steady-states for populations with increasingly asymmetric values ($\alpha\rightarrow 0.1$) for vital rates $r_{\rm max}$ and $\beta$ are shown in warmer tones.
-The lower-density values that appear between $m=0.06$ and $m=0.1$ represent populations that have been trapped in the low-density basins of attraction associated with regime I.
-Increasing the asymmetry in vital rates does not impact the qualitative nature of the dynamics.
+The effects of asymmetrical vital rates on the formation of dominant/subordinate states as a function of straying $m$. 
+(a) Steady state densities for populations with symmetrical values ($\alpha=0$) in vital rates $r_{\rm max}$ and $\beta$ are shown with cool (blue) tones.
+As the asymmetry among populations between sites increases ($\alpha>0$), vital rates diverge, such that the maximal growth at sites 1 and 2 is now $r_{\rm max,1}=(1 + \alpha)r_{\rm max,2}$ and $\beta_1=(1+\alpha)\beta_2$ where $\alpha$ is increased from $0$ to $0.1$.
+Steady-states for populations with increasingly asymmetric values are shown in warmer (red) tones.
+(b) Steady states for populations with higher asymmetric values ($0.1 \leq \alpha \leq 0.4$) for vital rates $r_{\rm max}$ and $\beta$. Note the change in scale compared to panel (a). 
+If asymmetry in vital rates is very high, the pitchfork bifurcation occurs at values of $m<0$.
+We observe that increasing the asymmetry in vital rates -- even to very high levels -- does not impact the qualitative nature of the dynamics.
 } \label{fig:symmetry}
@@
 \caption{
-Comparison of portfolio effects vs. recovery time following the near-collapse of both populations.
+Comparison of portfolio effects vs. recovery time following the near-collapse of both populations where straying is assumed to be density-independent.
+Parameter values used are those reported in table 1, with values of $ 0 < h^2 \leq 0.5$ and for $0 < m \leq 0.5$.
 } \label{fig:pevsrt}

### 2018-01-26 model/model2/Figure_4regimes_ddm.jl

@@ -185,8 +185,10 @@ boxplot(L)
 """
 
 @everywhere using Distributions, RCall, JLD, HDF5
-@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolveSS.jl")
-@everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolveSS_ddm.jl")
+# @everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolveSS.jl")
+# @everywhere include("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/src/KevinEvolveSS_ddm.jl")
+@everywhere include("$(homedir())/src/KevinEvolveSS.jl")
+@everywhere include("$(homedir())/src/KevinEvolveSS_ddm.jl")
 
 #Analysis over m and C
 tmax=10000;
@@ -283,8 +285,8 @@ burnin=0.80
       aggmean[i] = mean(n1trim+n2trim);
       aggsd[i] = std(n1trim+n2trim);
 
-      x1mean[i] = theta1-mean(x1trim);
-      x2mean[i] = (theta1+thetadiff)-mean(x2trim);
+      x1mean[i] = mean(x1trim);
+      x2mean[i] = mean(x2trim);
       
 
         
@@ -347,10 +349,17 @@ burnin=0.80
   
 end
 
-save(string("$(homedir())/Dropbox/PostDoc/2017_SalmonStrays/model/data3/data_meandiff.jld"),"n1mean",n1mean,"n2mean",n2mean,"x1mean",x1mean,"x2mean",x2mean,"n1mean_ddm",n1mean_ddm,"n2mean_ddm",n2mean_ddm,"x1mean_ddm",x1mean_ddm,"x2mean_ddm",x2mean_ddm,"m1mean_ddm",m1mean_ddm,"m2mean_ddm",m2mean_ddm,"pe_ddm",pe_ddm);
[…]
